### 2020-11-16 Analysis.Rmd
initial commit

---
title             : "National Longitudinal Evidence for Growth in Well-being From Spiritual Beliefs"
shorttitle        : "Title"
author: 
  - name          : "Ben Highland"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Postal address"
    email         : "my@email.com"
    role:         # First author
      - Conceptualization
      - Writing - Original Draft Preparation
      - Writing - Review & editing
  - name          : "Joseph Bulbulia"
    affiliation   : "1,2"
    role:
      - Writing - 
      - Analysis 
affiliation:
  - id            : "1"
    institution   : "Wake Forest Med School "
  - id            : "2"
    institution   : "Victoria University"
authornote: |
  Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.
  Enter author note here.
abstract: |
 A growing body of research finds an association between spirituality and subjective well-being. However, the use of tautological scales and a reliance on cross-sectional data compromise inference. Here, drawing on nine years of panel responses from a nationally diverse longitudinal study. We systematically test whether having spiritual beliefs improves subjective well-being (N = 21,705, New Zealand, 2010-2018). Supporting previous studies, we observe that belief in a spirit or life force predicts increasing life satisfaction over time. However, we also observe that relative to the disbelieving population, belief in a spirit or life force is initially associated with lower personal well-being. Such effects, which hold, after adjusting for known demographic and personal influences on subjective well-being, are consistent with findings that spiritual beliefs are compelling to those with lower well-being. However, despite the scale and resolution of these longitudinal data, the causal interplay between spiritual beliefs and subjective well-being remain matters for future investigations.

  <!-- https://tinyurl.com/ybremelq -->
keywords          : "keywords"
wordcount         : "X"
bibliography      : ["r-references.bib"]
floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
documentclass     : "apa6"
classoption       : "man"
output            : "html_document" 
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
library("papaja")
r_refs("r-references.bib")
```

```{r global_options, echo=FALSE, include = FALSE}
knitr::opts_chunk$set(fig.path="Figs/", 
                      message=FALSE, 
                      warning=FALSE,
                      collapse =TRUE,
                      echo=TRUE, 
                      #results="hide", 
                      fig.width= 10,
                      fig.height=8,
                      tidy = "styler")
```
```{r load_libraries, echo=FALSE, include=FALSE}
library("styler")
library("tidyverse")
library("lme4")
library("ggplot2")
library("tidyr")
library("ggeffects")
library("Amelia")
library("table1")
library("patchwork")
library("parameters")
library("prettycode")
library("table1")
library("gghighlight")
library("sjPlot")
library("see")
#library("viridis")
#library("ggpubr")
#library("wesanderson")
library("ggsci")
library("papaja")
#library("here")
library("citr")
library("texreg")
```

# Methods
 <!-- 21-word solution (Simmons, Nelson & Simonsohn, 2012; retrieved from http://ssrn.com/abstract=2160588) -->
We used `r cite_r("r-references.bib")` for all our analyses.

Load data (this is sepcific to my machine)
```{r load_data}
ldf.5<-readRDS("~/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/00Bulbulia\ Pubs/2020/ldf.5") 
```

Processing of data.
```{r data processing, cache = TRUE, dependson = "ldf.5"}
# filter only those who are measured 
dfben  <-ldf.5 %>%
  dplyr:: filter(Wave !=2009)%>% # exclude wave 1 (no spirit measures that wave)
  dplyr::filter(YearMeasured==1)%>%
  dplyr::group_by(Id) %>% filter(n() > 2)%>% # select those who have responded to at least 3 waves # Also works for 5 and 7
  dplyr::filter(n() !=0)%>%
  dplyr::ungroup(Id)%>%
  dplyr:: mutate(Years = as.numeric(years))%>%
  dplyr::mutate(Age.10yrs = (Age/10))%>%
  dplyr::mutate(Religion.Church.Log = log(Religion.Church+ 1))%>%
  dplyr::mutate(Pol.Orient.S = scale(Pol.Orient, center = TRUE, scale = TRUE),
         Age.10yrs.C = scale(Age.10yrs, center = TRUE, scale = FALSE),
         Male = factor(Gender),
         Employed = factor(Employed),
         Partner = factor(Partner),
         Edu.S = scale(as.numeric(Edu) ,scale=TRUE,center=TRUE),
         Education = as.numeric(Edu),
         Urban = factor(Urban),
         Believe.Spirit = as.factor(Believe.Spirit),
         Believe.God = as.factor(Believe.God),
         EthnicCategories = factor(EthnicCats),
         LIFESAT.S = scale(LIFESAT, center = TRUE, scale = TRUE),
         PWI.S = scale(PWI, center = TRUE, scale = TRUE),
         Deprivation.S = scale(NZdep, scale=TRUE, center=TRUE),
         Relid.C = scale(Relid, scale = TRUE,center=TRUE),
         Religious = as.factor(Religious),
         Religion.Church.Log.C = scale(Religion.Church.Log, center = TRUE, scale = FALSE))
```
```{r}
# nice labels
table1::label(dfben$Age.10yrs.C) <- "Age in Decades (C)"
table1::label(dfben$PWI) <- "Personal Wellbeing"
table1::label(dfben$LIFESAT) <- "Life Satisfaction"

table1::label(dfben$Deprivation.S) <- "Deprivation (S)"
table1::label(dfben$Pol.Orient) <- "Political Conservative"
table1::label(dfben$Pol.Orient.S) <- "Political Conservative (S)"
table1::label(dfben$NZdep) <- "NZ Deprivation Index"
dfben$Partner <- factor(dfben$Partner, labels = c("No Partner","Has Partner"))
dfben$Employed <- factor(dfben$Employed, labels = c("Not Employed","Employed"))
table1::label(dfben$Edu.S) <- "Education (S)"

# beliefs indicator
dfben$Beliefs<- factor(ifelse(dfben$Believe.God == "Not Believe God" & dfben$Believe.Spirit == "Not Believe Spirit", "_Skeptic_", 
                                        ifelse(dfben$Believe.God == "Not Believe God" & dfben$Believe.Spirit == "Believe Spirit", "_SpiritExcludesGod_",
                                               ifelse(dfben$Believe.God == "Believe God" & dfben$Believe.Spirit == "Believe Spirit","GodAndSpirit","GodExcludesSpirit"))))
```

Table of belief types
```{r}
table(dfben$Beliefs)
```

```{r echo=FALSE}
# nice labels
#dfben$GodSpiritBeliefs <- factor(dfben$GodSpiritBeliefs, labels = c("Skeptic","SpiritExcludesGod","GodAndSpirit","GodExcludesSpirit"))
# check to make sure labels are right
#table(dfben$GodSpiritBeliefs) == table(dfben$GodSpiritBeliefs)
```

There are `r  length(unique(dfben$Id))` participants.

This is the number of times participants returned responses: 
```{r}
dplyr::count(tally(group_by(dfben, Id), sort = TRUE, name="number_waves"),number_waves)
```

Table.

```{r}
table1::table1(~  Age + NZdep + Education + Employed + EthnicCategories + Male  + Partner  + Pol.Orient + Urban +  Beliefs + LIFESAT + PWI|Wave, data = dfben, overall = F)
```

```{r, include =FALSE}
furniture::table1(dfben, Age, NZdep,Education,Employed,EthnicCategories,Male,Partner, Pol.Orient,Urban, Beliefs, LIFESAT,PWI,
                    splitby = ~Wave, overall=F, output = "latex2", booktabs=TRUE)
```

```{r, include=FALSE}
r<-report::report_participants(
  dfben,
  age = "Age",
  sex = "Male",
  education = "Education",
  participants = "Id",
  group = "Wave",
  spell_n = FALSE,
)
```

We demean variables following https://easystats.github.io/parameters/articles/demean.html

```{r}
dfben <- cbind(
  dfben,
  parameters::demean(dfben, select = c("Beliefs"), group = "Id")
)
```

# Results

```{r models, cache = TRUE, dependson = "dfben"}
### Spirit Belief Models ### 
mod01<-lmer (LIFESAT ~ Years *(
                Beliefs__Skeptic__between  +  
                Beliefs__SpiritExcludesGod__between + 
                Beliefs_GodAndSpirit_between  + 
                Beliefs_GodExcludesSpirit_between +
                Beliefs__Skeptic__within  +  
                Beliefs__SpiritExcludesGod__within + 
                Beliefs_GodAndSpirit_within  + 
                Beliefs_GodExcludesSpirit_within) +
                (1|Id), data= dfben, na.action = na.omit)

parameters::check_heterogeneity(mod01, group = "ID")

mod02<-lmer (LIFESAT ~ Years *( 
                Beliefs__Skeptic__between  +  
                Beliefs__SpiritExcludesGod__between + 
                Beliefs_GodAndSpirit_between  + 
                Beliefs_GodExcludesSpirit_between +
                Beliefs__Skeptic__within  +  
                Beliefs__SpiritExcludesGod__within + 
                Beliefs_GodAndSpirit_within  + 
                Beliefs_GodExcludesSpirit_within) +
                (1|Id), data= dfben, na.action = na.omit)

parameters::check_heterogeneity(mod02, group = "ID")

# inference holds, but can we use this method with factors?  In any case, we keep to the simple model
tab_model(mod01)
tab_model(mod02)

# Inference holds  FOCe INTERCEEpTS TO BE 
mod.1 <-lmer (PWI ~  1 +  Years * Beliefs +  Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id), data= dfben, na.action = na.omit)
mod.2 <-lmer (LIFESAT ~ 1 + Years * Beliefs+ Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id), data= dfben, na.action = na.omit)
summary(mod.1)
summary(mod.2)

## Try models with generation as rand effect. Reason: how to interpret differences in intercepts?  # USING glmmTMB because lmer is crashing SAME RESULT SO IGNORE
# mod.1 <-glmmTMB(LIFESAT ~ Years * Beliefs + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id) + (1|GenCohort), data= dfben)
# 
# mod.2 <-glmmTMB(PWI ~ Years * Beliefs + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id) + (1|GenCohort), data= dfben)
```

Tables

```{r cache = TRUE, dependson="mod.2", dependson="mod.1"}
tab_model(mod.2, mod.1) # Table
```

```{r, echo = FALSE, include=FALSE}
te00 <-texreg::extract(
  mod.1,
  level = 0.95,
  include.random = TRUE,
  include.rsquared = T,
  include.nobs = T,
  include.loo.ic = T,
  include.waic = F)

texreg(list(te00),
                custom.model.names = c("PWI"),
                caption = "Personal Wellbeing",
                sideways = F,
                scalebox = .5,
                #fontsize= "footnotesize",
                label = "tab:REGRESS_1",
                ci.force.level = 0.95, bold = 0.05,
                settingstars = 0,
                booktabs = TRUE,
                custom.note ="")
```

```{r, echo = FALSE, include=FALSE}
te02 <-texreg::extract(
  mod.2,
  level = 0.95,
  include.random = TRUE,
  include.rsquared = T,
  include.nobs = T,
  include.loo.ic = T,
  include.waic = F)

texreg(list(te02),
                custom.model.names = c("PWI"),
                caption = "LifeSatisfaction",
                sideways = F,
                scalebox = .5,
                #fontsize= "footnotesize",
                label = "tab:REGRESS_1",
                ci.force.level = 0.95, bold = 0.05,
                settingstars = 0,
                booktabs = TRUE,
                custom.note ="")
texreg(list(te02),
                custom.model.names = c("PWI"),
                caption = "LifeSatisfaction",
                sideways = F,
                scalebox = .5,
                #fontsize= "footnotesize",
                label = "tab:REGRESS_1",
                ci.force.level = 0.95, bold = 0.05,
                settingstars = 0,
                booktabs = TRUE,
                custom.note ="")

texreg(list(te00, te02), dcolumn = F, booktabs = TRUE, scalebox = .6,ci.force.level = 0.95, bold = 0.05,
                settingstars = 0, caption = "Regression Models")
```

```{r, echo = FALSE, include=FALSE}
te00 <-texreg::extract(
  mod.1,
  level = 0.95,
  include.random = TRUE,
  include.rsquared = T,
  include.nobs = T,
  include.loo.ic = T,
  include.waic = F)
texreg(list(te00),
                custom.model.names = c("PWI"),
                caption = "Personal Wellbeing",
                sideways = F,
                scalebox = .5,
                #fontsize= "footnotesize",
                label = "tab:REGRESS_1",
                ci.force.level = 0.95, bold = 0.05,
                settingstars = 0,
                booktabs = TRUE,
                custom.note ="")
```

Plot models
```{r USEcoefficientplots}
plot_models(mod.2, mod.1) + ylim(-.4, .75) + theme_abyss() + scale_colour_brewer(labels = c("Personal Wellbeing", "Life Satisfaction")) # Plots
```

```{r cache = TRUE, dependson = "mod.1"}
#### Predicted Effects ####
pr.PWI <- ggpredict(model = mod.1, terms = c("Years [0:9]","Beliefs"), 
                    ci.lvl = 0.95,
                    type = "fe",
                    typical = "mean",
                   # condition = c(Euro =1),
                    back.transform = TRUE,
                    ppd = FALSE,
                    interval = "confidence")
p1<-plot(pr.PWI, facets = T) + 
 # scale_y_continuous(limits=c(4.75,5.25) )  + 
  gghighlight()  +  theme_blank()+ 
  ggtitle("Predicted Values of Personal Wellbeing") 
```
```{r pwi_expected}
```

```{r}
pr.LIFESAT <- ggpredict(model = mod.2, terms = c("Years [0:9]", "Beliefs"),
                        ci.lvl = 0.95,
                        type = "fe",
                        typical = "mean",
                     #   condition = c(Euro =1),
                        back.transform = TRUE,
                        ppd = FALSE,
                        interval = "confidence")
p2<-plot(pr.LIFESAT,facets = T) + 
 # scale_y_continuous(limits=c(6.35,6.85) ) + 
  gghighlight() + theme_blank() + 
  ggtitle("Predicted Values of Life Satisfaction") #+   scale_colour_brewer(palette = "Zissou1")
```

## Reporting

### equations
Lifesatisfaction Model 1 equation

```{r}
library(equatiomatic)
print(extract_eq(mod.1, wrap = TRUE,terms_per_line =3))
```

PWI Model 2 equation
```{r}
library(equatiomatic)
print(extract_eq(mod.2, wrap = TRUE, use_coefs=TRUE,terms_per_line = 3))
```

## report

### Model 1
model 1 report: for publication change all "significants" to "reliable"

short report:
```{r, cache =TRUE, dependson=mod.1,include=FALSE}
report::model_text(mod.1)
```

short report
```{r, cache=TRUE, include=FALSE }
library(report)
r1<- report::report(mod.1)
table_short(r1)
text_short(r1)
```

```{r, cache =TRUE, dependson=mod.1, include=FALSE}
r1<- report::report(mod.1)
report::text_long(r1)
```

### Model 2
For publication change all "significants" to "reliable"

short report
```{r, cache =TRUE, dependson=mod.2,include=FALSE}
report::model_text(mod.2)
```

short report
```{r,include=FALSE}
r2 <- report::report(mod.2)
report::text_short(r2)
report::table_short(r2)
```

long report
```{r,include=FALSE}
r2<- report::report(mod.2)
report::text_long(r2)
```

Graphs of predictions.

```{r USEPWI_expected}
p1 #+ plot_annotation(title = 'All cases Plot')
```

```{r USElifsat_expected}
p2 #+ plot_annotation(title = 'All cases Plot')
```

## Imputed Analysis

```{r echo = TRUE,  cache = TRUE, dependson = "dfben"}
# Imputed data
prp<- dfben %>% # Remove what we do not need anymore
  dplyr::select(c(Beliefs, Age, Id, Education, NZdep,KESSLER6,LIFESAT,PWI,Pol.Orient,Relid,EthnicCats,Partner,Employed,Male,Urban,Years,Wave))

# total ids = 21,705
length(unique(prp$Id))

#Make dataframes
prp<-as.data.frame(prp)

#Bifurcated imputation - This is presented in the main text
set.seed(1234)
imputed <- amelia(
  prp, #dataset to impute
  m = 10, # number of imputations
  cs= c("Id"),
  ts= c("Years"),
  noms = c("EthnicCats",
           "Urban",
           "Partner",
           "Male",
           "Employed",
           "Beliefs"),
  idvars=c("Wave","PWI","LIFESAT"), # not imputing outcomes
  polytime = 3) #https://stackoverflow.com/questions/56218702/missing-data-warning-r
#saveRDS(imputed,"imputed")
#imputed<- readRDS("imputed")

# center and scale indiators
imputed.2 <- transform.amelia(imputed,
                              Age.10yrs = (Age/10),
                              Pol.Orient.S =scale(Pol.Orient, center=TRUE,scale=TRUE),
                              Employed = factor(Employed),
                              Ethnicity = as.factor(EthnicCats),
                              Urban = factor(Urban),
                              Deprivation.S = scale(NZdep, scale=TRUE, center=TRUE),
                              Edu.S = scale(Education, scale =TRUE,center=TRUE),
                              Male = as.factor(Male),
                              PWI = as.numeric(PWI),
                              LIFESAT = as.numeric(LIFESAT),
                              Beliefs = as.factor(Beliefs),
                              Id =as.factor(Id))
# center an d scale age
imputed3 <- transform.amelia(imputed.2,
                                     Age.10yrs.C = scale(Age.10yrs,scale =FALSE, center=TRUE))
```

```{r cache = TRUE}
# run models iterating over imputed data
m <- 10
models3 <- NULL
for(i in 1:m) {
  models3[[i]] <- lmer(PWI ~ Years * Beliefs + Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id), data=imputed3$imputations[[i]])
}
```

```{r, cache=TRUE, dependson= models3}
#devtools::install_github("easystats/parameters") Get development version
library(parameters)
mps3 <- lapply(models3, model_parameters)
parameters::pool_parameters(mps3)
mp3<-pool_parameters(mps3)
```

```{r coef_PWI_imputed}
# coeff plot
plot(mp3) + ggtitle("Personal Wellbeing Coefficients (Imputed)") # pick your  own title
```

```{r cache = TRUE, dependson = models3}
## expectation
m<-10
out<-NULL
for(i in 1:m) {
  out[[i]] <- ggpredict(models3[[i]], terms =c("Years [0:9]","Beliefs"))
}

m<-10
plots<-NULL
for(i in 1:m) {
  plots[[i]] <- plot(out[[i]], facets = T)# + scale_y_continuous(limits=c(6.35,6.85) )
}
```
```{r, echo=FALSE}
# Inspect all  # No differences
# plots[[1]]
# plots[[2]]
# plots[[3]]
# plots[[4]]
# plots[[5]]
# plots[[6]]
# plots[[7]]
# plots[[8]]
# plots[[9]]
# plots[[10]]
```
```{r predicted_PWI_impute}
# to highlight differencs
explectationGraphPWI <-plots[[10]] + gghighlight() + ggtitle("Predicted Values of Personal Wellbeing (imputed)")
explectationGraphPWI
```
```{r cache =TRUE}
## life sat models
m <- 10
modelsL3 <- NULL

for(i in 1:m) {
  modelsL3[[i]] <- lmer(LIFESAT ~ Years * Beliefs + Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban +  (1|Id), data=imputed3$imputations[[i]])
}
```
```{r, cache =TRUE, dependson =  modelsL3}
mpl3 <- lapply(modelsL3, model_parameters)
parameters::pool_parameters(mpl3)
ml3<-pool_parameters(mpl3)
ml3 # table
p13<-plot(ml3)

p13<- p13 +  labs(title ="Coefficients Life Satisafaction (imputed)")
# get estimates for PWI from above
p23 <- plot(mp3) + labs(title ="Coefficients Personal Wellbeing (imputed)")
```

```{r coefficient_plot_both_imputed}
 p23/p13 +  plot_annotation(tag_levels = 'i')
```

```{r, cache =TRUE, dependson = "modelsL3"}
# Expectation plots
m<-10
out2<-NULL
for(i in 1:m) {
  out2[[i]] <- ggpredict(modelsL3[[i]], terms = c("Years [0:9]", "Beliefs"),
                         ci.lvl = 0.95,
                         type = "fe",
                         typical = "mean",
                         #   condition = c(Euro =1),
                         back.transform = TRUE,
                         ppd = FALSE,
                         interval = "confidence")
}

m<-10
plots2<-NULL
for(i in 1:m) {
  plots2[[i]] <- plot(out2[[i]], facets = T) #+ scale_y_continuous(limits=c(4.75,5.25) )
}
```
```{r echo = FALSE}
# Inspect all  # No differences
# plots2[[1]]
# plots2[[2]]
# plots2[[3]]
# plots2[[4]]
# plots2[[5]]
# plots2[[6]]
# plots2[[7]]
# plots2[[8]]
# plots2[[9]]
# plots2[[10]]

library(gghighlight)

# Just using one graph
explectationGraphLS <-plots2[[10]] + gghighlight() + ggtitle("Predicted Values of Life Satisfaction (imputed)")
```

```{r predicted_lifesat_imputed, cache=TRUE}
explectationGraphLS
```
```{r combograph_imputed}
# Combo_Graph
#explectationGraphPWI/explectationGraphLS + plot_annotation(title = "Imputed DataSet")
##
sessionInfo()
```

# Discussion
To be continued.

\newpage

# References

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

<div id="refs" custom-style="Bibliography"></div>
\endgroup

### 2020-11-16 Manuscript.Rmd
initial commit

---
title             : "National Longitudinal Study Reveals Growth in Wellbeing Among Those with Spiritual Beliefs"
shorttitle        : "Beliefs and Wellbeing"
author: 
  - name          : "Ben Highland"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Postal address"
    email         : "my@email.com"
    role:         # First author
      - Conceptualization
      - Literature Review
      - Writing - First draft
      - Writing - Review & editing
  - name          : "Ev l Worthing"
    affiliation   : "2"
    address       : "Postal address"
    email         : "my@email.com"
    role:        
      - Writing review & editing
  - name          : "Don E Davis"
    affiliation   : "3"
    address       : "Postal address"
    email         : "my@email.com"
    role:         
      - Writing review & editing
  - name          : "Chris G. Sibley"
    affiliation   : "4"
    address       : "Postal address"
    email         : "my@email.com"
    role:        
      - Writing review & editing
  - name          : "Joseph Bulbulia"
    affiliation   : "5"
    role:         
      - Data analysis
      - Writing second and subsequent drafts
      - Writing review & editing

authornote: |
  Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.
  Enter author note here.
abstract: |
 Previous research finds an association between spirituality and subjective well-being. However, the widespread use of tautological scales and the reliance on cross-sectional data cast doubts. Here, we leverage nine years of panel response data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction ($N$ =  20979, New Zealand, 2010-2018). Contrary to previous studies, we find that belief in a spirit or life force predicts initially lower personal wellbeing and life-satisfaction. However, in support of theories that sprituality improves psychological health, beliefs in a spirit or life force predict increasing life satisfaction over time relative to disbelief. These finding holds after adjusting for known demographic influences. Intriguingly these findings also hold among those who believe in a God but disbelieve in a spirit or life force. The causal mechanisms by which beliefs in a spirit or life force increases well-being remain an important horizon for future investigations.

  <!-- https://tinyurl.com/ybremelq -->
keywords          : "Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being"
wordcount         : no
bibliography      : ["b-references.bib"]
floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : no
mask              : no
draft             : no
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
extra_dependencies: ["rotating", "booktabs", "graphicx","longtable"]
header-includes:
   - \usepackage{tcolorbox}
   - \usepackage{paralist}
   - \usepackage{multicol}
---

```{r include = FALSE}
library(knitr)
opts_chunk$set(
    fig.align = "center",
    root.dir = "../"
)
library("papaja")
library("tidyverse")
library("tinytex")
```

```{r load_libraries, echo=FALSE, include=FALSE}
# load libraries
library("styler")
library("tidyverse")
library("lme4")
library("ggplot2")
library("tidyr")
library("sjPlot")
library("ggeffects")
library("Amelia")
library("table1")
library("patchwork")
library("parameters")
library("prettycode")
library("gghighlight")
library("sjPlot")
library("see")
library("viridis")
library("ggpubr")
library("wesanderson")
library("ggsci")
library("here")
library("citr")
#library("furniture")
library("knitr")
library("texreg")
```
```{r load_data, include=FALSE}
ldf.5<-readRDS("~/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/00Bulbulia\ Pubs/2020/ldf.5") 
```
```{r data processing, cache = TRUE, dependson = "ldf.5"}
# filter only those who are measured 
dfben  <-ldf.5 %>%
  dplyr:: filter(Wave !=2009)%>% # exclude wave 1 (no spirit measures that wave)
  dplyr::filter(YearMeasured==1)%>%
  dplyr::group_by(Id) %>% filter(n() > 2)%>% # select those who have responded to at least 3 waves # Also works for 5 and 7
  dplyr::filter(n() !=0)%>%
  dplyr::ungroup(Id)%>%
  dplyr:: mutate(Years = as.numeric(years))%>%
  dplyr::mutate(Age.10yrs = (Age/10))%>%
  dplyr::mutate(Religion.Church.Log = log(Religion.Church+ 1))%>%
  dplyr::mutate(Pol.Orient.S = scale(Pol.Orient, center = TRUE, scale = TRUE),
         Age.10yrs.C = scale(Age.10yrs, center = TRUE, scale = FALSE),
         Male = factor(Gender),
         Employed = factor(Employed),
         Partner = factor(Partner),
         Edu.S = scale(as.numeric(Edu) ,scale=TRUE,center=TRUE),
         Education = as.numeric(Edu),
         Urban = factor(Urban),
         Believe.Spirit = as.factor(Believe.Spirit),
         Believe.God = as.factor(Believe.God),
         EthnicCategories = factor(EthnicCats),
         LIFESAT.S = scale(LIFESAT, center = TRUE, scale = TRUE),
         PWI.S = scale(PWI, center = TRUE, scale = TRUE),
         Deprivation.S = scale(NZdep, scale=TRUE, center=TRUE),
         Relid.C = scale(Relid, scale = TRUE,center=TRUE),
         Religious = as.factor(Religious),
         Religion.Church.Log.C = scale(Religion.Church.Log, center = TRUE, scale = FALSE))
```
```{r}
# nice labels
table1::label(dfben$Age.10yrs.C) <- "Age in Decades (C)"
table1::label(dfben$PWI) <- "Personal Wellbeing"
table1::label(dfben$LIFESAT) <- "Life Satisfaction"

table1::label(dfben$Deprivation.S) <- "Deprivation (S)"
table1::label(dfben$Pol.Orient) <- "Political Conservative"
table1::label(dfben$Pol.Orient.S) <- "Political Conservative (S)"
table1::label(dfben$NZdep) <- "NZ Deprivation Index"
dfben$Partner <- factor(dfben$Partner, labels = c("No Partner","Has Partner"))
dfben$Employed <- factor(dfben$Employed, labels = c("Not Employed","Employed"))
table1::label(dfben$Edu.S) <- "Education (S)"

# create beliefs indicator
dfben$Beliefs<- factor(ifelse(dfben$Believe.God == "Not Believe God" & dfben$Believe.Spirit == "Not Believe Spirit", "_Skeptic_", 
                                        ifelse(dfben$Believe.God == "Not Believe God" & dfben$Believe.Spirit == "Believe Spirit", "_SpiritExcludesGod_",
                                               ifelse(dfben$Believe.God == "Believe God" & dfben$Believe.Spirit == "Believe Spirit","GodAndSpirit","GodExcludesSpirit"))))
```

The relationship between spirituality and psychological well-being is a matter of enduring human fascination [@Ellsworth2010-yu]. Previous attempts to quantify the relationship between these two domains find a positiven association between spirituality and well-being [@Russinova2002-rq; @Ginsburg1995-jr; @Zaza2005-ac; @Koenig2010-gk; @Smith2003-re; @Smith2003-re; @Koenig2001-ow]. However, critics have raised three challenges.

<!-- In spirituality/psychological wellbeing research,sSsubjective in spirituality health research, psychological well-being is typically decomposed into two categories: emotional well-being (i.e., the presence or absence of positive feelings about life) and functional well-being (i.e., psychological and social well-being) [@Keyes2002-bb]. -->

First, previous research has frequently operationalised spirituality using measures of well-being. For example, The Daily Spiritual Experience Scale, which is widely used in spiritual mental health research includes items such as "I feel thankful for my blessings" and "I feel deep inner peace or harmony" to measure spirituality [@Underwood2002-hg]. As Koenig observes, it is unsurprising that well-being predicts well-being –. what Koenig describes as a  "tautology" in spirituality/mental health research [@Koenig2008-lv]. Indeed, recent systematic-review of spirituality/mental health research found that nearly 45% of previous studies investigating spirituality and emotional well-being operationalise spirituality scales using constructs for positive emotional well-being [@Garssen2016-kb]. The first challenge, then, is to define measures of spirituality that are not well-being measures reclocked in other words.

How, then, should researchers operationalise spirituality. We can look to folk traditions for guidance. The term "spirituality" was first used to describe the religious practices of ascetics and monks. more recently, however, the term has taken on a family of broader meaning [@Swinton2001-vr]. Religion is typically operationalized to mean beliefs and practices respecting a God or gods (JB). A recurring element in the use of ‘spirituality,’ as it is distinguished from religion, is belief in a spirit or life force without commitment to a belief in a God or gods [@Hill2000-cl; @Davis2015-tx]. However, the term "spirituality" overlaps with the uses of "religion" in the sense that God or gods are instances of beliefs in spirits – that is one may be both religious and spiritual. Here we operationalize religious belief as either belief in a God or belief in both a God and a spirit or life force. We operationalize spiritual belief as either belief in a spirit or life force without any corresponding belief in a God. However, the term "spirituality" overlaps with the uses of "religion" in the sense that God or gods are instances of beliefs in spirits – that is one may be both religious and spiritual. Here we operationalize religious belief as either belief in a God or belief in both a God and a spirit or life force.  isNotably, previous spirituality mental health research has been disproportionately conducted on North American samples. In the United States, especially, samples include a majority of people who believe in both a God as well as and life forces other than God [@Smith2003-re; @Koenig2010-gk]. Although such studies drawing from such samples mightmay clarify the effects of religious spirituality on psychological subjective well-being, the question of how non-religious spirituality affects psychological well-being that is – or spiritual beliefs without any belief in a God or gods – affects subjective well-being remain unclear.  (For discussion, see: Therefore, to better understand possible distinctions between religious spirituality and non-religious spirituality for subjective well-being outcomes, samples collected from cultures outside the North Atlantic that contain a more equal ratio of the two are needed [@Garssen2016-km]).

Third, most research investigating the relationship between spirituality and subjective well-being has relied on cross-sectional samples. These studies include high-quality reviews and meta-analyses [@Ano2005-hx; @Hackney2003-rs; @Sawatzky2005-rw; @Smith2003-re; @Visser2010-kq; @Yonker2012-zg]. However, to assess whetherto whether spiritual beliefs lead to a growth, stability, or decline in psychological well-being relative to people who lack such beliefs , including people who believe in God but not a spirit or lifeforce, requires repeatedly measuring the same individuals over time. clarify causal effects requires measuring the same individuals over time and assessing how life turns out for demographically similar people who differ in their spiritual and religious beliefs over their lifespans  and orientations [@Garssen2016-km].

Our study addresses these three limitations in previous research.improves on past methods the following ways: First, we avoid a tautological elision of "spirituality" and "well-being"y by adopting a non-affective measure of spirituality through use of general, non-affective measures: "Do you believe in some form of a spirit or life force?" Our measures of psychological well being are [BEN SAY HERE] and Second we adopt a four-level categorical indicator that disentangles spiritual and religious beliefs: (1) Skeptics: those who neither believe in a God nor a life force; (2) SpiritExcludeGod: those who no not believe in a God  but believe in a spirit or life force; (3) GodAndSpirit: those who believe in a God and spirit; (4) GodExcludesSpirit: those who believe in both a God and a spirit or life force "Do you believe in a god?". Second, we address the prior dependence on cross-sectional samples taken from the United States by analyzing data taken from By sampling from respondents in New Zealand, a country with [GIVE NON_RELIGIOUS SPIRITUAL STATS HERE).  Third, we are able to assess whether spirit beliefs are associate with a growth in psychological wellbeing address causation by weusing we use longitudinal responses from athis nationally diverse probability sample year over an nine-yearseight-year period with nine measurement points  to assess whether spirit beliefs lead to growth, stability, or decline in psychological well-beingwellbeing (2010 to 2018 New Zealand Attitudes and Values Study, NZAVS).

# Method

<!-- We rigorously assess the relationship between our non-affective measures and two constructs of subjective well-being (personal well-being and life satisfaction), adjusting for known demographic and personal influences, including age, sex, deprivation, and political orientation (Okulicz-Kozaryn, Holmes, and Avery 2014; Cuesta and Budría 2012; Røysamb et al. 2002; Hansen and Slagsvold 2012; Chen and van Ours, n.d.; Vaughan et al. 1985; Easterlin, Angelescu, and Zweig 2011; Stam et al. 2016; Yakovlev and Leguizamon 2012). Collectively, then, our approach avoids the limitations of many previous studies and helps to clarify existing literature concerning how best to study the relationship between spirituality and subjective well-being. -->

## Ethics Statement

The New Zealand Attitudes and Values Study (NZAVS) is reviewed every three years by the University of Auckland Human Participants Ethics Committee. Our most recent ethics approval statement is as follows: The New Zealand Attitudes and Values Study was approved by The University of Auckland Human Participants Ethics Committee on 03-June-2015 until 03-June-2018, and renewed on 05-September-2017 until 03-June-2021. Reference Number: 014889. Our previous ethics approval statement for the 2009-2015 period is: The New Zealand Attitudes and Values Study was approved by The University of Auckland Human Participants Ethics Committee on 09-September-2009 until 09-September-2012, and renewed on 17-February-2012 until 09-September-2015. Reference Number: 6171. All participants granted informed written consent and The University of Auckland Human Participants Ethics Committee approved all procedures.

## Sampling Procedure

The NZAVS is an annual, longitudinal national probability sample of registered New Zealand voters, which was started in 2009. The Time 10 wave of the NZAVS contained responses from 47,951 participants (17,981 retained from one or more previous waves and 29,970 new additions from booster sampling and/or unmatched participants or unsolicited opt-ins). Participants who provided an email address were also emailed and invited to complete an online version if they preferred. We offered a prize draw for participation, non-respondents were emailed and phoned multiple times, and all participants were mailed a Season’s Greetings card from the NZAVS research team and informed that they had been automatically entered into a bonus seasonal grocery voucher prize draw. We also mailed our yearly pamphlet summarizing key research findings published during the current wave of the study.

## Participants

The Time 10 (2018) wave of the NZAVS included 47,951 respondents. We analyzed data from participants who responded to our survey at least three four times between Time 2 (2010) and Time 10 (2018), resulting in a sample of N = 21,705 (N = 20,979 with complete responses).

## Measures

**Life Satisfaction** Life satisfaction is a measure of emotional well-being that was assessed using a 2-item version of the Satisfaction With Life Scale, which has previously been shown to correlate with aspects of religiosity [@Diener1985-xy]. Participants rated their agreement with the statements (a) "I am satisfied with my life"; and (b) "In most ways my life is close to ideal." The items were rated on 7-point response options ranging from 1 = strongly disagree to 7 = strongly agree. The mean Cronbach’s alpha for this 2-item scale was $\alpha$ = 0.90. Higher scores on this scale indicate higher life satisfaction. Overall scores were means of the 2-items. For detailed information pertaining to means, standard deviations, and missingness, see Table 1.

**Personal well-being** Personal well-being is a measure of function all well-being that was assessed using a 4-item version of the Australian Unity well-being Index (Cummins et al. 2017). Participants rated their satisfaction with (a) "Your standard of living"; (b) "Your health"; (c) "Your future security"; (d) "Your personal relationships." Items were rated using a 10-point response option ranging from 1 = completely dissatisfied to 10 = completely satisfied. The mean Cronbach’s alpha for this scale was $\alpha$ = 0.90. Higher scores on this scale indicate higher personal well-being. Overall scores were means of the 4-items. For detailed information pertaining to means, standard deviations, and missingness, see Table 1.

***Belief in a Spirit or a God***  We assess belief in a spirit or a God by asking two non-affective questions: "Do you believe in a spirit or life force" and "Do you believe in a god?" Responses were coded as "Belief in a spirit that excludes a god" for those who indicated a belief in both a spirit or life force and a god, "Belief that includes a god" for those who indicated a belief in a god with or without an accompanying belief in a spirit or life force, and "skeptics" for those who indicated belief in neither a spirit or life force nor a god. These items appear in the NZAVS questionnaire and were developed for the 2005 Eurobarometer. For detailed information pertaining to beliefs, see Table 1.

Table 1 about here:
\[
\begin{table}[ ht ] 
\centering 
\caption{}\label{}
\scalebox{.6}{
\begin{tabular}{ l c c c c c c c c c }
\toprule
 &   \multicolumn{ 9 }{c}{ Wave }\\ 
  & 2010 & 2011 & 2012 & 2013 & 2014 & 2015 & 2016 & 2017 & 2018 \\ 
 & n = \emph{missing} & n = \emph{missing} & n = 8547 & n = 13472 & n = \emph{missing} & n = 11687 & n = 16048 & n = 14174 & n = 14883 \\ 
 \midrule
Age &   &   &   &   &   &   &   &   &  \\ 
NZdep &   &   &   &   &   &   &   &   &  \\ 
Education &   &   &   &   &   &   &   &   &  \\ 
Employed &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    0 &  (\%) &  (\%) & 2094 (24.5\%) & 2955 (21.9\%) &  (\%) & 2525 (21.6\%) & 3486 (21.7\%) & 3223 (22.7\%) & 3404 (22.9\%)\\ 
\hspace{6pt}    1 &  (\%) &  (\%) & 6453 (75.5\%) & 10517 (78.1\%) &  (\%) & 9162 (78.4\%) & 12562 (78.3\%) & 10951 (77.3\%) & 11479 (77.1\%)\\ 
EthnicCategories &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    Euro &  (\%) &  (\%) & 6731 (78.8\%) & 11024 (81.8\%) &  (\%) & 9671 (82.8\%) & 13389 (83.4\%) & 11814 (83.3\%) & 12427 (83.5\%)\\ 
\hspace{6pt}    Maori &  (\%) &  (\%) & 1188 (13.9\%) & 1608 (11.9\%) &  (\%) & 1341 (11.5\%) & 1749 (10.9\%) & 1620 (11.4\%) & 1677 (11.3\%)\\ 
\hspace{6pt}    Pacific &  (\%) &  (\%) & 284 (3.3\%) & 339 (2.5\%) &  (\%) & 277 (2.4\%) & 331 (2.1\%) & 243 (1.7\%) & 261 (1.8\%)\\ 
\hspace{6pt}    Asian &  (\%) &  (\%) & 344 (4\%) & 501 (3.7\%) &  (\%) & 398 (3.4\%) & 579 (3.6\%) & 497 (3.5\%) & 518 (3.5\%)\\ 
Male &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    0 &  (\%) &  (\%) & 5302 (62\%) & 8457 (62.8\%) &  (\%) & 7325 (62.7\%) & 10148 (63.2\%) & 9011 (63.6\%) & 9433 (63.4\%)\\ 
\hspace{6pt}    1 &  (\%) &  (\%) & 3245 (38\%) & 5015 (37.2\%) &  (\%) & 4362 (37.3\%) & 5900 (36.8\%) & 5163 (36.4\%) & 5450 (36.6\%)\\ 
Partner &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    0 &  (\%) &  (\%) & 2408 (28.2\%) & 3546 (26.3\%) &  (\%) & 2815 (24.1\%) & 3773 (23.5\%) & 3330 (23.5\%) & 3556 (23.9\%)\\ 
\hspace{6pt}    1 &  (\%) &  (\%) & 6139 (71.8\%) & 9926 (73.7\%) &  (\%) & 8872 (75.9\%) & 12275 (76.5\%) & 10844 (76.5\%) & 11327 (76.1\%)\\ 
Pol Orient &   &   &   &   &   &   &   &   &  \\ 
Urban &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    Not\_Urban &  (\%) &  (\%) & 2865 (33.5\%) & 4278 (31.8\%) &  (\%) & 3951 (33.8\%) & 5538 (34.5\%) & 2610 (18.4\%) & 2661 (17.9\%)\\ 
\hspace{6pt}    Urban &  (\%) &  (\%) & 5682 (66.5\%) & 9194 (68.2\%) &  (\%) & 7736 (66.2\%) & 10510 (65.5\%) & 11564 (81.6\%) & 12222 (82.1\%)\\ 
Beliefs &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    \_Skeptic\_ &  (\%) &  (\%) & 2048 (24\%) & 3286 (24.4\%) &  (\%) & 2951 (25.3\%) & 4116 (25.6\%) & 3916 (27.6\%) & 4120 (27.7\%)\\ 
\hspace{6pt}    \_SpiritExcludesGod\_ &  (\%) &  (\%) & 2295 (26.9\%) & 3735 (27.7\%) &  (\%) & 3169 (27.1\%) & 4501 (28\%) & 3918 (27.6\%) & 3970 (26.7\%)\\ 
\hspace{6pt}    GodAndSpirit &  (\%) &  (\%) & 3707 (43.4\%) & 5499 (40.8\%) &  (\%) & 5119 (43.8\%) & 6453 (40.2\%) & 5464 (38.5\%) & 6210 (41.7\%)\\ 
\hspace{6pt}    GodExcludesSpirit &  (\%) &  (\%) & 497 (5.8\%) & 952 (7.1\%) &  (\%) & 448 (3.8\%) & 978 (6.1\%) & 876 (6.2\%) & 583 (3.9\%)\\ 
LIFESAT &   &   &   &   &   &   &   &   &  \\ 
PWI &   &   &   &   &   &   &   &   &  \\ 
\bottomrule
\end{tabular}
}
\end{table}
\]

## Demographic Indicators

**Age** Age was put into units of 10 years and centered at its mean. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 2.
Gender. Gender was assessed by asking participants if they were "Male" was coded as "1" and "Female" was coded as "0." For detailed information pertaining to yearly responses and missingness, see Table 1.

**Education**  Education level was measured using an 11-point rating developed by the New Zealand Qualification Authority known as the New Zealand Qualification Framework (NZQF; 0 = no qualification, 10 = doctoral degree). Education was centered at its mean and standardized. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 1.

**Deprivation** We measured the socio-economic status of participants’ immediate (small area) neighborhood using the 2013 New Zealand Deprivation Index, which uses aggregate census information about the residents of small neighborhood-type units to assign a decile-rank index from 1 (most affluent) to 10 (most impoverished) [@Atkinson2014-ex]. The index is based on a Principal Components Analysis of the following nine variables (in weighted order): the proportion of adults who received a means-tested benefit, household income, proportion not owning own home, proportion single-parent families, proportion unemployed, proportion lacking qualifications, proportion household crowding, proportion no telephone access, and proportion no car access. Thus, the index reflects the average level of deprivation for small neighborhood-type units (or small community areas of about 80–90 people each) across the entire country. Our sample had a mean deprivation index of 4.80 (SD = 2.79). Deprivation was centered at its mean and standardized. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 1.

**Employed** Employment status was assessed by asking participants if they were currently working, "yes" was coded as "1" and "no" was coded as "0." For detailed information pertaining to yearly responses and missingness, see Table 1.

**Partner** Participants were asked if they were in a relationship, "yes" was coded as "1" and no was coded as "0." For detailed information pertaining to yearly responses and missingness, see Table 1.

**Ethnicity** Ethnicity was assessed using four basic categories: (1) New Zealand European/Pakeha, (2) Maori, (3) Pacific Islander, and (4) Asian. For detailed information pertaining to yearly responses and missingness, see Table 1.

**Urban** People were coded as either residing in an urban "1" or rural "0" area based on New Zealand census data. For detailed information pertaining to yearly responses and missingness, see Table 1.

**Political Orientation** To assess political orientation, we asked people to rate their political orientation using seven-point response options (1 = Liberal; 7 = Conservative). Higher values indicate more conservative political beliefs. Political Orientation was standardized and centered at its mean. For detailed information pertaining to means, standard deviations, and missingness, see Table 1.

## Statistical Analyses

Statistical analysis was performed using R version 4.0.2 R version 3.6.1 (2020-06-22) 2019-07-05). We analyzed data from participants who responded to the NZAVS survey at least three four times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of  21,705 (N = 20,979 with complete responses)N = 15,040.  To model predict growth in levels of self-reported psychological subjective well-being among these participants we use generalized linear mixed models. In our models, we account for the non-linear effects of time within participants by including its interaction with both of our predictor variables: "Belief in a spirit that excludes god" and "Belief that includes a god." To handle the dependencies and heterogeneity introduced from the repeated measures, we We also included individual ID as an effect modelled as random-effect. To improve interpretability of our models, we centered and scaled the education, deprivation, and political orientation covariates. Additionally, age was centred at and put into decade-long intervals. 
The model equation(s) are described in equation 1.

\[
\begin{aligned}
\operatorname{{\small LIFESAT/PWI}}  &\sim N \left(\mu,\sigma^2 \right) \\ \mu &=\alpha_{j[i]} + \beta_{1}(\operatorname{Years}) + \beta_{2}(\operatorname{Beliefs}_{\operatorname{\_SpiritExcludesGod\_}})\ + \\
&\quad \beta_{3}(\operatorname{Beliefs}_{\operatorname{GodAndSpirit}}) + \beta_{4}(\operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) + \beta_{5}(\operatorname{Age.10yrs.C})\ + \\
&\quad \beta_{6}(\operatorname{Deprivation.S}) + \beta_{7}(\operatorname{Edu.S}) + \beta_{8}(\operatorname{Employed})\ + \\
&\quad \beta_{9}(\operatorname{EthnicCats}_{\operatorname{Maori}}) + \beta_{10}(\operatorname{EthnicCats}_{\operatorname{Pacific}}) + \beta_{11}(\operatorname{EthnicCats}_{\operatorname{Asian}})\ + \\
&\quad \beta_{12}(\operatorname{Male}_{\operatorname{1}}) + \beta_{13}(\operatorname{Partner}_{\operatorname{Has\ }}) + \beta_{14}(\operatorname{Pol.Orient.S})\ + \\
&\quad \beta_{15}(\operatorname{Urban}) + \beta_{16}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{\_SpiritExcludesGod\_}}) + \beta_{17}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ + \\
&\quad \beta_{18}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) \\ \alpha_{j} &\sim N \left(\mu_{\alpha_{j}},\sigma^2_{\alpha_{j}} \right) , \operatorname{ for  Id } j  = 1, \dots ,  J
\end{aligned}
\]

Because missing values can result in biased estimates [@Blackwell2017-oq], we multiply imputed 10 datasets using the AmeliaII package in R[@Honaker2011-yu].  We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences.

## Code

All code for this analysis and a full report of the missing data imputation can be found at https://github.com/jacopastorius/NZAVS_Highland_SpiritHealth2.Table 1. Descriptive Statistics for Theoretical Variables in Each Wave. Belief in a god and belief in a spirit or life force were not measured during Wave 6 (2014). "BXG" indicates a belief in a spirit that excludes a god. "BIG" indicates a belief that includes a god. Many individuals appear in multiple waves.

Table 2. Descriptive Statistics for Demographic Variables in Each Wave. Education level was not measured during Waves 2 (2010) and 3 (2011). Many individuals appear in multiple waves.

# Results

Figure 2 about here:

```{r, cache =TRUE, include =FALSE}
# model 2
mod.1 <-lmer (PWI ~  1 +  Years * Beliefs +  Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id), data= dfben, na.action = na.omit)
```

```{r, cache=TRUE, include=FALSE}
# model 1
mod.1 <-lmer (LIFESAT ~ 1 + Years * Beliefs+ Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id), data= dfben, na.action = na.omit)
summary(mod.1)
```

Results for both the personal well-being and life satisfaction models are presented in Table 2 and Figure 1.

```{r fig.cap = "Combination Coefficient Plots for Personal Well-being and Life Satisfaction Spirit Belief Longitudinal Models"}
knitr::include_graphics("figs/USEcoefficientplots-1.png")
```

Table 2 about here:

\[
\begin{table}
\centering 
\caption{"Regression Models"}\label{}
\scalebox{0.6}{
\begin{tabular}{l c c}
\toprule
 & Model 1 & Model 2 \\
\midrule
(Intercept)                        & $\mathbf{6.63}^{***}$  & $\mathbf{4.94}^{***}$  \\
                                   & $(0.03)$               & $(0.02)$               \\
Years                              & $0.00$                 & $-0.00$                \\
                                   & $(0.00)$               & $(0.00)$               \\
Beliefs\_SpiritExcludesGod\_       & $\mathbf{-0.14}^{***}$ & $\mathbf{-0.08}^{***}$ \\
                                   & $(0.03)$               & $(0.02)$               \\
BeliefsGodAndSpirit                & $\mathbf{-0.16}^{***}$ & $-0.04$                \\
                                   & $(0.03)$               & $(0.02)$               \\
BeliefsGodExcludesSpirit           & $0.02$                 & $0.07$                 \\
                                   & $(0.05)$               & $(0.04)$               \\
Age.10yrs.C                        & $\mathbf{0.08}^{***}$  & $\mathbf{0.09}^{***}$  \\
                                   & $(0.01)$               & $(0.01)$               \\
Deprivation.S                      & $\mathbf{-0.14}^{***}$ & $\mathbf{-0.07}^{***}$ \\
                                   & $(0.01)$               & $(0.01)$               \\
Edu.S                              & $\mathbf{0.12}^{***}$  & $\mathbf{0.06}^{***}$  \\
                                   & $(0.01)$               & $(0.01)$               \\
Employed1                          & $\mathbf{0.09}^{***}$  & $\mathbf{0.08}^{***}$  \\
                                   & $(0.01)$               & $(0.01)$               \\
EthnicCatsMaori                    & $\mathbf{-0.30}^{***}$ & $\mathbf{-0.06}^{**}$  \\
                                   & $(0.03)$               & $(0.02)$               \\
EthnicCatsPacific                  & $\mathbf{-0.29}^{***}$ & $\mathbf{-0.08}^{*}$   \\
                                   & $(0.05)$               & $(0.04)$               \\
EthnicCatsAsian                    & $-0.08$                & $-0.04$                \\
                                   & $(0.05)$               & $(0.03)$               \\
Male1                              & $\mathbf{-0.11}^{***}$ & $\mathbf{-0.19}^{***}$ \\
                                   & $(0.02)$               & $(0.01)$               \\
Partner1                           & $\mathbf{0.61}^{***}$  & $\mathbf{0.42}^{***}$  \\
                                   & $(0.02)$               & $(0.01)$               \\
Pol.Orient.S                       & $\mathbf{0.02}^{***}$  & $\mathbf{0.01}^{**}$   \\
                                   & $(0.01)$               & $(0.00)$               \\
UrbanUrban                         & $\mathbf{-0.02}^{*}$   & $\mathbf{-0.02}^{*}$   \\
                                   & $(0.01)$               & $(0.01)$               \\
Years:Beliefs\_SpiritExcludesGod\_ & $\mathbf{0.02}^{**}$   & $\mathbf{0.02}^{***}$  \\
                                   & $(0.00)$               & $(0.00)$               \\
Years:BeliefsGodAndSpirit          & $\mathbf{0.02}^{***}$  & $\mathbf{0.01}^{***}$  \\
                                   & $(0.00)$               & $(0.00)$               \\
Years:BeliefsGodExcludesSpirit     & $0.00$                 & $-0.00$                \\
                                   & $(0.01)$               & $(0.01)$               \\
\midrule
AIC                                & $248580.18$            & $200599.16$            \\
BIC                                & $248775.08$            & $200793.96$            \\
Log Likelihood                     & $-124269.09$           & $-100278.58$           \\
Num. obs.                          & $79270$                & $78888$                \\
Num. groups: Id                    & $20979$                & $20972$                \\
Var: Id (Intercept)                & $1.78$                 & $0.88$                 \\
Var: Residual                      & $0.74$                 & $0.42$                 \\
\bottomrule
\multicolumn{3}{l}{\scriptsize{$^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$}}
\end{tabular}
}
\end{table}
\]

```{r, echo = FALSE, include=FALSE}
te00 <-texreg::extract(
  mod.1,
  level = 0.95,
  include.random = TRUE,
  include.rsquared = T,
  include.nobs = T,
  include.loo.ic = T,
  include.waic = F)
texreg(list(te00),
                custom.model.names = c("PWI"),
                caption = "Personal Wellbeing",
                sideways = F,
                scalebox = .5,
                #fontsize= "footnotesize",
                label = "tab:REGRESS_1",
                ci.force.level = 0.95, bold = 0.05,
                settingstars = 0,
                booktabs = TRUE,
                custom.note ="")
```

## Personal Well-Being

We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict personal well-being (PWI) with Years (0-9), Beliefs (four categories), Age.10yrs.C, Deprivation.S, Edu.S, Employed, EthnicCats, Male, Partner, Pol.Orient.S and Urban, with  Id as random effects. The model included Id as random effects.  The model's intercept, corresponding to Years = 0, Beliefs = Skeptic, Age = (sample mean), Deprivation.S = (sample mean), Edu.S =(sample mean), Employed = Not Employed, EthnicCats = Euro, Male = Not Mail, Partner = No Partner, Pol.Orient.S = (sample mean),, Urban = Not Urban and Id = 2, is at 4.94 (SE = 0.02, 95% CI [4.89, 4.99], p < .001) is at 6.63 (SE = 0.03, 95% CI [6.57, 6.70], p < .001). Within this model:

  - The effect of Years is positive is not reliable (beta = 2.66e-03, SE = 3.58e-03, std. beta = 3.19e-03, p = 0.458).  We infer that Skeptics personal well-being is not increasing in the Skeptic population
  - The effect of Beliefs_SpiritExcludesGod_is reliably negative (beta = -0.14, SE = 0.03, std. beta = -0.03, p < .001). We infer that overall the population that believes in a Spirit but does not believe in a God is expected to have a lower personal well-being than skeptics. 
  - The effect of BeliefsGodAndSpirit is reliably negative (beta = -0.16, SE = 0.03, std. beta = -0.02, p < .001). We infer that overall the population that believes in God and a spirit or life force is expected to have a lower Personal Well-being than skeptics. 
  - The effect of BeliefsGodExcludesSpirit is not reliably different from the skeptic population (beta = 0.02, SE = 0.05, std. beta = 0.02, p = 0.669).
  - The effect of Age.10yrs.C is reliably positive  (beta = 0.08, SE = 7.38e-03, std. beta = 0.07, p < .001). Older people have greater personal well-beingwellbeing.
  - The effect of Deprivation.S is reliably negative (beta = -0.14, SE = 7.50e-03, std. beta = -0.08, p < .001). Deprived people have lower personal well-beingwellbeing.
  - The effect of Edu.S is reliably positive  (beta = 0.12, SE = 8.74e-03, std. beta = 0.07, p < .001). Educated people greater personal well-beingwellbeing
  - The effect of Employed is reliably positive  (beta = 0.09, SE = 0.01, std. beta = 0.05, p < .001). Employed people have greater personal well-beingwellbeing.
  - The effect of EthnicCatsMaori is reliably negative (beta = -0.30, SE = 0.03, std. beta = -0.18, p < .001).
  - The effect of EthnicCatsPacific is reliably negative (beta = -0.29, SE = 0.05, std. beta = -0.17, p < .001).
  - The effect of EthnicCatsAsian is not reliable (beta = -0.08, SE = 0.05, std. beta = -0.05, p = 0.089).
  - The effect of Male is reliably negative (beta = -0.11, SE = 0.02, std. beta = -0.06, p < .001).
  - The effect of Having a Partner is reliably positive (beta = 0.61, SE = 0.02, std. beta = 0.37, p < .001).
  - The effect of Pol.Orient.S is reliably positive  (beta = 0.02, SE = 5.67e-03, std. beta = 0.01, p < .001).
  - The effect of Urban is reliably negative (beta = -0.02, SE = 0.01, std. beta = -0.01, p < .05).

**Interactions**

```{r fig.cap = "Combination Coefficient Plots for Personal Well-being and Life Satisfaction Spirit Belief Longitudinal Models"}
knitr::include_graphics("Figs/USEPWI_expected-1.png")
```
  - The effect of Years X Beliefs_SpiritExcludesGod_ is reliably positive  (beta = 0.02, SE = 4.86e-03, std. beta = 0.02, p < .01). We infer that, unlike the skeptic population, people with beliefs in a spirit or life force who do not believe in a God tend to grow in their personal well-beingwellbeing over time.
  - The effect of Years X BeliefsGodAndSpirit is reliably positive  (beta = 0.02, SE = 4.41e-03, std. beta = 0.03, p < .001). We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their personal well-beingwellbeing over time
  - The effect of Years X BeliefsGodExcludesSpirit is not reliable(beta = 1.96e-03, SE = 8.29e-03, std. beta = 2.36e-03, p = 0.813).  Although people who believe in a God but do not believe in a spirit or Life Force tend to have greater personal well-beingwellbeing than those who do not believe in a spirit or life force  (and greater life satisfaction than Skeptics)  this population is not expected to grow in Life Satisfaction over time.

## Life Satisfaction

We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict Life Satisfaction (LIFESAT) with Years (0-9), Beliefs (four categories), Age.10yrs.C, Deprivation.S, Edu.S, Employed, EthnicCats, Male, Partner, Pol.Orient.S and Urban, with  Id as random effects. Standardized parameters were obtained by fitting the model on a standardized version of the dataset.The model's total explanatory power is substantial (conditional R2 = 0.69). The model's intercept, corresponding to Years = 0, Beliefs = Skeptic, Age = (sample mean),, Deprivation.S = (sample mean), Edu.S =(sample mean), Employed = Not Employed, EthnicCats = Euro, Male = Not Mail, Partner = No Partner, Pol.Orient.S = (sample mean), Urban = Not Urban and Id = 2, is at 4.94 (SE = 0.02, 95% CI [4.89, 4.99], p < .001). Within this model:

  - The effect of Years is not reliable (beta = -5.50e-04, SE = 2.69e-03, std. beta = -9.32e-04, p = 0.838).  We do not infer growth over time in Life Satisfaction among the Skeptic population
  - The effect of BeliefsSpiritExcludesGod  is reliably negative (beta = -0.08, SE = 0.02, std. beta = 0.01, p < .001). We infer that overall the population that believes in a Spirit but does not believe in a God is expected to have a lower Life Satisfaction than skeptics. 
  - The effect of BeliefsGodAndSpirit is not reliable (beta = -0.04, SE = 0.02, std. beta = 0.04, p = 0.105). We do not infer growth over time in Life Satisfaction among this (Skeptic) population
  - The effect of BeliefsGodExcludesSpirit is reliably positive (beta = 0.07, SE = 0.04, std. beta = 0.05, p = 0.062).We infer that overall the population that believes in a God but does not believe in a spirit is not reliably different from  the Skeptics population in Life Satisfaction. 
  - The effect of Age.10yrs.C is reliably positive (beta = 0.09, SE = 5.26e-03, std. beta = 0.10, p < .001). Older people have greater Life Satisfaction
  - The effect of Deprivation.S is  is reliably negative (beta = -0.07, SE = 5.50e-03, std. beta = -0.06, p < .001). People who are more deprived have lower Life Satisfaction.
  - The effect of Edu.S is reliably positive (beta = 0.06, SE = 6.33e-03, std. beta = 0.05, p < .001). Educated people have greater Life Satisfaction
  - The effect of Employed is reliably positive (beta = 0.08, SE = 9.69e-03, std. beta = 0.07, p < .001). People who are employed tend to have greater Life Satisfaction
  - The effect of EthnicCatsMaori is reliably negative (beta = -0.06, SE = 0.02, std. beta = -0.05, p < .01).  People who belong to the minority Maori group tend to have lower Life Satisfaction.
  - The effect of EthnicCatsPacific is reliably negative (beta = -0.08, SE = 0.04, std. beta = -0.07, p < .05).  People who belong to the minority Pacific peoples group tend to have lower life satisfaction.
  - The effect of EthnicCatsAsian is not reliably different from people of European ethnicity in life satisfaction(beta = -0.04, SE = 0.03, std. beta = -0.03, p = 0.223).
  - The effect of Male is reliably negative (beta = -0.19, SE = 0.01, std. beta = -0.16, p < .001).
  - The effect of HasPartner is reliably positive (beta = 0.42, SE = 0.01, std. beta = 0.35, p < .001).
  - The effect of Pol.Orient.S is reliably positive (beta = 0.01, SE = 4.23e-03, std. beta = 0.01, p < .01).
  - The effect of Urban is reliably negative (beta = -0.02, SE = 8.85e-03, std. beta = -0.02, p < .05).

**Interactions**

```{r fig.cap = "Prediction Plot Life Satisfaction Spirit Belief Longitudinal Models"}
knitr::include_graphics("Figs/USElifsat_expected-1.png")
```
  - The interaction of Years X Beliefs_SpiritExcludesGod_ is reliably positive (beta = 0.02, SE = 3.66e-03, std. beta = 0.03, p < .001). We infer that, unlike the Skeptic population, people with beliefs in a Spirit or Life Force who do not believe in a God tend to grow in their Life Satisfaction over time.
  - The interaction of Years X BeliefsGodAndSpirit is reliably positive (beta = 0.01, SE = 3.32e-03, std. beta = 0.02, p < .001). We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their Life Satisfaction over time
  - The effect of Years X BeliefsGodExcludesSpirit is not reliable (beta = -1.59e-03, SE = 6.25e-03, std. beta = -2.70e-03, p = 0.799).  Although people who believe in a God but do not believe in a spirit or Life Force tend to have greater life satisfaction than those who do not believe in a Spirit or life force  (and greater life satisfaction than Skeptics)  this population is not expected to grow in Life Satisfaction over time.

<!-- ```{r} -->
<!-- knitr::include_graphics("Figs/Usecoefficientplots-1.png") -->
<!-- ``` -->

Figure 2. Subjective Well-being Predicted Probabilities for Belief in a Spirit or Life Force Over Time in a Listwise Deleted Dataset.

Table 3. Subjective Well-being Coefficient Table for Belief in a Spirit or Life Force. Age was put into units of 10 years and centered at its mean. Education, Deprivation, and Political Orientation were centered at their means and standardized.

# Discussion

The purpose of this study has been to clarify the relationship between belief in a spirit/life force and psychological well-being. Spirituality as a distinct category, separate from religion, is a little-studied phenomenon in the psychology of religion literature.

These findings demonstrate the importance ofinvestigating spiritual beliefs as a domain of interest in its own right. In many Western countries, people are increasingly identifying as "spiritual but not religious" [@Zinnbauer2005-vz; @Swinton2001-vr]. Other studies find that those who have a spiritual understanding of life without a religious framework are at an increased risk of having mental disorders [@King2013-cg]. As such, it is important to understand the ramifications of belief in a spirit for psychological health. 
To assess the relationship between a belief in a spirit or life force and subjective well-being, we improve upon past methods in three ways.

Our findings support critiques of the reliability of past methods. The majority of studies have found positive relationships between religion and spirituality and subjective well-being. These previous studies primarily use samples taken from within the North Atlantic region and are cross-sectional in design (Ano and Vasconcelles 2005; Hackney and Sanders 2003; Sawatzky, Ratner, and Chiu 2005; Smith, McCullough, and Poll 2003; Visser, Garssen, and Vingerhoets 2010; Yonker, Schnabelrauch, and Dehaan 2012). However, our results find a negative association between both belief in a spirit or life force that excludes a god (BXG) and belief that includes a god (BIG). These longitudinal results persist independent of time. Notably, our findings do partially support other previous work that shows that spirituality outside of a religious framework is associated with worse emotional well-being, though we also find a negative association within a religious framework (King et al. 2013). Still, further research is needed to determine if other aspects of religion and spirituality, such as practices and rituals, might yield different results.

Despite the opposition of our findings to those of many previous studies, there is room for agreement. We show that over time both "BXG" and "BIG" are associated with marginally better subjective well-being. As seen in Figure 2., both "BXG" and "BIG" predict marginally higher personal well-being and life satisfaction as time progresses. So, while those who adhere to these beliefs may start out below their skeptic counterparts in terms of their subjective well-being, they do eventually catch up. We also find that their life satisfaction is actually improving at a greater rate than skeptics. Thus, in the long term, both sustained "BXG" and "BIG" may actually be associated with better subjective well-being.
Our results also help to clarify the relationship among religion, spirituality, and subjective well-being outcomes. BXG and BIG performed almost identically in terms of their ability to predict subjective well-being outcomes. This trend suggests that any kind of existential belief - regardless of whether a god is involved - might affect subjective well-being in much the same way. This result is good news as increasing numbers of individuals continue to report spiritual but not necessarily religious beliefs.

Taken as a whole, our results cannot refute the entire narrative of the positive relationship between spirituality and emotional well-being. First, our analysis focuses only on belief. While belief is an important part of spirituality it is not by any means spirituality itself. Individuals may identify as spiritual without any accompanying belief in a spirit or life force. Moreover, practices and rituals associated with different forms of spirituality that were not tested in this study may have large effects. Second, despite flawed methods, past studies may still be accurate. New Zealand is a largely secular country, so belief in a spirit or life force may have different impacts on subjective well-being outcomes in this context than in other, more religious contexts. Overall, though, our findings add to the existing literature, helping to untangle the multifaceted relationship among religion, spirituality, and subjective well-being.

## Limitations

This article makes useful contributions to the continued debate regarding the relationship between religion and spirituality and mental health outcomes, but it is important to acknowledge several limitations. First, we relied upon self-reported data from the NZAVS. Although this is an accurate method for measuring belief in a spirit or life force, it may be a less accurate manner of measuring subjective well-being. It is possible that self-administration of our emotional well-being (personal well-being and life satisfaction) scales could result in mismeasurement. Second, we may not capture the full range spiritualities, which likely extend beyond belief and might even be inherently emotional. Thus, while our method avoided tautology, it might have also excluded participants who consider themselves to be spiritual but do not subscribe to a belief in a spirit or life force.

### Authors contributions
CS leads and manages the NZAVS, and collected the data for this study. BH and JB conceived
of the study. BH wrote the first draft of the manuscript and performed the analysis with input from JB. BH conducted background research and scientific literature review. All authors contributed substantially to the intellectual content of the study and the final manuscript.

### Funding statement
The New Zealand Attitudes and Values Study is supported by a grant from the Templeton
Religion Trust (TRT0196). The funders had no role in the preparation of the article or the
decision to publish.

# Supplemental Material
S1. Statistical Analysis in R
Statistical analysis was performed using R version 3.6.1 (2019-07-05). Platform: x86_64-w64-mingw32 (64-bit), running under: Windows 10 (Action of the Toes). Generalized Linear Mixed Models were generated using the glmmTMB (Brooks et al. 2017; Knowles and Frederick 2019) package in R. In addition, we used the following R packages: dplyr (Hadley Wickham et al. 2019), tidyr (H. Wickham and Henry 2019), sjPlot (D. Lüdecke 2020), ggeffects (Daniel Lüdecke 2018), table1  and their dependencies.
Tables and coefficient plots were generated using the glmmTMB, ggeffects, and sjPlot packages in R. Models for both subjective well-being variables used family = gaussian, assuming a normal distribution. We ran each individual’s unique ID as a random-effect. The tab_model and plot_model commands in sjPlot allowed for the creation of tables and coefficient plots. Predicted probability plots were generated using the ggpredict command in the ggeffects package. The table1 package in R was used for the creation of descriptive statistics tables.

\newpage

# References

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

<div id="refs" custom-style="Bibliography"></div>
\endgroup

### 2020-11-16 Manuscript.tex
initial commit

--- 2020-11-16 Manuscript.Rmd
@@
----
-title             : "National Longitudinal Study Reveals Growth in Wellbeing Among Those with Spiritual Beliefs"
-shorttitle        : "Beliefs and Wellbeing"
-author: 
-  - name          : "Ben Highland"
-    affiliation   : "1"
-    corresponding : yes    # Define only one corresponding author
-    address       : "Postal address"
-    email         : "my@email.com"
-    role:         # First author
-      - Conceptualization
-      - Literature Review
-      - Writing - First draft
-      - Writing - Review & editing
-  - name          : "Ev l Worthing"
-    affiliation   : "2"
-    address       : "Postal address"
-    email         : "my@email.com"
-    role:        
-      - Writing review & editing
-  - name          : "Don E Davis"
-    affiliation   : "3"
-    address       : "Postal address"
-    email         : "my@email.com"
-    role:         
-      - Writing review & editing
-  - name          : "Chris G. Sibley"
-    affiliation   : "4"
-    address       : "Postal address"
-    email         : "my@email.com"
-    role:        
-      - Writing review & editing
-  - name          : "Joseph Bulbulia"
-    affiliation   : "5"
-    role:         
-      - Data analysis
-      - Writing second and subsequent drafts
-      - Writing review & editing
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+%
+\documentclass[
+  english,
+  man]{apa6}
+\usepackage{lmodern}
+\usepackage{amssymb,amsmath}
+\usepackage{ifxetex,ifluatex}
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
+  \usepackage{unicode-math}
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+% Use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\IfFileExists{microtype.sty}{% use microtype if available
+  \usepackage[]{microtype}
+  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
+  \IfFileExists{parskip.sty}{%
+    \usepackage{parskip}
+  }{% else
+    \setlength{\parindent}{0pt}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\hypersetup{
+  pdftitle={National Longitudinal Study Reveals Growth in Wellbeing Among Those with Spiritual Beliefs},
+  pdfauthor={Ben Highland1, Ev l Worthing2, Don E Davis3, Chris G. Sibley4, \& Joseph Bulbulia5},
+  pdflang={en-EN},
+  pdfkeywords={Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being},
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
+\urlstyle{same} % disable monospaced font for URLs
+\usepackage{graphicx,grffile}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+% Make \paragraph and \subparagraph free-standing
+\ifx\paragraph\undefined\else
+  \let\oldparagraph\paragraph
+  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+  \let\oldsubparagraph\subparagraph
+  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
+% Manuscript styling
+\usepackage{upgreek}
+\captionsetup{font=singlespacing,justification=justified}
 
-authornote: |
+% Table formatting
+\usepackage{longtable}
+\usepackage{lscape}
+% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
+\usepackage{multirow}		% Table styling
+\usepackage{tabularx}		% Control Column width
+\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
+\usepackage{threeparttablex}            % Lets threeparttable work with longtable
+
+% Create new environments so endfloat can handle them
+% \newenvironment{ltable}
+%   {\begin{landscape}\begin{center}\begin{threeparttable}}
+%   {\end{threeparttable}\end{center}\end{landscape}}
+\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}
+
+% Enables adjusting longtable caption width to table width
+% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
+\makeatletter
+\newcommand\LastLTentrywidth{1em}
+\newlength\longtablewidth
+\setlength{\longtablewidth}{1in}
+\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}
+
+% \setlength{\parindent}{0.5in}
+% \setlength{\parskip}{0pt plus 0pt minus 0pt}
+
+% \usepackage{etoolbox}
+\makeatletter
+\patchcmd{\HyOrg@maketitle}
+  {\section{\normalfont\normalsize\abstractname}}
+  {\section*{\normalfont\normalsize\abstractname}}
+  {}{\typeout{Failed to patch abstract.}}
+\patchcmd{\HyOrg@maketitle}
+  {\section{\protect\normalfont{\@title}}}
+  {\section*{\protect\normalfont{\@title}}}
+  {}{\typeout{Failed to patch title.}}
+\makeatother
+\shorttitle{Beliefs and Wellbeing}
+\keywords{Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being\newline\indent Word count: FALSE}
+\DeclareDelayedFloatFlavor{ThreePartTable}{table}
+\DeclareDelayedFloatFlavor{lltable}{table}
+\DeclareDelayedFloatFlavor*{longtable}{table}
+\makeatletter
+\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
+\makeatother
+\usepackage{csquotes}
+\usepackage{tcolorbox}
+\usepackage{paralist}
+\usepackage{multicol}
+\ifxetex
+  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
+  \usepackage{polyglossia}
+  \setmainlanguage[]{english}
+\else
+  \usepackage[shorthands=off,main=english]{babel}
+\fi
+
+\title{National Longitudinal Study Reveals Growth in Wellbeing Among Those with Spiritual Beliefs}
+\author{Ben Highland\textsuperscript{1}, Ev l Worthing\textsuperscript{2}, Don E Davis\textsuperscript{3}, Chris G. Sibley\textsuperscript{4}, \& Joseph Bulbulia\textsuperscript{5}}
+\date{}
+
+\authornote{
+
 Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.
 Enter author note here.
-abstract: |
+
+The authors made the following contributions.
+Ben Highland: Conceptualization, Literature Review, Writing - First draft, Writing - Review \& editing; Ev l Worthing: Writing review \& editing; Don E Davis: Writing review \& editing; Chris G.
+Sibley: Writing review \& editing; Joseph Bulbulia: Data analysis, Writing second and subsequent drafts, Writing review \& editing.
+
+Correspondence concerning this article should be addressed to Ben Highland, Postal address. E-mail: \href{mailto:my@email.com}{\nolinkurl{my@email.com}}
+
+}
+
+\affiliation{\phantom{0}}
+
+\abstract{
 Previous research finds an association between spirituality and subjective well-being.
 However, the widespread use of tautological scales and the reliance on cross-sectional data cast doubts.
-Here, we leverage nine years of panel response data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction ($N$ =  20979, New Zealand, 2010-2018).
+Here, we leverage nine years of panel response data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction (\(N\) = 20979, New Zealand, 2010-2018).
 Contrary to previous studies, we find that belief in a spirit or life force predicts initially lower personal wellbeing and life-satisfaction.
@@
 The causal mechanisms by which beliefs in a spirit or life force increases well-being remain an important horizon for future investigations.
+}
 
-  <!-- https://tinyurl.com/ybremelq -->
-keywords          : "Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being"
-wordcount         : no
-bibliography      : ["b-references.bib"]
-floatsintext      : no
-figurelist        : no
-tablelist         : no
-footnotelist      : no
-linenumbers       : no
-mask              : no
-draft             : no
-documentclass     : "apa6"
-classoption       : "man"
-output            : papaja::apa6_pdf
-extra_dependencies: ["rotating", "booktabs", "graphicx","longtable"]
-header-includes:
-   - \usepackage{tcolorbox}
-   - \usepackage{paralist}
-   - \usepackage{multicol}
----
+\begin{document}
+\maketitle
 
-```{r include = FALSE}
-library(knitr)
-opts_chunk$set(
-    fig.align = "center",
-    root.dir = "../"
-)
-library("papaja")
-library("tidyverse")
-library("tinytex")
-```
-
-```{r load_libraries, echo=FALSE, include=FALSE}
-# load libraries
-library("styler")
-library("tidyverse")
-library("lme4")
-library("ggplot2")
-library("tidyr")
-library("sjPlot")
-library("ggeffects")
-library("Amelia")
-library("table1")
-library("patchwork")
-library("parameters")
-library("prettycode")
-library("gghighlight")
-library("sjPlot")
-library("see")
-library("viridis")
-library("ggpubr")
-library("wesanderson")
-library("ggsci")
-library("here")
-library("citr")
-#library("furniture")
-library("knitr")
-library("texreg")
-```
-```{r load_data, include=FALSE}
-ldf.5<-readRDS("~/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/00Bulbulia\ Pubs/2020/ldf.5") 
-```
-```{r data processing, cache = TRUE, dependson = "ldf.5"}
-# filter only those who are measured 
-dfben  <-ldf.5 %>%
-  dplyr:: filter(Wave !=2009)%>% # exclude wave 1 (no spirit measures that wave)
-  dplyr::filter(YearMeasured==1)%>%
-  dplyr::group_by(Id) %>% filter(n() > 2)%>% # select those who have responded to at least 3 waves # Also works for 5 and 7
-  dplyr::filter(n() !=0)%>%
-  dplyr::ungroup(Id)%>%
-  dplyr:: mutate(Years = as.numeric(years))%>%
-  dplyr::mutate(Age.10yrs = (Age/10))%>%
-  dplyr::mutate(Religion.Church.Log = log(Religion.Church+ 1))%>%
-  dplyr::mutate(Pol.Orient.S = scale(Pol.Orient, center = TRUE, scale = TRUE),
-         Age.10yrs.C = scale(Age.10yrs, center = TRUE, scale = FALSE),
-         Male = factor(Gender),
-         Employed = factor(Employed),
-         Partner = factor(Partner),
-         Edu.S = scale(as.numeric(Edu) ,scale=TRUE,center=TRUE),
-         Education = as.numeric(Edu),
-         Urban = factor(Urban),
-         Believe.Spirit = as.factor(Believe.Spirit),
-         Believe.God = as.factor(Believe.God),
-         EthnicCategories = factor(EthnicCats),
-         LIFESAT.S = scale(LIFESAT, center = TRUE, scale = TRUE),
-         PWI.S = scale(PWI, center = TRUE, scale = TRUE),
-         Deprivation.S = scale(NZdep, scale=TRUE, center=TRUE),
-         Relid.C = scale(Relid, scale = TRUE,center=TRUE),
-         Religious = as.factor(Religious),
-         Religion.Church.Log.C = scale(Religion.Church.Log, center = TRUE, scale = FALSE))
-```
-```{r}
-# nice labels
-table1::label(dfben$Age.10yrs.C) <- "Age in Decades (C)"
-table1::label(dfben$PWI) <- "Personal Wellbeing"
-table1::label(dfben$LIFESAT) <- "Life Satisfaction"
-
-table1::label(dfben$Deprivation.S) <- "Deprivation (S)"
-table1::label(dfben$Pol.Orient) <- "Political Conservative"
-table1::label(dfben$Pol.Orient.S) <- "Political Conservative (S)"
-table1::label(dfben$NZdep) <- "NZ Deprivation Index"
-dfben$Partner <- factor(dfben$Partner, labels = c("No Partner","Has Partner"))
-dfben$Employed <- factor(dfben$Employed, labels = c("Not Employed","Employed"))
-table1::label(dfben$Edu.S) <- "Education (S)"
-
-# create beliefs indicator
-dfben$Beliefs<- factor(ifelse(dfben$Believe.God == "Not Believe God" & dfben$Believe.Spirit == "Not Believe Spirit", "_Skeptic_", 
-                                        ifelse(dfben$Believe.God == "Not Believe God" & dfben$Believe.Spirit == "Believe Spirit", "_SpiritExcludesGod_",
-                                               ifelse(dfben$Believe.God == "Believe God" & dfben$Believe.Spirit == "Believe Spirit","GodAndSpirit","GodExcludesSpirit"))))
-```
-
-The relationship between spirituality and psychological well-being is a matter of enduring human fascination [@Ellsworth2010-yu].
-Previous attempts to quantify the relationship between these two domains find a positiven association between spirituality and well-being [@Russinova2002-rq; @Ginsburg1995-jr; @Zaza2005-ac; @Koenig2010-gk; @Smith2003-re; @Smith2003-re; @Koenig2001-ow].
+The relationship between spirituality and psychological well-being is a matter of enduring human fascination (Ellsworth \& Ellsworth, 2010).
+Previous attempts to quantify the relationship between these two domains find a positiven association between spirituality and well-being (Ginsburg, Quirt, Ginsburg, \& MacKillop, 1995; Koenig, 2010; Koenig, Associate Professor of Psychology and Religious Studies Michael E McCullough, PhD, McCullough, Larson, \& Adjunct Professor of Psychiatry David B Larson, 2001; Russinova, Wewiorski, \& Cash, 2002; Smith, McCullough, \& Poll, 2003; Smith et al., 2003; Zaza, Sellick, \& Hillier, 2005).
 However, critics have raised three challenges.
 
-<!-- In spirituality/psychological wellbeing research,sSsubjective in spirituality health research, psychological well-being is typically decomposed into two categories: emotional well-being (i.e., the presence or absence of positive feelings about life) and functional well-being (i.e., psychological and social well-being) [@Keyes2002-bb].
--->
-
 First, previous research has frequently operationalised spirituality using measures of well-being.
-For example, The Daily Spiritual Experience Scale, which is widely used in spiritual mental health research includes items such as "I feel thankful for my blessings" and "I feel deep inner peace or harmony" to measure spirituality [@Underwood2002-hg].
-As Koenig observes, it is unsurprising that well-being predicts well-being –.
-what Koenig describes as a  "tautology" in spirituality/mental health research [@Koenig2008-lv].
-Indeed, recent systematic-review of spirituality/mental health research found that nearly 45% of previous studies investigating spirituality and emotional well-being operationalise spirituality scales using constructs for positive emotional well-being [@Garssen2016-kb].
+For example, The Daily Spiritual Experience Scale, which is widely used in spiritual mental health research includes items such as \enquote{I feel thankful for my blessings} and \enquote{I feel deep inner peace or harmony} to measure spirituality (Underwood \& Teresi, 2002).
+As Koenig observes, it is unsurprising that well-being predicts well-being --.
+what Koenig describes as a \enquote{tautology} in spirituality/mental health research (Koenig, 2008).
+Indeed, recent systematic-review of spirituality/mental health research found that nearly 45\% of previous studies investigating spirituality and emotional well-being operationalise spirituality scales using constructs for positive emotional well-being (Garssen et al., 2016).
 The first challenge, then, is to define measures of spirituality that are not well-being measures reclocked in other words.
@@
 We can look to folk traditions for guidance.
-The term "spirituality" was first used to describe the religious practices of ascetics and monks.
-more recently, however, the term has taken on a family of broader meaning [@Swinton2001-vr].
+The term \enquote{spirituality} was first used to describe the religious practices of ascetics and monks.
+more recently, however, the term has taken on a family of broader meaning (Swinton, 2001).
 Religion is typically operationalized to mean beliefs and practices respecting a God or gods (JB).
-A recurring element in the use of ‘spirituality,’ as it is distinguished from religion, is belief in a spirit or life force without commitment to a belief in a God or gods [@Hill2000-cl; @Davis2015-tx].
-However, the term "spirituality" overlaps with the uses of "religion" in the sense that God or gods are instances of beliefs in spirits – that is one may be both religious and spiritual.
+A recurring element in the use of \enquote{spirituality,} as it is distinguished from religion, is belief in a spirit or life force without commitment to a belief in a God or gods (Davis et al., 2015; Hill et al., 2000).
+However, the term \enquote{spirituality} overlaps with the uses of \enquote{religion} in the sense that God or gods are instances of beliefs in spirits -- that is one may be both religious and spiritual.
 Here we operationalize religious belief as either belief in a God or belief in both a God and a spirit or life force.
 We operationalize spiritual belief as either belief in a spirit or life force without any corresponding belief in a God.
-However, the term "spirituality" overlaps with the uses of "religion" in the sense that God or gods are instances of beliefs in spirits – that is one may be both religious and spiritual.
+However, the term \enquote{spirituality} overlaps with the uses of \enquote{religion} in the sense that God or gods are instances of beliefs in spirits -- that is one may be both religious and spiritual.
 Here we operationalize religious belief as either belief in a God or belief in both a God and a spirit or life force.
 isNotably, previous spirituality mental health research has been disproportionately conducted on North American samples.
-In the United States, especially, samples include a majority of people who believe in both a God as well as and life forces other than God [@Smith2003-re; @Koenig2010-gk].
-Although such studies drawing from such samples mightmay clarify the effects of religious spirituality on psychological subjective well-being, the question of how non-religious spirituality affects psychological well-being that is – or spiritual beliefs without any belief in a God or gods – affects subjective well-being remain unclear.
-(For discussion, see: Therefore, to better understand possible distinctions between religious spirituality and non-religious spirituality for subjective well-being outcomes, samples collected from cultures outside the North Atlantic that contain a more equal ratio of the two are needed [@Garssen2016-km]).
+In the United States, especially, samples include a majority of people who believe in both a God as well as and life forces other than God (Koenig, 2010; Smith et al., 2003).
+Although such studies drawing from such samples mightmay clarify the effects of religious spirituality on psychological subjective well-being, the question of how non-religious spirituality affects psychological well-being that is -- or spiritual beliefs without any belief in a God or gods -- affects subjective well-being remain unclear.
+(For discussion, see: Therefore, to better understand possible distinctions between religious spirituality and non-religious spirituality for subjective well-being outcomes, samples collected from cultures outside the North Atlantic that contain a more equal ratio of the two are needed (Garssen \& Visser, 2016)).
 
 Third, most research investigating the relationship between spirituality and subjective well-being has relied on cross-sectional samples.
-These studies include high-quality reviews and meta-analyses [@Ano2005-hx; @Hackney2003-rs; @Sawatzky2005-rw; @Smith2003-re; @Visser2010-kq; @Yonker2012-zg].
+These studies include high-quality reviews and meta-analyses (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky, Ratner, \& Chiu, 2005; Smith et al., 2003; Visser, Garssen, \& Vingerhoets, 2010; Yonker, Schnabelrauch, \& Dehaan, 2012).
 However, to assess whetherto whether spiritual beliefs lead to a growth, stability, or decline in psychological well-being relative to people who lack such beliefs , including people who believe in God but not a spirit or lifeforce, requires repeatedly measuring the same individuals over time.
-clarify causal effects requires measuring the same individuals over time and assessing how life turns out for demographically similar people who differ in their spiritual and religious beliefs over their lifespans  and orientations [@Garssen2016-km].
+clarify causal effects requires measuring the same individuals over time and assessing how life turns out for demographically similar people who differ in their spiritual and religious beliefs over their lifespans and orientations (Garssen \& Visser, 2016).
 
-Our study addresses these three limitations in previous research.improves on past methods the following ways: First, we avoid a tautological elision of "spirituality" and "well-being"y by adopting a non-affective measure of spirituality through use of general, non-affective measures: "Do you believe in some form of a spirit or life force?" Our measures of psychological well being are [BEN SAY HERE] and Second we adopt a four-level categorical indicator that disentangles spiritual and religious beliefs: (1) Skeptics: those who neither believe in a God nor a life force; (2) SpiritExcludeGod: those who no not believe in a God  but believe in a spirit or life force; (3) GodAndSpirit: those who believe in a God and spirit; (4) GodExcludesSpirit: those who believe in both a God and a spirit or life force "Do you believe in a god?".
-Second, we address the prior dependence on cross-sectional samples taken from the United States by analyzing data taken from By sampling from respondents in New Zealand, a country with [GIVE NON_RELIGIOUS SPIRITUAL STATS HERE).
+Our study addresses these three limitations in previous research.improves on past methods the following ways: First, we avoid a tautological elision of \enquote{spirituality} and \enquote{well-being}y by adopting a non-affective measure of spirituality through use of general, non-affective measures: \enquote{Do you believe in some form of a spirit or life force?} Our measures of psychological well being are {[}BEN SAY HERE{]} and Second we adopt a four-level categorical indicator that disentangles spiritual and religious beliefs: (1) Skeptics: those who neither believe in a God nor a life force; (2) SpiritExcludeGod: those who no not believe in a God but believe in a spirit or life force; (3) GodAndSpirit: those who believe in a God and spirit; (4) GodExcludesSpirit: those who believe in both a God and a spirit or life force \enquote{Do you believe in a god?}.
+Second, we address the prior dependence on cross-sectional samples taken from the United States by analyzing data taken from By sampling from respondents in New Zealand, a country with {[}GIVE NON\_RELIGIOUS SPIRITUAL STATS HERE).
 Third, we are able to assess whether spirit beliefs are associate with a growth in psychological wellbeing address causation by weusing we use longitudinal responses from athis nationally diverse probability sample year over an nine-yearseight-year period with nine measurement points to assess whether spirit beliefs lead to growth, stability, or decline in psychological well-beingwellbeing (2010 to 2018 New Zealand Attitudes and Values Study, NZAVS).
 
-# Method
+\hypertarget{method}{%
+\section{Method}\label{method}}
 
-<!-- We rigorously assess the relationship between our non-affective measures and two constructs of subjective well-being (personal well-being and life satisfaction), adjusting for known demographic and personal influences, including age, sex, deprivation, and political orientation (Okulicz-Kozaryn, Holmes, and Avery 2014; Cuesta and Budría 2012; Røysamb et al.
-2002; Hansen and Slagsvold 2012; Chen and van Ours, n.d.; Vaughan et al.
-1985; Easterlin, Angelescu, and Zweig 2011; Stam et al.
-2016; Yakovlev and Leguizamon 2012).
-Collectively, then, our approach avoids the limitations of many previous studies and helps to clarify existing literature concerning how best to study the relationship between spirituality and subjective well-being.
--->
-
-## Ethics Statement
+\hypertarget{ethics-statement}{%
+\subsection{Ethics Statement}\label{ethics-statement}}
 
@@
 
-## Sampling Procedure
+\hypertarget{sampling-procedure}{%
+\subsection{Sampling Procedure}\label{sampling-procedure}}
 
@@
 Participants who provided an email address were also emailed and invited to complete an online version if they preferred.
-We offered a prize draw for participation, non-respondents were emailed and phoned multiple times, and all participants were mailed a Season’s Greetings card from the NZAVS research team and informed that they had been automatically entered into a bonus seasonal grocery voucher prize draw.
+We offered a prize draw for participation, non-respondents were emailed and phoned multiple times, and all participants were mailed a Season's Greetings card from the NZAVS research team and informed that they had been automatically entered into a bonus seasonal grocery voucher prize draw.
 We also mailed our yearly pamphlet summarizing key research findings published during the current wave of the study.
 
-## Participants
+\hypertarget{participants}{%
+\subsection{Participants}\label{participants}}
 
@@
 
-## Measures
+\hypertarget{measures}{%
+\subsection{Measures}\label{measures}}
 
-**Life Satisfaction** Life satisfaction is a measure of emotional well-being that was assessed using a 2-item version of the Satisfaction With Life Scale, which has previously been shown to correlate with aspects of religiosity [@Diener1985-xy].
-Participants rated their agreement with the statements (a) "I am satisfied with my life"; and (b) "In most ways my life is close to ideal." The items were rated on 7-point response options ranging from 1 = strongly disagree to 7 = strongly agree.
-The mean Cronbach’s alpha for this 2-item scale was $\alpha$ = 0.90.
+\textbf{Life Satisfaction} Life satisfaction is a measure of emotional well-being that was assessed using a 2-item version of the Satisfaction With Life Scale, which has previously been shown to correlate with aspects of religiosity (Diener, Emmons, Larsen, \& Griffin, 1985).
+Participants rated their agreement with the statements (a) \enquote{I am satisfied with my life}; and (b) \enquote{In most ways my life is close to ideal.} The items were rated on 7-point response options ranging from 1 = strongly disagree to 7 = strongly agree.
+The mean Cronbach's alpha for this 2-item scale was \(\alpha\) = 0.90.
 Higher scores on this scale indicate higher life satisfaction.
@@
 
-**Personal well-being** Personal well-being is a measure of function all well-being that was assessed using a 4-item version of the Australian Unity well-being Index (Cummins et al.
-2017).
-Participants rated their satisfaction with (a) "Your standard of living"; (b) "Your health"; (c) "Your future security"; (d) "Your personal relationships." Items were rated using a 10-point response option ranging from 1 = completely dissatisfied to 10 = completely satisfied.
-The mean Cronbach’s alpha for this scale was $\alpha$ = 0.90.
+\textbf{Personal well-being} Personal well-being is a measure of function all well-being that was assessed using a 4-item version of the Australian Unity well-being Index (Cummins et al.~2017).
+Participants rated their satisfaction with (a) \enquote{Your standard of living}; (b) \enquote{Your health}; (c) \enquote{Your future security}; (d) \enquote{Your personal relationships.} Items were rated using a 10-point response option ranging from 1 = completely dissatisfied to 10 = completely satisfied.
+The mean Cronbach's alpha for this scale was \(\alpha\) = 0.90.
 Higher scores on this scale indicate higher personal well-being.
@@
 
-***Belief in a Spirit or a God***  We assess belief in a spirit or a God by asking two non-affective questions: "Do you believe in a spirit or life force" and "Do you believe in a god?" Responses were coded as "Belief in a spirit that excludes a god" for those who indicated a belief in both a spirit or life force and a god, "Belief that includes a god" for those who indicated a belief in a god with or without an accompanying belief in a spirit or life force, and "skeptics" for those who indicated belief in neither a spirit or life force nor a god.
+\textbf{\emph{Belief in a Spirit or a God}} We assess belief in a spirit or a God by asking two non-affective questions: \enquote{Do you believe in a spirit or life force} and \enquote{Do you believe in a god?} Responses were coded as \enquote{Belief in a spirit that excludes a god} for those who indicated a belief in both a spirit or life force and a god, \enquote{Belief that includes a god} for those who indicated a belief in a god with or without an accompanying belief in a spirit or life force, and \enquote{skeptics} for those who indicated belief in neither a spirit or life force nor a god.
 These items appear in the NZAVS questionnaire and were developed for the 2005 Eurobarometer.
@@
 
-## Demographic Indicators
+\hypertarget{demographic-indicators}{%
+\subsection{Demographic Indicators}\label{demographic-indicators}}
 
-**Age** Age was put into units of 10 years and centered at its mean. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 2.
+\textbf{Age} Age was put into units of 10 years and centered at its mean. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 2.
 Gender.
-Gender was assessed by asking participants if they were "Male" was coded as "1" and "Female" was coded as "0." For detailed information pertaining to yearly responses and missingness, see Table 1.
+Gender was assessed by asking participants if they were \enquote{Male} was coded as \enquote{1} and \enquote{Female} was coded as \enquote{0.} For detailed information pertaining to yearly responses and missingness, see Table 1.
 
-**Education**  Education level was measured using an 11-point rating developed by the New Zealand Qualification Authority known as the New Zealand Qualification Framework (NZQF; 0 = no qualification, 10 = doctoral degree).
+\textbf{Education} Education level was measured using an 11-point rating developed by the New Zealand Qualification Authority known as the New Zealand Qualification Framework (NZQF; 0 = no qualification, 10 = doctoral degree).
 Education was centered at its mean and standardized.
@@
 
-**Deprivation** We measured the socio-economic status of participants’ immediate (small area) neighborhood using the 2013 New Zealand Deprivation Index, which uses aggregate census information about the residents of small neighborhood-type units to assign a decile-rank index from 1 (most affluent) to 10 (most impoverished) [@Atkinson2014-ex].
+\textbf{Deprivation} We measured the socio-economic status of participants' immediate (small area) neighborhood using the 2013 New Zealand Deprivation Index, which uses aggregate census information about the residents of small neighborhood-type units to assign a decile-rank index from 1 (most affluent) to 10 (most impoverished) (Atkinson, Salmond, \& Crampton, 2014).
 The index is based on a Principal Components Analysis of the following nine variables (in weighted order): the proportion of adults who received a means-tested benefit, household income, proportion not owning own home, proportion single-parent families, proportion unemployed, proportion lacking qualifications, proportion household crowding, proportion no telephone access, and proportion no car access.
-Thus, the index reflects the average level of deprivation for small neighborhood-type units (or small community areas of about 80–90 people each) across the entire country.
+Thus, the index reflects the average level of deprivation for small neighborhood-type units (or small community areas of about 80--90 people each) across the entire country.
 Our sample had a mean deprivation index of 4.80 (SD = 2.79).
@@
 
-**Employed** Employment status was assessed by asking participants if they were currently working, "yes" was coded as "1" and "no" was coded as "0." For detailed information pertaining to yearly responses and missingness, see Table 1.
+\textbf{Employed} Employment status was assessed by asking participants if they were currently working, \enquote{yes} was coded as \enquote{1} and \enquote{no} was coded as \enquote{0.} For detailed information pertaining to yearly responses and missingness, see Table 1.
 
-**Partner** Participants were asked if they were in a relationship, "yes" was coded as "1" and no was coded as "0." For detailed information pertaining to yearly responses and missingness, see Table 1.
+\textbf{Partner} Participants were asked if they were in a relationship, \enquote{yes} was coded as \enquote{1} and no was coded as \enquote{0.} For detailed information pertaining to yearly responses and missingness, see Table 1.
 
-**Ethnicity** Ethnicity was assessed using four basic categories: (1) New Zealand European/Pakeha, (2) Maori, (3) Pacific Islander, and (4) Asian.
+\textbf{Ethnicity} Ethnicity was assessed using four basic categories: (1) New Zealand European/Pakeha, (2) Maori, (3) Pacific Islander, and (4) Asian.
 For detailed information pertaining to yearly responses and missingness, see Table 1.
 
-**Urban** People were coded as either residing in an urban "1" or rural "0" area based on New Zealand census data. For detailed information pertaining to yearly responses and missingness, see Table 1.
+\textbf{Urban} People were coded as either residing in an urban \enquote{1} or rural \enquote{0} area based on New Zealand census data.
+For detailed information pertaining to yearly responses and missingness, see Table 1.
 
-**Political Orientation** To assess political orientation, we asked people to rate their political orientation using seven-point response options (1 = Liberal; 7 = Conservative).
+\textbf{Political Orientation} To assess political orientation, we asked people to rate their political orientation using seven-point response options (1 = Liberal; 7 = Conservative).
 Higher values indicate more conservative political beliefs.
@@
 
-## Statistical Analyses
+\hypertarget{statistical-analyses}{%
+\subsection{Statistical Analyses}\label{statistical-analyses}}
 
@@
 To model predict growth in levels of self-reported psychological subjective well-being among these participants we use generalized linear mixed models.
-In our models, we account for the non-linear effects of time within participants by including its interaction with both of our predictor variables: "Belief in a spirit that excludes god" and "Belief that includes a god." To handle the dependencies and heterogeneity introduced from the repeated measures, we We also included individual ID as an effect modelled as random-effect.
+In our models, we account for the non-linear effects of time within participants by including its interaction with both of our predictor variables: \enquote{Belief in a spirit that excludes god} and \enquote{Belief that includes a god.} To handle the dependencies and heterogeneity introduced from the repeated measures, we We also included individual ID as an effect modelled as random-effect.
 To improve interpretability of our models, we centered and scaled the education, deprivation, and political orientation covariates.
@@
 
-Because missing values can result in biased estimates [@Blackwell2017-oq], we multiply imputed 10 datasets using the AmeliaII package in R[@Honaker2011-yu].
+Because missing values can result in biased estimates (Blackwell, Honaker, \& King, 2017), we multiply imputed 10 datasets using the AmeliaII package in R(Honaker, King, Blackwell, \& Others, 2011).
 We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences.
 
-## Code
+\hypertarget{code}{%
+\subsection{Code}\label{code}}
 
-All code for this analysis and a full report of the missing data imputation can be found at https://github.com/jacopastorius/NZAVS_Highland_SpiritHealth2.Table 1.
+All code for this analysis and a full report of the missing data imputation can be found at \url{https://github.com/jacopastorius/NZAVS_Highland_SpiritHealth2.Table} 1.
 Descriptive Statistics for Theoretical Variables in Each Wave.
 Belief in a god and belief in a spirit or life force were not measured during Wave 6 (2014).
-"BXG" indicates a belief in a spirit that excludes a god.
-"BIG" indicates a belief that includes a god.
+\enquote{BXG} indicates a belief in a spirit that excludes a god.
+\enquote{BIG} indicates a belief that includes a god.
 Many individuals appear in multiple waves.
@@
 
-# Results
+\hypertarget{results}{%
+\section{Results}\label{results}}
 
@@
 
-```{r, cache =TRUE, include =FALSE}
-# model 2
-mod.1 <-lmer (PWI ~  1 +  Years * Beliefs +  Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id), data= dfben, na.action = na.omit)
-```
-
-```{r, cache=TRUE, include=FALSE}
-# model 1
-mod.1 <-lmer (LIFESAT ~ 1 + Years * Beliefs+ Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id), data= dfben, na.action = na.omit)
-summary(mod.1)
-```
-
 Results for both the personal well-being and life satisfaction models are presented in Table 2 and Figure 1.
 
-```{r fig.cap = "Combination Coefficient Plots for Personal Well-being and Life Satisfaction Spirit Belief Longitudinal Models"}
-knitr::include_graphics("figs/USEcoefficientplots-1.png")
-```
+\begin{figure}
+
+{\centering \includegraphics[width=6.4in]{figs/USEcoefficientplots-1}
+
+}
+
+\caption{Combination Coefficient Plots for Personal Well-being and Life Satisfaction Spirit Belief Longitudinal Models}\label{fig:unnamed-chunk-5}
+\end{figure}
 
@@
 
-```{r, echo = FALSE, include=FALSE}
-te00 <-texreg::extract(
-  mod.1,
-  level = 0.95,
-  include.random = TRUE,
-  include.rsquared = T,
-  include.nobs = T,
-  include.loo.ic = T,
-  include.waic = F)
-texreg(list(te00),
-                custom.model.names = c("PWI"),
-                caption = "Personal Wellbeing",
-                sideways = F,
-                scalebox = .5,
-                #fontsize= "footnotesize",
-                label = "tab:REGRESS_1",
-                ci.force.level = 0.95, bold = 0.05,
-                settingstars = 0,
-                booktabs = TRUE,
-                custom.note ="")
-```
-
-## Personal Well-Being
+\hypertarget{personal-well-being}{%
+\subsection{Personal Well-Being}\label{personal-well-being}}
 
@@
 The model included Id as random effects.
-The model's intercept, corresponding to Years = 0, Beliefs = Skeptic, Age = (sample mean), Deprivation.S = (sample mean), Edu.S =(sample mean), Employed = Not Employed, EthnicCats = Euro, Male = Not Mail, Partner = No Partner, Pol.Orient.S = (sample mean),, Urban = Not Urban and Id = 2, is at 4.94 (SE = 0.02, 95% CI [4.89, 4.99], p < .001) is at 6.63 (SE = 0.03, 95% CI [6.57, 6.70], p < .001).
+The model's intercept, corresponding to Years = 0, Beliefs = Skeptic, Age = (sample mean), Deprivation.S = (sample mean), Edu.S =(sample mean), Employed = Not Employed, EthnicCats = Euro, Male = Not Mail, Partner = No Partner, Pol.Orient.S = (sample mean),, Urban = Not Urban and Id = 2, is at 4.94 (SE = 0.02, 95\% CI {[}4.89, 4.99{]}, p \textless{} .001) is at 6.63 (SE = 0.03, 95\% CI {[}6.57, 6.70{]}, p \textless{} .001).
 Within this model:
 
-  - The effect of Years is positive is not reliable (beta = 2.66e-03, SE = 3.58e-03, std.
+\begin{itemize}
+\tightlist
+\item
+  The effect of Years is positive is not reliable (beta = 2.66e-03, SE = 3.58e-03, std.
 beta = 3.19e-03, p = 0.458).
 We infer that Skeptics personal well-being is not increasing in the Skeptic population
-  - The effect of Beliefs_SpiritExcludesGod_is reliably negative (beta = -0.14, SE = 0.03, std.
-beta = -0.03, p < .001).
+\item
+  The effect of Beliefs\_SpiritExcludesGod\_is reliably negative (beta = -0.14, SE = 0.03, std.
+beta = -0.03, p \textless{} .001).
 We infer that overall the population that believes in a Spirit but does not believe in a God is expected to have a lower personal well-being than skeptics.
-  - The effect of BeliefsGodAndSpirit is reliably negative (beta = -0.16, SE = 0.03, std.
-beta = -0.02, p < .001).
+\item
+  The effect of BeliefsGodAndSpirit is reliably negative (beta = -0.16, SE = 0.03, std.
+beta = -0.02, p \textless{} .001).
 We infer that overall the population that believes in God and a spirit or life force is expected to have a lower Personal Well-being than skeptics.
-  - The effect of BeliefsGodExcludesSpirit is not reliably different from the skeptic population (beta = 0.02, SE = 0.05, std. beta = 0.02, p = 0.669).
-  - The effect of Age.10yrs.C is reliably positive  (beta = 0.08, SE = 7.38e-03, std. beta = 0.07, p < .001). Older people have greater personal well-beingwellbeing.
-  - The effect of Deprivation.S is reliably negative (beta = -0.14, SE = 7.50e-03, std. beta = -0.08, p < .001). Deprived people have lower personal well-beingwellbeing.
-  - The effect of Edu.S is reliably positive  (beta = 0.12, SE = 8.74e-03, std. beta = 0.07, p < .001). Educated people greater personal well-beingwellbeing
-  - The effect of Employed is reliably positive  (beta = 0.09, SE = 0.01, std. beta = 0.05, p < .001). Employed people have greater personal well-beingwellbeing.
-  - The effect of EthnicCatsMaori is reliably negative (beta = -0.30, SE = 0.03, std. beta = -0.18, p < .001).
-  - The effect of EthnicCatsPacific is reliably negative (beta = -0.29, SE = 0.05, std. beta = -0.17, p < .001).
-  - The effect of EthnicCatsAsian is not reliable (beta = -0.08, SE = 0.05, std. beta = -0.05, p = 0.089).
-  - The effect of Male is reliably negative (beta = -0.11, SE = 0.02, std. beta = -0.06, p < .001).
-  - The effect of Having a Partner is reliably positive (beta = 0.61, SE = 0.02, std. beta = 0.37, p < .001).
-  - The effect of Pol.Orient.S is reliably positive  (beta = 0.02, SE = 5.67e-03, std. beta = 0.01, p < .001).
-  - The effect of Urban is reliably negative (beta = -0.02, SE = 0.01, std. beta = -0.01, p < .05).
+\item
+  The effect of BeliefsGodExcludesSpirit is not reliably different from the skeptic population (beta = 0.02, SE = 0.05, std. beta = 0.02, p = 0.669).
+\item
+  The effect of Age.10yrs.C is reliably positive (beta = 0.08, SE = 7.38e-03, std. beta = 0.07, p \textless{} .001). Older people have greater personal well-beingwellbeing.
+\item
+  The effect of Deprivation.S is reliably negative (beta = -0.14, SE = 7.50e-03, std. beta = -0.08, p \textless{} .001). Deprived people have lower personal well-beingwellbeing.
+\item
+  The effect of Edu.S is reliably positive (beta = 0.12, SE = 8.74e-03, std. beta = 0.07, p \textless{} .001). Educated people greater personal well-beingwellbeing
+\item
+  The effect of Employed is reliably positive (beta = 0.09, SE = 0.01, std. beta = 0.05, p \textless{} .001). Employed people have greater personal well-beingwellbeing.
+\item
+  The effect of EthnicCatsMaori is reliably negative (beta = -0.30, SE = 0.03, std. beta = -0.18, p \textless{} .001).
+\item
+  The effect of EthnicCatsPacific is reliably negative (beta = -0.29, SE = 0.05, std. beta = -0.17, p \textless{} .001).
+\item
+  The effect of EthnicCatsAsian is not reliable (beta = -0.08, SE = 0.05, std. beta = -0.05, p = 0.089).
+\item
+  The effect of Male is reliably negative (beta = -0.11, SE = 0.02, std. beta = -0.06, p \textless{} .001).
+\item
+  The effect of Having a Partner is reliably positive (beta = 0.61, SE = 0.02, std. beta = 0.37, p \textless{} .001).
+\item
+  The effect of Pol.Orient.S is reliably positive (beta = 0.02, SE = 5.67e-03, std. beta = 0.01, p \textless{} .001).
+\item
+  The effect of Urban is reliably negative (beta = -0.02, SE = 0.01, std. beta = -0.01, p \textless{} .05).
+\end{itemize}
 
-**Interactions**
+\textbf{Interactions}
 
-```{r fig.cap = "Combination Coefficient Plots for Personal Well-being and Life Satisfaction Spirit Belief Longitudinal Models"}
-knitr::include_graphics("Figs/USEPWI_expected-1.png")
-```
-  - The effect of Years X Beliefs_SpiritExcludesGod_ is reliably positive  (beta = 0.02, SE = 4.86e-03, std.
-beta = 0.02, p < .01).
+\begin{figure}
+
+{\centering \includegraphics[width=6.4in]{Figs/USEPWI_expected-1}
+
+}
+
+\caption{Combination Coefficient Plots for Personal Well-being and Life Satisfaction Spirit Belief Longitudinal Models}\label{fig:unnamed-chunk-7}
+\end{figure}
+
+\begin{itemize}
+\tightlist
+\item
+  The effect of Years X Beliefs\_SpiritExcludesGod\_ is reliably positive (beta = 0.02, SE = 4.86e-03, std.
+beta = 0.02, p \textless{} .01).
 We infer that, unlike the skeptic population, people with beliefs in a spirit or life force who do not believe in a God tend to grow in their personal well-beingwellbeing over time.
-  - The effect of Years X BeliefsGodAndSpirit is reliably positive  (beta = 0.02, SE = 4.41e-03, std.
-beta = 0.03, p < .001).
+
+  \begin{itemize}
+  \tightlist
+  \item
+    The effect of Years X BeliefsGodAndSpirit is reliably positive (beta = 0.02, SE = 4.41e-03, std.
+beta = 0.03, p \textless{} .001).
 We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their personal well-beingwellbeing over time
-  - The effect of Years X BeliefsGodExcludesSpirit is not reliable(beta = 1.96e-03, SE = 8.29e-03, std.
+  \item
+    The effect of Years X BeliefsGodExcludesSpirit is not reliable(beta = 1.96e-03, SE = 8.29e-03, std.
 beta = 2.36e-03, p = 0.813).
 Although people who believe in a God but do not believe in a spirit or Life Force tend to have greater personal well-beingwellbeing than those who do not believe in a spirit or life force (and greater life satisfaction than Skeptics) this population is not expected to grow in Life Satisfaction over time.
+  \end{itemize}
+\end{itemize}
 
-## Life Satisfaction
+\hypertarget{life-satisfaction}{%
+\subsection{Life Satisfaction}\label{life-satisfaction}}
 
@@
 Standardized parameters were obtained by fitting the model on a standardized version of the dataset.The model's total explanatory power is substantial (conditional R2 = 0.69).
-The model's intercept, corresponding to Years = 0, Beliefs = Skeptic, Age = (sample mean),, Deprivation.S = (sample mean), Edu.S =(sample mean), Employed = Not Employed, EthnicCats = Euro, Male = Not Mail, Partner = No Partner, Pol.Orient.S = (sample mean), Urban = Not Urban and Id = 2, is at 4.94 (SE = 0.02, 95% CI [4.89, 4.99], p < .001).
+The model's intercept, corresponding to Years = 0, Beliefs = Skeptic, Age = (sample mean),, Deprivation.S = (sample mean), Edu.S =(sample mean), Employed = Not Employed, EthnicCats = Euro, Male = Not Mail, Partner = No Partner, Pol.Orient.S = (sample mean), Urban = Not Urban and Id = 2, is at 4.94 (SE = 0.02, 95\% CI {[}4.89, 4.99{]}, p \textless{} .001).
 Within this model:
 
-  - The effect of Years is not reliable (beta = -5.50e-04, SE = 2.69e-03, std. beta = -9.32e-04, p = 0.838).  We do not infer growth over time in Life Satisfaction among the Skeptic population
-  - The effect of BeliefsSpiritExcludesGod  is reliably negative (beta = -0.08, SE = 0.02, std.
-beta = 0.01, p < .001).
+\begin{itemize}
+\tightlist
+\item
+  The effect of Years is not reliable (beta = -5.50e-04, SE = 2.69e-03, std. beta = -9.32e-04, p = 0.838). We do not infer growth over time in Life Satisfaction among the Skeptic population
+\item
+  The effect of BeliefsSpiritExcludesGod is reliably negative (beta = -0.08, SE = 0.02, std.
+beta = 0.01, p \textless{} .001).
 We infer that overall the population that believes in a Spirit but does not believe in a God is expected to have a lower Life Satisfaction than skeptics.
-  - The effect of BeliefsGodAndSpirit is not reliable (beta = -0.04, SE = 0.02, std. beta = 0.04, p = 0.105). We do not infer growth over time in Life Satisfaction among this (Skeptic) population
-  - The effect of BeliefsGodExcludesSpirit is reliably positive (beta = 0.07, SE = 0.04, std.
+\item
+  The effect of BeliefsGodAndSpirit is not reliable (beta = -0.04, SE = 0.02, std. beta = 0.04, p = 0.105). We do not infer growth over time in Life Satisfaction among this (Skeptic) population
+\item
+  The effect of BeliefsGodExcludesSpirit is reliably positive (beta = 0.07, SE = 0.04, std.
 beta = 0.05, p = 0.062).We infer that overall the population that believes in a God but does not believe in a spirit is not reliably different from the Skeptics population in Life Satisfaction.
-  - The effect of Age.10yrs.C is reliably positive (beta = 0.09, SE = 5.26e-03, std. beta = 0.10, p < .001). Older people have greater Life Satisfaction
-  - The effect of Deprivation.S is  is reliably negative (beta = -0.07, SE = 5.50e-03, std. beta = -0.06, p < .001). People who are more deprived have lower Life Satisfaction.
-  - The effect of Edu.S is reliably positive (beta = 0.06, SE = 6.33e-03, std. beta = 0.05, p < .001). Educated people have greater Life Satisfaction
-  - The effect of Employed is reliably positive (beta = 0.08, SE = 9.69e-03, std. beta = 0.07, p < .001). People who are employed tend to have greater Life Satisfaction
-  - The effect of EthnicCatsMaori is reliably negative (beta = -0.06, SE = 0.02, std. beta = -0.05, p < .01).  People who belong to the minority Maori group tend to have lower Life Satisfaction.
-  - The effect of EthnicCatsPacific is reliably negative (beta = -0.08, SE = 0.04, std.
-beta = -0.07, p < .05).
+\item
+  The effect of Age.10yrs.C is reliably positive (beta = 0.09, SE = 5.26e-03, std. beta = 0.10, p \textless{} .001). Older people have greater Life Satisfaction
+\item
+  The effect of Deprivation.S is is reliably negative (beta = -0.07, SE = 5.50e-03, std. beta = -0.06, p \textless{} .001). People who are more deprived have lower Life Satisfaction.
+\item
+  The effect of Edu.S is reliably positive (beta = 0.06, SE = 6.33e-03, std. beta = 0.05, p \textless{} .001). Educated people have greater Life Satisfaction
+\item
+  The effect of Employed is reliably positive (beta = 0.08, SE = 9.69e-03, std. beta = 0.07, p \textless{} .001). People who are employed tend to have greater Life Satisfaction
+\item
+  The effect of EthnicCatsMaori is reliably negative (beta = -0.06, SE = 0.02, std.
+beta = -0.05, p \textless{} .01).
+People who belong to the minority Maori group tend to have lower Life Satisfaction.
+\item
+  The effect of EthnicCatsPacific is reliably negative (beta = -0.08, SE = 0.04, std.
+beta = -0.07, p \textless{} .05).
 People who belong to the minority Pacific peoples group tend to have lower life satisfaction.
-  - The effect of EthnicCatsAsian is not reliably different from people of European ethnicity in life satisfaction(beta = -0.04, SE = 0.03, std. beta = -0.03, p = 0.223).
-  - The effect of Male is reliably negative (beta = -0.19, SE = 0.01, std. beta = -0.16, p < .001).
-  - The effect of HasPartner is reliably positive (beta = 0.42, SE = 0.01, std. beta = 0.35, p < .001).
-  - The effect of Pol.Orient.S is reliably positive (beta = 0.01, SE = 4.23e-03, std. beta = 0.01, p < .01).
-  - The effect of Urban is reliably negative (beta = -0.02, SE = 8.85e-03, std. beta = -0.02, p < .05).
+\item
+  The effect of EthnicCatsAsian is not reliably different from people of European ethnicity in life satisfaction(beta = -0.04, SE = 0.03, std. beta = -0.03, p = 0.223).
+\item
+  The effect of Male is reliably negative (beta = -0.19, SE = 0.01, std. beta = -0.16, p \textless{} .001).
+\item
+  The effect of HasPartner is reliably positive (beta = 0.42, SE = 0.01, std. beta = 0.35, p \textless{} .001).
+\item
+  The effect of Pol.Orient.S is reliably positive (beta = 0.01, SE = 4.23e-03, std. beta = 0.01, p \textless{} .01).
+\item
+  The effect of Urban is reliably negative (beta = -0.02, SE = 8.85e-03, std. beta = -0.02, p \textless{} .05).
+\end{itemize}
 
-**Interactions**
+\textbf{Interactions}
 
-```{r fig.cap = "Prediction Plot Life Satisfaction Spirit Belief Longitudinal Models"}
-knitr::include_graphics("Figs/USElifsat_expected-1.png")
-```
-  - The interaction of Years X Beliefs_SpiritExcludesGod_ is reliably positive (beta = 0.02, SE = 3.66e-03, std.
-beta = 0.03, p < .001).
+\begin{figure}
+
+{\centering \includegraphics[width=6.4in]{Figs/USElifsat_expected-1}
+
+}
+
+\caption{Prediction Plot Life Satisfaction Spirit Belief Longitudinal Models}\label{fig:unnamed-chunk-8}
+\end{figure}
+
+\begin{itemize}
+\tightlist
+\item
+  The interaction of Years X Beliefs\_SpiritExcludesGod\_ is reliably positive (beta = 0.02, SE = 3.66e-03, std.
+beta = 0.03, p \textless{} .001).
 We infer that, unlike the Skeptic population, people with beliefs in a Spirit or Life Force who do not believe in a God tend to grow in their Life Satisfaction over time.
-  - The interaction of Years X BeliefsGodAndSpirit is reliably positive (beta = 0.01, SE = 3.32e-03, std.
-beta = 0.02, p < .001).
+
+  \begin{itemize}
+  \tightlist
+  \item
+    The interaction of Years X BeliefsGodAndSpirit is reliably positive (beta = 0.01, SE = 3.32e-03, std.
+beta = 0.02, p \textless{} .001).
 We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their Life Satisfaction over time
-  - The effect of Years X BeliefsGodExcludesSpirit is not reliable (beta = -1.59e-03, SE = 6.25e-03, std.
+  \item
+    The effect of Years X BeliefsGodExcludesSpirit is not reliable (beta = -1.59e-03, SE = 6.25e-03, std.
 beta = -2.70e-03, p = 0.799).
 Although people who believe in a God but do not believe in a spirit or Life Force tend to have greater life satisfaction than those who do not believe in a Spirit or life force (and greater life satisfaction than Skeptics) this population is not expected to grow in Life Satisfaction over time.
-
-<!-- ```{r} -->
-<!-- knitr::include_graphics("Figs/Usecoefficientplots-1.png") -->
-<!-- ``` -->
+  \end{itemize}
+\end{itemize}
 
@@
 
-# Discussion
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
 
@@
 These findings demonstrate the importance ofinvestigating spiritual beliefs as a domain of interest in its own right.
-In many Western countries, people are increasingly identifying as "spiritual but not religious" [@Zinnbauer2005-vz; @Swinton2001-vr].
-Other studies find that those who have a spiritual understanding of life without a religious framework are at an increased risk of having mental disorders [@King2013-cg].
+In many Western countries, people are increasingly identifying as \enquote{spiritual but not religious} (Swinton, 2001; Zinnbauer \& Pargament, 2005).
+Other studies find that those who have a spiritual understanding of life without a religious framework are at an increased risk of having mental disorders (King et al., 2013).
 As such, it is important to understand the ramifications of belief in a spirit for psychological health.
@@
 These longitudinal results persist independent of time.
-Notably, our findings do partially support other previous work that shows that spirituality outside of a religious framework is associated with worse emotional well-being, though we also find a negative association within a religious framework (King et al.
-2013).
+Notably, our findings do partially support other previous work that shows that spirituality outside of a religious framework is associated with worse emotional well-being, though we also find a negative association within a religious framework (King et al.~2013).
 Still, further research is needed to determine if other aspects of religion and spirituality, such as practices and rituals, might yield different results.
@@
 Despite the opposition of our findings to those of many previous studies, there is room for agreement.
-We show that over time both "BXG" and "BIG" are associated with marginally better subjective well-being.
-As seen in Figure 2., both "BXG" and "BIG" predict marginally higher personal well-being and life satisfaction as time progresses.
+We show that over time both \enquote{BXG} and \enquote{BIG} are associated with marginally better subjective well-being.
+As seen in Figure 2., both \enquote{BXG} and \enquote{BIG} predict marginally higher personal well-being and life satisfaction as time progresses.
 So, while those who adhere to these beliefs may start out below their skeptic counterparts in terms of their subjective well-being, they do eventually catch up.
 We also find that their life satisfaction is actually improving at a greater rate than skeptics.
-Thus, in the long term, both sustained "BXG" and "BIG" may actually be associated with better subjective well-being.
+Thus, in the long term, both sustained \enquote{BXG} and \enquote{BIG} may actually be associated with better subjective well-being.
 Our results also help to clarify the relationship among religion, spirituality, and subjective well-being outcomes.
@@
 
-## Limitations
+\hypertarget{limitations}{%
+\subsection{Limitations}\label{limitations}}
 
@@
 
-### Authors contributions
+\hypertarget{authors-contributions}{%
+\subsubsection{Authors contributions}\label{authors-contributions}}
+
 CS leads and manages the NZAVS, and collected the data for this study. BH and JB conceived
@@
 
-### Funding statement
+\hypertarget{funding-statement}{%
+\subsubsection{Funding statement}\label{funding-statement}}
+
 The New Zealand Attitudes and Values Study is supported by a grant from the Templeton
@@
 
-# Supplemental Material
+\hypertarget{supplemental-material}{%
+\section{Supplemental Material}\label{supplemental-material}}
+
 S1. Statistical Analysis in R
 Statistical analysis was performed using R version 3.6.1 (2019-07-05).
-Platform: x86_64-w64-mingw32 (64-bit), running under: Windows 10 (Action of the Toes).
-Generalized Linear Mixed Models were generated using the glmmTMB (Brooks et al.
-2017; Knowles and Frederick 2019) package in R.
-In addition, we used the following R packages: dplyr (Hadley Wickham et al.
-2019), tidyr (H.
+Platform: x86\_64-w64-mingw32 (64-bit), running under: Windows 10 (Action of the Toes).
+Generalized Linear Mixed Models were generated using the glmmTMB (Brooks et al.~2017; Knowles and Frederick 2019) package in R.
+In addition, we used the following R packages: dplyr (Hadley Wickham et al.~2019), tidyr (H.
 Wickham and Henry 2019), sjPlot (D.
@@
 Models for both subjective well-being variables used family = gaussian, assuming a normal distribution.
-We ran each individual’s unique ID as a random-effect.
-The tab_model and plot_model commands in sjPlot allowed for the creation of tables and coefficient plots.
+We ran each individual's unique ID as a random-effect.
+The tab\_model and plot\_model commands in sjPlot allowed for the creation of tables and coefficient plots.
 Predicted probability plots were generated using the ggpredict command in the ggeffects package.
@@
 
-# References
+\hypertarget{references}{%
+\section{References}\label{references}}
 
@@
 
-<div id="refs" custom-style="Bibliography"></div>
+\hypertarget{refs}{}
+\leavevmode\hypertarget{ref-Ano2005-hx}{}%
+Ano, G. G., \& Vasconcelles, E. B. (2005). Religious coping and psychological adjustment to stress: A meta‐analysis. \emph{J. Clin. Psychol.}
+
+\leavevmode\hypertarget{ref-Atkinson2014-ex}{}%
+Atkinson, J., Salmond, C., \& Crampton, P. (2014). NZDep2013 index of deprivation. \emph{Wellington: Department of Public Health, University of Otago}.
+
+\leavevmode\hypertarget{ref-Blackwell2017-oq}{}%
+Blackwell, M., Honaker, J., \& King, G. (2017). A unified approach to measurement error and missing data: Details and extensions. \emph{Sociological Methods \& Research}.
+
+\leavevmode\hypertarget{ref-Davis2015-tx}{}%
+Davis, D. E., Rice, K., Hook, J. N., Van Tongeren, D. R., DeBlaere, C., Choe, E., \& Worthington, E. L. (2015). Sources of spirituality scale. \emph{PsycTESTS Dataset}.
+
+\leavevmode\hypertarget{ref-Diener1985-xy}{}%
+Diener, E., Emmons, R. A., Larsen, R. J., \& Griffin, S. (1985). The satisfaction with life scale. \emph{Journal of Personality Assessment}.
+
+\leavevmode\hypertarget{ref-Ellsworth2010-yu}{}%
+Ellsworth, R. B., \& Ellsworth, J. B. (2010). Churches that enhance spirituality and wellbeing. \emph{Int. J. Appl. Psychoanal. Studies}, \emph{6}.
+
+\leavevmode\hypertarget{ref-Garssen2016-km}{}%
+Garssen, B., \& Visser, A. (2016). The association between religion/spirituality and mental health in cancer. \emph{Cancer}, \emph{122}(15), 2440.
+
+\leavevmode\hypertarget{ref-Garssen2016-kb}{}%
+Garssen, B., Visser, A., \& Jager Meezenbroek, E.
+de.
+(2016).
+Examining whether spirituality predicts subjective well-being: How to avoid tautology.
+\emph{Psycholog.
+Relig.
+Spiritual.}, \emph{8}(2), 141.
+
+\leavevmode\hypertarget{ref-Ginsburg1995-jr}{}%
+Ginsburg, M.
+L., Quirt, C., Ginsburg, A.
+D., \& MacKillop, W.
+J.
+(1995).
+Psychiatric illness and psychosocial concerns of patients with newly diagnosed lung cancer.
+\emph{CMAJ}, \emph{152}(5), 701--708.
+
+\leavevmode\hypertarget{ref-Hackney2003-rs}{}%
+Hackney, C. H., \& Sanders, G. S. (2003). Religiosity and mental health: A meta--analysis of recent studies. \emph{J. Sci. Study Relig.}
+
+\leavevmode\hypertarget{ref-Hill2000-cl}{}%
+Hill, P.
+C., Pargament, K.
+I.
+I., Hood, R.
+W., McCullough, M.
+E., Jr, Swyers, J.
+P., Larson, D.
+B., \& Zinnbauer, B.
+J.
+(2000).
+Conceptualizing religion and spirituality: Points of commonality, points of departure.
+\emph{J.
+Theory Soc.
+Behav.}, \emph{30}(1), 51--77.
+
+\leavevmode\hypertarget{ref-Honaker2011-yu}{}%
+Honaker, J., King, G., Blackwell, M., \& Others. (2011). Amelia II: A program for missing data. \emph{J. Stat. Softw.}, \emph{45}(7), 1--47.
+
+\leavevmode\hypertarget{ref-King2013-cg}{}%
+King, M., Marston, L., McManus, S., Brugha, T., Meltzer, H., \& Bebbington, P.
+(2013).
+Religion, spirituality and mental health: Results from a national study of english households.
+\emph{British Journal of Psychiatry}.
+
+\leavevmode\hypertarget{ref-Koenig2008-lv}{}%
+Koenig, H. G. (2008). Concerns about measuring ``spirituality'' in research. \emph{J. Nerv. Ment. Dis.}, \emph{196}(5), 349.
+
+\leavevmode\hypertarget{ref-Koenig2010-gk}{}%
+Koenig, H. G. (2010). Spirituality and mental health. \emph{Int. J. Appl. Psychoanal. Studies}, \emph{15}.
+
+\leavevmode\hypertarget{ref-Koenig2001-ow}{}%
+Koenig, H.
+G., Associate Professor of Psychology and Religious Studies Michael E McCullough, PhD, McCullough, M.
+E., Larson, D.
+B., \& Adjunct Professor of Psychiatry David B Larson.
+(2001).
+\emph{Handbook of religion and health}.
+Oxford University Press.
+
+\leavevmode\hypertarget{ref-Russinova2002-rq}{}%
+Russinova, Z., Wewiorski, N.
+J., \& Cash, D.
+(2002).
+Use of alternative health care practices by persons with serious mental illness: Perceived benefits.
+\emph{Am.
+J.
+Public Health}, \emph{92}(10), 1600--1603.
+
+\leavevmode\hypertarget{ref-Sawatzky2005-rw}{}%
+Sawatzky, R., Ratner, P. A., \& Chiu, L. (2005). A Meta-Analysis of the relationship between spirituality and quality of life. \emph{Soc. Indic. Res.}, \emph{72}(2), 153--188.
+
+\leavevmode\hypertarget{ref-Smith2003-re}{}%
+Smith, T.
+B., McCullough, M.
+E., \& Poll, J.
+(2003).
+Religiousness and depression: Evidence for a main effect and the moderating influence of stressful life events.
+\emph{Psychol.
+Bull.}, \emph{129}(4), 614--636.
+
+\leavevmode\hypertarget{ref-Swinton2001-vr}{}%
+Swinton, J. (2001). \emph{Spirituality and mental health care: Rediscovering a 'forgotten' dimension}. Jessica Kingsley Publishers.
+
+\leavevmode\hypertarget{ref-Underwood2002-hg}{}%
+Underwood, L.
+G., \& Teresi, J.
+A.
+(2002).
+The daily spiritual experience scale: Development, theoretical description, reliability, exploratory factor analysis, and preliminary construct validity using health-related data.
+\emph{Ann.
+Behav.
+Med.}, \emph{24}(1), 22--33.
+
+\leavevmode\hypertarget{ref-Visser2010-kq}{}%
+Visser, A., Garssen, B., \& Vingerhoets, A. (2010). Spirituality and well-being in cancer patients: A review. \emph{Psychooncology}, \emph{19}(6), 565--572.
+
+\leavevmode\hypertarget{ref-Yonker2012-zg}{}%
+Yonker, J.
+E., Schnabelrauch, C.
+A., \& Dehaan, L.
+G.
+(2012).
+The relationship between spirituality and religiosity on psychological outcomes in adolescents and emerging adults: A meta-analytic review.
+\emph{J.
+Adolesc.}, \emph{35}(2), 299--314.
+
+\leavevmode\hypertarget{ref-Zaza2005-ac}{}%
+Zaza, C., Sellick, S. M., \& Hillier, L. M. (2005). Coping with cancer. \emph{Journal of Psychosocial Oncology}.
+
+\leavevmode\hypertarget{ref-Zinnbauer2005-vz}{}%
+Zinnbauer, B. J., \& Pargament, K. I. (2005). Religiousness and spirituality {[}in:{]} Handbook of the psychology of religion and spirituality. \emph{New York}, 21--42.
+
 \endgroup
+
+\end{document}

### 2020-11-16 Manuscript.Rmd
Adding figures, updating manuscript (clarified introduction)

@@
     corresponding : yes    # Define only one corresponding author
-    address       : "Postal address"
-    email         : "my@email.com"
-    role:         # First author
-      - Conceptualization
-      - Literature Review
-      - Writing - First draft
-      - Writing - Review & editing
-  - name          : "Ev l Worthing"
+    email         : "benhighland6420@gmail.com"
+  - name          : "Everett L. Worthington"
     affiliation   : "2"
-    address       : "Postal address"
-    email         : "my@email.com"
-    role:        
-      - Writing review & editing
+    institution   : "Department of Psychology Virginia Commonwealth University"
   - name          : "Don E Davis"
     affiliation   : "3"
-    address       : "Postal address"
-    email         : "my@email.com"
-    role:         
-      - Writing review & editing
+    institution   : "Department of Psychology Georgia State University"
   - name          : "Chris G. Sibley"
     affiliation   : "4"
-    address       : "Postal address"
-    email         : "my@email.com"
-    role:        
-      - Writing review & editing
+    institution   : "School of Psychology University of Auckland"
   - name          : "Joseph Bulbulia"
     affiliation   : "5"
-    role:         
-      - Data analysis
-      - Writing second and subsequent drafts
-      - Writing review & editing
+    institution   : "School of Psychology Victoria University"
 
 authornote: |
-  Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.
-  Enter author note here.
+  1 Wake Forest Medical School 2 Department of Psychology Virginia Commonwealth University 3 Department of Psychology Georgia State University 4 School of Psychology University of Auckland 5 School of Psychology Victoria University
+
 abstract: |
  Previous research finds an association between spirituality and subjective well-being.
-However, the widespread use of tautological scales and the reliance on cross-sectional data cast doubts.
-Here, we leverage nine years of panel response data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction ($N$ =  20979, New Zealand, 2010-2018).
-Contrary to previous studies, we find that belief in a spirit or life force predicts initially lower personal wellbeing and life-satisfaction.
-However, in support of theories that sprituality improves psychological health, beliefs in a spirit or life force predict increasing life satisfaction over time relative to disbelief.
-These finding holds after adjusting for known demographic influences.
-Intriguingly these findings also hold among those who believe in a God but disbelieve in a spirit or life force.
-The causal mechanisms by which beliefs in a spirit or life force increases well-being remain an important horizon for future investigations.
+However, the widespread use of tautological scales and reliance on cross-sectional data cast doubts.
+Here, we leverage nine years of panel data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction ($N$ =  20979, New Zealand, 2010-2018).
+We find that belief in a spirit or life force predicts lower personal wellbeing and life-satisfaction.
+However, in support of previous theories, beliefs in a spirit or life force predict increasing personal well-being and life satisfaction over time relative to disbelief.
+These finding holds after adjusting for known demographic influences, and intriguingly, among those who believe in a God but disbelieve in a spirit or life force.
+The recent growth in spiritual beliefs and decline in traditional religion across many industrial societies motivates futher causal nvestigations of the social and psychological mechanisms through which spiritual beliefs lead to growth in subjective wellbeing.
+  <!-- https://tinyurl.com/ybremelq -->
 
-  <!-- https://tinyurl.com/ybremelq -->
 keywords          : "Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being"
-wordcount         : no
+wordcount         : "4778"
 bibliography      : ["b-references.bib"]
+
 floatsintext      : no
 figurelist        : no
+figsintext        : no
 tablelist         : no
@@
 draft             : no
+
 documentclass     : "apa6"
@@
    - \usepackage{multicol}
+   - \usepackage{graphics}
+   - \usepackage{rotating}
+   - \usepackage{longtable}
 ---
@@
 ```
-```{r data processing, cache = TRUE, dependson = "ldf.5"}
+```{r data processing, cache = TRUE, dependson = "ldf.5", include=FALSE}
 # filter only those who are measured 
@@
 
-The relationship between spirituality and psychological well-being is a matter of enduring human fascination [@Ellsworth2010-yu].
-Previous attempts to quantify the relationship between these two domains find a positiven association between spirituality and well-being [@Russinova2002-rq; @Ginsburg1995-jr; @Zaza2005-ac; @Koenig2010-gk; @Smith2003-re; @Smith2003-re; @Koenig2001-ow].
-However, critics have raised three challenges.
+The relationship between spirituality and psychological well-being is a matter of enduring fascination [@Ellsworth2010-yu].
+Previous attempts to quantify the relationship between these two domains reveal a positive association: spirituality predicts well-being [@Russinova2002-rq; @Ginsburg1995-jr; @Zaza2005-ac; @Koenig2010-gk; @Smith2003-re; @Smith2003-re; @Koenig2001-ow].
+However, critics raise three challenges.
 
@@
 First, previous research has frequently operationalised spirituality using measures of well-being.
-For example, The Daily Spiritual Experience Scale, which is widely used in spiritual mental health research includes items such as "I feel thankful for my blessings" and "I feel deep inner peace or harmony" to measure spirituality [@Underwood2002-hg].
-As Koenig observes, it is unsurprising that well-being predicts well-being –.
-what Koenig describes as a  "tautology" in spirituality/mental health research [@Koenig2008-lv].
-Indeed, recent systematic-review of spirituality/mental health research found that nearly 45% of previous studies investigating spirituality and emotional well-being operationalise spirituality scales using constructs for positive emotional well-being [@Garssen2016-kb].
+For example, The Daily Spiritual Experience Scale, which is widely used in spiritual mental health research, includes items such as "I feel thankful for my blessings" and "I feel deep inner peace or harmony" to measure spirituality [@Underwood2002-hg].
+As Koenig observes, it is unsurprising that spirituality predict well-being when the same items are used to measure both constructs.
+Koenig aptly describes this features as a "tautology" in spirituality/mental health research [@Koenig2008-lv].
+Indeed, recent systematic-review of spirituality/mental health research found that nearly 45% of previous studies employe tautological scales [@Garssen2016-kb].
 The first challenge, then, is to define measures of spirituality that are not well-being measures reclocked in other words.
 
-How, then, should researchers operationalise spirituality.
-We can look to folk traditions for guidance.
-The term "spirituality" was first used to describe the religious practices of ascetics and monks.
-more recently, however, the term has taken on a family of broader meaning [@Swinton2001-vr].
-Religion is typically operationalized to mean beliefs and practices respecting a God or gods (JB).
-A recurring element in the use of ‘spirituality,’ as it is distinguished from religion, is belief in a spirit or life force without commitment to a belief in a God or gods [@Hill2000-cl; @Davis2015-tx].
-However, the term "spirituality" overlaps with the uses of "religion" in the sense that God or gods are instances of beliefs in spirits – that is one may be both religious and spiritual.
-Here we operationalize religious belief as either belief in a God or belief in both a God and a spirit or life force.
-We operationalize spiritual belief as either belief in a spirit or life force without any corresponding belief in a God.
-However, the term "spirituality" overlaps with the uses of "religion" in the sense that God or gods are instances of beliefs in spirits – that is one may be both religious and spiritual.
-Here we operationalize religious belief as either belief in a God or belief in both a God and a spirit or life force.
-isNotably, previous spirituality mental health research has been disproportionately conducted on North American samples.
-In the United States, especially, samples include a majority of people who believe in both a God as well as and life forces other than God [@Smith2003-re; @Koenig2010-gk].
-Although such studies drawing from such samples mightmay clarify the effects of religious spirituality on psychological subjective well-being, the question of how non-religious spirituality affects psychological well-being that is – or spiritual beliefs without any belief in a God or gods – affects subjective well-being remain unclear.
-(For discussion, see: Therefore, to better understand possible distinctions between religious spirituality and non-religious spirituality for subjective well-being outcomes, samples collected from cultures outside the North Atlantic that contain a more equal ratio of the two are needed [@Garssen2016-km]).
+A second challenge is to disentangle concepts of spirituality from concepts of religion.
+The term "spirituality" was first used to describe the religious practices of ascetics and monks, however, more recently, the term has taken on a broader family of meanings [@Swinton2001-vr].
+However previous research does not converge on a unique set of meanings.
+Most studies of spirituality and religion have been conducted among (cross-sectional) North American samples among people who are affiliate with traditional religion [@Ano2005-hx; @Hackney2003-rs; @Sawatzky2005-rw; @Smith2003-re;  @Visser2010-kq; @Yonker2012-zg].
+However, when spirituality has been operationalised as distinct from religion, some researchers have found a negative association within subjective well-being [@King2013-cg].
+Thus, in addition to disentangling the concept of spirituality from concepts of subjective well-being, it is also important to disentangel the concept of spirituality from concepts of religion.
 
-Third, most research investigating the relationship between spirituality and subjective well-being has relied on cross-sectional samples.
-These studies include high-quality reviews and meta-analyses [@Ano2005-hx; @Hackney2003-rs; @Sawatzky2005-rw; @Smith2003-re; @Visser2010-kq; @Yonker2012-zg].
-However, to assess whetherto whether spiritual beliefs lead to a growth, stability, or decline in psychological well-being relative to people who lack such beliefs , including people who believe in God but not a spirit or lifeforce, requires repeatedly measuring the same individuals over time.
-clarify causal effects requires measuring the same individuals over time and assessing how life turns out for demographically similar people who differ in their spiritual and religious beliefs over their lifespans  and orientations [@Garssen2016-km].
+A third challenge is to investigate the relationship between spirituality and subjective well-being over the life-span in nationally diverse samples.
+To assess whether spiritual beliefs lead to growth, stability, or decline in psychological well-being requires comparing requires repeatedly measuring individuals who differ in spirituality, demography, and religion as these people age, and investigating how their perceive their lives to be turning out [@Garssen2016-km].
+However, previous investigations of spirituality and wellbeing, including high-quality reviews and meta-analyses, have relied on cross-sectional samples [@Ano2005-hx; @Hackney2003-rs; @Sawatzky2005-rw; @Smith2003-re; @Visser2010-kq; @Yonker2012-zg].
+To address the key questions in spirituality psychological health research, then, requires longitudional investigations.
 
-Our study addresses these three limitations in previous research.improves on past methods the following ways: First, we avoid a tautological elision of "spirituality" and "well-being"y by adopting a non-affective measure of spirituality through use of general, non-affective measures: "Do you believe in some form of a spirit or life force?" Our measures of psychological well being are [BEN SAY HERE] and Second we adopt a four-level categorical indicator that disentangles spiritual and religious beliefs: (1) Skeptics: those who neither believe in a God nor a life force; (2) SpiritExcludeGod: those who no not believe in a God  but believe in a spirit or life force; (3) GodAndSpirit: those who believe in a God and spirit; (4) GodExcludesSpirit: those who believe in both a God and a spirit or life force "Do you believe in a god?".
-Second, we address the prior dependence on cross-sectional samples taken from the United States by analyzing data taken from By sampling from respondents in New Zealand, a country with [GIVE NON_RELIGIOUS SPIRITUAL STATS HERE).
-Third, we are able to assess whether spirit beliefs are associate with a growth in psychological wellbeing address causation by weusing we use longitudinal responses from athis nationally diverse probability sample year over an nine-yearseight-year period with nine measurement points  to assess whether spirit beliefs lead to growth, stability, or decline in psychological well-beingwellbeing (2010 to 2018 New Zealand Attitudes and Values Study, NZAVS).
+Here, we addresses the three challenges of previous inthe following ways: First, we avoid a tautological elision of "spirituality" and "well-being" by adopting a non-affective measure of spirituality as a believe in a Spirit or Life Force.
+To  measure subjective well-being we use a well-validated scale for Personal Well-being and a well-validated scale for Life-Satisfaction (see Method).
+This design avoids measuring spirituality with an indicator of spirituality that is cloacked as psychological well-being, thereby avoiding tautology.
+Second, we adopt a four-level categorical indicator for belief-states to clarify differences among people with spiritual, religious, and skeptical beliefs.
+We construct this four-level categorical indicator by combining responses to the Spiritual Belief measure with responses to the question: "Do you believe in some form of a spirit or life force?" The levels of our four-level categorical indicator of spiritual and religious beliefs are as follows: (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who nevertheless believe in a Spirit or Life Force (henceforth, SpiritExcludeGod); (3) Those who believe in a God but do not believe in a Spirit or Life force (henceforth, GodAndSpirit); (4) Those who who believe in a both God and a Spirit or Life Force, (henceforth GodExcludesSpirit).
+Narrowing the concept of spirituality to belief state comes at a price.
+Such an operationalisation, clearly, cannot simultaneously address all interests in previous research.
+However, this cost to scope is paid for with the benefits of precision.
+The spiritual indicators measure cognitive states in which distinct varieties of skeptical beliefs, traditional religious beliefs, and non-traditional spirit beliefs can are disentangled.
+Third, to assess whether spirit beliefs lead to growth, stability, or decline in psychological well-being, we leverage longitudinal responses from a nationally diverse probability sample year over an nine--year period with nine measurement points (2010 to 2018 New Zealand Attitudes and Values Study, NZAVS).
 
 # Method
-
-<!-- We rigorously assess the relationship between our non-affective measures and two constructs of subjective well-being (personal well-being and life satisfaction), adjusting for known demographic and personal influences, including age, sex, deprivation, and political orientation (Okulicz-Kozaryn, Holmes, and Avery 2014; Cuesta and Budría 2012; Røysamb et al.
-2002; Hansen and Slagsvold 2012; Chen and van Ours, n.d.; Vaughan et al.
-1985; Easterlin, Angelescu, and Zweig 2011; Stam et al.
-2016; Yakovlev and Leguizamon 2012).
-Collectively, then, our approach avoids the limitations of many previous studies and helps to clarify existing literature concerning how best to study the relationship between spirituality and subjective well-being.
--->
-
-## Ethics Statement
 
@@
 
-**Personal well-being** Personal well-being is a measure of function all well-being that was assessed using a 4-item version of the Australian Unity well-being Index (Cummins et al.
-2017).
+**Personal well-being** Personal well-being is a measure of function all well-being that was assessed using a 4-item version of the Australian Unity well-being Index [@Cummins2017-ur].
 Participants rated their satisfaction with (a) "Your standard of living"; (b) "Your health"; (c) "Your future security"; (d) "Your personal relationships." Items were rated using a 10-point response option ranging from 1 = completely dissatisfied to 10 = completely satisfied.
@@
 Our findings support critiques of the reliability of past methods.
-The majority of studies have found positive relationships between religion and spirituality and subjective well-being.
-These previous studies primarily use samples taken from within the North Atlantic region and are cross-sectional in design (Ano and Vasconcelles 2005; Hackney and Sanders 2003; Sawatzky, Ratner, and Chiu 2005; Smith, McCullough, and Poll 2003; Visser, Garssen, and Vingerhoets 2010; Yonker, Schnabelrauch, and Dehaan 2012).
-However, our results find a negative association between both belief in a spirit or life force that excludes a god (BXG) and belief that includes a god (BIG).
-These longitudinal results persist independent of time.
-Notably, our findings do partially support other previous work that shows that spirituality outside of a religious framework is associated with worse emotional well-being, though we also find a negative association within a religious framework (King et al.
-2013).
 Still, further research is needed to determine if other aspects of religion and spirituality, such as practices and rituals, might yield different results.

### 2020-11-16 Manuscript.tex
Adding figures, updating manuscript (clarified introduction)

@@
   pdftitle={National Longitudinal Study Reveals Growth in Wellbeing Among Those with Spiritual Beliefs},
-  pdfauthor={Ben Highland1, Ev l Worthing2, Don E Davis3, Chris G. Sibley4, \& Joseph Bulbulia5},
+  pdfauthor={Ben Highland1, Everett L. Worthington2, Don E Davis3, Chris G. Sibley4, \& Joseph Bulbulia5},
   pdflang={en-EN},
@@
 \shorttitle{Beliefs and Wellbeing}
-\keywords{Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being\newline\indent Word count: FALSE}
+\keywords{Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being\newline\indent Word count: 4778}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 \usepackage{multicol}
+\usepackage{graphics}
+\usepackage{rotating}
+\usepackage{longtable}
 \ifxetex
@@
 \title{National Longitudinal Study Reveals Growth in Wellbeing Among Those with Spiritual Beliefs}
-\author{Ben Highland\textsuperscript{1}, Ev l Worthing\textsuperscript{2}, Don E Davis\textsuperscript{3}, Chris G. Sibley\textsuperscript{4}, \& Joseph Bulbulia\textsuperscript{5}}
+\author{Ben Highland\textsuperscript{1}, Everett L. Worthington\textsuperscript{2}, Don E Davis\textsuperscript{3}, Chris G. Sibley\textsuperscript{4}, \& Joseph Bulbulia\textsuperscript{5}}
 \date{}
@@
 
-Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.
-Enter author note here.
-
-The authors made the following contributions.
-Ben Highland: Conceptualization, Literature Review, Writing - First draft, Writing - Review \& editing; Ev l Worthing: Writing review \& editing; Don E Davis: Writing review \& editing; Chris G.
-Sibley: Writing review \& editing; Joseph Bulbulia: Data analysis, Writing second and subsequent drafts, Writing review \& editing.
-
-Correspondence concerning this article should be addressed to Ben Highland, Postal address. E-mail: \href{mailto:my@email.com}{\nolinkurl{my@email.com}}
+1 Wake Forest Medical School 2 Department of Psychology Virginia Commonwealth University 3 Department of Psychology Georgia State University 4 School of Psychology University of Auckland 5 School of Psychology Victoria University
 
@@
 Previous research finds an association between spirituality and subjective well-being.
-However, the widespread use of tautological scales and the reliance on cross-sectional data cast doubts.
-Here, we leverage nine years of panel response data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction (\(N\) = 20979, New Zealand, 2010-2018).
-Contrary to previous studies, we find that belief in a spirit or life force predicts initially lower personal wellbeing and life-satisfaction.
-However, in support of theories that sprituality improves psychological health, beliefs in a spirit or life force predict increasing life satisfaction over time relative to disbelief.
-These finding holds after adjusting for known demographic influences.
-Intriguingly these findings also hold among those who believe in a God but disbelieve in a spirit or life force.
-The causal mechanisms by which beliefs in a spirit or life force increases well-being remain an important horizon for future investigations.
+However, the widespread use of tautological scales and reliance on cross-sectional data cast doubts.
+Here, we leverage nine years of panel data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction (\(N\) = 20979, New Zealand, 2010-2018).
+We find that belief in a spirit or life force predicts lower personal wellbeing and life-satisfaction.
+However, in support of previous theories, beliefs in a spirit or life force predict increasing personal well-being and life satisfaction over time relative to disbelief.
+These finding holds after adjusting for known demographic influences, and intriguingly, among those who believe in a God but disbelieve in a spirit or life force.
+The recent growth in spiritual beliefs and decline in traditional religion across many industrial societies motivates futher causal nvestigations of the social and psychological mechanisms through which spiritual beliefs lead to growth in subjective wellbeing.
 }
@@
 
-The relationship between spirituality and psychological well-being is a matter of enduring human fascination (Ellsworth \& Ellsworth, 2010).
-Previous attempts to quantify the relationship between these two domains find a positiven association between spirituality and well-being (Ginsburg, Quirt, Ginsburg, \& MacKillop, 1995; Koenig, 2010; Koenig, Associate Professor of Psychology and Religious Studies Michael E McCullough, PhD, McCullough, Larson, \& Adjunct Professor of Psychiatry David B Larson, 2001; Russinova, Wewiorski, \& Cash, 2002; Smith, McCullough, \& Poll, 2003; Smith et al., 2003; Zaza, Sellick, \& Hillier, 2005).
-However, critics have raised three challenges.
+The relationship between spirituality and psychological well-being is a matter of enduring fascination (Ellsworth \& Ellsworth, 2010).
+Previous attempts to quantify the relationship between these two domains reveal a positive association: spirituality predicts well-being (Ginsburg, Quirt, Ginsburg, \& MacKillop, 1995; Koenig, 2010; Koenig, Associate Professor of Psychology and Religious Studies Michael E McCullough, PhD, McCullough, Larson, \& Adjunct Professor of Psychiatry David B Larson, 2001; Russinova, Wewiorski, \& Cash, 2002; Smith, McCullough, \& Poll, 2003; Smith et al., 2003; Zaza, Sellick, \& Hillier, 2005).
+However, critics raise three challenges.
 
 First, previous research has frequently operationalised spirituality using measures of well-being.
-For example, The Daily Spiritual Experience Scale, which is widely used in spiritual mental health research includes items such as \enquote{I feel thankful for my blessings} and \enquote{I feel deep inner peace or harmony} to measure spirituality (Underwood \& Teresi, 2002).
-As Koenig observes, it is unsurprising that well-being predicts well-being --.
-what Koenig describes as a \enquote{tautology} in spirituality/mental health research (Koenig, 2008).
-Indeed, recent systematic-review of spirituality/mental health research found that nearly 45\% of previous studies investigating spirituality and emotional well-being operationalise spirituality scales using constructs for positive emotional well-being (Garssen et al., 2016).
+For example, The Daily Spiritual Experience Scale, which is widely used in spiritual mental health research, includes items such as \enquote{I feel thankful for my blessings} and \enquote{I feel deep inner peace or harmony} to measure spirituality (Underwood \& Teresi, 2002).
+As Koenig observes, it is unsurprising that spirituality predict well-being when the same items are used to measure both constructs.
+Koenig aptly describes this features as a \enquote{tautology} in spirituality/mental health research (Koenig, 2008).
+Indeed, recent systematic-review of spirituality/mental health research found that nearly 45\% of previous studies employe tautological scales (Garssen et al., 2016).
 The first challenge, then, is to define measures of spirituality that are not well-being measures reclocked in other words.
 
-How, then, should researchers operationalise spirituality.
-We can look to folk traditions for guidance.
-The term \enquote{spirituality} was first used to describe the religious practices of ascetics and monks.
-more recently, however, the term has taken on a family of broader meaning (Swinton, 2001).
-Religion is typically operationalized to mean beliefs and practices respecting a God or gods (JB).
-A recurring element in the use of \enquote{spirituality,} as it is distinguished from religion, is belief in a spirit or life force without commitment to a belief in a God or gods (Davis et al., 2015; Hill et al., 2000).
-However, the term \enquote{spirituality} overlaps with the uses of \enquote{religion} in the sense that God or gods are instances of beliefs in spirits -- that is one may be both religious and spiritual.
-Here we operationalize religious belief as either belief in a God or belief in both a God and a spirit or life force.
-We operationalize spiritual belief as either belief in a spirit or life force without any corresponding belief in a God.
-However, the term \enquote{spirituality} overlaps with the uses of \enquote{religion} in the sense that God or gods are instances of beliefs in spirits -- that is one may be both religious and spiritual.
-Here we operationalize religious belief as either belief in a God or belief in both a God and a spirit or life force.
-isNotably, previous spirituality mental health research has been disproportionately conducted on North American samples.
-In the United States, especially, samples include a majority of people who believe in both a God as well as and life forces other than God (Koenig, 2010; Smith et al., 2003).
-Although such studies drawing from such samples mightmay clarify the effects of religious spirituality on psychological subjective well-being, the question of how non-religious spirituality affects psychological well-being that is -- or spiritual beliefs without any belief in a God or gods -- affects subjective well-being remain unclear.
-(For discussion, see: Therefore, to better understand possible distinctions between religious spirituality and non-religious spirituality for subjective well-being outcomes, samples collected from cultures outside the North Atlantic that contain a more equal ratio of the two are needed (Garssen \& Visser, 2016)).
+A second challenge is to disentangle concepts of spirituality from concepts of religion.
+The term \enquote{spirituality} was first used to describe the religious practices of ascetics and monks, however, more recently, the term has taken on a broader family of meanings (Swinton, 2001).
+However previous research does not converge on a unique set of meanings.
+Most studies of spirituality and religion have been conducted among (cross-sectional) North American samples among people who are affiliate with traditional religion (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky, Ratner, \& Chiu, 2005; Smith et al., 2003; Visser, Garssen, \& Vingerhoets, 2010; Yonker, Schnabelrauch, \& Dehaan, 2012).
+However, when spirituality has been operationalised as distinct from religion, some researchers have found a negative association within subjective well-being (King et al., 2013).
+Thus, in addition to disentangling the concept of spirituality from concepts of subjective well-being, it is also important to disentangel the concept of spirituality from concepts of religion.
 
-Third, most research investigating the relationship between spirituality and subjective well-being has relied on cross-sectional samples.
-These studies include high-quality reviews and meta-analyses (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky, Ratner, \& Chiu, 2005; Smith et al., 2003; Visser, Garssen, \& Vingerhoets, 2010; Yonker, Schnabelrauch, \& Dehaan, 2012).
-However, to assess whetherto whether spiritual beliefs lead to a growth, stability, or decline in psychological well-being relative to people who lack such beliefs , including people who believe in God but not a spirit or lifeforce, requires repeatedly measuring the same individuals over time.
-clarify causal effects requires measuring the same individuals over time and assessing how life turns out for demographically similar people who differ in their spiritual and religious beliefs over their lifespans and orientations (Garssen \& Visser, 2016).
+A third challenge is to investigate the relationship between spirituality and subjective well-being over the life-span in nationally diverse samples.
+To assess whether spiritual beliefs lead to growth, stability, or decline in psychological well-being requires comparing requires repeatedly measuring individuals who differ in spirituality, demography, and religion as these people age, and investigating how their perceive their lives to be turning out (Garssen \& Visser, 2016).
+However, previous investigations of spirituality and wellbeing, including high-quality reviews and meta-analyses, have relied on cross-sectional samples (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky et al., 2005; Smith et al., 2003; Visser et al., 2010; Yonker et al., 2012).
+To address the key questions in spirituality psychological health research, then, requires longitudional investigations.
 
-Our study addresses these three limitations in previous research.improves on past methods the following ways: First, we avoid a tautological elision of \enquote{spirituality} and \enquote{well-being}y by adopting a non-affective measure of spirituality through use of general, non-affective measures: \enquote{Do you believe in some form of a spirit or life force?} Our measures of psychological well being are {[}BEN SAY HERE{]} and Second we adopt a four-level categorical indicator that disentangles spiritual and religious beliefs: (1) Skeptics: those who neither believe in a God nor a life force; (2) SpiritExcludeGod: those who no not believe in a God but believe in a spirit or life force; (3) GodAndSpirit: those who believe in a God and spirit; (4) GodExcludesSpirit: those who believe in both a God and a spirit or life force \enquote{Do you believe in a god?}.
-Second, we address the prior dependence on cross-sectional samples taken from the United States by analyzing data taken from By sampling from respondents in New Zealand, a country with {[}GIVE NON\_RELIGIOUS SPIRITUAL STATS HERE).
-Third, we are able to assess whether spirit beliefs are associate with a growth in psychological wellbeing address causation by weusing we use longitudinal responses from athis nationally diverse probability sample year over an nine-yearseight-year period with nine measurement points to assess whether spirit beliefs lead to growth, stability, or decline in psychological well-beingwellbeing (2010 to 2018 New Zealand Attitudes and Values Study, NZAVS).
+Here, we addresses the three challenges of previous inthe following ways: First, we avoid a tautological elision of \enquote{spirituality} and \enquote{well-being} by adopting a non-affective measure of spirituality as a believe in a Spirit or Life Force.
+To measure subjective well-being we use a well-validated scale for Personal Well-being and a well-validated scale for Life-Satisfaction (see Method).
+This design avoids measuring spirituality with an indicator of spirituality that is cloacked as psychological well-being, thereby avoiding tautology.
+Second, we adopt a four-level categorical indicator for belief-states to clarify differences among people with spiritual, religious, and skeptical beliefs.
+We construct this four-level categorical indicator by combining responses to the Spiritual Belief measure with responses to the question: \enquote{Do you believe in some form of a spirit or life force?} The levels of our four-level categorical indicator of spiritual and religious beliefs are as follows: (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who nevertheless believe in a Spirit or Life Force (henceforth, SpiritExcludeGod); (3) Those who believe in a God but do not believe in a Spirit or Life force (henceforth, GodAndSpirit); (4) Those who who believe in a both God and a Spirit or Life Force, (henceforth GodExcludesSpirit).
+Narrowing the concept of spirituality to belief state comes at a price.
+Such an operationalisation, clearly, cannot simultaneously address all interests in previous research.
+However, this cost to scope is paid for with the benefits of precision.
+The spiritual indicators measure cognitive states in which distinct varieties of skeptical beliefs, traditional religious beliefs, and non-traditional spirit beliefs can are disentangled.
+Third, to assess whether spirit beliefs lead to growth, stability, or decline in psychological well-being, we leverage longitudinal responses from a nationally diverse probability sample year over an nine--year period with nine measurement points (2010 to 2018 New Zealand Attitudes and Values Study, NZAVS).
 
@@
 \section{Method}\label{method}}
-
-\hypertarget{ethics-statement}{%
-\subsection{Ethics Statement}\label{ethics-statement}}
 
@@
 
-\textbf{Personal well-being} Personal well-being is a measure of function all well-being that was assessed using a 4-item version of the Australian Unity well-being Index (Cummins et al.~2017).
+\textbf{Personal well-being} Personal well-being is a measure of function all well-being that was assessed using a 4-item version of the Australian Unity well-being Index (Cummins, Eckerseley, Pallant, Van Vugt, \& Misajon, 2017).
 Participants rated their satisfaction with (a) \enquote{Your standard of living}; (b) \enquote{Your health}; (c) \enquote{Your future security}; (d) \enquote{Your personal relationships.} Items were rated using a 10-point response option ranging from 1 = completely dissatisfied to 10 = completely satisfied.
@@
 Our findings support critiques of the reliability of past methods.
-The majority of studies have found positive relationships between religion and spirituality and subjective well-being.
-These previous studies primarily use samples taken from within the North Atlantic region and are cross-sectional in design (Ano and Vasconcelles 2005; Hackney and Sanders 2003; Sawatzky, Ratner, and Chiu 2005; Smith, McCullough, and Poll 2003; Visser, Garssen, and Vingerhoets 2010; Yonker, Schnabelrauch, and Dehaan 2012).
-However, our results find a negative association between both belief in a spirit or life force that excludes a god (BXG) and belief that includes a god (BIG).
-These longitudinal results persist independent of time.
-Notably, our findings do partially support other previous work that shows that spirituality outside of a religious framework is associated with worse emotional well-being, though we also find a negative association within a religious framework (King et al.~2013).
 Still, further research is needed to determine if other aspects of religion and spirituality, such as practices and rituals, might yield different results.
@@
 
-\leavevmode\hypertarget{ref-Davis2015-tx}{}%
-Davis, D. E., Rice, K., Hook, J. N., Van Tongeren, D. R., DeBlaere, C., Choe, E., \& Worthington, E. L. (2015). Sources of spirituality scale. \emph{PsycTESTS Dataset}.
+\leavevmode\hypertarget{ref-Cummins2017-ur}{}%
+Cummins, R. A., Eckerseley, R., Pallant, J., Van Vugt, J., \& Misajon, R. (2017). Australian unity wellbeing index. \emph{PsycTESTS Dataset}.
 
@@
 Hackney, C. H., \& Sanders, G. S. (2003). Religiosity and mental health: A meta--analysis of recent studies. \emph{J. Sci. Study Relig.}
-
-\leavevmode\hypertarget{ref-Hill2000-cl}{}%
-Hill, P.
-C., Pargament, K.
-I.
-I., Hood, R.
-W., McCullough, M.
-E., Jr, Swyers, J.
-P., Larson, D.
-B., \& Zinnbauer, B.
-J.
-(2000).
-Conceptualizing religion and spirituality: Points of commonality, points of departure.
-\emph{J.
-Theory Soc.
-Behav.}, \emph{30}(1), 51--77.
 

### 2020-11-16 Manuscript.Rmd
fixed "figs" --> "Figs" and testing overleaf push.

@@
 ```{r fig.cap = "Combination Coefficient Plots for Personal Well-being and Life Satisfaction Spirit Belief Longitudinal Models"}
-knitr::include_graphics("figs/USEcoefficientplots-1.png")
+knitr::include_graphics("Figs/USEcoefficientplots-1.png")
 ```

### 2020-11-16 Manuscript.Rmd
further edits

@@
  Previous research finds an association between spirituality and subjective well-being.
-However, the widespread use of tautological scales and reliance on cross-sectional data cast doubts.
+However, the widespread use of tautological spirituality scales, poorly defined concepts of spirituality, and reliance on cross-sectional samples cast doubts.
 Here, we leverage nine years of panel data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction ($N$ = 20979, New Zealand, 2010-2018).
-We find that belief in a spirit or life force predicts lower personal wellbeing and life-satisfaction.
-However, in support of previous theories, beliefs in a spirit or life force predict increasing personal well-being and life satisfaction over time relative to disbelief.
+Contrary to previous research, we find that belief in a spirit or life force predicts lower personal well-being and life-satisfaction.
+However, in support of previous speculation, beliefs in a spirit or life force predict increasing personal well-being and life satisfaction over time relative to disbelief.
 These finding holds after adjusting for known demographic influences, and intriguingly, among those who believe in a God but disbelieve in a spirit or life force.
-The recent growth in spiritual beliefs and decline in traditional religion across many industrial societies motivates futher causal nvestigations of the social and psychological mechanisms through which spiritual beliefs lead to growth in subjective wellbeing.
+The recent growth in spiritual beliefs and decline in traditional religion across many industrial societies motivates further causal investigations of the social and psychological mechanisms through which spiritual beliefs lead to growth in subjective well-being.
   <!-- https://tinyurl.com/ybremelq -->
@@
 The relationship between spirituality and psychological well-being is a matter of enduring fascination [@Ellsworth2010-yu].
-Previous attempts to quantify the relationship between these two domains reveal a positive association: spirituality predicts well-being [@Russinova2002-rq; @Ginsburg1995-jr; @Zaza2005-ac; @Koenig2010-gk; @Smith2003-re; @Smith2003-re; @Koenig2001-ow].
+Most previous attempts to quantify the relationship between these two domains find a positive association: spirituality predicts well-being [@Russinova2002-rq; @Ginsburg1995-jr; @Zaza2005-ac; @Koenig2010-gk; @Smith2003-re; @Smith2003-re; @Koenig2001-ow].
 However, critics raise three challenges.
@@
 Indeed, recent systematic-review of spirituality/mental health research found that nearly 45% of previous studies employe tautological scales [@Garssen2016-kb].
-The first challenge, then, is to define measures of spirituality that are not well-being measures reclocked in other words.
+The first challenge, then, is to define measures of spirituality that are not well-being measures dressed in other words.
 
@@
 However, when spirituality has been operationalised as distinct from religion, some researchers have found a negative association within subjective well-being [@King2013-cg].
-Thus, in addition to disentangling the concept of spirituality from concepts of subjective well-being, it is also important to disentangel the concept of spirituality from concepts of religion.
+Thus, in addition to disentangling the concept of spirituality from concepts of subjective well-being, it is also important to disentangle the concept of spirituality from concepts of religion.
 
@@
 To assess whether spiritual beliefs lead to growth, stability, or decline in psychological well-being requires comparing requires repeatedly measuring individuals who differ in spirituality, demography, and religion as these people age, and investigating how their perceive their lives to be turning out [@Garssen2016-km].
-However, previous investigations of spirituality and wellbeing, including high-quality reviews and meta-analyses, have relied on cross-sectional samples [@Ano2005-hx; @Hackney2003-rs; @Sawatzky2005-rw; @Smith2003-re; @Visser2010-kq; @Yonker2012-zg].
-To address the key questions in spirituality psychological health research, then, requires longitudional investigations.
+However, previous investigations of spirituality and well-being, including high-quality reviews and meta-analyses, have relied on cross-sectional samples [@Ano2005-hx; @Hackney2003-rs; @Sawatzky2005-rw; @Smith2003-re; @Visser2010-kq; @Yonker2012-zg].
+To address the key questions in spirituality psychological health research, then, requires longitudinal investigations.
 
-Here, we addresses the three challenges of previous inthe following ways: First, we avoid a tautological elision of "spirituality" and "well-being" by adopting a non-affective measure of spirituality as a believe in a Spirit or Life Force.
-To  measure subjective well-being we use a well-validated scale for Personal Well-being and a well-validated scale for Life-Satisfaction (see Method).
-This design avoids measuring spirituality with an indicator of spirituality that is cloacked as psychological well-being, thereby avoiding tautology.
-Second, we adopt a four-level categorical indicator for belief-states to clarify differences among people with spiritual, religious, and skeptical beliefs.
-We construct this four-level categorical indicator by combining responses to the Spiritual Belief measure with responses to the question: "Do you believe in some form of a spirit or life force?" The levels of our four-level categorical indicator of spiritual and religious beliefs are as follows: (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who nevertheless believe in a Spirit or Life Force (henceforth, SpiritExcludeGod); (3) Those who believe in a God but do not believe in a Spirit or Life force (henceforth, GodAndSpirit); (4) Those who who believe in a both God and a Spirit or Life Force, (henceforth GodExcludesSpirit).
-Narrowing the concept of spirituality to belief state comes at a price.
-Such an operationalisation, clearly, cannot simultaneously address all interests in previous research.
-However, this cost to scope is paid for with the benefits of precision.
-The spiritual indicators measure cognitive states in which distinct varieties of skeptical beliefs, traditional religious beliefs, and non-traditional spirit beliefs can are disentangled.
+Here, we addresses the three challenges of previous in the following ways: First, to avoid a tautological elision of "spirituality" and "well-being" we adopt a non-affective measure of spirituality as a "believe in a Spirit or Life Force." On the other side, to measure subjective well-being, we use a well-validated scale for Personal Well-being and a well-validated scale for Life-Satisfaction (see Method).
+Second, to disentangle spiritual beliefs from traditional religious beliefs and skeptical beliefs we adopt a four-level categorical indicator combining the binary indicator of belief in a Spirit or Life Force with the binary indicator "Do you believe in some form of a spirit or life force?" The levels of this four-level categorical indicator for spiritual, religious, and skeptical beliefs are as follows: (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who nevertheless believe in a Spirit or Life Force; (3) Those who believe in a God but do not believe in a Spirit or Life force; (4) Those who who believe in a both God and a Spirit or Life Force.
+Clearly such an approach cannot simultaneously address all interests in previous spirituality/well-being research.
+For example we do not combine our cognitive measure of belief with behavioral measures or attitudinal measures.
+However, this cost to scope is paid for with benefits to precision.
+The spiritual indicators measure cognitive states in which distinct varieties of skeptical beliefs, traditional religious beliefs, and non-traditional spirit beliefs are straightforwardly disentangled, and measured within people.
 Third, to assess whether spirit beliefs lead to growth, stability, or decline in psychological well-being, we leverage longitudinal responses from a nationally diverse probability sample year over an nine--year period with nine measurement points (2010 to 2018 New Zealand Attitudes and Values Study, NZAVS).
@@
 
-***Belief in a Spirit or a God***  We assess belief in a spirit or a God by asking two non-affective questions: "Do you believe in a spirit or life force" and "Do you believe in a god?" Responses were coded as "Belief in a spirit that excludes a god" for those who indicated a belief in both a spirit or life force and a god, "Belief that includes a god" for those who indicated a belief in a god with or without an accompanying belief in a spirit or life force, and "skeptics" for those who indicated belief in neither a spirit or life force nor a god.
-These items appear in the NZAVS questionnaire and were developed for the 2005 Eurobarometer.
+***Belief in a Spirit or a God***  We assess belief in a spirit or a God by asking two non-affective questions: "Do you believe in a spirit or life force" and "Do you believe in a god?" Responses were coded as (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who believe in a Spirit or Life Force; (3) Believers in a God who disbelieve in a Spirit or Life force; (4) Believers in a God and Spirit or Life Force.
+These believe were developed for the NZAVS from the 2005 Eurobarometer.
 For detailed information pertaining to beliefs, see Table 1.
@@
 Statistical analysis was performed using R version 4.0.2 R version 3.6.1 (2020-06-22) 2019-07-05).
-We analyzed data from participants who responded to the NZAVS survey at least three four times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of  21,705 (N = 20,979 with complete responses)N = 15,040.
-To model predict growth in levels of self-reported psychological subjective well-being among these participants we use generalized linear mixed models.
-In our models, we account for the non-linear effects of time within participants by including its interaction with both of our predictor variables: "Belief in a spirit that excludes god" and "Belief that includes a god." To handle the dependencies and heterogeneity introduced from the repeated measures, we We also included individual ID as an effect modelled as random-effect.
-To improve interpretability of our models, we centered and scaled the education, deprivation, and political orientation covariates.
-Additionally, age was centred at and put into decade-long intervals.
-The model equation(s) are described in equation 1.
+We analyze data from participants who responded to the NZAVS survey at least three four times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of  21,705 (N = 20,979 with complete responses).
+To model growth in subjective well-being we use generalized linear mixed models.
+In our models, we account for the non-linear effects of time within participants by including the interaction of years (a continuous variable) with Beliefs (a four-level categorical indicator).
+The baseline for this indicator is Skepticism about a Spirit/Life Forces and Skepticism about a God.
+To handle the dependencies and heterogeneity introduced from the repeated measures, we include individual ID as an effect modelled as random.
+We centered and scaled the education, deprivation, and political conservativism.
+We centred age was centred at and put into decade-long intervals.
+The model equation(s) are described in Equation 1.
+
+**Equation 1**
 
@@
 \begin{aligned}
-\operatorname{{\small LIFESAT/PWI}}  &\sim N \left(\mu,\sigma^2 \right) \\ \mu &=\alpha_{j[i]} + \beta_{1}(\operatorname{Years}) + \beta_{2}(\operatorname{Beliefs}_{\operatorname{\_SpiritExcludesGod\_}})\ + \\
+\operatorname{{\small LIFESAT/PWI}}  &\sim N \left(\mu,\sigma^2 \right) \\ \mu &=\alpha_{j[i]} + \beta_{1}(\operatorname{Years}) + \beta_{2}(\operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\
 &\quad \beta_{3}(\operatorname{Beliefs}_{\operatorname{GodAndSpirit}}) + \beta_{4}(\operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) + \beta_{5}(\operatorname{Age.10yrs.C})\ + \\
@@
 &\quad \beta_{12}(\operatorname{Male}_{\operatorname{1}}) + \beta_{13}(\operatorname{Partner}_{\operatorname{Has\ }}) + \beta_{14}(\operatorname{Pol.Orient.S})\ + \\
-&\quad \beta_{15}(\operatorname{Urban}) + \beta_{16}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{\_SpiritExcludesGod\_}}) + \beta_{17}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ + \\
-&\quad \beta_{18}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) \\ \alpha_{j} &\sim N \left(\mu_{\alpha_{j}},\sigma^2_{\alpha_{j}} \right) , \operatorname{ for  Id } j  = 1, \dots ,  J
+&\quad \beta_{15}(\operatorname{Urban}) + \beta_{16}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}}) + \beta_{17}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ + \\
+&\quad \beta_{18}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) \\ \alpha_{j} &\sim N \left(\mu_{\alpha_{j}},\sigma^2_{\alpha_{j}} \right) , \operatorname{ for  Id }~j~= 1, \dots~J = 20,979~\operatorname{Ids}
 \end{aligned}
@@
 
-All code for this analysis and a full report of the missing data imputation can be found at https://github.com/jacopastorius/NZAVS_Highland_SpiritHealth2.Table 1.
-Descriptive Statistics for Theoretical Variables in Each Wave.
-Belief in a god and belief in a spirit or life force were not measured during Wave 6 (2014).
-"BXG" indicates a belief in a spirit that excludes a god.
-"BIG" indicates a belief that includes a god.
-Many individuals appear in multiple waves.
-
-Table 2. Descriptive Statistics for Demographic Variables in Each Wave. Education level was not measured during Waves 2 (2010) and 3 (2011). Many individuals appear in multiple waves.
+All code for this analysis and a full report of the missing data imputation can be found at https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing.
 
 # Results
-
-Figure 2 about here:
 
@@
 ```
-
 ```{r, cache=TRUE, include=FALSE}
@@
 Results for both the personal well-being and life satisfaction models are presented in Table 2 and Figure 1.
+
+Figure 1 about here:
 
@@
 \centering 
-\caption{"Regression Models"}\label{}
+\caption{"Longitudinal Models"}\label{}
 \scalebox{0.6}{
@@
 \toprule
- & Model 1 & Model 2 \\
+ & Personal Well-Being & Life Satisfaction \\
 \midrule
@@
 ```
-  - The effect of Years X Beliefs_SpiritExcludesGod_ is reliably positive  (beta = 0.02, SE = 4.86e-03, std.
+Focusing on the interaction of time with belief-type on personal well-being we observe:
+
+  - The effect of Years X SpiritExcludesGod_ is reliably positive  (beta = 0.02, SE = 4.86e-03, std.
 beta = 0.02, p < .01).
-We infer that, unlike the skeptic population, people with beliefs in a spirit or life force who do not believe in a God tend to grow in their personal well-beingwellbeing over time.
-  - The effect of Years X BeliefsGodAndSpirit is reliably positive  (beta = 0.02, SE = 4.41e-03, std.
+We infer that, unlike the skeptic population, people with beliefs in a spirit or life force who do not believe in a God tend to grow in their personal well-being over time.
+  - The effect of Years X GodAndSpirit is reliably positive  (beta = 0.02, SE = 4.41e-03, std.
 beta = 0.03, p < .001).
-We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their personal well-beingwellbeing over time
-  - The effect of Years X BeliefsGodExcludesSpirit is not reliable(beta = 1.96e-03, SE = 8.29e-03, std.
+We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their personal well-being over time
+  - The effect of Years X GodExcludesSpirit is not reliable(beta = 1.96e-03, SE = 8.29e-03, std.
 beta = 2.36e-03, p = 0.813).
-Although people who believe in a God but do not believe in a spirit or Life Force tend to have greater personal well-beingwellbeing than those who do not believe in a spirit or life force  (and greater life satisfaction than Skeptics)  this population is not expected to grow in Life Satisfaction over time.
+Thus, God believers who doubt in a Spirit or Life Force start maintain higher personal well-being than do other groups, but they are not expected to grow in personal well-being over time.
 
@@
 beta = -2.70e-03, p = 0.799).
-Although people who believe in a God but do not believe in a spirit or Life Force tend to have greater life satisfaction than those who do not believe in a Spirit or life force  (and greater life satisfaction than Skeptics)  this population is not expected to grow in Life Satisfaction over time.
+Thus, God believers who doubt in a Spirit or Life Force start maintain higher Life-Satisfaction than do other groups, but they are not expected to grow in Life-Satisfaction over time.
 
@@
 
-Figure 2. Subjective Well-being Predicted Probabilities for Belief in a Spirit or Life Force Over Time in a Listwise Deleted Dataset.
+## Imputed data analysis
 
-Table 3.
-Subjective Well-being Coefficient Table for Belief in a Spirit or Life Force.
-Age was put into units of 10 years and centered at its mean.
-Education, Deprivation, and Political Orientation were centered at their means and standardized.
+To adjust for potential biases from missing responses, we conducted a supplementary analysis in which we imputed missing values over the the longitudinal sample.
+This analysis is describe in Supplement 1.
+Averaging over the missing responses does not change inference, however we observe growth across all four groups in both personal well-being and life satisfaction.
+Additionally we observe substantially stronger effects for the slope of Spirit Beliefs on both personal well-being and life-satisfaction.
+Figures 3 and Figures 4 describe the predicted slopes of change.
+The inference from this analysis remains the same as from the analysis in which we delete missing values row-wise, by panel wave.
+In all analyses those who believe in a spirit or life-force are expected to experience greater growth in well-being over time.
+
+Figure 3 about here
+
+Figure 4 about here
 
@@
 
-The purpose of this study has been to clarify the relationship between belief in a spirit/life force and psychological well-being.
-Spirituality as a distinct category, separate from religion, is a little-studied phenomenon in the psychology of religion literature.
+The purpose of this study has been to clarify the relationship between belief in a spirit/life force and subjective well-being, addressing three challenges from previous research: (1) tautological measures of spirituality and subjective well-being; (2) insufficiently distinct measures of spirituality and religion; (3) cross-sectional samples.
+Here, we ask whether beliefs in a Spirit or Life Force predict growth in personal well-being and life-satisfaction over eight years in populations who differ in the this belief, and who also differ in their belief in a God.
+We find that
 
-These findings demonstrate the importance ofinvestigating spiritual beliefs as a domain of interest in its own right.
+These findings demonstrate the importance of investigating spiritual beliefs as a domain of interest in its own right.
 In many Western countries, people are increasingly identifying as "spiritual but not religious" [@Zinnbauer2005-vz; @Swinton2001-vr].
@@
 To assess the relationship between a belief in a spirit or life force and subjective well-being, we improve upon past methods in three ways.
-
-Our findings support critiques of the reliability of past methods.
-Still, further research is needed to determine if other aspects of religion and spirituality, such as practices and rituals, might yield different results.
 
@@
 Thus, in the long term, both sustained "BXG" and "BIG" may actually be associated with better subjective well-being.
+
 Our results also help to clarify the relationship among religion, spirituality, and subjective well-being outcomes.

### 2020-11-17 Analysis.Rmd
- include a source file to make Rmd manuscript more manageable - revised text

@@
 
-# Inference holds  FOCe INTERCEEpTS TO BE 
+# Inference holds 
 mod.1 <-lmer (PWI ~  1 +  Years * Beliefs +  Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id), data= dfben, na.action = na.omit)
@@
 ```
-
-Tables
 
@@
 ```
+```{r}
+library(parameters)
+g1<-plot(parameters::model_parameters(mod.1))+ ggtitle("Personal Wellbeing Coefficients")
+g2<-plot(parameters::model_parameters(mod.2))+ ggtitle("Life Satisfaction Coefficients")
+```
+
+```{r USE_2_coefficientplots}
+g1/g2
+```
 
@@
 #### Predicted Effects ####
-pr.PWI <- ggpredict(model = mod.1, terms = c("Years [0:9]","Beliefs"), 
+pr.PWI <- ggpredict(model = mod.1, terms = c("Years [7:9]","Beliefs"), 
                     ci.lvl = 0.95,
@@
 
-```{r}
+```{r predictlifesat}
 pr.LIFESAT <- ggpredict(model = mod.2, terms = c("Years [0:9]", "Beliefs"),
@@
   ggtitle("Predicted Values of Life Satisfaction") #+   scale_colour_brewer(palette = "Zissou1")
+p2
 ```
@@
 ### equations
-Lifesatisfaction Model 1 equation
+Life satisfaction Model 1 equation
 
-```{r}
+```{r eq1}
 library(equatiomatic)
@@
 PWI Model 2 equation
-```{r}
+```{r eq2}
 library(equatiomatic)
@@
 short report:
-```{r, cache =TRUE, dependson=mod.1,include=FALSE}
+```{r, report1, cache =TRUE, dependson=mod.1,include=FALSE}
 report::model_text(mod.1)
@@
 ```
-
-### Model 2
-For publication change all "significants" to "reliable"
-
 short report
 ```{r, cache =TRUE, dependson=mod.2,include=FALSE}
+# Model 2
+# For publication change all "significants" to "reliable"
+
 report::model_text(mod.2)

### 2020-11-17 Manuscript.Rmd
- include a source file to make Rmd manuscript more manageable - revised text

@@
 ---
-
-```{r include = FALSE}
-library(knitr)
-opts_chunk$set(
-    fig.align = "center",
-    root.dir = "../"
-)
+```{r libraries}
+library("tidyverse")
 library("papaja")
-library("tidyverse")
-library("tinytex")
+library("here")
 ```
 
-```{r load_libraries, echo=FALSE, include=FALSE}
-# load libraries
-library("styler")
-library("tidyverse")
-library("lme4")
-library("ggplot2")
-library("tidyr")
-library("sjPlot")
-library("ggeffects")
-library("Amelia")
-library("table1")
-library("patchwork")
-library("parameters")
-library("prettycode")
-library("gghighlight")
-library("sjPlot")
-library("see")
-library("viridis")
-library("ggpubr")
-library("wesanderson")
-library("ggsci")
-library("here")
-library("citr")
-#library("furniture")
-library("knitr")
-library("texreg")
+```{r runanalysis, include = FALSE}
+source('short.R')
 ```
-```{r load_data, include=FALSE}
-ldf.5<-readRDS("~/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/00Bulbulia\ Pubs/2020/ldf.5") 
-```
-```{r data processing, cache = TRUE, dependson = "ldf.5", include=FALSE}
-# filter only those who are measured 
-dfben  <-ldf.5 %>%
-  dplyr:: filter(Wave !=2009)%>% # exclude wave 1 (no spirit measures that wave)
-  dplyr::filter(YearMeasured==1)%>%
-  dplyr::group_by(Id) %>% filter(n() > 2)%>% # select those who have responded to at least 3 waves # Also works for 5 and 7
-  dplyr::filter(n() !=0)%>%
-  dplyr::ungroup(Id)%>%
-  dplyr:: mutate(Years = as.numeric(years))%>%
-  dplyr::mutate(Age.10yrs = (Age/10))%>%
-  dplyr::mutate(Religion.Church.Log = log(Religion.Church+ 1))%>%
-  dplyr::mutate(Pol.Orient.S = scale(Pol.Orient, center = TRUE, scale = TRUE),
-         Age.10yrs.C = scale(Age.10yrs, center = TRUE, scale = FALSE),
-         Male = factor(Gender),
-         Employed = factor(Employed),
-         Partner = factor(Partner),
-         Edu.S = scale(as.numeric(Edu) ,scale=TRUE,center=TRUE),
-         Education = as.numeric(Edu),
-         Urban = factor(Urban),
-         Believe.Spirit = as.factor(Believe.Spirit),
-         Believe.God = as.factor(Believe.God),
-         EthnicCategories = factor(EthnicCats),
-         LIFESAT.S = scale(LIFESAT, center = TRUE, scale = TRUE),
-         PWI.S = scale(PWI, center = TRUE, scale = TRUE),
-         Deprivation.S = scale(NZdep, scale=TRUE, center=TRUE),
-         Relid.C = scale(Relid, scale = TRUE,center=TRUE),
-         Religious = as.factor(Religious),
-         Religion.Church.Log.C = scale(Religion.Church.Log, center = TRUE, scale = FALSE))
-```
-```{r}
-# nice labels
-table1::label(dfben$Age.10yrs.C) <- "Age in Decades (C)"
-table1::label(dfben$PWI) <- "Personal Wellbeing"
-table1::label(dfben$LIFESAT) <- "Life Satisfaction"
-
-table1::label(dfben$Deprivation.S) <- "Deprivation (S)"
-table1::label(dfben$Pol.Orient) <- "Political Conservative"
-table1::label(dfben$Pol.Orient.S) <- "Political Conservative (S)"
-table1::label(dfben$NZdep) <- "NZ Deprivation Index"
-dfben$Partner <- factor(dfben$Partner, labels = c("No Partner","Has Partner"))
-dfben$Employed <- factor(dfben$Employed, labels = c("Not Employed","Employed"))
-table1::label(dfben$Edu.S) <- "Education (S)"
-
-# create beliefs indicator
-dfben$Beliefs<- factor(ifelse(dfben$Believe.God == "Not Believe God" & dfben$Believe.Spirit == "Not Believe Spirit", "_Skeptic_", 
-                                        ifelse(dfben$Believe.God == "Not Believe God" & dfben$Believe.Spirit == "Believe Spirit", "_SpiritExcludesGod_",
-                                               ifelse(dfben$Believe.God == "Believe God" & dfben$Believe.Spirit == "Believe Spirit","GodAndSpirit","GodExcludesSpirit"))))
-```
-
 The relationship between spirituality and psychological well-being is a matter of enduring fascination [@Ellsworth2010-yu].
@@
 
-**Life Satisfaction** Life satisfaction is a measure of emotional well-being that was assessed using a 2-item version of the Satisfaction With Life Scale, which has previously been shown to correlate with aspects of religiosity [@Diener1985-xy].
+### Life Satisfaction
+Life satisfaction is a measure of emotional well-being that was assessed using a 2-item version of the Satisfaction With Life Scale, which has previously been shown to correlate with aspects of religiosity [@Diener1985-xy].
 Participants rated their agreement with the statements (a) "I am satisfied with my life"; and (b) "In most ways my life is close to ideal." The items were rated on 7-point response options ranging from 1 = strongly disagree to 7 = strongly agree.
@@
 
-**Personal well-being** Personal well-being is a measure of function all well-being that was assessed using a 4-item version of the Australian Unity well-being Index [@Cummins2017-ur].
+### Personal well-being 
+Personal well-being is a measure of function all well-being that was assessed using a 4-item version of the Australian Unity well-being Index [@Cummins2017-ur].
 Participants rated their satisfaction with (a) "Your standard of living"; (b) "Your health"; (c) "Your future security"; (d) "Your personal relationships." Items were rated using a 10-point response option ranging from 1 = completely dissatisfied to 10 = completely satisfied.
@@
 
-***Belief in a Spirit or a God***  We assess belief in a spirit or a God by asking two non-affective questions: "Do you believe in a spirit or life force" and "Do you believe in a god?" Responses were coded as (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who believe in a Spirit or Life Force; (3) Believers in a God who disbelieve in a Spirit or Life force; (4) Believers in a God and Spirit or Life Force.
+### Belief in a Spirit or a God  
+We assess belief in a spirit or a God by asking two non-affective questions: "Do you believe in a spirit or life force" and "Do you believe in a god?" Responses were coded as (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who believe in a Spirit or Life Force; (3) Believers in a God who disbelieve in a Spirit or Life force; (4) Believers in a God and Spirit or Life Force.
 These believe were developed for the NZAVS from the 2005 Eurobarometer.
@@
 
-**Age** Age was put into units of 10 years and centered at its mean. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 2.
+### Age 
+Age was put into units of 10 years and centered at its mean. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 2.
 Gender.
@@
 
-**Education**  Education level was measured using an 11-point rating developed by the New Zealand Qualification Authority known as the New Zealand Qualification Framework (NZQF; 0 = no qualification, 10 = doctoral degree).
+### Education
+Education level was measured using an 11-point rating developed by the New Zealand Qualification Authority known as the New Zealand Qualification Framework (NZQF; 0 = no qualification, 10 = doctoral degree).
 Education was centered at its mean and standardized.
@@
 
-**Deprivation** We measured the socio-economic status of participants’ immediate (small area) neighborhood using the 2013 New Zealand Deprivation Index, which uses aggregate census information about the residents of small neighborhood-type units to assign a decile-rank index from 1 (most affluent) to 10 (most impoverished) [@Atkinson2014-ex].
+### Deprivation
+We measured the socio-economic status of participants’ immediate (small area) neighborhood using the 2013 New Zealand Deprivation Index, which uses aggregate census information about the residents of small neighborhood-type units to assign a decile-rank index from 1 (most affluent) to 10 (most impoverished) [@Atkinson2014-ex].
 The index is based on a Principal Components Analysis of the following nine variables (in weighted order): the proportion of adults who received a means-tested benefit, household income, proportion not owning own home, proportion single-parent families, proportion unemployed, proportion lacking qualifications, proportion household crowding, proportion no telephone access, and proportion no car access.
@@
 
-**Employed** Employment status was assessed by asking participants if they were currently working, "yes" was coded as "1" and "no" was coded as "0." For detailed information pertaining to yearly responses and missingness, see Table 1.
+### Employed
+Employment status was assessed by asking participants if they were currently working, "yes" was coded as "1" and "no" was coded as "0." For detailed information pertaining to yearly responses and missingness, see Table 1.
 
-**Partner** Participants were asked if they were in a relationship, "yes" was coded as "1" and no was coded as "0." For detailed information pertaining to yearly responses and missingness, see Table 1.
+### Partner
+Participants were asked if they were in a relationship, "yes" was coded as "1" and no was coded as "0." For detailed information pertaining to yearly responses and missingness, see Table 1.
 
-**Ethnicity** Ethnicity was assessed using four basic categories: (1) New Zealand European/Pakeha, (2) Maori, (3) Pacific Islander, and (4) Asian.
+### Ethnicity
+Ethnicity was assessed using four basic categories: (1) New Zealand European/Pakeha, (2) Maori, (3) Pacific Islander, and (4) Asian.
 For detailed information pertaining to yearly responses and missingness, see Table 1.
 
-**Urban** People were coded as either residing in an urban "1" or rural "0" area based on New Zealand census data. For detailed information pertaining to yearly responses and missingness, see Table 1.
+### Urban
+People were coded as either residing in an urban "1" or rural "0" area based on New Zealand census data. For detailed information pertaining to yearly responses and missingness, see Table 1.
 
-**Political Orientation** To assess political orientation, we asked people to rate their political orientation using seven-point response options (1 = Liberal; 7 = Conservative).
+### Political Orientation
+To assess political orientation, we asked people to rate their political orientation using seven-point response options (1 = Liberal; 7 = Conservative).
 Higher values indicate more conservative political beliefs.
@@
 
-**Equation 1**
+## Equation 1
 
@@
 # Results
-
-```{r, cache =TRUE, include =FALSE}
-# model 2
-mod.1 <-lmer (PWI ~  1 +  Years * Beliefs +  Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id), data= dfben, na.action = na.omit)
-```
-```{r, cache=TRUE, include=FALSE}
-# model 1
-mod.1 <-lmer (LIFESAT ~ 1 + Years * Beliefs+ Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id), data= dfben, na.action = na.omit)
-summary(mod.1)
-```
-
-Results for both the personal well-being and life satisfaction models are presented in Table 2 and Figure 1.
-
-Figure 1 about here:
 
@@
 
-```{r, echo = FALSE, include=FALSE}
-te00 <-texreg::extract(
-  mod.1,
-  level = 0.95,
-  include.random = TRUE,
-  include.rsquared = T,
-  include.nobs = T,
-  include.loo.ic = T,
-  include.waic = F)
-texreg(list(te00),
-                custom.model.names = c("PWI"),
-                caption = "Personal Wellbeing",
-                sideways = F,
-                scalebox = .5,
-                #fontsize= "footnotesize",
-                label = "tab:REGRESS_1",
-                ci.force.level = 0.95, bold = 0.05,
-                settingstars = 0,
-                booktabs = TRUE,
-                custom.note ="")
-```
-
 ## Personal Well-Being
@@
 
-**Interactions**
+##  Personal well-being: years X beliefs
 
@@
 
-**Interactions**
+## Life Satisfaction: years X beliefs
 
@@
 
-<!-- ```{r} -->
-<!-- knitr::include_graphics("Figs/Usecoefficientplots-1.png") -->
-<!-- ``` -->
-
 ## Imputed data analysis
@@
 
+```{r fig.cap = "Imputed data analysis: coefficient plot"}
+knitr::include_graphics("Figs/coefficient_plot_both_imputed-1.png")
+```
+
 Figure 4 about here
+
+```{r fig.cap = "Imputed data analysis: predicted personal wellbeing by years X beliefs"}
+knitr::include_graphics("Figs/predicted_PWI_impute-1.png")
+```
+
+Figure 5 about here
+
+```{r fig.cap = "Imputed data analysis: predicted personal wellbeing by years X beliefs"}
+knitr::include_graphics("Figs/predicted_lifesat_imputed-1.png")
+```
 
@@
 Here, we ask whether beliefs in a Spirit or Life Force predict growth in personal well-being and life-satisfaction over eight years in populations who differ in the this belief, and who also differ in their belief in a God.
-We find that
+
+We find that \dots
+1. Lower subjective well-being among spirit beliefs population at outset of study, surpassed (LS) or surpassing (PWI) by the end of the interval
+2. Holds after adjusting for demograph
+3. Holds for Belief in God
+
+Questions:
+1. Why initially lower well-being in NZ? 
+2. Why the slope for spirit beliefs (and not God)?
+
+Importance:
+1. Methods: Disentangling
+2. Growing population: practical importance
 

### 2020-11-17 Manuscript.tex
fixed "figs" --> "Figs" and testing overleaf push.
- include a source file to make Rmd manuscript more manageable - revised text

@@
 Previous research finds an association between spirituality and subjective well-being.
-However, the widespread use of tautological scales and reliance on cross-sectional data cast doubts.
+However, the widespread use of tautological spirituality scales, poorly defined concepts of spirituality, and reliance on cross-sectional samples cast doubts.
 Here, we leverage nine years of panel data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction (\(N\) = 20979, New Zealand, 2010-2018).
-We find that belief in a spirit or life force predicts lower personal wellbeing and life-satisfaction.
-However, in support of previous theories, beliefs in a spirit or life force predict increasing personal well-being and life satisfaction over time relative to disbelief.
+Contrary to previous research, we find that belief in a spirit or life force predicts lower personal well-being and life-satisfaction.
+However, in support of previous speculation, beliefs in a spirit or life force predict increasing personal well-being and life satisfaction over time relative to disbelief.
 These finding holds after adjusting for known demographic influences, and intriguingly, among those who believe in a God but disbelieve in a spirit or life force.
-The recent growth in spiritual beliefs and decline in traditional religion across many industrial societies motivates futher causal nvestigations of the social and psychological mechanisms through which spiritual beliefs lead to growth in subjective wellbeing.
+The recent growth in spiritual beliefs and decline in traditional religion across many industrial societies motivates further causal investigations of the social and psychological mechanisms through which spiritual beliefs lead to growth in subjective well-being.
 }
@@
 The relationship between spirituality and psychological well-being is a matter of enduring fascination (Ellsworth \& Ellsworth, 2010).
-Previous attempts to quantify the relationship between these two domains reveal a positive association: spirituality predicts well-being (Ginsburg, Quirt, Ginsburg, \& MacKillop, 1995; Koenig, 2010; Koenig, Associate Professor of Psychology and Religious Studies Michael E McCullough, PhD, McCullough, Larson, \& Adjunct Professor of Psychiatry David B Larson, 2001; Russinova, Wewiorski, \& Cash, 2002; Smith, McCullough, \& Poll, 2003; Smith et al., 2003; Zaza, Sellick, \& Hillier, 2005).
+Most previous attempts to quantify the relationship between these two domains find a positive association: spirituality predicts well-being (Ginsburg, Quirt, Ginsburg, \& MacKillop, 1995; Koenig, 2010; Koenig, Associate Professor of Psychology and Religious Studies Michael E McCullough, PhD, McCullough, Larson, \& Adjunct Professor of Psychiatry David B Larson, 2001; Russinova, Wewiorski, \& Cash, 2002; Smith, McCullough, \& Poll, 2003; Smith et al., 2003; Zaza, Sellick, \& Hillier, 2005).
 However, critics raise three challenges.
@@
 Indeed, recent systematic-review of spirituality/mental health research found that nearly 45\% of previous studies employe tautological scales (Garssen et al., 2016).
-The first challenge, then, is to define measures of spirituality that are not well-being measures reclocked in other words.
+The first challenge, then, is to define measures of spirituality that are not well-being measures dressed in other words.
 
@@
 However, when spirituality has been operationalised as distinct from religion, some researchers have found a negative association within subjective well-being (King et al., 2013).
-Thus, in addition to disentangling the concept of spirituality from concepts of subjective well-being, it is also important to disentangel the concept of spirituality from concepts of religion.
+Thus, in addition to disentangling the concept of spirituality from concepts of subjective well-being, it is also important to disentangle the concept of spirituality from concepts of religion.
 
@@
 To assess whether spiritual beliefs lead to growth, stability, or decline in psychological well-being requires comparing requires repeatedly measuring individuals who differ in spirituality, demography, and religion as these people age, and investigating how their perceive their lives to be turning out (Garssen \& Visser, 2016).
-However, previous investigations of spirituality and wellbeing, including high-quality reviews and meta-analyses, have relied on cross-sectional samples (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky et al., 2005; Smith et al., 2003; Visser et al., 2010; Yonker et al., 2012).
-To address the key questions in spirituality psychological health research, then, requires longitudional investigations.
+However, previous investigations of spirituality and well-being, including high-quality reviews and meta-analyses, have relied on cross-sectional samples (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky et al., 2005; Smith et al., 2003; Visser et al., 2010; Yonker et al., 2012).
+To address the key questions in spirituality psychological health research, then, requires longitudinal investigations.
 
-Here, we addresses the three challenges of previous inthe following ways: First, we avoid a tautological elision of \enquote{spirituality} and \enquote{well-being} by adopting a non-affective measure of spirituality as a believe in a Spirit or Life Force.
-To measure subjective well-being we use a well-validated scale for Personal Well-being and a well-validated scale for Life-Satisfaction (see Method).
-This design avoids measuring spirituality with an indicator of spirituality that is cloacked as psychological well-being, thereby avoiding tautology.
-Second, we adopt a four-level categorical indicator for belief-states to clarify differences among people with spiritual, religious, and skeptical beliefs.
-We construct this four-level categorical indicator by combining responses to the Spiritual Belief measure with responses to the question: \enquote{Do you believe in some form of a spirit or life force?} The levels of our four-level categorical indicator of spiritual and religious beliefs are as follows: (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who nevertheless believe in a Spirit or Life Force (henceforth, SpiritExcludeGod); (3) Those who believe in a God but do not believe in a Spirit or Life force (henceforth, GodAndSpirit); (4) Those who who believe in a both God and a Spirit or Life Force, (henceforth GodExcludesSpirit).
-Narrowing the concept of spirituality to belief state comes at a price.
-Such an operationalisation, clearly, cannot simultaneously address all interests in previous research.
-However, this cost to scope is paid for with the benefits of precision.
-The spiritual indicators measure cognitive states in which distinct varieties of skeptical beliefs, traditional religious beliefs, and non-traditional spirit beliefs can are disentangled.
+Here, we addresses the three challenges of previous in the following ways: First, to avoid a tautological elision of \enquote{spirituality} and \enquote{well-being} we adopt a non-affective measure of spirituality as a \enquote{believe in a Spirit or Life Force.} On the other side, to measure subjective well-being, we use a well-validated scale for Personal Well-being and a well-validated scale for Life-Satisfaction (see Method).
+Second, to disentangle spiritual beliefs from traditional religious beliefs and skeptical beliefs we adopt a four-level categorical indicator combining the binary indicator of belief in a Spirit or Life Force with the binary indicator \enquote{Do you believe in some form of a spirit or life force?} The levels of this four-level categorical indicator for spiritual, religious, and skeptical beliefs are as follows: (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who nevertheless believe in a Spirit or Life Force; (3) Those who believe in a God but do not believe in a Spirit or Life force; (4) Those who who believe in a both God and a Spirit or Life Force.
+Clearly such an approach cannot simultaneously address all interests in previous spirituality/well-being research.
+For example we do not combine our cognitive measure of belief with behavioral measures or attitudinal measures.
+However, this cost to scope is paid for with benefits to precision.
+The spiritual indicators measure cognitive states in which distinct varieties of skeptical beliefs, traditional religious beliefs, and non-traditional spirit beliefs are straightforwardly disentangled, and measured within people.
 Third, to assess whether spirit beliefs lead to growth, stability, or decline in psychological well-being, we leverage longitudinal responses from a nationally diverse probability sample year over an nine--year period with nine measurement points (2010 to 2018 New Zealand Attitudes and Values Study, NZAVS).
@@
 
-\textbf{\emph{Belief in a Spirit or a God}} We assess belief in a spirit or a God by asking two non-affective questions: \enquote{Do you believe in a spirit or life force} and \enquote{Do you believe in a god?} Responses were coded as \enquote{Belief in a spirit that excludes a god} for those who indicated a belief in both a spirit or life force and a god, \enquote{Belief that includes a god} for those who indicated a belief in a god with or without an accompanying belief in a spirit or life force, and \enquote{skeptics} for those who indicated belief in neither a spirit or life force nor a god.
-These items appear in the NZAVS questionnaire and were developed for the 2005 Eurobarometer.
+\textbf{\emph{Belief in a Spirit or a God}} We assess belief in a spirit or a God by asking two non-affective questions: \enquote{Do you believe in a spirit or life force} and \enquote{Do you believe in a god?} Responses were coded as (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who believe in a Spirit or Life Force; (3) Believers in a God who disbelieve in a Spirit or Life force; (4) Believers in a God and Spirit or Life Force.
+These believe were developed for the NZAVS from the 2005 Eurobarometer.
 For detailed information pertaining to beliefs, see Table 1.
@@
 Statistical analysis was performed using R version 4.0.2 R version 3.6.1 (2020-06-22) 2019-07-05).
-We analyzed data from participants who responded to the NZAVS survey at least three four times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of 21,705 (N = 20,979 with complete responses)N = 15,040.
-To model predict growth in levels of self-reported psychological subjective well-being among these participants we use generalized linear mixed models.
-In our models, we account for the non-linear effects of time within participants by including its interaction with both of our predictor variables: \enquote{Belief in a spirit that excludes god} and \enquote{Belief that includes a god.} To handle the dependencies and heterogeneity introduced from the repeated measures, we We also included individual ID as an effect modelled as random-effect.
-To improve interpretability of our models, we centered and scaled the education, deprivation, and political orientation covariates.
-Additionally, age was centred at and put into decade-long intervals.
-The model equation(s) are described in equation 1.
+We analyze data from participants who responded to the NZAVS survey at least three four times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of 21,705 (N = 20,979 with complete responses).
+To model growth in subjective well-being we use generalized linear mixed models.
+In our models, we account for the non-linear effects of time within participants by including the interaction of years (a continuous variable) with Beliefs (a four-level categorical indicator).
+The baseline for this indicator is Skepticism about a Spirit/Life Forces and Skepticism about a God.
+To handle the dependencies and heterogeneity introduced from the repeated measures, we include individual ID as an effect modelled as random.
+We centered and scaled the education, deprivation, and political conservativism.
+We centred age was centred at and put into decade-long intervals.
+The model equation(s) are described in Equation 1.
+
+\hypertarget{equation-1}{%
+\subsection{Equation 1}\label{equation-1}}
 
@@
 
-All code for this analysis and a full report of the missing data imputation can be found at \url{https://github.com/jacopastorius/NZAVS_Highland_SpiritHealth2.Table} 1.
-Descriptive Statistics for Theoretical Variables in Each Wave.
-Belief in a god and belief in a spirit or life force were not measured during Wave 6 (2014).
-\enquote{BXG} indicates a belief in a spirit that excludes a god.
-\enquote{BIG} indicates a belief that includes a god.
-Many individuals appear in multiple waves.
-
-Table 2. Descriptive Statistics for Demographic Variables in Each Wave. Education level was not measured during Waves 2 (2010) and 3 (2011). Many individuals appear in multiple waves.
+All code for this analysis and a full report of the missing data imputation can be found at \url{https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing}.
 
@@
 
-Figure 2 about here:
-
-Results for both the personal well-being and life satisfaction models are presented in Table 2 and Figure 1.
-
 \begin{figure}
-
-{\centering \includegraphics[width=6.4in]{figs/USEcoefficientplots-1}
-
-}
-
-\caption{Combination Coefficient Plots for Personal Well-being and Life Satisfaction Spirit Belief Longitudinal Models}\label{fig:unnamed-chunk-5}
+\includegraphics[width=6.4in]{Figs/USEcoefficientplots-1} \caption{Combination Coefficient Plots for Personal Well-being and Life Satisfaction Spirit Belief Longitudinal Models}\label{fig:unnamed-chunk-1}
 \end{figure}
@@
 
-\textbf{Interactions}
+\hypertarget{personal-well-being-years-x-beliefs}{%
+\subsection{Personal well-being: years X beliefs}\label{personal-well-being-years-x-beliefs}}
 
 \begin{figure}
+\includegraphics[width=6.4in]{Figs/USEPWI_expected-1} \caption{Combination Coefficient Plots for Personal Well-being and Life Satisfaction Spirit Belief Longitudinal Models}\label{fig:unnamed-chunk-2}
+\end{figure}
 
-{\centering \includegraphics[width=6.4in]{Figs/USEPWI_expected-1}
-
-}
-
-\caption{Combination Coefficient Plots for Personal Well-being and Life Satisfaction Spirit Belief Longitudinal Models}\label{fig:unnamed-chunk-7}
-\end{figure}
+Focusing on the interaction of time with belief-type on personal well-being we observe:
 
@@
 \item
-  The effect of Years X Beliefs\_SpiritExcludesGod\_ is reliably positive (beta = 0.02, SE = 4.86e-03, std.
+  The effect of Years X SpiritExcludesGod\_ is reliably positive (beta = 0.02, SE = 4.86e-03, std.
 beta = 0.02, p \textless{} .01).
-We infer that, unlike the skeptic population, people with beliefs in a spirit or life force who do not believe in a God tend to grow in their personal well-beingwellbeing over time.
-
-  \begin{itemize}
-  \tightlist
+We infer that, unlike the skeptic population, people with beliefs in a spirit or life force who do not believe in a God tend to grow in their personal well-being over time.
 \item
-    The effect of Years X BeliefsGodAndSpirit is reliably positive (beta = 0.02, SE = 4.41e-03, std.
+  The effect of Years X GodAndSpirit is reliably positive (beta = 0.02, SE = 4.41e-03, std.
 beta = 0.03, p \textless{} .001).
-We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their personal well-beingwellbeing over time
+We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their personal well-being over time
 \item
-    The effect of Years X BeliefsGodExcludesSpirit is not reliable(beta = 1.96e-03, SE = 8.29e-03, std.
+  The effect of Years X GodExcludesSpirit is not reliable(beta = 1.96e-03, SE = 8.29e-03, std.
 beta = 2.36e-03, p = 0.813).
-Although people who believe in a God but do not believe in a spirit or Life Force tend to have greater personal well-beingwellbeing than those who do not believe in a spirit or life force (and greater life satisfaction than Skeptics) this population is not expected to grow in Life Satisfaction over time.
-  \end{itemize}
+Thus, God believers who doubt in a Spirit or Life Force start maintain higher personal well-being than do other groups, but they are not expected to grow in personal well-being over time.
 \end{itemize}
@@
 
-\textbf{Interactions}
+\hypertarget{life-satisfaction-years-x-beliefs}{%
+\subsection{Life Satisfaction: years X beliefs}\label{life-satisfaction-years-x-beliefs}}
 
 \begin{figure}
-
-{\centering \includegraphics[width=6.4in]{Figs/USElifsat_expected-1}
-
-}
-
-\caption{Prediction Plot Life Satisfaction Spirit Belief Longitudinal Models}\label{fig:unnamed-chunk-8}
+\includegraphics[width=6.4in]{Figs/USElifsat_expected-1} \caption{Prediction Plot Life Satisfaction Spirit Belief Longitudinal Models}\label{fig:unnamed-chunk-3}
 \end{figure}
@@
 beta = -2.70e-03, p = 0.799).
-Although people who believe in a God but do not believe in a spirit or Life Force tend to have greater life satisfaction than those who do not believe in a Spirit or life force (and greater life satisfaction than Skeptics) this population is not expected to grow in Life Satisfaction over time.
+Thus, God believers who doubt in a Spirit or Life Force start maintain higher Life-Satisfaction than do other groups, but they are not expected to grow in Life-Satisfaction over time.
   \end{itemize}
@@
 
-Figure 2. Subjective Well-being Predicted Probabilities for Belief in a Spirit or Life Force Over Time in a Listwise Deleted Dataset.
+\hypertarget{imputed-data-analysis}{%
+\subsection{Imputed data analysis}\label{imputed-data-analysis}}
 
-Table 3.
-Subjective Well-being Coefficient Table for Belief in a Spirit or Life Force.
-Age was put into units of 10 years and centered at its mean.
-Education, Deprivation, and Political Orientation were centered at their means and standardized.
+To adjust for potential biases from missing responses, we conducted a supplementary analysis in which we imputed missing values over the the longitudinal sample.
+This analysis is describe in Supplement 1.
+Averaging over the missing responses does not change inference, however we observe growth across all four groups in both personal well-being and life satisfaction.
+Additionally we observe substantially stronger effects for the slope of Spirit Beliefs on both personal well-being and life-satisfaction.
+Figures 3 and Figures 4 describe the predicted slopes of change.
+The inference from this analysis remains the same as from the analysis in which we delete missing values row-wise, by panel wave.
+In all analyses those who believe in a spirit or life-force are expected to experience greater growth in well-being over time.
+
+Figure 3 about here
+
+\begin{figure}
+\includegraphics[width=6.4in]{Figs/coefficient_plot_both_imputed-1} \caption{Imputed data analysis: coefficient plot}\label{fig:unnamed-chunk-4}
+\end{figure}
+
+Figure 4 about here
+
+\begin{figure}
+\includegraphics[width=6.4in]{Figs/predicted_PWI_impute-1} \caption{Imputed data analysis: predicted personal wellbeing by years X beliefs}\label{fig:unnamed-chunk-5}
+\end{figure}
+
+Figure 5 about here
+
+\begin{figure}
+\includegraphics[width=6.4in]{Figs/predicted_lifesat_imputed-1} \caption{Imputed data analysis: predicted personal wellbeing by years X beliefs}\label{fig:unnamed-chunk-6}
+\end{figure}
 
@@
 
-The purpose of this study has been to clarify the relationship between belief in a spirit/life force and psychological well-being.
-Spirituality as a distinct category, separate from religion, is a little-studied phenomenon in the psychology of religion literature.
+The purpose of this study has been to clarify the relationship between belief in a spirit/life force and subjective well-being, addressing three challenges from previous research: (1) tautological measures of spirituality and subjective well-being; (2) insufficiently distinct measures of spirituality and religion; (3) cross-sectional samples.
+Here, we ask whether beliefs in a Spirit or Life Force predict growth in personal well-being and life-satisfaction over eight years in populations who differ in the this belief, and who also differ in their belief in a God.
+We find that
 
-These findings demonstrate the importance ofinvestigating spiritual beliefs as a domain of interest in its own right.
+These findings demonstrate the importance of investigating spiritual beliefs as a domain of interest in its own right.
 In many Western countries, people are increasingly identifying as \enquote{spiritual but not religious} (Swinton, 2001; Zinnbauer \& Pargament, 2005).
@@
 Thus, in the long term, both sustained \enquote{BXG} and \enquote{BIG} may actually be associated with better subjective well-being.
+
 Our results also help to clarify the relationship among religion, spirituality, and subjective well-being outcomes.

### 2020-11-17 short.R

@@ -1,0 +1,79 @@
+library("styler")
+library("tidyverse")
+library("lme4")
+library("ggplot2")
+library("tidyr")
+library("ggeffects")
+library("Amelia")
+library("table1")
+library("patchwork")
+library("parameters")
+library("prettycode")
+library("table1")
+library("gghighlight")
+library("sjPlot")
+library("see")
+#library("viridis")
+#library("ggpubr")
+#library("wesanderson")
+library("ggsci")
+library("papaja")
+#library("here")
+library("citr")
+library("texreg")
+
+
+ldf.5<-readRDS("~/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/00Bulbulia\ Pubs/2020/ldf.5") 
+
+# filter only those who are measured 
+dfben  <-ldf.5 %>%
+  dplyr:: filter(Wave !=2009)%>% # exclude wave 1 (no spirit measures that wave)
+  dplyr::filter(YearMeasured==1)%>%
+  dplyr::group_by(Id) %>% filter(n() > 2)%>% # select those who have responded to at least 3 waves # Also works for 5 and 7
+  dplyr::filter(n() !=0)%>%
+  dplyr::ungroup(Id)%>%
+  dplyr:: mutate(Years = as.numeric(years))%>%
+  dplyr::mutate(Age.10yrs = (Age/10))%>%
+  dplyr::mutate(Religion.Church.Log = log(Religion.Church+ 1))%>%
+  dplyr::mutate(Pol.Orient.S = scale(Pol.Orient, center = TRUE, scale = TRUE),
+         Age.10yrs.C = scale(Age.10yrs, center = TRUE, scale = FALSE),
+         Male = factor(Gender),
+         Employed = factor(Employed),
+         Partner = factor(Partner),
+         Edu.S = scale(as.numeric(Edu) ,scale=TRUE,center=TRUE),
+         Education = as.numeric(Edu),
+         Urban = factor(Urban),
+         Believe.Spirit = as.factor(Believe.Spirit),
+         Believe.God = as.factor(Believe.God),
+         EthnicCategories = factor(EthnicCats),
+         LIFESAT.S = scale(LIFESAT, center = TRUE, scale = TRUE),
+         PWI.S = scale(PWI, center = TRUE, scale = TRUE),
+         Deprivation.S = scale(NZdep, scale=TRUE, center=TRUE),
[…]

### 2020-11-17 Analysis.Rmd
revisions to manuscript, worked out practical effect sizes

@@
 # Inference holds 
+library(lme4)
 mod.1 <-lmer (PWI ~  1 +  Years * Beliefs +  Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id), data= dfben, na.action = na.omit)
@@
 #### Predicted Effects ####
-pr.PWI <- ggpredict(model = mod.1, terms = c("Years [7:9]","Beliefs"), 
+pr.PWI <- ggpredict(model = mod.1, terms = c("Years [0:9]","Beliefs"), 
                     ci.lvl = 0.95,
@@
   ggtitle("Predicted Values of Personal Wellbeing")
+
+pr.PWI
 ```
@@
                         interval = "confidence")
+# print expected intervals for the y-hats reported in the conclusion
+pr.LIFESAT
+
 p2<-plot(pr.LIFESAT,facets = T) + 

### 2020-11-17 Manuscript.Rmd
revisions to manuscript, worked out practical effect sizes

@@
 ---
-title             : "National Longitudinal Study Reveals Growth in Wellbeing Among Those with Spiritual Beliefs"
-shorttitle        : "Beliefs and Wellbeing"
+title             : "National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Beliefs"
+shorttitle        : "Longitudinal Wellbeing from Spiritual Beliefs"
 author: 
@@
 abstract: |
+
   Previous research finds an association between spirituality and subjective well-being.
-However, the widespread use of tautological spirituality scales, poorly defined concepts of spirituality, and reliance on cross-sectional samples cast doubts.
-Here, we leverage nine years of panel data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction ($N$ = 20979, New Zealand, 2010-2018).
+However, the widespread use of tautological spirituality scales, poorly defined concepts of spirituality, and heavy reliance on cross-sectional samples cast doubts.
+Here, we leverage nine waves of panel data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction ($N$ = 20979, New Zealand, 2010-2018).
 Contrary to previous research, we find that belief in a spirit or life force predicts lower personal well-being and life-satisfaction.
 However, in support of previous speculation, beliefs in a spirit or life force predict increasing personal well-being and life satisfaction over time relative to disbelief.
-These finding holds after adjusting for known demographic influences, and intriguingly, among those who believe in a God but disbelieve in a spirit or life force.
-The recent growth in spiritual beliefs and decline in traditional religion across many industrial societies motivates further causal investigations of the social and psychological mechanisms through which spiritual beliefs lead to growth in subjective well-being.
+These finding are robust after to known demographic influences, and intriguingly, hold among those who believe in a God but disbelieve in a Spirit or Life Force.
+The recent growth in spiritual beliefs and decline in traditional religion across many industrial societies motivates further causal investigations of the mechanisms by which spiritual beliefs lead to growth in subjective well-being.
   <!-- https://tinyurl.com/ybremelq -->
@@
 keywords          : "Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being"
-wordcount         : "4778"
+wordcount         : "5015"
 bibliography      : ["b-references.bib"]
@@
 Most previous attempts to quantify the relationship between these two domains find a positive association: spirituality predicts well-being [@Russinova2002-rq; @Ginsburg1995-jr; @Zaza2005-ac; @Koenig2010-gk; @Smith2003-re; @Smith2003-re; @Koenig2001-ow].
-However, critics raise three challenges.
+However, critics have raised three credible challenges.
 
@@
 As Koenig observes, it is unsurprising that spirituality predict well-being when the same items are used to measure both constructs.
-Koenig aptly describes this features as a "tautology" in spirituality/mental health research [@Koenig2008-lv].
+Koenig aptly describes this approach as "tautological" [@Koenig2008-lv].
 Indeed, recent systematic-review of spirituality/mental health research found that nearly 45% of previous studies employe tautological scales [@Garssen2016-kb].
-The first challenge, then, is to define measures of spirituality that are not well-being measures dressed in other words.
+The first challenge, then, is to define measures of spirituality that are not well-being measures dressed up in other words.
 
@@
 The term "spirituality" was first used to describe the religious practices of ascetics and monks, however, more recently, the term has taken on a broader family of meanings [@Swinton2001-vr].
-However previous research does not converge on a unique set of meanings.
-Most studies of spirituality and religion have been conducted among (cross-sectional) North American samples among people who are affiliate with traditional religion [@Ano2005-hx; @Hackney2003-rs; @Sawatzky2005-rw; @Smith2003-re;  @Visser2010-kq; @Yonker2012-zg].
-However, when spirituality has been operationalised as distinct from religion, some researchers have found a negative association within subjective well-being [@King2013-cg].
-Thus, in addition to disentangling the concept of spirituality from concepts of subjective well-being, it is also important to disentangle the concept of spirituality from concepts of religion.
+Spirituality and health research does not converge on a unique set of meanings from the wider universe of popular meanings.
+Although it is commonplace across most areas of psychological science for researchers to explore different operationalisations of focal concepts, the psychology of spirituality has mostly been conducted with cross-sectional North American samples among participants who affiliate with traditional religion [@Ano2005-hx; @Hackney2003-rs; @Sawatzky2005-rw; @Smith2003-re;  @Visser2010-kq; @Yonker2012-zg].
+As such, the extent to which the psychology of spirituality and health has repeated or extended the psychology of religion and health remains unclear.
+Notably, researchers who have operationalised spirituality constructs as distinct from religion constructs have observed a negative association between spirituality and subjective well-being [@King2013-cg].
+Thus to infer how spirituality affects subjective well-being, it is important to disentangle the concepts of spirituality from concepts of religion, and to carefully investigate features of contructs in populations where these dimensions overlap, partially overlap, and diverge.
 
-A third challenge is to investigate the relationship between spirituality and subjective well-being over the life-span in nationally diverse samples.
-To assess whether spiritual beliefs lead to growth, stability, or decline in psychological well-being requires comparing requires repeatedly measuring individuals who differ in spirituality, demography, and religion as these people age, and investigating how their perceive their lives to be turning out [@Garssen2016-km].
-However, previous investigations of spirituality and well-being, including high-quality reviews and meta-analyses, have relied on cross-sectional samples [@Ano2005-hx; @Hackney2003-rs; @Sawatzky2005-rw; @Smith2003-re; @Visser2010-kq; @Yonker2012-zg].
-To address the key questions in spirituality psychological health research, then, requires longitudinal investigations.
+A third challenge is to investigate the relationship between spirituality and subjective well-being in nationally diverse samples over individual life-spans.
+Reliable inference requires observing change within people who differ dimensions of spirituality over time, and assing how their lives turn out [@Garssen2016-km].
+Previous investigations of spirituality and well-being, including high-quality reviews and meta-analyses, have relied on cross-sectional samples [@Ano2005-hx; @Hackney2003-rs; @Sawatzky2005-rw; @Smith2003-re; @Visser2010-kq; @Yonker2012-zg].
+However, To address the key questions in spirituality psychological health research requires longitudinal investigations.
+Unfortunately the relevant datasets are rare.
 
@@
 The Time 10 (2018) wave of the NZAVS included 47,951 respondents.
-We analyzed data from participants who responded to our survey at least three four times between Time 2 (2010) and Time 10 (2018), resulting in a sample of N = 21,705 (N = 20,979 with complete responses).
+We analyzed data from participants who responded to our survey at least three four times between Time 2 (2010) and Time 10 (2018), resulting in a sample of $N$ = 21,705 ($N$ = 20,979 with complete responses).
 
@@
 ## Statistical Analyses
-
 Statistical analysis was performed using R version 4.0.2 R version 3.6.1 (2020-06-22) 2019-07-05).
@@
 ## Equation 1
-
 \[
@@
 &\quad \beta_{15}(\operatorname{Urban}) + \beta_{16}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}}) + \beta_{17}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ + \\
-&\quad \beta_{18}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) \\ \alpha_{j} &\sim N \left(\mu_{\alpha_{j}},\sigma^2_{\alpha_{j}} \right) , \operatorname{ for  Id }~j~= 1, \dots~J = 20,979~\operatorname{Ids}
+&\quad \beta_{18}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) \\ \alpha_{j} &\sim N \left(\mu_{\alpha_{j}},\sigma^2_{\alpha_{j}} \right) , \operatorname{ for  Id }~j~= 1, \dots~J = 20,979~\operatorname{Individuals}
 \end{aligned}
@@
 
-We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict personal well-being (PWI) with Years (0-9), Beliefs (four categories), Age.10yrs.C, Deprivation.S, Edu.S, Employed, EthnicCats, Male, Partner, Pol.Orient.S and Urban, with  Id as random effects.
+We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict personal well-being (PWI) with Years (0-9), Beliefs (four categories), Age.10yrs.C, Deprivation.S, Edu.S, Employed, EthnicCats, Male, Partner, Pol.Orient.S and Urban, withId as random effects.
 The model included Id as random effects.
@@
 We infer that Skeptics personal well-being is not increasing in the Skeptic population
-  - The effect of Beliefs_SpiritExcludesGod_is reliably negative (beta = -0.14, SE = 0.03, std.
+  - The effect of Beliefs in SpiritExcludesGod is reliably negative (beta = -0.14, SE = 0.03, std.
 beta = -0.03, p < .001).
 We infer that overall the population that believes in a Spirit but does not believe in a God is expected to have a lower personal well-being than skeptics.
-  - The effect of BeliefsGodAndSpirit is reliably negative (beta = -0.16, SE = 0.03, std.
+  - The effect of Beliefs in GodAndSpirit is reliably negative (beta = -0.16, SE = 0.03, std.
 beta = -0.02, p < .001).
 We infer that overall the population that believes in God and a spirit or life force is expected to have a lower Personal Well-being than skeptics.
-  - The effect of BeliefsGodExcludesSpirit is not reliably different from the skeptic population (beta = 0.02, SE = 0.05, std. beta = 0.02, p = 0.669).
+  - The effect of Beliefs in GodExcludesSpirit is not reliably different from the skeptic population (beta = 0.02, SE = 0.05, std. beta = 0.02, p = 0.669).
   - The effect of Age.10yrs.C is reliably positive  (beta = 0.08, SE = 7.38e-03, std. beta = 0.07, p < .001). Older people have greater personal well-beingwellbeing.
-  - The effect of Deprivation.S is reliably negative (beta = -0.14, SE = 7.50e-03, std. beta = -0.08, p < .001). Deprived people have lower personal well-beingwellbeing.
-  - The effect of Edu.S is reliably positive  (beta = 0.12, SE = 8.74e-03, std. beta = 0.07, p < .001). Educated people greater personal well-beingwellbeing
+  - The effect of Deprivation (S) is reliably negative (beta = -0.14, SE = 7.50e-03, std. beta = -0.08, p < .001). Deprived people have lower personal well-beingwellbeing.
+  - The effect of Education (S) is reliably positive  (beta = 0.12, SE = 8.74e-03, std. beta = 0.07, p < .001). Educated people greater personal well-beingwellbeing
   - The effect of Employed is reliably positive  (beta = 0.09, SE = 0.01, std. beta = 0.05, p < .001). Employed people have greater personal well-beingwellbeing.
-  - The effect of EthnicCatsMaori is reliably negative (beta = -0.30, SE = 0.03, std. beta = -0.18, p < .001).
-  - The effect of EthnicCatsPacific is reliably negative (beta = -0.29, SE = 0.05, std. beta = -0.17, p < .001).
-  - The effect of EthnicCatsAsian is not reliable (beta = -0.08, SE = 0.05, std. beta = -0.05, p = 0.089).
-  - The effect of Male is reliably negative (beta = -0.11, SE = 0.02, std. beta = -0.06, p < .001).
-  - The effect of Having a Partner is reliably positive (beta = 0.61, SE = 0.02, std. beta = 0.37, p < .001).
-  - The effect of Pol.Orient.S is reliably positive  (beta = 0.02, SE = 5.67e-03, std. beta = 0.01, p < .001).
-  - The effect of Urban is reliably negative (beta = -0.02, SE = 0.01, std. beta = -0.01, p < .05).
+  - The effect of identifying as Maori is reliably negative (beta = -0.30, SE = 0.03, std. beta = -0.18, p < .001).
+  - The effect of identifying as Pacific is reliably negative (beta = -0.29, SE = 0.05, std. beta = -0.17, p < .001).
+  - The effect of identifying as Asian is not reliable (beta = -0.08, SE = 0.05, std. beta = -0.05, p = 0.089).
+  - The effect of identifying as Male is reliably negative (beta = -0.11, SE = 0.02, std. beta = -0.06, p < .001).
+  - The effect of having a Partner is reliably positive (beta = 0.61, SE = 0.02, std. beta = 0.37, p < .001).
+  - The effect of Political Conservativism (S) is reliably positive  (beta = 0.02, SE = 5.67e-03, std. beta = 0.01, p < .001).
+  - The effect of living in an Urban area is reliably negative (beta = -0.02, SE = 0.01, std. beta = -0.01, p < .05).
 
@@
 
-  - The effect of Years X SpiritExcludesGod_ is reliably positive  (beta = 0.02, SE = 4.86e-03, std.
+  - The effect of years $\times$ belief in a SpiritExcludesGod_ is reliably positive  (beta = 0.02, SE = 4.86e-03, std.
 beta = 0.02, p < .01).
 We infer that, unlike the skeptic population, people with beliefs in a spirit or life force who do not believe in a God tend to grow in their personal well-being over time.
-  - The effect of Years X GodAndSpirit is reliably positive  (beta = 0.02, SE = 4.41e-03, std.
+  - The effect of years $\times$ in GodAndSpirit is reliably positive  (beta = 0.02, SE = 4.41e-03, std.
 beta = 0.03, p < .001).
 We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their personal well-being over time
-  - The effect of Years X GodExcludesSpirit is not reliable(beta = 1.96e-03, SE = 8.29e-03, std.
+  - The effect of Years $\times$ belief in GodExcludesSpirit is not reliable(beta = 1.96e-03, SE = 8.29e-03, std.
 beta = 2.36e-03, p = 0.813).
@@
 
-  - The effect of Years is not reliable (beta = -5.50e-04, SE = 2.69e-03, std. beta = -9.32e-04, p = 0.838).  We do not infer growth over time in Life Satisfaction among the Skeptic population
-  - The effect of BeliefsSpiritExcludesGod  is reliably negative (beta = -0.08, SE = 0.02, std.
+  - The effect of years is not reliable (beta = -5.50e-04, SE = 2.69e-03, std. beta = -9.32e-04, p = 0.838).  We do not infer growth over time in Life Satisfaction among the Skeptic population
+  - The effect of beliefs in SpiritExcludesGodis reliably negative (beta = -0.08, SE = 0.02, std.
 beta = 0.01, p < .001).
 We infer that overall the population that believes in a Spirit but does not believe in a God is expected to have a lower Life Satisfaction than skeptics.
-  - The effect of BeliefsGodAndSpirit is not reliable (beta = -0.04, SE = 0.02, std. beta = 0.04, p = 0.105). We do not infer growth over time in Life Satisfaction among this (Skeptic) population
-  - The effect of BeliefsGodExcludesSpirit is reliably positive (beta = 0.07, SE = 0.04, std.
+  - The effect of beliefs in GodAndSpirit is not reliable (beta = -0.04, SE = 0.02, std. beta = 0.04, p = 0.105). We do not infer growth over time in Life Satisfaction among this (Skeptic) population
+  - The effect of beliefs in GodExcludesSpirit is reliably positive (beta = 0.07, SE = 0.04, std.
 beta = 0.05, p = 0.062).We infer that overall the population that believes in a God but does not believe in a spirit is not reliably different from  the Skeptics population in Life Satisfaction.
   - The effect of Age.10yrs.C is reliably positive (beta = 0.09, SE = 5.26e-03, std. beta = 0.10, p < .001). Older people have greater Life Satisfaction
-  - The effect of Deprivation.S is  is reliably negative (beta = -0.07, SE = 5.50e-03, std. beta = -0.06, p < .001). People who are more deprived have lower Life Satisfaction.
-  - The effect of Edu.S is reliably positive (beta = 0.06, SE = 6.33e-03, std. beta = 0.05, p < .001). Educated people have greater Life Satisfaction
-  - The effect of Employed is reliably positive (beta = 0.08, SE = 9.69e-03, std. beta = 0.07, p < .001). People who are employed tend to have greater Life Satisfaction
-  - The effect of EthnicCatsMaori is reliably negative (beta = -0.06, SE = 0.02, std. beta = -0.05, p < .01).  People who belong to the minority Maori group tend to have lower Life Satisfaction.
-  - The effect of EthnicCatsPacific is reliably negative (beta = -0.08, SE = 0.04, std.
+  - The effect of Deprivation (S) is  is reliably negative (beta = -0.07, SE = 5.50e-03, std. beta = -0.06, p < .001). People who are more deprived have lower Life Satisfaction.
+  - The effect of Education (S) is reliably positive (beta = 0.06, SE = 6.33e-03, std. beta = 0.05, p < .001). Educated people have greater Life Satisfaction
+  - The effect of being Employed is reliably positive (beta = 0.08, SE = 9.69e-03, std. beta = 0.07, p < .001). People who are employed tend to have greater Life Satisfaction
+  - The effect of identifying as Maori is reliably negative (beta = -0.06, SE = 0.02, std. beta = -0.05, p < .01).  People who belong to the minority Maori group tend to have lower Life Satisfaction.
+  - The effect of identifying as Pacific is reliably negative (beta = -0.08, SE = 0.04, std.
 beta = -0.07, p < .05).
 People who belong to the minority Pacific peoples group tend to have lower life satisfaction.
-  - The effect of EthnicCatsAsian is not reliably different from people of European ethnicity in life satisfaction(beta = -0.04, SE = 0.03, std. beta = -0.03, p = 0.223).
-  - The effect of Male is reliably negative (beta = -0.19, SE = 0.01, std. beta = -0.16, p < .001).
-  - The effect of HasPartner is reliably positive (beta = 0.42, SE = 0.01, std. beta = 0.35, p < .001).
-  - The effect of Pol.Orient.S is reliably positive (beta = 0.01, SE = 4.23e-03, std. beta = 0.01, p < .01).
-  - The effect of Urban is reliably negative (beta = -0.02, SE = 8.85e-03, std. beta = -0.02, p < .05).
+  - The effect of identifying as Asian is not reliably different from people of European ethnicity in life satisfaction(beta = -0.04, SE = 0.03, std. beta = -0.03, p = 0.223).
+  - The effect of identifying as Male is reliably negative (beta = -0.19, SE = 0.01, std. beta = -0.16, p < .001).
+  - The effect of having a Parnter is reliably positive (beta = 0.42, SE = 0.01, std. beta = 0.35, p < .001).
+  - The effect of Political Conservativism (S) is reliably positive (beta = 0.01, SE = 4.23e-03, std. beta = 0.01, p < .01).
+  - The effect of living in an Urban area is reliably negative (beta = -0.02, SE = 8.85e-03, std. beta = -0.02, p < .05).
 
@@
 ```
-  - The interaction of Years X Beliefs_SpiritExcludesGod_ is reliably positive (beta = 0.02, SE = 3.66e-03, std.
+  - The interaction of years $\times$ beliefs in SpiritExcludesGod is reliably positive (beta = 0.02, SE = 3.66e-03, std.
 beta = 0.03, p < .001).
 We infer that, unlike the Skeptic population, people with beliefs in a Spirit or Life Force who do not believe in a God tend to grow in their Life Satisfaction over time.
-  - The interaction of Years X BeliefsGodAndSpirit is reliably positive (beta = 0.01, SE = 3.32e-03, std.
+  - The interaction of years $\times$ beliefsGodAndSpirit is reliably positive (beta = 0.01, SE = 3.32e-03, std.
 beta = 0.02, p < .001).
 We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their Life Satisfaction over time
-  - The effect of Years X BeliefsGodExcludesSpirit is not reliable (beta = -1.59e-03, SE = 6.25e-03, std.
+  - The effect of years $\times$ beliefs in GodExcludesSpirit is not reliable (beta = -1.59e-03, SE = 6.25e-03, std.
 beta = -2.70e-03, p = 0.799).
@@
 ## Imputed data analysis
-
 To adjust for potential biases from missing responses, we conducted a supplementary analysis in which we imputed missing values over the the longitudinal sample.
@@
 
-Figure 3 about here
+Figure 3 about here:
 
@@
 
-Figure 4 about here
+Figure 4 about here:
 
@@
 
-Figure 5 about here
+Figure 5 about here:
 
@@
 The purpose of this study has been to clarify the relationship between belief in a spirit/life force and subjective well-being, addressing three challenges from previous research: (1) tautological measures of spirituality and subjective well-being; (2) insufficiently distinct measures of spirituality and religion; (3) cross-sectional samples.
-Here, we ask whether beliefs in a Spirit or Life Force predict growth in personal well-being and life-satisfaction over eight years in populations who differ in the this belief, and who also differ in their belief in a God.
+Here, we ask whether beliefs in a Spirit or Life Force predict growth in personal well-being and life-satisfaction over 9 waves (8 years) in nationally diverse western populations who differ in the dimensions of spiritual and religious beliefs.
+We observe that:
 
-We find that \dots
-1. Lower subjective well-being among spirit beliefs population at outset of study, surpassed (LS) or surpassing (PWI) by the end of the interval
-2. Holds after adjusting for demograph
-3. Holds for Belief in God
+1.
+At the outset of the study, in 2010, the expected subjective well-being as measured using the Personal Well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God ($\hat{y}$ =  6.50, se =.03) and lower among those who believed in both a Spirit/Life-Force and a God ($\hat{y}$ =  6.48, se =.03) in comparison both to the skeptical population who neither believed in a God nor a life force ($\hat{y}$ = 6.64,se =.03) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 6.66, se =.05).
+(See Figure 2)
+
+Similarly,the expected subjective well-being as measured using the Life Satisfaction Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God ($\hat{y}$ = 4.86, se =.02) and lower among those who believed in both a Spirit/Life-Force and a God  ($\hat{y}$ = 4.91, se =.02) in comparison both with the skeptical population who neither believed in a God nor a life force ($\hat{y}$ = 4.94,se =.02) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 5.02, se =.02).
+(See Figure 3)
+
+2.
+However, over time, belief in a Spirit or Life Force predicted a trajectory of growth in Personal Well-being.
+In 2018, the expected subjective well-being as measured using the Personal Well-being Inventory among those who believed in Spirit or Life Force but not a God was expected to grow to ($\hat{y}$ = 6.67, se =.03, $\Delta \hat{y} = 0.17$).
+Among those who believed in both a Spirit/Life-Force and a God Personal Well-Being was expected to grow to ($\hat{y}$ = 6.69, se = 0.02,$\Delta \hat{y} = .21$).
+In practical terms, such growth from spiritual beliefs among to the expected gain in Personal Well-being about twice as large as the expected benefit of being employed and about about a third as large as the expected benefit of having a romantic partner.
+In comparison, Personal Well-Being among skeptical population who neither believed in a God nor a Life Force remained stable $\hat{y} = 6.66, se = 0.03, \Delta \hat{y} = .02$).
+We also observe stability in Personal Well-being among the population that believed in God but was skeptical about a Spirit or Life Force $\hat{y} =  6.70, se = 0.04, \Delta \hat{y} = .04$).
+(See Figure 2).
+
+Similarly, over time, belief in a Spirit or Life Force predicted a trajectory of growth in Life Satisfaction.
+In 2018, the expected subjective well-being as measured using the Life Satisfaction Inventory among those who believed in Spirit or Life Force but not a God was expected to grow to ($\hat{y} = 5.01, se = 0.02, \Delta \hat{y} = 0.07$).
+In practical terms, such growth from spiritual beliefs is approximately equivalent to the expected gain in Life Satisfaction from being employed and a sixth as large as the expected benefit to Life Satisfaction from having a romantic partner.
+Among those who believed in both a Spirit/Life-Force and a God Personal Well-Being was expected to grow to ($\hat{y} = 5.03, se = 0.02, \Delta \hat{y} = 0.12$).
+In practical terms, such growth from spiritual beliefs is approximately equivalent to the 1.5 times the expected gain in Life Satisfaction from being employed and almost a third the expected benefit to Life Satisfaction from having a romantic partner.
+In comparison, Life Satisfaction among skeptical population who neither believed in a God nor a Life Force remained stable $\hat{y} =  4.94, se =0.02, \Delta \hat{y} = 0$).
+We also observe stability in Personal Well-being among the population that believed in God but was skeptical about a Spirit or Life Force $\hat{y} =  5.00, se = 0.03, \Delta \hat{y} = -.02$) (See Figure 3).
+
+3.
+Such observed growth in subjective well-being among the population that believes in a Spirit or Life Force holds even after adjusting for demographic factors that are known to affect well-being.
+Additionally we infer growth from  9 annual waves of data collection (8 years) in a nationally diverse sample.
+On this basis of this window of time and sample, we infer that such growth in Life Satisifacton and in Personal Well-being reflects changes in the populations that believed in Spirit or Life Force over that period.
 
 Questions:
-1. Why initially lower well-being in NZ? 
-2. Why the slope for spirit beliefs (and not God)?
-
-Importance:
-1. Methods: Disentangling
-2. Growing population: practical importance
-
-These findings demonstrate the importance of investigating spiritual beliefs as a domain of interest in its own right.
-In many Western countries, people are increasingly identifying as "spiritual but not religious" [@Zinnbauer2005-vz; @Swinton2001-vr].
-Other studies find that those who have a spiritual understanding of life without a religious framework are at an increased risk of having mental disorders [@King2013-cg].
-As such, it is important to understand the ramifications of belief in a spirit for psychological health.
-To assess the relationship between a belief in a spirit or life force and subjective well-being, we improve upon past methods in three ways.
-
-Despite the opposition of our findings to those of many previous studies, there is room for agreement.
-We show that over time both "BXG" and "BIG" are associated with marginally better subjective well-being.
-As seen in Figure 2., both "BXG" and "BIG" predict marginally higher personal well-being and life satisfaction as time progresses.
-So, while those who adhere to these beliefs may start out below their skeptic counterparts in terms of their subjective well-being, they do eventually catch up.
-We also find that their life satisfaction is actually improving at a greater rate than skeptics.
-Thus, in the long term, both sustained "BXG" and "BIG" may actually be associated with better subjective well-being.
-
-Our results also help to clarify the relationship among religion, spirituality, and subjective well-being outcomes.
-BXG and BIG performed almost identically in terms of their ability to predict subjective well-being outcomes.
-This trend suggests that any kind of existential belief - regardless of whether a god is involved - might affect subjective well-being in much the same way.
-This result is good news as increasing numbers of individuals continue to report spiritual but not necessarily religious beliefs.
-
+1. Why initially lower well-being in NZ? $\dots$
+  + historical contingencies? 
+  + possibility of non-linear futures, are we observing a regression to the mean?
+  + individual factors: earthquakes, however..why the slope for spirit beliefs (and not God)?
+  + other things besides beliefs? behaviors, social interactions, perspectives... 
   Taken as a whole, our results cannot refute the entire narrative of the positive relationship between spirituality and emotional well-being.
 First, our analysis focuses only on belief.
-While belief is an important part of spirituality it is not by any means spirituality itself.
+Although belief is an important part of spirituality it is not by any means spirituality itself.
 Individuals may identify as spiritual without any accompanying belief in a spirit or life force.
@@
 Overall, though, our findings add to the existing literature, helping to untangle the multifaceted relationship among religion, spirituality, and subjective well-being.
+$\dots$ fruitful horizons for future research.
+
+Importance:
+1. Methods: Disentangling $\dots$ concepts/ longitudinal 
+2. Growing population: practical importance. In many Western countries, people are increasingly identifying as "spiritual but not religious" [@Zinnbauer2005-vz; @Swinton2001-vr]. ... 
+ + risks: Risks? ... Other studies find that those who have a spiritual understanding of life without a religious framework are at an increased risk of having mental disorders [@King2013-cg].
+ + However, the observed growth suggests that any kind of existential belief - regardless of whether a god is involved - might affect subjective well-being in much the same way.
+This result is good news as increasing numbers of individuals continue to report spiritual but not necessarily religious beliefs.
 

### 2020-11-17 Manuscript.tex
revisions to manuscript, worked out practical effect sizes

@@
 \hypersetup{
-  pdftitle={National Longitudinal Study Reveals Growth in Wellbeing Among Those with Spiritual Beliefs},
+  pdftitle={National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Beliefs},
   pdfauthor={Ben Highland1, Everett L. Worthington2, Don E Davis3, Chris G. Sibley4, \& Joseph Bulbulia5},
@@
 \makeatother
-\shorttitle{Beliefs and Wellbeing}
-\keywords{Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being\newline\indent Word count: 4778}
+\shorttitle{Longitudinal Wellbeing from Spiritual Beliefs}
+\keywords{Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being\newline\indent Word count: 5015}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 
-\title{National Longitudinal Study Reveals Growth in Wellbeing Among Those with Spiritual Beliefs}
+\title{National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Beliefs}
 \author{Ben Highland\textsuperscript{1}, Everett L. Worthington\textsuperscript{2}, Don E Davis\textsuperscript{3}, Chris G. Sibley\textsuperscript{4}, \& Joseph Bulbulia\textsuperscript{5}}
@@
 Previous research finds an association between spirituality and subjective well-being.
-However, the widespread use of tautological spirituality scales, poorly defined concepts of spirituality, and reliance on cross-sectional samples cast doubts.
-Here, we leverage nine years of panel data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction (\(N\) = 20979, New Zealand, 2010-2018).
+However, the widespread use of tautological spirituality scales, poorly defined concepts of spirituality, and heavy reliance on cross-sectional samples cast doubts.
+Here, we leverage nine waves of panel data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction (\(N\) = 20979, New Zealand, 2010-2018).
 Contrary to previous research, we find that belief in a spirit or life force predicts lower personal well-being and life-satisfaction.
 However, in support of previous speculation, beliefs in a spirit or life force predict increasing personal well-being and life satisfaction over time relative to disbelief.
-These finding holds after adjusting for known demographic influences, and intriguingly, among those who believe in a God but disbelieve in a spirit or life force.
-The recent growth in spiritual beliefs and decline in traditional religion across many industrial societies motivates further causal investigations of the social and psychological mechanisms through which spiritual beliefs lead to growth in subjective well-being.
+These finding are robust after to known demographic influences, and intriguingly, hold among those who believe in a God but disbelieve in a Spirit or Life Force.
+The recent growth in spiritual beliefs and decline in traditional religion across many industrial societies motivates further causal investigations of the mechanisms by which spiritual beliefs lead to growth in subjective well-being.
 }
@@
 Most previous attempts to quantify the relationship between these two domains find a positive association: spirituality predicts well-being (Ginsburg, Quirt, Ginsburg, \& MacKillop, 1995; Koenig, 2010; Koenig, Associate Professor of Psychology and Religious Studies Michael E McCullough, PhD, McCullough, Larson, \& Adjunct Professor of Psychiatry David B Larson, 2001; Russinova, Wewiorski, \& Cash, 2002; Smith, McCullough, \& Poll, 2003; Smith et al., 2003; Zaza, Sellick, \& Hillier, 2005).
-However, critics raise three challenges.
+However, critics have raised three credible challenges.
 
@@
 As Koenig observes, it is unsurprising that spirituality predict well-being when the same items are used to measure both constructs.
-Koenig aptly describes this features as a \enquote{tautology} in spirituality/mental health research (Koenig, 2008).
+Koenig aptly describes this approach as \enquote{tautological} (Koenig, 2008).
 Indeed, recent systematic-review of spirituality/mental health research found that nearly 45\% of previous studies employe tautological scales (Garssen et al., 2016).
-The first challenge, then, is to define measures of spirituality that are not well-being measures dressed in other words.
+The first challenge, then, is to define measures of spirituality that are not well-being measures dressed up in other words.
 
@@
 The term \enquote{spirituality} was first used to describe the religious practices of ascetics and monks, however, more recently, the term has taken on a broader family of meanings (Swinton, 2001).
-However previous research does not converge on a unique set of meanings.
-Most studies of spirituality and religion have been conducted among (cross-sectional) North American samples among people who are affiliate with traditional religion (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky, Ratner, \& Chiu, 2005; Smith et al., 2003; Visser, Garssen, \& Vingerhoets, 2010; Yonker, Schnabelrauch, \& Dehaan, 2012).
-However, when spirituality has been operationalised as distinct from religion, some researchers have found a negative association within subjective well-being (King et al., 2013).
-Thus, in addition to disentangling the concept of spirituality from concepts of subjective well-being, it is also important to disentangle the concept of spirituality from concepts of religion.
+Spirituality and health research does not converge on a unique set of meanings from the wider universe of popular meanings.
+Although it is commonplace across most areas of psychological science for researchers to explore different operationalisations of focal concepts, the psychology of spirituality has mostly been conducted with cross-sectional North American samples among participants who affiliate with traditional religion (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky, Ratner, \& Chiu, 2005; Smith et al., 2003; Visser, Garssen, \& Vingerhoets, 2010; Yonker, Schnabelrauch, \& Dehaan, 2012).
+As such, the extent to which the psychology of spirituality and health has repeated or extended the psychology of religion and health remains unclear.
+Notably, researchers who have operationalised spirituality constructs as distinct from religion constructs have observed a negative association between spirituality and subjective well-being (King et al., 2013).
+Thus to infer how spirituality affects subjective well-being, it is important to disentangle the concepts of spirituality from concepts of religion, and to carefully investigate features of contructs in populations where these dimensions overlap, partially overlap, and diverge.
 
-A third challenge is to investigate the relationship between spirituality and subjective well-being over the life-span in nationally diverse samples.
-To assess whether spiritual beliefs lead to growth, stability, or decline in psychological well-being requires comparing requires repeatedly measuring individuals who differ in spirituality, demography, and religion as these people age, and investigating how their perceive their lives to be turning out (Garssen \& Visser, 2016).
-However, previous investigations of spirituality and well-being, including high-quality reviews and meta-analyses, have relied on cross-sectional samples (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky et al., 2005; Smith et al., 2003; Visser et al., 2010; Yonker et al., 2012).
-To address the key questions in spirituality psychological health research, then, requires longitudinal investigations.
+A third challenge is to investigate the relationship between spirituality and subjective well-being in nationally diverse samples over individual life-spans.
+Reliable inference requires observing change within people who differ dimensions of spirituality over time, and assing how their lives turn out (Garssen \& Visser, 2016).
+Previous investigations of spirituality and well-being, including high-quality reviews and meta-analyses, have relied on cross-sectional samples (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky et al., 2005; Smith et al., 2003; Visser et al., 2010; Yonker et al., 2012).
+However, To address the key questions in spirituality psychological health research requires longitudinal investigations.
+Unfortunately the relevant datasets are rare.
 
@@
 The Time 10 (2018) wave of the NZAVS included 47,951 respondents.
-We analyzed data from participants who responded to our survey at least three four times between Time 2 (2010) and Time 10 (2018), resulting in a sample of N = 21,705 (N = 20,979 with complete responses).
+We analyzed data from participants who responded to our survey at least three four times between Time 2 (2010) and Time 10 (2018), resulting in a sample of \(N\) = 21,705 (\(N\) = 20,979 with complete responses).
 
@@
 
-\textbf{Life Satisfaction} Life satisfaction is a measure of emotional well-being that was assessed using a 2-item version of the Satisfaction With Life Scale, which has previously been shown to correlate with aspects of religiosity (Diener, Emmons, Larsen, \& Griffin, 1985).
+\hypertarget{life-satisfaction}{%
+\subsubsection{Life Satisfaction}\label{life-satisfaction}}
+
+Life satisfaction is a measure of emotional well-being that was assessed using a 2-item version of the Satisfaction With Life Scale, which has previously been shown to correlate with aspects of religiosity (Diener, Emmons, Larsen, \& Griffin, 1985).
 Participants rated their agreement with the statements (a) \enquote{I am satisfied with my life}; and (b) \enquote{In most ways my life is close to ideal.} The items were rated on 7-point response options ranging from 1 = strongly disagree to 7 = strongly agree.
@@
 
-\textbf{Personal well-being} Personal well-being is a measure of function all well-being that was assessed using a 4-item version of the Australian Unity well-being Index (Cummins, Eckerseley, Pallant, Van Vugt, \& Misajon, 2017).
+\hypertarget{personal-well-being}{%
+\subsubsection{Personal well-being}\label{personal-well-being}}
+
+Personal well-being is a measure of function all well-being that was assessed using a 4-item version of the Australian Unity well-being Index (Cummins, Eckerseley, Pallant, Van Vugt, \& Misajon, 2017).
 Participants rated their satisfaction with (a) \enquote{Your standard of living}; (b) \enquote{Your health}; (c) \enquote{Your future security}; (d) \enquote{Your personal relationships.} Items were rated using a 10-point response option ranging from 1 = completely dissatisfied to 10 = completely satisfied.
@@
 
-\textbf{\emph{Belief in a Spirit or a God}} We assess belief in a spirit or a God by asking two non-affective questions: \enquote{Do you believe in a spirit or life force} and \enquote{Do you believe in a god?} Responses were coded as (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who believe in a Spirit or Life Force; (3) Believers in a God who disbelieve in a Spirit or Life force; (4) Believers in a God and Spirit or Life Force.
+\hypertarget{belief-in-a-spirit-or-a-god}{%
+\subsubsection{Belief in a Spirit or a God}\label{belief-in-a-spirit-or-a-god}}
+
+We assess belief in a spirit or a God by asking two non-affective questions: \enquote{Do you believe in a spirit or life force} and \enquote{Do you believe in a god?} Responses were coded as (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who believe in a Spirit or Life Force; (3) Believers in a God who disbelieve in a Spirit or Life force; (4) Believers in a God and Spirit or Life Force.
 These believe were developed for the NZAVS from the 2005 Eurobarometer.
@@
 
-\textbf{Age} Age was put into units of 10 years and centered at its mean. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 2.
+\hypertarget{age}{%
+\subsubsection{Age}\label{age}}
+
+Age was put into units of 10 years and centered at its mean. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 2.
 Gender.
@@
 
-\textbf{Education} Education level was measured using an 11-point rating developed by the New Zealand Qualification Authority known as the New Zealand Qualification Framework (NZQF; 0 = no qualification, 10 = doctoral degree).
+\hypertarget{education}{%
+\subsubsection{Education}\label{education}}
+
+Education level was measured using an 11-point rating developed by the New Zealand Qualification Authority known as the New Zealand Qualification Framework (NZQF; 0 = no qualification, 10 = doctoral degree).
 Education was centered at its mean and standardized.
@@
 
-\textbf{Deprivation} We measured the socio-economic status of participants' immediate (small area) neighborhood using the 2013 New Zealand Deprivation Index, which uses aggregate census information about the residents of small neighborhood-type units to assign a decile-rank index from 1 (most affluent) to 10 (most impoverished) (Atkinson, Salmond, \& Crampton, 2014).
+\hypertarget{deprivation}{%
+\subsubsection{Deprivation}\label{deprivation}}
+
+We measured the socio-economic status of participants' immediate (small area) neighborhood using the 2013 New Zealand Deprivation Index, which uses aggregate census information about the residents of small neighborhood-type units to assign a decile-rank index from 1 (most affluent) to 10 (most impoverished) (Atkinson, Salmond, \& Crampton, 2014).
 The index is based on a Principal Components Analysis of the following nine variables (in weighted order): the proportion of adults who received a means-tested benefit, household income, proportion not owning own home, proportion single-parent families, proportion unemployed, proportion lacking qualifications, proportion household crowding, proportion no telephone access, and proportion no car access.
@@
 
-\textbf{Employed} Employment status was assessed by asking participants if they were currently working, \enquote{yes} was coded as \enquote{1} and \enquote{no} was coded as \enquote{0.} For detailed information pertaining to yearly responses and missingness, see Table 1.
+\hypertarget{employed}{%
+\subsubsection{Employed}\label{employed}}
 
-\textbf{Partner} Participants were asked if they were in a relationship, \enquote{yes} was coded as \enquote{1} and no was coded as \enquote{0.} For detailed information pertaining to yearly responses and missingness, see Table 1.
+Employment status was assessed by asking participants if they were currently working, \enquote{yes} was coded as \enquote{1} and \enquote{no} was coded as \enquote{0.} For detailed information pertaining to yearly responses and missingness, see Table 1.
 
-\textbf{Ethnicity} Ethnicity was assessed using four basic categories: (1) New Zealand European/Pakeha, (2) Maori, (3) Pacific Islander, and (4) Asian.
+\hypertarget{partner}{%
+\subsubsection{Partner}\label{partner}}
+
+Participants were asked if they were in a relationship, \enquote{yes} was coded as \enquote{1} and no was coded as \enquote{0.} For detailed information pertaining to yearly responses and missingness, see Table 1.
+
+\hypertarget{ethnicity}{%
+\subsubsection{Ethnicity}\label{ethnicity}}
+
+Ethnicity was assessed using four basic categories: (1) New Zealand European/Pakeha, (2) Maori, (3) Pacific Islander, and (4) Asian.
 For detailed information pertaining to yearly responses and missingness, see Table 1.
 
-\textbf{Urban} People were coded as either residing in an urban \enquote{1} or rural \enquote{0} area based on New Zealand census data.
+\hypertarget{urban}{%
+\subsubsection{Urban}\label{urban}}
+
+People were coded as either residing in an urban \enquote{1} or rural \enquote{0} area based on New Zealand census data.
 For detailed information pertaining to yearly responses and missingness, see Table 1.
 
-\textbf{Political Orientation} To assess political orientation, we asked people to rate their political orientation using seven-point response options (1 = Liberal; 7 = Conservative).
+\hypertarget{political-orientation}{%
+\subsubsection{Political Orientation}\label{political-orientation}}
+
+To assess political orientation, we asked people to rate their political orientation using seven-point response options (1 = Liberal; 7 = Conservative).
 Higher values indicate more conservative political beliefs.
@@
 
-\hypertarget{personal-well-being}{%
-\subsection{Personal Well-Being}\label{personal-well-being}}
+\hypertarget{personal-well-being-1}{%
+\subsection{Personal Well-Being}\label{personal-well-being-1}}
 
-We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict personal well-being (PWI) with Years (0-9), Beliefs (four categories), Age.10yrs.C, Deprivation.S, Edu.S, Employed, EthnicCats, Male, Partner, Pol.Orient.S and Urban, with Id as random effects.
+We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict personal well-being (PWI) with Years (0-9), Beliefs (four categories), Age.10yrs.C, Deprivation.S, Edu.S, Employed, EthnicCats, Male, Partner, Pol.Orient.S and Urban, withId as random effects.
 The model included Id as random effects.
@@
 \item
-  The effect of Beliefs\_SpiritExcludesGod\_is reliably negative (beta = -0.14, SE = 0.03, std.
+  The effect of Beliefs in SpiritExcludesGod is reliably negative (beta = -0.14, SE = 0.03, std.
 beta = -0.03, p \textless{} .001).
@@
 \item
-  The effect of BeliefsGodAndSpirit is reliably negative (beta = -0.16, SE = 0.03, std.
+  The effect of Beliefs in GodAndSpirit is reliably negative (beta = -0.16, SE = 0.03, std.
 beta = -0.02, p \textless{} .001).
@@
 \item
-  The effect of BeliefsGodExcludesSpirit is not reliably different from the skeptic population (beta = 0.02, SE = 0.05, std. beta = 0.02, p = 0.669).
+  The effect of Beliefs in GodExcludesSpirit is not reliably different from the skeptic population (beta = 0.02, SE = 0.05, std. beta = 0.02, p = 0.669).
 \item
@@
 \item
-  The effect of Deprivation.S is reliably negative (beta = -0.14, SE = 7.50e-03, std. beta = -0.08, p \textless{} .001). Deprived people have lower personal well-beingwellbeing.
+  The effect of Deprivation (S) is reliably negative (beta = -0.14, SE = 7.50e-03, std. beta = -0.08, p \textless{} .001). Deprived people have lower personal well-beingwellbeing.
 \item
-  The effect of Edu.S is reliably positive (beta = 0.12, SE = 8.74e-03, std. beta = 0.07, p \textless{} .001). Educated people greater personal well-beingwellbeing
+  The effect of Education (S) is reliably positive (beta = 0.12, SE = 8.74e-03, std. beta = 0.07, p \textless{} .001). Educated people greater personal well-beingwellbeing
 \item
@@
 \item
-  The effect of EthnicCatsMaori is reliably negative (beta = -0.30, SE = 0.03, std. beta = -0.18, p \textless{} .001).
+  The effect of identifying as Maori is reliably negative (beta = -0.30, SE = 0.03, std. beta = -0.18, p \textless{} .001).
 \item
-  The effect of EthnicCatsPacific is reliably negative (beta = -0.29, SE = 0.05, std. beta = -0.17, p \textless{} .001).
+  The effect of identifying as Pacific is reliably negative (beta = -0.29, SE = 0.05, std. beta = -0.17, p \textless{} .001).
 \item
-  The effect of EthnicCatsAsian is not reliable (beta = -0.08, SE = 0.05, std. beta = -0.05, p = 0.089).
+  The effect of identifying as Asian is not reliable (beta = -0.08, SE = 0.05, std. beta = -0.05, p = 0.089).
 \item
-  The effect of Male is reliably negative (beta = -0.11, SE = 0.02, std. beta = -0.06, p \textless{} .001).
+  The effect of identifying as Male is reliably negative (beta = -0.11, SE = 0.02, std. beta = -0.06, p \textless{} .001).
 \item
-  The effect of Having a Partner is reliably positive (beta = 0.61, SE = 0.02, std. beta = 0.37, p \textless{} .001).
+  The effect of having a Partner is reliably positive (beta = 0.61, SE = 0.02, std. beta = 0.37, p \textless{} .001).
 \item
-  The effect of Pol.Orient.S is reliably positive (beta = 0.02, SE = 5.67e-03, std. beta = 0.01, p \textless{} .001).
+  The effect of Political Conservativism (S) is reliably positive (beta = 0.02, SE = 5.67e-03, std. beta = 0.01, p \textless{} .001).
 \item
-  The effect of Urban is reliably negative (beta = -0.02, SE = 0.01, std. beta = -0.01, p \textless{} .05).
+  The effect of living in an Urban area is reliably negative (beta = -0.02, SE = 0.01, std. beta = -0.01, p \textless{} .05).
 \end{itemize}
@@
 \item
-  The effect of Years X SpiritExcludesGod\_ is reliably positive (beta = 0.02, SE = 4.86e-03, std.
+  The effect of years \(\times\) belief in a SpiritExcludesGod\_ is reliably positive (beta = 0.02, SE = 4.86e-03, std.
 beta = 0.02, p \textless{} .01).
@@
 \item
-  The effect of Years X GodAndSpirit is reliably positive (beta = 0.02, SE = 4.41e-03, std.
+  The effect of years \(\times\) in GodAndSpirit is reliably positive (beta = 0.02, SE = 4.41e-03, std.
 beta = 0.03, p \textless{} .001).
@@
 \item
-  The effect of Years X GodExcludesSpirit is not reliable(beta = 1.96e-03, SE = 8.29e-03, std.
+  The effect of Years \(\times\) belief in GodExcludesSpirit is not reliable(beta = 1.96e-03, SE = 8.29e-03, std.
 beta = 2.36e-03, p = 0.813).
@@
 
-\hypertarget{life-satisfaction}{%
-\subsection{Life Satisfaction}\label{life-satisfaction}}
+\hypertarget{life-satisfaction-1}{%
+\subsection{Life Satisfaction}\label{life-satisfaction-1}}
 
@@
 \item
-  The effect of Years is not reliable (beta = -5.50e-04, SE = 2.69e-03, std. beta = -9.32e-04, p = 0.838). We do not infer growth over time in Life Satisfaction among the Skeptic population
+  The effect of years is not reliable (beta = -5.50e-04, SE = 2.69e-03, std. beta = -9.32e-04, p = 0.838). We do not infer growth over time in Life Satisfaction among the Skeptic population
 \item
-  The effect of BeliefsSpiritExcludesGod is reliably negative (beta = -0.08, SE = 0.02, std.
+  The effect of beliefs in SpiritExcludesGodis reliably negative (beta = -0.08, SE = 0.02, std.
 beta = 0.01, p \textless{} .001).
@@
 \item
-  The effect of BeliefsGodAndSpirit is not reliable (beta = -0.04, SE = 0.02, std. beta = 0.04, p = 0.105). We do not infer growth over time in Life Satisfaction among this (Skeptic) population
+  The effect of beliefs in GodAndSpirit is not reliable (beta = -0.04, SE = 0.02, std. beta = 0.04, p = 0.105). We do not infer growth over time in Life Satisfaction among this (Skeptic) population
 \item
-  The effect of BeliefsGodExcludesSpirit is reliably positive (beta = 0.07, SE = 0.04, std.
+  The effect of beliefs in GodExcludesSpirit is reliably positive (beta = 0.07, SE = 0.04, std.
 beta = 0.05, p = 0.062).We infer that overall the population that believes in a God but does not believe in a spirit is not reliably different from the Skeptics population in Life Satisfaction.
@@
 \item
-  The effect of Deprivation.S is is reliably negative (beta = -0.07, SE = 5.50e-03, std. beta = -0.06, p \textless{} .001). People who are more deprived have lower Life Satisfaction.
+  The effect of Deprivation (S) is is reliably negative (beta = -0.07, SE = 5.50e-03, std. beta = -0.06, p \textless{} .001). People who are more deprived have lower Life Satisfaction.
 \item
-  The effect of Edu.S is reliably positive (beta = 0.06, SE = 6.33e-03, std. beta = 0.05, p \textless{} .001). Educated people have greater Life Satisfaction
+  The effect of Education (S) is reliably positive (beta = 0.06, SE = 6.33e-03, std. beta = 0.05, p \textless{} .001). Educated people have greater Life Satisfaction
 \item
-  The effect of Employed is reliably positive (beta = 0.08, SE = 9.69e-03, std. beta = 0.07, p \textless{} .001). People who are employed tend to have greater Life Satisfaction
+  The effect of being Employed is reliably positive (beta = 0.08, SE = 9.69e-03, std. beta = 0.07, p \textless{} .001). People who are employed tend to have greater Life Satisfaction
 \item
-  The effect of EthnicCatsMaori is reliably negative (beta = -0.06, SE = 0.02, std.
+  The effect of identifying as Maori is reliably negative (beta = -0.06, SE = 0.02, std.
 beta = -0.05, p \textless{} .01).
@@
 \item
-  The effect of EthnicCatsPacific is reliably negative (beta = -0.08, SE = 0.04, std.
+  The effect of identifying as Pacific is reliably negative (beta = -0.08, SE = 0.04, std.
 beta = -0.07, p \textless{} .05).
@@
 \item
-  The effect of EthnicCatsAsian is not reliably different from people of European ethnicity in life satisfaction(beta = -0.04, SE = 0.03, std. beta = -0.03, p = 0.223).
+  The effect of identifying as Asian is not reliably different from people of European ethnicity in life satisfaction(beta = -0.04, SE = 0.03, std. beta = -0.03, p = 0.223).
 \item
-  The effect of Male is reliably negative (beta = -0.19, SE = 0.01, std. beta = -0.16, p \textless{} .001).
+  The effect of identifying as Male is reliably negative (beta = -0.19, SE = 0.01, std. beta = -0.16, p \textless{} .001).
 \item
-  The effect of HasPartner is reliably positive (beta = 0.42, SE = 0.01, std. beta = 0.35, p \textless{} .001).
+  The effect of having a Parnter is reliably positive (beta = 0.42, SE = 0.01, std. beta = 0.35, p \textless{} .001).
 \item
-  The effect of Pol.Orient.S is reliably positive (beta = 0.01, SE = 4.23e-03, std. beta = 0.01, p \textless{} .01).
+  The effect of Political Conservativism (S) is reliably positive (beta = 0.01, SE = 4.23e-03, std. beta = 0.01, p \textless{} .01).
 \item
-  The effect of Urban is reliably negative (beta = -0.02, SE = 8.85e-03, std. beta = -0.02, p \textless{} .05).
+  The effect of living in an Urban area is reliably negative (beta = -0.02, SE = 8.85e-03, std. beta = -0.02, p \textless{} .05).
 \end{itemize}
@@
 \item
-  The interaction of Years X Beliefs\_SpiritExcludesGod\_ is reliably positive (beta = 0.02, SE = 3.66e-03, std.
+  The interaction of years \(\times\) beliefs in SpiritExcludesGod is reliably positive (beta = 0.02, SE = 3.66e-03, std.
 beta = 0.03, p \textless{} .001).
@@
   \item
-    The interaction of Years X BeliefsGodAndSpirit is reliably positive (beta = 0.01, SE = 3.32e-03, std.
+    The interaction of years \(\times\) beliefsGodAndSpirit is reliably positive (beta = 0.01, SE = 3.32e-03, std.
 beta = 0.02, p \textless{} .001).
@@
   \item
-    The effect of Years X BeliefsGodExcludesSpirit is not reliable (beta = -1.59e-03, SE = 6.25e-03, std.
+    The effect of years \(\times\) beliefs in GodExcludesSpirit is not reliable (beta = -1.59e-03, SE = 6.25e-03, std.
 beta = -2.70e-03, p = 0.799).
@@
 The purpose of this study has been to clarify the relationship between belief in a spirit/life force and subjective well-being, addressing three challenges from previous research: (1) tautological measures of spirituality and subjective well-being; (2) insufficiently distinct measures of spirituality and religion; (3) cross-sectional samples.
-Here, we ask whether beliefs in a Spirit or Life Force predict growth in personal well-being and life-satisfaction over eight years in populations who differ in the this belief, and who also differ in their belief in a God.
-We find that
+Here, we ask whether beliefs in a Spirit or Life Force predict growth in personal well-being and life-satisfaction over 9 waves (8 years) in nationally diverse western populations who differ in the dimensions of spiritual and religious beliefs.
+We observe that:
 
-These findings demonstrate the importance of investigating spiritual beliefs as a domain of interest in its own right.
-In many Western countries, people are increasingly identifying as \enquote{spiritual but not religious} (Swinton, 2001; Zinnbauer \& Pargament, 2005).
-Other studies find that those who have a spiritual understanding of life without a religious framework are at an increased risk of having mental disorders (King et al., 2013).
-As such, it is important to understand the ramifications of belief in a spirit for psychological health.
-To assess the relationship between a belief in a spirit or life force and subjective well-being, we improve upon past methods in three ways.
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  At the outset of the study, in 2010, the expected subjective well-being as measured using the Personal Well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God (\(\hat{y}\) = 6.50, se =.03) and lower among those who believed in both a Spirit/Life-Force and a God (\(\hat{y}\) = 6.48, se =.03) in comparison both to the skeptical population who neither believed in a God nor a life force (\(\hat{y}\) = 6.64,se =.03) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force (\(\hat{y}\) 6.66, se =.05).
+(See Figure 2)
+\end{enumerate}
 
-Despite the opposition of our findings to those of many previous studies, there is room for agreement.
-We show that over time both \enquote{BXG} and \enquote{BIG} are associated with marginally better subjective well-being.
-As seen in Figure 2., both \enquote{BXG} and \enquote{BIG} predict marginally higher personal well-being and life satisfaction as time progresses.
-So, while those who adhere to these beliefs may start out below their skeptic counterparts in terms of their subjective well-being, they do eventually catch up.
-We also find that their life satisfaction is actually improving at a greater rate than skeptics.
-Thus, in the long term, both sustained \enquote{BXG} and \enquote{BIG} may actually be associated with better subjective well-being.
+Similarly,the expected subjective well-being as measured using the Life Satisfaction Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God (\(\hat{y}\) = 4.86, se =.02) and lower among those who believed in both a Spirit/Life-Force and a God (\(\hat{y}\) = 4.91, se =.02) in comparison both with the skeptical population who neither believed in a God nor a life force (\(\hat{y}\) = 4.94,se =.02) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force (\(\hat{y}\) 5.02, se =.02).
+(See Figure 3)
 
-Our results also help to clarify the relationship among religion, spirituality, and subjective well-being outcomes.
-BXG and BIG performed almost identically in terms of their ability to predict subjective well-being outcomes.
-This trend suggests that any kind of existential belief - regardless of whether a god is involved - might affect subjective well-being in much the same way.
-This result is good news as increasing numbers of individuals continue to report spiritual but not necessarily religious beliefs.
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{1}
+\tightlist
+\item
+  However, over time, belief in a Spirit or Life Force predicted a trajectory of growth in Personal Well-being.
+In 2018, the expected subjective well-being as measured using the Personal Well-being Inventory among those who believed in Spirit or Life Force but not a God was expected to grow to (\(\hat{y}\) = 6.67, se =.03, \(\Delta \hat{y} = 0.17\)).
+Among those who believed in both a Spirit/Life-Force and a God Personal Well-Being was expected to grow to (\(\hat{y}\) = 6.69, se = 0.02,\(\Delta \hat{y} = .21\)).
+In practical terms, such growth from spiritual beliefs among to the expected gain in Personal Well-being about twice as large as the expected benefit of being employed and about about a third as large as the expected benefit of having a romantic partner.
+In comparison, Personal Well-Being among skeptical population who neither believed in a God nor a Life Force remained stable \(\hat{y} = 6.66, se = 0.03, \Delta \hat{y} = .02\)).
+We also observe stability in Personal Well-being among the population that believed in God but was skeptical about a Spirit or Life Force \(\hat{y} = 6.70, se = 0.04, \Delta \hat{y} = .04\)).
+(See Figure 2).
+\end{enumerate}
 
+Similarly, over time, belief in a Spirit or Life Force predicted a trajectory of growth in Life Satisfaction.
+In 2018, the expected subjective well-being as measured using the Life Satisfaction Inventory among those who believed in Spirit or Life Force but not a God was expected to grow to (\(\hat{y} = 5.01, se = 0.02, \Delta \hat{y} = 0.07\)).
+In practical terms, such growth from spiritual beliefs is approximately equivalent to the expected gain in Life Satisfaction from being employed and a sixth as large as the expected benefit to Life Satisfaction from having a romantic partner.
+Among those who believed in both a Spirit/Life-Force and a God Personal Well-Being was expected to grow to (\(\hat{y} = 5.03, se = 0.02, \Delta \hat{y} = 0.12\)).
+In practical terms, such growth from spiritual beliefs is approximately equivalent to the 1.5 times the expected gain in Life Satisfaction from being employed and almost a third the expected benefit to Life Satisfaction from having a romantic partner.
+In comparison, Life Satisfaction among skeptical population who neither believed in a God nor a Life Force remained stable \(\hat{y} = 4.94, se =0.02, \Delta \hat{y} = 0\)).
+We also observe stability in Personal Well-being among the population that believed in God but was skeptical about a Spirit or Life Force \(\hat{y} = 5.00, se = 0.03, \Delta \hat{y} = -.02\)) (See Figure 3).
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{2}
+\tightlist
+\item
+  Such observed growth in subjective well-being among the population that believes in a Spirit or Life Force holds even after adjusting for demographic factors that are known to affect well-being.
+Additionally we infer growth from 9 annual waves of data collection (8 years) in a nationally diverse sample.
+On this basis of this window of time and sample, we infer that such growth in Life Satisifacton and in Personal Well-being reflects changes in the populations that believed in Spirit or Life Force over that period.
+\end{enumerate}
+
+Questions:
+1. Why initially lower well-being in NZ? \(\dots\)
++ historical contingencies?
++ possibility of non-linear futures, are we observing a regression to the mean?
++ individual factors: earthquakes, however..why the slope for spirit beliefs (and not God)?
++ other things besides beliefs? behaviors, social interactions, perspectives\ldots{}
 Taken as a whole, our results cannot refute the entire narrative of the positive relationship between spirituality and emotional well-being.
 First, our analysis focuses only on belief.
-While belief is an important part of spirituality it is not by any means spirituality itself.
+Although belief is an important part of spirituality it is not by any means spirituality itself.
 Individuals may identify as spiritual without any accompanying belief in a spirit or life force.
@@
 Overall, though, our findings add to the existing literature, helping to untangle the multifaceted relationship among religion, spirituality, and subjective well-being.
+\(\dots\) fruitful horizons for future research.
+
+Importance:
+1. Methods: Disentangling \(\dots\) concepts/ longitudinal
+2.
+Growing population: practical importance.
+In many Western countries, people are increasingly identifying as \enquote{spiritual but not religious} (Swinton, 2001; Zinnbauer \& Pargament, 2005).
+\ldots{}
++ risks: Risks? \ldots{} Other studies find that those who have a spiritual understanding of life without a religious framework are at an increased risk of having mental disorders (King et al., 2013).
++ However, the observed growth suggests that any kind of existential belief - regardless of whether a god is involved - might affect subjective well-being in much the same way.
+This result is good news as increasing numbers of individuals continue to report spiritual but not necessarily religious beliefs.
 

### 2020-11-18 Manuscript.Rmd
revisions

@@
 
-We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict Life Satisfaction (LIFESAT) with Years (0-9), Beliefs (four categories), Age.10yrs.C, Deprivation.S, Edu.S, Employed, EthnicCats, Male, Partner, Pol.Orient.S and Urban, with  Id as random effects.
-Standardized parameters were obtained by fitting the model on a standardized version of the dataset.The model's total explanatory power is substantial (conditional R2 = 0.69).
-The model's intercept, corresponding to Years = 0, Beliefs = Skeptic, Age = (sample mean),, Deprivation.S = (sample mean), Edu.S =(sample mean), Employed = Not Employed, EthnicCats = Euro, Male = Not Mail, Partner = No Partner, Pol.Orient.S = (sample mean), Urban = Not Urban and Id = 2, is at 4.94 (SE = 0.02, 95% CI [4.89, 4.99], p < .001).
+We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict Life Satisfaction (LIFESAT) with Years (0-9), Beliefs (four categories), Age.10yrs.C, Deprivation (S), Education (S), Employed, EthnicCats, Male, Partner, Pol.Orient.S and Urban, with  Id as random effects.
+Standardized parameters were obtained by fitting the model on a standardized version of the dataset.
+The model's total explanatory power is substantial (conditional R2 = 0.69).
+The model's intercept, corresponding to Years = 0, Beliefs = Skeptic, Age = (sample mean),, Deprivation.S = (sample mean), Education =(sample mean), Employed = Not Employed, EthnicCats = Euro, Male = Not Mail, Partner = No Partner, Pol.Orient.S = (sample mean), Urban = Not Urban and Id = 2, is at 4.94 (SE = 0.02, 95% CI [4.89, 4.99], p < .001).
 Within this model:
@@
   - The effect of years is not reliable (beta = -5.50e-04, SE = 2.69e-03, std. beta = -9.32e-04, p = 0.838).  We do not infer growth over time in Life Satisfaction among the Skeptic population
-  - The effect of beliefs in SpiritExcludesGodis reliably negative (beta = -0.08, SE = 0.02, std.
+  - The effect of beliefs in SpiritExcludesGod is reliably negative (beta = -0.08, SE = 0.02, std.
 beta = 0.01, p < .001).
@@
 ## Imputed data analysis
-To adjust for potential biases from missing responses, we conducted a supplementary analysis in which we imputed missing values over the the longitudinal sample.
+To adjust for potential bias from missing responses, we conducted a supplementary analysis in which we imputed missing values over the the longitudinal sample.
 This analysis is describe in Supplement 1.
@@
 The inference from this analysis remains the same as from the analysis in which we delete missing values row-wise, by panel wave.
-In all analyses those who believe in a spirit or life-force are expected to experience greater growth in well-being over time.
+In all analyses, those who believe in a spirit or life-force experience greater growth in both Personal Well-being and Life Satisfaction over time.
 
@@
 At the outset of the study, in 2010, the expected subjective well-being as measured using the Personal Well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God ($\hat{y}$ =  6.50, se =.03) and lower among those who believed in both a Spirit/Life-Force and a God ($\hat{y}$ =  6.48, se =.03) in comparison both to the skeptical population who neither believed in a God nor a life force ($\hat{y}$ = 6.64,se =.03) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 6.66, se =.05).
-(See Figure 2)
-
-Similarly,the expected subjective well-being as measured using the Life Satisfaction Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God ($\hat{y}$ = 4.86, se =.02) and lower among those who believed in both a Spirit/Life-Force and a God  ($\hat{y}$ = 4.91, se =.02) in comparison both with the skeptical population who neither believed in a God nor a life force ($\hat{y}$ = 4.94,se =.02) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 5.02, se =.02).
-(See Figure 3)
+(See Figure 2)  Similarly,the expected subjective well-being as measured using the Life Satisfaction Inventory is lower among those who believed in Spirit or Life Force but not a God ($\hat{y}$ = 4.86, se =.02) and lower among those who believed in both a Spirit/Life-Force and a God  ($\hat{y}$ = 4.91, se =.02) in comparison both with the skeptical population who neither believed in a God nor a life force ($\hat{y}$ = 4.94,se =.02) and the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 5.02, se =.02) (see Figure 3)
 
 2.
-However, over time, belief in a Spirit or Life Force predicted a trajectory of growth in Personal Well-being.
-In 2018, the expected subjective well-being as measured using the Personal Well-being Inventory among those who believed in Spirit or Life Force but not a God was expected to grow to ($\hat{y}$ = 6.67, se =.03, $\Delta \hat{y} = 0.17$).
-Among those who believed in both a Spirit/Life-Force and a God Personal Well-Being was expected to grow to ($\hat{y}$ = 6.69, se = 0.02,$\Delta \hat{y} = .21$).
-In practical terms, such growth from spiritual beliefs among to the expected gain in Personal Well-being about twice as large as the expected benefit of being employed and about about a third as large as the expected benefit of having a romantic partner.
-In comparison, Personal Well-Being among skeptical population who neither believed in a God nor a Life Force remained stable $\hat{y} = 6.66, se = 0.03, \Delta \hat{y} = .02$).
-We also observe stability in Personal Well-being among the population that believed in God but was skeptical about a Spirit or Life Force $\hat{y} =  6.70, se = 0.04, \Delta \hat{y} = .04$).
-(See Figure 2).
-
-Similarly, over time, belief in a Spirit or Life Force predicted a trajectory of growth in Life Satisfaction.
-In 2018, the expected subjective well-being as measured using the Life Satisfaction Inventory among those who believed in Spirit or Life Force but not a God was expected to grow to ($\hat{y} = 5.01, se = 0.02, \Delta \hat{y} = 0.07$).
-In practical terms, such growth from spiritual beliefs is approximately equivalent to the expected gain in Life Satisfaction from being employed and a sixth as large as the expected benefit to Life Satisfaction from having a romantic partner.
-Among those who believed in both a Spirit/Life-Force and a God Personal Well-Being was expected to grow to ($\hat{y} = 5.03, se = 0.02, \Delta \hat{y} = 0.12$).
-In practical terms, such growth from spiritual beliefs is approximately equivalent to the 1.5 times the expected gain in Life Satisfaction from being employed and almost a third the expected benefit to Life Satisfaction from having a romantic partner.
-In comparison, Life Satisfaction among skeptical population who neither believed in a God nor a Life Force remained stable $\hat{y} =  4.94, se =0.02, \Delta \hat{y} = 0$).
-We also observe stability in Personal Well-being among the population that believed in God but was skeptical about a Spirit or Life Force $\hat{y} =  5.00, se = 0.03, \Delta \hat{y} = -.02$) (See Figure 3).
+However, over time, belief in a Spirit or Life Force predicts a trajectory of growth in personal well-being.
+In 2018, personal well-being among who believe in Spirit or Life Force but not a God had grown to ($\hat{y}$ = 6.67, se =.03, $\Delta \hat{y} = 0.17$).
+Among those who believed in both a Spirit/Life-Force and a God Personal Well-Being had grown to ($\hat{y}$ = 6.69, se = 0.02,$\Delta \hat{y} = .21$).
+To place this effect in practical perspective, such growth in personal well-being is about twice as large as the expected benefit of being employed and about a third as large as the expected benefit from having a romantic partner.
+By contrast, personal well-being among the skeptical population who neither believed in a God nor a Life Force remained stable across the eight year interval $\hat{y} = 6.66, se = 0.03, \Delta \hat{y} = .02$).
+By the same token, personal well-being among those who believe in a God but are skeptical about a Spirit/Life Force was constant $\hat{y} =  6.70, se = 0.04, \Delta \hat{y} = .04$).
+Similarly, over time, belief in a Spirit or Life Force predicts a trajectory of growth in Life Satisfaction.
+In 2018, life satisfaction among those who believe in Spirit or Life Force but not a God had grown to ($\hat{y} = 5.01, se = 0.02, \Delta \hat{y} = 0.07$).
+In practical terms, such growth is equivalent to the expected gain in Life Satisfaction from being employed and a sixth as large as the expected benefit to Life Satisfaction from having a romantic partner.
+Among those who believed in both a Spirit/Life-Force and a God, life satisfaction had grown to ($\hat{y} = 5.03, se = 0.02, \Delta \hat{y} = 0.12$).
+In practical terms, such growth is equivalent to 1.5 times the expected gain to life satisfaction from being employed and almost a third benefit to life satisfaction from having a romantic partner.
+By contrast, life satisfaction among the skeptical population who neither believed in a God nor a Life Force remained stable over the eight year interval $\hat{y} =  4.94, se =0.02, \Delta \hat{y} = 0$).
+Relatedly, life satisfaction in the population that believe in God but are skeptical about a Spirit/Life Force was constant $\hat{y} =  5.00, se = 0.03, \Delta \hat{y} = -.02$) (See Figure 3).
 
 3.
-Such observed growth in subjective well-being among the population that believes in a Spirit or Life Force holds even after adjusting for demographic factors that are known to affect well-being.
-Additionally we infer growth from  9 annual waves of data collection (8 years) in a nationally diverse sample.
-On this basis of this window of time and sample, we infer that such growth in Life Satisifacton and in Personal Well-being reflects changes in the populations that believed in Spirit or Life Force over that period.
+Inferred growth from spiritual beliefs is evident from 9 annual waves of data collection (8 years), and across a nationally diverse sample.
+On this basis of this longitudinal information accumulated across nationally, we are confident that growth in personal well-being and in life-satisfaction among those who believe in a Spirit and Life force are evident.
+Such effects co-occur in people who lack a belief in a God as well as among people who believe in a God as well as a Spirit or Life force.
 
-Questions:
-1. Why initially lower well-being in NZ? $\dots$
-  + historical contingencies? 
-  + possibility of non-linear futures, are we observing a regression to the mean?
-  + individual factors: earthquakes, however..why the slope for spirit beliefs (and not God)?
-  + other things besides beliefs? behaviors, social interactions, perspectives... 
-  Taken as a whole, our results cannot refute the entire narrative of the positive relationship between spirituality and emotional well-being.
-First, our analysis focuses only on belief.
-Although belief is an important part of spirituality it is not by any means spirituality itself.
-Individuals may identify as spiritual without any accompanying belief in a spirit or life force.
-Moreover, practices and rituals associated with different forms of spirituality that were not tested in this study may have large effects.
-Second, despite flawed methods, past studies may still be accurate.
-New Zealand is a largely secular country, so belief in a spirit or life force may have different impacts on subjective well-being outcomes in this context than in other, more religious contexts.
-Overall, though, our findings add to the existing literature, helping to untangle the multifaceted relationship among religion, spirituality, and subjective well-being.
-$\dots$ fruitful horizons for future research.
+## Key Questions
+Two fundamental question arise from the patterns we observe in this study.
+The first question is why those who believed in Spirit or Life Force were initially lower in their personal well-being and life-satisfaction in the initial years of this study in comparison to religious and non-religious skeptics?
+During the early years of New Zealand Attitudes and Values data collection New Zealand was emerging from the twin impacts of a global financial recession and a series of devastating earthquakes that destroyed 30% of Christchurch, New Zealand's second largest city [CITE].
+Orthogonal research observes that economic and ecological distress are associated with increases in traditional religious beliefs [CITE, CITE].
+Previous research finds an increase in traditional religious affiliation in the wake of the Christchurch earthquakes, with no immediate effect in subjective well-being.
+Experimental research has found among religious people death anxiety provokes acceptance of non-traditional religious beliefs [cite].
+Because spiritual belief items were only introduced during the year of the earthquakes, we cannot here resolve whether the shifting relationship between spirit beliefs and well-being were the result of demand for spiritual beliefs among those secular and religious people who were adversely affected by the financial crisis or natural disaster.
+On the other hand, such exogenous influences might have adversely affected those with Spirit/Life force beliefs in the early phases of this study, and the patterns observed here are better understood as a regression to a population mean.
+Though beyond the scope of this study, how subjective well-being changes in response to changes in spiritual beliefs, as well as special sensitivities to distress among people with spiritual beliefs are important questions for future investigations.
+Despite the longitudinal quality of this study, a causal explanation remains elusive.
 
-Importance:
-1. Methods: Disentangling $\dots$ concepts/ longitudinal 
-2. Growing population: practical importance. In many Western countries, people are increasingly identifying as "spiritual but not religious" [@Zinnbauer2005-vz; @Swinton2001-vr]. ... 
- + risks: Risks? ... Other studies find that those who have a spiritual understanding of life without a religious framework are at an increased risk of having mental disorders [@King2013-cg].
- + However, the observed growth suggests that any kind of existential belief - regardless of whether a god is involved - might affect subjective well-being in much the same way.
-This result is good news as increasing numbers of individuals continue to report spiritual but not necessarily religious beliefs.
+Second, our task here has been to clarify whether beliefs predict growth in well-being.
+However, as we have indicated at the outset, the term "spirituality" applies to a family of concepts.
+We have not attempted to model the diverse philosophical, social, and behavioral attributes within people who adhere to spirit beliefs.
+Moreover, people might identify as spiritual without having any accompanying belief in a Spirit or Life force.
+The task of clarifying how distinct features across a diverse family of "spirituality" concepts affect perceptions of individual well-being remain intriguing horizons.
 
-## Limitations
-
-This article makes useful contributions to the continued debate regarding the relationship between religion and spirituality and mental health outcomes, but it is important to acknowledge several limitations.
-First, we relied upon self-reported data from the NZAVS.
-Although this is an accurate method for measuring belief in a spirit or life force, it may be a less accurate manner of measuring subjective well-being.
-It is possible that self-administration of our emotional well-being (personal well-being and life satisfaction) scales could result in mismeasurement.
-Second, we may not capture the full range spiritualities, which likely extend beyond belief and might even be inherently emotional.
-Thus, while our method avoided tautology, it might have also excluded participants who consider themselves to be spiritual but do not subscribe to a belief in a spirit or life force.
+## Importance
+Despite its limitations, our study is important both for its methods and its findings.
+We demonstrate that spirituality can be operationalised to avoid the pitfalls of tautology evident in previous research by focusing on states of beliefs, that concepts of such belief states can be refined to enable distinctions necessary to investigate spirituality in mixed secular and religious populations, and that the question of whether beliefs states predict growth in subjective well-being can be investigated in a longitudional setting.
+Our finding that, indeed, spiritual beliefs predict growth in perceived well-being and life satisfaction may hold practical importance.
+Previous research indicates that religious beliefs are associated with having a rewarding life.
+Western populations are increasingly identifying as "spiritual but not religious" [@Zinnbauer2005-vz; @Swinton2001-vr], raising the specter that the psychological from traditional might be receding.
+Here, we observe that even among people who lack a traditional religious belief in God, believing in a Spirit or Life force predicts incremental improvements in well-being that sum to practically large effects.
+Though the contextual and psychological mechanisms that underpin such growth will require many decades of research, this preliminary finding is good news for those increasing numbers of individuals continue to report spiritual beliefs but who do not affirm traditional religious beliefs.
 
 ### Authors contributions
-CS leads and manages the NZAVS, and collected the data for this study. BH and JB conceived
-of the study.
-BH wrote the first draft of the manuscript and performed the analysis with input from JB.
-BH conducted background research and scientific literature review.
-All authors contributed substantially to the intellectual content of the study and the final manuscript.
+BH conceived of the study, wrote the first draft of the manuscript, and performed the initial data analysis with input from JB.
+EW developed the study with BH and JB and revised the manuscript.
+DD revised the manuscript.
+CS was responsible for NZAVS data collection and revised the manuscript.
+JB developed the study with BH and EW, performed the data re-analysis, and revised the manuscript.
 
 ### Funding statement
-The New Zealand Attitudes and Values Study is supported by a grant from the Templeton
-Religion Trust (TRT0196). The funders had no role in the preparation of the article or the
-decision to publish.
+The New Zealand Attitudes and Values Study is supported by a grant from the Templeton Religion Trust (TRT0196). The funders had no role in the preparation of the article or the decision to publish.
 
 # Supplemental Material
-S1. Statistical Analysis in R
 Statistical analysis was performed using R version 3.6.1 (2019-07-05).

### 2020-11-18 Analysis.Rmd
jb finished first pass at re-write. To do: - ben to polish - jb and ben to redo revision letter. - ben to assimilate futher comments to overleaf draft - jb to remind ben to push to github from overleaf

@@
 wordcount         : "X"
-bibliography      : ["r-references.bib"]
+bibliography      : ["b-references.bib"]
 floatsintext      : no
@@
 
+```{r, sessioninfo, include=FALSE}
+
+citation(package = "base", lib.loc = "/Users/josephbulbulia/Dropbox/BIBS")
+toLatex(sessionInfo(), locale = FALSE)
+sapply(names(sessionInfo()$otherPkgs), function(x) print(citation(x), style = "Bibtex"))
+
+out <- sapply(names(sessionInfo()$otherPkgs),
+              function(x) print(citation(x), style = "Bibtex"))
+print(out)
+```
+
 # Methods
  <!-- 21-word solution (Simmons, Nelson & Simonsohn, 2012; retrieved from http://ssrn.com/abstract=2160588) -->
-We used `r cite_r("r-references.bib")` for all our analyses.
+We used `r cite_r("b-references.bib")` for all our analyses.
 

### 2020-11-18 Manuscript.Rmd
jb finished first pass at re-write. To do: - ben to polish - jb and ben to redo revision letter. - ben to assimilate futher comments to overleaf draft - jb to remind ben to push to github from overleaf

@@
     corresponding : yes    # Define only one corresponding author
+    address       : "Ben's Postal address"
     email         : "benhighland6420@gmail.com"
@@
 keywords          : "Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being"
-wordcount         : "5015"
+wordcount         : "5269 words in text body, 1078 words in reference section"
 bibliography      : ["b-references.bib"]
@@
 
-Because missing values can result in biased estimates [@Blackwell2017-oq], we multiply imputed 10 datasets using the AmeliaII package in R[@Honaker2011-yu].
-We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences.
+### System and packages
+The analysis was performed using R version 4.0.2 (2020-06-22). The Platform was x86_64-apple-darwin17.0 (64-bit) Running under: macOS Catalina 10.15.6
+We are greatful to the contributors and mantainers of the following packages:
+R [Version 4.0.2; @R-base] and the R-packages *Amelia* [Version 1.7.6; @R-Amelia], *citr* [Version 0.3.2; @R-citr], *dplyr* [Version 1.0.2; @R-dplyr], *equatiomatic* [Version 0.1.0.9000; @R-equatiomatic], *forcats* [Version 0.5.0; @R-forcats], *ggeffects* [Version 0.15.0; @R-ggeffects], *gghighlight* [Version 0.3.0; @R-gghighlight], *ggplot2* [Version 3.3.2; @R-ggplot2], *ggsci* [Version 2.9; @R-ggsci], *lme4* [Version 1.1.23; @R-lme4], *Matrix* [Version 1.2.18; @R-Matrix], *papaja* [Version 0.1.0.9997; @R-papaja], *parameters* [Version 0.8.6.1; @R-parameters], *patchwork* [Version 1.0.1; @R-patchwork], *prettycode* [Version 1.1.0; @R-prettycode], *purrr* [Version 0.3.4; @R-purrr], *Rcpp* [Version 1.0.5.1; @R-Rcpp_a; @R-Rcpp_b], *readr* [Version 1.3.1; @R-readr], *report* [Version 0.1.0; @R-report], *see* [Version 0.5.2.1; @R-see], *sjPlot* [Version 2.8.4; @R-sjPlot], *stringr* [Version 1.4.0; @R-stringr], *styler* [Version 1.3.2; @R-styler], *table1* [Version 1.2; @R-table1], *texreg* [Version 1.37.5; @R-texreg], *tibble* [Version 3.0.4; @R-tibble], *tidyr* [Version 1.1.2; @R-tidyr], and *tidyverse* [Version 1.3.0; @R-tidyverse]"
+
+### Missing data
+Because missing values can result in biased estimates [@Blackwell2017-oq], we multiply imputed 10 datasets using the AmeliaII package in R[@Honaker2011-yu].We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences.
+Details of this analysis can be found at: https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing.
 
@@
 ```{r fig.cap = "Combination Coefficient Plots for Personal Well-being and Life Satisfaction Spirit Belief Longitudinal Models"}
-knitr::include_graphics("Figs/USEcoefficientplots-1.png")
+knitr::include_graphics("Figs/USE_2_coefficientplots-1.png")
 ```
@@
 
-# Supplemental Material
-Statistical analysis was performed using R version 3.6.1 (2019-07-05).
-Platform: x86_64-w64-mingw32 (64-bit), running under: Windows 10 (Action of the Toes).
-Generalized Linear Mixed Models were generated using the glmmTMB (Brooks et al.
-2017; Knowles and Frederick 2019) package in R.
-In addition, we used the following R packages: dplyr (Hadley Wickham et al.
-2019), tidyr (H.
-Wickham and Henry 2019), sjPlot (D.
-Lüdecke 2020), ggeffects (Daniel Lüdecke 2018), table1  and their dependencies.
-Tables and coefficient plots were generated using the glmmTMB, ggeffects, and sjPlot packages in R.
-Models for both subjective well-being variables used family = gaussian, assuming a normal distribution.
-We ran each individual’s unique ID as a random-effect.
-The tab_model and plot_model commands in sjPlot allowed for the creation of tables and coefficient plots.
-Predicted probability plots were generated using the ggpredict command in the ggeffects package.
-The table1 package in R was used for the creation of descriptive statistics tables.
-
 \newpage

### 2020-11-18 Manuscript.tex
jb finished first pass at re-write. To do: - ben to polish - jb and ben to redo revision letter. - ben to assimilate futher comments to overleaf draft - jb to remind ben to push to github from overleaf

@@
 \shorttitle{Longitudinal Wellbeing from Spiritual Beliefs}
-\keywords{Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being\newline\indent Word count: 5015}
+\keywords{Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being\newline\indent Word count: 5269 words in text body, 1078 words in reference section}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 1 Wake Forest Medical School 2 Department of Psychology Virginia Commonwealth University 3 Department of Psychology Georgia State University 4 School of Psychology University of Auckland 5 School of Psychology Victoria University
+
+Correspondence concerning this article should be addressed to Ben Highland, Ben's Postal address. E-mail: \href{mailto:benhighland6420@gmail.com}{\nolinkurl{benhighland6420@gmail.com}}
 
@@
 
-Because missing values can result in biased estimates (Blackwell, Honaker, \& King, 2017), we multiply imputed 10 datasets using the AmeliaII package in R(Honaker, King, Blackwell, \& Others, 2011).
-We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences.
+\hypertarget{system-and-packages}{%
+\subsubsection{System and packages}\label{system-and-packages}}
+
+The analysis was performed using R version 4.0.2 (2020-06-22). The Platform was x86\_64-apple-darwin17.0 (64-bit) Running under: macOS Catalina 10.15.6
+We are greatful to the contributors and mantainers of the following packages:
+R (Version 4.0.2; R Core Team, 2020) and the R-packages \emph{Amelia} (Version 1.7.6; Honaker et al., 2011a), \emph{citr} (Version 0.3.2; Aust, 2019), \emph{dplyr} (Version 1.0.2; Wickham et al., 2020), \emph{equatiomatic} (Version 0.1.0.9000; Anderson \& Heiss, 2020), \emph{forcats} (Version 0.5.0; Wickham, 2020a), \emph{ggeffects} (Version 0.15.0; Lüdecke, 2018), \emph{gghighlight} (Version 0.3.0; Yutani, 2020), \emph{ggplot2} (Version 3.3.2; Wickham, 2016), \emph{ggsci} (Version 2.9; Xiao, 2018), \emph{lme4} (Version 1.1.23; Bates, Mächler, Bolker, \& Walker, 2015), \emph{Matrix} (Version 1.2.18; Bates \& Maechler, 2019), \emph{papaja} (Version 0.1.0.9997; Aust \& Barth, 2020), \emph{parameters} (Version 0.8.6.1; D.
+Lüdecke et al., 2020a), \emph{patchwork} (Version 1.0.1; Pedersen, 2020), \emph{prettycode} (Version 1.1.0; Csárdi, 2019), \emph{purrr} (Version 0.3.4; Henry \& Wickham, 2020), \emph{Rcpp} (Version 1.0.5.1; Eddelbuettel \& François, 2011; Eddelbuettel \& Balamuta, 2017), \emph{readr} (Version 1.3.1; Wickham, Hester, \& Francois, 2018), \emph{report} (Version 0.1.0; Makowski, Dominique, Lüdecke, \& Daniel, 2019), \emph{see} (Version 0.5.2.1; D.
+Lüdecke et al., 2020b), \emph{sjPlot} (Version 2.8.4; D.
+Lüdecke, 2020), \emph{stringr} (Version 1.4.0; Wickham, 2019), \emph{styler} (Version 1.3.2; Müller \& Walthert, 2020), \emph{table1} (Version 1.2; Rich, 2020), \emph{texreg} (Version 1.37.5; Leifeld, 2013), \emph{tibble} (Version 3.0.4; Müller \& Wickham, 2020), \emph{tidyr} (Version 1.1.2; Wickham, 2020b), and \emph{tidyverse} (Version 1.3.0; Wickham, Averick, et al., 2019)"
+
+\hypertarget{missing-data}{%
+\subsubsection{Missing data}\label{missing-data}}
+
+Because missing values can result in biased estimates (Blackwell, Honaker, \& King, 2017), we multiply imputed 10 datasets using the AmeliaII package in R(Honaker et al., 2011b).We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences.
+Details of this analysis can be found at: \url{https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing}.
 
@@
 \begin{figure}
-\includegraphics[width=6.4in]{Figs/USEcoefficientplots-1} \caption{Combination Coefficient Plots for Personal Well-being and Life Satisfaction Spirit Belief Longitudinal Models}\label{fig:unnamed-chunk-1}
+\includegraphics[width=6.4in]{Figs/USE_2_coefficientplots-1} \caption{Combination Coefficient Plots for Personal Well-being and Life Satisfaction Spirit Belief Longitudinal Models}\label{fig:unnamed-chunk-1}
 \end{figure}
@@
 
-We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict Life Satisfaction (LIFESAT) with Years (0-9), Beliefs (four categories), Age.10yrs.C, Deprivation.S, Edu.S, Employed, EthnicCats, Male, Partner, Pol.Orient.S and Urban, with Id as random effects.
-Standardized parameters were obtained by fitting the model on a standardized version of the dataset.The model's total explanatory power is substantial (conditional R2 = 0.69).
-The model's intercept, corresponding to Years = 0, Beliefs = Skeptic, Age = (sample mean),, Deprivation.S = (sample mean), Edu.S =(sample mean), Employed = Not Employed, EthnicCats = Euro, Male = Not Mail, Partner = No Partner, Pol.Orient.S = (sample mean), Urban = Not Urban and Id = 2, is at 4.94 (SE = 0.02, 95\% CI {[}4.89, 4.99{]}, p \textless{} .001).
+We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict Life Satisfaction (LIFESAT) with Years (0-9), Beliefs (four categories), Age.10yrs.C, Deprivation (S), Education (S), Employed, EthnicCats, Male, Partner, Pol.Orient.S and Urban, with Id as random effects.
+Standardized parameters were obtained by fitting the model on a standardized version of the dataset.
+The model's total explanatory power is substantial (conditional R2 = 0.69).
+The model's intercept, corresponding to Years = 0, Beliefs = Skeptic, Age = (sample mean),, Deprivation.S = (sample mean), Education =(sample mean), Employed = Not Employed, EthnicCats = Euro, Male = Not Mail, Partner = No Partner, Pol.Orient.S = (sample mean), Urban = Not Urban and Id = 2, is at 4.94 (SE = 0.02, 95\% CI {[}4.89, 4.99{]}, p \textless{} .001).
 Within this model:
@@
 \item
-  The effect of beliefs in SpiritExcludesGodis reliably negative (beta = -0.08, SE = 0.02, std.
+  The effect of beliefs in SpiritExcludesGod is reliably negative (beta = -0.08, SE = 0.02, std.
 beta = 0.01, p \textless{} .001).
@@
 
-To adjust for potential biases from missing responses, we conducted a supplementary analysis in which we imputed missing values over the the longitudinal sample.
+To adjust for potential bias from missing responses, we conducted a supplementary analysis in which we imputed missing values over the the longitudinal sample.
 This analysis is describe in Supplement 1.
@@
 \def\labelenumi{\arabic{enumi}.}
-\tightlist
 \item
   At the outset of the study, in 2010, the expected subjective well-being as measured using the Personal Well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God (\(\hat{y}\) = 6.50, se =.03) and lower among those who believed in both a Spirit/Life-Force and a God (\(\hat{y}\) = 6.48, se =.03) in comparison both to the skeptical population who neither believed in a God nor a life force (\(\hat{y}\) = 6.64,se =.03) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force (\(\hat{y}\) 6.66, se =.05).
-(See Figure 2)
+(See Figure 2) Similarly,the expected subjective well-being as measured using the Life Satisfaction Inventory is lower among those who believed in Spirit or Life Force but not a God (\(\hat{y}\) = 4.86, se =.02) and lower among those who believed in both a Spirit/Life-Force and a God (\(\hat{y}\) = 4.91, se =.02) in comparison both with the skeptical population who neither believed in a God nor a life force (\(\hat{y}\) = 4.94,se =.02) and the population that believed in a God but was skeptical about belief in a Spirit or Life Force (\(\hat{y}\) 5.02, se =.02) (see Figure 3)
+\item
+  However, over time, belief in a Spirit or Life Force predicts a trajectory of growth in personal well-being.
+In 2018, personal well-being among who believe in Spirit or Life Force but not a God had grown to (\(\hat{y}\) = 6.67, se =.03, \(\Delta \hat{y} = 0.17\)).
+Among those who believed in both a Spirit/Life-Force and a God Personal Well-Being had grown to (\(\hat{y}\) = 6.69, se = 0.02,\(\Delta \hat{y} = .21\)).
+To place this effect in practical perspective, such growth in personal well-being is about twice as large as the expected benefit of being employed and about a third as large as the expected benefit from having a romantic partner.
+By contrast, personal well-being among the skeptical population who neither believed in a God nor a Life Force remained stable across the eight year interval \(\hat{y} = 6.66, se = 0.03, \Delta \hat{y} = .02\)).
+By the same token, personal well-being among those who believe in a God but are skeptical about a Spirit/Life Force was constant \(\hat{y} = 6.70, se = 0.04, \Delta \hat{y} = .04\)).
+Similarly, over time, belief in a Spirit or Life Force predicts a trajectory of growth in Life Satisfaction.
+In 2018, life satisfaction among those who believe in Spirit or Life Force but not a God had grown to (\(\hat{y} = 5.01, se = 0.02, \Delta \hat{y} = 0.07\)).
+In practical terms, such growth is equivalent to the expected gain in Life Satisfaction from being employed and a sixth as large as the expected benefit to Life Satisfaction from having a romantic partner.
+Among those who believed in both a Spirit/Life-Force and a God, life satisfaction had grown to (\(\hat{y} = 5.03, se = 0.02, \Delta \hat{y} = 0.12\)).
+In practical terms, such growth is equivalent to 1.5 times the expected gain to life satisfaction from being employed and almost a third benefit to life satisfaction from having a romantic partner.
+By contrast, life satisfaction among the skeptical population who neither believed in a God nor a Life Force remained stable over the eight year interval \(\hat{y} = 4.94, se =0.02, \Delta \hat{y} = 0\)).
+Relatedly, life satisfaction in the population that believe in God but are skeptical about a Spirit/Life Force was constant \(\hat{y} = 5.00, se = 0.03, \Delta \hat{y} = -.02\)) (See Figure 3).
+\item
+  Inferred growth from spiritual beliefs is evident from 9 annual waves of data collection (8 years), and across a nationally diverse sample.
+On this basis of this longitudinal information accumulated across nationally, we are confident that growth in personal well-being and in life-satisfaction among those who believe in a Spirit and Life force are evident.
+Such effects co-occur in people who lack a belief in a God as well as among people who believe in a God as well as a Spirit or Life force.
 \end{enumerate}
 
-Similarly,the expected subjective well-being as measured using the Life Satisfaction Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God (\(\hat{y}\) = 4.86, se =.02) and lower among those who believed in both a Spirit/Life-Force and a God (\(\hat{y}\) = 4.91, se =.02) in comparison both with the skeptical population who neither believed in a God nor a life force (\(\hat{y}\) = 4.94,se =.02) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force (\(\hat{y}\) 5.02, se =.02).
-(See Figure 3)
+\hypertarget{key-questions}{%
+\subsection{Key Questions}\label{key-questions}}
 
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{1}
-\tightlist
-\item
-  However, over time, belief in a Spirit or Life Force predicted a trajectory of growth in Personal Well-being.
-In 2018, the expected subjective well-being as measured using the Personal Well-being Inventory among those who believed in Spirit or Life Force but not a God was expected to grow to (\(\hat{y}\) = 6.67, se =.03, \(\Delta \hat{y} = 0.17\)).
-Among those who believed in both a Spirit/Life-Force and a God Personal Well-Being was expected to grow to (\(\hat{y}\) = 6.69, se = 0.02,\(\Delta \hat{y} = .21\)).
-In practical terms, such growth from spiritual beliefs among to the expected gain in Personal Well-being about twice as large as the expected benefit of being employed and about about a third as large as the expected benefit of having a romantic partner.
-In comparison, Personal Well-Being among skeptical population who neither believed in a God nor a Life Force remained stable \(\hat{y} = 6.66, se = 0.03, \Delta \hat{y} = .02\)).
-We also observe stability in Personal Well-being among the population that believed in God but was skeptical about a Spirit or Life Force \(\hat{y} = 6.70, se = 0.04, \Delta \hat{y} = .04\)).
-(See Figure 2).
-\end{enumerate}
+Two fundamental question arise from the patterns we observe in this study.
+The first question is why those who believed in Spirit or Life Force were initially lower in their personal well-being and life-satisfaction in the initial years of this study in comparison to religious and non-religious skeptics?
+During the early years of New Zealand Attitudes and Values data collection New Zealand was emerging from the twin impacts of a global financial recession and a series of devastating earthquakes that destroyed 30\% of Christchurch, New Zealand's second largest city {[}CITE{]}.
+Orthogonal research observes that economic and ecological distress are associated with increases in traditional religious beliefs {[}CITE, CITE{]}.
+Previous research finds an increase in traditional religious affiliation in the wake of the Christchurch earthquakes, with no immediate effect in subjective well-being.
+Experimental research has found among religious people death anxiety provokes acceptance of non-traditional religious beliefs {[}cite{]}.
+Because spiritual belief items were only introduced during the year of the earthquakes, we cannot here resolve whether the shifting relationship between spirit beliefs and well-being were the result of demand for spiritual beliefs among those secular and religious people who were adversely affected by the financial crisis or natural disaster.
+On the other hand, such exogenous influences might have adversely affected those with Spirit/Life force beliefs in the early phases of this study, and the patterns observed here are better understood as a regression to a population mean.
+Though beyond the scope of this study, how subjective well-being changes in response to changes in spiritual beliefs, as well as special sensitivities to distress among people with spiritual beliefs are important questions for future investigations.
+Despite the longitudinal quality of this study, a causal explanation remains elusive.
 
-Similarly, over time, belief in a Spirit or Life Force predicted a trajectory of growth in Life Satisfaction.
-In 2018, the expected subjective well-being as measured using the Life Satisfaction Inventory among those who believed in Spirit or Life Force but not a God was expected to grow to (\(\hat{y} = 5.01, se = 0.02, \Delta \hat{y} = 0.07\)).
-In practical terms, such growth from spiritual beliefs is approximately equivalent to the expected gain in Life Satisfaction from being employed and a sixth as large as the expected benefit to Life Satisfaction from having a romantic partner.
-Among those who believed in both a Spirit/Life-Force and a God Personal Well-Being was expected to grow to (\(\hat{y} = 5.03, se = 0.02, \Delta \hat{y} = 0.12\)).
-In practical terms, such growth from spiritual beliefs is approximately equivalent to the 1.5 times the expected gain in Life Satisfaction from being employed and almost a third the expected benefit to Life Satisfaction from having a romantic partner.
-In comparison, Life Satisfaction among skeptical population who neither believed in a God nor a Life Force remained stable \(\hat{y} = 4.94, se =0.02, \Delta \hat{y} = 0\)).
-We also observe stability in Personal Well-being among the population that believed in God but was skeptical about a Spirit or Life Force \(\hat{y} = 5.00, se = 0.03, \Delta \hat{y} = -.02\)) (See Figure 3).
+Second, our task here has been to clarify whether beliefs predict growth in well-being.
+However, as we have indicated at the outset, the term \enquote{spirituality} applies to a family of concepts.
+We have not attempted to model the diverse philosophical, social, and behavioral attributes within people who adhere to spirit beliefs.
+Moreover, people might identify as spiritual without having any accompanying belief in a Spirit or Life force.
+The task of clarifying how distinct features across a diverse family of \enquote{spirituality} concepts affect perceptions of individual well-being remain intriguing horizons.
 
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{2}
-\tightlist
-\item
-  Such observed growth in subjective well-being among the population that believes in a Spirit or Life Force holds even after adjusting for demographic factors that are known to affect well-being.
-Additionally we infer growth from 9 annual waves of data collection (8 years) in a nationally diverse sample.
-On this basis of this window of time and sample, we infer that such growth in Life Satisifacton and in Personal Well-being reflects changes in the populations that believed in Spirit or Life Force over that period.
-\end{enumerate}
+\hypertarget{importance}{%
+\subsection{Importance}\label{importance}}
 
-Questions:
-1. Why initially lower well-being in NZ? \(\dots\)
-+ historical contingencies?
-+ possibility of non-linear futures, are we observing a regression to the mean?
-+ individual factors: earthquakes, however..why the slope for spirit beliefs (and not God)?
-+ other things besides beliefs? behaviors, social interactions, perspectives\ldots{}
-Taken as a whole, our results cannot refute the entire narrative of the positive relationship between spirituality and emotional well-being.
-First, our analysis focuses only on belief.
-Although belief is an important part of spirituality it is not by any means spirituality itself.
-Individuals may identify as spiritual without any accompanying belief in a spirit or life force.
-Moreover, practices and rituals associated with different forms of spirituality that were not tested in this study may have large effects.
-Second, despite flawed methods, past studies may still be accurate.
-New Zealand is a largely secular country, so belief in a spirit or life force may have different impacts on subjective well-being outcomes in this context than in other, more religious contexts.
-Overall, though, our findings add to the existing literature, helping to untangle the multifaceted relationship among religion, spirituality, and subjective well-being.
-\(\dots\) fruitful horizons for future research.
-
-Importance:
-1. Methods: Disentangling \(\dots\) concepts/ longitudinal
-2.
-Growing population: practical importance.
-In many Western countries, people are increasingly identifying as \enquote{spiritual but not religious} (Swinton, 2001; Zinnbauer \& Pargament, 2005).
-\ldots{}
-+ risks: Risks? \ldots{} Other studies find that those who have a spiritual understanding of life without a religious framework are at an increased risk of having mental disorders (King et al., 2013).
-+ However, the observed growth suggests that any kind of existential belief - regardless of whether a god is involved - might affect subjective well-being in much the same way.
-This result is good news as increasing numbers of individuals continue to report spiritual but not necessarily religious beliefs.
-
-\hypertarget{limitations}{%
-\subsection{Limitations}\label{limitations}}
-
-This article makes useful contributions to the continued debate regarding the relationship between religion and spirituality and mental health outcomes, but it is important to acknowledge several limitations.
-First, we relied upon self-reported data from the NZAVS.
-Although this is an accurate method for measuring belief in a spirit or life force, it may be a less accurate manner of measuring subjective well-being.
-It is possible that self-administration of our emotional well-being (personal well-being and life satisfaction) scales could result in mismeasurement.
-Second, we may not capture the full range spiritualities, which likely extend beyond belief and might even be inherently emotional.
-Thus, while our method avoided tautology, it might have also excluded participants who consider themselves to be spiritual but do not subscribe to a belief in a spirit or life force.
+Despite its limitations, our study is important both for its methods and its findings.
+We demonstrate that spirituality can be operationalised to avoid the pitfalls of tautology evident in previous research by focusing on states of beliefs, that concepts of such belief states can be refined to enable distinctions necessary to investigate spirituality in mixed secular and religious populations, and that the question of whether beliefs states predict growth in subjective well-being can be investigated in a longitudional setting.
+Our finding that, indeed, spiritual beliefs predict growth in perceived well-being and life satisfaction may hold practical importance.
+Previous research indicates that religious beliefs are associated with having a rewarding life.
+Western populations are increasingly identifying as \enquote{spiritual but not religious} (Swinton, 2001; Zinnbauer \& Pargament, 2005), raising the specter that the psychological from traditional might be receding.
+Here, we observe that even among people who lack a traditional religious belief in God, believing in a Spirit or Life force predicts incremental improvements in well-being that sum to practically large effects.
+Though the contextual and psychological mechanisms that underpin such growth will require many decades of research, this preliminary finding is good news for those increasing numbers of individuals continue to report spiritual beliefs but who do not affirm traditional religious beliefs.
 
@@
 
-CS leads and manages the NZAVS, and collected the data for this study. BH and JB conceived
-of the study.
-BH wrote the first draft of the manuscript and performed the analysis with input from JB.
-BH conducted background research and scientific literature review.
-All authors contributed substantially to the intellectual content of the study and the final manuscript.
+BH conceived of the study, wrote the first draft of the manuscript, and performed the initial data analysis with input from JB.
+EW developed the study with BH and JB and revised the manuscript.
+DD revised the manuscript.
+CS was responsible for NZAVS data collection and revised the manuscript.
+JB developed the study with BH and EW, performed the data re-analysis, and revised the manuscript.
 
@@
 
-The New Zealand Attitudes and Values Study is supported by a grant from the Templeton
-Religion Trust (TRT0196). The funders had no role in the preparation of the article or the
-decision to publish.
-
-\hypertarget{supplemental-material}{%
-\section{Supplemental Material}\label{supplemental-material}}
-
-S1. Statistical Analysis in R
-Statistical analysis was performed using R version 3.6.1 (2019-07-05).
-Platform: x86\_64-w64-mingw32 (64-bit), running under: Windows 10 (Action of the Toes).
-Generalized Linear Mixed Models were generated using the glmmTMB (Brooks et al.~2017; Knowles and Frederick 2019) package in R.
-In addition, we used the following R packages: dplyr (Hadley Wickham et al.~2019), tidyr (H.
-Wickham and Henry 2019), sjPlot (D.
-Lüdecke 2020), ggeffects (Daniel Lüdecke 2018), table1 and their dependencies.
-Tables and coefficient plots were generated using the glmmTMB, ggeffects, and sjPlot packages in R.
-Models for both subjective well-being variables used family = gaussian, assuming a normal distribution.
-We ran each individual's unique ID as a random-effect.
-The tab\_model and plot\_model commands in sjPlot allowed for the creation of tables and coefficient plots.
-Predicted probability plots were generated using the ggpredict command in the ggeffects package.
-The table1 package in R was used for the creation of descriptive statistics tables.
+The New Zealand Attitudes and Values Study is supported by a grant from the Templeton Religion Trust (TRT0196). The funders had no role in the preparation of the article or the decision to publish.
 
@@
 \hypertarget{refs}{}
+\leavevmode\hypertarget{ref-R-equatiomatic}{}%
+Anderson, D., \& Heiss, A. (2020). \emph{Equatiomatic: Transform models into 'latex' equations}. Retrieved from \url{https://github.com/datalorax/equatiomatic}
+
 \leavevmode\hypertarget{ref-Ano2005-hx}{}%
@@
 
+\leavevmode\hypertarget{ref-R-citr}{}%
+Aust, F. (2019). \emph{Citr: 'RStudio' add-in to insert markdown citations}. Retrieved from \url{https://github.com/crsh/citr}
+
+\leavevmode\hypertarget{ref-R-papaja}{}%
+Aust, F., \& Barth, M. (2020). \emph{papaja: Create APA manuscripts with R Markdown}. Retrieved from \url{https://github.com/crsh/papaja}
+
+\leavevmode\hypertarget{ref-R-Matrix}{}%
+Bates, D., \& Maechler, M. (2019). \emph{Matrix: Sparse and dense matrix classes and methods}. Retrieved from \url{https://CRAN.R-project.org/package=Matrix}
+
+\leavevmode\hypertarget{ref-R-lme4}{}%
+Bates, D., Mächler, M., Bolker, B., \& Walker, S.
+(2015).
+Fitting linear mixed-effects models using lme4.
+\emph{Journal of Statistical Software}, \emph{67}(1), 1--48.
+\url{https://doi.org/10.18637/jss.v067.i01}
+
 \leavevmode\hypertarget{ref-Blackwell2017-oq}{}%
 Blackwell, M., Honaker, J., \& King, G. (2017). A unified approach to measurement error and missing data: Details and extensions. \emph{Sociological Methods \& Research}.
+
+\leavevmode\hypertarget{ref-R-prettycode}{}%
+Csárdi, G. (2019). \emph{Prettycode: Pretty print r code in the terminal}. Retrieved from \url{https://CRAN.R-project.org/package=prettycode}
 
@@
 Diener, E., Emmons, R. A., Larsen, R. J., \& Griffin, S. (1985). The satisfaction with life scale. \emph{Journal of Personality Assessment}.
+
+\leavevmode\hypertarget{ref-R-Rcpp_b}{}%
+Eddelbuettel, D., \& Balamuta, J.
+J.
+(2017).
+Extending extitR with extitC++: A Brief Introduction to extitRcpp.
+\emph{PeerJ Preprints}, \emph{5}, e3188v1.
+\url{https://doi.org/10.7287/peerj.preprints.3188v1}
+
+\leavevmode\hypertarget{ref-R-Rcpp_a}{}%
+Eddelbuettel, D., \& François, R. (2011). Rcpp: Seamless R and C++ integration. \emph{Journal of Statistical Software}, \emph{40}(8), 1--18. \url{https://doi.org/10.18637/jss.v040.i08}
 
@@
 
+\leavevmode\hypertarget{ref-R-purrr}{}%
+Henry, L., \& Wickham, H. (2020). \emph{Purrr: Functional programming tools}. Retrieved from \url{https://CRAN.R-project.org/package=purrr}
+
+\leavevmode\hypertarget{ref-R-Amelia}{}%
+Honaker, J., King, G., \& Blackwell, M.
+(2011a).
+Amelia II: A program for missing data.
+\emph{Journal of Statistical Software}, \emph{45}(7), 1--47.
+Retrieved from \url{http://www.jstatsoft.org/v45/i07/}
+
 \leavevmode\hypertarget{ref-Honaker2011-yu}{}%
-Honaker, J., King, G., Blackwell, M., \& Others. (2011). Amelia II: A program for missing data. \emph{J. Stat. Softw.}, \emph{45}(7), 1--47.
+Honaker, J., King, G., Blackwell, M., \& Others. (2011b). Amelia II: A program for missing data. \emph{J. Stat. Softw.}, \emph{45}(7), 1--47.
 
@@
 Oxford University Press.
+
+\leavevmode\hypertarget{ref-R-texreg}{}%
+Leifeld, P.
+(2013).
+texreg: Conversion of statistical model output in R to LaTeX and HTML tables.
+\emph{Journal of Statistical Software}, \emph{55}(8), 1--24.
+Retrieved from \url{http://dx.doi.org/10.18637/jss.v055.i08}
+
+\leavevmode\hypertarget{ref-R-ggeffects}{}%
+Lüdecke, D. (2018). Ggeffects: Tidy data frames of marginal effects from regression models. \emph{Journal of Open Source Software}, \emph{3}(26), 772. \url{https://doi.org/10.21105/joss.00772}
+
+\leavevmode\hypertarget{ref-R-sjPlot}{}%
+Lüdecke, D. (2020). \emph{SjPlot: Data visualization for statistics in social science}. Retrieved from \url{https://CRAN.R-project.org/package=sjPlot}
+
+\leavevmode\hypertarget{ref-R-parameters}{}%
+Lüdecke, D., Ben-Shachar, M.
+S., Patil, I., \& Makowski, D.
+(2020a).
+Parameters: Extracting, computing and exploring the parameters of statistical models using R.
+\emph{Journal of Open Source Software}, \emph{5}(53), 2445.
+\url{https://doi.org/10.21105/joss.02445}
+
+\leavevmode\hypertarget{ref-R-see}{}%
+Lüdecke, D., Ben-Shachar, M.
+S., Waggoner, P., \& Makowski, D.
+(2020b).
+See: Visualisation toolbox for 'easystats' and extra geoms, themes and color palettes for 'ggplot2'.
+\emph{CRAN}.
+\url{https://doi.org/10.5281/zenodo.3952153}
+
+\leavevmode\hypertarget{ref-R-report}{}%
+Makowski, Dominique, Lüdecke, \& Daniel.
+(2019).
+The report package for r: Ensuring the use of best practices for results reporting.
+\emph{CRAN}.
+Retrieved from \url{https://github.com/easystats/report}
+
+\leavevmode\hypertarget{ref-R-styler}{}%
+Müller, K., \& Walthert, L. (2020). \emph{Styler: Non-invasive pretty printing of r code}. Retrieved from \url{https://CRAN.R-project.org/package=styler}
+
+\leavevmode\hypertarget{ref-R-tibble}{}%
+Müller, K., \& Wickham, H. (2020). \emph{Tibble: Simple data frames}. Retrieved from \url{https://CRAN.R-project.org/package=tibble}
+
+\leavevmode\hypertarget{ref-R-patchwork}{}%
+Pedersen, T. L. (2020). \emph{Patchwork: The composer of plots}. Retrieved from \url{https://CRAN.R-project.org/package=patchwork}
+
+\leavevmode\hypertarget{ref-R-base}{}%
+R Core Team. (2020). \emph{R: A language and environment for statistical computing}. Vienna, Austria: R Foundation for Statistical Computing. Retrieved from \url{https://www.R-project.org/}
+
+\leavevmode\hypertarget{ref-R-table1}{}%
+Rich, B. (2020). \emph{Table1: Tables of descriptive statistics in html}. Retrieved from \url{https://CRAN.R-project.org/package=table1}
 
@@
 
+\leavevmode\hypertarget{ref-R-ggplot2}{}%
+Wickham, H. (2016). \emph{Ggplot2: Elegant graphics for data analysis}. Springer-Verlag New York. Retrieved from \url{https://ggplot2.tidyverse.org}
+
+\leavevmode\hypertarget{ref-R-stringr}{}%
+Wickham, H. (2019). \emph{Stringr: Simple, consistent wrappers for common string operations}. Retrieved from \url{https://CRAN.R-project.org/package=stringr}
+
+\leavevmode\hypertarget{ref-R-forcats}{}%
+Wickham, H. (2020a). \emph{Forcats: Tools for working with categorical variables (factors)}. Retrieved from \url{https://CRAN.R-project.org/package=forcats}
+
+\leavevmode\hypertarget{ref-R-tidyr}{}%
+Wickham, H. (2020b). \emph{Tidyr: Tidy messy data}. Retrieved from \url{https://CRAN.R-project.org/package=tidyr}
+
+\leavevmode\hypertarget{ref-R-tidyverse}{}%
+Wickham, H., Averick, M., Bryan, J., Chang, W., McGowan, L.
+D., François, R., \ldots{} Yutani, H.
+(2019).
+Welcome to the tidyverse.
+\emph{Journal of Open Source Software}, \emph{4}(43), 1686.
+\url{https://doi.org/10.21105/joss.01686}
+
+\leavevmode\hypertarget{ref-R-dplyr}{}%
+Wickham, H., François, R., Henry, L., \& Müller, K. (2020). \emph{Dplyr: A grammar of data manipulation}. Retrieved from \url{https://CRAN.R-project.org/package=dplyr}
+
+\leavevmode\hypertarget{ref-R-readr}{}%
+Wickham, H., Hester, J., \& Francois, R. (2018). \emph{Readr: Read rectangular text data}. Retrieved from \url{https://CRAN.R-project.org/package=readr}
+
+\leavevmode\hypertarget{ref-R-ggsci}{}%
+Xiao, N. (2018). \emph{Ggsci: Scientific journal and sci-fi themed color palettes for 'ggplot2'}. Retrieved from \url{https://CRAN.R-project.org/package=ggsci}
+
 \leavevmode\hypertarget{ref-Yonker2012-zg}{}%
@@
 
+\leavevmode\hypertarget{ref-R-gghighlight}{}%
+Yutani, H. (2020). \emph{Gghighlight: Highlight lines and points in 'ggplot2'}. Retrieved from \url{https://CRAN.R-project.org/package=gghighlight}
+
 \leavevmode\hypertarget{ref-Zaza2005-ac}{}%

### 2020-11-18 Manuscript.Rmd
small fixes

@@
 First, previous research has frequently operationalised spirituality using measures of well-being.
-For example, The Daily Spiritual Experience Scale, which is widely used in spiritual mental health research, includes items such as "I feel thankful for my blessings" and "I feel deep inner peace or harmony" to measure spirituality [@Underwood2002-hg].
+For example, The *Daily Spiritual Experience Scale*, which is widely used in spiritual mental health research, includes items such as "I feel thankful for my blessings" and "I feel deep inner peace or harmony" to measure spirituality [@Underwood2002-hg].
 As Koenig observes, it is unsurprising that spirituality predict well-being when the same items are used to measure both constructs.

### 2020-11-18 Manuscript.tex
small fixes

@@
 First, previous research has frequently operationalised spirituality using measures of well-being.
-For example, The Daily Spiritual Experience Scale, which is widely used in spiritual mental health research, includes items such as \enquote{I feel thankful for my blessings} and \enquote{I feel deep inner peace or harmony} to measure spirituality (Underwood \& Teresi, 2002).
+For example, The \emph{Daily Spiritual Experience Scale}, which is widely used in spiritual mental health research, includes items such as \enquote{I feel thankful for my blessings} and \enquote{I feel deep inner peace or harmony} to measure spirituality (Underwood \& Teresi, 2002).
 As Koenig observes, it is unsurprising that spirituality predict well-being when the same items are used to measure both constructs.

### 2020-11-19 Manuscript.tex
Updates from Overleaf

@@
 \title{National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Beliefs}
-\author{Ben Highland\textsuperscript{1}, Everett L. Worthington\textsuperscript{2}, Don E Davis\textsuperscript{3}, Chris G. Sibley\textsuperscript{4}, \& Joseph Bulbulia\textsuperscript{5}}
+\author{Benjamin Highland\textsuperscript{1}, Everett L. Worthington\textsuperscript{2}, Don E Davis\textsuperscript{3}, Chris G. Sibley\textsuperscript{4}, \& Joseph Bulbulia\textsuperscript{5}}
 \date{}
@@
 
-1 Wake Forest Medical School 2 Department of Psychology Virginia Commonwealth University 3 Department of Psychology Georgia State University 4 School of Psychology University of Auckland 5 School of Psychology Victoria University
+1 Wake Forest School of Medicine 2 Department of Psychology Virginia Commonwealth University 3 Department of Psychology Georgia State University 4 School of Psychology University of Auckland 5 School of Psychology Victoria University
 
-Correspondence concerning this article should be addressed to Ben Highland, Ben's Postal address. E-mail: \href{mailto:benhighland6420@gmail.com}{\nolinkurl{benhighland6420@gmail.com}}
+Correspondence concerning this article should be addressed to Benjamin Highland, Benjamin's E-mail: \href{mailto:benhighland6420@gmail.com}{\nolinkurl{benhighland6420@gmail.com}}
 
@@
 
-\abstract{
-Previous research finds an association between spirituality and subjective well-being.
+\abstract{Previous research finds an association between spirituality and subjective well-being.
 However, the widespread use of tautological spirituality scales, poorly defined concepts of spirituality, and heavy reliance on cross-sectional samples cast doubts.
@@
 However, in support of previous speculation, beliefs in a spirit or life force predict increasing personal well-being and life satisfaction over time relative to disbelief.
-These finding are robust after to known demographic influences, and intriguingly, hold among those who believe in a God but disbelieve in a Spirit or Life Force.
+These finding are robust after adjusting for known demographic influences, and intriguingly, hold among those who believe in a God but disbelieve in a Spirit or Life Force.
 The recent growth in spiritual beliefs and decline in traditional religion across many industrial societies motivates further causal investigations of the mechanisms by which spiritual beliefs lead to growth in subjective well-being.
@@
 Koenig aptly describes this approach as \enquote{tautological} (Koenig, 2008).
-Indeed, recent systematic-review of spirituality/mental health research found that nearly 45\% of previous studies employe tautological scales (Garssen et al., 2016).
+Indeed, recent systematic-review of spirituality/mental health research found that nearly 45\% of previous studies employ tautological scales (Garssen et al., 2016).
 The first challenge, then, is to define measures of spirituality that are not well-being measures dressed up in other words.
@@
 Notably, researchers who have operationalised spirituality constructs as distinct from religion constructs have observed a negative association between spirituality and subjective well-being (King et al., 2013).
-Thus to infer how spirituality affects subjective well-being, it is important to disentangle the concepts of spirituality from concepts of religion, and to carefully investigate features of contructs in populations where these dimensions overlap, partially overlap, and diverge.
+Thus to infer how spirituality affects subjective well-being, it is important to disentangle the concepts of spirituality from concepts of religion, and to carefully investigate features of constructs in populations where these dimensions overlap, partially overlap, and diverge.
 
 A third challenge is to investigate the relationship between spirituality and subjective well-being in nationally diverse samples over individual life-spans.
-Reliable inference requires observing change within people who differ dimensions of spirituality over time, and assing how their lives turn out (Garssen \& Visser, 2016).
+Reliable inference requires observing change within people who differ in dimensions of spirituality over time, and assessing how their lives turn out (Garssen \& Visser, 2016).
 Previous investigations of spirituality and well-being, including high-quality reviews and meta-analyses, have relied on cross-sectional samples (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky et al., 2005; Smith et al., 2003; Visser et al., 2010; Yonker et al., 2012).
@@
 
-Statistical analysis was performed using R version 4.0.2 R version 3.6.1 (2020-06-22) 2019-07-05).
+Statistical analysis was performed using R version 4.0.2 (2020-06-22).
 We analyze data from participants who responded to the NZAVS survey at least three four times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of 21,705 (N = 20,979 with complete responses).
@@
 We centered and scaled the education, deprivation, and political conservativism.
-We centred age was centred at and put into decade-long intervals.
+We centred and put into decade-long intervals.
 The model equation(s) are described in Equation 1.
@@
 
-Because missing values can result in biased estimates (Blackwell, Honaker, \& King, 2017), we multiply imputed 10 datasets using the AmeliaII package in R(Honaker et al., 2011b).We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences.
+Because missing values can result in biased estimates (Blackwell, Honaker, \& King, 2017), we multiply imputed 10 datasets using the Amelia II package in R (Honaker et al., 2011b).
+We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences.
 Details of this analysis can be found at: \url{https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing}.
@@
 \item
-  The effect of Age.10yrs.C is reliably positive (beta = 0.08, SE = 7.38e-03, std. beta = 0.07, p \textless{} .001). Older people have greater personal well-beingwellbeing.
+  The effect of Age.10yrs.C is reliably positive (beta = 0.08, SE = 7.38e-03, std. beta = 0.07, p \textless{} .001). Older people have greater personal well-being.
 \item
-  The effect of Deprivation (S) is reliably negative (beta = -0.14, SE = 7.50e-03, std. beta = -0.08, p \textless{} .001). Deprived people have lower personal well-beingwellbeing.
+  The effect of Deprivation (S) is reliably negative (beta = -0.14, SE = 7.50e-03, std. beta = -0.08, p \textless{} .001). Deprived people have lower personal well-being.
 \item
-  The effect of Education (S) is reliably positive (beta = 0.12, SE = 8.74e-03, std. beta = 0.07, p \textless{} .001). Educated people greater personal well-beingwellbeing
+  The effect of Education (S) is reliably positive (beta = 0.12, SE = 8.74e-03, std. beta = 0.07, p \textless{} .001). Educated people greater personal well-being.
 \item
-  The effect of Employed is reliably positive (beta = 0.09, SE = 0.01, std. beta = 0.05, p \textless{} .001). Employed people have greater personal well-beingwellbeing.
+  The effect of Employed is reliably positive (beta = 0.09, SE = 0.01, std. beta = 0.05, p \textless{} .001). Employed people have greater personal well-being.
 \item
@@
 \hypertarget{personal-well-being-years-x-beliefs}{%
-\subsection{Personal well-being: years X beliefs}\label{personal-well-being-years-x-beliefs}}
+\subsection{Personal Well-Being: Years X Beliefs}\label{personal-well-being-years-x-beliefs}}
 
@@
 \item
-  The effect of years \(\times\) belief in a SpiritExcludesGod\_ is reliably positive (beta = 0.02, SE = 4.86e-03, std.
+  The effect of years \(\times\) belief in a SpiritExcludesGod\ is reliably positive (beta = 0.02, SE = 4.86e-03, std.
 beta = 0.02, p \textless{} .01).
@@
 beta = 0.03, p \textless{} .001).
-We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their personal well-being over time
+We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their personal well-being over time.
 \item
-  The effect of Years \(\times\) belief in GodExcludesSpirit is not reliable(beta = 1.96e-03, SE = 8.29e-03, std.
+  The effect of Years \(\times\) belief in GodExcludesSpirit is not reliable (beta = 1.96e-03, SE = 8.29e-03, std.
 beta = 2.36e-03, p = 0.813).
@@
 \item
-  The effect of beliefs in GodAndSpirit is not reliable (beta = -0.04, SE = 0.02, std. beta = 0.04, p = 0.105). We do not infer growth over time in Life Satisfaction among this (Skeptic) population
+  The effect of beliefs in GodAndSpirit is not reliable (beta = -0.04, SE = 0.02, std. beta = 0.04, p = 0.105). We do not infer growth over time in Life Satisfaction among this (Skeptic) population.
 \item
   The effect of beliefs in GodExcludesSpirit is reliably positive (beta = 0.07, SE = 0.04, std.
-beta = 0.05, p = 0.062).We infer that overall the population that believes in a God but does not believe in a spirit is not reliably different from the Skeptics population in Life Satisfaction.
+beta = 0.05, p = 0.062).
+We infer that overall the population that believes in a God but does not believe in a spirit is not reliably different from the Skeptics population in Life Satisfaction.
 \item
@@
 \item
-  The effect of identifying as Asian is not reliably different from people of European ethnicity in life satisfaction(beta = -0.04, SE = 0.03, std. beta = -0.03, p = 0.223).
+  The effect of identifying as Asian is not reliably different from people of European ethnicity in life satisfaction (beta = -0.04, SE = 0.03, std. beta = -0.03, p = 0.223).
 \item
@@
 \item
-  The effect of having a Parnter is reliably positive (beta = 0.42, SE = 0.01, std. beta = 0.35, p \textless{} .001).
+  The effect of having a Partner is reliably positive (beta = 0.42, SE = 0.01, std. beta = 0.35, p \textless{} .001).
 \item
@@
 \hypertarget{life-satisfaction-years-x-beliefs}{%
-\subsection{Life Satisfaction: years X beliefs}\label{life-satisfaction-years-x-beliefs}}
+\subsection{Life Satisfaction: Years X Beliefs}\label{life-satisfaction-years-x-beliefs}}
 
@@
 beta = 0.03, p \textless{} .001).
-We infer that, unlike the Skeptic population, people with beliefs in a Spirit or Life Force who do not believe in a God tend to grow in their Life Satisfaction over time.
+We infer that, unlike the Skeptic population, people with beliefs in a Spirit or Life Force who do not believe in a God tend to grow in their life satisfaction over time.
 
@@
 beta = 0.02, p \textless{} .001).
-We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their Life Satisfaction over time
+We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their life satisfaction over time.
   \item
@@
 beta = -2.70e-03, p = 0.799).
-Thus, God believers who doubt in a Spirit or Life Force start maintain higher Life-Satisfaction than do other groups, but they are not expected to grow in Life-Satisfaction over time.
+Thus, God believers who doubt in a Spirit or Life Force start maintain higher life-satisfaction than do other groups, but they are not expected to grow in life-satisfaction over time.
   \end{itemize}
@@
 The inference from this analysis remains the same as from the analysis in which we delete missing values row-wise, by panel wave.
-In all analyses, those who believe in a spirit or life-force experience greater growth in both Personal Well-being and Life Satisfaction over time.
+In all analyses, those who believe in a spirit or life-force experience greater growth in both personal well-being and life satisfaction over time.
 
@@
   At the outset of the study, in 2010, the expected subjective well-being as measured using the Personal Well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God (\(\hat{y}\) = 6.50, se =.03) and lower among those who believed in both a Spirit/Life-Force and a God (\(\hat{y}\) = 6.48, se =.03) in comparison both to the skeptical population who neither believed in a God nor a life force (\(\hat{y}\) = 6.64,se =.03) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force (\(\hat{y}\) 6.66, se =.05).
-(See Figure 2) Similarly,the expected subjective well-being as measured using the Life Satisfaction Inventory is lower among those who believed in Spirit or Life Force but not a God (\(\hat{y}\) = 4.86, se =.02) and lower among those who believed in both a Spirit/Life-Force and a God (\(\hat{y}\) = 4.91, se =.02) in comparison both with the skeptical population who neither believed in a God nor a life force (\(\hat{y}\) = 4.94,se =.02) and the population that believed in a God but was skeptical about belief in a Spirit or Life Force (\(\hat{y}\) 5.02, se =.02) (see Figure 3)
+(See Figure 2) Similarly, the expected subjective well-being as measured using the Life Satisfaction Inventory is lower among those who believed in Spirit or Life Force but not a God (\(\hat{y}\) = 4.86, se =.02) and lower among those who believed in both a Spirit/Life-Force and a God (\(\hat{y}\) = 4.91, se =.02) in comparison both with the skeptical population who neither believed in a God nor a life force (\(\hat{y}\) = 4.94,se =.02) and the population that believed in a God but was skeptical about belief in a Spirit or Life Force (\(\hat{y}\) 5.02, se =.02) (see Figure 3)
 \item
   However, over time, belief in a Spirit or Life Force predicts a trajectory of growth in personal well-being.
-In 2018, personal well-being among who believe in Spirit or Life Force but not a God had grown to (\(\hat{y}\) = 6.67, se =.03, \(\Delta \hat{y} = 0.17\)).
+In 2018, personal well-being among those who believe in Spirit or Life Force but not a God had grown to (\(\hat{y}\) = 6.67, se =.03, \(\Delta \hat{y} = 0.17\)).
 Among those who believed in both a Spirit/Life-Force and a God Personal Well-Being had grown to (\(\hat{y}\) = 6.69, se = 0.02,\(\Delta \hat{y} = .21\)).
@@
 By the same token, personal well-being among those who believe in a God but are skeptical about a Spirit/Life Force was constant \(\hat{y} = 6.70, se = 0.04, \Delta \hat{y} = .04\)).
-Similarly, over time, belief in a Spirit or Life Force predicts a trajectory of growth in Life Satisfaction.
+Similarly, over time, belief in a Spirit or Life Force predicts a trajectory of growth in life satisfaction.
 In 2018, life satisfaction among those who believe in Spirit or Life Force but not a God had grown to (\(\hat{y} = 5.01, se = 0.02, \Delta \hat{y} = 0.07\)).
-In practical terms, such growth is equivalent to the expected gain in Life Satisfaction from being employed and a sixth as large as the expected benefit to Life Satisfaction from having a romantic partner.
+In practical terms, such growth is equivalent to the expected gain in life satisfaction from being employed and a sixth as large as the expected benefit to life satisfaction from having a romantic partner.
 Among those who believed in both a Spirit/Life-Force and a God, life satisfaction had grown to (\(\hat{y} = 5.03, se = 0.02, \Delta \hat{y} = 0.12\)).
@@
 Despite its limitations, our study is important both for its methods and its findings.
-We demonstrate that spirituality can be operationalised to avoid the pitfalls of tautology evident in previous research by focusing on states of beliefs, that concepts of such belief states can be refined to enable distinctions necessary to investigate spirituality in mixed secular and religious populations, and that the question of whether beliefs states predict growth in subjective well-being can be investigated in a longitudional setting.
+We demonstrate that spirituality can be operationalised to avoid the pitfalls of tautology evident in previous research by focusing on states of beliefs, that concepts of such belief states can be refined to enable distinctions necessary to investigate spirituality in mixed secular and religious populations, and that the question of whether beliefs states predict growth in subjective well-being can be investigated in a longitudinal setting.
 Our finding that, indeed, spiritual beliefs predict growth in perceived well-being and life satisfaction may hold practical importance.

### 2020-11-19 Manuscript.Rmd
revisions

@@
     institution   : "School of Psychology University of Auckland"
-  - name          : "Joseph Bulbulia"
+  - name          : "Joseph A. Bulbulia"
     affiliation   : "5"
@@
 The relationship between spirituality and psychological well-being is a matter of enduring fascination [@Ellsworth2010-yu].
-Most previous attempts to quantify the relationship between these two domains find a positive association: spirituality predicts well-being [@Russinova2002-rq; @Ginsburg1995-jr; @Zaza2005-ac; @Koenig2010-gk; @Smith2003-re; @Smith2003-re; @Koenig2001-ow].
+Most previous attempts to quantify the relationship between these two domains find a positive association: spirituality predicts well-being [@Russinova2002-rq; @Ginsburg1995-jr; @Zaza2005-ac; @Koenig2010-gk; @Smith2003-re; @Koenig2001-ow].
 However, critics have raised three credible challenges.
 
-<!-- In spirituality/psychological wellbeing research,sSsubjective in spirituality health research, psychological well-being is typically decomposed into two categories: emotional well-being (i.e., the presence or absence of positive feelings about life) and functional well-being (i.e., psychological and social well-being) [@Keyes2002-bb].
+<!-- In spirituality/psychological wellbeing research,subjective in spirituality health research, psychological well-being is typically decomposed into two categories: emotional well-being (i.e., the presence or absence of positive feelings about life) and functional well-being (i.e., psychological and social well-being) [@Keyes2002-bb].
 -->
@@
 Koenig aptly describes this approach as "tautological" [@Koenig2008-lv].
-Indeed, recent systematic-review of spirituality/mental health research found that nearly 45% of previous studies employe tautological scales [@Garssen2016-kb].
+Indeed, recent systematic-review of spirituality/mental health research found that nearly 45% of previous studies employ tautological scales [@Garssen2016-kb].
 The first challenge, then, is to define measures of spirituality that are not well-being measures dressed up in other words.
@@
 Notably, researchers who have operationalised spirituality constructs as distinct from religion constructs have observed a negative association between spirituality and subjective well-being [@King2013-cg].
-Thus to infer how spirituality affects subjective well-being, it is important to disentangle the concepts of spirituality from concepts of religion, and to carefully investigate features of contructs in populations where these dimensions overlap, partially overlap, and diverge.
+Thus to infer how spirituality affects subjective well-being, it is important to disentangle the concepts of spirituality from concepts of religion, and to carefully investigate features of constructs in populations where these dimensions overlap, partially overlap, and diverge.
 
 A third challenge is to investigate the relationship between spirituality and subjective well-being in nationally diverse samples over individual life-spans.
-Reliable inference requires observing change within people who differ dimensions of spirituality over time, and assing how their lives turn out [@Garssen2016-km].
+Reliable inference requires observing change within people who differ dimensions of spirituality over time, and assessing how their lives turn out [@Garssen2016-km].
 Previous investigations of spirituality and well-being, including high-quality reviews and meta-analyses, have relied on cross-sectional samples [@Ano2005-hx; @Hackney2003-rs; @Sawatzky2005-rw; @Smith2003-re; @Visser2010-kq; @Yonker2012-zg].
@@
 
-Here, we addresses the three challenges of previous in the following ways: First, to avoid a tautological elision of "spirituality" and "well-being" we adopt a non-affective measure of spirituality as a "believe in a Spirit or Life Force." On the other side, to measure subjective well-being, we use a well-validated scale for Personal Well-being and a well-validated scale for Life-Satisfaction (see Method).
+Here, we addresses the three challenges of previous in the following ways: First, to avoid a tautological elision of "spirituality" and "well-being" we adopt a non-affective measure of spirituality as a "believe in a Spirit or Life Force." On the other side, to measure subjective well-being, we use a well-validated scale for personal well-being and a well-validated scale for Life-Satisfaction (see Method).
 Second, to disentangle spiritual beliefs from traditional religious beliefs and skeptical beliefs we adopt a four-level categorical indicator combining the binary indicator of belief in a Spirit or Life Force with the binary indicator "Do you believe in some form of a spirit or life force?" The levels of this four-level categorical indicator for spiritual, religious, and skeptical beliefs are as follows: (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who nevertheless believe in a Spirit or Life Force; (3) Those who believe in a God but do not believe in a Spirit or Life force; (4) Those who who believe in a both God and a Spirit or Life Force.
@@
 The Time 10 (2018) wave of the NZAVS included 47,951 respondents.
-We analyzed data from participants who responded to our survey at least three four times between Time 2 (2010) and Time 10 (2018), resulting in a sample of $N$ = 21,705 ($N$ = 20,979 with complete responses).
+We analyzed data from participants who responded to our survey at least three times between Time 2 (2010) and Time 10 (2018), resulting in a sample of $N$ = 21,705 ($N$ = 20,979 with complete responses).
 
@@
 \hspace{6pt}    Euro &  (\%) &  (\%) & 6731 (78.8\%) & 11024 (81.8\%) &  (\%) & 9671 (82.8\%) & 13389 (83.4\%) & 11814 (83.3\%) & 12427 (83.5\%)\\ 
-\hspace{6pt}    Maori &  (\%) &  (\%) & 1188 (13.9\%) & 1608 (11.9\%) &  (\%) & 1341 (11.5\%) & 1749 (10.9\%) & 1620 (11.4\%) & 1677 (11.3\%)\\ 
+\hspace{6pt}    Māori &  (\%) &  (\%) & 1188 (13.9\%) & 1608 (11.9\%) &  (\%) & 1341 (11.5\%) & 1749 (10.9\%) & 1620 (11.4\%) & 1677 (11.3\%)\\ 
 \hspace{6pt}    Pacific &  (\%) &  (\%) & 284 (3.3\%) & 339 (2.5\%) &  (\%) & 277 (2.4\%) & 331 (2.1\%) & 243 (1.7\%) & 261 (1.8\%)\\ 
@@
 ### Deprivation
-We measured the socio-economic status of participants’ immediate (small area) neighborhood using the 2013 New Zealand Deprivation Index, which uses aggregate census information about the residents of small neighborhood-type units to assign a decile-rank index from 1 (most affluent) to 10 (most impoverished) [@Atkinson2014-ex].
+We measured the socioeconomic status of participants’ immediate (small area) neighborhood using the 2013 New Zealand Deprivation Index, which uses aggregate census information about the residents of small neighborhood-type units to assign a decile-rank index from 1 (most affluent) to 10 (most impoverished) [@Atkinson2014-ex].
 The index is based on a Principal Components Analysis of the following nine variables (in weighted order): the proportion of adults who received a means-tested benefit, household income, proportion not owning own home, proportion single-parent families, proportion unemployed, proportion lacking qualifications, proportion household crowding, proportion no telephone access, and proportion no car access.
@@
 ### Ethnicity
-Ethnicity was assessed using four basic categories: (1) New Zealand European/Pakeha, (2) Maori, (3) Pacific Islander, and (4) Asian.
+Ethnicity was assessed using four basic categories: (1) New Zealand European/Pakeha, (2) Māori, (3) Pacific Islander, and (4) Asian.
 For detailed information pertaining to yearly responses and missingness, see Table 1.
@@
 Statistical analysis was performed using R version 4.0.2 R version 3.6.1 (2020-06-22) 2019-07-05).
-We analyze data from participants who responded to the NZAVS survey at least three four times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of  21,705 (N = 20,979 with complete responses).
+We analyze data from participants who responded to the NZAVS survey at least three times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of  21,705 (N = 20,979 with complete responses).
 To model growth in subjective well-being we use generalized linear mixed models.
@@
 &\quad \beta_{6}(\operatorname{Deprivation.S}) + \beta_{7}(\operatorname{Edu.S}) + \beta_{8}(\operatorname{Employed})\ + \\
-&\quad \beta_{9}(\operatorname{EthnicCats}_{\operatorname{Maori}}) + \beta_{10}(\operatorname{EthnicCats}_{\operatorname{Pacific}}) + \beta_{11}(\operatorname{EthnicCats}_{\operatorname{Asian}})\ + \\
+&\quad \beta_{9}(\operatorname{EthnicCats}_{\operatorname{Māori}}) + \beta_{10}(\operatorname{EthnicCats}_{\operatorname{Pacific}}) + \beta_{11}(\operatorname{EthnicCats}_{\operatorname{Asian}})\ + \\
 &\quad \beta_{12}(\operatorname{Male}_{\operatorname{1}}) + \beta_{13}(\operatorname{Partner}_{\operatorname{Has\ }}) + \beta_{14}(\operatorname{Pol.Orient.S})\ + \\
@@
 
-```{r fig.cap = "Combination Coefficient Plots for Personal Well-being and Life Satisfaction Spirit Belief Longitudinal Models"}
+```{r fig.cap = "Combination Coefficient Plots for personal well-being and life satisfaction longitudinal growth models"}
 knitr::include_graphics("Figs/USE_2_coefficientplots-1.png")
@@
                                    & $(0.01)$               & $(0.01)$               \\
-EthnicCatsMaori                    & $\mathbf{-0.30}^{***}$ & $\mathbf{-0.06}^{**}$  \\
+EthnicCatsMāori                    & $\mathbf{-0.30}^{***}$ & $\mathbf{-0.06}^{**}$  \\
                                    & $(0.03)$               & $(0.02)$               \\
@@
 
-We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict personal well-being (PWI) with Years (0-9), Beliefs (four categories), Age.10yrs.C, Deprivation.S, Edu.S, Employed, EthnicCats, Male, Partner, Pol.Orient.S and Urban, withId as random effects.
-The model included Id as random effects.
-The model's intercept, corresponding to Years = 0, Beliefs = Skeptic, Age = (sample mean), Deprivation.S = (sample mean), Edu.S =(sample mean), Employed = Not Employed, EthnicCats = Euro, Male = Not Mail, Partner = No Partner, Pol.Orient.S = (sample mean),, Urban = Not Urban and Id = 2, is at 4.94 (SE = 0.02, 95% CI [4.89, 4.99], p < .001) is at 6.63 (SE = 0.03, 95% CI [6.57, 6.70], p < .001).
+We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict personal well-being (PWI) with years (0-9), beliefs (four categories), Age.10yrs.C, deprivation (S), education (S), employed, ethnic categories, male, partner, political conservativism (S) and urban, with Id as a random effect.
+The model's intercept, corresponding to years = 0, beliefs = Skeptic, age = (sample mean), deprivation.S = (sample mean), education (S) =(sample mean), employed = not employed, ethnic Categories = European, male = not male, partner = no partner, political conservativism (S) = (sample mean), urban = not urban, is at 4.94 (SE = 0.02, 95% CI [4.89, 4.99], p < .001) is at 6.63 (SE = 0.03, 95% CI [6.57, 6.70], p < .001).
 Within this model:
@@
 beta = 3.19e-03, p = 0.458).
-We infer that Skeptics personal well-being is not increasing in the Skeptic population
-  - The effect of Beliefs in SpiritExcludesGod is reliably negative (beta = -0.14, SE = 0.03, std.
+We infer that personal well-being did not increase in the Skeptic population between 2010 and 2018.
+  - The effect of Beliefs in *Spirit Excludes God*  is reliably negative (beta = -0.14, SE = 0.03, std.
 beta = -0.03, p < .001).
-We infer that overall the population that believes in a Spirit but does not believe in a God is expected to have a lower personal well-being than skeptics.
-  - The effect of Beliefs in GodAndSpirit is reliably negative (beta = -0.16, SE = 0.03, std.
+We infer that in 2010, the population that believed in a Spirit but did not believe in a God had lower personal well-being than did Skeptics.
+  - The effect of Beliefs in *God And Spirit*  is reliably negative (beta = -0.16, SE = 0.03, std.
 beta = -0.02, p < .001).
-We infer that overall the population that believes in God and a spirit or life force is expected to have a lower Personal Well-being than skeptics.
-  - The effect of Beliefs in GodExcludesSpirit is not reliably different from the skeptic population (beta = 0.02, SE = 0.05, std. beta = 0.02, p = 0.669).
-  - The effect of Age.10yrs.C is reliably positive  (beta = 0.08, SE = 7.38e-03, std. beta = 0.07, p < .001). Older people have greater personal well-beingwellbeing.
-  - The effect of Deprivation (S) is reliably negative (beta = -0.14, SE = 7.50e-03, std. beta = -0.08, p < .001). Deprived people have lower personal well-beingwellbeing.
-  - The effect of Education (S) is reliably positive  (beta = 0.12, SE = 8.74e-03, std. beta = 0.07, p < .001). Educated people greater personal well-beingwellbeing
-  - The effect of Employed is reliably positive  (beta = 0.09, SE = 0.01, std. beta = 0.05, p < .001). Employed people have greater personal well-beingwellbeing.
-  - The effect of identifying as Maori is reliably negative (beta = -0.30, SE = 0.03, std. beta = -0.18, p < .001).
+We infer that in 2010, the population that believed in a God and a Spirit or Life Force had lower Personal Well-being than did Skeptics.
+  - The effect of Beliefs in *God Excludes Spirit*  is not reliable (beta = 0.02, SE = 0.05, std.
+beta = 0.02, p = 0.669).
+We infer that in 2010, the population that believed in a God and doubted a Spirit or Life Force similar levels of personal well-being to Skeptics.
+  - The effect of Age.10yrs.C is reliably positive  (beta = 0.08, SE = 7.38e-03, std. beta = 0.07, p < .001). Older people tend to have greater personal well-being.
+  - The effect of Deprivation (S) is reliably negative (beta = -0.14, SE = 7.50e-03, std. beta = -0.08, p < .001). Deprived people tend to have lower personal well-being.
+  - The effect of Education (S) is reliably positive  (beta = 0.12, SE = 8.74e-03, std. beta = 0.07, p < .001). Educated people tend to have greater personal well-being.
+  - The effect of Employed is reliably positive  (beta = 0.09, SE = 0.01, std. beta = 0.05, p < .001).
+  - The effect of identifying as M\={a}ori is reliably negative (beta = -0.30, SE = 0.03, std. beta = -0.18, p < .001).
   - The effect of identifying as Pacific is reliably negative (beta = -0.29, SE = 0.05, std. beta = -0.17, p < .001). 
   - The effect of identifying as Asian is not reliable (beta = -0.08, SE = 0.05, std. beta = -0.05, p = 0.089).
-  - The effect of identifying as Male is reliably negative (beta = -0.11, SE = 0.02, std. beta = -0.06, p < .001).
-  - The effect of having a Partner is reliably positive (beta = 0.61, SE = 0.02, std. beta = 0.37, p < .001).
-  - The effect of Political Conservativism (S) is reliably positive  (beta = 0.02, SE = 5.67e-03, std. beta = 0.01, p < .001).
-  - The effect of living in an Urban area is reliably negative (beta = -0.02, SE = 0.01, std. beta = -0.01, p < .05).
+  - The effect of identifying as male is reliably negative (beta = -0.11, SE = 0.02, std. beta = -0.06, p < .001). 
+  - The effect of having a partner is reliably positive (beta = 0.61, SE = 0.02, std. beta = 0.37, p < .001).
+  - The effect of political conservativism (S) is reliably positive  (beta = 0.02, SE = 5.67e-03, std. beta = 0.01, p < .001). 
+  - The effect of living in an Urban area is reliably negative (beta = -0.02, SE = 0.01, std. beta = -0.01, p < .05). T
 
@@
 
-```{r fig.cap = "Combination Coefficient Plots for Personal Well-being and Life Satisfaction Spirit Belief Longitudinal Models"}
+```{r fig.cap = "Combination Coefficient plots for personal well-being and life satisfaction longitudinal growth models"}
 knitr::include_graphics("Figs/USEPWI_expected-1.png")
@@
 
-  - The effect of years $\times$ belief in a SpiritExcludesGod_ is reliably positive  (beta = 0.02, SE = 4.86e-03, std.
+  - The effect of years $\times$ belief in a *Spirit Excludes God*  is reliably positive  (beta = 0.02, SE = 4.86e-03, std.
 beta = 0.02, p < .01).
 We infer that, unlike the skeptic population, people with beliefs in a spirit or life force who do not believe in a God tend to grow in their personal well-being over time.
-  - The effect of years $\times$ in GodAndSpirit is reliably positive  (beta = 0.02, SE = 4.41e-03, std.
+  - The effect of years $\times$ in *God And Spirit*  is reliably positive  (beta = 0.02, SE = 4.41e-03, std.
 beta = 0.03, p < .001).
 We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their personal well-being over time
-  - The effect of Years $\times$ belief in GodExcludesSpirit is not reliable(beta = 1.96e-03, SE = 8.29e-03, std.
+  - The effect of Years $\times$ belief in *God Excludes Spirit*  is not reliable(beta = 1.96e-03, SE = 8.29e-03, std.
 beta = 2.36e-03, p = 0.813).
@@
 
-We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict Life Satisfaction (LIFESAT) with Years (0-9), Beliefs (four categories), Age.10yrs.C, Deprivation (S), Education (S), Employed, EthnicCats, Male, Partner, Pol.Orient.S and Urban, with  Id as random effects.
+We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict personal well-being (PWI) with years (0-9), beliefs (four categories), Age.10yrs.C, deprivation (S), education (S), employed, ethnic categories, male, partner, political conservativism (S) and urban, with Id as a random effect.
 Standardized parameters were obtained by fitting the model on a standardized version of the dataset.
 The model's total explanatory power is substantial (conditional R2 = 0.69).
-The model's intercept, corresponding to Years = 0, Beliefs = Skeptic, Age = (sample mean),, Deprivation.S = (sample mean), Education =(sample mean), Employed = Not Employed, EthnicCats = Euro, Male = Not Mail, Partner = No Partner, Pol.Orient.S = (sample mean), Urban = Not Urban and Id = 2, is at 4.94 (SE = 0.02, 95% CI [4.89, 4.99], p < .001).
+The model's intercept, corresponding to years = 0, beliefs = Skeptic, age = (sample mean), deprivation.S = (sample mean), education (S) =(sample mean), employed = not employed, ethnic Categories = European, male = not male, partner = no partner, political conservativism (S) = (sample mean), urban = not urban is at 4.94 (SE = 0.02, 95% CI [4.89, 4.99], p < .001).
 Within this model:
 
-  - The effect of years is not reliable (beta = -5.50e-04, SE = 2.69e-03, std. beta = -9.32e-04, p = 0.838).  We do not infer growth over time in Life Satisfaction among the Skeptic population
-  - The effect of beliefs in SpiritExcludesGod is reliably negative (beta = -0.08, SE = 0.02, std.
+  - The effect of years is not reliable (beta = -5.50e-04, SE = 2.69e-03, std.
+beta = -9.32e-04, p = 0.838).
+We infer that life satisfaction did not increase in the Skeptic population between 2010 and 2018.
+  - The effect of beliefs in *Spirit Excludes God* is reliably negative (beta = -0.08, SE = 0.02, std.
 beta = 0.01, p < .001).
 We infer that overall the population that believes in a Spirit but does not believe in a God is expected to have a lower Life Satisfaction than skeptics.
-  - The effect of beliefs in GodAndSpirit is not reliable (beta = -0.04, SE = 0.02, std. beta = 0.04, p = 0.105). We do not infer growth over time in Life Satisfaction among this (Skeptic) population
-  - The effect of beliefs in GodExcludesSpirit is reliably positive (beta = 0.07, SE = 0.04, std.
+  - The effect of beliefs in *God And Spirit* is not reliable (beta = -0.04, SE = 0.02, std.
+beta = 0.04, p = 0.105).
+People who believe in God but doubt a Spirit/Life force are expected to have had similar levels of life satisfaction in Wave 10 as skeptics.
+  - The effect of beliefs in *God Excludes Spirit* is reliably positive (beta = 0.07, SE = 0.04, std.
 beta = 0.05, p = 0.062).We infer that overall the population that believes in a God but does not believe in a spirit is not reliably different from  the Skeptics population in Life Satisfaction.
@@
   - The effect of being Employed is reliably positive (beta = 0.08, SE = 9.69e-03, std. beta = 0.07, p < .001). People who are employed tend to have greater Life Satisfaction
-  - The effect of identifying as Maori is reliably negative (beta = -0.06, SE = 0.02, std. beta = -0.05, p < .01).  People who belong to the minority Maori group tend to have lower Life Satisfaction.
+  - The effect of identifying as Māori is reliably negative (beta = -0.06, SE = 0.02, std. beta = -0.05, p < .01).  People who belong to the minority Māori group tend to have lower Life Satisfaction.
   - The effect of identifying as Pacific is reliably negative (beta = -0.08, SE = 0.04, std.
@@
   - The effect of identifying as Asian is not reliably different from people of European ethnicity in life satisfaction(beta = -0.04, SE = 0.03, std. beta = -0.03, p = 0.223).
-  - The effect of identifying as Male is reliably negative (beta = -0.19, SE = 0.01, std. beta = -0.16, p < .001).
-  - The effect of having a Parnter is reliably positive (beta = 0.42, SE = 0.01, std. beta = 0.35, p < .001).
-  - The effect of Political Conservativism (S) is reliably positive (beta = 0.01, SE = 4.23e-03, std. beta = 0.01, p < .01).
-  - The effect of living in an Urban area is reliably negative (beta = -0.02, SE = 8.85e-03, std. beta = -0.02, p < .05).
+  - The effect of identifying as male is reliably negative (beta = -0.19, SE = 0.01, std. beta = -0.16, p < .001).
+  - The effect of having a partner is reliably positive (beta = 0.42, SE = 0.01, std. beta = 0.35, p < .001).
+  - The effect of political conservativism (S) is reliably positive (beta = 0.01, SE = 4.23e-03, std. beta = 0.01, p < .01).
+  - The effect of living in an urban area is reliably negative (beta = -0.02, SE = 8.85e-03, std. beta = -0.02, p < .05).
 
@@
 ```
-  - The interaction of years $\times$ beliefs in SpiritExcludesGod is reliably positive (beta = 0.02, SE = 3.66e-03, std.
+  - The interaction of years $\times$ beliefs in *Spirit Excludes God* ludesGod is reliably positive (beta = 0.02, SE = 3.66e-03, std.
 beta = 0.03, p < .001).
 We infer that, unlike the Skeptic population, people with beliefs in a Spirit or Life Force who do not believe in a God tend to grow in their Life Satisfaction over time.
-  - The interaction of years $\times$ beliefsGodAndSpirit is reliably positive (beta = 0.01, SE = 3.32e-03, std.
+  - The interaction of years $\times$ *God And Spirit*  is reliably positive (beta = 0.01, SE = 3.32e-03, std.
 beta = 0.02, p < .001).
-We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their Life Satisfaction over time
-  - The effect of years $\times$ beliefs in GodExcludesSpirit is not reliable (beta = -1.59e-03, SE = 6.25e-03, std.
+We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their life satisfaction over time
+  - The effect of years $\times$ beliefs in *God Excludes Spirit*  is not reliable (beta = -1.59e-03, SE = 6.25e-03, std.
 beta = -2.70e-03, p = 0.799).
@@
 1.
-At the outset of the study, in 2010, the expected subjective well-being as measured using the Personal Well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God ($\hat{y}$ =  6.50, se =.03) and lower among those who believed in both a Spirit/Life-Force and a God ($\hat{y}$ =  6.48, se =.03) in comparison both to the skeptical population who neither believed in a God nor a life force ($\hat{y}$ = 6.64,se =.03) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 6.66, se =.05).
-(See Figure 2)  Similarly,the expected subjective well-being as measured using the Life Satisfaction Inventory is lower among those who believed in Spirit or Life Force but not a God ($\hat{y}$ = 4.86, se =.02) and lower among those who believed in both a Spirit/Life-Force and a God  ($\hat{y}$ = 4.91, se =.02) in comparison both with the skeptical population who neither believed in a God nor a life force ($\hat{y}$ = 4.94,se =.02) and the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 5.02, se =.02) (see Figure 3)
+At the outset of the study, in 2010, the expected subjective well-being as measured using the personal well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God ($\hat{y}$ =  6.50, se =.03) and lower among those who believed in both a Spirit/Life-Force and a God ($\hat{y}$ =  6.48, se =.03) in comparison both to the skeptical population who neither believed in a God nor a life force ($\hat{y}$ = 6.64,se =.03) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 6.66, se =.05).
+(See Figure 2)  Similarly,the expected subjective well-being as measured using the life satisfaction Inventory is lower among those who believed in Spirit or Life Force but not a God ($\hat{y}$ = 4.86, se =.02) and lower among those who believed in both a Spirit/Life-Force and a God  ($\hat{y}$ = 4.91, se =.02) in comparison both with the skeptical population who neither believed in a God nor a life force ($\hat{y}$ = 4.94,se =.02) and the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 5.02, se =.02) (see Figure 3)
 
@@
 In 2018, personal well-being among who believe in Spirit or Life Force but not a God had grown to ($\hat{y}$ = 6.67, se =.03, $\Delta \hat{y} = 0.17$).
-Among those who believed in both a Spirit/Life-Force and a God Personal Well-Being had grown to ($\hat{y}$ = 6.69, se = 0.02,$\Delta \hat{y} = .21$).
+Among those who believed in both a Spirit/Life-Force and a God, personal well-being had grown to ($\hat{y}$ = 6.69, se = 0.02,$\Delta \hat{y} = .21$).
 To place this effect in practical perspective, such growth in personal well-being is about twice as large as the expected benefit of being employed and about a third as large as the expected benefit from having a romantic partner.
@@
 By the same token, personal well-being among those who believe in a God but are skeptical about a Spirit/Life Force was constant $\hat{y} =  6.70, se = 0.04, \Delta \hat{y} = .04$).
-Similarly, over time, belief in a Spirit or Life Force predicts a trajectory of growth in Life Satisfaction.
+Similarly, over time, belief in a Spirit or Life Force predicts a trajectory of growth in life satisfaction.
 In 2018, life satisfaction among those who believe in Spirit or Life Force but not a God had grown to ($\hat{y} = 5.01, se = 0.02, \Delta \hat{y} = 0.07$).
-In practical terms, such growth is equivalent to the expected gain in Life Satisfaction from being employed and a sixth as large as the expected benefit to Life Satisfaction from having a romantic partner.
+In practical terms, such growth is equivalent to the expected gain in life satisfaction from being employed and a sixth as large as the expected benefit to life satisfaction from having a romantic partner.
 Among those who believed in both a Spirit/Life-Force and a God, life satisfaction had grown to ($\hat{y} = 5.03, se = 0.02, \Delta \hat{y} = 0.12$).
@@
 ## Key Questions
+
 Two fundamental question arise from the patterns we observe in this study.
@@
 Previous research finds an increase in traditional religious affiliation in the wake of the Christchurch earthquakes, with no immediate effect in subjective well-being.
-Experimental research has found among religious people death anxiety provokes acceptance of non-traditional religious beliefs [cite].
+Experimental research has found among religious people death anxiety provokes acceptance of non-traditional religious beliefs [CITE].
 Because spiritual belief items were only introduced during the year of the earthquakes, we cannot here resolve whether the shifting relationship between spirit beliefs and well-being were the result of demand for spiritual beliefs among those secular and religious people who were adversely affected by the financial crisis or natural disaster.
@@
 Despite its limitations, our study is important both for its methods and its findings.
-We demonstrate that spirituality can be operationalised to avoid the pitfalls of tautology evident in previous research by focusing on states of beliefs, that concepts of such belief states can be refined to enable distinctions necessary to investigate spirituality in mixed secular and religious populations, and that the question of whether beliefs states predict growth in subjective well-being can be investigated in a longitudional setting.
+We demonstrate that spirituality can be operationalised to avoid the pitfalls of tautology evident in previous research by focusing on states of beliefs, that concepts of such belief states can be refined to enable distinctions necessary to investigate spirituality in mixed secular and religious populations, and that the question of whether beliefs states predict growth in subjective well-being can be investigated in a longitudinal setting.
 Our finding that, indeed, spiritual beliefs predict growth in perceived well-being and life satisfaction may hold practical importance.
-Previous research indicates that religious beliefs are associated with having a rewarding life.
-Western populations are increasingly identifying as "spiritual but not religious" [@Zinnbauer2005-vz; @Swinton2001-vr], raising the specter that the psychological from traditional might be receding.
+Previous research indicates that religious beliefs are associated with psychological well-being [@Smith2003-re].
+Western populations are increasingly identifying as "spiritual but not religious" [@Zinnbauer2005-vz; @Swinton2001-vr].
 Here, we observe that even among people who lack a traditional religious belief in God, believing in a Spirit or Life force predicts incremental improvements in well-being that sum to practically large effects.
@@
 ### Authors contributions
-BH conceived of the study, wrote the first draft of the manuscript, and performed the initial data analysis with input from JB.
-EW developed the study with BH and JB and revised the manuscript.
+BH conceived of the study.
+BH wrote the first draft of the manuscript, and performed the initial data analysis with input from JB.
+EW improved the approach with BH and JB and revised the manuscript.
 DD revised the manuscript.
 CS was responsible for NZAVS data collection and revised the manuscript.
-JB developed the study with BH and EW, performed the data re-analysis, and revised the manuscript.
+JB improved the approach with BH and EW, performed a data re-analysis, and revised the manuscript.
 

### 2020-11-19 Analysis.Rmd
revised manuscript, bib

@@
 library("patchwork")
+library("easystats")
 library("parameters")
@@
 library("sjPlot")
+library("texreg")
 library("see")
@@
 library("citr")
-library("texreg")
+
 ```
@@
   dplyr:: mutate(Years = as.numeric(years))%>%
+  dplyr:: mutate(yearsC = scale(years, center=TRUE,scale=FALSE))%>%
   dplyr::mutate(Age.10yrs = (Age/10))%>%
@@
 table(dfben$Beliefs)
+range(dfben$years)
+range(dfben$Years)
 ```
@@
 
-```{r, include =FALSE}
-furniture::table1(dfben, Age, NZdep,Education,Employed,EthnicCategories,Male,Partner, Pol.Orient,Urban, Beliefs, LIFESAT,PWI,
-                    splitby = ~Wave, overall=F, output = "latex2", booktabs=TRUE)
+```{r}
+table1::table1(~ Beliefs|Wave, data = dfben, overall = T)
 ```
 
-```{r, include=FALSE}
-r<-report::report_participants(
-  dfben,
-  age = "Age",
-  sex = "Male",
-  education = "Education",
-  participants = "Id",
-  group = "Wave",
-  spell_n = FALSE,
-)
+```{r, include =FALSE}
+#furniture::table1(dfben, Age, NZdep,Education,Employed,EthnicCategories,Male,Partner, Pol.Orient,Urban, Beliefs, LIFESAT,PWI,
+                  #  splitby = ~Wave, overall=F, output = "latex2", booktabs=TRUE)
 ```
+
+<!-- ```{r, include=FALSE} -->
+<!-- r<-easystats::report_participants( -->
+<!--   dfben, -->
+<!--   age = "Age", -->
+<!--   sex = "Male", -->
+<!--   education = "Education", -->
+<!--   participants = "Id", -->
+<!--   group = "Wave", -->
+<!--   spell_n = FALSE, -->
+<!-- ) -->
+<!-- ??report_participants -->
+<!-- ``` -->
 
@@
 # mod.2 <-glmmTMB(PWI ~ Years * Beliefs + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id) + (1|GenCohort), data= dfben)
+```
+
+```{r check collinearity}
+# Note High VIFS from 
+performance::check_collinearity(mod.1)
+performance::check_collinearity(mod.2)
+```
+
+# Inference holds but VIFS for years x beliefs
+
+```{r model centering years, include = FALSE}
+library(lme4)
+mod.1a <-lmer (PWI ~  1 +  yearsC * Beliefs +  Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id), data= dfben, na.action = na.omit)
+mod.2a <-lmer (LIFESAT ~ 1 + yearsC * Beliefs+ Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id), data= dfben, na.action = na.omit)
+summary(mod.1)
+summary(mod.2)
+# identical result
+performance::check_collinearity(mod.1a)
+performance::check_collinearity(mod.2a)
+
+# address reviewer worry about deprivation and partner being correlated
+m<-lmer (NZdep ~ Partner + (1|Id), data= dfben, na.action = na.omit)
+tab_model(m)
+model_parameters(m)
 ```
@@
 g1/g2
+range(dfben$yearsC)
 ```
@@
   ggtitle("Predicted Values of Personal Wellbeing") 
-
+p1
 pr.PWI
@@
 
-```{r eq1}
-library(equatiomatic)
-print(extract_eq(mod.1, wrap = TRUE,terms_per_line =3))
-```
+<!-- ```{r eq1} -->
+<!-- library(equatiomatic) -->
+<!-- print(extract_eq(mod.1, wrap = TRUE,terms_per_line =3)) -->
+<!-- ``` -->
 
-PWI Model 2 equation
-```{r eq2}
-library(equatiomatic)
-print(extract_eq(mod.2, wrap = TRUE, use_coefs=TRUE,terms_per_line = 3))
-```
+<!-- PWI Model 2 equation -->
+<!-- ```{r eq2} -->
+<!-- library(equatiomatic) -->
+<!-- print(extract_eq(mod.2, wrap = TRUE, use_coefs=TRUE,terms_per_line = 3)) -->
+<!-- ``` -->
 
-## report
+<!-- ## report -->
 
-### Model 1
-model 1 report: for publication change all "significants" to "reliable"
+<!-- ### Model 1 -->
+<!-- model 1 report: for publication change all "significants" to "reliable" -->
 
-short report:
-```{r, report1, cache =TRUE, dependson=mod.1,include=FALSE}
-report::model_text(mod.1)
-```
+<!-- short report: -->
+<!-- ```{r, report1, cache =TRUE, dependson=mod.1,include=FALSE} -->
+<!-- easystats::model_text(mod.1) -->
+<!-- ``` -->
 
-short report
-```{r, cache=TRUE, include=FALSE }
-library(report)
-r1<- report::report(mod.1)
-table_short(r1)
-text_short(r1)
-```
+<!-- short report -->
+<!-- ```{r, cache=TRUE, include=FALSE } -->
+<!-- library(report) -->
+<!-- r1<- easystats::report(mod.1) -->
+<!-- easystats::table_short(r1) -->
+<!-- easystats::text_short(r1) -->
+<!-- ``` -->
 
-```{r, cache =TRUE, dependson=mod.1, include=FALSE}
-r1<- report::report(mod.1)
-report::text_long(r1)
-```
-short report
-```{r, cache =TRUE, dependson=mod.2,include=FALSE}
-# Model 2
-# For publication change all "significants" to "reliable"
+<!-- ```{r, cache =TRUE, dependson=mod.1, include=FALSE} -->
+<!-- r1<- easystats::report(mod.1) -->
+<!-- easystats::text_long(r1) -->
+<!-- ``` -->
+<!-- short report -->
+<!-- ```{r, cache =TRUE, dependson=mod.2,include=FALSE} -->
+<!-- # Model 2 -->
+<!-- # For publication change all "significants" to "reliable" -->
 
-report::model_text(mod.2)
-```
+<!-- easystats::model_text(mod.2) -->
+<!-- ``` -->
 
-short report
-```{r,include=FALSE}
-r2 <- report::report(mod.2)
-report::text_short(r2)
-report::table_short(r2)
-```
+<!-- short report -->
+<!-- ```{r,include=FALSE} -->
+<!-- r2 <- easystats::report(mod.2) -->
+<!-- easystats::text_short(r2) -->
+<!-- easystats::table_short(r2) -->
+<!-- ``` -->
 
-long report
-```{r,include=FALSE}
-r2<- report::report(mod.2)
-report::text_long(r2)
-```
+<!-- long report -->
+<!-- ```{r,include=FALSE} -->
+<!-- r2<- easystats::report(mod.2) -->
+<!-- easystats::text_long(r2) -->
+<!-- ``` -->
 
@@
 ##
-sessionInfo()
 ```

### 2020-11-19 Manuscript.Rmd
revised manuscript, bib

@@
 output            : papaja::apa6_pdf
-extra_dependencies: ["rotating", "booktabs", "graphicx","longtable"]
+extra_dependencies: ["rotating", "booktabs", "graphicx","longtable","graphicx"]
 header-includes:
@@
 \hspace{6pt}    Euro &  (\%) &  (\%) & 6731 (78.8\%) & 11024 (81.8\%) &  (\%) & 9671 (82.8\%) & 13389 (83.4\%) & 11814 (83.3\%) & 12427 (83.5\%)\\ 
-\hspace{6pt}    Māori &  (\%) &  (\%) & 1188 (13.9\%) & 1608 (11.9\%) &  (\%) & 1341 (11.5\%) & 1749 (10.9\%) & 1620 (11.4\%) & 1677 (11.3\%)\\ 
+\hspace{6pt}    Maori &  (\%) &  (\%) & 1188 (13.9\%) & 1608 (11.9\%) &  (\%) & 1341 (11.5\%) & 1749 (10.9\%) & 1620 (11.4\%) & 1677 (11.3\%)\\ 
 \hspace{6pt}    Pacific &  (\%) &  (\%) & 284 (3.3\%) & 339 (2.5\%) &  (\%) & 277 (2.4\%) & 331 (2.1\%) & 243 (1.7\%) & 261 (1.8\%)\\ 
@@
 ## Demographic Indicators
+We included standard demographic indicators used in New Zealand Attitudes and Values Research: [NZAVS modelling standards](https://cdn.auckland.ac.nz/assets/psych/about/our-research/nzavs/NZAVSTechnicalDocuments/NZAVS-Technical-Documents-e11-Default-Statistical-Standards.pdf)
 
@@
 ### Ethnicity
-Ethnicity was assessed using four basic categories: (1) New Zealand European/Pakeha, (2) Māori, (3) Pacific Islander, and (4) Asian.
+Ethnicity was assessed using four basic categories: (1) New Zealand European/Pakeha, (2) Maori, (3) Pacific Islander, and (4) Asian.
 For detailed information pertaining to yearly responses and missingness, see Table 1.
@@
 &\quad \beta_{6}(\operatorname{Deprivation.S}) + \beta_{7}(\operatorname{Edu.S}) + \beta_{8}(\operatorname{Employed})\ + \\
-&\quad \beta_{9}(\operatorname{EthnicCats}_{\operatorname{Māori}}) + \beta_{10}(\operatorname{EthnicCats}_{\operatorname{Pacific}}) + \beta_{11}(\operatorname{EthnicCats}_{\operatorname{Asian}})\ + \\
+&\quad \beta_{9}(\operatorname{EthnicCats}_{\operatorname{Maori}}) + \beta_{10}(\operatorname{EthnicCats}_{\operatorname{Pacific}}) + \beta_{11}(\operatorname{EthnicCats}_{\operatorname{Asian}})\ + \\
 &\quad \beta_{12}(\operatorname{Male}_{\operatorname{1}}) + \beta_{13}(\operatorname{Partner}_{\operatorname{Has\ }}) + \beta_{14}(\operatorname{Pol.Orient.S})\ + \\
@@
 Because missing values can result in biased estimates [@Blackwell2017-oq], we multiply imputed 10 datasets using the AmeliaII package in R[@Honaker2011-yu].We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences.
-Details of this analysis can be found at: https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing.
+Additionally the results were robust to filtering of participants of N >3 to N = 9 waves.
+Additionally, the findings reported below were robust to modeling only the effects of years and beliefs, without adjusting for the standard NZAVS demographic covariates.
+The supplement describing these analysis is located at: https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing.
 
@@
 ```{r fig.cap = "Combination Coefficient Plots for personal well-being and life satisfaction longitudinal growth models"}
-knitr::include_graphics("Figs/USE_2_coefficientplots-1.png")
+knitr::include_graphics("/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Figs/USE_2_coefficientplots-1.png")
 ```
@@
 \[
+
 \begin{table}
-\centering 
-\caption{"Longitudinal Models"}\label{}
+\begin{center}
 \scalebox{0.6}{
@@
 \toprule
- & Personal Well-Being & Life Satisfaction \\
+ & PWI & Life Satisfaction \\
 \midrule
@@
                                    & $(0.01)$               & $(0.01)$               \\
-Employed1                          & $\mathbf{0.09}^{***}$  & $\mathbf{0.08}^{***}$  \\
+EmployedEmployed                   & $\mathbf{0.09}^{***}$  & $\mathbf{0.08}^{***}$  \\
                                    & $(0.01)$               & $(0.01)$               \\
-EthnicCatsMāori                    & $\mathbf{-0.30}^{***}$ & $\mathbf{-0.06}^{**}$  \\
+EthnicCatsMaori                    & $\mathbf{-0.30}^{***}$ & $\mathbf{-0.06}^{**}$  \\
                                    & $(0.03)$               & $(0.02)$               \\
@@
                                    & $(0.02)$               & $(0.01)$               \\
-Partner1                           & $\mathbf{0.61}^{***}$  & $\mathbf{0.42}^{***}$  \\
+PartnerHas Partner                 & $\mathbf{0.61}^{***}$  & $\mathbf{0.42}^{***}$  \\
                                    & $(0.02)$               & $(0.01)$               \\
@@
 }
+\caption{Regression Models}
+\label{table:coefficients}
+\end{center}
 \end{table}
@@
   - The effect of being Employed is reliably positive (beta = 0.08, SE = 9.69e-03, std. beta = 0.07, p < .001). People who are employed tend to have greater Life Satisfaction
-  - The effect of identifying as Māori is reliably negative (beta = -0.06, SE = 0.02, std. beta = -0.05, p < .01).  People who belong to the minority Māori group tend to have lower Life Satisfaction.
+  - The effect of identifying as Maori is reliably negative (beta = -0.06, SE = 0.02, std. beta = -0.05, p < .01).  People who belong to the minority Maori group tend to have lower Life Satisfaction.
   - The effect of identifying as Pacific is reliably negative (beta = -0.08, SE = 0.04, std.
@@
 ```{r fig.cap = "Prediction Plot Life Satisfaction Spirit Belief Longitudinal Models"}
-knitr::include_graphics("Figs/USElifsat_expected-1.png")
+knitr::include_graphics("/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Figs/USElifsat_expected-1.png")
 ```
@@
 ```{r fig.cap = "Imputed data analysis: predicted personal wellbeing by years X beliefs"}
-knitr::include_graphics("Figs/predicted_PWI_impute-1.png")
+knitr::include_graphics("/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Figs/predicted_PWI_impute-1.png")
 ```
@@
 ```{r fig.cap = "Imputed data analysis: predicted personal wellbeing by years X beliefs"}
-knitr::include_graphics("Figs/predicted_lifesat_imputed-1.png")
+knitr::include_graphics("/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Figs/predicted_lifesat_imputed-1.png")
 ```
@@
 In 2018, personal well-being among who believe in Spirit or Life Force but not a God had grown to ($\hat{y}$ = 6.67, se =.03, $\Delta \hat{y} = 0.17$).
-Among those who believed in both a Spirit/Life-Force and a God, personal well-being had grown to ($\hat{y}$ = 6.69, se = 0.02,$\Delta \hat{y} = .21$).
+Among those who believed in both a Spirit/Life-Force and a God, personal well-being had grown to ($\hat{y} = 6.69, se = 0.02, \Delta \hat{y} = .21$).
 To place this effect in practical perspective, such growth in personal well-being is about twice as large as the expected benefit of being employed and about a third as large as the expected benefit from having a romantic partner.
@@
 The first question is why those who believed in Spirit or Life Force were initially lower in their personal well-being and life-satisfaction in the initial years of this study in comparison to religious and non-religious skeptics?
-During the early years of New Zealand Attitudes and Values data collection New Zealand was emerging from the twin impacts of a global financial recession and a series of devastating earthquakes that destroyed 30% of Christchurch, New Zealand's second largest city [CITE].
-Orthogonal research observes that economic and ecological distress are associated with increases in traditional religious beliefs [CITE, CITE].
-Previous research finds an increase in traditional religious affiliation in the wake of the Christchurch earthquakes, with no immediate effect in subjective well-being.
-Experimental research has found among religious people death anxiety provokes acceptance of non-traditional religious beliefs [CITE].
+During the early years of New Zealand Attitudes and Values data collection New Zealand was emerging from the twin impacts of a global financial recession and a series of devastating earthquakes that destroyed 30% of Christchurch, New Zealand's second largest city [@greaves2015regional].
+Orthogonal research observes that economic and ecological distress are associated with increases in traditional religious beliefs [@sibley2012faith].
+Experimental research has found among religious people death anxiety provokes acceptance of non-traditional religious beliefs [@norenzayan2006belief].
 Because spiritual belief items were only introduced during the year of the earthquakes, we cannot here resolve whether the shifting relationship between spirit beliefs and well-being were the result of demand for spiritual beliefs among those secular and religious people who were adversely affected by the financial crisis or natural disaster.

### 2020-11-19 Manuscript.tex
include latex

--- 2020-11-18 Manuscript.tex
@@
   pdftitle={National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Beliefs},
-  pdfauthor={Ben Highland1, Everett L. Worthington2, Don E Davis3, Chris G. Sibley4, \& Joseph Bulbulia5},
+  pdfauthor={Ben Highland1, Everett L. Worthington2, Don E Davis3, Chris G. Sibley4, \& Joseph A. Bulbulia5},
   pdflang={en-EN},
@@
 
+% Overwrite redefinition of paragraph and subparagraph by the default LaTeX template
+% See https://github.com/crsh/papaja/issues/292
+\makeatletter
+\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
+  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
+  {-1em}%
+  {\normalfont\normalsize\bfseries\itshape\typesectitle}}
+
+\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
+  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
+  {-\z@\relax}%
+  {\normalfont\normalsize\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
+\makeatother
+
 % \usepackage{etoolbox}
@@
 \title{National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Beliefs}
-\author{Ben Highland\textsuperscript{1}, Everett L. Worthington\textsuperscript{2}, Don E Davis\textsuperscript{3}, Chris G. Sibley\textsuperscript{4}, \& Joseph Bulbulia\textsuperscript{5}}
+\author{Ben Highland\textsuperscript{1}, Everett L. Worthington\textsuperscript{2}, Don E Davis\textsuperscript{3}, Chris G. Sibley\textsuperscript{4}, \& Joseph A. Bulbulia\textsuperscript{5}}
 \date{}
@@
 The relationship between spirituality and psychological well-being is a matter of enduring fascination (Ellsworth \& Ellsworth, 2010).
-Most previous attempts to quantify the relationship between these two domains find a positive association: spirituality predicts well-being (Ginsburg, Quirt, Ginsburg, \& MacKillop, 1995; Koenig, 2010; Koenig, Associate Professor of Psychology and Religious Studies Michael E McCullough, PhD, McCullough, Larson, \& Adjunct Professor of Psychiatry David B Larson, 2001; Russinova, Wewiorski, \& Cash, 2002; Smith, McCullough, \& Poll, 2003; Smith et al., 2003; Zaza, Sellick, \& Hillier, 2005).
+Most previous attempts to quantify the relationship between these two domains find a positive association: spirituality predicts well-being (Ginsburg, Quirt, Ginsburg, \& MacKillop, 1995; Koenig, 2010; Koenig, Associate Professor of Psychology and Religious Studies Michael E McCullough, PhD, McCullough, Larson, \& Adjunct Professor of Psychiatry David B Larson, 2001; Russinova, Wewiorski, \& Cash, 2002; Smith, McCullough, \& Poll, 2003; Zaza, Sellick, \& Hillier, 2005).
 However, critics have raised three credible challenges.
@@
 Notably, researchers who have operationalised spirituality constructs as distinct from religion constructs have observed a negative association between spirituality and subjective well-being (King et al., 2013).
-Thus to infer how spirituality affects subjective well-being, it is important to disentangle the concepts of spirituality from concepts of religion, and to carefully investigate features of contructs in populations where these dimensions overlap, partially overlap, and diverge.
+Thus to infer how spirituality affects subjective well-being, it is important to disentangle the concepts of spirituality from concepts of religion, and to carefully investigate features of constructs in populations where these dimensions overlap, partially overlap, and diverge.
 
 A third challenge is to investigate the relationship between spirituality and subjective well-being in nationally diverse samples over individual life-spans.
-Reliable inference requires observing change within people who differ dimensions of spirituality over time, and assing how their lives turn out (Garssen \& Visser, 2016).
+Reliable inference requires observing change within people who differ dimensions of spirituality over time, and assessing how their lives turn out (Garssen \& Visser, 2016).
 Previous investigations of spirituality and well-being, including high-quality reviews and meta-analyses, have relied on cross-sectional samples (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky et al., 2005; Smith et al., 2003; Visser et al., 2010; Yonker et al., 2012).
@@
 
-Here, we addresses the three challenges of previous in the following ways: First, to avoid a tautological elision of \enquote{spirituality} and \enquote{well-being} we adopt a non-affective measure of spirituality as a \enquote{believe in a Spirit or Life Force.} On the other side, to measure subjective well-being, we use a well-validated scale for Personal Well-being and a well-validated scale for Life-Satisfaction (see Method).
+Here, we addresses the three challenges of previous in the following ways: First, to avoid a tautological elision of \enquote{spirituality} and \enquote{well-being} we adopt a non-affective measure of spirituality as a \enquote{believe in a Spirit or Life Force.} On the other side, to measure subjective well-being, we use a well-validated scale for personal well-being and a well-validated scale for Life-Satisfaction (see Method).
 Second, to disentangle spiritual beliefs from traditional religious beliefs and skeptical beliefs we adopt a four-level categorical indicator combining the binary indicator of belief in a Spirit or Life Force with the binary indicator \enquote{Do you believe in some form of a spirit or life force?} The levels of this four-level categorical indicator for spiritual, religious, and skeptical beliefs are as follows: (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who nevertheless believe in a Spirit or Life Force; (3) Those who believe in a God but do not believe in a Spirit or Life force; (4) Those who who believe in a both God and a Spirit or Life Force.
@@
 The Time 10 (2018) wave of the NZAVS included 47,951 respondents.
-We analyzed data from participants who responded to our survey at least three four times between Time 2 (2010) and Time 10 (2018), resulting in a sample of \(N\) = 21,705 (\(N\) = 20,979 with complete responses).
+We analyzed data from participants who responded to our survey at least three times between Time 2 (2010) and Time 10 (2018), resulting in a sample of \(N\) = 21,705 (\(N\) = 20,979 with complete responses).
 
@@
 
+We included standard demographic indicators used in New Zealand Attitudes and Values Research: \href{https://cdn.auckland.ac.nz/assets/psych/about/our-research/nzavs/NZAVSTechnicalDocuments/NZAVS-Technical-Documents-e11-Default-Statistical-Standards.pdf}{NZAVS modelling standards}
+
 \hypertarget{age}{%
@@
 
-We measured the socio-economic status of participants' immediate (small area) neighborhood using the 2013 New Zealand Deprivation Index, which uses aggregate census information about the residents of small neighborhood-type units to assign a decile-rank index from 1 (most affluent) to 10 (most impoverished) (Atkinson, Salmond, \& Crampton, 2014).
+We measured the socioeconomic status of participants' immediate (small area) neighborhood using the 2013 New Zealand Deprivation Index, which uses aggregate census information about the residents of small neighborhood-type units to assign a decile-rank index from 1 (most affluent) to 10 (most impoverished) (Atkinson, Salmond, \& Crampton, 2014).
 The index is based on a Principal Components Analysis of the following nine variables (in weighted order): the proportion of adults who received a means-tested benefit, household income, proportion not owning own home, proportion single-parent families, proportion unemployed, proportion lacking qualifications, proportion household crowding, proportion no telephone access, and proportion no car access.
@@
 Statistical analysis was performed using R version 4.0.2 R version 3.6.1 (2020-06-22) 2019-07-05).
-We analyze data from participants who responded to the NZAVS survey at least three four times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of 21,705 (N = 20,979 with complete responses).
+We analyze data from participants who responded to the NZAVS survey at least three times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of 21,705 (N = 20,979 with complete responses).
 To model growth in subjective well-being we use generalized linear mixed models.
@@
 Because missing values can result in biased estimates (Blackwell, Honaker, \& King, 2017), we multiply imputed 10 datasets using the AmeliaII package in R(Honaker et al., 2011b).We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences.
-Details of this analysis can be found at: \url{https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing}.
+Additionally the results were robust to filtering of participants of N \textgreater3 to N = 9 waves.
+Additionally, the findings reported below were robust to modeling only the effects of years and beliefs, without adjusting for the standard NZAVS demographic covariates.
+The supplement describing these analysis is located at: \url{https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing}.
 
@@
 \begin{figure}
-\includegraphics[width=6.4in]{Figs/USE_2_coefficientplots-1} \caption{Combination Coefficient Plots for Personal Well-being and Life Satisfaction Spirit Belief Longitudinal Models}\label{fig:unnamed-chunk-1}
+\centering
+\includegraphics{/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Figs/USE_2_coefficientplots-1.png}
+\caption{\label{fig:unnamed-chunk-1}Combination Coefficient Plots for personal well-being and life satisfaction longitudinal growth models}
 \end{figure}
@@
 
-\[
+{[}
+
 \begin{table}
-\centering 
-\caption{"Longitudinal Models"}\label{}
+\begin{center}
 \scalebox{0.6}{
@@
                                    & $(0.01)$               & $(0.01)$               \\
-Employed1                          & $\mathbf{0.09}^{***}$  & $\mathbf{0.08}^{***}$  \\
+EmployedEmployed                   & $\mathbf{0.09}^{***}$  & $\mathbf{0.08}^{***}$  \\
                                    & $(0.01)$               & $(0.01)$               \\
@@
 }
+\caption{Regression Models}
+\label{table:coefficients}
+\end{center}
 \end{table}
-\]
+
+{]}
 
@@
 
-We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict personal well-being (PWI) with Years (0-9), Beliefs (four categories), Age.10yrs.C, Deprivation.S, Edu.S, Employed, EthnicCats, Male, Partner, Pol.Orient.S and Urban, withId as random effects.
-The model included Id as random effects.
-The model's intercept, corresponding to Years = 0, Beliefs = Skeptic, Age = (sample mean), Deprivation.S = (sample mean), Edu.S =(sample mean), Employed = Not Employed, EthnicCats = Euro, Male = Not Mail, Partner = No Partner, Pol.Orient.S = (sample mean),, Urban = Not Urban and Id = 2, is at 4.94 (SE = 0.02, 95\% CI {[}4.89, 4.99{]}, p \textless{} .001) is at 6.63 (SE = 0.03, 95\% CI {[}6.57, 6.70{]}, p \textless{} .001).
+We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict personal well-being (PWI) with years (0-9), beliefs (four categories), Age.10yrs.C, deprivation (S), education (S), employed, ethnic categories, male, partner, political conservativism (S) and urban, with Id as a random effect.
+The model's intercept, corresponding to years = 0, beliefs = Skeptic, age = (sample mean), deprivation.S = (sample mean), education (S) =(sample mean), employed = not employed, ethnic Categories = European, male = not male, partner = no partner, political conservativism (S) = (sample mean), urban = not urban, is at 4.94 (SE = 0.02, 95\% CI {[}4.89, 4.99{]}, p \textless{} .001) is at 6.63 (SE = 0.03, 95\% CI {[}6.57, 6.70{]}, p \textless{} .001).
 Within this model:
@@
 beta = 3.19e-03, p = 0.458).
-We infer that Skeptics personal well-being is not increasing in the Skeptic population
-\item
-  The effect of Beliefs in SpiritExcludesGod is reliably negative (beta = -0.14, SE = 0.03, std.
+We infer that personal well-being did not increase in the Skeptic population between 2010 and 2018.
+\item
+  The effect of Beliefs in \emph{Spirit Excludes God} is reliably negative (beta = -0.14, SE = 0.03, std.
 beta = -0.03, p \textless{} .001).
-We infer that overall the population that believes in a Spirit but does not believe in a God is expected to have a lower personal well-being than skeptics.
-\item
-  The effect of Beliefs in GodAndSpirit is reliably negative (beta = -0.16, SE = 0.03, std.
+We infer that in 2010, the population that believed in a Spirit but did not believe in a God had lower personal well-being than did Skeptics.
+\item
+  The effect of Beliefs in \emph{God And Spirit} is reliably negative (beta = -0.16, SE = 0.03, std.
 beta = -0.02, p \textless{} .001).
-We infer that overall the population that believes in God and a spirit or life force is expected to have a lower Personal Well-being than skeptics.
-\item
-  The effect of Beliefs in GodExcludesSpirit is not reliably different from the skeptic population (beta = 0.02, SE = 0.05, std. beta = 0.02, p = 0.669).
-\item
-  The effect of Age.10yrs.C is reliably positive (beta = 0.08, SE = 7.38e-03, std. beta = 0.07, p \textless{} .001). Older people have greater personal well-beingwellbeing.
-\item
-  The effect of Deprivation (S) is reliably negative (beta = -0.14, SE = 7.50e-03, std. beta = -0.08, p \textless{} .001). Deprived people have lower personal well-beingwellbeing.
-\item
-  The effect of Education (S) is reliably positive (beta = 0.12, SE = 8.74e-03, std. beta = 0.07, p \textless{} .001). Educated people greater personal well-beingwellbeing
-\item
-  The effect of Employed is reliably positive (beta = 0.09, SE = 0.01, std. beta = 0.05, p \textless{} .001). Employed people have greater personal well-beingwellbeing.
-\item
-  The effect of identifying as Maori is reliably negative (beta = -0.30, SE = 0.03, std. beta = -0.18, p \textless{} .001).
+We infer that in 2010, the population that believed in a God and a Spirit or Life Force had lower Personal Well-being than did Skeptics.
+\item
+  The effect of Beliefs in \emph{God Excludes Spirit} is not reliable (beta = 0.02, SE = 0.05, std.
+beta = 0.02, p = 0.669).
+We infer that in 2010, the population that believed in a God and doubted a Spirit or Life Force similar levels of personal well-being to Skeptics.
+\item
+  The effect of Age.10yrs.C is reliably positive (beta = 0.08, SE = 7.38e-03, std. beta = 0.07, p \textless{} .001). Older people tend to have greater personal well-being.
+\item
+  The effect of Deprivation (S) is reliably negative (beta = -0.14, SE = 7.50e-03, std. beta = -0.08, p \textless{} .001). Deprived people tend to have lower personal well-being.
+\item
+  The effect of Education (S) is reliably positive (beta = 0.12, SE = 8.74e-03, std. beta = 0.07, p \textless{} .001). Educated people tend to have greater personal well-being.
+\item
+  The effect of Employed is reliably positive (beta = 0.09, SE = 0.01, std. beta = 0.05, p \textless{} .001).
+\item
+  The effect of identifying as M=\{a\}ori is reliably negative (beta = -0.30, SE = 0.03, std. beta = -0.18, p \textless{} .001).
 \item
@@
 \item
-  The effect of identifying as Male is reliably negative (beta = -0.11, SE = 0.02, std. beta = -0.06, p \textless{} .001).
-\item
-  The effect of having a Partner is reliably positive (beta = 0.61, SE = 0.02, std. beta = 0.37, p \textless{} .001).
-\item
-  The effect of Political Conservativism (S) is reliably positive (beta = 0.02, SE = 5.67e-03, std. beta = 0.01, p \textless{} .001).
-\item
-  The effect of living in an Urban area is reliably negative (beta = -0.02, SE = 0.01, std. beta = -0.01, p \textless{} .05).
+  The effect of identifying as male is reliably negative (beta = -0.11, SE = 0.02, std. beta = -0.06, p \textless{} .001).
+\item
+  The effect of having a partner is reliably positive (beta = 0.61, SE = 0.02, std. beta = 0.37, p \textless{} .001).
+\item
+  The effect of political conservativism (S) is reliably positive (beta = 0.02, SE = 5.67e-03, std. beta = 0.01, p \textless{} .001).
+\item
+  The effect of living in an Urban area is reliably negative (beta = -0.02, SE = 0.01, std. beta = -0.01, p \textless{} .05). T
 \end{itemize}
@@
 
-\begin{figure}
-\includegraphics[width=6.4in]{Figs/USEPWI_expected-1} \caption{Combination Coefficient Plots for Personal Well-being and Life Satisfaction Spirit Belief Longitudinal Models}\label{fig:unnamed-chunk-2}
-\end{figure}
-
+\includegraphics{Figs/USEPWI_expected-1.png}
 Focusing on the interaction of time with belief-type on personal well-being we observe:
@@
 \item
-  The effect of years \(\times\) belief in a SpiritExcludesGod\_ is reliably positive (beta = 0.02, SE = 4.86e-03, std.
+  The effect of years \(\times\) belief in a \emph{Spirit Excludes God} is reliably positive (beta = 0.02, SE = 4.86e-03, std.
 beta = 0.02, p \textless{} .01).
@@
 \item
-  The effect of years \(\times\) in GodAndSpirit is reliably positive (beta = 0.02, SE = 4.41e-03, std.
+  The effect of years \(\times\) in \emph{God And Spirit} is reliably positive (beta = 0.02, SE = 4.41e-03, std.
 beta = 0.03, p \textless{} .001).
@@
 \item
-  The effect of Years \(\times\) belief in GodExcludesSpirit is not reliable(beta = 1.96e-03, SE = 8.29e-03, std.
+  The effect of Years \(\times\) belief in \emph{God Excludes Spirit} is not reliable(beta = 1.96e-03, SE = 8.29e-03, std.
 beta = 2.36e-03, p = 0.813).
@@
 
-We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict Life Satisfaction (LIFESAT) with Years (0-9), Beliefs (four categories), Age.10yrs.C, Deprivation (S), Education (S), Employed, EthnicCats, Male, Partner, Pol.Orient.S and Urban, with Id as random effects.
+We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict personal well-being (PWI) with years (0-9), beliefs (four categories), Age.10yrs.C, deprivation (S), education (S), employed, ethnic categories, male, partner, political conservativism (S) and urban, with Id as a random effect.
 Standardized parameters were obtained by fitting the model on a standardized version of the dataset.
 The model's total explanatory power is substantial (conditional R2 = 0.69).
-The model's intercept, corresponding to Years = 0, Beliefs = Skeptic, Age = (sample mean),, Deprivation.S = (sample mean), Education =(sample mean), Employed = Not Employed, EthnicCats = Euro, Male = Not Mail, Partner = No Partner, Pol.Orient.S = (sample mean), Urban = Not Urban and Id = 2, is at 4.94 (SE = 0.02, 95\% CI {[}4.89, 4.99{]}, p \textless{} .001).
+The model's intercept, corresponding to years = 0, beliefs = Skeptic, age = (sample mean), deprivation.S = (sample mean), education (S) =(sample mean), employed = not employed, ethnic Categories = European, male = not male, partner = no partner, political conservativism (S) = (sample mean), urban = not urban is at 4.94 (SE = 0.02, 95\% CI {[}4.89, 4.99{]}, p \textless{} .001).
 Within this model:
@@
 \item
-  The effect of years is not reliable (beta = -5.50e-04, SE = 2.69e-03, std. beta = -9.32e-04, p = 0.838). We do not infer growth over time in Life Satisfaction among the Skeptic population
-\item
-  The effect of beliefs in SpiritExcludesGod is reliably negative (beta = -0.08, SE = 0.02, std.
+  The effect of years is not reliable (beta = -5.50e-04, SE = 2.69e-03, std.
+beta = -9.32e-04, p = 0.838).
+We infer that life satisfaction did not increase in the Skeptic population between 2010 and 2018.
+\item
+  The effect of beliefs in \emph{Spirit Excludes God} is reliably negative (beta = -0.08, SE = 0.02, std.
 beta = 0.01, p \textless{} .001).
@@
 \item
-  The effect of beliefs in GodAndSpirit is not reliable (beta = -0.04, SE = 0.02, std. beta = 0.04, p = 0.105). We do not infer growth over time in Life Satisfaction among this (Skeptic) population
-\item
-  The effect of beliefs in GodExcludesSpirit is reliably positive (beta = 0.07, SE = 0.04, std.
+  The effect of beliefs in \emph{God And Spirit} is not reliable (beta = -0.04, SE = 0.02, std.
+beta = 0.04, p = 0.105).
+People who believe in God but doubt a Spirit/Life force are expected to have had similar levels of life satisfaction in Wave 10 as skeptics.
+\item
+  The effect of beliefs in \emph{God Excludes Spirit} is reliably positive (beta = 0.07, SE = 0.04, std.
 beta = 0.05, p = 0.062).We infer that overall the population that believes in a God but does not believe in a spirit is not reliably different from the Skeptics population in Life Satisfaction.
@@
 \item
-  The effect of identifying as Male is reliably negative (beta = -0.19, SE = 0.01, std. beta = -0.16, p \textless{} .001).
-\item
-  The effect of having a Parnter is reliably positive (beta = 0.42, SE = 0.01, std. beta = 0.35, p \textless{} .001).
-\item
-  The effect of Political Conservativism (S) is reliably positive (beta = 0.01, SE = 4.23e-03, std. beta = 0.01, p \textless{} .01).
-\item
-  The effect of living in an Urban area is reliably negative (beta = -0.02, SE = 8.85e-03, std. beta = -0.02, p \textless{} .05).
+  The effect of identifying as male is reliably negative (beta = -0.19, SE = 0.01, std. beta = -0.16, p \textless{} .001).
+\item
+  The effect of having a partner is reliably positive (beta = 0.42, SE = 0.01, std. beta = 0.35, p \textless{} .001).
+\item
+  The effect of political conservativism (S) is reliably positive (beta = 0.01, SE = 4.23e-03, std. beta = 0.01, p \textless{} .01).
+\item
+  The effect of living in an urban area is reliably negative (beta = -0.02, SE = 8.85e-03, std. beta = -0.02, p \textless{} .05).
 \end{itemize}
@@
 
-\begin{figure}
-\includegraphics[width=6.4in]{Figs/USElifsat_expected-1} \caption{Prediction Plot Life Satisfaction Spirit Belief Longitudinal Models}\label{fig:unnamed-chunk-3}
-\end{figure}
-
-\begin{itemize}
-\tightlist
-\item
-  The interaction of years \(\times\) beliefs in SpiritExcludesGod is reliably positive (beta = 0.02, SE = 3.66e-03, std.
+\includegraphics{/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Figs/USElifsat_expected-1.png}
+- The interaction of years \(\times\) beliefs in \emph{Spirit Excludes God} ludesGod is reliably positive (beta = 0.02, SE = 3.66e-03, std.
 beta = 0.03, p \textless{} .001).
 We infer that, unlike the Skeptic population, people with beliefs in a Spirit or Life Force who do not believe in a God tend to grow in their Life Satisfaction over time.
-
-  \begin{itemize}
-  \tightlist
-  \item
-    The interaction of years \(\times\) beliefsGodAndSpirit is reliably positive (beta = 0.01, SE = 3.32e-03, std.
+- The interaction of years \(\times\) \emph{God And Spirit} is reliably positive (beta = 0.01, SE = 3.32e-03, std.
 beta = 0.02, p \textless{} .001).
-We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their Life Satisfaction over time
-  \item
-    The effect of years \(\times\) beliefs in GodExcludesSpirit is not reliable (beta = -1.59e-03, SE = 6.25e-03, std.
+We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their life satisfaction over time
+- The effect of years \(\times\) beliefs in \emph{God Excludes Spirit} is not reliable (beta = -1.59e-03, SE = 6.25e-03, std.
 beta = -2.70e-03, p = 0.799).
 Thus, God believers who doubt in a Spirit or Life Force start maintain higher Life-Satisfaction than do other groups, but they are not expected to grow in Life-Satisfaction over time.
-  \end{itemize}
-\end{itemize}
 
@@
 \begin{figure}
-\includegraphics[width=6.4in]{Figs/coefficient_plot_both_imputed-1} \caption{Imputed data analysis: coefficient plot}\label{fig:unnamed-chunk-4}
+\centering
+\includegraphics{Figs/coefficient_plot_both_imputed-1.png}
+\caption{\label{fig:unnamed-chunk-4}Imputed data analysis: coefficient plot}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=6.4in]{Figs/predicted_PWI_impute-1} \caption{Imputed data analysis: predicted personal wellbeing by years X beliefs}\label{fig:unnamed-chunk-5}
+\centering
+\includegraphics{/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Figs/predicted_PWI_impute-1.png}
+\caption{\label{fig:unnamed-chunk-5}Imputed data analysis: predicted personal wellbeing by years X beliefs}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=6.4in]{Figs/predicted_lifesat_imputed-1} \caption{Imputed data analysis: predicted personal wellbeing by years X beliefs}\label{fig:unnamed-chunk-6}
+\centering
+\includegraphics{/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Figs/predicted_lifesat_imputed-1.png}
+\caption{\label{fig:unnamed-chunk-6}Imputed data analysis: predicted personal wellbeing by years X beliefs}
 \end{figure}
@@
 \item
-  At the outset of the study, in 2010, the expected subjective well-being as measured using the Personal Well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God (\(\hat{y}\) = 6.50, se =.03) and lower among those who believed in both a Spirit/Life-Force and a God (\(\hat{y}\) = 6.48, se =.03) in comparison both to the skeptical population who neither believed in a God nor a life force (\(\hat{y}\) = 6.64,se =.03) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force (\(\hat{y}\) 6.66, se =.05).
-(See Figure 2) Similarly,the expected subjective well-being as measured using the Life Satisfaction Inventory is lower among those who believed in Spirit or Life Force but not a God (\(\hat{y}\) = 4.86, se =.02) and lower among those who believed in both a Spirit/Life-Force and a God (\(\hat{y}\) = 4.91, se =.02) in comparison both with the skeptical population who neither believed in a God nor a life force (\(\hat{y}\) = 4.94,se =.02) and the population that believed in a God but was skeptical about belief in a Spirit or Life Force (\(\hat{y}\) 5.02, se =.02) (see Figure 3)
+  At the outset of the study, in 2010, the expected subjective well-being as measured using the personal well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God (\(\hat{y}\) = 6.50, se =.03) and lower among those who believed in both a Spirit/Life-Force and a God (\(\hat{y}\) = 6.48, se =.03) in comparison both to the skeptical population who neither believed in a God nor a life force (\(\hat{y}\) = 6.64,se =.03) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force (\(\hat{y}\) 6.66, se =.05).
+(See Figure 2) Similarly,the expected subjective well-being as measured using the life satisfaction Inventory is lower among those who believed in Spirit or Life Force but not a God (\(\hat{y}\) = 4.86, se =.02) and lower among those who believed in both a Spirit/Life-Force and a God (\(\hat{y}\) = 4.91, se =.02) in comparison both with the skeptical population who neither believed in a God nor a life force (\(\hat{y}\) = 4.94,se =.02) and the population that believed in a God but was skeptical about belief in a Spirit or Life Force (\(\hat{y}\) 5.02, se =.02) (see Figure 3)
 \item
@@
 In 2018, personal well-being among who believe in Spirit or Life Force but not a God had grown to (\(\hat{y}\) = 6.67, se =.03, \(\Delta \hat{y} = 0.17\)).
-Among those who believed in both a Spirit/Life-Force and a God Personal Well-Being had grown to (\(\hat{y}\) = 6.69, se = 0.02,\(\Delta \hat{y} = .21\)).
+Among those who believed in both a Spirit/Life-Force and a God, personal well-being had grown to (\(\hat{y} = 6.69, se = 0.02, \Delta \hat{y} = .21\)).
 To place this effect in practical perspective, such growth in personal well-being is about twice as large as the expected benefit of being employed and about a third as large as the expected benefit from having a romantic partner.
@@
 The first question is why those who believed in Spirit or Life Force were initially lower in their personal well-being and life-satisfaction in the initial years of this study in comparison to religious and non-religious skeptics?
-During the early years of New Zealand Attitudes and Values data collection New Zealand was emerging from the twin impacts of a global financial recession and a series of devastating earthquakes that destroyed 30\% of Christchurch, New Zealand's second largest city {[}CITE{]}.
-Orthogonal research observes that economic and ecological distress are associated with increases in traditional religious beliefs {[}CITE, CITE{]}.
-Previous research finds an increase in traditional religious affiliation in the wake of the Christchurch earthquakes, with no immediate effect in subjective well-being.
-Experimental research has found among religious people death anxiety provokes acceptance of non-traditional religious beliefs {[}cite{]}.
+During the early years of New Zealand Attitudes and Values data collection New Zealand was emerging from the twin impacts of a global financial recession and a series of devastating earthquakes that destroyed 30\% of Christchurch, New Zealand's second largest city (Greaves et al., 2015).
+Orthogonal research observes that economic and ecological distress are associated with increases in traditional religious beliefs (Sibley \& Bulbulia, 2012).
+Experimental research has found among religious people death anxiety provokes acceptance of non-traditional religious beliefs (Norenzayan \& Hansen, 2006).
 Because spiritual belief items were only introduced during the year of the earthquakes, we cannot here resolve whether the shifting relationship between spirit beliefs and well-being were the result of demand for spiritual beliefs among those secular and religious people who were adversely affected by the financial crisis or natural disaster.
@@
 Despite its limitations, our study is important both for its methods and its findings.
-We demonstrate that spirituality can be operationalised to avoid the pitfalls of tautology evident in previous research by focusing on states of beliefs, that concepts of such belief states can be refined to enable distinctions necessary to investigate spirituality in mixed secular and religious populations, and that the question of whether beliefs states predict growth in subjective well-being can be investigated in a longitudional setting.
+We demonstrate that spirituality can be operationalised to avoid the pitfalls of tautology evident in previous research by focusing on states of beliefs, that concepts of such belief states can be refined to enable distinctions necessary to investigate spirituality in mixed secular and religious populations, and that the question of whether beliefs states predict growth in subjective well-being can be investigated in a longitudinal setting.
 Our finding that, indeed, spiritual beliefs predict growth in perceived well-being and life satisfaction may hold practical importance.
-Previous research indicates that religious beliefs are associated with having a rewarding life.
-Western populations are increasingly identifying as \enquote{spiritual but not religious} (Swinton, 2001; Zinnbauer \& Pargament, 2005), raising the specter that the psychological from traditional might be receding.
+Previous research indicates that religious beliefs are associated with psychological well-being (Smith et al., 2003).
+Western populations are increasingly identifying as \enquote{spiritual but not religious} (Swinton, 2001; Zinnbauer \& Pargament, 2005).
 Here, we observe that even among people who lack a traditional religious belief in God, believing in a Spirit or Life force predicts incremental improvements in well-being that sum to practically large effects.
@@
 
-BH conceived of the study, wrote the first draft of the manuscript, and performed the initial data analysis with input from JB.
-EW developed the study with BH and JB and revised the manuscript.
+BH conceived of the study.
+BH wrote the first draft of the manuscript, and performed the initial data analysis with input from JB.
+EW improved the approach with BH and JB and revised the manuscript.
 DD revised the manuscript.
 CS was responsible for NZAVS data collection and revised the manuscript.
-JB developed the study with BH and EW, performed the data re-analysis, and revised the manuscript.
+JB improved the approach with BH and EW, performed a data re-analysis, and revised the manuscript.
 
@@
 \leavevmode\hypertarget{ref-Blackwell2017-oq}{}%
-Blackwell, M., Honaker, J., \& King, G. (2017). A unified approach to measurement error and missing data: Details and extensions. \emph{Sociological Methods \& Research}.
+Blackwell, M., Honaker, J., \& King, G. (2017). A unified approach to measurement error and missing data: Details and extensions.
 
@@
 \leavevmode\hypertarget{ref-Cummins2017-ur}{}%
-Cummins, R. A., Eckerseley, R., Pallant, J., Van Vugt, J., \& Misajon, R. (2017). Australian unity wellbeing index. \emph{PsycTESTS Dataset}.
+Cummins, R. A., Eckerseley, R., Pallant, J., Van Vugt, J., \& Misajon, R. (2017). Australian unity wellbeing index.
 
 \leavevmode\hypertarget{ref-Diener1985-xy}{}%
-Diener, E., Emmons, R. A., Larsen, R. J., \& Griffin, S. (1985). The satisfaction with life scale. \emph{Journal of Personality Assessment}.
+Diener, E., Emmons, R. A., Larsen, R. J., \& Griffin, S. (1985). The satisfaction with life scale.
 
@@
 
+\leavevmode\hypertarget{ref-greaves2015regional}{}%
+Greaves, L.
+M., Milojev, P., Huang, Y., Stronge, S., Osborne, D., Bulbulia, J., \ldots{} Sibley, C.
+G.
+(2015).
+Regional differences in the psychological recovery of christchurch residents following the 2010/2011 earthquakes: A longitudinal study.
+\emph{PLoS One}, \emph{10}(5), e0124278.
+
 \leavevmode\hypertarget{ref-Hackney2003-rs}{}%
@@
 \leavevmode\hypertarget{ref-King2013-cg}{}%
-King, M., Marston, L., McManus, S., Brugha, T., Meltzer, H., \& Bebbington, P.
-(2013).
-Religion, spirituality and mental health: Results from a national study of english households.
-\emph{British Journal of Psychiatry}.
+King, M., Marston, L., McManus, S., Brugha, T., Meltzer, H., \& Bebbington, P. (2013). Religion, spirituality and mental health: Results from a national study of english households.
 
@@
 
+\leavevmode\hypertarget{ref-norenzayan2006belief}{}%
+Norenzayan, A., \& Hansen, I. G. (2006). Belief in supernatural agents in the face of death. \emph{Personality and Social Psychology Bulletin}, \emph{32}(2), 174--187.
+
 \leavevmode\hypertarget{ref-R-patchwork}{}%
@@
 Sawatzky, R., Ratner, P. A., \& Chiu, L. (2005). A Meta-Analysis of the relationship between spirituality and quality of life. \emph{Soc. Indic. Res.}, \emph{72}(2), 153--188.
+
+\leavevmode\hypertarget{ref-sibley2012faith}{}%
+Sibley, C.
+G., \& Bulbulia, J.
+(2012).
+Faith after an earthquake: A longitudinal study of religion and perceived health before and after the 2011 christchurch new zealand earthquake.
+\emph{PloS One}, \emph{7}(12), e49648.
 
@@
 \leavevmode\hypertarget{ref-Zaza2005-ac}{}%
-Zaza, C., Sellick, S. M., \& Hillier, L. M. (2005). Coping with cancer. \emph{Journal of Psychosocial Oncology}.
+Zaza, C., Sellick, S. M., \& Hillier, L. M. (2005). Coping with cancer.
 

### 2020-11-19 Analysis.Rmd
committing figures

@@
 
-# Inference holds but VIFS for years x beliefs
+Inference holds but VIFS years x beliefs when years are centered.
 
-```{r model centering years, include = FALSE}
+```{r model centering years}
 library(lme4)
@@
 mod.2a <-lmer (LIFESAT ~ 1 + yearsC * Beliefs+ Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id), data= dfben, na.action = na.omit)
-summary(mod.1)
-summary(mod.2)
+summary(mod.1a)
+summary(mod.2a)
 # identical result
@@
 model_parameters(m)
+```
+
+Model is robust without covariates
+
+```{r robust to years and beliefs only}
+library(lme4)
+mod.1aa <-lmer (PWI ~  1 +  years * Beliefs + (1|Id), data= dfben, na.action = na.omit)
+mod.2aa <-lmer (LIFESAT ~ 1 + years * Beliefs + (1|Id), data= dfben, na.action = na.omit)
+)
+# identical result
+sjPlot::tab_model(mod.1aa,mod.2aa)
 ```

### 2020-11-19 Analysis.Rmd
added analysis with centred years for the imputed data set. Inference is the same

@@
 #imputed<- readRDS("imputed")
+```
+```{r}
 
@@
                               Age.10yrs = (Age/10),
+                              yearsC = scale(Years, center=TRUE,scale=FALSE),
                               Pol.Orient.S =scale(Pol.Orient, center=TRUE,scale=TRUE),
@@
 
+## analysis with centred time (to avoid multicollinearity)
+
+```{r cache = FALSE}
+# run models iterating over imputed data
+m <- 10
+models33 <- NULL
+for(i in 1:m) {
+  models33[[i]] <- lmer(PWI ~ yearsC * Beliefs + Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id), data=imputed3$imputations[[i]])
+}
+```
+
+```{r, cache=FALSE, dependson= models33}
+#devtools::install_github("easystats/parameters") Get development version
+library(parameters)
+mps33 <- lapply(models33, model_parameters)
+parameters::pool_parameters(mps33)
+mp33<-pool_parameters(mps33)
+
+```
+
+```{r coef_PWI_imputed_C}
+# coeff plot
+plot(mp33) + ggtitle("Personal Wellbeing Coefficients (Imputed)") # pick your  own title
+```
+
+```{r cache = FALSE, dependson = models3}
+## expectation
+m<-10
+out<-NULL
+for(i in 1:m) {
+  out[[i]] <- ggpredict(models33[[i]], terms =c("yearsC [minmax]","Beliefs"))
+}
+
+m<-10
+plots<-NULL
+for(i in 1:m) {
+  plots[[i]] <- plot(out[[i]], facets = T)# + scale_y_continuous(limits=c(6.35,6.85) )
+}
+```
+```{r, echo=FALSE}
+# Inspect all  # No differences
+# plots[[1]]
+# plots[[2]]
+# plots[[3]]
+# plots[[4]]
+# plots[[5]]
+# plots[[6]]
+# plots[[7]]
+# plots[[8]]
+# plots[[9]]
+# plots[[10]]
+```
+```{r predicted_PWI_impute_C}
+# to highlight differencs
+explectationGraphPWI_C <-plots[[10]] + gghighlight() + ggtitle("Predicted Values of Personal Wellbeing (imputed)")
+explectationGraphPWI_C
+```
+```{r cache =TRUE}
+## life sat models
+m <- 10
+modelsL33 <- NULL
+
+for(i in 1:m) {
+  modelsL33[[i]] <- lmer(LIFESAT ~ yearsC * Beliefs + Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban +  (1|Id), data=imputed3$imputations[[i]])
+}
+```
+```{r, cache =FALSE, dependson =  modelsL33}
+mpl33 <- lapply(modelsL33, model_parameters)
+parameters::pool_parameters(modelsL33)
+ml33<-pool_parameters(mpl33)
+ml33 # table
+p133<-plot(ml3)
+
+p133<- p133 +  labs(title ="Coefficients Life Satisafaction (imputed)")
+# get estimates for PWI from above
+p233 <- plot(mp133) + labs(title ="Coefficients Personal Wellbeing (imputed)")
+```
+
+```{r coefficient_plot_both_imputed_C}
+ p233/p133 +  plot_annotation(tag_levels = 'i')
+```
+
+```{r, cache =FALSE, dependson = "modelsL3"}
+# Expectation plots
+m<-10
+out2<-NULL
+for(i in 1:m) {
+  out2[[i]] <- ggpredict(modelsL33[[i]], terms = c("yearsC [minmax]", "Beliefs"),
+                         ci.lvl = 0.95,
+                         type = "fe",
+                         typical = "mean",
+                         #   condition = c(Euro =1),
+                         back.transform = TRUE,
+                         ppd = FALSE,
+                         interval = "confidence")
+}
+
+m<-10
+plots2<-NULL
+for(i in 1:m) {
+  plots2[[i]] <- plot(out2[[i]], facets = T) #+ scale_y_continuous(limits=c(4.75,5.25) )
+}
+```
+```{r echo = FALSE}
+# Inspect all  # No differences
+# plots2[[1]]
+# plots2[[2]]
+# plots2[[3]]
+# plots2[[4]]
+# plots2[[5]]
+# plots2[[6]]
+# plots2[[7]]
+# plots2[[8]]
+# plots2[[9]]
+# plots2[[10]]
+
+library(gghighlight)
+```
+
+```{r explectationGraphLS_C}
+# Just using one graph
+explectationGraphLS_C <-plots2[[10]] + gghighlight() + ggtitle("Predicted Values of Life Satisfaction (imputed)")
+```
+
 # Discussion

### 2020-11-20 Manuscript.Rmd

@@
 author: 
-  - name          : "Ben Highland"
+  - name          : "Benjamin Highland"
     affiliation   : "1"
     corresponding : yes    # Define only one corresponding author
-    address       : "Ben's Postal address"
+    address       : "520 Power Plant Circle Apartment 105 Winston-Salem, NC 27101 USA"
     email         : "benhighland6420@gmail.com"
@@
 authornote: |
-  1 Wake Forest Medical School 2 Department of Psychology Virginia Commonwealth University 3 Department of Psychology Georgia State University 4 School of Psychology University of Auckland 5 School of Psychology Victoria University
+  1 Wake Forest School of Medicine 2 Department of Psychology Virginia Commonwealth University 3 Department of Psychology Georgia State University 4 School of Psychology University of Auckland 5 School of Psychology Victoria University
 
@@
 
-<!-- In spirituality/psychological wellbeing research,subjective in spirituality health research, psychological well-being is typically decomposed into two categories: emotional well-being (i.e., the presence or absence of positive feelings about life) and functional well-being (i.e., psychological and social well-being) [@Keyes2002-bb].
+<!-- In spirituality/psychological wellbeing research, subjective in spirituality health research, psychological well-being is typically decomposed into two categories: emotional well-being (i.e., the presence or absence of positive feelings about life) and functional well-being (i.e., psychological and social well-being) [@Keyes2002-bb].
 -->

### 2020-11-20 Analysis.Rmd
final second submit manuscript.

@@
 ml33 # table
-p133<-plot(ml3)
+p133<-plot(ml33)
 
@@
 # get estimates for PWI from above
-p233 <- plot(mp133) + labs(title ="Coefficients Personal Wellbeing (imputed)")
+p233 <- plot(ml33) + labs(title ="Coefficients Personal Wellbeing (imputed)")
 ```
@@
 explectationGraphLS_C <-plots2[[10]] + gghighlight() + ggtitle("Predicted Values of Life Satisfaction (imputed)")
+explectationGraphLS_C
 ```

### 2020-11-20 Manuscript.Rmd
final second submit manuscript.

@@
 keywords          : "Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being"
-wordcount         : "5269 words in text body, 1078 words in reference section"
+wordcount         : "5415 words in text body, 1155 words in reference section"
 bibliography      : ["b-references.bib"]
@@
 \[
-
 \begin{table}

### 2020-11-20 Manuscript.tex
final second submit manuscript.

@@
   pdftitle={National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Beliefs},
-  pdfauthor={Ben Highland1, Everett L. Worthington2, Don E Davis3, Chris G. Sibley4, \& Joseph A. Bulbulia5},
+  pdfauthor={Benjamin Highland1, Everett L. Worthington2, Don E Davis3, Chris G. Sibley4, \& Joseph A. Bulbulia5},
   pdflang={en-EN},
@@
 \shorttitle{Longitudinal Wellbeing from Spiritual Beliefs}
-\keywords{Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being\newline\indent Word count: 5269 words in text body, 1078 words in reference section}
+\keywords{Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being\newline\indent Word count: 5415 words in text body, 1155 words in reference section}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 \title{National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Beliefs}
-\author{Ben Highland\textsuperscript{1}, Everett L. Worthington\textsuperscript{2}, Don E Davis\textsuperscript{3}, Chris G. Sibley\textsuperscript{4}, \& Joseph A. Bulbulia\textsuperscript{5}}
+\author{Benjamin Highland\textsuperscript{1}, Everett L. Worthington\textsuperscript{2}, Don E Davis\textsuperscript{3}, Chris G. Sibley\textsuperscript{4}, \& Joseph A. Bulbulia\textsuperscript{5}}
 \date{}
@@
 
-1 Wake Forest Medical School 2 Department of Psychology Virginia Commonwealth University 3 Department of Psychology Georgia State University 4 School of Psychology University of Auckland 5 School of Psychology Victoria University
-
-Correspondence concerning this article should be addressed to Ben Highland, Ben's Postal address. E-mail: \href{mailto:benhighland6420@gmail.com}{\nolinkurl{benhighland6420@gmail.com}}
+1 Wake Forest School of Medicine 2 Department of Psychology Virginia Commonwealth University 3 Department of Psychology Georgia State University 4 School of Psychology University of Auckland 5 School of Psychology Victoria University
+
+Correspondence concerning this article should be addressed to Benjamin Highland, 520 Power Plant Circle Apartment 105 Winston-Salem, NC 27101 USA.
+E-mail: \href{mailto:benhighland6420@gmail.com}{\nolinkurl{benhighland6420@gmail.com}}
 
@@
 
-{[}
-
+\[
 \begin{table}
@@
 \end{table}
-
-{]}
+\]
 

### 2020-11-21 Analysis.Rmd
small changes

--- 2020-11-19 Analysis.Rmd
@@
 library("patchwork")
-library("easystats")
 library("parameters")
@@
 library("sjPlot")
-library("texreg")
 library("see")
@@
 library("citr")
-
+library("texreg")
 ```
@@
   dplyr::ungroup(Id)%>%
-  dplyr:: mutate(Years = as.numeric(years))%>%
-  dplyr:: mutate(yearsC = scale(years, center=TRUE,scale=FALSE))%>%
+  dplyr::mutate(Years = as.numeric(years))%>%
   dplyr::mutate(Age.10yrs = (Age/10))%>%
@@
          Religion.Church.Log.C = scale(Religion.Church.Log, center = TRUE, scale = FALSE))
-```
-```{r}
+
 # nice labels
@@
 table(dfben$Beliefs)
-range(dfben$years)
-range(dfben$Years)
 ```
@@
 
-```{r}
-table1::table1(~ Beliefs|Wave, data = dfben, overall = T)
+```{r, include =FALSE}
+furniture::table1(dfben, Age, NZdep,Education,Employed,EthnicCategories,Male,Partner, Pol.Orient,Urban, Beliefs, LIFESAT,PWI,
+                    splitby = ~Wave, overall=F, output = "latex2", booktabs=TRUE)
 ```
 
-```{r, include =FALSE}
-#furniture::table1(dfben, Age, NZdep,Education,Employed,EthnicCategories,Male,Partner, Pol.Orient,Urban, Beliefs, LIFESAT,PWI,
-                  #  splitby = ~Wave, overall=F, output = "latex2", booktabs=TRUE)
+```{r, include=FALSE}
+r<-report::report_participants(
+  dfben,
+  age = "Age",
+  sex = "Male",
+  education = "Education",
+  participants = "Id",
+  group = "Wave",
+  spell_n = FALSE,
+)
 ```
-
-<!-- ```{r, include=FALSE} -->
-<!-- r<-easystats::report_participants( -->
-<!--   dfben, -->
-<!--   age = "Age", -->
-<!--   sex = "Male", -->
-<!--   education = "Education", -->
-<!--   participants = "Id", -->
-<!--   group = "Wave", -->
-<!--   spell_n = FALSE, -->
-<!-- ) -->
-<!-- ??report_participants -->
-<!-- ``` -->
 
@@
 # mod.2 <-glmmTMB(PWI ~ Years * Beliefs + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id) + (1|GenCohort), data= dfben)
-```
-
-```{r check collinearity}
-# Note High VIFS from 
-performance::check_collinearity(mod.1)
-performance::check_collinearity(mod.2)
-```
-
-Inference holds but VIFS years x beliefs when years are centered.
-
-```{r model centering years}
-library(lme4)
-mod.1a <-lmer (PWI ~  1 +  yearsC * Beliefs +  Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id), data= dfben, na.action = na.omit)
-mod.2a <-lmer (LIFESAT ~ 1 + yearsC * Beliefs+ Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id), data= dfben, na.action = na.omit)
-summary(mod.1a)
-summary(mod.2a)
-# identical result
-performance::check_collinearity(mod.1a)
-performance::check_collinearity(mod.2a)
-
-# address reviewer worry about deprivation and partner being correlated
-m<-lmer (NZdep ~ Partner + (1|Id), data= dfben, na.action = na.omit)
-tab_model(m)
-model_parameters(m)
-```
-
-Model is robust without covariates
-
-```{r robust to years and beliefs only}
-library(lme4)
-mod.1aa <-lmer (PWI ~  1 +  years * Beliefs + (1|Id), data= dfben, na.action = na.omit)
-mod.2aa <-lmer (LIFESAT ~ 1 + years * Beliefs + (1|Id), data= dfben, na.action = na.omit)
-)
-# identical result
-sjPlot::tab_model(mod.1aa,mod.2aa)
 ```
@@
 g1/g2
-range(dfben$yearsC)
 ```
@@
   ggtitle("Predicted Values of Personal Wellbeing")
-p1
+
 pr.PWI
@@
 
-<!-- ```{r eq1} -->
-<!-- library(equatiomatic) -->
-<!-- print(extract_eq(mod.1, wrap = TRUE,terms_per_line =3)) -->
-<!-- ``` -->
+```{r eq1}
+library(equatiomatic)
+print(extract_eq(mod.1, wrap = TRUE,terms_per_line =3))
+```
 
-<!-- PWI Model 2 equation -->
-<!-- ```{r eq2} -->
-<!-- library(equatiomatic) -->
-<!-- print(extract_eq(mod.2, wrap = TRUE, use_coefs=TRUE,terms_per_line = 3)) -->
-<!-- ``` -->
+PWI Model 2 equation
+```{r eq2}
+library(equatiomatic)
+print(extract_eq(mod.2, wrap = TRUE, use_coefs=TRUE,terms_per_line = 3))
+```
 
-<!-- ## report -->
+## report
 
-<!-- ### Model 1 -->
-<!-- model 1 report: for publication change all "significants" to "reliable" -->
+### Model 1
+model 1 report: for publication change all "significants" to "reliable"
 
-<!-- short report: -->
-<!-- ```{r, report1, cache =TRUE, dependson=mod.1,include=FALSE} -->
-<!-- easystats::model_text(mod.1) -->
-<!-- ``` -->
+short report:
+```{r, report1, cache =TRUE, dependson=mod.1,include=FALSE}
+report::model_text(mod.1)
+```
 
-<!-- short report -->
-<!-- ```{r, cache=TRUE, include=FALSE } -->
-<!-- library(report) -->
-<!-- r1<- easystats::report(mod.1) -->
-<!-- easystats::table_short(r1) -->
-<!-- easystats::text_short(r1) -->
-<!-- ``` -->
+short report
+```{r, cache=TRUE, include=FALSE }
+library(report)
+r1<- report::report(mod.1)
+table_short(r1)
+text_short(r1)
+```
 
-<!-- ```{r, cache =TRUE, dependson=mod.1, include=FALSE} -->
-<!-- r1<- easystats::report(mod.1) -->
-<!-- easystats::text_long(r1) -->
-<!-- ``` -->
-<!-- short report -->
-<!-- ```{r, cache =TRUE, dependson=mod.2,include=FALSE} -->
-<!-- # Model 2 -->
-<!-- # For publication change all "significants" to "reliable" -->
+```{r, cache =TRUE, dependson=mod.1, include=FALSE}
+r1<- report::report(mod.1)
+report::text_long(r1)
+```
+short report
+```{r, cache =TRUE, dependson=mod.2,include=FALSE}
+# Model 2
+# For publication change all "significants" to "reliable"
 
-<!-- easystats::model_text(mod.2) -->
-<!-- ``` -->
+report::model_text(mod.2)
+```
 
-<!-- short report -->
-<!-- ```{r,include=FALSE} -->
-<!-- r2 <- easystats::report(mod.2) -->
-<!-- easystats::text_short(r2) -->
-<!-- easystats::table_short(r2) -->
-<!-- ``` -->
+short report
+```{r,include=FALSE}
+r2 <- report::report(mod.2)
+report::text_short(r2)
+report::table_short(r2)
+```
 
-<!-- long report -->
-<!-- ```{r,include=FALSE} -->
-<!-- r2<- easystats::report(mod.2) -->
-<!-- easystats::text_long(r2) -->
-<!-- ``` -->
+long report
+```{r,include=FALSE}
+r2<- report::report(mod.2)
+report::text_long(r2)
+```
 
@@
 ##
+sessionInfo()
 ```

### 2020-11-21 Analysis.Rmd
added bayesian estimation

--- 2020-11-20 Analysis.Rmd
@@
 library("patchwork")
-library("easystats")
+#library("easystats")
 library("parameters")
@@
 #library("here")
+library(parameters) # great tables, plots
 library("citr")
+library("brms")
+library("rstan")
+rstan_options(auto_write=TRUE)
+options(mc.cores=parallel::detectCores ())
+#theme_set(theme_few())
 
@@
 table(dfben$Beliefs)
+# years
 range(dfben$years)
-range(dfben$Years)
 ```
@@
 
-Inference holds but VIFS years x beliefs when years are centered.
+VIF high,  this is the years x factor interaction. See if it makes a difference by centering years.
 
@@
 library(lme4)
-mod.1a <-lmer (PWI ~  1 +  yearsC * Beliefs +  Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id), data= dfben, na.action = na.omit)
-mod.2a <-lmer (LIFESAT ~ 1 + yearsC * Beliefs+ Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id), data= dfben, na.action = na.omit)
+mod.1a <-lmer (PWI ~  1 +  yearsC * Beliefs +  Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id), data= dfben)
+mod.2a <-lmer (LIFESAT ~ 1 + yearsC * Beliefs+ Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id), data= dfben)
 summary(mod.1a)
@@
 performance::check_collinearity(mod.2a)
+```
+Centering years makes no difference to inference.
 
-# address reviewer worry about deprivation and partner being correlated
+Check reviewers interest in correlation of deprivation and having a partner.
+```{r}
 m<-lmer (NZdep ~ Partner + (1|Id), data= dfben, na.action = na.omit)
@@
 
-Model is robust without covariates
+Check whether model is robust without control covariates
 
@@
 library(lme4)
-mod.1aa <-lmer (PWI ~  1 +  years * Beliefs + (1|Id), data= dfben, na.action = na.omit)
-mod.2aa <-lmer (LIFESAT ~ 1 + years * Beliefs + (1|Id), data= dfben, na.action = na.omit)
+mod.1aa <-lmer (PWI ~  1 +  yearsC * Beliefs + (1|Id), data= dfben)
+mod.2aa <-lmer (LIFESAT ~ 1 + yearsC * Beliefs + (1|Id), data= dfben)
 )
@@
 
+Indeed, models are robust without control covariates
 ```{r cache = TRUE, dependson="mod.2", dependson="mod.1"}
 tab_model(mod.2, mod.1) # Table
+```
+
+```{r}
+# check model with bayesian estimator (preferred)
+mod.1aaB <-brm(PWI ~  1 +  years * Beliefs + (1|Id), data= dfben, file ="mod.1aaB")
+mod.2aaB <-brm (LIFESAT ~ 1 + years * Beliefs + (1|Id), data= dfben,
+                file ="mod.2aaB")
+```
+
+note that the bayesian model recovers the time slope
+
+```{r bayesian_pwi}
+conditional_effects(mod.1aaB)[[3]]
+```
+
+```{r bayesian_lifesat}
+conditional_effects(mod.2aaB)[[3]]
+
+```
+
+Try with a tensor spline
+
+```{r}
+# complete cases
+
+dfbenTS<-dfben%>%
+  select(PWI,LIFESAT,yearsC,Beliefs,Id)%>%
+  filter(complete.cases(.))
+```
+
+```{r}
+# check model with bayesian estimator (preferred)
+mod.1aaB_ts <-brm(PWI ~  1 +  ts(years) + (1|Id), data= dfbenTS, file ="mod.1aaB_ts")
+
+mod.2aaB_ts <-brm (LIFESAT ~ 1 + ts(Beliefs,years) + (1|Id), data= dfbenTS,
+                file ="mod.2aaB_ts")
+```
+
+```{r}
+str(dfben$Beliefs)
+conditional_effects(mod.1aaB_ts, surface = TRUE)
+conditional_smooths(mod.1aaB_ts)
+```
+
+```{r}
+conditional_effects(mod.2aaB_ts, surface = TRUE)
+```
+
+```{r bayesian_pwi_plot }
+pr.PWI_b <- ggpredict(model = mod.1aaB, terms = c("years [0:9]","Beliefs"), 
+                    ci.lvl = 0.95,
+                    type = "fe",
+                    typical = "mean",
+                   # condition = c(Euro =1),
+                    back.transform = TRUE,
+                    ppd = FALSE,
+                    interval = "confidence")
+p1<-plot(pr.PWI_b, facets = T) + 
+ # scale_y_continuous(limits=c(4.75,5.25) )  + 
+  gghighlight()  +  theme_blank()+ 
+  ggtitle("Bayesian Model Predicted Values of Personal Wellbeing") 
+p1
+pr.PWI_b
+```
+
+```{r bayesian_pwi_plot }
+pr.LS_b <- ggpredict(model = mod.2aaB, terms = c("years [0:9]","Beliefs"), 
+                    ci.lvl = 0.95,
+                    type = "fe",
+                    typical = "mean",
+                   # condition = c(Euro =1),
+                    back.transform = TRUE,
+                    ppd = FALSE,
+                    interval = "confidence")
+p1<-plot(pr.LS_b, facets = T) + 
+ # scale_y_continuous(limits=c(4.75,5.25) )  + 
+  gghighlight()  +  theme_blank()+ 
+  ggtitle("Bayesian Model Predicted Values of Personal Wellbeing") 
+p1
+pr.LS_b
+```
+
+```{r bayesian_pwi_plot }
+pr.PWI_b <- ggpredict(model = mod.21aaB, terms = c("years [0:9]","Beliefs"), 
+                    ci.lvl = 0.95,
+                    type = "fe",
+                    typical = "mean",
+                   # condition = c(Euro =1),
+                    back.transform = TRUE,
+                    ppd = FALSE,
+                    interval = "confidence")
+p1<-plot(pr.PWI_b, facets = T) + 
+ # scale_y_continuous(limits=c(4.75,5.25) )  + 
+  gghighlight()  +  theme_blank()+ 
+  ggtitle("Vayesian Model Predicted Values of Personal Wellbeing") 
+p1
+pr.PWI_b
+```
+
+```{r bayesian_pwi_plot }
+pr.LS_b <- ggpredict(model = mod.2aaB, terms = c("years [0:9]","Beliefs"), 
+                    ci.lvl = 0.95,
+                    type = "fe",
+                    typical = "mean",
+                   # condition = c(Euro =1),
+                    back.transform = TRUE,
+                    ppd = FALSE,
+                    interval = "confidence")
+p1<-plot(pr.LS_b, facets = T) + 
+ # scale_y_continuous(limits=c(4.75,5.25) )  + 
+  gghighlight()  +  theme_blank()+ 
+  ggtitle("Bayesian Model Predicted Values of Personal Wellbeing") 
+p1
+pr.LS_b
 ```

### 2020-11-22 Analysis.Rmd
Further within and between analysis

@@
 ```{r}
+# must remove Furniture package before running table 1
 table1::table1(~  Age + NZdep + Education + Employed + EthnicCategories + Male  + Partner  + Pol.Orient + Urban +  Beliefs + LIFESAT + PWI|Wave, data = dfben, overall = F)
@@
 )
+
 ```
@@
 ### Spirit Belief Models ### 
-mod01<-lmer (LIFESAT ~ Years *(
+mod01<-lmer (PWI ~ Years +
                 Beliefs__Skeptic__between  +  
@@
                 Beliefs_GodAndSpirit_within  + 
-                Beliefs_GodExcludesSpirit_within) +
+                Beliefs_GodExcludesSpirit_within +
                 (1|Id), data= dfben, na.action = na.omit)
@@
 parameters::check_heterogeneity(mod01, group = "ID")
+parameters::check_heterogeneity(mod01, group = "ID")
 
-mod02<-lmer (LIFESAT ~ Years *( 
+mod02<-lmer (LIFESAT ~ Years +
                 Beliefs__Skeptic__between  +  
@@
 
-# inference holds, but can we use this method with factors?  In any case, we keep to the simple model
+# inference holds, but can we use this method with factors? What do the contrasts mean? 
 tab_model(mod01)
 tab_model(mod02)
+plot_models(mod01,mod02)
+```
 
+```{r models, cache = TRUE, dependson = "dfben"}
+### Spirit Belief Models ### 
+mod01<-lmer (PWI ~ yearsC +
+                Beliefs__Skeptic__between  +  
+                Beliefs__SpiritExcludesGod__between + 
+                Beliefs_GodAndSpirit_between  + 
+                Beliefs_GodExcludesSpirit_between +
+                Beliefs__Skeptic__within  +  
+                Beliefs__SpiritExcludesGod__within + 
+                Beliefs_GodAndSpirit_within  + 
+                Beliefs_GodExcludesSpirit_within +
+                (1|Id), data= dfben, na.action = na.omit)
+
+parameters::check_heterogeneity(mod01, group = "ID")
+
+mod02<- lmer(LIFESAT ~ yearsC +
+                Beliefs__Skeptic__between  +  
+                Beliefs__SpiritExcludesGod__between + 
+                Beliefs_GodAndSpirit_between  + 
+                Beliefs_GodExcludesSpirit_between +
+                Beliefs__Skeptic__within  +  
+                Beliefs__SpiritExcludesGod__within + 
+                Beliefs_GodAndSpirit_within  + 
+                Beliefs_GodExcludesSpirit_within +
+                (1|Id), data= dfben, na.action = na.omit)
+
+parameters::check_heterogeneity(mod02, group = "ID")
+
+# inference holds, but can we use this method with factors? What do the contrasts mean? 
+tab_model(mod01)
+tab_model(mod02)
+plot_models(mod01,mod02)
+
+performance::check_collinearity(mod01)
+performance::check_collinearity(mod02)
+
+```
+
+parameters::check_heterogeneity(mod02, group = "ID")
+
+# inference holds, but can we use this method with factors? What do the contrasts mean? 
+tab_model(mod01)
+tab_model
+
+## BAYES
+
+```{r bayesian_pwi_plot }
+pr.LS_b <- ggpredict(model = mod.2aaB, terms = c("years [0:9]","Beliefs"), 
+                    ci.lvl = 0.95,
+                    type = "fe",
+                    typical = "mean",
+                   # condition = c(Euro =1),
+                    back.transform = TRUE,
+                    ppd = FALSE,
+                    interval = "confidence")
+p1<-plot(pr.LS_b, facets = T) + 
+ # scale_y_continuous(limits=c(4.75,5.25) )  + 
+  gghighlight()  +  theme_blank()+ 
+  ggtitle("Bayesian Model Predicted Values of Personal Wellbeing") 
+p1
+pr.LS_b
+```
+```
+
+```{r lme model}
 # Inference holds 
@@
 ```
-Centering years makes no difference to inference.
 
@@
 sjPlot::tab_model(mod.1aa,mod.2aa)
+```
+
+```{r check collinearity}
+# Note High VIFS from 
+performance::check_collinearity(mod.1aa)
+performance::check_collinearity(mod.2aa)
+
+parameters::check_heterogeneity(mod.1aa, group = "ID")
+parameters::check_heterogeneity(mod.2aa, group = "ID")
+
 ```

### 2020-11-25 Manuscript_Bayesian.Rmd
Add files via upload
bayesian replication document

--- 2020-11-20 Manuscript.Rmd
@@
 ---
-title             : "National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Beliefs"
+title             : "Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Belief"
 shorttitle        : "Longitudinal Wellbeing from Spiritual Beliefs"
@@
 abstract: | 
-
   Previous research finds an association between spirituality and subjective well-being.
@@
 bibliography      : ["b-references.bib"]
-
 floatsintext      : no
@@
 draft             : no
-
 documentclass     : "apa6"
@@
 output            : papaja::apa6_pdf
-extra_dependencies: ["rotating", "booktabs", "graphicx","longtable","graphicx"]
-header-includes:
-   - \usepackage{tcolorbox}
-   - \usepackage{paralist}
-   - \usepackage{multicol}
-   - \usepackage{graphics}
-   - \usepackage{rotating}
-   - \usepackage{longtable}
+extra_dependencies: ["rotating", "booktabs", "graphicx","longtable","amsmath","amsfonts","amssymb"]
+# header-includes:
+#    - \usepackage{tcolorbox}
+#    - \usepackage{paralist}
+#    - \usepackage{multicol}
+#    - \usepackage{graphics}
+#    - \usepackage{rotating}
+#    - \usepackage{longtable}
+#    - \usepackage[amsmath]
 ---
@@
 
-<!-- In spirituality/psychological wellbeing research, subjective in spirituality health research, psychological well-being is typically decomposed into two categories: emotional well-being (i.e., the presence or absence of positive feelings about life) and functional well-being (i.e., psychological and social well-being) [@Keyes2002-bb].
--->
-
 First, previous research has frequently operationalised spirituality using measures of well-being.
 For example, The *Daily Spiritual Experience Scale*, which is widely used in spiritual mental health research, includes items such as "I feel thankful for my blessings" and "I feel deep inner peace or harmony" to measure spirituality [@Underwood2002-hg].
-As Koenig observes, it is unsurprising that spirituality predict well-being when the same items are used to measure both constructs.
-Koenig aptly describes this approach as "tautological" [@Koenig2008-lv].
-Indeed, recent systematic-review of spirituality/mental health research found that nearly 45% of previous studies employ tautological scales [@Garssen2016-kb].
+These items are used both to measure spirituality and subjective well-being.
+As Koenig observes, it is unsurprising that spirituality is associated with well-being when the same items are used to measure both constructs.
+Koenig aptly describes this problematic approach as "tautological" [@Koenig2008-lv].
+How pervasive is this problem?
+A recent systematic-review of spirituality/mental health research found that nearly 45% of previous studies employ tautological scales [@Garssen2016-kb].
 The first challenge, then, is to define measures of spirituality that are not well-being measures dressed up in other words.
@@
 
-Here, we addresses the three challenges of previous in the following ways: First, to avoid a tautological elision of "spirituality" and "well-being" we adopt a non-affective measure of spirituality as a "believe in a Spirit or Life Force." On the other side, to measure subjective well-being, we use a well-validated scale for personal well-being and a well-validated scale for Life-Satisfaction (see Method).
+Here, we addresses these three challenges of previous research in the following ways: First, to avoid a tautological elision of "spirituality" and "well-being" we adopt a non-affective measure of spirituality as a "believe in a Spirit or Life Force." On the other side, to measure subjective well-being, we use a well-validated scale for personal well-being and a well-validated scale for Life-Satisfaction (see Method).
 Second, to disentangle spiritual beliefs from traditional religious beliefs and skeptical beliefs we adopt a four-level categorical indicator combining the binary indicator of belief in a Spirit or Life Force with the binary indicator "Do you believe in some form of a spirit or life force?" The levels of this four-level categorical indicator for spiritual, religious, and skeptical beliefs are as follows: (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who nevertheless believe in a Spirit or Life Force; (3) Those who believe in a God but do not believe in a Spirit or Life force; (4) Those who who believe in a both God and a Spirit or Life Force.
@@
 The spiritual indicators measure cognitive states in which distinct varieties of skeptical beliefs, traditional religious beliefs, and non-traditional spirit beliefs are straightforwardly disentangled, and measured within people.
-Third, to assess whether spirit beliefs lead to growth, stability, or decline in psychological well-being, we leverage longitudinal responses from a nationally diverse probability sample year over an nine--year period with nine measurement points (2010 to 2018 New Zealand Attitudes and Values Study, NZAVS).
+Third, to understand how spirit beliefs affect growth, stability, or decline in psychological well-being over time, we leverage longitudinal responses from a nationally diverse probability sample year over nine years (2010 to 2018 New Zealand Attitudes and Values Study, NZAVS).
 
@@
 Table 1 about here:
-\[
-\begin{table}[ ht ] 
-\centering 
-\caption{}\label{}
-\scalebox{.6}{
-\begin{tabular}{ l c c c c c c c c c }
-\toprule
- &   \multicolumn{ 9 }{c}{ Wave }\\ 
-  & 2010 & 2011 & 2012 & 2013 & 2014 & 2015 & 2016 & 2017 & 2018 \\ 
- & n = \emph{missing} & n = \emph{missing} & n = 8547 & n = 13472 & n = \emph{missing} & n = 11687 & n = 16048 & n = 14174 & n = 14883 \\ 
- \midrule
-Age &   &   &   &   &   &   &   &   &  \\ 
-NZdep &   &   &   &   &   &   &   &   &  \\ 
-Education &   &   &   &   &   &   &   &   &  \\ 
-Employed &   &   &   &   &   &   &   &   &  \\ 
-\hspace{6pt}    0 &  (\%) &  (\%) & 2094 (24.5\%) & 2955 (21.9\%) &  (\%) & 2525 (21.6\%) & 3486 (21.7\%) & 3223 (22.7\%) & 3404 (22.9\%)\\ 
-\hspace{6pt}    1 &  (\%) &  (\%) & 6453 (75.5\%) & 10517 (78.1\%) &  (\%) & 9162 (78.4\%) & 12562 (78.3\%) & 10951 (77.3\%) & 11479 (77.1\%)\\ 
-EthnicCategories &   &   &   &   &   &   &   &   &  \\ 
-\hspace{6pt}    Euro &  (\%) &  (\%) & 6731 (78.8\%) & 11024 (81.8\%) &  (\%) & 9671 (82.8\%) & 13389 (83.4\%) & 11814 (83.3\%) & 12427 (83.5\%)\\ 
-\hspace{6pt}    Maori &  (\%) &  (\%) & 1188 (13.9\%) & 1608 (11.9\%) &  (\%) & 1341 (11.5\%) & 1749 (10.9\%) & 1620 (11.4\%) & 1677 (11.3\%)\\ 
-\hspace{6pt}    Pacific &  (\%) &  (\%) & 284 (3.3\%) & 339 (2.5\%) &  (\%) & 277 (2.4\%) & 331 (2.1\%) & 243 (1.7\%) & 261 (1.8\%)\\ 
-\hspace{6pt}    Asian &  (\%) &  (\%) & 344 (4\%) & 501 (3.7\%) &  (\%) & 398 (3.4\%) & 579 (3.6\%) & 497 (3.5\%) & 518 (3.5\%)\\ 
-Male &   &   &   &   &   &   &   &   &  \\ 
-\hspace{6pt}    0 &  (\%) &  (\%) & 5302 (62\%) & 8457 (62.8\%) &  (\%) & 7325 (62.7\%) & 10148 (63.2\%) & 9011 (63.6\%) & 9433 (63.4\%)\\ 
-\hspace{6pt}    1 &  (\%) &  (\%) & 3245 (38\%) & 5015 (37.2\%) &  (\%) & 4362 (37.3\%) & 5900 (36.8\%) & 5163 (36.4\%) & 5450 (36.6\%)\\ 
-Partner &   &   &   &   &   &   &   &   &  \\ 
-\hspace{6pt}    0 &  (\%) &  (\%) & 2408 (28.2\%) & 3546 (26.3\%) &  (\%) & 2815 (24.1\%) & 3773 (23.5\%) & 3330 (23.5\%) & 3556 (23.9\%)\\ 
-\hspace{6pt}    1 &  (\%) &  (\%) & 6139 (71.8\%) & 9926 (73.7\%) &  (\%) & 8872 (75.9\%) & 12275 (76.5\%) & 10844 (76.5\%) & 11327 (76.1\%)\\ 
-Pol Orient &   &   &   &   &   &   &   &   &  \\ 
-Urban &   &   &   &   &   &   &   &   &  \\ 
-\hspace{6pt}    Not\_Urban &  (\%) &  (\%) & 2865 (33.5\%) & 4278 (31.8\%) &  (\%) & 3951 (33.8\%) & 5538 (34.5\%) & 2610 (18.4\%) & 2661 (17.9\%)\\ 
-\hspace{6pt}    Urban &  (\%) &  (\%) & 5682 (66.5\%) & 9194 (68.2\%) &  (\%) & 7736 (66.2\%) & 10510 (65.5\%) & 11564 (81.6\%) & 12222 (82.1\%)\\ 
-Beliefs &   &   &   &   &   &   &   &   &  \\ 
-\hspace{6pt}    \_Skeptic\_ &  (\%) &  (\%) & 2048 (24\%) & 3286 (24.4\%) &  (\%) & 2951 (25.3\%) & 4116 (25.6\%) & 3916 (27.6\%) & 4120 (27.7\%)\\ 
-\hspace{6pt}    \_SpiritExcludesGod\_ &  (\%) &  (\%) & 2295 (26.9\%) & 3735 (27.7\%) &  (\%) & 3169 (27.1\%) & 4501 (28\%) & 3918 (27.6\%) & 3970 (26.7\%)\\ 
-\hspace{6pt}    GodAndSpirit &  (\%) &  (\%) & 3707 (43.4\%) & 5499 (40.8\%) &  (\%) & 5119 (43.8\%) & 6453 (40.2\%) & 5464 (38.5\%) & 6210 (41.7\%)\\ 
-\hspace{6pt}    GodExcludesSpirit &  (\%) &  (\%) & 497 (5.8\%) & 952 (7.1\%) &  (\%) & 448 (3.8\%) & 978 (6.1\%) & 876 (6.2\%) & 583 (3.9\%)\\ 
-LIFESAT &   &   &   &   &   &   &   &   &  \\ 
-PWI &   &   &   &   &   &   &   &   &  \\ 
-\bottomrule
-\end{tabular}
-}
-\end{table}
-\]
 
-## Demographic Indicators
-We included standard demographic indicators used in New Zealand Attitudes and Values Research: [NZAVS modelling standards](https://cdn.auckland.ac.nz/assets/psych/about/our-research/nzavs/NZAVSTechnicalDocuments/NZAVS-Technical-Documents-e11-Default-Statistical-Standards.pdf)
+## Sample Demographics
 
@@
 ## Statistical Analyses
+
 Statistical analysis was performed using R version 4.0.2 R version 3.6.1 (2020-06-22) 2019-07-05).
@@
 To handle the dependencies and heterogeneity introduced from the repeated measures, we include individual ID as an effect modelled as random.
+
+## Potential Confounders and Missing Data
+
+In a supplemental analysis we include standard demographic indicators used in New Zealand Attitudes and Values Research: [NZAVS modelling standards](https://cdn.auckland.ac.nz/assets/psych/about/our-research/nzavs/NZAVSTechnicalDocuments/NZAVS-Technical-Documents-e11-Default-Statistical-Standards.pdf).
 We centered and scaled the education, deprivation, and political conservativism.
 We centred age was centred at and put into decade-long intervals.
-The model equation(s) are described in Equation 1.
+The model equation(s) are described in Equation 2.
 
-## Equation 1
-\[
-\begin{aligned}
-\operatorname{{\small LIFESAT/PWI}}  &\sim N \left(\mu,\sigma^2 \right) \\ \mu &=\alpha_{j[i]} + \beta_{1}(\operatorname{Years}) + \beta_{2}(\operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\
-&\quad \beta_{3}(\operatorname{Beliefs}_{\operatorname{GodAndSpirit}}) + \beta_{4}(\operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) + \beta_{5}(\operatorname{Age.10yrs.C})\ + \\
-&\quad \beta_{6}(\operatorname{Deprivation.S}) + \beta_{7}(\operatorname{Edu.S}) + \beta_{8}(\operatorname{Employed})\ + \\
-&\quad \beta_{9}(\operatorname{EthnicCats}_{\operatorname{Maori}}) + \beta_{10}(\operatorname{EthnicCats}_{\operatorname{Pacific}}) + \beta_{11}(\operatorname{EthnicCats}_{\operatorname{Asian}})\ + \\
-&\quad \beta_{12}(\operatorname{Male}_{\operatorname{1}}) + \beta_{13}(\operatorname{Partner}_{\operatorname{Has\ }}) + \beta_{14}(\operatorname{Pol.Orient.S})\ + \\
-&\quad \beta_{15}(\operatorname{Urban}) + \beta_{16}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}}) + \beta_{17}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ + \\
-&\quad \beta_{18}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) \\ \alpha_{j} &\sim N \left(\mu_{\alpha_{j}},\sigma^2_{\alpha_{j}} \right) , \operatorname{ for  Id }~j~= 1, \dots~J = 20,979~\operatorname{Individuals}
-\end{aligned}
-\]
-
-### System and packages
-The analysis was performed using R version 4.0.2 (2020-06-22). The Platform was x86_64-apple-darwin17.0 (64-bit) Running under: macOS Catalina 10.15.6
-We are greatful to the contributors and mantainers of the following packages:
-R [Version 4.0.2; @R-base] and the R-packages *Amelia* [Version 1.7.6; @R-Amelia], *citr* [Version 0.3.2; @R-citr], *dplyr* [Version 1.0.2; @R-dplyr], *equatiomatic* [Version 0.1.0.9000; @R-equatiomatic], *forcats* [Version 0.5.0; @R-forcats], *ggeffects* [Version 0.15.0; @R-ggeffects], *gghighlight* [Version 0.3.0; @R-gghighlight], *ggplot2* [Version 3.3.2; @R-ggplot2], *ggsci* [Version 2.9; @R-ggsci], *lme4* [Version 1.1.23; @R-lme4], *Matrix* [Version 1.2.18; @R-Matrix], *papaja* [Version 0.1.0.9997; @R-papaja], *parameters* [Version 0.8.6.1; @R-parameters], *patchwork* [Version 1.0.1; @R-patchwork], *prettycode* [Version 1.1.0; @R-prettycode], *purrr* [Version 0.3.4; @R-purrr], *Rcpp* [Version 1.0.5.1; @R-Rcpp_a; @R-Rcpp_b], *readr* [Version 1.3.1; @R-readr], *report* [Version 0.1.0; @R-report], *see* [Version 0.5.2.1; @R-see], *sjPlot* [Version 2.8.4; @R-sjPlot], *stringr* [Version 1.4.0; @R-stringr], *styler* [Version 1.3.2; @R-styler], *table1* [Version 1.2; @R-table1], *texreg* [Version 1.37.5; @R-texreg], *tibble* [Version 3.0.4; @R-tibble], *tidyr* [Version 1.1.2; @R-tidyr], and *tidyverse* [Version 1.3.0; @R-tidyverse]"
-
-### Missing data
 Because missing values can result in biased estimates [@Blackwell2017-oq], we multiply imputed 10 datasets using the AmeliaII package in R[@Honaker2011-yu].We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences.
@@
 The supplement describing these analysis is located at: https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing.
+
+### System and packages
+
+The analysis was performed using R version 4.0.2 (2020-06-22). The Platform was x86_64-apple-darwin17.0 (64-bit) Running under: macOS Catalina 10.15.6
+We are greatful to the contributors and mantainers of the following packages:
+R [Version 4.0.2; @R-base] and the R-packages *Amelia* [Version 1.7.6; @R-Amelia], *citr* [Version 0.3.2; @R-citr], *dplyr* [Version 1.0.2; @R-dplyr], *equatiomatic* [Version 0.1.0.9000; @R-equatiomatic], *forcats* [Version 0.5.0; @R-forcats], *ggeffects* [Version 0.15.0; @R-ggeffects], *gghighlight* [Version 0.3.0; @R-gghighlight], *ggplot2* [Version 3.3.2; @R-ggplot2], *ggsci* [Version 2.9; @R-ggsci], *lme4* [Version 1.1.23; @R-lme4], *Matrix* [Version 1.2.18; @R-Matrix], *papaja* [Version 0.1.0.9997; @R-papaja], *parameters* [Version 0.8.6.1; @R-parameters], *patchwork* [Version 1.0.1; @R-patchwork], *prettycode* [Version 1.1.0; @R-prettycode], *purrr* [Version 0.3.4; @R-purrr], *Rcpp* [Version 1.0.5.1; @R-Rcpp_a; @R-Rcpp_b], *readr* [Version 1.3.1; @R-readr], *report* [Version 0.1.0; @R-report], *see* [Version 0.5.2.1; @R-see], *sjPlot* [Version 2.8.4; @R-sjPlot], *stringr* [Version 1.4.0; @R-stringr], *styler* [Version 1.3.2; @R-styler], *table1* [Version 1.2; @R-table1], *texreg* [Version 1.37.5; @R-texreg], *tibble* [Version 3.0.4; @R-tibble], *tidyr* [Version 1.1.2; @R-tidyr], and *tidyverse* [Version 1.3.0; @R-tidyverse]"
 
@@
 
-\[
-\begin{table}
-\begin{center}
-\scalebox{0.6}{
-\begin{tabular}{l c c}
-\toprule
- & PWI & Life Satisfaction \\
-\midrule
-(Intercept)                        & $\mathbf{6.63}^{***}$  & $\mathbf{4.94}^{***}$  \\
-                                   & $(0.03)$               & $(0.02)$               \\
-Years                              & $0.00$                 & $-0.00$                \\
-                                   & $(0.00)$               & $(0.00)$               \\
-Beliefs\_SpiritExcludesGod\_       & $\mathbf{-0.14}^{***}$ & $\mathbf{-0.08}^{***}$ \\
-                                   & $(0.03)$               & $(0.02)$               \\
-BeliefsGodAndSpirit                & $\mathbf{-0.16}^{***}$ & $-0.04$                \\
-                                   & $(0.03)$               & $(0.02)$               \\
-BeliefsGodExcludesSpirit           & $0.02$                 & $0.07$                 \\
-                                   & $(0.05)$               & $(0.04)$               \\
-Age.10yrs.C                        & $\mathbf{0.08}^{***}$  & $\mathbf{0.09}^{***}$  \\
-                                   & $(0.01)$               & $(0.01)$               \\
-Deprivation.S                      & $\mathbf{-0.14}^{***}$ & $\mathbf{-0.07}^{***}$ \\
-                                   & $(0.01)$               & $(0.01)$               \\
-Edu.S                              & $\mathbf{0.12}^{***}$  & $\mathbf{0.06}^{***}$  \\
-                                   & $(0.01)$               & $(0.01)$               \\
-EmployedEmployed                   & $\mathbf{0.09}^{***}$  & $\mathbf{0.08}^{***}$  \\
-                                   & $(0.01)$               & $(0.01)$               \\
-EthnicCatsMaori                    & $\mathbf{-0.30}^{***}$ & $\mathbf{-0.06}^{**}$  \\
-                                   & $(0.03)$               & $(0.02)$               \\
-EthnicCatsPacific                  & $\mathbf{-0.29}^{***}$ & $\mathbf{-0.08}^{*}$   \\
-                                   & $(0.05)$               & $(0.04)$               \\
-EthnicCatsAsian                    & $-0.08$                & $-0.04$                \\
-                                   & $(0.05)$               & $(0.03)$               \\
-Male1                              & $\mathbf{-0.11}^{***}$ & $\mathbf{-0.19}^{***}$ \\
-                                   & $(0.02)$               & $(0.01)$               \\
-PartnerHas Partner                 & $\mathbf{0.61}^{***}$  & $\mathbf{0.42}^{***}$  \\
-                                   & $(0.02)$               & $(0.01)$               \\
-Pol.Orient.S                       & $\mathbf{0.02}^{***}$  & $\mathbf{0.01}^{**}$   \\
-                                   & $(0.01)$               & $(0.00)$               \\
-UrbanUrban                         & $\mathbf{-0.02}^{*}$   & $\mathbf{-0.02}^{*}$   \\
-                                   & $(0.01)$               & $(0.01)$               \\
-Years:Beliefs\_SpiritExcludesGod\_ & $\mathbf{0.02}^{**}$   & $\mathbf{0.02}^{***}$  \\
-                                   & $(0.00)$               & $(0.00)$               \\
-Years:BeliefsGodAndSpirit          & $\mathbf{0.02}^{***}$  & $\mathbf{0.01}^{***}$  \\
-                                   & $(0.00)$               & $(0.00)$               \\
-Years:BeliefsGodExcludesSpirit     & $0.00$                 & $-0.00$                \\
-                                   & $(0.01)$               & $(0.01)$               \\
-\midrule
-AIC                                & $248580.18$            & $200599.16$            \\
-BIC                                & $248775.08$            & $200793.96$            \\
-Log Likelihood                     & $-124269.09$           & $-100278.58$           \\
-Num. obs.                          & $79270$                & $78888$                \\
-Num. groups: Id                    & $20979$                & $20972$                \\
-Var: Id (Intercept)                & $1.78$                 & $0.88$                 \\
-Var: Residual                      & $0.74$                 & $0.42$                 \\
-\bottomrule
-\multicolumn{3}{l}{\scriptsize{$^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$}}
-\end{tabular}
-}
-\caption{Regression Models}
-\label{table:coefficients}
-\end{center}
-\end{table}
-\]
+# Discussion
+
+The purpose of this study has been to clarify the relationship between belief in a spirit/life force and subjective well-being, addressing three challenges from previous research: (1) tautological measures of spirituality and subjective well-being; (2) insufficiently distinct measures of spirituality and religion; (3) cross-sectional samples.
+Here, we ask whether beliefs in a Spirit or Life Force predict growth in personal well-being and life-satisfaction over 9 waves (8 years) in nationally diverse western populations who differ in the dimensions of spiritual and religious beliefs.
+We observe that:
+
+1.
+At the outset of the study, in 2010, the expected subjective well-being as measured using the personal well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God ($\hat{y}$ =  6.50, se =.03) and lower among those who believed in both a Spirit/Life-Force and a God ($\hat{y}$ =  6.48, se =.03) in comparison both to the skeptical population who neither believed in a God nor a life force ($\hat{y}$ = 6.64,se =.03) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 6.66, se =.05).
+(See Figure 2)  Similarly,the expected subjective well-being as measured using the life satisfaction Inventory is lower among those who believed in Spirit or Life Force but not a God ($\hat{y}$ = 4.86, se =.02) and lower among those who believed in both a Spirit/Life-Force and a God  ($\hat{y}$ = 4.91, se =.02) in comparison both with the skeptical population who neither believed in a God nor a life force ($\hat{y}$ = 4.94,se =.02) and the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 5.02, se =.02) (see Figure 3)
+
+2.
+However, over time, belief in a Spirit or Life Force predicts a trajectory of growth in personal well-being.
+In 2018, personal well-being among who believe in Spirit or Life Force but not a God had grown to ($\hat{y}$ = 6.67, se =.03, $\Delta \hat{y} = 0.17$).
+Among those who believed in both a Spirit/Life-Force and a God, personal well-being had grown to ($\hat{y} = 6.69, se = 0.02, \Delta \hat{y} = .21$).
+To place this effect in practical perspective, such growth in personal well-being is about twice as large as the expected benefit of being employed and about a third as large as the expected benefit from having a romantic partner.
+By contrast, personal well-being among the skeptical population who neither believed in a God nor a Life Force remained stable across the eight year interval $\hat{y} = 6.66, se = 0.03, \Delta \hat{y} = .02$).
+By the same token, personal well-being among those who believe in a God but are skeptical about a Spirit/Life Force was constant $\hat{y} =  6.70, se = 0.04, \Delta \hat{y} = .04$).
+Similarly, over time, belief in a Spirit or Life Force predicts a trajectory of growth in life satisfaction.
+In 2018, life satisfaction among those who believe in Spirit or Life Force but not a God had grown to ($\hat{y} = 5.01, se = 0.02, \Delta \hat{y} = 0.07$).
+In practical terms, such growth is equivalent to the expected gain in life satisfaction from being employed and a sixth as large as the expected benefit to life satisfaction from having a romantic partner.
+Among those who believed in both a Spirit/Life-Force and a God, life satisfaction had grown to ($\hat{y} = 5.03, se = 0.02, \Delta \hat{y} = 0.12$).
+In practical terms, such growth is equivalent to 1.5 times the expected gain to life satisfaction from being employed and almost a third benefit to life satisfaction from having a romantic partner.
+By contrast, life satisfaction among the skeptical population who neither believed in a God nor a Life Force remained stable over the eight year interval $\hat{y} =  4.94, se =0.02, \Delta \hat{y} = 0$).
+Relatedly, life satisfaction in the population that believe in God but are skeptical about a Spirit/Life Force was constant $\hat{y} =  5.00, se = 0.03, \Delta \hat{y} = -.02$) (See Figure 3).
+
+3.
+Inferred growth from spiritual beliefs is evident from 9 annual waves of data collection (8 years), and across a nationally diverse sample.
+On this basis of this longitudinal information accumulated across nationally, we are confident that growth in personal well-being and in life-satisfaction among those who believe in a Spirit and Life force are evident.
+Such effects co-occur in people who lack a belief in a God as well as among people who believe in a God as well as a Spirit or Life force.
+
+## Key Questions
+
+Two fundamental question arise from the patterns we observe in this study.
+The first question is why those who believed in Spirit or Life Force were initially lower in their personal well-being and life-satisfaction in the initial years of this study in comparison to religious and non-religious skeptics?
+During the early years of New Zealand Attitudes and Values data collection New Zealand was emerging from the twin impacts of a global financial recession and a series of devastating earthquakes that destroyed 30% of Christchurch, New Zealand's second largest city [@greaves2015regional].
+Orthogonal research observes that economic and ecological distress are associated with increases in traditional religious beliefs [@sibley2012faith].
+Experimental research has found among religious people death anxiety provokes acceptance of non-traditional religious beliefs [@norenzayan2006belief].
+Because spiritual belief items were only introduced during the year of the earthquakes, we cannot here resolve whether the shifting relationship between spirit beliefs and well-being were the result of demand for spiritual beliefs among those secular and religious people who were adversely affected by the financial crisis or natural disaster.
+On the other hand, such exogenous influences might have adversely affected those with Spirit/Life force beliefs in the early phases of this study, and the patterns observed here are better understood as a regression to a population mean.
+Though beyond the scope of this study, how subjective well-being changes in response to changes in spiritual beliefs, as well as special sensitivities to distress among people with spiritual beliefs are important questions for future investigations.
+Despite the longitudinal quality of this study, a causal explanation remains elusive.
+
+Second, our task here has been to clarify whether beliefs predict growth in well-being.
+However, as we have indicated at the outset, the term "spirituality" applies to a family of concepts.
+We have not attempted to model the diverse philosophical, social, and behavioral attributes within people who adhere to spirit beliefs.
+Moreover, people might identify as spiritual without having any accompanying belief in a Spirit or Life force.
+The task of clarifying how distinct features across a diverse family of "spirituality" concepts affect perceptions of individual well-being remain intriguing horizons.
+
+## Importance
+Despite its limitations, our study is important both for its methods and its findings.
+We demonstrate that spirituality can be operationalised to avoid the pitfalls of tautology evident in previous research by focusing on states of beliefs, that concepts of such belief states can be refined to enable distinctions necessary to investigate spirituality in mixed secular and religious populations, and that the question of whether beliefs states predict growth in subjective well-being can be investigated in a longitudinal setting.
+Our finding that, indeed, spiritual beliefs predict growth in perceived well-being and life satisfaction may hold practical importance.
+Previous research indicates that religious beliefs are associated with psychological well-being [@Smith2003-re].
+Western populations are increasingly identifying as "spiritual but not religious" [@Zinnbauer2005-vz; @Swinton2001-vr].
+Here, we observe that even among people who lack a traditional religious belief in God, believing in a Spirit or Life force predicts incremental improvements in well-being that sum to practically large effects.
+Though the contextual and psychological mechanisms that underpin such growth will require many decades of research, this preliminary finding is good news for those increasing numbers of individuals continue to report spiritual beliefs but who do not affirm traditional religious beliefs.
+
+### Authors contributions
+BH conceived of the study.
+BH wrote the first draft of the manuscript, and performed the initial data analysis with input from JB.
+EW improved the approach with BH and JB and revised the manuscript.
+DD revised the manuscript.
+CS was responsible for NZAVS data collection and revised the manuscript.
+JB improved the approach with BH and EW, performed a data re-analysis, and revised the manuscript.
+
+### Funding statement
+The New Zealand Attitudes and Values Study is supported by a grant from the Templeton Religion Trust (TRT0196). The funders had no role in the preparation of the article or the decision to publish.
+
+\newpage
+
+## Supplement
 
@@
 
-# Discussion
+\newpage
+# Supplement Model Equations
 
-The purpose of this study has been to clarify the relationship between belief in a spirit/life force and subjective well-being, addressing three challenges from previous research: (1) tautological measures of spirituality and subjective well-being; (2) insufficiently distinct measures of spirituality and religion; (3) cross-sectional samples.
-Here, we ask whether beliefs in a Spirit or Life Force predict growth in personal well-being and life-satisfaction over 9 waves (8 years) in nationally diverse western populations who differ in the dimensions of spiritual and religious beliefs.
-We observe that:
+## Priors for the Personal Wellbeing model
 
-1.
-At the outset of the study, in 2010, the expected subjective well-being as measured using the personal well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God ($\hat{y}$ =  6.50, se =.03) and lower among those who believed in both a Spirit/Life-Force and a God ($\hat{y}$ =  6.48, se =.03) in comparison both to the skeptical population who neither believed in a God nor a life force ($\hat{y}$ = 6.64,se =.03) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 6.66, se =.05).
-(See Figure 2)  Similarly,the expected subjective well-being as measured using the life satisfaction Inventory is lower among those who believed in Spirit or Life Force but not a God ($\hat{y}$ = 4.86, se =.02) and lower among those who believed in both a Spirit/Life-Force and a God  ($\hat{y}$ = 4.91, se =.02) in comparison both with the skeptical population who neither believed in a God nor a life force ($\hat{y}$ = 4.94,se =.02) and the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 5.02, se =.02) (see Figure 3)
+Model priors for Personal Wellbeing including group-level ID intercept and beliefs slopes are as follows
 
-2.
-However, over time, belief in a Spirit or Life Force predicts a trajectory of growth in personal well-being.
-In 2018, personal well-being among who believe in Spirit or Life Force but not a God had grown to ($\hat{y}$ = 6.67, se =.03, $\Delta \hat{y} = 0.17$).
-Among those who believed in both a Spirit/Life-Force and a God, personal well-being had grown to ($\hat{y} = 6.69, se = 0.02, \Delta \hat{y} = .21$).
-To place this effect in practical perspective, such growth in personal well-being is about twice as large as the expected benefit of being employed and about a third as large as the expected benefit from having a romantic partner.
-By contrast, personal well-being among the skeptical population who neither believed in a God nor a Life Force remained stable across the eight year interval $\hat{y} = 6.66, se = 0.03, \Delta \hat{y} = .02$).
-By the same token, personal well-being among those who believe in a God but are skeptical about a Spirit/Life Force was constant $\hat{y} =  6.70, se = 0.04, \Delta \hat{y} = .04$).
-Similarly, over time, belief in a Spirit or Life Force predicts a trajectory of growth in life satisfaction.
-In 2018, life satisfaction among those who believe in Spirit or Life Force but not a God had grown to ($\hat{y} = 5.01, se = 0.02, \Delta \hat{y} = 0.07$).
-In practical terms, such growth is equivalent to the expected gain in life satisfaction from being employed and a sixth as large as the expected benefit to life satisfaction from having a romantic partner.
-Among those who believed in both a Spirit/Life-Force and a God, life satisfaction had grown to ($\hat{y} = 5.03, se = 0.02, \Delta \hat{y} = 0.12$).
-In practical terms, such growth is equivalent to 1.5 times the expected gain to life satisfaction from being employed and almost a third benefit to life satisfaction from having a romantic partner.
-By contrast, life satisfaction among the skeptical population who neither believed in a God nor a Life Force remained stable over the eight year interval $\hat{y} =  4.94, se =0.02, \Delta \hat{y} = 0$).
-Relatedly, life satisfaction in the population that believe in God but are skeptical about a Spirit/Life Force was constant $\hat{y} =  5.00, se = 0.03, \Delta \hat{y} = -.02$) (See Figure 3).
+$$
+\begin{aligned}
+\operatorname{Personal Well Being}_{ij} \sim \text{Normal}(\mu_{ij},\sigma^2) \nonumber\\
+\mu_{ij}\sim \boldsymbol{\alpha +\beta} \nonumber\\
+\boldsymbol{\alpha} \sim \alpha_{0} +\alpha_j \nonumber\\
+\alpha_{0} \sim \text{StudentT}(3,7.2,2.5)\nonumber\\
+\boldsymbol{\beta}\sim \boldsymbol{b} + \beta_j \nonumber\\
+\boldsymbol{b}\sim\text{Normal}(0,2.5) \nonumber\\
+\begin{bmatrix}
+\alpha_j \\
+\beta_j
+\end{bmatrix}
+\sim 
+\text{MVNormal}
+\begin{pmatrix}
+\begin{bmatrix}
+0\\
+0
+\end{bmatrix}
+,\boldsymbol{S} \nonumber
+\end{pmatrix}\\
+\boldsymbol{S} = 
+\begin{pmatrix}
+0 & \sigma_{\beta_j} \nonumber
+\end{pmatrix} \boldsymbol{R} \begin{pmatrix}
+\sigma_{\alpha_j} & 0 \\
+0 & \sigma_{\beta_j} \nonumber
+\end{pmatrix}\\
+\sigma_{\alpha_j} \sim 
+\text{HalfCauchy}(0,2.5) \nonumber\\
+\sigma_{\beta_j} \sim 
+\text{HalfCauchy}(0,2.5) \nonumber\\
+\boldsymbol{R}\sim \text{LKJcorr}(1) \nonumber\\
+\sigma  \sim  \operatorname{HalfStudent T}^{+}(3, 0, 2.5) 
+\end{aligned}
+$$
 
-3.
-Inferred growth from spiritual beliefs is evident from 9 annual waves of data collection (8 years), and across a nationally diverse sample.
-On this basis of this longitudinal information accumulated across nationally, we are confident that growth in personal well-being and in life-satisfaction among those who believe in a Spirit and Life force are evident.
-Such effects co-occur in people who lack a belief in a God as well as among people who believe in a God as well as a Spirit or Life force.
+\newpage
+## Priors for the Life Satisfaction model
 
-## Key Questions
+Model priors for Life Satisfaction Model with including group-level ID intercept and belief slopes are as follows:
+$$
+\begin{aligned}
+\operatorname{Life Satisfaction}_{ij} \sim \text{Normal}(\mu_{ij},\sigma^2) \nonumber\\
+\mu_{ij}\sim \boldsymbol{\alpha +\beta} \nonumber\\
+\boldsymbol{\alpha} \sim \alpha_{0} +\alpha_j \nonumber\\
+\alpha_{0} \sim \text{StudentT}(3,5.5, 2.5)\nonumber\\
+\alpha_j \sim \text{StudentT}+(3,0,10)\nonumber\\
+\boldsymbol{\beta}\sim \boldsymbol{b} + \beta_j \nonumber\\
+\boldsymbol{b}\sim\text{Normal}(0,2.5) \nonumber\\
+\begin{bmatrix}
+\alpha_j \\
+\beta_j
+\end{bmatrix}
+\sim 
+\text{MVNormal}
+\begin{pmatrix}
+\begin{bmatrix}
+0\\
+0
+\end{bmatrix}
+,\boldsymbol{S} \nonumber
+\end{pmatrix}\\
+\boldsymbol{S} = 
+\begin{pmatrix}
+\sigma_{\alpha_j} & 0 \\
+0 & \sigma_{\beta_j} \nonumber
+\end{pmatrix} \boldsymbol{R} \begin{pmatrix}
+\sigma_{\alpha_j} & 0 \\
+0 & \sigma_{\beta_j} \nonumber
+\end{pmatrix}\\
+\sigma_{\alpha_j} \sim 
+\text{HalfCauchy}(0,1) \nonumber\\
+\sigma_{\beta_j} \sim 
+\text{HalfCauchy}(0,1) \nonumber\\
+\boldsymbol{R}\sim \text{LKJcorr}(1) \nonumber\\
+\sigma  \sim  \operatorname{HalfStudent T}^{+}(3, 0, 2.5) 
+\end{aligned}
+$$
+\newpage
 
-Two fundamental question arise from the patterns we observe in this study.
-The first question is why those who believed in Spirit or Life Force were initially lower in their personal well-being and life-satisfaction in the initial years of this study in comparison to religious and non-religious skeptics?
-During the early years of New Zealand Attitudes and Values data collection New Zealand was emerging from the twin impacts of a global financial recession and a series of devastating earthquakes that destroyed 30% of Christchurch, New Zealand's second largest city [@greaves2015regional].
-Orthogonal research observes that economic and ecological distress are associated with increases in traditional religious beliefs [@sibley2012faith].
-Experimental research has found among religious people death anxiety provokes acceptance of non-traditional religious beliefs [@norenzayan2006belief].
-Because spiritual belief items were only introduced during the year of the earthquakes, we cannot here resolve whether the shifting relationship between spirit beliefs and well-being were the result of demand for spiritual beliefs among those secular and religious people who were adversely affected by the financial crisis or natural disaster.
-On the other hand, such exogenous influences might have adversely affected those with Spirit/Life force beliefs in the early phases of this study, and the patterns observed here are better understood as a regression to a population mean.
-Though beyond the scope of this study, how subjective well-being changes in response to changes in spiritual beliefs, as well as special sensitivities to distress among people with spiritual beliefs are important questions for future investigations.
-Despite the longitudinal quality of this study, a causal explanation remains elusive.
+## Deterministic model
 
-Second, our task here has been to clarify whether beliefs predict growth in well-being.
-However, as we have indicated at the outset, the term "spirituality" applies to a family of concepts.
-We have not attempted to model the diverse philosophical, social, and behavioral attributes within people who adhere to spirit beliefs.
-Moreover, people might identify as spiritual without having any accompanying belief in a Spirit or Life force.
-The task of clarifying how distinct features across a diverse family of "spirituality" concepts affect perceptions of individual well-being remain intriguing horizons.
-
-## Importance
-Despite its limitations, our study is important both for its methods and its findings.
-We demonstrate that spirituality can be operationalised to avoid the pitfalls of tautology evident in previous research by focusing on states of beliefs, that concepts of such belief states can be refined to enable distinctions necessary to investigate spirituality in mixed secular and religious populations, and that the question of whether beliefs states predict growth in subjective well-being can be investigated in a longitudinal setting.
-Our finding that, indeed, spiritual beliefs predict growth in perceived well-being and life satisfaction may hold practical importance.
-Previous research indicates that religious beliefs are associated with psychological well-being [@Smith2003-re].
-Western populations are increasingly identifying as "spiritual but not religious" [@Zinnbauer2005-vz; @Swinton2001-vr].
-Here, we observe that even among people who lack a traditional religious belief in God, believing in a Spirit or Life force predicts incremental improvements in well-being that sum to practically large effects.
-Though the contextual and psychological mechanisms that underpin such growth will require many decades of research, this preliminary finding is good news for those increasing numbers of individuals continue to report spiritual beliefs but who do not affirm traditional religious beliefs.
-
-### Authors contributions
-BH conceived of the study.
-BH wrote the first draft of the manuscript, and performed the initial data analysis with input from JB.
-EW improved the approach with BH and JB and revised the manuscript.
-DD revised the manuscript.
-CS was responsible for NZAVS data collection and revised the manuscript.
-JB improved the approach with BH and EW, performed a data re-analysis, and revised the manuscript.
-
-### Funding statement
-The New Zealand Attitudes and Values Study is supported by a grant from the Templeton Religion Trust (TRT0196). The funders had no role in the preparation of the article or the decision to publish.
-
+The deterministic portion of both the Personal Wellbeing and the Life Satisfaction models is as follows: 
+$$
+\begin{aligned}
+& \quad  \boldsymbol{b} = \\
+& \quad \beta_{1}(\operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\
+&\quad \beta_{2}(\operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ +  \beta_{3}(\operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}})\ +\\
+& \quad \beta_{4}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\ 
+& \quad \beta_{4}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ + \\
+&\quad \beta_{6}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) \\ 
+\end{aligned}
+$$
 \newpage

### 2020-11-25 Supplement_And_Analysis.Rmd
bayesian analysis and supplement.
functions and library file to simplify

---
title             : "Supplement:Analysis for Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Belief"
shorttitle        : "Supplement 1"
author: 
  - name          : "Benjamin Highland"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "520 Power Plant Circle Apartment 105 Winston-Salem, NC 27101 USA"
    email         : "benhighland6420@gmail.com"
  - name          : "Everett L. Worthington"
    affiliation   : "2"
    institution   : "Department of Psychology Virginia Commonwealth University"
  - name          : "Don E Davis"
    affiliation   : "3"
    institution   : "Department of Psychology Georgia State University"
  - name          : "Chris G. Sibley"
    affiliation   : "4"
    institution   : "School of Psychology University of Auckland"
  - name          : "Joseph A. Bulbulia"
    affiliation   : "5"
    institution   : "School of Psychology Victoria University"

authornote: |
  1 Wake Forest School of Medicine 2 Department of Psychology Virginia Commonwealth University 3 Department of Psychology Georgia State University 4 School of Psychology University of Auckland 5 School of Psychology Victoria University

  <!-- https://tinyurl.com/ybremelq -->
# keywords          : "keywords"
wordcount         : no
bibliography      : ["b-references.bib"]
floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : no
mask              : no
draft             : no
# documentclass     : "apa6"
classoption       : "man"
output            : "html_document" 
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
library("papaja")
r_refs("r-references.bib")
```

```{r global_options, echo=FALSE, include = FALSE}
knitr::opts_chunk$set(fig.path="Figs/", 
                      message=FALSE, 
                      warning=FALSE,
                      collapse =TRUE,
                      echo=TRUE, 
                      #results="hide", 
                      fig.width= 10,
                      fig.height=8,
                      tidy = "styler")
```

This supplement offers a step-by-step description our full analysis of the "Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Belief."  This analysis can be found on github [https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing](https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing)

First we load our functions. We use functions to automate repetative coding, to avoid errors from repetation, and to make the data analysis easier to follow.

```{r load_libraries, echo=TRUE, include=FALSE}
source("libs.R")
source("funs.R")
```

We need to give credit to those who have developed the packages we are using. This script returns a bibliography file.
```{r, sessioninfo, include=FALSE}
#  custom function for citing packages. 
cite_packages()
```

This reads the data into my machine. Note that the NZAVS is an ongoing longitudinal study. Our ethics requires us to protect the privacy of our participants. For this reason, we cannot post the raw data. However, we provide data to qualified researchers for the purposes of replication. Additionally we are interested in collaboration with qualified researchers. For data requests, please contact Joseph Bulbulia [joseph.bulbulia@gmail.com](joseph.bulbulia@gmail.com), Chris G. Sibley  [c.sibley@auckland.ac.nz](c.sibley@auckland.ac.nz).

```{r, echo = TRUE}
# custom function for my machine to read data. you'll need to read data to your machine. tweak function in the `funs.R' script for your machine
d<-data_read_mymachine()

# function to clean data, specify number of waves as 3 
d_3<-data_clean_spirit_wellbeing(d,3) # d = dataframe, 3 = number of waves
# function to count uinque ides
show_unique_id(d_3$Id) 
# function to count waves for participants
count_waves_participants(d_3)
# table of beliefs 
table(d_3$Beliefs)
# years Note that we use a continuous response measure particpants. 
range(d_3$years)
```

This code will create a demographic table.

```{r, include=FALSE}
# must remove Furniture package before running table 1
demographic_table(d_3)
```

This table shows the  belief responses over the 9 years.  
```{r, include=FALSE}
beliefs_table(d_3)
```

We wrote the publication in \LaTeX. For the data table in LaTeX run this code.

```{r latex_data_table, include=FALSE}
# you might need to install the dev verion.//remotes::install_github("tysonstanley/furniture")
demographic_table_latex(d_3)
```

We select a minimum of three waves because we want to recover a longitudinal sample in which a slope for the effect of Beliefs might change (or not). However, we want to consider whether our selection of three waves is sensitive to that number. So we also prepare 4-7 years of data frames for separate analyses, and we check the n's for each of those datasets.

```{r, include = FALSE}
# create datasets to test robustness to years
d_4<-data_clean_spirit_wellbeing(d,4) # 4 years
d_5<-data_clean_spirit_wellbeing(d,5) # 5 years
d_6<-data_clean_spirit_wellbeing(d,6) # 6 years
d_7<-data_clean_spirit_wellbeing(d,7) # 7 years
d_8<-data_clean_spirit_wellbeing(d,8) # 7 years

# function to count uinque number of ID's
show_unique_id(d_3$Id) 
show_unique_id(d_4$Id)
show_unique_id(d_5$Id) 
show_unique_id(d_6$Id) 
show_unique_id(d_7$Id) 
show_unique_id(d_8$Id)

# function to count waves for participants
count_waves_participants(d_3)
count_waves_participants(d_4)
count_waves_participants(d_5)
count_waves_participants(d_6)
count_waves_participants(d_7)
count_waves_participants(d_8)
count_waves_participants(d_9)
```

In the first set of analyses we included covariates to adjust for potential confounding effects from usual suspect factors associated with our two measures of wellbeing that might explain the relationship of spirit-beliefs to well-being. We initially included a test of Variance Inflation Factors to rule out mult-collinearity. However, upon further reading, we became aware of at least three other types of confounding that might arise from longitudinal analysis (see: [@mcelreath2020statistical]). We report our controlled analysis below, which does not change the inference in this study. However, because adding covariates can *open* a relationship between beliefs and wellbeing we have switched to reporting the simple time series model in the main text. Additionally, we are careful to state that the present student cannot infer a causal relationshp between spirit beliefs and wellbeing. Rather, we can only infer how the expected relationship between beliefs and wellbeing changes between those who differ in such beliefs over time. A causal understanding inference requires testing specific mechanistic theories. This is a matter for future research.

First we use a maximum Likelihood estimator. Such an estimator is faster than our preferred bayesian estimator.  We consider whether the slope of beliefs changes as a function of time, or equivalently whether the time effect differs depending on one's leve of belief. We also create a table for the two analyses (personal wellbeing and lifesatisfaction.)

```{r robust to years and beliefs only, cache = TRUE, dependson=dat_3,cache =TRUE}
library(lme4)
pwi_formula_simple <- 'PWI ~  years * Beliefs + (1+ Beliefs|Id)'
ls_formula_simple <- 'LIFESAT ~  years * Beliefs + (1+ Beliefs|Id)'
mod.1aa <-lmer(pwi_formula_simple, data= d_3)
mod.2aa <-lmer(ls_formula_simple, data= d_3)

# table
sjPlot::tab_model(mod.1aa,mod.2aa)
```

We next produce a predictive graph for beliefs by time on PWI. Note, again, the function for producing this graph is in the "funs.R" file.

```{r prediction plot of ML longitudinal model, cache =TRUE, dependson=mod.1aa}
```{r bayesian_pwi_predict, cache =TRUE}
graph_predictions(mod.1aa, "Predicted Values of Personal Wellbeing")
```

We next produce a predictive graph for beliefs by time on life satisfaction. 
```{r bayesian_lifesat_predict, cache =TRUE}
graph_predictions(mod.2aa, "Predicted Values of Life Satisfaction")
```

For both outcomes, the slope for beliefs in a spirit or life fource predicts a greater increase in  wellbeing.

Next we attempt a model with that is adjusted by "usual suspect" demographic inicators. We find that the relationship is a little stronger. Because these strengthened relationship might owe to confounding, we are wearing of the adjusted model.  The idea that skeptics are not increasing in well-being, whereas other groups are increasing strikes us as improbable. We do not find evidence for a such a stall in the bayesian analysis; nor do we find any such evidence for a stall in the multiply-imputed dataset analysis (which handles missing responses).  Ultimately we favour, and report the bayesian analysis because bayesian data analysis recovers an explicit estimate of posterior uncertianty.

```{r lmer_model_with_full_covariates, cache =TRUE}
library(lme4)
pwi_formula <- 'PWI ~  years * Beliefs +  Age_in_Decades_C + Deprivation_S + Education_S  + Employed + Ethnic_Categories + Male  + Has_Partner  + Political_Conservativism_S + Urban + (1 + Beliefs|Id)'
ls_formula <- 'LIFESAT ~  years * Beliefs +  Age_in_Decades_C + Deprivation_S + Education_S  + Employed + Ethnic_Categories + Male  + Has_Partner  + Political_Conservativism_S + Urban + (1 + Beliefs|Id)'

mod.1 <-lmer (pwi_formula, data= d_3)
mod.2 <-lmer (ls_formula, data= d_3)
```

Here is the table.
```{r saturated model, cache = TRUE, dependson=d_3}
sjPlot::tab_model(mod.1,mod.2)
```

Compare complex with simple models, and we find the inference doesn't change. 
```{r}
# identical result
sjPlot::tab_model(mod.1,mod.1aa)
sjPlot::tab_model(mod.2,mod.2aa)
```

Next we check for multi-collinearity.

```{r check collinearity, cache =TRUE}
# Note High VIFS from 
performance::check_collinearity(mod.1)
performance::check_collinearity(mod.2)
```

We observe that VIFs are high, for the years x beliefs interaction. However, this is likely to be an artifact of how we code the variables (we do not center by default), but to make sure let's center the years indicator and run the models again. We we do this we find that collinearity disappears, and without any effect on inference.

```{r lmer model centering years}
library(lme4)
pwi_formulaC <- 'PWI ~  yearsC * Beliefs +  Age_in_Decades_C + Deprivation_S + Education_S  + Employed + Ethnic_Categories + Male  + Has_Partner  + Political_Conservativism_S + Urban + (1 |Id)'

ls_formulaC <- 'LIFESAT ~  yearsC * Beliefs +  Age_in_Decades_C + Deprivation_S + Education_S  + Employed + Ethnic_Categories + Male  + Has_Partner  + Political_Conservativism_S + Urban + (1 |Id)'

mod.1a <-lmer (pwi_formulaC, data= d_3)
mod.2a <-lmer (ls_formulaC, data= d_3)

summary(mod.1a)
summary(mod.2a)
# identical result
performance::check_collinearity(mod.1a)
performance::check_collinearity(mod.2a)
```

Here's the table, again, inference is robust.

```{r table of saturated models, cached =TRUE}
sjPlot::tab_model(mod.1a,mod.2a)
```

Her are thecoefficient plots for the full models
```{r coefficientplots_LMER_FULL, cache=TRUE}
plot_models(mod.2, mod.1) + ylim(-.4, .75) + theme_abyss() + scale_colour_brewer(labels = c("Personal Wellbeing", "Life Satisfaction")) # Plots
```

Here is the code for the coefficient plots of our preferred simple model
```{r cache =TRUE}
g1<-plot(parameters::model_parameters(mod.1aa))+ ggtitle("Personal Wellbeing Coefficients")
g2<-plot(parameters::model_parameters(mod.2aa))+ ggtitle("Life Satisfaction Coefficients")
```

And here's the graph.

```{r plot_coefficientplots_LMER_FULL, cache =TRUE}
g1/g2
```

Next, recall we want to check whether inference is  robustness to the researcher degree of freedom in selecting the number of waves.
```{r run all the models with different ns, cache=TRUE}
mod.1_3 <-lmer (pwi_formula_simple, data= d_4)
mod.2_3 <-lmer (ls_formula_simple, data= d_4)

mod.1_4 <-lmer (pwi_formula_simple, data= d_4)
mod.2_4 <-lmer (ls_formula_simple, data= d_4)

mod.1_5 <-lmer (pwi_formula_simple, data= d_5)
mod.2_5 <-lmer (ls_formula_simple, data= d_5)

mod.1_6 <-lmer (pwi_formula_simple, data= d_6)
mod.2_6 <-lmer (ls_formula_simple, data= d_6)

mod.1_7 <-lmer (pwi_formula_simple, data= d_7)
mod.2_7 <-lmer (ls_formula_simple, data= d_7)

mod.1_8 <-lmer (pwi_formula_simple, data= d_8)
mod.2_8 <-lmer (ls_formula_simple, data= d_8)
```

Let's plot all the models.  We we do so, we find that uncertainty increases, but that the inference remains robust.

Let's graph the extremes of the degree-of-freedom choice to be certain that the inference remains satisfying to the eye (and it is.)

```{r extremes_of_researcher_degree_of_freedom_test, cache =TRUE}
library(parameters)
g1a_3<-plot(parameters::model_parameters(mod.1_3))+ ggtitle("Personal Wellbeing Coefficients: 8 Waves")
g2a_3<-plot(parameters::model_parameters(mod.2_3))+ ggtitle("Life Satisfaction Coefficients: 8 Waves")

g1a_8<-plot(parameters::model_parameters(mod.1_8))+ ggtitle("Personal Wellbeing Coefficients: 8 Waves")
g2a_8<-plot(parameters::model_parameters(mod.2_8))+ ggtitle("Life Satisfaction Coefficients: 8 Waves")
```

Estimates robust at 8 waves
```{r USE_2_coefficientplots_LMER_FULL,cache =TRUE}
library(patchwork)
g1a_3/g1a_8
```

Finally we model the relationshp bewteen having a partner and deprivation, to address the reviewer's worry that we might have introduced correlation of between deprivation and having a partner that could spoil inference. We indeed find such a relationship. This is an important potential confound. Again we emphasise that in a longitudinal setting the reviewers concern is even greater, as we might open a path to the outcome variable by including the demographic indicator.

```{r checkforref model,cache =TRUE}
m<-lmer (NZdep ~ Partner + (1|Id), data= d_3, na.action = na.omit)
tab_model(m)
model_parameters(m)
```

# Bayesian estimation

Bayesian inference is time intensive. However, a bayesian model will pool information to produce a better, and natural estimate of uncertainty. The posterior distributions of the model describe conditional probabilities. So where we can we want to use bayesian models. Here we use the default weakly informative priors that the BRMS package produces. It is generally better to use informative priors. However, the size or our dataset does not make a difference. We report the priors at the end.

```{r setmodelformula}
forumula_bayesPWI <- bf(PWI ~ years * Beliefs + (1 + Beliefs|Id))
forumula_bayesLIFESAT <- bf(LIFESAT ~  years * Beliefs + (1 + Beliefs|Id))
```
```{r pwi_bayesmodel}
# PWI
mod.1_b <-brm(forumula_bayesPWI, data= d_3, file ="mod.1_b.rds") # note these are very large files
summary(mod.1_b)
#saveRDS(mod.1_b,"backup_mod.1_b") this is excessive, but we save another backup to avoid running this model again in case of SNAFU.
```
```{r lsbayesmodel}
# LIFE SAT
mod.2_b <-brm (forumula_bayesLIFESAT, data= d_3,cores = 8, file ="mod.2_b.rds") # note these are very large files
# saveRDS(mod.2_b,"backup_mod.2_b") this is excessive, but we save another backup to avoid running this model again in case of SNAFU.
```

Tables
```{r lifesat bayesian model table:pwi, cache =TRUE}
tab_model(mod.1_b)
```

```{r lifesat bayesian model table:ls lifesat, cache =TRUE}
tab_model(mod.2_b)
```

Here we extract the information about the priors

```{r stancode for models}
#model equation
stancode(mod.1_b)
stancode(mod.2_b)
```

```{r prior summaries}
brms::prior_summary(mod.1_b)
brms::prior_summary(mod.2_b)
```

```{r bayesian_model_table, cache=TRUE}
#tab_model(mod.1_b, show.ci= .89) # we really should use something other than 95 but the team consensus is to use 95
tab_model(mod.1_b, mod.1_b show.ci= .95) # no difference to inference
```

```{r mixing of sd_pwi}
plot(mod.1_b, pars ="b_") #requires stronger priors
```

```{r mixing of ls_pwi}
plot(mod.1_b, pars ="sd_") # requires stronger priors
```

```{r posterior predictive checks bayes:pwi, cache=TRUE}

pp_1<-NULL
pp_1<-rstanarm::pp_check(mod.1_b) + ylab("Personal Wellbeing")

pp_2<-NULL
pp_2<-rstanarm::pp_check(mod.2_b) + ylab("Life Satisfaction")

# plot
pp_1 + pp_2
```

Here are the prediction plots that we use in the main publication text.

Prediction plots
```{r bayesian_pwi_predicted, cache=TRUE, dependson= mod.1_b}
graph_predictions(mod.1_b, "Predicted Values of Personal Wellbeing")
```
```{r bayesian_lifesat_predicted, cache=TRUE, dependson= mod.2_b}
graph_predictions(mod.2_b, "Predicted Values of Life Satisfaction")
```

## Imputed Analysis

Finally we run the data analysis to test whether our results are robust to missing data. Missing responses can be an important source of bias in a longitudinal study. Again we write some functions to deal with all the work. I have saved the imputed files and will read them back to for analysis.

```{r echo = TRUE,  cache = TRUE, dependson = "dfben"}
# Imputed data
# imputed3<-amelia_imputation_clean(d_3)
# saveRDS(imputed3,"imputed3")
imputed3<-readRDS(imputed3)
```

We rerun the imputed dat analysis.

```{r cache = TRUE, imputedsimplePWI}
# run models iterating over imputed data
mod_imputed_pwi<-loop_lmer_model(pwi_formula_simple,imputed3)
```

```{r imputed pwi table, cache=TRUE}
tab_mods_ls<-loop_lmer_model_tab(mod_imputed_ls)
create_hux_table(tab_mods_ls)
```

We graph the coefficients.
```{r coef_PWI_imputed coefplot, cache =TRUE}
# coeff plot
imp_pwi <- plot(tab_mods_pwi) + ggtitle("Personal Wellbeing Coefficients (Imputed)") # pick your  own title
imp_pwi
```

We create a graph for the predictions.
```{r r coef_PWI_imputed coefplot, cache=TRUE}
# function for prediction
graph_predictions_imputed(mod_imputed_pwi,"Predicted Values of Personal Wellbeing (imputed)")
```

The inference is robust in the imputation model.

## Life Satisfaction Imputation Model

The following code returns the analysis for the life statisfaction model (again the functions are in the "funs.R" file).

```{r cache = TRUE}
# run models iterating over imputed data
mod_imputed_ls<-loop_lmer_model(ls_formula_simple,imputed3)

```{r table_ls_imputed,cache=TRUE}
# table
tab_mods_ls<-loop_lmer_model_tab(mod_imputed_ls)
create_hux_table(tab_mods_ls)
```

Here's the graph:

```{r coef_PWI_imputed, cache = TRUE}
# coeff plot
imp_ls<-plot(tab_mods_ls) + ggtitle("Life Satisfaction Coefficients (Imputed)") 
imp_ls
```

```{r pred_LS_imputed,  cache = TRUE}
# function for prediction
graph_predictions_imputed(mod_imputed_ls,"Predicted Values of Life Satisfaction (imputed)")
```

```{r coefficient_plot_both_imputed,  cache = TRUE}
 imp_pwi/imp_ls +  plot_annotation(tag_levels = 'i')
```

## Bayesian Multiple imputation models

We want to run a bayesian MI model. However, we we do this, we blow up the RAM in JB's machine.  So we must satisfy ourselves with a frequentist test here.

<!-- ```{r} won't work for missing categorical predictors--> 
<!-- form_pwi <- bf(PWI | mi() ~ years * mi(Beliefs)) + -->
<!--   bf(Beliefs | mi() ~ years) + set_rescor(FALSE) -->
<!-- imp_bayes_pwi <- brm(form_pwi, data = d_3) -->
<!-- ``` -->

```{r create data frame of lists for BRMS imputation}
# brms requires a list of the imputations
brmdat <-list(imputed3$imputations$imp1,
              imputed3$imputations$imp2,
              imputed3$imputations$imp3,
              imputed3$imputations$imp4,
              imputed3$imputations$imp5,
              imputed3$imputations$imp6,  
              imputed3$imputations$imp7,
              imputed3$imputations$imp8,
              imputed3$imputations$imp9,
              imputed3$imputations$imp10)

```

<!-- ```{r bayesian imputed model_pwi} -->
<!-- # not run -->
<!-- #fit_imp1 <- brm_multiple(PWI ~ years * Beliefs + (1 + years|Id), data = brmdat) -->
<!-- #summary(fit_imp1) -->
<!-- #save(fit_imp1) -->
<!-- ``` -->

<!-- ```{r bayesian imputation model_ls} -->
<!-- # not run -->
<!-- #fit_imp2 <- brm_multiple(LIFESAT ~ years * Beliefs + (1+years|Id), data = brmdat, file = "fit_imp2") -->
<!-- ``` -->

In the commends of the "Analysis.Rmd" file, there are further analyeses that we ran (e.g. within/between subject effects). None affected inference.

# model equations
```{r}
stancode(mod.1_b)
```

\newpage
## Priors for the Personal Wellbeing model

Model priors for Personal Wellbeing including group-level ID intercept and beliefs slopes:

\[
\begin{aligned}
\operatorname{Personal Well Being}_{ij} \sim \text{Normal}(\mu_{ij},\sigma^2) \nonumber\\
\mu_{ij}\sim \boldsymbol{\alpha +\beta} \nonumber\\
\boldsymbol{\alpha} \sim \alpha_{0} +\alpha_j \nonumber\\
\alpha_{0} \sim \text{StudentT}(3,7.2,2.5)\nonumber\\
\boldsymbol{\beta}\sim \boldsymbol{b} + \beta_j \nonumber\\
\boldsymbol{b}\sim\text{Normal}(0,2.5) \nonumber\\
\begin{bmatrix}
\alpha_j \\
\beta_j
\end{bmatrix}
\sim 
\text{MVNormal}
\begin{pmatrix}
\begin{bmatrix}
0\\
0
\end{bmatrix}
,\boldsymbol{S} \nonumber
\end{pmatrix}\\
\boldsymbol{S} = 
\begin{pmatrix}
0 & \sigma_{\beta_j} \nonumber
\end{pmatrix} \boldsymbol{R} \begin{pmatrix}
\sigma_{\alpha_j} & 0 \\
0 & \sigma_{\beta_j} \nonumber
\end{pmatrix}\\
\sigma_{\alpha_j} \sim 
\text{HalfCauchy}(0,2.5) \nonumber\\
\sigma_{\beta_j} \sim 
\text{HalfCauchy}(0,2.5) \nonumber\\
\boldsymbol{R}\sim \text{LKJcorr}(1) \nonumber\\
\sigma  \sim  \operatorname{HalfStudent T}^{+}(3, 0, 2.5) 
\end{aligned}
\]

\newpage
## Priors for the Life Satisfaction model:
Model priors for Life Satisfaction Model with including group-level ID intercept and belief slopes:

\[
\begin{aligned}
\operatorname{Life Satisfaction}_{ij} \sim \text{Normal}(\mu_{ij},\sigma^2) \nonumber\\
\mu_{ij}\sim \boldsymbol{\alpha +\beta} \nonumber\\
\boldsymbol{\alpha} \sim \alpha_{0} +\alpha_j \nonumber\\
\alpha_{0} \sim \text{StudentT}(3,5.5, 2.5)\nonumber\\
\alpha_j \sim \text{StudentT}+(3,0,10)\nonumber\\
\boldsymbol{\beta}\sim \boldsymbol{b} + \beta_j \nonumber\\
\boldsymbol{b}\sim\text{Normal}(0,2.5) \nonumber\\
\begin{bmatrix}
\alpha_j \\
\beta_j
\end{bmatrix}
\sim 
\text{MVNormal}
\begin{pmatrix}
\begin{bmatrix}
0\\
0
\end{bmatrix}
,\boldsymbol{S} \nonumber
\end{pmatrix}\\
\boldsymbol{S} = 
\begin{pmatrix}
\sigma_{\alpha_j} & 0 \\
0 & \sigma_{\beta_j} \nonumber
\end{pmatrix} \boldsymbol{R} \begin{pmatrix}
\sigma_{\alpha_j} & 0 \\
0 & \sigma_{\beta_j} \nonumber
\end{pmatrix}\\
\sigma_{\alpha_j} \sim 
\text{HalfCauchy}(0,1) \nonumber\\
\sigma_{\beta_j} \sim 
\text{HalfCauchy}(0,1) \nonumber\\
\boldsymbol{R}\sim \text{LKJcorr}(1) \nonumber\\
\sigma  \sim  \operatorname{HalfStudent T}^{+}(3, 0, 2.5) 
\end{aligned}
\]

\newpage
## Deterministic model

For the deterministic part of both the Personal Wellbeing and the Life Satisfaction models:

\[
\begin{aligned}
& \quad  \boldsymbol{b} = \\
& \quad \beta_{1}(\operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\
&\quad \beta_{2}(\operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ +  \beta_{3}(\operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}})\ +\\
& \quad \beta_{4}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\ 
& \quad \beta_{4}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ + \\
&\quad \beta_{6}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) \\ 
\end{aligned}
\]

<!-- Frequentist model: -->

<!-- ```{r eq1} -->
<!-- library(equatiomatic) -->
<!-- summary(mod.1aa) -->
<!-- print(extract_eq(mod.1aa, wrap = TRUE)) -->
<!-- ``` -->

<!-- ```{r eq2} -->
<!-- library(equatiomatic) -->
<!-- print(extract_eq(mod.1aa, wrap = TRUE, use_coefs=TRUE,terms_per_line = 3)) -->
<!-- ``` -->

<!-- We might want to look at within/between indivual change (i.e. do people who change in degree of spirit beliefs manifest different wellbeing.  I ran this as below, however, what we really want is a markov model, where we compare state changes between the different types of belief. So this model is not useful. For more on demeaning see:  -->
<!-- https://easystats.github.io/parameters/articles/demean.html -->

<!-- ```{r} -->
<!-- dfben <- cbind( -->
<!--   dfben, -->
<!--   parameters::demean(dfben, select = c("Beliefs"), group = "Id") -->
<!-- ) -->
<!-- ``` -->
<!-- ```{r models, cache = TRUE, dependson = "dfben"} -->
<!-- ### Spirit Belief Models ###  -->
<!-- mod01<-lmer (PWI ~ Years + -->
<!--                 Beliefs__Skeptic__between  +   -->
<!--                 Beliefs__SpiritExcludesGod__between +  -->
<!--                 Beliefs_GodAndSpirit_between  +  -->
<!--                 Beliefs_GodExcludesSpirit_between + -->
<!--                 Beliefs__Skeptic__within  +   -->
<!--                 Beliefs__SpiritExcludesGod__within +  -->
<!--                 Beliefs_GodAndSpirit_within  +  -->
<!--                 Beliefs_GodExcludesSpirit_within + -->
<!--                 (1|Id), data= dfben, na.action = na.omit) -->

<!-- parameters::check_heterogeneity(mod01, group = "ID") -->
<!-- parameters::check_heterogeneity(mod01, group = "ID") -->

<!-- mod02<-lmer (LIFESAT ~ Years + -->
<!--                 Beliefs__Skeptic__between  +   -->
<!--                 Beliefs__SpiritExcludesGod__between +  -->
<!--                 Beliefs_GodAndSpirit_between  +  -->
<!--                 Beliefs_GodExcludesSpirit_between + -->
<!--                 Beliefs__Skeptic__within  +   -->
<!--                 Beliefs__SpiritExcludesGod__within +  -->
<!--                 Beliefs_GodAndSpirit_within  +  -->
<!--                 Beliefs_GodExcludesSpirit_within + -->
<!--                 (1|Id), data= dfben, na.action = na.omit) -->
<!-- parameters::check_heterogeneity(mod02, group = "ID") -->
<!-- # inference holds, but can we use this method with factors? What do the contrasts mean?  -->
<!-- tab_model(mod01) -->
<!-- tab_model(mod02) -->
<!-- plot_models(mod01,mod02) -->
<!-- ``` -->

<!-- ```{r eq1} -->
<!-- #for model equation -->
<!-- library(equatiomatic) -->
<!-- print(extract_eq(mod.1, wrap = TRUE,terms_per_line =3)) -->
<!-- ``` -->

<!-- PWI Model 2 equation -->
<!-- ```{r eq2} -->
<!-- library(equatiomatic) -->
<!-- print(extract_eq(mod.2, wrap = TRUE, use_coefs=TRUE,terms_per_line = 3)) -->
<!-- ``` -->
<!-- ## report -->
<!-- ### Model 1 -->
<!-- model 1 report: for publication change all "significants" to "reliable" -->

<!-- short report: -->
<!-- ```{r, report1, cache =TRUE, dependson=mod.1,include=FALSE} -->
<!-- easystats::model_text(mod.1) -->
<!-- ``` -->

<!-- short report -->
<!-- ```{r, cache=TRUE, include=FALSE } -->
<!-- library(report) -->
<!-- r1<- easystats::report(mod.1) -->
<!-- easystats::table_short(r1) -->
<!-- easystats::text_short(r1) -->
<!-- ``` -->

<!-- ```{r, cache =TRUE, dependson=mod.1, include=FALSE} -->
<!-- r1<- easystats::report(mod.1) -->
<!-- easystats::text_long(r1) -->
<!-- ``` -->
<!-- short report -->
<!-- ```{r, cache =TRUE, dependson=mod.2,include=FALSE} -->
<!-- # Model 2 -->
<!-- # For publication change all "significants" to "reliable" -->

<!-- easystats::model_text(mod.2) -->
<!-- ``` -->

<!-- short report -->
<!-- ```{r,include=FALSE} -->
<!-- r2 <- easystats::report(mod.2) -->
<!-- easystats::text_short(r2) -->
<!-- easystats::table_short(r2) -->
<!-- ``` -->

<!-- long report -->
<!-- ```{r,include=FALSE} -->
<!-- r2<- easystats::report(mod.2) -->
<!-- easystats::text_long(r2) -->
<!-- ``` -->
<!-- ## Equation 1: Personal Well-being  -->
<!-- model equations for models without random intercept slope correlations; not correct, or reported\[ -->
<!-- \begin{aligned} -->
<!-- \operatorname{{\small PWI}}  &\sim N \left(\mu,\sigma^2 \right) \\  -->
<!-- \mu & =  \boldsymbol{a} +  \boldsymbol{b} \\ -->
<!-- & \quad \boldsymbol{a} =  \alpha_{0} + \alpha_{j[i]} \\ -->
<!-- & \quad \alpha_{0} \sim \operatorname{Student T}^{+}(3,7.2,2.5) \\ -->
<!-- & \quad  \alpha_{j[i]}  \sim \operatorname{Normal}(0, \sigma_j) \\  -->
<!-- & \quad \sigma_j \sim \operatorname{HalfStudent T}^{+}(3, 0, 2.5) \\ -->
<!-- & \quad   \boldsymbol{b}   \sim \operatorname{Student T}^{+}(0, 2.5) \\  -->
<!-- & \quad \sigma  \sim  \operatorname{HalfStudent T}^{+}(3, 0, 2.5) \\ -->
<!-- & \quad  \boldsymbol{b} = \\ -->
<!-- & \quad \beta_{2}(\operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\ -->
<!-- &\quad \beta_{3}(\operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ +  \beta_{4}(\operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}})\ +\\ -->
<!-- & \quad \beta_{5}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\  -->
<!-- & \quad \beta_{17}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ + \\ -->
<!-- &\quad \beta_{18}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) \\  -->
<!-- \end{aligned} -->
<!-- \] -->
<!-- ## Equation 2: Life Satisfaction. -->
<!-- \[ -->
<!-- \begin{aligned} -->
<!-- \operatorname{{\small LifeSatisfaction}}  &\sim N \left(\mu,\sigma^2 \right) \\  -->
<!-- \mu & =  \boldsymbol{a} +  \boldsymbol{b} \\ -->
<!-- & \quad \boldsymbol{a} =  \alpha_{0} + \alpha_{j[i]} \\ -->
<!-- & \quad \alpha_{0} \sim \operatorname{Student T}^{+}(3,5.5, 2.5) \\ -->
<!-- & \quad  \alpha_{j[i]}  \sim \operatorname{Normal}(0, \sigma_j) \\  -->
<!-- & \quad \sigma_j \sim \operatorname{HalfStudent T}^{+}(3, 0, 2.5) \\ -->
<!-- & \quad   \boldsymbol{b}   \sim \operatorname{Student T}^{+}(0, 2.5) \\  -->
<!-- & \quad \sigma  \sim  \operatorname{HalfStudent T}^{+}(3, 0, 2.5) \\ -->
<!-- & \quad  \boldsymbol{b} = \\ -->
<!-- & \quad \beta_{2}(\operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\ -->
<!-- &\quad \beta_{3}(\operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ +  \beta_{4}(\operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}})\ +\\ -->
<!-- & \quad \beta_{5}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\  -->
<!-- & \quad \beta_{17}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ + \\ -->
<!-- &\quad \beta_{18}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) \\  -->
<!-- \end{aligned} -->
<!-- \] -->
<!-- ## Equation 2 -->
<!-- \[ -->
<!-- \begin{aligned} -->
<!-- \operatorname{{\small LIFESAT/PWI}}  &\sim N \left(\mu,\sigma^2 \right) \\ \mu &=\alpha_{j[i]} + \beta_{1}(\operatorname{Years}) + \beta_{2}(\operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\ -->
<!-- &\quad \beta_{3}(\operatorname{Beliefs}_{\operatorname{GodAndSpirit}}) + \beta_{4}(\operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) + \beta_{5}(\operatorname{Age.10yrs.C})\ + \\ -->
<!-- &\quad \beta_{6}(\operatorname{Deprivation.S}) + \beta_{7}(\operatorname{Edu.S}) + \beta_{8}(\operatorname{Employed})\ + \\ -->
<!-- &\quad \beta_{9}(\operatorname{EthnicCats}_{\operatorname{Maori}}) + \beta_{10}(\operatorname{EthnicCats}_{\operatorname{Pacific}}) + \beta_{11}(\operatorname{EthnicCats}_{\operatorname{Asian}})\ + \\ -->
<!-- &\quad \beta_{12}(\operatorname{Male}_{\operatorname{1}}) + \beta_{13}(\operatorname{Partner}_{\operatorname{Has\ }}) + \beta_{14}(\operatorname{Pol.Orient.S})\ + \\ -->
<!-- &\quad \beta_{15}(\operatorname{Urban}) + \beta_{16}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}}) + \beta_{17}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ + \\ -->
<!-- &\quad \beta_{18}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) \\ \alpha_{j} &\sim N \left(\mu_{\alpha_{j}},\sigma^2_{\alpha_{j}} \right) , \operatorname{ for  Id }~j~= 1, \dots~J = 20,979~\operatorname{Individuals} -->
<!-- \end{aligned} -->
<!-- \] -->

\[
\begin{table}[ ht ] 
\centering 
\caption{}\label{}
\scalebox{.6}{
\begin{tabular}{ l c c c c c c c c c }
\toprule
 &   \multicolumn{ 9 }{c}{ Wave }\\ 
  & 2010 & 2011 & 2012 & 2013 & 2014 & 2015 & 2016 & 2017 & 2018 \\ 
 & n = \emph{missing} & n = \emph{missing} & n = 8547 & n = 13472 & n = \emph{missing} & n = 11687 & n = 16048 & n = 14174 & n = 14883 \\ 
 \midrule
Age &   &   &   &   &   &   &   &   &  \\ 
NZdep &   &   &   &   &   &   &   &   &  \\ 
Education &   &   &   &   &   &   &   &   &  \\ 
Employed &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    0 &  (\%) &  (\%) & 2094 (24.5\%) & 2955 (21.9\%) &  (\%) & 2525 (21.6\%) & 3486 (21.7\%) & 3223 (22.7\%) & 3404 (22.9\%)\\ 
\hspace{6pt}    1 &  (\%) &  (\%) & 6453 (75.5\%) & 10517 (78.1\%) &  (\%) & 9162 (78.4\%) & 12562 (78.3\%) & 10951 (77.3\%) & 11479 (77.1\%)\\ 
EthnicCategories &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    Euro &  (\%) &  (\%) & 6731 (78.8\%) & 11024 (81.8\%) &  (\%) & 9671 (82.8\%) & 13389 (83.4\%) & 11814 (83.3\%) & 12427 (83.5\%)\\ 
\hspace{6pt}    Maori &  (\%) &  (\%) & 1188 (13.9\%) & 1608 (11.9\%) &  (\%) & 1341 (11.5\%) & 1749 (10.9\%) & 1620 (11.4\%) & 1677 (11.3\%)\\ 
\hspace{6pt}    Pacific &  (\%) &  (\%) & 284 (3.3\%) & 339 (2.5\%) &  (\%) & 277 (2.4\%) & 331 (2.1\%) & 243 (1.7\%) & 261 (1.8\%)\\ 
\hspace{6pt}    Asian &  (\%) &  (\%) & 344 (4\%) & 501 (3.7\%) &  (\%) & 398 (3.4\%) & 579 (3.6\%) & 497 (3.5\%) & 518 (3.5\%)\\ 
Male &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    0 &  (\%) &  (\%) & 5302 (62\%) & 8457 (62.8\%) &  (\%) & 7325 (62.7\%) & 10148 (63.2\%) & 9011 (63.6\%) & 9433 (63.4\%)\\ 
\hspace{6pt}    1 &  (\%) &  (\%) & 3245 (38\%) & 5015 (37.2\%) &  (\%) & 4362 (37.3\%) & 5900 (36.8\%) & 5163 (36.4\%) & 5450 (36.6\%)\\ 
Partner &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    0 &  (\%) &  (\%) & 2408 (28.2\%) & 3546 (26.3\%) &  (\%) & 2815 (24.1\%) & 3773 (23.5\%) & 3330 (23.5\%) & 3556 (23.9\%)\\ 
\hspace{6pt}    1 &  (\%) &  (\%) & 6139 (71.8\%) & 9926 (73.7\%) &  (\%) & 8872 (75.9\%) & 12275 (76.5\%) & 10844 (76.5\%) & 11327 (76.1\%)\\ 
Pol Orient &   &   &   &   &   &   &   &   &  \\ 
Urban &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    Not\_Urban &  (\%) &  (\%) & 2865 (33.5\%) & 4278 (31.8\%) &  (\%) & 3951 (33.8\%) & 5538 (34.5\%) & 2610 (18.4\%) & 2661 (17.9\%)\\ 
\hspace{6pt}    Urban &  (\%) &  (\%) & 5682 (66.5\%) & 9194 (68.2\%) &  (\%) & 7736 (66.2\%) & 10510 (65.5\%) & 11564 (81.6\%) & 12222 (82.1\%)\\ 
Beliefs &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    \_Skeptic\_ &  (\%) &  (\%) & 2048 (24\%) & 3286 (24.4\%) &  (\%) & 2951 (25.3\%) & 4116 (25.6\%) & 3916 (27.6\%) & 4120 (27.7\%)\\ 
\hspace{6pt}    \_SpiritExcludesGod\_ &  (\%) &  (\%) & 2295 (26.9\%) & 3735 (27.7\%) &  (\%) & 3169 (27.1\%) & 4501 (28\%) & 3918 (27.6\%) & 3970 (26.7\%)\\ 
\hspace{6pt}    GodAndSpirit &  (\%) &  (\%) & 3707 (43.4\%) & 5499 (40.8\%) &  (\%) & 5119 (43.8\%) & 6453 (40.2\%) & 5464 (38.5\%) & 6210 (41.7\%)\\ 
\hspace{6pt}    GodExcludesSpirit &  (\%) &  (\%) & 497 (5.8\%) & 952 (7.1\%) &  (\%) & 448 (3.8\%) & 978 (6.1\%) & 876 (6.2\%) & 583 (3.9\%)\\ 
LIFESAT &   &   &   &   &   &   &   &   &  \\ 
PWI &   &   &   &   &   &   &   &   &  \\ 
\bottomrule
\end{tabular}
}
\end{table}
\]

\[
\begin{table}
\begin{center}
\scalebox{0.6}{
\begin{tabular}{l c c}
\toprule
 & PWI & Life Satisfaction \\
\midrule
(Intercept)                        & $\mathbf{6.63}^{***}$  & $\mathbf{4.94}^{***}$  \\
                                   & $(0.03)$               & $(0.02)$               \\
Years                              & $0.00$                 & $-0.00$                \\
                                   & $(0.00)$               & $(0.00)$               \\
Beliefs\_SpiritExcludesGod\_       & $\mathbf{-0.14}^{***}$ & $\mathbf{-0.08}^{***}$ \\
                                   & $(0.03)$               & $(0.02)$               \\
BeliefsGodAndSpirit                & $\mathbf{-0.16}^{***}$ & $-0.04$                \\
                                   & $(0.03)$               & $(0.02)$               \\
BeliefsGodExcludesSpirit           & $0.02$                 & $0.07$                 \\
                                   & $(0.05)$               & $(0.04)$               \\
Age.10yrs.C                        & $\mathbf{0.08}^{***}$  & $\mathbf{0.09}^{***}$  \\
                                   & $(0.01)$               & $(0.01)$               \\
Deprivation.S                      & $\mathbf{-0.14}^{***}$ & $\mathbf{-0.07}^{***}$ \\
                                   & $(0.01)$               & $(0.01)$               \\
Edu.S                              & $\mathbf{0.12}^{***}$  & $\mathbf{0.06}^{***}$  \\
                                   & $(0.01)$               & $(0.01)$               \\
EmployedEmployed                   & $\mathbf{0.09}^{***}$  & $\mathbf{0.08}^{***}$  \\
                                   & $(0.01)$               & $(0.01)$               \\
EthnicCatsMaori                    & $\mathbf{-0.30}^{***}$ & $\mathbf{-0.06}^{**}$  \\
                                   & $(0.03)$               & $(0.02)$               \\
EthnicCatsPacific                  & $\mathbf{-0.29}^{***}$ & $\mathbf{-0.08}^{*}$   \\
                                   & $(0.05)$               & $(0.04)$               \\
EthnicCatsAsian                    & $-0.08$                & $-0.04$                \\
                                   & $(0.05)$               & $(0.03)$               \\
Male1                              & $\mathbf{-0.11}^{***}$ & $\mathbf{-0.19}^{***}$ \\
                                   & $(0.02)$               & $(0.01)$               \\
PartnerHas Partner                 & $\mathbf{0.61}^{***}$  & $\mathbf{0.42}^{***}$  \\
                                   & $(0.02)$               & $(0.01)$               \\
Pol.Orient.S                       & $\mathbf{0.02}^{***}$  & $\mathbf{0.01}^{**}$   \\
                                   & $(0.01)$               & $(0.00)$               \\
UrbanUrban                         & $\mathbf{-0.02}^{*}$   & $\mathbf{-0.02}^{*}$   \\
                                   & $(0.01)$               & $(0.01)$               \\
Years:Beliefs\_SpiritExcludesGod\_ & $\mathbf{0.02}^{**}$   & $\mathbf{0.02}^{***}$  \\
                                   & $(0.00)$               & $(0.00)$               \\
Years:BeliefsGodAndSpirit          & $\mathbf{0.02}^{***}$  & $\mathbf{0.01}^{***}$  \\
                                   & $(0.00)$               & $(0.00)$               \\
Years:BeliefsGodExcludesSpirit     & $0.00$                 & $-0.00$                \\
                                   & $(0.01)$               & $(0.01)$               \\
\midrule
AIC                                & $248580.18$            & $200599.16$            \\
BIC                                & $248775.08$            & $200793.96$            \\
Log Likelihood                     & $-124269.09$           & $-100278.58$           \\
Num. obs.                          & $79270$                & $78888$                \\
Num. groups: Id                    & $20979$                & $20972$                \\
Var: Id (Intercept)                & $1.78$                 & $0.88$                 \\
Var: Residual                      & $0.74$                 & $0.42$                 \\
\bottomrule
\multicolumn{3}{l}{\scriptsize{$^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$}}
\end{tabular}
}
\caption{Regression Models}
\label{table:coefficients}
\end{center}
\end{table}
\]

\newpage

# References

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

<div id="refs" custom-style="Bibliography"></div>
\endgroup

### 2020-11-25 funs.R

@@ -1,0 +1,309 @@
+# funs.R
+# functions used in this analysis
+data_read_mymachine <- function(){
+  out <-readRDS("~/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/00Bulbulia\ Pubs/2020/ldf.5")
+}
+
+# this code is for citing packages 
+cite_packages <- function() {
+  citation(package = "base", lib.loc = "/Users/josephbulbulia/Dropbox/BIBS")
+toLatex(sessionInfo(), locale = FALSE)
+sapply(names(sessionInfo()$otherPkgs), function(x) print(citation(x), style = "Bibtex"))
+out <- sapply(names(sessionInfo()$otherPkgs),
+              function(x) print(citation(x), style = "Bibtex"))
+print(out)
+}
+
+
+# data clean
+data_clean_spirit_wellbeing <- function(df,y) {  
+ #function for cleaning this dataset for any arbitrary number of waves
+  n_waves =  as.numeric(y)
+  out <- df %>%
+    dplyr::mutate(Beliefs = factor(ifelse(df$Believe.God == "Not Believe God" & df$Believe.Spirit == "Not Believe Spirit", "_Skeptic_", 
+                                          ifelse(df$Believe.God == "Not Believe God" & df$Believe.Spirit == "Believe Spirit", "_SpiritExcludesGod_",
+                                                 ifelse(df$Believe.God == "Believe God" & df$Believe.Spirit == "Believe Spirit","GodAndSpirit","GodExcludesSpirit")))))%>%
+    dplyr::filter(Wave !=2009)%>% # exclude wave 1 (no spirit measures that wave)
+    dplyr::filter(YearMeasured==1)%>%
+    dplyr::group_by(Id) %>% 
+    dplyr::filter(n() > n_waves-1)%>% # select those who have responded to at least 3 waves # 
+    dplyr::filter(n() !=0)%>%
+    dplyr::ungroup(Id)%>%
+    dplyr:: mutate(Years = as.numeric(years))%>%
+    dplyr:: mutate(yearsC = scale(years, center=TRUE,scale=FALSE))%>%
+    dplyr::mutate(Age.10yrs = (Age/10))%>%
+    dplyr::mutate(Religion.Church.Log = log(Religion.Church+ 1))%>%
+    dplyr::mutate(Political_Conservativism_S = scale(Pol.Orient, center = TRUE, scale = TRUE),
[…]

### 2020-11-25 libs.R

@@ -1,0 +1,31 @@
+# libraries and settings
+library("tidyverse")
+library("easystats")
+library("lme4")
+library("table1")
+library("ggplot2")
+library("ggeffects")
+library("gghighlight")
+library("Amelia")
+library("magrittr")
+library("patchwork")
+library("furniture")
+library("table1")
+library("sjPlot")
+library("texreg")
+#library("viridis")
+#library("ggpubr")
+#library("wesanderson")
+library("ggsci")
+library("papaja")
+#library("here")
+library("citr")
+library("here")
+library("huxtable")
+library("brms")
+library("rstan")
+library("rstanarm")
+library("tidybayes")
+rstan_options(auto_write=TRUE)
+options(mc.cores=parallel::detectCores ())
+#theme_set(theme_few())

### 2020-11-25 Manuscript_Bayesian.Rmd
added expectations. need to fix all figures and rewrite results.

@@
 
-```{r runanalysis, include = FALSE}
-source('short.R')
-```
+# ```{r runanalysis, include = FALSE}
+# source('short.R')
+# ```
 The relationship between spirituality and psychological well-being is a matter of enduring fascination [@Ellsworth2010-yu].
@@
 
+```{r fig.cap = "Combination Coefficient Plots for personal well-being and life satisfaction longitudinal growth models"}
+knitr::include_graphics("/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Figs/USE_2_coefficientplots-1.png")
+
 # Discussion
@@
 1.
-At the outset of the study, in 2010, the expected subjective well-being as measured using the personal well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God ($\hat{y}$ =  6.50, se =.03) and lower among those who believed in both a Spirit/Life-Force and a God ($\hat{y}$ =  6.48, se =.03) in comparison both to the skeptical population who neither believed in a God nor a life force ($\hat{y}$ = 6.64,se =.03) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 6.66, se =.05).
-(See Figure 2)  Similarly,the expected subjective well-being as measured using the life satisfaction Inventory is lower among those who believed in Spirit or Life Force but not a God ($\hat{y}$ = 4.86, se =.02) and lower among those who believed in both a Spirit/Life-Force and a God  ($\hat{y}$ = 4.91, se =.02) in comparison both with the skeptical population who neither believed in a God nor a life force ($\hat{y}$ = 4.94,se =.02) and the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 5.02, se =.02) (see Figure 3)
+At the outset of the study, in 2010, the expected subjective well-being as measured using the personal well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God ($\hat{y}$ =  6.78 [6.74, 6.82]) and lower among those who believed in both a Spirit/Life-Force and a God ($\hat{y}$ =  6.75 [6.72, 6.78]) in comparison both to the skeptical population who neither believed in a God nor a life force ($\hat{y}$ = 6.91 [6.86, 6.95]) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 6.88 [6.81, 6.96]).
+(See Figure 2)  Similarly,the expected subjective well-being as measured using the life satisfaction Inventory was lowest in 2010 among those who believed in Spirit or Life Force but not a God ($\hat{y}$ = 5.07 [5.04, 5.10]).
+Life satisfaction was similar among those who believed in both a Spirit/Life-Force and a God  ($\hat{y}$ =  5.11 [5.08, 5.13]) in comparison both with the skeptical population who neither believed in a God nor a life force ($\hat{y}$ = 5.12 [5.08, 5.15]).
+Life satisfaction was highest among the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 5.19 [5.14, 5.25]) (see Figure 3)
 
@@
 However, over time, belief in a Spirit or Life Force predicts a trajectory of growth in personal well-being.
-In 2018, personal well-being among who believe in Spirit or Life Force but not a God had grown to ($\hat{y}$ = 6.67, se =.03, $\Delta \hat{y} = 0.17$).
-Among those who believed in both a Spirit/Life-Force and a God, personal well-being had grown to ($\hat{y} = 6.69, se = 0.02, \Delta \hat{y} = .21$).
-To place this effect in practical perspective, such growth in personal well-being is about twice as large as the expected benefit of being employed and about a third as large as the expected benefit from having a romantic partner.
-By contrast, personal well-being among the skeptical population who neither believed in a God nor a Life Force remained stable across the eight year interval $\hat{y} = 6.66, se = 0.03, \Delta \hat{y} = .02$).
-By the same token, personal well-being among those who believe in a God but are skeptical about a Spirit/Life Force was constant $\hat{y} =  6.70, se = 0.04, \Delta \hat{y} = .04$).
+In 2018, personal well-being among who believe in Spirit or Life Force but not a God had grown to ($\hat{y} = 7.13$ [7.09, 7.16], $\Delta \hat{y} = 0.35$).
+Among those who believed in both a Spirit/Life-Force and a God, personal well-being had grown to ($\hat{y} =  7.14$, [7.11, 7.16], $\Delta \hat{y} = 0.39$).
+By contrast, personal well-being among the skeptical population who neither believed in a God nor a Life Force grew less steeply across the eight year interval $\hat{y} = 7.15 [7.12, 7.19]$, $\Delta \hat{y} = 0.24$).
+By the same token, personal well-being among those who believe in a God but are skeptical about a Spirit/Life Force grew less steeply $\hat{y} = 7.15$, [7.09, 7.21] $\Delta \hat{y} = 0.27$).
 Similarly, over time, belief in a Spirit or Life Force predicts a trajectory of growth in life satisfaction.
-In 2018, life satisfaction among those who believe in Spirit or Life Force but not a God had grown to ($\hat{y} = 5.01, se = 0.02, \Delta \hat{y} = 0.07$).
+In 2018, the expected effect of spiritual beliefs on life satisfaction among those who believe in Spirit or Life Force but not a God had grown to ($\hat{y} =  5.30$ [5.28, 5.33], $\Delta \hat{y} = 0.23$).
 In practical terms, such growth is equivalent to the expected gain in life satisfaction from being employed and a sixth as large as the expected benefit to life satisfaction from having a romantic partner.
-Among those who believed in both a Spirit/Life-Force and a God, life satisfaction had grown to ($\hat{y} = 5.03, se = 0.02, \Delta \hat{y} = 0.12$).
-In practical terms, such growth is equivalent to 1.5 times the expected gain to life satisfaction from being employed and almost a third benefit to life satisfaction from having a romantic partner.
-By contrast, life satisfaction among the skeptical population who neither believed in a God nor a Life Force remained stable over the eight year interval $\hat{y} =  4.94, se =0.02, \Delta \hat{y} = 0$).
-Relatedly, life satisfaction in the population that believe in God but are skeptical about a Spirit/Life Force was constant $\hat{y} =  5.00, se = 0.03, \Delta \hat{y} = -.02$) (See Figure 3).
+In practical termss, the rate of increase in personal wellbeing among those with Spirit/Life force beliefs was about one and half times greater than it was among those who lacked such beliefs (See Figure 2).
+Among those who believed in both a Spirit/Life-Force and a God, life satisfaction had grown to ($\hat{y} = 5.33 [5.31, 5.35]$, $\Delta \hat{y} = 0.22$).
+By contrast, life satisfaction among the skeptical population who neither believed in a God nor a Life Force lagged behind over the eight year interval $\hat{y} =  5.24 [5.21, 5.27], \Delta \hat{y} = 0.12$).
+Relatedly, life satisfaction in the population that believe in God but are skeptical about a Spirit/Life Force grew less steeply $\hat{y} =  5.32 [5.27, 5.36], \Delta \hat{y} = 0.13$).
+In practical terms, the rate of increase in life satisfaction among those with Spirit/Life force beliefs was twice as large (See Figure 3).
 

### 2020-11-25 Supplement_And_Analysis.Rmd
added expectations. need to fix all figures and rewrite results.

@@
     institution   : "School of Psychology Victoria University"
-
 authornote: |
@@
 ```{r setup, include = FALSE}
-library("papaja")
-r_refs("r-references.bib")
+#library("papaja")
+#r_refs("r-references.bib")
 ```
@@
 
-```{r, echo = TRUE}
+```{r, read_data, cache=TRUE}
 # custom function for my machine to read data. you'll need to read data to your machine. tweak function in the `funs.R' script for your machine
@@
 # you might need to install the dev verion.//remotes::install_github("tysonstanley/furniture")
-demographic_table_latex(d_3)
+#demographic_table_latex(d_3)
 ```
@@
 count_waves_participants(d_8)
-count_waves_participants(d_9)
 ```
@@
 
-```{r robust to years and beliefs only, cache = TRUE, dependson=dat_3,cache =TRUE}
+```{r robust to years and beliefs only, cache = TRUE, dependson=d_3,cache =TRUE}
 library(lme4)
@@
 
-```{r prediction plot of ML longitudinal model, cache =TRUE, dependson=mod.1aa}
+```{r prediction plot of ML longitudinal model, cache =TRUE}
 ```{r bayesian_pwi_predict, cache =TRUE}
@@
 
-```{r table of saturated models, cached =TRUE}
+```{r table of saturated models, cached = TRUE}
 sjPlot::tab_model(mod.1a,mod.2a)
@@
 
-```{r plot_coefficientplots_LMER_FULL, cache =TRUE}
+```{r plot_coefficientplots_LMER_FULL, cache = TRUE}
 g1/g2
@@
 Next, recall we want to check whether inference is  robustness to the researcher degree of freedom in selecting the number of waves.
-```{r run all the models with different ns, cache=TRUE}
+```{r run all the models with different ns, cache = TRUE}
 mod.1_3 <-lmer (pwi_formula_simple, data= d_4)
@@
 #saveRDS(mod.1_b,"backup_mod.1_b") this is excessive, but we save another backup to avoid running this model again in case of SNAFU.
+rm(mod.1_b)
 ```
+
 ```{r lsbayesmodel}
@@
 
-Tables
+Tables:
 ```{r lifesat bayesian model table:pwi, cache =TRUE}
@@
 ```{r mixing of sd_pwi}
-plot(mod.1_b, pars ="b_") #requires stronger priors
+plot(mod.1_b, pars ="b_") 
 ```
@@
 
-```{r posterior predictive checks bayes:pwi, cache=TRUE}
-
-pp_1<-NULL
-pp_1<-rstanarm::pp_check(mod.1_b) + ylab("Personal Wellbeing")
-
-pp_2<-NULL
-pp_2<-rstanarm::pp_check(mod.2_b) + ylab("Life Satisfaction")
-
-# plot
-pp_1 + pp_2
-```
+# ```{r posterior predictive checks bayes:pwi, cache=TRUE}
+# # cannot run on my laptop
+# # pp_1<-rstanarm::pp_check(mod.1_b) + ylab("Personal Wellbeing")
+# # pp_2<-rstanarm::pp_check(mod.2_b) + ylab("Life Satisfaction")
+# # 
+# # # plot
+# pp_1 + pp_2
+#```
 
@@
 
-Prediction plots
+Predictions and plots of them. First we get the predictions.
+
+```{r}
+preds_1b<-get_predictions(mod.1_b)
+preds_1b
+```
+
+```{r}
+preds_2b<-get_predictions(mod.2_b)
+preds_2b
+```
+
 ```{r bayesian_pwi_predicted, cache=TRUE, dependson= mod.1_b}

### 2020-11-25 funs.R
fix functions

@@ -5,19 +5,20 @@ data_read_mymachine <- function(){
 }
 
 # this code is for citing packages 
+
 cite_packages <- function() {
   citation(package = "base", lib.loc = "/Users/josephbulbulia/Dropbox/BIBS")
-toLatex(sessionInfo(), locale = FALSE)
-sapply(names(sessionInfo()$otherPkgs), function(x) print(citation(x), style = "Bibtex"))
-out <- sapply(names(sessionInfo()$otherPkgs),
-              function(x) print(citation(x), style = "Bibtex"))
-print(out)
+  toLatex(sessionInfo(), locale = FALSE)
+  sapply(names(sessionInfo()$otherPkgs), function(x) print(citation(x), style = "Bibtex"))
+  out <- sapply(names(sessionInfo()$otherPkgs),
+                function(x) print(citation(x), style = "Bibtex"))
+  print(out)
 }
 
 
 # data clean
 data_clean_spirit_wellbeing <- function(df,y) {  
- #function for cleaning this dataset for any arbitrary number of waves
+  #function for cleaning this dataset for any arbitrary number of waves
   n_waves =  as.numeric(y)
   out <- df %>%
     dplyr::mutate(Beliefs = factor(ifelse(df$Believe.God == "Not Believe God" & df$Believe.Spirit == "Not Believe Spirit", "_Skeptic_", 
@@ -50,7 +51,7 @@ data_clean_spirit_wellbeing <- function(df,y) {
                   Relid_C = scale(Relid, scale = TRUE,center=TRUE),
                   Religious = as.factor(Religious),
                   Religion_Church_Log_C = scale(Religion.Church.Log, center = TRUE, scale = FALSE))%>%
-   dplyr::rename(Political_Conservativism = Pol.Orient)
+    dplyr::rename(Political_Conservativism = Pol.Orient)
   
   return(out)
 }
@@ -82,8 +83,8 @@ demographic_table <- function(x){
                    Urban +  
                    Beliefs + 
                    LIFESAT + PWI|Wave, data = x, 
-               overall = F)
-  }
+                 overall = F)
+}
 
 beliefs_table <- function(x){
   table1::table1(~ Beliefs|Wave, data = x, 
@@ -113,69 +114,78 @@ demographic_table_latex <- function(x){
 }
 
 
-# function for predicting PWI
+# function for predicting PWI  Not used
 
-graph_predictions_pwi <- function(x){ 
[…]

### 2020-11-26 Supplement_And_Analysis.Rmd
few fixes
rewrote supplement

@@
 ---
-title             : "Supplement:Analysis for Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Belief"
-shorttitle        : "Supplement 1"
-author: 
-  - name          : "Benjamin Highland"
-    affiliation   : "1"
-    corresponding : yes    # Define only one corresponding author
-    address       : "520 Power Plant Circle Apartment 105 Winston-Salem, NC 27101 USA"
-    email         : "benhighland6420@gmail.com"
-  - name          : "Everett L. Worthington"
-    affiliation   : "2"
-    institution   : "Department of Psychology Virginia Commonwealth University"
-  - name          : "Don E Davis"
-    affiliation   : "3"
-    institution   : "Department of Psychology Georgia State University"
-  - name          : "Chris G. Sibley"
-    affiliation   : "4"
-    institution   : "School of Psychology University of Auckland"
-  - name          : "Joseph A. Bulbulia"
-    affiliation   : "5"
-    institution   : "School of Psychology Victoria University"
+title             : "Supplement: Analysis for Longitudinal Evidence for Growth in Subjective well-being from Spiritual Belief"
+author: "contact: joseph.bulbulia@gmail.com on behalf of Highland et al."
+date: "27/11/2020"
 authornote: |
-  1 Wake Forest School of Medicine 2 Department of Psychology Virginia Commonwealth University 3 Department of Psychology Georgia State University 4 School of Psychology University of Auckland 5 School of Psychology Victoria University
-
   <!-- https://tinyurl.com/ybremelq -->
@@
 ```{r setup, include = FALSE}
-#library("papaja")
+# library("papaja")
 #r_refs("r-references.bib")
@@
 
-This supplement offers a step-by-step description our full analysis of the "Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Belief."  This analysis can be found on github [https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing](https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing)
+This supplement offers a step-by-step analysis for the "Longitudinal Evidence for Growth in Subjective Well-being from Spiritual Beliefs." The functions and other materials for this analysis can be found at the following GitHub link: [https://github.com/jacopastorius/nzavs_spiritbeliefwell-being](https://github.com/jacopastorius/nzavs_spiritbeliefwell-being)
 
-First we load our functions. We use functions to automate repetative coding, to avoid errors from repetation, and to make the data analysis easier to follow.
+First, we load our functions.
 
-```{r load_libraries, echo=TRUE, include=FALSE}
+```{r load_libraries, echo=TRUE, include=TRUE}
 source("libs.R")
@@
 
-We need to give credit to those who have developed the packages we are using. This script returns a bibliography file.
-```{r, sessioninfo, include=FALSE}
-#  custom function for citing packages. 
-cite_packages()
-```
+We use functions to automate repetition in coding, to avoid errors from such repetition, and to make the data analysis easier to follow.
 
-This reads the data into my machine.
-Note that the NZAVS is an ongoing longitudinal study.
-Our ethics requires us to protect the privacy of our participants.
-For this reason, we cannot post the raw data.
-However, we provide data to qualified researchers for the purposes of replication.
-Additionally we are interested in collaboration with qualified researchers.
-For data requests, please contact Joseph Bulbulia [joseph.bulbulia@gmail.com](joseph.bulbulia@gmail.com), Chris G.
-Sibley  [c.sibley@auckland.ac.nz](c.sibley@auckland.ac.nz).
-
+This function reads our data into my machine. 
 ```{r, read_data, cache=TRUE}
-# custom function for my machine to read data. you'll need to read data to your machine. tweak function in the `funs.R' script for your machine
+# custom function for my machine to read data. you'll need to read data to your machine. tweak function in the `funs.R` script for your machine
 d<-data_read_mymachine()
@@
 
-This code will create a demographic table.
+## Sample demographic table
 
@@
 
-This table shows the  belief responses over the 9 years.  
 ```{r, include=FALSE}
@@
 
-We wrote the publication in \LaTeX. For the data table in LaTeX run this code.
+We wrote the publication in $\LaTeX$, which requires converting the tables to .tex files; this conversion used a custom function we wrote in the `funs.R` file.
 
-```{r latex_data_table, include=FALSE}
+```{r latex_data_table, include=TRUE}
 # you might need to install the dev verion.//remotes::install_github("tysonstanley/furniture")
-#demographic_table_latex(d_3)
+#demographic_table_latex(d_3) # note the furniture package conflicts with the table1 package. 
 ```
 
-We select a minimum of three waves because we want to recover a longitudinal sample in which a slope for the effect of Beliefs might change (or not).
-However, we want to consider whether our selection of three waves is sensitive to that number.
-So we also prepare 4-7 years of data frames for separate analyses, and we check the n's for each of those datasets.
+## How many responses waves to select?
 
-```{r, include = FALSE}
+We selected a minimum of three waves because we want to recover a longitudinal sample in which a slope for the effect of the years $\times$ beliefs effect might be recovered.
+However, we needed to consider whether our selection of three waves was sensitive to that degree of researcher freedom.
+So we also prepared 4-8 years of data frames for separate analyses.
+
+```{r, include = TRUE}
 # create datasets to test robustness to years
@@
 
-In the first set of analyses we included covariates to adjust for potential confounding effects from usual suspect factors associated with our two measures of wellbeing that might explain the relationship of spirit-beliefs to well-being.
-We initially included a test of Variance Inflation Factors to rule out mult-collinearity.
-However, upon further reading, we became aware of at least three other types of confounding that might arise from longitudinal analysis (see: [@mcelreath2020statistical]).
-We report our controlled analysis below, which does not change the inference in this study.
-However, because adding covariates can *open* a relationship between beliefs and wellbeing we have switched to reporting the simple time series model in the main text.
-Additionally, we are careful to state that the present student cannot infer a causal relationshp between spirit beliefs and wellbeing.
-Rather, we can only infer how the expected relationship between beliefs and wellbeing changes between those who differ in such beliefs over time.
-A causal understanding inference requires testing specific mechanistic theories.
+The analyses for these waves is described below.
+These analyses reveal that the inference of the paper is not sensitive to the threshold number of waves for participant selection.
+Which is to say, we recover the same inference with an 8-wave selection criterion.
+
+## Covariates
+
+In the the reported analyses, we included covariates to adjust for potential confounding effects from usual suspect factors associated with our two measures of well-being (Personal Well-Being, Life Satisfaction).
+Where $Z$ is a third variable, $X$ is the predictor, and $Y$ is the outcome, there are at least four types of confounding that might arise:
+1. $Z$ causes $X$ and  $Z$ causes $Y$: $Z \leftarrow Z \rightarrow Y$
+2. $X$ causes $Z$ and  $Z$ causes $Y$: $X \rightarrow Z \rightarrow Y$
+3. $X$ causes $Z$ and $Y$ causes $Z$: $X \rightarrow Z; Y \rightarrow Z$
+4. $X$ causes $Z$; $Z$ causes $D$; and we condition on D (the descendent of $Z$: $Y$ causes $Z$: $Z$ \rightarrow D; $X \rightarrow Z; Y \rightarrow Z$ (see: [@mcelreath2020statistical], Chapter 6).
+
+We report our usual-suspect analysis below, which does not change the inference in this study.
+However, the third and fourth type of confounding illustrate how conditioning on a covariates may confound by *opening* a relationship between the $X$ and $Y$ indicators in a regression analysis.
+Such confounding is particularly worrying in the case of longitudinal research.
+The second type of counfounding can mask a relationship.
+We do not wish to exclude third variable explanations of the kind that arise from the first type of confounding.
+Indeed, we are careful to state in the main text that our study cannot infer a causal relationship between spirit beliefs and well-being.
+Rather, the contribution of the study comes as an inference about how the expected relationship between beliefs and well-being changes between those who differ in such beliefs over time.
+By contrast, a causal inference requires testing specific mechanistic theory.
 This is a matter for future research.
+Finally, the usual suspect analysis that we report below does not affect inference.
+We find that the relationships that we observe in the simple analysis also hold there.
 
-First we use a maximum Likelihood estimator.
-Such an estimator is faster than our preferred bayesian estimator.
-We consider whether the slope of beliefs changes as a function of time, or equivalently whether the time effect differs depending on one's leve of belief.
-We also create a table for the two analyses (personal wellbeing and lifesatisfaction.)
+## Bayesian data analysis
 
-```{r robust to years and beliefs only, cache = TRUE, dependson=d_3,cache =TRUE}
-library(lme4)
-pwi_formula_simple <- 'PWI ~  years * Beliefs + (1+ Beliefs|Id)'
-ls_formula_simple <- 'LIFESAT ~  years * Beliefs + (1+ Beliefs|Id)'
-mod.1aa <-lmer(pwi_formula_simple, data= d_3)
-mod.2aa <-lmer(ls_formula_simple, data= d_3)
+Bayesian inference require substantial computational resources.
+The cost of time is repaid with a model that efficiently pools information to produce a more accurate, and transparent, estimate of uncertainty.
+The posterior distributions of the model describe intuitive conditional probabilities.
 
-# table
-sjPlot::tab_model(mod.1aa,mod.2aa)
-```
+Below we write the model in a notation that explicitly states the the default weakly informative priors that the BRMS package produces.
+And indicated in the model we include group-level indicator for variation in subject-level intercepts $\alpha_j$, a group-level indicator for subject-level time (or "growth")  slope  $\Beta_j$, and we model the co-variance of $\alpha_j$, Below we write the model in a notation that explicitly states the the default weakly informative priors that the BRMS package produces.
+And indicated in the model we include group-level indicator for variation in subject-level intercepts $\alpha_j$, a group-level indicator for subject-level time (or "growth")  slope  $\Beta_j$, and we model the co-variance of $\alpha_j$ and  $\Beta_j$ which for computational reasons is estimate as a correlation matrix of (${\bf S}$} and ${\bf R}$).
 
-We next produce a predictive graph for beliefs by time on PWI. Note, again, the function for producing this graph is in the "funs.R" file.
+### Priors for the Personal Well-Being model
 
-```{r prediction plot of ML longitudinal model, cache =TRUE}
-```{r bayesian_pwi_predict, cache =TRUE}
-graph_predictions(mod.1aa, "Predicted Values of Personal Wellbeing")
-```
-
-We next produce a predictive graph for beliefs by time on life satisfaction. 
-```{r bayesian_lifesat_predict, cache =TRUE}
-graph_predictions(mod.2aa, "Predicted Values of Life Satisfaction")
-```
-
-For both outcomes, the slope for beliefs in a spirit or life fource predicts a greater increase in  wellbeing.
-
-Next we attempt a model with that is adjusted by "usual suspect" demographic inicators.
-We find that the relationship is a little stronger.
-Because these strengthened relationship might owe to confounding, we are wearing of the adjusted model.
-The idea that skeptics are not increasing in well-being, whereas other groups are increasing strikes us as improbable.
-We do not find evidence for a such a stall in the bayesian analysis; nor do we find any such evidence for a stall in the multiply-imputed dataset analysis (which handles missing responses).
-Ultimately we favour, and report the bayesian analysis because bayesian data analysis recovers an explicit estimate of posterior uncertianty.
-
-```{r lmer_model_with_full_covariates, cache =TRUE}
-library(lme4)
-pwi_formula <- 'PWI ~  years * Beliefs +  Age_in_Decades_C + Deprivation_S + Education_S  + Employed + Ethnic_Categories + Male  + Has_Partner  + Political_Conservativism_S + Urban + (1 + Beliefs|Id)'
-ls_formula <- 'LIFESAT ~  years * Beliefs +  Age_in_Decades_C + Deprivation_S + Education_S  + Employed + Ethnic_Categories + Male  + Has_Partner  + Political_Conservativism_S + Urban + (1 + Beliefs|Id)'
-
-mod.1 <-lmer (pwi_formula, data= d_3)
-mod.2 <-lmer (ls_formula, data= d_3)
-```
-
-Here is the table.
-```{r saturated model, cache = TRUE, dependson=d_3}
-sjPlot::tab_model(mod.1,mod.2)
-```
-
-Compare complex with simple models, and we find the inference doesn't change. 
-```{r}
-# identical result
-sjPlot::tab_model(mod.1,mod.1aa)
-sjPlot::tab_model(mod.2,mod.2aa)
-```
-
-Next we check for multi-collinearity.
-
-```{r check collinearity, cache =TRUE}
-# Note High VIFS from 
-performance::check_collinearity(mod.1)
-performance::check_collinearity(mod.2)
-```
-
-We observe that VIFs are high, for the years x beliefs interaction.
-However, this is likely to be an artifact of how we code the variables (we do not center by default), but to make sure let's center the years indicator and run the models again.
-We we do this we find that collinearity disappears, and without any effect on inference.
-
-```{r lmer model centering years}
-library(lme4)
-pwi_formulaC <- 'PWI ~  yearsC * Beliefs +  Age_in_Decades_C + Deprivation_S + Education_S  + Employed + Ethnic_Categories + Male  + Has_Partner  + Political_Conservativism_S + Urban + (1 |Id)'
-
-ls_formulaC <- 'LIFESAT ~  yearsC * Beliefs +  Age_in_Decades_C + Deprivation_S + Education_S  + Employed + Ethnic_Categories + Male  + Has_Partner  + Political_Conservativism_S + Urban + (1 |Id)'
-
-mod.1a <-lmer (pwi_formulaC, data= d_3)
-mod.2a <-lmer (ls_formulaC, data= d_3)
-
-summary(mod.1a)
-summary(mod.2a)
-# identical result
-performance::check_collinearity(mod.1a)
-performance::check_collinearity(mod.2a)
-```
-
-Here's the table, again, inference is robust.
-
-```{r table of saturated models, cached = TRUE}
-sjPlot::tab_model(mod.1a,mod.2a)
-```
-
-Her are thecoefficient plots for the full models
-```{r coefficientplots_LMER_FULL, cache=TRUE}
-plot_models(mod.2, mod.1) + ylim(-.4, .75) + theme_abyss() + scale_colour_brewer(labels = c("Personal Wellbeing", "Life Satisfaction")) # Plots
-```
-
-Here is the code for the coefficient plots of our preferred simple model
-```{r cache =TRUE}
-g1<-plot(parameters::model_parameters(mod.1aa))+ ggtitle("Personal Wellbeing Coefficients")
-g2<-plot(parameters::model_parameters(mod.2aa))+ ggtitle("Life Satisfaction Coefficients")
-```
-
-And here's the graph.
-
-```{r plot_coefficientplots_LMER_FULL, cache = TRUE}
-g1/g2
-```
-
-Next, recall we want to check whether inference is  robustness to the researcher degree of freedom in selecting the number of waves.
-```{r run all the models with different ns, cache = TRUE}
-mod.1_3 <-lmer (pwi_formula_simple, data= d_4)
-mod.2_3 <-lmer (ls_formula_simple, data= d_4)
-
-mod.1_4 <-lmer (pwi_formula_simple, data= d_4)
-mod.2_4 <-lmer (ls_formula_simple, data= d_4)
-
-mod.1_5 <-lmer (pwi_formula_simple, data= d_5)
-mod.2_5 <-lmer (ls_formula_simple, data= d_5)
-
-mod.1_6 <-lmer (pwi_formula_simple, data= d_6)
-mod.2_6 <-lmer (ls_formula_simple, data= d_6)
-
-mod.1_7 <-lmer (pwi_formula_simple, data= d_7)
-mod.2_7 <-lmer (ls_formula_simple, data= d_7)
-
-mod.1_8 <-lmer (pwi_formula_simple, data= d_8)
-mod.2_8 <-lmer (ls_formula_simple, data= d_8)
-```
-
-Let's plot all the models.  We we do so, we find that uncertainty increases, but that the inference remains robust.
-
-Let's graph the extremes of the degree-of-freedom choice to be certain that the inference remains satisfying to the eye (and it is.)
-
-```{r extremes_of_researcher_degree_of_freedom_test, cache =TRUE}
-library(parameters)
-g1a_3<-plot(parameters::model_parameters(mod.1_3))+ ggtitle("Personal Wellbeing Coefficients: 8 Waves")
-g2a_3<-plot(parameters::model_parameters(mod.2_3))+ ggtitle("Life Satisfaction Coefficients: 8 Waves")
-
-g1a_8<-plot(parameters::model_parameters(mod.1_8))+ ggtitle("Personal Wellbeing Coefficients: 8 Waves")
-g2a_8<-plot(parameters::model_parameters(mod.2_8))+ ggtitle("Life Satisfaction Coefficients: 8 Waves")
-```
-
-Estimates robust at 8 waves
-```{r USE_2_coefficientplots_LMER_FULL,cache =TRUE}
-library(patchwork)
-g1a_3/g1a_8
-```
-
-Finally we model the relationshp bewteen having a partner and deprivation, to address the reviewer's worry that we might have introduced correlation of between deprivation and having a partner that could spoil inference.
-We indeed find such a relationship.
-This is an important potential confound.
-Again we emphasise that in a longitudinal setting the reviewers concern is even greater, as we might open a path to the outcome variable by including the demographic indicator.
-
-```{r checkforref model,cache =TRUE}
-m<-lmer (NZdep ~ Partner + (1|Id), data= d_3, na.action = na.omit)
-tab_model(m)
-model_parameters(m)
-```
-
-# Bayesian estimation
-
-Bayesian inference is time intensive.
-However, a bayesian model will pool information to produce a better, and natural estimate of uncertainty.
-The posterior distributions of the model describe conditional probabilities.
-So where we can we want to use bayesian models.
-Here we use the default weakly informative priors that the BRMS package produces.
-It is generally better to use informative priors.
-However, the size or our dataset does not make a difference.
-We report the priors at the end.
-
-```{r setmodelformula}
-forumula_bayesPWI <- bf(PWI ~ years * Beliefs + (1 + Beliefs|Id))
-forumula_bayesLIFESAT <- bf(LIFESAT ~  years * Beliefs + (1 + Beliefs|Id))
-```
-```{r pwi_bayesmodel}
-# PWI
-mod.1_b <-brm(forumula_bayesPWI, data= d_3, file ="mod.1_b.rds") # note these are very large files
-summary(mod.1_b)
-#saveRDS(mod.1_b,"backup_mod.1_b") this is excessive, but we save another backup to avoid running this model again in case of SNAFU.
-rm(mod.1_b)
-```
-
-```{r lsbayesmodel}
-# LIFE SAT
-mod.2_b <-brm (forumula_bayesLIFESAT, data= d_3,cores = 8, file ="mod.2_b.rds") # note these are very large files
-# saveRDS(mod.2_b,"backup_mod.2_b") this is excessive, but we save another backup to avoid running this model again in case of SNAFU.
-```
-
-Tables:
-```{r lifesat bayesian model table:pwi, cache =TRUE}
-tab_model(mod.1_b)
-```
-
-```{r lifesat bayesian model table:ls lifesat, cache =TRUE}
-tab_model(mod.2_b)
-```
-
-Here we extract the information about the priors
-
-```{r stancode for models}
-#model equation
-stancode(mod.1_b)
-stancode(mod.2_b)
-```
-
-```{r prior summaries}
-brms::prior_summary(mod.1_b)
-brms::prior_summary(mod.2_b)
-```
-
-```{r bayesian_model_table, cache=TRUE}
-#tab_model(mod.1_b, show.ci= .89) # we really should use something other than 95 but the team consensus is to use 95
-tab_model(mod.1_b, mod.1_b show.ci= .95) # no difference to inference
-```
-
-```{r mixing of sd_pwi}
-plot(mod.1_b, pars ="b_") 
-```
-
-```{r mixing of ls_pwi}
-plot(mod.1_b, pars ="sd_") # requires stronger priors
-```
-
-# ```{r posterior predictive checks bayes:pwi, cache=TRUE}
-# # cannot run on my laptop
-# # pp_1<-rstanarm::pp_check(mod.1_b) + ylab("Personal Wellbeing")
-# # pp_2<-rstanarm::pp_check(mod.2_b) + ylab("Life Satisfaction")
-# # 
-# # # plot
-# pp_1 + pp_2
-#```
-
-Here are the prediction plots that we use in the main publication text.
-
-Predictions and plots of them. First we get the predictions.
-
-```{r}
-preds_1b<-get_predictions(mod.1_b)
-preds_1b
-```
-
-```{r}
-preds_2b<-get_predictions(mod.2_b)
-preds_2b
-```
-
-```{r bayesian_pwi_predicted, cache=TRUE, dependson= mod.1_b}
-graph_predictions(mod.1_b, "Predicted Values of Personal Wellbeing")
-```
-```{r bayesian_lifesat_predicted, cache=TRUE, dependson= mod.2_b}
-graph_predictions(mod.2_b, "Predicted Values of Life Satisfaction")
-```
-
-## Imputed Analysis
-
-Finally we run the data analysis to test whether our results are robust to missing data.
-Missing responses can be an important source of bias in a longitudinal study.
-Again we write some functions to deal with all the work.
-I have saved the imputed files and will read them back to for analysis.
-
-```{r echo = TRUE,  cache = TRUE, dependson = "dfben"}
-# Imputed data
-# imputed3<-amelia_imputation_clean(d_3)
-# saveRDS(imputed3,"imputed3")
-imputed3<-readRDS(imputed3)
-```
-
-We rerun the imputed dat analysis.
-
-```{r cache = TRUE, imputedsimplePWI}
-# run models iterating over imputed data
-mod_imputed_pwi<-loop_lmer_model(pwi_formula_simple,imputed3)
-```
-
-```{r imputed pwi table, cache=TRUE}
-tab_mods_ls<-loop_lmer_model_tab(mod_imputed_ls)
-create_hux_table(tab_mods_ls)
-```
-
-We graph the coefficients.
-```{r coef_PWI_imputed coefplot, cache =TRUE}
-# coeff plot
-imp_pwi <- plot(tab_mods_pwi) + ggtitle("Personal Wellbeing Coefficients (Imputed)") # pick your  own title
-imp_pwi
-```
-
-We create a graph for the predictions.
-```{r r coef_PWI_imputed coefplot, cache=TRUE}
-# function for prediction
-graph_predictions_imputed(mod_imputed_pwi,"Predicted Values of Personal Wellbeing (imputed)")
-```
-
-The inference is robust in the imputation model.
-
-## Life Satisfaction Imputation Model
-
-The following code returns the analysis for the life statisfaction model (again the functions are in the "funs.R" file).
-
-```{r cache = TRUE}
-# run models iterating over imputed data
-mod_imputed_ls<-loop_lmer_model(ls_formula_simple,imputed3)
-
-```{r table_ls_imputed,cache=TRUE}
-# table
-tab_mods_ls<-loop_lmer_model_tab(mod_imputed_ls)
-create_hux_table(tab_mods_ls)
-```
-
-Here's the graph:
-
-```{r coef_PWI_imputed, cache = TRUE}
-# coeff plot
-imp_ls<-plot(tab_mods_ls) + ggtitle("Life Satisfaction Coefficients (Imputed)") 
-imp_ls
-```
-
-```{r pred_LS_imputed,  cache = TRUE}
-# function for prediction
-graph_predictions_imputed(mod_imputed_ls,"Predicted Values of Life Satisfaction (imputed)")
-```
-
-```{r coefficient_plot_both_imputed,  cache = TRUE}
- imp_pwi/imp_ls +  plot_annotation(tag_levels = 'i')
-```
-
-## Bayesian Multiple imputation models
-
-We want to run a bayesian MI model. However, we we do this, we blow up the RAM in JB's machine.  So we must satisfy ourselves with a frequentist test here.
-
-<!-- ```{r} won't work for missing categorical predictors--> 
-<!-- form_pwi <- bf(PWI | mi() ~ years * mi(Beliefs)) + -->
-<!--   bf(Beliefs | mi() ~ years) + set_rescor(FALSE) -->
-<!-- imp_bayes_pwi <- brm(form_pwi, data = d_3) -->
-<!-- ``` -->
-
-```{r create data frame of lists for BRMS imputation}
-# brms requires a list of the imputations
-brmdat <-list(imputed3$imputations$imp1,
-              imputed3$imputations$imp2,
-              imputed3$imputations$imp3,
-              imputed3$imputations$imp4,
-              imputed3$imputations$imp5,
-              imputed3$imputations$imp6,  
-              imputed3$imputations$imp7,
-              imputed3$imputations$imp8,
-              imputed3$imputations$imp9,
-              imputed3$imputations$imp10)
-
-```
-
-<!-- ```{r bayesian imputed model_pwi} -->
-<!-- # not run -->
-<!-- #fit_imp1 <- brm_multiple(PWI ~ years * Beliefs + (1 + years|Id), data = brmdat) -->
-<!-- #summary(fit_imp1) -->
-<!-- #save(fit_imp1) -->
-<!-- ``` -->
-
-<!-- ```{r bayesian imputation model_ls} -->
-<!-- # not run -->
-<!-- #fit_imp2 <- brm_multiple(LIFESAT ~ years * Beliefs + (1+years|Id), data = brmdat, file = "fit_imp2") -->
-<!-- ``` -->
-
-In the commends of the "Analysis.Rmd" file, there are further analyeses that we ran (e.g. within/between subject effects). None affected inference.
-
-# model equations
-```{r}
-stancode(mod.1_b)
-```
-
-\newpage
-## Priors for the Personal Wellbeing model
-
-Model priors for Personal Wellbeing including group-level ID intercept and beliefs slopes:
+First we report priors for the Personal Well-Being model. Note that we model the correlation of the id-level intercept and id-level year-slope.
 
@@
 
-\newpage
-## Priors for the Life Satisfaction model:
-Model priors for Life Satisfaction Model with including group-level ID intercept and belief slopes:
+### Priors for the Life Satisfaction model:
+
+Model priors for the  Life Satisfaction model. Here, again, we model the correlation of the id-level intercept and id-level year-slope.
 
@@
 \alpha_{0} \sim \text{StudentT}(3,5.5, 2.5)\nonumber\\
-\alpha_j \sim \text{StudentT}+(3,0,10)\nonumber\\
+\alpha_j \sim \text{StudentT}+(3,0,2.5)\nonumber\\
 \boldsymbol{\beta}\sim \boldsymbol{b} + \beta_j \nonumber\\
@@
 
-\newpage
-## Deterministic model
+### Theoretical model
 
-For the deterministic part of both the Personal Wellbeing and the Life Satisfaction models:
+The deterministic part of both the Personal Well-Being and the Life Satisfaction models is as follows.
 
@@
 
-<!-- Frequentist model: -->
+```{r setmodelformula}
+forumula_bayesPWI <- bf(PWI ~ yearsC * Beliefs + (1 + yearsC|Id))
+forumula_bayesLIFESAT <- bf(LIFESAT ~  yearsC * Beliefs + (1 + yearsC|Id))
+```
+```{r pwi_bayesmodel}
+# PWI
+mod.1_b <-brm(forumula_bayesPWI, data= d_3, file ="mod.1_b.rds", cores = 8) # note these are very large files
+```
 
+```{r lsbayesmodel}
+# LIFE SAT
+mod.2_b <-brm (forumula_bayesLIFESAT, data= d_3,cores = 8, file ="mod.2_b.rds") 
+# note these are very large files
+```
+
+<!-- Tables: -->
+<!-- # ```{r pwi_Bayesian_model_table_pwi, cache =TRUE} -->
+<!-- # tab_model(mod.1_b) -->
+<!-- # ``` -->
+<!-- # -->
+<!-- # ```{r lifesat_Bayesian_model_table_ls lifesat, cache =TRUE} -->
+<!-- # tab_model(mod.2_b) -->
+<!-- # ``` -->
+
+```{r texreg_bayes_table:pwi,cache=TRUE, include=FALSE}
+#Results for publication.
+table_model_latex_pwi(mod.1_b)
+```
+
+```{r texreg_bayes_table:ls, cache=TRUE, include=FALSE}
+table_model_latex_ls(mod.2_b)
+```
+
+<!-- Coefficient plots used in manuscript-->
+<!-- ```{r coefficientplotsbayes, cache=TRUE} -->
+<!-- plot_coefs(mod.1_b,mod.2_b) -->
+<!-- ``` -->
+
+```{r stancode for models, include=FALSE}
+#model equation
+#Here we extract our priors for reporting
+
+stancode(mod.1_b)
+stancode(mod.2_b)
+```
+
+```{r prior summaries, include =FALSE}
+#this does the same as the stancode
+brms::prior_summary(mod.1_b)
+brms::prior_summary(mod.2_b)
+```
+
+```{r Bayesian_model_table, cache=TRUE}
+#tab_model(mod.1_b, show.ci= .89) # we really should use something other than 95 but the team consensus is to use 95
+tab_model(mod.1_b, mod.2_b, show.ci= .95) # no difference to inference
+```
+<!-- ```{r Bayesian_model_table, cache=TRUE, include =FALSE} -->
+<!-- table_model_latex_dual() -->
+<!-- ``` -->
+
+We inspected the posterior distributions, which look good.
+
+```{r mixing_pwi, cache=TRUE}
+plot(mod.1_b, pars ="b_") 
+```
+```{r mixing_ls, cache=TRUE}
+plot(mod.2_b, pars ="b_") 
+```
+
+As do the random components
+
+```{r mixing_sd_pwi}
+plot(mod.1_b, pars ="sd_") # requires stronger priors
+```
+
+```{r mixing_ls_pwi}
+plot(mod.2_b, pars ="sd_") # requires stronger priors
+```
+
+The posterior prediction plots look reasonable.
+
+```{r posteriorpredictivechecks:pwi, cache=TRUE}
+# cannot run on my laptop
+pp_1<-rstanarm::pp_check(mod.1_b) + ylab("Personal well-being")
+pp_2<-rstanarm::pp_check(mod.2_b) + ylab("Life Satisfaction") ## life sat suggests an ordinal model
+pp_1 + pp_2
+```
+
+### Predictions graphs
+
+Personal Well being predictions.
+
+```{r}
+preds_1b<-get_predictions(mod.1_b)
+preds_1b
+```
+
+Life Satisfaction predictions.
+
+```{r}
+preds_2b<-get_predictions(mod.2_b)
+preds_2b
+```
+
+Next we graph the expected effects.
+
+```{r Bayesian_pwi_predicted, cache=TRUE, dependson= mod.1_b}
+graph_predictions(mod.1_b, "Predicted Values of Personal Well-Being")
+```
+```{r Bayesian_lifesat_predicted, cache=TRUE, dependson= mod.2_b}
+graph_predictions(mod.2_b, "Predicted Values of Life Satisfaction")
+```
+
+## ML estimation
+
+Many researchers have yet to familiarise themselves with Bayesian estimation.
+We therefore report results from models using a Maximum Likelihood estimator.
+We do so to avoid the perception of chicanery, as our experience is that such a choice does not affect inference for a dataset such as ours.
+ML estimatation is also faster than our preferred Bayesian estimator.
+
+Results are virtually identical to the Bayesian model.
+
+```{r robust to years and beliefs only, cache = TRUE, dependson=d_3,cache =TRUE}
+library(lme4)
+pwi_formula_simple <- 'PWI ~  yearsC * Beliefs + (1+ yearsC|Id)'
+ls_formula_simple <- 'LIFESAT ~  yearsC * Beliefs + (1+ yearsC|Id)'
+mod.1aa <-lmer(pwi_formula_simple, data= d_3)
+mod.2aa <-lmer(ls_formula_simple, data= d_3)
+
+# table
+sjPlot::tab_model(mod.1aa,mod.2aa)
+```
+
+Here is a predictive graph for years $\times$ beliefs interaction on PWI. Rhe function for producing this graph is in the `funs.R` file.
+
+```{r Bayesian_pwi_predict, cache =TRUE}
+graph_predictions(mod.1aa, "Predicted Values of Personal well-being")
+```
+
+Here is a predictive graph for the years $\times$ beliefs interaction on life satisfaction.
+
+```{r Bayesian_lifesat_predict, cache =TRUE}
+graph_predictions(mod.2aa, "Predicted Values of Life Satisfaction")
+```
+
+## Adjusting for Demographics.
+
+Next we attempted a model with that is adjusted by "usual suspect" demographic indicator. The inference remains unchanged.
+
+```{r lmer_model_with_full_covariates, cache =TRUE}
+library(lme4)
+pwi_formula <- 'PWI ~  yearsC * Beliefs +  Age_in_Decades_C + Deprivation_S + Education_S  + Employed + Ethnic_Categories + Male  + Has_Partner  + Political_Conservativism_S + Urban + (1 + yearsC|Id)'
+ls_formula <- 'LIFESAT ~  yearsC * Beliefs +  Age_in_Decades_C + Deprivation_S + Education_S  + Employed + Ethnic_Categories + Male  + Has_Partner  + Political_Conservativism_S + Urban + (1 + yearsC|Id)'
+
+mod.1 <-lmer (pwi_formula, data= d_3)
+mod.2 <-lmer (ls_formula, data= d_3)
+```
+
+Here is the table.
+```{r saturated model, cache = TRUE, dependson=d_3}
+sjPlot::tab_model(mod.1,mod.2)
+```
+
+We compared complex with simple models, and we found the inference does not change. 
+```{r}
+# identical result
+sjPlot::tab_model(mod.1,mod.1aa)
+sjPlot::tab_model(mod.2,mod.2aa)
+```
+
+Next we checked for multi-collinearity. Multi-collinearity is low. (Note that we centre the "years" variable, had we not done this, mult-collinearity would have been high).
+
+```{r check collinearity, cache =TRUE}
+# Low VIFS from 
+performance::check_collinearity(mod.1)
+performance::check_collinearity(mod.2)
+```
+
+```{r lmer model centering years}
+library(lme4)
+pwi_formulaC <- 'PWI ~  yearsC * Beliefs +  Age_in_Decades_C + Deprivation_S + Education_S  + Employed + Ethnic_Categories + Male  + Has_Partner  + Political_Conservativism_S + Urban + (1+ yearsC|Id)'
+
+ls_formulaC <- 'LIFESAT ~  yearsC * Beliefs +  Age_in_Decades_C + Deprivation_S + Education_S  + Employed + Ethnic_Categories + Male  + Has_Partner  + Political_Conservativism_S + Urban + (1+ yearsC|Id)'
+
+mod.1a <-lmer (pwi_formulaC, data= d_3)
+mod.2a <-lmer (ls_formulaC, data= d_3)
+
+# identical result
+performance::check_collinearity(mod.1a)
+performance::check_collinearity(mod.2a)
+```
+
+Here's the table, again, we find inference is robust.
+
+```{r table of saturated models, cached = TRUE}
+sjPlot::tab_model(mod.1a,mod.2a)
+```
+
+<!-- Her are the coefficient plots for the full models -->
+<!-- ```{r coefficientplots_LMER_FULL, cache=TRUE} -->
+<!-- plot_models(mod.2, mod.1) + ylim(-.4, .75) + theme_plain()  # Plots -->
+<!-- ``` -->
+
+Here is the code for the coefficient plots of our preferred simple model.
+
+```{r cache =TRUE}
+g1<-plot(parameters::model_parameters(mod.1aa))+ ggtitle("Personal well-being Coefficients")
+g2<-plot(parameters::model_parameters(mod.2aa))+ ggtitle("Life Satisfaction Coefficients")
+```
+
+And here's the graph.
+
+```{r plot_coefficientplots_LMER_FULL, cache = TRUE}
+g1/g2
+```
+
+Next, recall we want to check whether inference is robustness to the researcher degree of freedom in selecting how many number of waves a participant must have responded to for inclusion.
+
+```{r run all the models with different ns, cache = TRUE}
+mod.1_3 <-lmer (pwi_formula_simple, data= d_4)
+mod.2_3 <-lmer (ls_formula_simple, data= d_4)
+
+mod.1_4 <-lmer (pwi_formula_simple, data= d_4)
+mod.2_4 <-lmer (ls_formula_simple, data= d_4)
+
+mod.1_5 <-lmer (pwi_formula_simple, data= d_5)
+mod.2_5 <-lmer (ls_formula_simple, data= d_5)
+
+mod.1_6 <-lmer (pwi_formula_simple, data= d_6)
+mod.2_6 <-lmer (ls_formula_simple, data= d_6)
+
+mod.1_7 <-lmer (pwi_formula_simple, data= d_7)
+mod.2_7 <-lmer (ls_formula_simple, data= d_7)
+
+mod.1_8 <-lmer (pwi_formula_simple, data= d_8)
+mod.2_8 <-lmer (ls_formula_simple, data= d_8)
+```
+
+Let's plot all the models. We find that uncertainty increases, but that, again, the inference remains robust.
+
+```{r extremes_of_researcher_degree_of_freedom_test, cache =TRUE}
+library(parameters)
+g1a_3<-plot(parameters::model_parameters(mod.1_3))+ ggtitle("Personal well-being Coefficients: 8 Waves")
+g2a_3<-plot(parameters::model_parameters(mod.2_3))+ ggtitle("Life Satisfaction Coefficients: 8 Waves")
+
+g1a_8<-plot(parameters::model_parameters(mod.1_8))+ ggtitle("Personal well-being Coefficients: 8 Waves")
+g2a_8<-plot(parameters::model_parameters(mod.2_8))+ ggtitle("Life Satisfaction Coefficients: 8 Waves")
+```
+
+```{r USE_2_coefficientplots_LMER_FULL_pwi,cache =TRUE}
+library(patchwork)
+g1a_3/g1a_8 +   plot_annotation(title = "Compare 3 X response model (top) with 8 X response model (bottom): Personal Well-Being")
+```
+
+```{r USE_2_coefficientplots_LMER_FULL_ls,cache =TRUE}
+library(patchwork)
+g2a_3/g2a_8 +   plot_annotation(title = "Compare 3 X response model (top) with 8 X response model (bottom) : Life Satisfaction")
+```
+
+Next, we modelled the relationship between having a partner and deprivation.
+We did so to address a reviewer's worry that we might have introduced correlation of between deprivation and having a partner that could spoil inference.
+We indeed found such a relationship.
+This is an important potential confound, and another reason to favour the simple longitudinal model that we used.
+
+```{r checkforref model,cache =TRUE}
+m<-lmer (NZdep ~ Partner + (1|Id), data= d_3, na.action = na.omit)
+tab_model(m)
+model_parameters(m)
+```
+
+# Missing Data Analysis
+
+Finally, we ran our analyses on 10 multiply imputed datasets.
+Missing responses can be an important source of bias in a longitudinal study.
+Again we wrote some functions to deal with all the work, you can find these in `funs.R` file.
+
+```{r echo = TRUE,  cache = TRUE}
+# Imputed data
+imputed3<-amelia_imputation_clean(d_3)
+# saveRDS(imputed3,"imputed3")
+# imputed3<-readRDS(imputed3)
+```
+
+We reran the imputed data analysis.
+
+```{r cache = TRUE, imputedsimplePWI, cache=TRUE}
+# run models iterating over imputed data
+mod_imputed_pwi<-loop_lmer_model(pwi_formula_simple, imputed3)
+```
+
+```{r imputed pwi table, cache=TRUE}
+tab_mods_pwi<-loop_lmer_model_tab(mod_imputed_pwi)
+#create_hux_table(tab_mods_pwi)
+```
+
+Next a graph of the coefficients.
+
+```{r coef_PWI_imputed_coefplot, cache =TRUE}
+# coeff plot
+imp_pwi <- plot(tab_mods_pwi) + ggtitle("Personal well-being Coefficients (Imputed)") # pick your  own title
+```
+
+The inference is robust in the imputation model.
+
+The following code returns the analysis for the life satisfaction model (again the functions are in the `funs.R` file).
+
+```{r cache = TRUE}
+# run models iterating over imputed data
+mod_imputed_ls<-loop_lmer_model(ls_formula_simple,imputed3)
+
+```{r table_ls_imputed,cache=TRUE}
+# table
+tab_mods_ls<-loop_lmer_model_tab(mod_imputed_ls)
+#create_hux_table(tab_mods_ls)
+```
+
+Here's the graph:
+
+```{r coef_imputed_ls, cache = TRUE}
+# coeff plot
+imp_ls<-plot(tab_mods_ls) + ggtitle("Life Satisfaction Coefficients (Imputed)") 
+```
+We found that the model is robust to multiple imputations over the missing data, though with greater uncertainty for the Personal Well-being effect.
+
+```{r coefficient_plot_both_imputed,  cache = TRUE}
+ imp_pwi/imp_ls +  plot_annotation(tag_levels = 'i')
+```
+
+Here are the prediction graphs.
+
+```{r  coef_PWI_imputed coefplot, cache=TRUE}
+# function for prediction
+graph_predictions_imputed(mod_imputed_pwi,"Predicted Values of Personal well-being (imputed)")
+```
+
+```{r pred_LS_imputed,  cache = TRUE}
+# function for prediction
+graph_predictions_imputed(mod_imputed_ls,"Predicted Values of Life Satisfaction (imputed)")
+```
+
+<!-- ## Bayesian Multiple imputation models -->
+<!-- We wanted to run a Bayesian MI model. However, we did this, we blew up the RAM in JB's machine. So we must satisfy ourselves with ML MI models.  -->
+<!-- ```{r} won't work for missing categorical predictors--> 
+<!-- form_pwi <- bf(PWI | mi() ~ years * mi(Beliefs)) + -->
+<!--   bf(Beliefs | mi() ~ years) + set_rescor(FALSE) -->
+<!-- imp_bayes_pwi <- brm(form_pwi, data = d_3) -->
+<!-- ``` -->
+<!-- ```{r create data frame of lists for BRMS imputation} -->
+<!-- # brms requires a list of the imputations -->
+<!-- brmdat <-list(imputed3$imputations$imp1, -->
+<!--               imputed3$imputations$imp2, -->
+<!--               imputed3$imputations$imp3, -->
+<!--               imputed3$imputations$imp4, -->
+<!--               imputed3$imputations$imp5, -->
+<!--               imputed3$imputations$imp6,   -->
+<!--               imputed3$imputations$imp7, -->
+<!--               imputed3$imputations$imp8, -->
+<!--               imputed3$imputations$imp9, -->
+<!--               imputed3$imputations$imp10) -->
+<!-- ``` -->
+<!-- ```{r Bayesian imputed model_pwi} -->
+<!-- # not run -->
+<!-- #fit_imp1 <- brm_multiple(PWI ~ years * Beliefs + (1 + years|Id), data = brmdat) -->
+<!-- #summary(fit_imp1) -->
+<!-- #save(fit_imp1) -->
+<!-- ``` -->
+<!-- ```{r Bayesian imputation model_ls} -->
+<!-- # not run -->
+<!-- #fit_imp2 <- brm_multiple(LIFESAT ~ years * Beliefs + (1+years|Id), data = brmdat, file = "fit_imp2") -->
+<!-- ``` -->
+
+There are further analyses that we ran (e.g. within/between subject effects). These can be found in the GitHub link, posted above. The inferences remain the same.
+
+## Data
+
+The NZAVS is an ongoing longitudinal study.
+Our ethics requires us to protect the privacy of our participants.
+For this reason, we cannot post the raw data, even anonymous data.
+However, we provide data by request to qualified researchers for the purposes of replication.
+We areare interested in collaboration with qualified researchers: please contact Chris G.
+Sibley: [c.sibley@auckland.ac.nz](c.sibley@auckland.ac.nz).
+
+## Credits
+Package credits:
+
+```{r, sessioninfo, include=FALSE}
+#  custom function for citing packages. 
+cite_packages()
+```
+
+# References
+
+\begingroup
+\setlength{\parindent}{-0.5in}
+\setlength{\leftskip}{0.5in}
+
+<div id="refs" custom-style="Bibliography"></div>
+\endgroup
+<!-- ## model equations -->
+<!-- ```{r} -->
+<!-- stancode(mod.1_b) -->
+<!-- ``` -->
+<!-- <!-- Frequentist model: -->
 <!-- ```{r eq1} -->
@@
 <!-- ``` -->
-
 <!-- ```{r eq2} -->
@@
 <!-- ``` -->
-
 <!-- We might want to look at within/between indivual change (i.e.
-do people who change in degree of spirit beliefs manifest different wellbeing.
+do people who change in degree of spirit beliefs manifest different well-being.
 I ran this as below, however, what we really want is a markov model, where we compare state changes between the different types of belief.
@@
 <!-- https://easystats.github.io/parameters/articles/demean.html -->
-
 <!-- ```{r} -->
@@
 <!--                 (1|Id), data= dfben, na.action = na.omit) -->
-
 <!-- parameters::check_heterogeneity(mod01, group = "ID") -->
 <!-- parameters::check_heterogeneity(mod01, group = "ID") -->
-
 <!-- mod02<-lmer (LIFESAT ~ Years + -->
@@
 <!-- ``` -->
-
 <!-- ```{r eq1} -->
@@
 <!-- ``` -->
-
 <!-- PWI Model 2 equation -->
@@
 <!-- model 1 report: for publication change all "significants" to "reliable" -->
-
 <!-- short report: -->
@@
 <!-- ``` -->
-
 <!-- short report -->
@@
 <!-- ``` -->
-
 <!-- ```{r, cache =TRUE, dependson=mod.1, include=FALSE} -->
@@
 <!-- # For publication change all "significants" to "reliable" -->
-
 <!-- easystats::model_text(mod.2) -->
 <!-- ``` -->
-
 <!-- short report -->
@@
 <!-- ``` -->
-
 <!-- long report -->
@@
 <!-- \] -->
+<!-- \[ -->
+<!-- \begin{table}[ ht ]  -->
+<!-- \centering  -->
+<!-- \caption{}\label{} -->
+<!-- \scalebox{.6}{ -->
+<!-- \begin{tabular}{ l c c c c c c c c c } -->
+<!-- \toprule -->
+<!--  &   \multicolumn{ 9 }{c}{ Wave }\\  -->
+<!--   & 2010 & 2011 & 2012 & 2013 & 2014 & 2015 & 2016 & 2017 & 2018 \\  -->
+<!--  & n = \emph{missing} & n = \emph{missing} & n = 8547 & n = 13472 & n = \emph{missing} & n = 11687 & n = 16048 & n = 14174 & n = 14883 \\  -->
+<!--  \midrule -->
+<!-- Age &   &   &   &   &   &   &   &   &  \\  -->
+<!-- NZdep &   &   &   &   &   &   &   &   &  \\  -->
+<!-- Education &   &   &   &   &   &   &   &   &  \\  -->
+<!-- Employed &   &   &   &   &   &   &   &   &  \\  -->
+<!-- \hspace{6pt}    0 &  (\%) &  (\%) & 2094 (24.5\%) & 2955 (21.9\%) &  (\%) & 2525 (21.6\%) & 3486 (21.7\%) & 3223 (22.7\%) & 3404 (22.9\%)\\  -->
+<!-- \hspace{6pt}    1 &  (\%) &  (\%) & 6453 (75.5\%) & 10517 (78.1\%) &  (\%) & 9162 (78.4\%) & 12562 (78.3\%) & 10951 (77.3\%) & 11479 (77.1\%)\\  -->
+<!-- EthnicCategories &   &   &   &   &   &   &   &   &  \\  -->
+<!-- \hspace{6pt}    Euro &  (\%) &  (\%) & 6731 (78.8\%) & 11024 (81.8\%) &  (\%) & 9671 (82.8\%) & 13389 (83.4\%) & 11814 (83.3\%) & 12427 (83.5\%)\\  -->
+<!-- \hspace{6pt}    Maori &  (\%) &  (\%) & 1188 (13.9\%) & 1608 (11.9\%) &  (\%) & 1341 (11.5\%) & 1749 (10.9\%) & 1620 (11.4\%) & 1677 (11.3\%)\\  -->
+<!-- \hspace{6pt}    Pacific &  (\%) &  (\%) & 284 (3.3\%) & 339 (2.5\%) &  (\%) & 277 (2.4\%) & 331 (2.1\%) & 243 (1.7\%) & 261 (1.8\%)\\  -->
+<!-- \hspace{6pt}    Asian &  (\%) &  (\%) & 344 (4\%) & 501 (3.7\%) &  (\%) & 398 (3.4\%) & 579 (3.6\%) & 497 (3.5\%) & 518 (3.5\%)\\  -->
+<!-- Male &   &   &   &   &   &   &   &   &  \\  -->
+<!-- \hspace{6pt}    0 &  (\%) &  (\%) & 5302 (62\%) & 8457 (62.8\%) &  (\%) & 7325 (62.7\%) & 10148 (63.2\%) & 9011 (63.6\%) & 9433 (63.4\%)\\  -->
+<!-- \hspace{6pt}    1 &  (\%) &  (\%) & 3245 (38\%) & 5015 (37.2\%) &  (\%) & 4362 (37.3\%) & 5900 (36.8\%) & 5163 (36.4\%) & 5450 (36.6\%)\\  -->
+<!-- Partner &   &   &   &   &   &   &   &   &  \\  -->
+<!-- \hspace{6pt}    0 &  (\%) &  (\%) & 2408 (28.2\%) & 3546 (26.3\%) &  (\%) & 2815 (24.1\%) & 3773 (23.5\%) & 3330 (23.5\%) & 3556 (23.9\%)\\  -->
+<!-- \hspace{6pt}    1 &  (\%) &  (\%) & 6139 (71.8\%) & 9926 (73.7\%) &  (\%) & 8872 (75.9\%) & 12275 (76.5\%) & 10844 (76.5\%) & 11327 (76.1\%)\\  -->
+<!-- Pol Orient &   &   &   &   &   &   &   &   &  \\  -->
+<!-- Urban &   &   &   &   &   &   &   &   &  \\  -->
+<!-- \hspace{6pt}    Not\_Urban &  (\%) &  (\%) & 2865 (33.5\%) & 4278 (31.8\%) &  (\%) & 3951 (33.8\%) & 5538 (34.5\%) & 2610 (18.4\%) & 2661 (17.9\%)\\  -->
+<!-- \hspace{6pt}    Urban &  (\%) &  (\%) & 5682 (66.5\%) & 9194 (68.2\%) &  (\%) & 7736 (66.2\%) & 10510 (65.5\%) & 11564 (81.6\%) & 12222 (82.1\%)\\  -->
+<!-- Beliefs &   &   &   &   &   &   &   &   &  \\  -->
+<!-- \hspace{6pt}    \_Skeptic\_ &  (\%) &  (\%) & 2048 (24\%) & 3286 (24.4\%) &  (\%) & 2951 (25.3\%) & 4116 (25.6\%) & 3916 (27.6\%) & 4120 (27.7\%)\\  -->
+<!-- \hspace{6pt}    \_SpiritExcludesGod\_ &  (\%) &  (\%) & 2295 (26.9\%) & 3735 (27.7\%) &  (\%) & 3169 (27.1\%) & 4501 (28\%) & 3918 (27.6\%) & 3970 (26.7\%)\\  -->
+<!-- \hspace{6pt}    GodAndSpirit &  (\%) &  (\%) & 3707 (43.4\%) & 5499 (40.8\%) &  (\%) & 5119 (43.8\%) & 6453 (40.2\%) & 5464 (38.5\%) & 6210 (41.7\%)\\  -->
+<!-- \hspace{6pt}    GodExcludesSpirit &  (\%) &  (\%) & 497 (5.8\%) & 952 (7.1\%) &  (\%) & 448 (3.8\%) & 978 (6.1\%) & 876 (6.2\%) & 583 (3.9\%)\\  -->
+<!-- LIFESAT &   &   &   &   &   &   &   &   &  \\  -->
+<!-- PWI &   &   &   &   &   &   &   &   &  \\  -->
+<!-- \bottomrule -->
+<!-- \end{tabular} -->
+<!-- } -->
+<!-- \end{table} -->
+<!-- \] -->
+<!-- \[ -->
+<!-- \begin{table} -->
+<!-- \begin{center} -->
+<!-- \scalebox{0.6}{ -->
+<!-- \begin{tabular}{l c c} -->
+<!-- \toprule -->
+<!--  & PWI & Life Satisfaction \\ -->
+<!-- \midrule -->
+<!-- (Intercept)                        & $\mathbf{6.63}^{***}$  & $\mathbf{4.94}^{***}$  \\ -->
+<!--                                    & $(0.03)$               & $(0.02)$               \\ -->
+<!-- Years                              & $0.00$                 & $-0.00$                \\ -->
+<!--                                    & $(0.00)$               & $(0.00)$               \\ -->
+<!-- Beliefs\_SpiritExcludesGod\_       & $\mathbf{-0.14}^{***}$ & $\mathbf{-0.08}^{***}$ \\ -->
+<!--                                    & $(0.03)$               & $(0.02)$               \\ -->
+<!-- BeliefsGodAndSpirit                & $\mathbf{-0.16}^{***}$ & $-0.04$                \\ -->
+<!--                                    & $(0.03)$               & $(0.02)$               \\ -->
+<!-- BeliefsGodExcludesSpirit           & $0.02$                 & $0.07$                 \\ -->
+<!--                                    & $(0.05)$               & $(0.04)$               \\ -->
+<!-- Age.10yrs.C                        & $\mathbf{0.08}^{***}$  & $\mathbf{0.09}^{***}$  \\ -->
+<!--                                    & $(0.01)$               & $(0.01)$               \\ -->
+<!-- Deprivation.S                      & $\mathbf{-0.14}^{***}$ & $\mathbf{-0.07}^{***}$ \\ -->
+<!--                                    & $(0.01)$               & $(0.01)$               \\ -->
+<!-- Edu.S                              & $\mathbf{0.12}^{***}$  & $\mathbf{0.06}^{***}$  \\ -->
+<!--                                    & $(0.01)$               & $(0.01)$               \\ -->
+<!-- EmployedEmployed                   & $\mathbf{0.09}^{***}$  & $\mathbf{0.08}^{***}$  \\ -->
+<!--                                    & $(0.01)$               & $(0.01)$               \\ -->
+<!-- EthnicCatsMaori                    & $\mathbf{-0.30}^{***}$ & $\mathbf{-0.06}^{**}$  \\ -->
+<!--                                    & $(0.03)$               & $(0.02)$               \\ -->
+<!-- EthnicCatsPacific                  & $\mathbf{-0.29}^{***}$ & $\mathbf{-0.08}^{*}$   \\ -->
+<!--                                    & $(0.05)$               & $(0.04)$               \\ -->
+<!-- EthnicCatsAsian                    & $-0.08$                & $-0.04$                \\ -->
+<!--                                    & $(0.05)$               & $(0.03)$               \\ -->
+<!-- Male1                              & $\mathbf{-0.11}^{***}$ & $\mathbf{-0.19}^{***}$ \\ -->
+<!--                                    & $(0.02)$               & $(0.01)$               \\ -->
+<!-- PartnerHas Partner                 & $\mathbf{0.61}^{***}$  & $\mathbf{0.42}^{***}$  \\ -->
+<!--                                    & $(0.02)$               & $(0.01)$               \\ -->
+<!-- Pol.Orient.S                       & $\mathbf{0.02}^{***}$  & $\mathbf{0.01}^{**}$   \\ -->
+<!--                                    & $(0.01)$               & $(0.00)$               \\ -->
+<!-- UrbanUrban                         & $\mathbf{-0.02}^{*}$   & $\mathbf{-0.02}^{*}$   \\ -->
+<!--                                    & $(0.01)$               & $(0.01)$               \\ -->
+<!-- Years:Beliefs\_SpiritExcludesGod\_ & $\mathbf{0.02}^{**}$   & $\mathbf{0.02}^{***}$  \\ -->
+<!--                                    & $(0.00)$               & $(0.00)$               \\ -->
+<!-- Years:BeliefsGodAndSpirit          & $\mathbf{0.02}^{***}$  & $\mathbf{0.01}^{***}$  \\ -->
+<!--                                    & $(0.00)$               & $(0.00)$               \\ -->
+<!-- Years:BeliefsGodExcludesSpirit     & $0.00$                 & $-0.00$                \\ -->
+<!--                                    & $(0.01)$               & $(0.01)$               \\ -->
+<!-- \midrule -->
+<!-- AIC                                & $248580.18$            & $200599.16$            \\ -->
+<!-- BIC                                & $248775.08$            & $200793.96$            \\ -->
+<!-- Log Likelihood                     & $-124269.09$           & $-100278.58$           \\ -->
+<!-- Num. obs.                          & $79270$                & $78888$                \\ -->
+<!-- Num. groups: Id                    & $20979$                & $20972$                \\ -->
+<!-- Var: Id (Intercept)                & $1.78$                 & $0.88$                 \\ -->
+<!-- Var: Residual                      & $0.74$                 & $0.42$                 \\ -->
+<!-- \bottomrule -->
+<!-- \multicolumn{3}{l}{\scriptsize{$^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$}} -->
+<!-- \end{tabular} -->
+<!-- } -->
+<!-- \caption{Regression Models} -->
+<!-- \label{table:coefficients} -->
+<!-- \end{center} -->
+<!-- \end{table} -->
+<!-- \] -->
+<!-- \newpage -->
+<!-- ## Personal Well-Being -->
+<!-- We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict personal well-being (PWI) with years (0-9), beliefs (four categories), Age.10yrs.C, deprivation (S), education (S), employed, ethnic categories, male, partner, political conservativism (S) and urban, with Id as a random effect.
+The model's intercept, corresponding to years = 0, beliefs = Skeptic, age = (sample mean), deprivation.S = (sample mean), education (S) =(sample mean), employed = not employed, ethnic Categories = European, male = not male, partner = no partner, political conservativism (S) = (sample mean), urban = not urban, is at 4.94 (SE = 0.02, 95% CI [4.89, 4.99], p < .001) is at 6.63 (SE = 0.03, 95% CI [6.57, 6.70], p < .001).
+Within this model: -->
+<!--   - The effect of Years is positive is not reliable (beta = 2.66e-03, SE = 3.58e-03, std.
+beta = 3.19e-03, p = 0.458).
+We infer that personal well-being did not increase in the Skeptic population between 2010 and 2018.
+-->
+<!--   - The effect of Beliefs in *Spirit Excludes God*  is reliably negative (beta = -0.14, SE = 0.03, std.
+beta = -0.03, p < .001).
+We infer that in 2010, the population that believed in a Spirit but did not believe in a God had lower personal well-being than did Skeptics.
+-->
+<!--   - The effect of Beliefs in *God And Spirit*  is reliably negative (beta = -0.16, SE = 0.03, std.
+beta = -0.02, p < .001).
+We infer that in 2010, the population that believed in a God and a Spirit or Life Force had lower Personal Well-being than did Skeptics.
+-->
+<!--   - The effect of Beliefs in *God Excludes Spirit*  is not reliable (beta = 0.02, SE = 0.05, std.
+beta = 0.02, p = 0.669).
+We infer that in 2010, the population that believed in a God and doubted a Spirit or Life Force similar levels of personal well-being to Skeptics.
+-->
+<!--   - The effect of Age.10yrs.C is reliably positive  (beta = 0.08, SE = 7.38e-03, std. beta = 0.07, p < .001). Older people tend to have greater personal well-being. -->
+<!--   - The effect of Deprivation (S) is reliably negative (beta = -0.14, SE = 7.50e-03, std. beta = -0.08, p < .001). Deprived people tend to have lower personal well-being. -->
+<!--   - The effect of Education (S) is reliably positive  (beta = 0.12, SE = 8.74e-03, std. beta = 0.07, p < .001). Educated people tend to have greater personal well-being. -->
+<!--   - The effect of Employed is reliably positive  (beta = 0.09, SE = 0.01, std. beta = 0.05, p < .001). -->
+<!--   - The effect of identifying as M\={a}ori is reliably negative (beta = -0.30, SE = 0.03, std. beta = -0.18, p < .001). -->
+<!--   - The effect of identifying as Pacific is reliably negative (beta = -0.29, SE = 0.05, std. beta = -0.17, p < .001).  -->
+<!--   - The effect of identifying as Asian is not reliable (beta = -0.08, SE = 0.05, std. beta = -0.05, p = 0.089). -->
+<!--   - The effect of identifying as male is reliably negative (beta = -0.11, SE = 0.02, std. beta = -0.06, p < .001).  -->
+<!--   - The effect of having a partner is reliably positive (beta = 0.61, SE = 0.02, std. beta = 0.37, p < .001). -->
+<!--   - The effect of political conservativism (S) is reliably positive  (beta = 0.02, SE = 5.67e-03, std. beta = 0.01, p < .001).  -->
+<!--   - The effect of living in an Urban area is reliably negative (beta = -0.02, SE = 0.01, std. beta = -0.01, p < .05). T -->
+<!-- ##  Personal well-being: years X beliefs  -->
+<!-- ```{r fig.cap = "Combination Coefficient plots for personal well-being and life satisfaction longitudinal growth models"} -->
+<!-- knitr::include_graphics("Figs/USEPWI_expected-1.png") -->
+<!-- ``` -->
+<!-- Focusing on the interaction of time with belief-type on personal well-being we observe:  -->
 
-\[
-\begin{table}[ ht ] 
-\centering 
-\caption{}\label{}
-\scalebox{.6}{
-\begin{tabular}{ l c c c c c c c c c }
-\toprule
- &   \multicolumn{ 9 }{c}{ Wave }\\ 
-  & 2010 & 2011 & 2012 & 2013 & 2014 & 2015 & 2016 & 2017 & 2018 \\ 
- & n = \emph{missing} & n = \emph{missing} & n = 8547 & n = 13472 & n = \emph{missing} & n = 11687 & n = 16048 & n = 14174 & n = 14883 \\ 
- \midrule
-Age &   &   &   &   &   &   &   &   &  \\ 
-NZdep &   &   &   &   &   &   &   &   &  \\ 
-Education &   &   &   &   &   &   &   &   &  \\ 
-Employed &   &   &   &   &   &   &   &   &  \\ 
-\hspace{6pt}    0 &  (\%) &  (\%) & 2094 (24.5\%) & 2955 (21.9\%) &  (\%) & 2525 (21.6\%) & 3486 (21.7\%) & 3223 (22.7\%) & 3404 (22.9\%)\\ 
-\hspace{6pt}    1 &  (\%) &  (\%) & 6453 (75.5\%) & 10517 (78.1\%) &  (\%) & 9162 (78.4\%) & 12562 (78.3\%) & 10951 (77.3\%) & 11479 (77.1\%)\\ 
-EthnicCategories &   &   &   &   &   &   &   &   &  \\ 
-\hspace{6pt}    Euro &  (\%) &  (\%) & 6731 (78.8\%) & 11024 (81.8\%) &  (\%) & 9671 (82.8\%) & 13389 (83.4\%) & 11814 (83.3\%) & 12427 (83.5\%)\\ 
-\hspace{6pt}    Maori &  (\%) &  (\%) & 1188 (13.9\%) & 1608 (11.9\%) &  (\%) & 1341 (11.5\%) & 1749 (10.9\%) & 1620 (11.4\%) & 1677 (11.3\%)\\ 
-\hspace{6pt}    Pacific &  (\%) &  (\%) & 284 (3.3\%) & 339 (2.5\%) &  (\%) & 277 (2.4\%) & 331 (2.1\%) & 243 (1.7\%) & 261 (1.8\%)\\ 
-\hspace{6pt}    Asian &  (\%) &  (\%) & 344 (4\%) & 501 (3.7\%) &  (\%) & 398 (3.4\%) & 579 (3.6\%) & 497 (3.5\%) & 518 (3.5\%)\\ 
-Male &   &   &   &   &   &   &   &   &  \\ 
-\hspace{6pt}    0 &  (\%) &  (\%) & 5302 (62\%) & 8457 (62.8\%) &  (\%) & 7325 (62.7\%) & 10148 (63.2\%) & 9011 (63.6\%) & 9433 (63.4\%)\\ 
-\hspace{6pt}    1 &  (\%) &  (\%) & 3245 (38\%) & 5015 (37.2\%) &  (\%) & 4362 (37.3\%) & 5900 (36.8\%) & 5163 (36.4\%) & 5450 (36.6\%)\\ 
-Partner &   &   &   &   &   &   &   &   &  \\ 
-\hspace{6pt}    0 &  (\%) &  (\%) & 2408 (28.2\%) & 3546 (26.3\%) &  (\%) & 2815 (24.1\%) & 3773 (23.5\%) & 3330 (23.5\%) & 3556 (23.9\%)\\ 
-\hspace{6pt}    1 &  (\%) &  (\%) & 6139 (71.8\%) & 9926 (73.7\%) &  (\%) & 8872 (75.9\%) & 12275 (76.5\%) & 10844 (76.5\%) & 11327 (76.1\%)\\ 
-Pol Orient &   &   &   &   &   &   &   &   &  \\ 
-Urban &   &   &   &   &   &   &   &   &  \\ 
-\hspace{6pt}    Not\_Urban &  (\%) &  (\%) & 2865 (33.5\%) & 4278 (31.8\%) &  (\%) & 3951 (33.8\%) & 5538 (34.5\%) & 2610 (18.4\%) & 2661 (17.9\%)\\ 
-\hspace{6pt}    Urban &  (\%) &  (\%) & 5682 (66.5\%) & 9194 (68.2\%) &  (\%) & 7736 (66.2\%) & 10510 (65.5\%) & 11564 (81.6\%) & 12222 (82.1\%)\\ 
-Beliefs &   &   &   &   &   &   &   &   &  \\ 
-\hspace{6pt}    \_Skeptic\_ &  (\%) &  (\%) & 2048 (24\%) & 3286 (24.4\%) &  (\%) & 2951 (25.3\%) & 4116 (25.6\%) & 3916 (27.6\%) & 4120 (27.7\%)\\ 
-\hspace{6pt}    \_SpiritExcludesGod\_ &  (\%) &  (\%) & 2295 (26.9\%) & 3735 (27.7\%) &  (\%) & 3169 (27.1\%) & 4501 (28\%) & 3918 (27.6\%) & 3970 (26.7\%)\\ 
-\hspace{6pt}    GodAndSpirit &  (\%) &  (\%) & 3707 (43.4\%) & 5499 (40.8\%) &  (\%) & 5119 (43.8\%) & 6453 (40.2\%) & 5464 (38.5\%) & 6210 (41.7\%)\\ 
-\hspace{6pt}    GodExcludesSpirit &  (\%) &  (\%) & 497 (5.8\%) & 952 (7.1\%) &  (\%) & 448 (3.8\%) & 978 (6.1\%) & 876 (6.2\%) & 583 (3.9\%)\\ 
-LIFESAT &   &   &   &   &   &   &   &   &  \\ 
-PWI &   &   &   &   &   &   &   &   &  \\ 
-\bottomrule
-\end{tabular}
-}
-\end{table}
-\]
-
-\[
-\begin{table}
-\begin{center}
-\scalebox{0.6}{
-\begin{tabular}{l c c}
-\toprule
- & PWI & Life Satisfaction \\
-\midrule
-(Intercept)                        & $\mathbf{6.63}^{***}$  & $\mathbf{4.94}^{***}$  \\
-                                   & $(0.03)$               & $(0.02)$               \\
-Years                              & $0.00$                 & $-0.00$                \\
-                                   & $(0.00)$               & $(0.00)$               \\
-Beliefs\_SpiritExcludesGod\_       & $\mathbf{-0.14}^{***}$ & $\mathbf{-0.08}^{***}$ \\
-                                   & $(0.03)$               & $(0.02)$               \\
-BeliefsGodAndSpirit                & $\mathbf{-0.16}^{***}$ & $-0.04$                \\
-                                   & $(0.03)$               & $(0.02)$               \\
-BeliefsGodExcludesSpirit           & $0.02$                 & $0.07$                 \\
-                                   & $(0.05)$               & $(0.04)$               \\
-Age.10yrs.C                        & $\mathbf{0.08}^{***}$  & $\mathbf{0.09}^{***}$  \\
-                                   & $(0.01)$               & $(0.01)$               \\
-Deprivation.S                      & $\mathbf{-0.14}^{***}$ & $\mathbf{-0.07}^{***}$ \\
-                                   & $(0.01)$               & $(0.01)$               \\
-Edu.S                              & $\mathbf{0.12}^{***}$  & $\mathbf{0.06}^{***}$  \\
-                                   & $(0.01)$               & $(0.01)$               \\
-EmployedEmployed                   & $\mathbf{0.09}^{***}$  & $\mathbf{0.08}^{***}$  \\
-                                   & $(0.01)$               & $(0.01)$               \\
-EthnicCatsMaori                    & $\mathbf{-0.30}^{***}$ & $\mathbf{-0.06}^{**}$  \\
-                                   & $(0.03)$               & $(0.02)$               \\
-EthnicCatsPacific                  & $\mathbf{-0.29}^{***}$ & $\mathbf{-0.08}^{*}$   \\
-                                   & $(0.05)$               & $(0.04)$               \\
-EthnicCatsAsian                    & $-0.08$                & $-0.04$                \\
-                                   & $(0.05)$               & $(0.03)$               \\
-Male1                              & $\mathbf{-0.11}^{***}$ & $\mathbf{-0.19}^{***}$ \\
-                                   & $(0.02)$               & $(0.01)$               \\
-PartnerHas Partner                 & $\mathbf{0.61}^{***}$  & $\mathbf{0.42}^{***}$  \\
-                                   & $(0.02)$               & $(0.01)$               \\
-Pol.Orient.S                       & $\mathbf{0.02}^{***}$  & $\mathbf{0.01}^{**}$   \\
-                                   & $(0.01)$               & $(0.00)$               \\
-UrbanUrban                         & $\mathbf{-0.02}^{*}$   & $\mathbf{-0.02}^{*}$   \\
-                                   & $(0.01)$               & $(0.01)$               \\
-Years:Beliefs\_SpiritExcludesGod\_ & $\mathbf{0.02}^{**}$   & $\mathbf{0.02}^{***}$  \\
-                                   & $(0.00)$               & $(0.00)$               \\
-Years:BeliefsGodAndSpirit          & $\mathbf{0.02}^{***}$  & $\mathbf{0.01}^{***}$  \\
-                                   & $(0.00)$               & $(0.00)$               \\
-Years:BeliefsGodExcludesSpirit     & $0.00$                 & $-0.00$                \\
-                                   & $(0.01)$               & $(0.01)$               \\
-\midrule
-AIC                                & $248580.18$            & $200599.16$            \\
-BIC                                & $248775.08$            & $200793.96$            \\
-Log Likelihood                     & $-124269.09$           & $-100278.58$           \\
-Num. obs.                          & $79270$                & $78888$                \\
-Num. groups: Id                    & $20979$                & $20972$                \\
-Var: Id (Intercept)                & $1.78$                 & $0.88$                 \\
-Var: Residual                      & $0.74$                 & $0.42$                 \\
-\bottomrule
-\multicolumn{3}{l}{\scriptsize{$^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$}}
-\end{tabular}
-}
-\caption{Regression Models}
-\label{table:coefficients}
-\end{center}
-\end{table}
-\]
-
-\newpage
-
-# References
-
-\begingroup
-\setlength{\parindent}{-0.5in}
-\setlength{\leftskip}{0.5in}
-
-<div id="refs" custom-style="Bibliography"></div>
-\endgroup
+<!--   - The effect of years $\times$ belief in a *Spirit Excludes God*  is reliably positive  (beta = 0.02, SE = 4.86e-03, std.
+beta = 0.02, p < .01).
+We infer that, unlike the skeptic population, people with beliefs in a spirit or life force who do not believe in a God tend to grow in their personal well-being over time.
+-->
+<!--   - The effect of years $\times$ in *God And Spirit*  is reliably positive  (beta = 0.02, SE = 4.41e-03, std.
+beta = 0.03, p < .001).
+We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their personal well-being over time -->
+<!--   - The effect of Years $\times$ belief in *God Excludes Spirit*  is not reliable(beta = 1.96e-03, SE = 8.29e-03, std.
+beta = 2.36e-03, p = 0.813).
+Thus, God believers who doubt in a Spirit or Life Force start maintain higher personal well-being than do other groups, but they are not expected to grow in personal well-being over time.
+-->
+<!-- ## Life Satisfaction  -->
+<!-- We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict personal well-being (PWI) with years (0-9), beliefs (four categories), Age.10yrs.C, deprivation (S), education (S), employed, ethnic categories, male, partner, political conservativism (S) and urban, with Id as a random effect.
+Standardized parameters were obtained by fitting the model on a standardized version of the dataset.
+The model's total explanatory power is substantial (conditional R2 = 0.69).
+The model's intercept, corresponding to years = 0, beliefs = Skeptic, age = (sample mean), deprivation.S = (sample mean), education (S) =(sample mean), employed = not employed, ethnic Categories = European, male = not male, partner = no partner, political conservativism (S) = (sample mean), urban = not urban is at 4.94 (SE = 0.02, 95% CI [4.89, 4.99], p < .001).
+Within this model: -->
+<!--   - The effect of years is not reliable (beta = -5.50e-04, SE = 2.69e-03, std.
+beta = -9.32e-04, p = 0.838).
+We infer that life satisfaction did not increase in the Skeptic population between 2010 and 2018.
+-->
+<!--   - The effect of beliefs in *Spirit Excludes God* is reliably negative (beta = -0.08, SE = 0.02, std.
+beta = 0.01, p < .001).
+We infer that overall the population that believes in a Spirit but does not believe in a God is expected to have a lower Life Satisfaction than skeptics.
+-->
+<!--   - The effect of beliefs in *God And Spirit* is not reliable (beta = -0.04, SE = 0.02, std.
+beta = 0.04, p = 0.105).
+People who believe in God but doubt a Spirit/Life force are expected to have had similar levels of life satisfaction in Wave 10 as skeptics.
+-->
+<!--   - The effect of beliefs in *God Excludes Spirit* is reliably positive (beta = 0.07, SE = 0.04, std.
+beta = 0.05, p = 0.062).We infer that overall the population that believes in a God but does not believe in a spirit is not reliably different from  the Skeptics population in Life Satisfaction.
+-->
+<!--   - The effect of Age.10yrs.C is reliably positive (beta = 0.09, SE = 5.26e-03, std. beta = 0.10, p < .001). Older people have greater Life Satisfaction -->
+<!--   - The effect of Deprivation (S) is  is reliably negative (beta = -0.07, SE = 5.50e-03, std. beta = -0.06, p < .001). People who are more deprived have lower Life Satisfaction. -->
+<!--   - The effect of Education (S) is reliably positive (beta = 0.06, SE = 6.33e-03, std. beta = 0.05, p < .001). Educated people have greater Life Satisfaction -->
+<!--   - The effect of being Employed is reliably positive (beta = 0.08, SE = 9.69e-03, std. beta = 0.07, p < .001). People who are employed tend to have greater Life Satisfaction -->
+<!--   - The effect of identifying as Maori is reliably negative (beta = -0.06, SE = 0.02, std.
+beta = -0.05, p < .01).
+People who belong to the minority Maori group tend to have lower Life Satisfaction.
+-->
+<!--   - The effect of identifying as Pacific is reliably negative (beta = -0.08, SE = 0.04, std.
+beta = -0.07, p < .05).
+People who belong to the minority Pacific peoples group tend to have lower life satisfaction.
+-->
+<!--   - The effect of identifying as Asian is not reliably different from people of European ethnicity in life satisfaction(beta = -0.04, SE = 0.03, std. beta = -0.03, p = 0.223). -->
+<!--   - The effect of identifying as male is reliably negative (beta = -0.19, SE = 0.01, std. beta = -0.16, p < .001). -->
+<!--   - The effect of having a partner is reliably positive (beta = 0.42, SE = 0.01, std. beta = 0.35, p < .001). -->
+<!--   - The effect of political conservativism (S) is reliably positive (beta = 0.01, SE = 4.23e-03, std. beta = 0.01, p < .01). -->
+<!--   - The effect of living in an urban area is reliably negative (beta = -0.02, SE = 8.85e-03, std. beta = -0.02, p < .05). -->
+<!-- ## Life Satisfaction: years X beliefs  -->
+<!-- ```{r fig.cap = "Prediction Plot Life Satisfaction Spirit Belief Longitudinal Models"} -->
+<!-- knitr::include_graphics("/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Figs/USElifsat_expected-1.png") -->
+<!-- ``` -->
+<!--   - The interaction of years $\times$ beliefs in *Spirit Excludes God* ludesGod is reliably positive (beta = 0.02, SE = 3.66e-03, std.
+beta = 0.03, p < .001).
+We infer that, unlike the Skeptic population, people with beliefs in a Spirit or Life Force who do not believe in a God tend to grow in their Life Satisfaction over time.
+-->
+<!--   - The interaction of years $\times$ *God And Spirit*  is reliably positive (beta = 0.01, SE = 3.32e-03, std.
+beta = 0.02, p < .001).
+We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their life satisfaction over time -->
+  <!--   - The effect of years $\times$ beliefs in *God Excludes Spirit*  is not reliable (beta = -1.59e-03, SE = 6.25e-03, std.
+beta = -2.70e-03, p = 0.799).
+Thus, God believers who doubt in a Spirit or Life Force start maintain higher Life-Satisfaction than do other groups, but they are not expected to grow in Life-Satisfaction over time.
+-->

### 2020-11-26 funs.R

@@ -93,7 +93,7 @@ beliefs_table <- function(x){
 
 
 # latex demographic table
-library("furniture")
+#library("furniture")
 demographic_table_latex <- function(x){
   furniture::table1(d_3, 
                     Age, 
@@ -143,7 +143,7 @@ demographic_table_latex <- function(x){
 
 
 graph_predictions <- function(x,y){ 
-  out <- ggeffects::ggpredict(model = x, terms = c("years [0:9]","Beliefs"), 
+  out <- ggeffects::ggpredict(model = x, terms = c("yearsC [minmax]","Beliefs"), 
                               ci.lvl = 0.95,
                               type = "fe",
                               typical = "mean",
@@ -154,7 +154,7 @@ graph_predictions <- function(x,y){
 }
 
 get_predictions <- function(x){ 
-  out <- ggeffects::ggpredict(model = x, terms = c("years [0:9]","Beliefs"), 
+  out <- ggeffects::ggpredict(model = x, terms = c("yearsC [minmax]","Beliefs"), 
                               ci.lvl = 0.95,
                               type = "fe",
                               typical = "mean",
@@ -210,6 +210,41 @@ table_model_latex_ls <- function(x){ # x is a model
          custom.note ="")
 }
 
+# dual table
+# 
+table_model_latex_dual <- function(x,y){
+  xtable <-texreg::extract(
+    x,
+    level = 0.95,
+    include.random = TRUE,
+    include.rsquared = F,
+    include.nobs = F,
+    include.loo.ic = F,
+    include.waic = F)
+ytable <-texreg::extract(
+  y,
+  level = 0.95,
+  include.random = TRUE,
+  include.rsquared = F,
+  include.nobs = F,
+  include.loo.ic = F,
+  include.waic = F)
+texreg(list(xtable,ytable),
+       custom.model.names = c("PWI","Life Sat"),
+       caption = "",
+       sideways = F,
+       scalebox = .5,
+       #fontsize= "footnotesize",
+       label = "tab:REGRESS_LS",
+       ci.force.level = 0.95, bold = 0.05,
+       settingstars = 0,
+       booktabs = TRUE,
+       custom.note ="")
+}
+
+
+
+
 # imputation functions
 amelia_imputation_clean <- function(x){
   set.seed(1234)
@@ -261,13 +296,12 @@ amelia_imputation_clean <- function(x){
[…]

### 2020-11-26 libs.R

@@ -9,10 +9,12 @@ library("gghighlight")
 library("Amelia")
 library("magrittr")
 library("patchwork")
-library("furniture")
+#library("furniture") # messes with table1
 library("table1")
 library("sjPlot")
 library("texreg")
+library("kableExtra")
+library("ggpubr")
 #library("viridis")
 #library("ggpubr")
 #library("wesanderson")

### 2020-11-26 Manuscript_Bayesian.Rmd
improved

@@
 The baseline for this indicator is Skepticism about a Spirit/Life Forces and Skepticism about a God.
-To handle the dependencies and heterogeneity introduced from the repeated measures, we include individual ID as an effect modelled as random.
+To handle the dependencies introduced from the repeated measures, we include individual ID as an effect modeled as random, and estimated a random effect for the slope for years.
+The model equations are given below and priors are given below.
 
@@
 
+Because missing values can result in biased estimates [@Blackwell2017-oq], we multiply imputed 10 datasets using the AmeliaII package in R[@Honaker2011-yu].We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences.
+Additionally the results were robust to filtering of participants of N > 3 to N = 8 waves.
 In a supplemental analysis we include standard demographic indicators used in New Zealand Attitudes and Values Research: [NZAVS modelling standards](https://cdn.auckland.ac.nz/assets/psych/about/our-research/nzavs/NZAVSTechnicalDocuments/NZAVS-Technical-Documents-e11-Default-Statistical-Standards.pdf).
@@
 We centred age was centred at and put into decade-long intervals.
-The model equation(s) are described in Equation 2.
+There are increasing worries
 
-Because missing values can result in biased estimates [@Blackwell2017-oq], we multiply imputed 10 datasets using the AmeliaII package in R[@Honaker2011-yu].We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences.
-Additionally the results were robust to filtering of participants of N >3 to N = 9 waves.
-Additionally, the findings reported below were robust to modeling only the effects of years and beliefs, without adjusting for the standard NZAVS demographic covariates.
-The supplement describing these analysis is located at: https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing.
+We provide the analyses of these additional analyses in the supplement.  The supplement describing these analysis is located at: https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing.
 
@@
 
-All code for this analysis and a full report of the missing data imputation can be found at https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing.
+All code for this analysis and a full report of the missing data imputation can be found at [https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing](https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing).
 
@@
 ```{r fig.cap = "Combination Coefficient Plots for personal well-being and life satisfaction longitudinal growth models"}
-knitr::include_graphics("/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Figs/USE_2_coefficientplots-1.png")
+knitr::include_graphics("/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Bayesian_pwi_predicted.png")
 ```
@@
 
+\begin{table}
+\begin{center}
+\scalebox{0.5}{
+\begin{tabular}{l c c}
+\toprule
+ & PWI & Life Sat \\
+\midrule
+Intercept                           & $\mathbf{7.04}$   & $\mathbf{5.19}$  \\
+                                    & $ [ 7.02;  7.07]$ & $ [ 5.17; 5.21]$ \\
+yearsC                              & $\mathbf{0.02}$   & $\mathbf{0.01}$  \\
+                                    & $ [ 0.02;  0.03]$ & $ [ 0.01; 0.02]$ \\
+Beliefs\_SpiritExcludesGod\_        & $\mathbf{-0.06}$  & $0.02$           \\
+                                    & $ [-0.09; -0.03]$ & $ [-0.00; 0.04]$ \\
+BeliefsGodAndSpirit                 & $\mathbf{-0.06}$  & $\mathbf{0.05}$  \\
+                                    & $ [-0.10; -0.03]$ & $ [ 0.03; 0.07]$ \\
+BeliefsGodExcludesSpirit            & $-0.00$           & $\mathbf{0.08}$  \\
+                                    & $ [-0.04;  0.04]$ & $ [ 0.04; 0.11]$ \\
+yearsC:Beliefs\_SpiritExcludesGod\_ & $\mathbf{0.01}$   & $\mathbf{0.01}$  \\
+                                    & $ [ 0.00;  0.02]$ & $ [ 0.01; 0.02]$ \\
+yearsC:BeliefsGodAndSpirit          & $\mathbf{0.01}$   & $\mathbf{0.01}$  \\
+                                    & $ [ 0.01;  0.02]$ & $ [ 0.00; 0.02]$ \\
+yearsC:BeliefsGodExcludesSpirit     & $0.00$            & $-0.00$          \\
+                                    & $ [-0.01;  0.02]$ & $ [-0.01; 0.01]$ \\
+\midrule
+SD: Idsd(Intercept)                 & $1.44$            & $0.99$           \\
+SD: Idsd(yearsC)                    & $0.13$            & $0.10$           \\
+SD: Idcor(Intercept,yearsC)         & $-0.02$           & $0.03$           \\
+\bottomrule
+\end{tabular}
+}
+\caption{}
+\label{tab:REGRESS}
+\end{center}
+\end{table}
+
 ```{r fig.cap = "Combination Coefficient Plots for personal well-being and life satisfaction longitudinal growth models"}
-knitr::include_graphics("/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Figs/USE_2_coefficientplots-1.png")
+knitr::include_graphics("/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Bayesian_pwi_predicted.png")
+```
+
+```{r fig.cap = "Combination Coefficient Plots for personal well-being and life satisfaction longitudinal growth models"}
+knitr::include_graphics("/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Bayesian_lifesat_predicted.png")
+```
 
@@
 1.
-At the outset of the study, in 2010, the expected subjective well-being as measured using the personal well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God ($\hat{y}$ =  6.78 [6.74, 6.82]) and lower among those who believed in both a Spirit/Life-Force and a God ($\hat{y}$ =  6.75 [6.72, 6.78]) in comparison both to the skeptical population who neither believed in a God nor a life force ($\hat{y}$ = 6.91 [6.86, 6.95]) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 6.88 [6.81, 6.96]).
+At the outset of the study, in 2010, the expected subjective well-being as measured using the personal well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God ($\hat{y} =$  6.80 [6.76, 6.85]) and lower among those who believed in both a Spirit/Life-Force and a God ($\hat{y}$ =  6.78 [6.74, 6.81]) in comparison both to the skeptical population who neither believed in a God nor a life force ($\hat{y}$ = 6.92 [6.87, 6.96]) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 6.90 [6.83, 6.98]).
 (See Figure 2)  Similarly,the expected subjective well-being as measured using the life satisfaction Inventory was lowest in 2010 among those who believed in Spirit or Life Force but not a God ($\hat{y}$ = 5.07 [5.04, 5.10]).
@@
 However, over time, belief in a Spirit or Life Force predicts a trajectory of growth in personal well-being.
-In 2018, personal well-being among who believe in Spirit or Life Force but not a God had grown to ($\hat{y} = 7.13$ [7.09, 7.16], $\Delta \hat{y} = 0.35$).
-Among those who believed in both a Spirit/Life-Force and a God, personal well-being had grown to ($\hat{y} =  7.14$, [7.11, 7.16], $\Delta \hat{y} = 0.39$).
-By contrast, personal well-being among the skeptical population who neither believed in a God nor a Life Force grew less steeply across the eight year interval $\hat{y} = 7.15 [7.12, 7.19]$, $\Delta \hat{y} = 0.24$).
-By the same token, personal well-being among those who believe in a God but are skeptical about a Spirit/Life Force grew less steeply $\hat{y} = 7.15$, [7.09, 7.21] $\Delta \hat{y} = 0.27$).
+In 2018, personal well-being among who believe in Spirit or Life Force but not a God had grown to ($\hat{y} =  7.12$ [7.09, 7.16], $\Delta \hat{y} = 0.32$).
+Among those who believed in both a Spirit/Life-Force and a God, personal well-being had grown to ($\hat{y} =  7.13$ [7.10, 7.16], $\Delta \hat{y} = 0.35$).
+By contrast, personal well-being among the skeptical population who neither believed in a God nor a Life Force grew less steeply across the eight year interval $\hat{y} = 7.14$ [7.11, 7.18], $\Delta \hat{y} = 0.22$).
+By the same token, personal well-being among those who believe in a God but are skeptical about a Spirit/Life Force grew less steeply $\hat{y} = 7.15$, [7.09, 7.21] $\Delta \hat{y} = 0.25$).
 Similarly, over time, belief in a Spirit or Life Force predicts a trajectory of growth in life satisfaction.
-In 2018, the expected effect of spiritual beliefs on life satisfaction among those who believe in Spirit or Life Force but not a God had grown to ($\hat{y} =  5.30$ [5.28, 5.33], $\Delta \hat{y} = 0.23$).
+In 2018, the expected effect of spiritual beliefs on life satisfaction among those who believe in Spirit or Life Force but not a God had grown to ($\hat{y} =  5.31$ [5.28, 5.33], $\Delta \hat{y} = 0.23$).
 In practical terms, such growth is equivalent to the expected gain in life satisfaction from being employed and a sixth as large as the expected benefit to life satisfaction from having a romantic partner.
-In practical termss, the rate of increase in personal wellbeing among those with Spirit/Life force beliefs was about one and half times greater than it was among those who lacked such beliefs (See Figure 2).
-Among those who believed in both a Spirit/Life-Force and a God, life satisfaction had grown to ($\hat{y} = 5.33 [5.31, 5.35]$, $\Delta \hat{y} = 0.22$).
-By contrast, life satisfaction among the skeptical population who neither believed in a God nor a Life Force lagged behind over the eight year interval $\hat{y} =  5.24 [5.21, 5.27], \Delta \hat{y} = 0.12$).
-Relatedly, life satisfaction in the population that believe in God but are skeptical about a Spirit/Life Force grew less steeply $\hat{y} =  5.32 [5.27, 5.36], \Delta \hat{y} = 0.13$).
+In practical terms, the rate of increase in personal wellbeing among those with Spirit/Life force beliefs was about one and half times greater than it was among those who lacked such beliefs (See Figure 2).
+Among those who believed in both a Spirit/Life-Force and a God, life satisfaction had grown to ($\hat{y} = 5.33 [5.31, 5.35]$, $\Delta \hat{y} = 0.24$).
+By contrast, life satisfaction among the skeptical population who neither believed in a God nor a Life Force lagged behind over the eight year interval $\hat{y} =  5.24 [5.21, 5.27], \Delta \hat{y} = 0.13$).
+Relatedly, life satisfaction in the population that believe in God but are skeptical about a Spirit/Life Force grew less steeply $\hat{y} =  5.31 [5.26, 5.36], \Delta \hat{y} = 0.11$).
 In practical terms, the rate of increase in life satisfaction among those with Spirit/Life force beliefs was twice as large (See Figure 3).
@@
 
-## Supplement
-
-## Personal Well-Being
-
-We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict personal well-being (PWI) with years (0-9), beliefs (four categories), Age.10yrs.C, deprivation (S), education (S), employed, ethnic categories, male, partner, political conservativism (S) and urban, with Id as a random effect.
-The model's intercept, corresponding to years = 0, beliefs = Skeptic, age = (sample mean), deprivation.S = (sample mean), education (S) =(sample mean), employed = not employed, ethnic Categories = European, male = not male, partner = no partner, political conservativism (S) = (sample mean), urban = not urban, is at 4.94 (SE = 0.02, 95% CI [4.89, 4.99], p < .001) is at 6.63 (SE = 0.03, 95% CI [6.57, 6.70], p < .001).
-Within this model:
-
-  - The effect of Years is positive is not reliable (beta = 2.66e-03, SE = 3.58e-03, std.
-beta = 3.19e-03, p = 0.458).
-We infer that personal well-being did not increase in the Skeptic population between 2010 and 2018.
-  - The effect of Beliefs in *Spirit Excludes God*  is reliably negative (beta = -0.14, SE = 0.03, std.
-beta = -0.03, p < .001).
-We infer that in 2010, the population that believed in a Spirit but did not believe in a God had lower personal well-being than did Skeptics.
-  - The effect of Beliefs in *God And Spirit*  is reliably negative (beta = -0.16, SE = 0.03, std.
-beta = -0.02, p < .001).
-We infer that in 2010, the population that believed in a God and a Spirit or Life Force had lower Personal Well-being than did Skeptics.
-  - The effect of Beliefs in *God Excludes Spirit*  is not reliable (beta = 0.02, SE = 0.05, std.
-beta = 0.02, p = 0.669).
-We infer that in 2010, the population that believed in a God and doubted a Spirit or Life Force similar levels of personal well-being to Skeptics.
-  - The effect of Age.10yrs.C is reliably positive  (beta = 0.08, SE = 7.38e-03, std. beta = 0.07, p < .001). Older people tend to have greater personal well-being.
-  - The effect of Deprivation (S) is reliably negative (beta = -0.14, SE = 7.50e-03, std. beta = -0.08, p < .001). Deprived people tend to have lower personal well-being.
-  - The effect of Education (S) is reliably positive  (beta = 0.12, SE = 8.74e-03, std. beta = 0.07, p < .001). Educated people tend to have greater personal well-being.
-  - The effect of Employed is reliably positive  (beta = 0.09, SE = 0.01, std. beta = 0.05, p < .001).
-  - The effect of identifying as M\={a}ori is reliably negative (beta = -0.30, SE = 0.03, std. beta = -0.18, p < .001).
-  - The effect of identifying as Pacific is reliably negative (beta = -0.29, SE = 0.05, std. beta = -0.17, p < .001). 
-  - The effect of identifying as Asian is not reliable (beta = -0.08, SE = 0.05, std. beta = -0.05, p = 0.089).
-  - The effect of identifying as male is reliably negative (beta = -0.11, SE = 0.02, std. beta = -0.06, p < .001). 
-  - The effect of having a partner is reliably positive (beta = 0.61, SE = 0.02, std. beta = 0.37, p < .001).
-  - The effect of political conservativism (S) is reliably positive  (beta = 0.02, SE = 5.67e-03, std. beta = 0.01, p < .001). 
-  - The effect of living in an Urban area is reliably negative (beta = -0.02, SE = 0.01, std. beta = -0.01, p < .05). T
-
-##  Personal well-being: years X beliefs
-
-```{r fig.cap = "Combination Coefficient plots for personal well-being and life satisfaction longitudinal growth models"}
-knitr::include_graphics("Figs/USEPWI_expected-1.png")
-```
-Focusing on the interaction of time with belief-type on personal well-being we observe:
-
-  - The effect of years $\times$ belief in a *Spirit Excludes God*  is reliably positive  (beta = 0.02, SE = 4.86e-03, std.
-beta = 0.02, p < .01).
-We infer that, unlike the skeptic population, people with beliefs in a spirit or life force who do not believe in a God tend to grow in their personal well-being over time.
-  - The effect of years $\times$ in *God And Spirit*  is reliably positive  (beta = 0.02, SE = 4.41e-03, std.
-beta = 0.03, p < .001).
-We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their personal well-being over time
-  - The effect of Years $\times$ belief in *God Excludes Spirit*  is not reliable(beta = 1.96e-03, SE = 8.29e-03, std.
-beta = 2.36e-03, p = 0.813).
-Thus, God believers who doubt in a Spirit or Life Force start maintain higher personal well-being than do other groups, but they are not expected to grow in personal well-being over time.
-
-## Life Satisfaction
-
-We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict personal well-being (PWI) with years (0-9), beliefs (four categories), Age.10yrs.C, deprivation (S), education (S), employed, ethnic categories, male, partner, political conservativism (S) and urban, with Id as a random effect.
-Standardized parameters were obtained by fitting the model on a standardized version of the dataset.
-The model's total explanatory power is substantial (conditional R2 = 0.69).
-The model's intercept, corresponding to years = 0, beliefs = Skeptic, age = (sample mean), deprivation.S = (sample mean), education (S) =(sample mean), employed = not employed, ethnic Categories = European, male = not male, partner = no partner, political conservativism (S) = (sample mean), urban = not urban is at 4.94 (SE = 0.02, 95% CI [4.89, 4.99], p < .001).
-Within this model:
-
-  - The effect of years is not reliable (beta = -5.50e-04, SE = 2.69e-03, std.
-beta = -9.32e-04, p = 0.838).
-We infer that life satisfaction did not increase in the Skeptic population between 2010 and 2018.
-  - The effect of beliefs in *Spirit Excludes God* is reliably negative (beta = -0.08, SE = 0.02, std.
-beta = 0.01, p < .001).
-We infer that overall the population that believes in a Spirit but does not believe in a God is expected to have a lower Life Satisfaction than skeptics.
-  - The effect of beliefs in *God And Spirit* is not reliable (beta = -0.04, SE = 0.02, std.
-beta = 0.04, p = 0.105).
-People who believe in God but doubt a Spirit/Life force are expected to have had similar levels of life satisfaction in Wave 10 as skeptics.
-  - The effect of beliefs in *God Excludes Spirit* is reliably positive (beta = 0.07, SE = 0.04, std.
-beta = 0.05, p = 0.062).We infer that overall the population that believes in a God but does not believe in a spirit is not reliably different from  the Skeptics population in Life Satisfaction.
-  - The effect of Age.10yrs.C is reliably positive (beta = 0.09, SE = 5.26e-03, std. beta = 0.10, p < .001). Older people have greater Life Satisfaction
-  - The effect of Deprivation (S) is  is reliably negative (beta = -0.07, SE = 5.50e-03, std. beta = -0.06, p < .001). People who are more deprived have lower Life Satisfaction.
-  - The effect of Education (S) is reliably positive (beta = 0.06, SE = 6.33e-03, std. beta = 0.05, p < .001). Educated people have greater Life Satisfaction
-  - The effect of being Employed is reliably positive (beta = 0.08, SE = 9.69e-03, std. beta = 0.07, p < .001). People who are employed tend to have greater Life Satisfaction
-  - The effect of identifying as Maori is reliably negative (beta = -0.06, SE = 0.02, std. beta = -0.05, p < .01).  People who belong to the minority Maori group tend to have lower Life Satisfaction.
-  - The effect of identifying as Pacific is reliably negative (beta = -0.08, SE = 0.04, std.
-beta = -0.07, p < .05).
-People who belong to the minority Pacific peoples group tend to have lower life satisfaction.
-  - The effect of identifying as Asian is not reliably different from people of European ethnicity in life satisfaction(beta = -0.04, SE = 0.03, std. beta = -0.03, p = 0.223).
-  - The effect of identifying as male is reliably negative (beta = -0.19, SE = 0.01, std. beta = -0.16, p < .001).
-  - The effect of having a partner is reliably positive (beta = 0.42, SE = 0.01, std. beta = 0.35, p < .001).
-  - The effect of political conservativism (S) is reliably positive (beta = 0.01, SE = 4.23e-03, std. beta = 0.01, p < .01).
-  - The effect of living in an urban area is reliably negative (beta = -0.02, SE = 8.85e-03, std. beta = -0.02, p < .05).
-
-## Life Satisfaction: years X beliefs
-
-```{r fig.cap = "Prediction Plot Life Satisfaction Spirit Belief Longitudinal Models"}
-knitr::include_graphics("/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Figs/USElifsat_expected-1.png")
-```
-  - The interaction of years $\times$ beliefs in *Spirit Excludes God* ludesGod is reliably positive (beta = 0.02, SE = 3.66e-03, std.
-beta = 0.03, p < .001).
-We infer that, unlike the Skeptic population, people with beliefs in a Spirit or Life Force who do not believe in a God tend to grow in their Life Satisfaction over time.
-  - The interaction of years $\times$ *God And Spirit*  is reliably positive (beta = 0.01, SE = 3.32e-03, std.
-beta = 0.02, p < .001).
-We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their life satisfaction over time
-  - The effect of years $\times$ beliefs in *God Excludes Spirit*  is not reliable (beta = -1.59e-03, SE = 6.25e-03, std.
-beta = -2.70e-03, p = 0.799).
-Thus, God believers who doubt in a Spirit or Life Force start maintain higher Life-Satisfaction than do other groups, but they are not expected to grow in Life-Satisfaction over time.
-
-## Imputed data analysis
-To adjust for potential bias from missing responses, we conducted a supplementary analysis in which we imputed missing values over the the longitudinal sample.
-This analysis is describe in Supplement 1.
-Averaging over the missing responses does not change inference, however we observe growth across all four groups in both personal well-being and life satisfaction.
-Additionally we observe substantially stronger effects for the slope of Spirit Beliefs on both personal well-being and life-satisfaction.
-Figures 3 and Figures 4 describe the predicted slopes of change.
-The inference from this analysis remains the same as from the analysis in which we delete missing values row-wise, by panel wave.
-In all analyses, those who believe in a spirit or life-force experience greater growth in both personal well-being and life satisfaction over time.
-
-Figure 3 about here:
-
-```{r fig.cap = "Imputed data analysis: coefficient plot"}
-knitr::include_graphics("Figs/coefficient_plot_both_imputed-1.png")
-```
-
-Figure 4 about here:
-
-```{r fig.cap = "Imputed data analysis: predicted personal wellbeing by years X beliefs"}
-knitr::include_graphics("/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Figs/predicted_PWI_impute-1.png")
-```
-
-Figure 5 about here:
-
-```{r fig.cap = "Imputed data analysis: predicted personal wellbeing by years X beliefs"}
-knitr::include_graphics("/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Figs/predicted_lifesat_imputed-1.png")
-```
+# Appendix
 
 \newpage
-# Supplement Model Equations
+## Model Equations
 
-## Priors for the Personal Wellbeing model
+### Priors for the Personal Wellbeing model
 
@@
 \newpage
-## Priors for the Life Satisfaction model
+
+### Priors for the Life Satisfaction model
 
 Model priors for Life Satisfaction Model with including group-level ID intercept and belief slopes are as follows:
+
 $$
@@
 \boldsymbol{\alpha} \sim \alpha_{0} +\alpha_j \nonumber\\
-\alpha_{0} \sim \text{StudentT}(3,5.5, 2.5)\nonumber\\
-\alpha_j \sim \text{StudentT}+(3,0,10)\nonumber\\
+\alpha_{0} \sim \text{StudentT}(3,5.5,2.5)\nonumber\\
+\alpha_j \sim \text{StudentT}+(3,0,2.5)\nonumber\\
 \boldsymbol{\beta}\sim \boldsymbol{b} + \beta_j \nonumber\\
@@
 
-## Deterministic model
+### Deterministic model
 

### 2020-11-27 Manuscript_Bayesian.Rmd
close to final revision - ben to proof on overleaf.

@@
 ---
-title             : "Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Belief"
+title             : "National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Belief"
 shorttitle        : "Longitudinal Wellbeing from Spiritual Beliefs"
@@
 keywords          : "Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being"
-wordcount         : "5415 words in text body, 1155 words in reference section"
+wordcount         : "4685 words in text body, 1191 words in reference section"
 bibliography      : ["b-references.bib"]
@@
 figurelist        : no
-figsintext        : no
+figsintext        : yes
 tablelist         : no
 footnotelist      : no
-linenumbers       : no
+linenumbers       : yes
 mask              : no
@@
 classoption       : "man"
-output            : papaja::apa6_pdf
+output            : papaja::apa6_word
 extra_dependencies: ["rotating", "booktabs", "graphicx","longtable","amsmath","amsfonts","amssymb"]
-# header-includes:
-#    - \usepackage{tcolorbox}
-#    - \usepackage{paralist}
-#    - \usepackage{multicol}
-#    - \usepackage{graphics}
-#    - \usepackage{rotating}
-#    - \usepackage{longtable}
-#    - \usepackage[amsmath]
+
 ---
@@
 ```
+```{r global_options, echo=FALSE, include = FALSE}
+knitr::opts_chunk$set(#fig.path="Figs/", 
+                      message=FALSE, 
+                      warning=FALSE,
+                      collapse =TRUE,
+                      echo=TRUE, 
+                      #results="hide", 
+                     # fig.width= 10,
+                     # fig.height=8,
+                      tidy = "styler")
+```
 
-# ```{r runanalysis, include = FALSE}
-# source('short.R')
-# ```
 The relationship between spirituality and psychological well-being is a matter of enduring fascination [@Ellsworth2010-yu].
@@
 ## Measures
+
+Table 1 about here
+
+\begin{table}[ ht ] 
+\centering 
+\caption{}\label{}
+\scalebox{0.65}{
+\begin{tabular}{ l c c c c c c c c c }
+\toprule
+ &   \multicolumn{ 9 }{c}{ Wave }\\ 
+  & 2010 & 2011 & 2012 & 2013 & 2014 & 2015 & 2016 & 2017 & 2018 \\ 
+ & n = \emph{missing} & n = \emph{missing} & n = 8547 & n = 13472 & n = \emph{missing} & n = 11687 & n = 16048 & n = 14174 & n = 14883 \\ 
+ \midrule
+Age &   &   &   &   &   &   &   &   &  \\ 
+NZdep &   &   &   &   &   &   &   &   &  \\ 
+Education &   &   &   &   &   &   &   &   &  \\ 
+Employed &   &   &   &   &   &   &   &   &  \\ 
+\hspace{6pt}    0 &  (\%) &  (\%) & 2094 (24.5\%) & 2955 (21.9\%) &  (\%) & 2525 (21.6\%) & 3486 (21.7\%) & 3223 (22.7\%) & 3404 (22.9\%)\\ 
+\hspace{6pt}    1 &  (\%) &  (\%) & 6453 (75.5\%) & 10517 (78.1\%) &  (\%) & 9162 (78.4\%) & 12562 (78.3\%) & 10951 (77.3\%) & 11479 (77.1\%)\\ 
+Ethnic\_Categories &   &   &   &   &   &   &   &   &  \\ 
+\hspace{6pt}    Euro &  (\%) &  (\%) & 6731 (78.8\%) & 11024 (81.8\%) &  (\%) & 9671 (82.8\%) & 13389 (83.4\%) & 11814 (83.3\%) & 12427 (83.5\%)\\ 
+\hspace{6pt}    Maori &  (\%) &  (\%) & 1188 (13.9\%) & 1608 (11.9\%) &  (\%) & 1341 (11.5\%) & 1749 (10.9\%) & 1620 (11.4\%) & 1677 (11.3\%)\\ 
+\hspace{6pt}    Pacific &  (\%) &  (\%) & 284 (3.3\%) & 339 (2.5\%) &  (\%) & 277 (2.4\%) & 331 (2.1\%) & 243 (1.7\%) & 261 (1.8\%)\\ 
+\hspace{6pt}    Asian &  (\%) &  (\%) & 344 (4\%) & 501 (3.7\%) &  (\%) & 398 (3.4\%) & 579 (3.6\%) & 497 (3.5\%) & 518 (3.5\%)\\ 
+Male &   &   &   &   &   &   &   &   &  \\ 
+\hspace{6pt}    0 &  (\%) &  (\%) & 5302 (62\%) & 8457 (62.8\%) &  (\%) & 7325 (62.7\%) & 10148 (63.2\%) & 9011 (63.6\%) & 9433 (63.4\%)\\ 
+\hspace{6pt}    1 &  (\%) &  (\%) & 3245 (38\%) & 5015 (37.2\%) &  (\%) & 4362 (37.3\%) & 5900 (36.8\%) & 5163 (36.4\%) & 5450 (36.6\%)\\ 
+Has\_Partner &   &   &   &   &   &   &   &   &  \\ 
+\hspace{6pt}    0 &  (\%) &  (\%) & 2408 (28.2\%) & 3546 (26.3\%) &  (\%) & 2815 (24.1\%) & 3773 (23.5\%) & 3330 (23.5\%) & 3556 (23.9\%)\\ 
+\hspace{6pt}    1 &  (\%) &  (\%) & 6139 (71.8\%) & 9926 (73.7\%) &  (\%) & 8872 (75.9\%) & 12275 (76.5\%) & 10844 (76.5\%) & 11327 (76.1\%)\\ 
+Political\_Conservativism &   &   &   &   &   &   &   &   &  \\ 
+Urban &   &   &   &   &   &   &   &   &  \\ 
+\hspace{6pt}    Not\_Urban &  (\%) &  (\%) & 2865 (33.5\%) & 4278 (31.8\%) &  (\%) & 3951 (33.8\%) & 5538 (34.5\%) & 2610 (18.4\%) & 2661 (17.9\%)\\ 
+\hspace{6pt}    Urban &  (\%) &  (\%) & 5682 (66.5\%) & 9194 (68.2\%) &  (\%) & 7736 (66.2\%) & 10510 (65.5\%) & 11564 (81.6\%) & 12222 (82.1\%)\\ 
+Beliefs &   &   &   &   &   &   &   &   &  \\ 
+\hspace{6pt}    \_Skeptic\_ &  (\%) &  (\%) & 2048 (24\%) & 3286 (24.4\%) &  (\%) & 2951 (25.3\%) & 4116 (25.6\%) & 3916 (27.6\%) & 4120 (27.7\%)\\ 
+\hspace{6pt}    \_SpiritExcludesGod\_ &  (\%) &  (\%) & 2295 (26.9\%) & 3735 (27.7\%) &  (\%) & 3169 (27.1\%) & 4501 (28\%) & 3918 (27.6\%) & 3970 (26.7\%)\\ 
+\hspace{6pt}    GodAndSpirit &  (\%) &  (\%) & 3707 (43.4\%) & 5499 (40.8\%) &  (\%) & 5119 (43.8\%) & 6453 (40.2\%) & 5464 (38.5\%) & 6210 (41.7\%)\\ 
+\hspace{6pt}    GodExcludesSpirit &  (\%) &  (\%) & 497 (5.8\%) & 952 (7.1\%) &  (\%) & 448 (3.8\%) & 978 (6.1\%) & 876 (6.2\%) & 583 (3.9\%)\\ 
+LIFESAT &   &   &   &   &   &   &   &   &  \\ 
+PWI &   &   &   &   &   &   &   &   &  \\ 
+\bottomrule
+
+\end{tabular}
+}
+\end{table}
 
@@
 ## Potential Confounders and Missing Data 
-
 Because missing values can result in biased estimates [@Blackwell2017-oq], we multiply imputed 10 datasets using the AmeliaII package in R[@Honaker2011-yu].We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences.
@@
 
-### System and packages
-
+## System and packages
 The analysis was performed using R version 4.0.2 (2020-06-22). The Platform was x86_64-apple-darwin17.0 (64-bit) Running under: macOS Catalina 10.15.6
@@
 ## Code
-
 All code for this analysis and a full report of the missing data imputation can be found at [https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing](https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing).
@@
 # Results
-
-```{r fig.cap = "Combination Coefficient Plots for personal well-being and life satisfaction longitudinal growth models"}
-knitr::include_graphics("/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Bayesian_pwi_predicted.png")
-```
 
@@
 \begin{center}
-\scalebox{0.5}{
+\scalebox{0.8}{
 \begin{tabular}{l c c}
@@
 
-```{r fig.cap = "Combination Coefficient Plots for personal well-being and life satisfaction longitudinal growth models"}
-knitr::include_graphics("/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Bayesian_pwi_predicted.png")
+Figure 1 about here
+```{r fig.cap = "Coefficient Graphs", echo =FALSE,out.width="\\textwidth"}
+knitr::include_graphics("/Users/jbul176/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/coefficientplotsbayes.png")
 ```
 
-```{r fig.cap = "Combination Coefficient Plots for personal well-being and life satisfaction longitudinal growth models"}
-knitr::include_graphics("/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Bayesian_lifesat_predicted.png")
+## Personal Well-Being
+We fitted a linear mixed effects model using the brms package [@BRMSpackage] in R [@R-base] to predict personal well-being (PWI) with years (0-9), beliefs (four categories), with Id as random intercept and years as a random slope effect (see below: model equations).
+
+Within this model:
+
+  - The effect of years is positive is reliable (posterior median = 0.025, HDI [0.020, 0.031] pd = 100.00%).
+We infer that personal well-being increased in the skeptic population between 2010 and 2018.
+  - The effect of beliefs in *Spirit Excludes God*  is reliably negative (posterior median = -0.058, HDI [-0.080, -0.035], pd = 100.00%).
+We infer that the population that believed in a Spirit but did not believe in a God was lower in personal well-being at the intercept compared to skeptics.
+  - The effect of beliefs in *God And Spirit*  is reliably negative (posterior median = -0.064, HDI [-0.091, -0.037], pd = 100.00%).
+We infer that the population that believed in a God and a Spirit or Life Force was lower in personal well-being at the intercept compared to skeptics.
+  - The effect of beliefs in *God Excludes Spirit*  is not reliable (posterior median = -0.002 HDI [-0.039,  0.033], pd = 53.50%).
+We infer that the population that believed in a God and doubted a Spirit or Life Force were similar were similar in personal well-being at the intercept to skeptics.
+
+## Personal Well-Being: years X beliefs
+
+```{r fig.cap = "Personal Well-being prediction graph", echo =FALSE, out.width="\\textwidth"}
+knitr::include_graphics("/Users/jbul176/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/Bayesian_pwi_predicted.png")
 ```
+
+The effect of years $\times$ belief in a *Spirit Excludes God*  is reliably positive  (posterior median = 0.010, HDI [ 0.003,  0.017], pd = 99.02%).
+We infer that people with beliefs in a spirit or life force who do not believe in a God tended to grow in their personal well-being over time at a faster rate than skeptics grew.
+  - The effect of years $\times$ in *God And Spirit*  is reliably positive  (posterior median = 0.014, HDI [0.008, 0.021], pd = 99.98%).
+We infer that people who believe in a spirit or life force who also believe in a God tended to grow in their personal well-being over time at a faster rate than skeptics grew.
+  - The effect of years $\times$ belief in *God Excludes Spirit*  is not reliable (posterior median = 0.002, HDI [-0.009,  0.015], pd =  63.73%).
+We infer that people who believe in a God but not a spirit or life force did not grow at a faster rate in personal well-being than skeptics grew.
+
+## Life Satisfaction
+
+We fitted a linear mixed effects model using the brms package [@BRMSpackage] in R [@R-base] to predict personal well-being (PWI) with years (0-9), beliefs (four categories), with Id as random intercept and years as a random slope effect (see model equations, below).
+
+Within this model:
+
+  - The effect of years is positive is reliable (posterior median =  0.014, HDI [ 0.010, 0.019], pd = 100.00%).
+We infer that life satisfaction increased in the skeptic population between 2010 and 2018.
+  -The effect of beliefs in *Spirit Excludes God*  is reliably positive (posterior median =   0.018, HDI [ 0.001, 0.035], pd = 96.05%).
+We infer that the population that believed in a Spirit but did not believe in a God was higher in life satisfaction at the intercept compared with skeptics.
+  - The effect of beliefs in *God And Spirit*  is reliably positive (posterior median = 0.051, HDI [ 0.031, 0.071], pd = 100.00%).
+We infer that the population that believed in a God and a Spirit or Life Force was higher in life satisfaction at the intercept compared with skeptics.
+  - The effect of beliefs in *God Excludes Spirit*  is reliably positive (posterior median =  0.076, HDI [ 0.048, 0.102], pd = 100.00%).
+We infer that the population that believed in a God and doubted a Spirit or Life Force  was higher in life satisfaction at the intercept compared with skeptics.
+
+## Life Satisfaction: years X beliefs
+
+```{r  fig.cap = "Life satisfaction prediction graph", echo = FALSE,out.width="\\textwidth"}
+knitr::include_graphics("/Users/jbul176/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/Bayesian_lifesat_predicted.png")
+```
+  - The interaction of years $\times$ beliefs in *Spirit Excludes God* is reliably positive (posterior median = 0.012, HDI [ 0.007, 0.017], pd = 100.00%).
+We infer that people with beliefs in a Spirit or Life Force who do not believe in a God tended to grow over time in their life satisfaction at a faster rate than skeptics.
+  - The interaction of years $\times$ *God And Spirit*  is reliably positive (posterior median =  0.010, HDI [ 0.004, 0.015], pd = 99.90%).
+We infer that people who believe in a spirit or life force who also believe in a God tended to grow over time in their life satisfaction at a faster rate than skeptics.
+  - The effect of years $\times$ beliefs in *God Excludes Spirit*  is not reliable (posterior median = -0.002, HDI [-0.011, 0.007], pd = 62.80%).
+We infer that people who believe in a God but not a spirit or life force tended to grow over time in their life satisfaction at a similar rate to skeptics.
 
@@
 ### Authors contributions
-BH conceived of the study.
-BH wrote the first draft of the manuscript, and performed the initial data analysis with input from JB.
+BH and JB conceived of the study.
+BH wrote the first draft of the manuscript.
 EW improved the approach with BH and JB and revised the manuscript.
@@
 CS was responsible for NZAVS data collection and revised the manuscript.
-JB improved the approach with BH and EW, performed a data re-analysis, and revised the manuscript.
+JB performed the data re-analysis and revised the manuscript.
 
@@
 # Appendix 
-
-\newpage
 ## Model Equations
-
 ### Priors for the Personal Wellbeing model

### 2020-11-27 Supplement_And_Analysis.Rmd
close to final revision - ben to proof on overleaf.

@@
 ---
-title             : "Supplement: Analysis for Longitudinal Evidence for Growth in Subjective well-being from Spiritual Belief"
+title             : "Supplement: Analysis for National Longitudinal Evidence for Growth in Subjective well-being from Spiritual Belief"
 author: "contact: joseph.bulbulia@gmail.com on behalf of Highland et al."
@@
 
-```{r latex_data_table, include=TRUE}
+```{r latex_data_table, include=FALSE}
 # you might need to install the dev verion.//remotes::install_github("tysonstanley/furniture")
-#demographic_table_latex(d_3) # note the furniture package conflicts with the table1 package. 
+demographic_table_latex(d_3) # note the furniture package conflicts with the table1 package. 
+conflicts()
+furniture::table1(d_3, splitby = ~Wave,
+                    Age, 
+                    NZdep,
+                    Education,
+                    Employed,
+                    Ethnic_Categories,
+                    Male,Has_Partner, 
+                    Political_Conservativism,
+                    Urban, 
+                    Beliefs, 
+                    LIFESAT,
+                    PWI,
+                    overall=F, 
+                    output = "latex2", 
+                    booktabs=TRUE)
 ```
@@
 
+```{r simple_table, cache =TRUE}
+bayestestR::describe_posterior(mod.1_b, ci = 0.90,  test = c("p_direction"),diagnostic = "ESS")
+```
+
+```{r simple_table_ls, cache =TRUE}
+bayestestR::describe_posterior(mod.2_b, ci = 0.90, test = c("p_direction"), diagnostic = "ESS")
+```
+
 <!-- Tables: -->
@@
 
+```{r}
+print(citation("brms"), bibtex=TRUE)
+```
+
 ## Data
@@
 We areare interested in collaboration with qualified researchers: please contact Chris G.
-Sibley: [c.sibley@auckland.ac.nz](c.sibley@auckland.ac.nz).
-
-## Credits
-Package credits:
-
-```{r, sessioninfo, include=FALSE}
-#  custom function for citing packages. 
-cite_packages()
-```
+Sibley [c.sibley@auckland.ac.nz](c.sibley@auckland.ac.nz).
 

### 2020-11-27 Manuscript_Bayesian.tex
ends and pieces committing

--- 2020-11-20 Manuscript.tex
@@
   english,
-  man]{apa6}
+  man,floatsintext]{apa6}
 \usepackage{lmodern}
@@
 \hypersetup{
-  pdftitle={National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Beliefs},
+  pdftitle={National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Belief},
   pdfauthor={Benjamin Highland1, Everett L. Worthington2, Don E Davis3, Chris G. Sibley4, \& Joseph A. Bulbulia5},
@@
 
-% Overwrite redefinition of paragraph and subparagraph by the default LaTeX template
-% See https://github.com/crsh/papaja/issues/292
-\makeatletter
-\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
-  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
-  {-1em}%
-  {\normalfont\normalsize\bfseries\itshape\typesectitle}}
-
-\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
-  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
-  {-\z@\relax}%
-  {\normalfont\normalsize\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
-\makeatother
-
 % \usepackage{etoolbox}
@@
 \shorttitle{Longitudinal Wellbeing from Spiritual Beliefs}
-\keywords{Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being\newline\indent Word count: 5415 words in text body, 1155 words in reference section}
+\keywords{Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being\newline\indent Word count: 4685 words in text body, 1191 words in reference section}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 \makeatother
+\usepackage{lineno}
+
+\linenumbers
 \usepackage{csquotes}
-\usepackage{tcolorbox}
-\usepackage{paralist}
-\usepackage{multicol}
-\usepackage{graphics}
-\usepackage{rotating}
-\usepackage{longtable}
 \ifxetex
@@
 
-\title{National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Beliefs}
+\title{National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Belief}
 \author{Benjamin Highland\textsuperscript{1}, Everett L. Worthington\textsuperscript{2}, Don E Davis\textsuperscript{3}, Chris G. Sibley\textsuperscript{4}, \& Joseph A. Bulbulia\textsuperscript{5}}
@@
 For example, The \emph{Daily Spiritual Experience Scale}, which is widely used in spiritual mental health research, includes items such as \enquote{I feel thankful for my blessings} and \enquote{I feel deep inner peace or harmony} to measure spirituality (Underwood \& Teresi, 2002).
-As Koenig observes, it is unsurprising that spirituality predict well-being when the same items are used to measure both constructs.
-Koenig aptly describes this approach as \enquote{tautological} (Koenig, 2008).
-Indeed, recent systematic-review of spirituality/mental health research found that nearly 45\% of previous studies employ tautological scales (Garssen et al., 2016).
+These items are used both to measure spirituality and subjective well-being.
+As Koenig observes, it is unsurprising that spirituality is associated with well-being when the same items are used to measure both constructs.
+Koenig aptly describes this problematic approach as \enquote{tautological} (Koenig, 2008).
+How pervasive is this problem?
+A recent systematic-review of spirituality/mental health research found that nearly 45\% of previous studies employ tautological scales (Garssen et al., 2016).
 The first challenge, then, is to define measures of spirituality that are not well-being measures dressed up in other words.
@@
 
-Here, we addresses the three challenges of previous in the following ways: First, to avoid a tautological elision of \enquote{spirituality} and \enquote{well-being} we adopt a non-affective measure of spirituality as a \enquote{believe in a Spirit or Life Force.} On the other side, to measure subjective well-being, we use a well-validated scale for personal well-being and a well-validated scale for Life-Satisfaction (see Method).
+Here, we addresses these three challenges of previous research in the following ways: First, to avoid a tautological elision of \enquote{spirituality} and \enquote{well-being} we adopt a non-affective measure of spirituality as a \enquote{believe in a Spirit or Life Force.} On the other side, to measure subjective well-being, we use a well-validated scale for personal well-being and a well-validated scale for Life-Satisfaction (see Method).
 Second, to disentangle spiritual beliefs from traditional religious beliefs and skeptical beliefs we adopt a four-level categorical indicator combining the binary indicator of belief in a Spirit or Life Force with the binary indicator \enquote{Do you believe in some form of a spirit or life force?} The levels of this four-level categorical indicator for spiritual, religious, and skeptical beliefs are as follows: (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who nevertheless believe in a Spirit or Life Force; (3) Those who believe in a God but do not believe in a Spirit or Life force; (4) Those who who believe in a both God and a Spirit or Life Force.
@@
 
-\hypertarget{life-satisfaction}{%
-\subsubsection{Life Satisfaction}\label{life-satisfaction}}
-
-Life satisfaction is a measure of emotional well-being that was assessed using a 2-item version of the Satisfaction With Life Scale, which has previously been shown to correlate with aspects of religiosity (Diener, Emmons, Larsen, \& Griffin, 1985).
-Participants rated their agreement with the statements (a) \enquote{I am satisfied with my life}; and (b) \enquote{In most ways my life is close to ideal.} The items were rated on 7-point response options ranging from 1 = strongly disagree to 7 = strongly agree.
-The mean Cronbach's alpha for this 2-item scale was \(\alpha\) = 0.90.
-Higher scores on this scale indicate higher life satisfaction.
-Overall scores were means of the 2-items.
-For detailed information pertaining to means, standard deviations, and missingness, see Table 1.
-
-\hypertarget{personal-well-being}{%
-\subsubsection{Personal well-being}\label{personal-well-being}}
-
-Personal well-being is a measure of function all well-being that was assessed using a 4-item version of the Australian Unity well-being Index (Cummins, Eckerseley, Pallant, Van Vugt, \& Misajon, 2017).
-Participants rated their satisfaction with (a) \enquote{Your standard of living}; (b) \enquote{Your health}; (c) \enquote{Your future security}; (d) \enquote{Your personal relationships.} Items were rated using a 10-point response option ranging from 1 = completely dissatisfied to 10 = completely satisfied.
-The mean Cronbach's alpha for this scale was \(\alpha\) = 0.90.
-Higher scores on this scale indicate higher personal well-being.
-Overall scores were means of the 4-items.
-For detailed information pertaining to means, standard deviations, and missingness, see Table 1.
-
-\hypertarget{belief-in-a-spirit-or-a-god}{%
-\subsubsection{Belief in a Spirit or a God}\label{belief-in-a-spirit-or-a-god}}
-
-We assess belief in a spirit or a God by asking two non-affective questions: \enquote{Do you believe in a spirit or life force} and \enquote{Do you believe in a god?} Responses were coded as (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who believe in a Spirit or Life Force; (3) Believers in a God who disbelieve in a Spirit or Life force; (4) Believers in a God and Spirit or Life Force.
-These believe were developed for the NZAVS from the 2005 Eurobarometer.
-For detailed information pertaining to beliefs, see Table 1.
-
-Table 1 about here:
-\[
+Table 1 about here
+
 \begin{table}[ ht ] 
@@
 \caption{}\label{}
-\scalebox{.6}{
+\scalebox{0.65}{
 \begin{tabular}{ l c c c c c c c c c }
@@
 \hspace{6pt}    1 &  (\%) &  (\%) & 6453 (75.5\%) & 10517 (78.1\%) &  (\%) & 9162 (78.4\%) & 12562 (78.3\%) & 10951 (77.3\%) & 11479 (77.1\%)\\ 
-EthnicCategories &   &   &   &   &   &   &   &   &  \\ 
+Ethnic\_Categories &   &   &   &   &   &   &   &   &  \\ 
 \hspace{6pt}    Euro &  (\%) &  (\%) & 6731 (78.8\%) & 11024 (81.8\%) &  (\%) & 9671 (82.8\%) & 13389 (83.4\%) & 11814 (83.3\%) & 12427 (83.5\%)\\ 
@@
 \hspace{6pt}    1 &  (\%) &  (\%) & 3245 (38\%) & 5015 (37.2\%) &  (\%) & 4362 (37.3\%) & 5900 (36.8\%) & 5163 (36.4\%) & 5450 (36.6\%)\\ 
-Partner &   &   &   &   &   &   &   &   &  \\ 
+Has\_Partner &   &   &   &   &   &   &   &   &  \\ 
 \hspace{6pt}    0 &  (\%) &  (\%) & 2408 (28.2\%) & 3546 (26.3\%) &  (\%) & 2815 (24.1\%) & 3773 (23.5\%) & 3330 (23.5\%) & 3556 (23.9\%)\\ 
 \hspace{6pt}    1 &  (\%) &  (\%) & 6139 (71.8\%) & 9926 (73.7\%) &  (\%) & 8872 (75.9\%) & 12275 (76.5\%) & 10844 (76.5\%) & 11327 (76.1\%)\\ 
-Pol Orient &   &   &   &   &   &   &   &   &  \\ 
+Political\_Conservativism &   &   &   &   &   &   &   &   &  \\ 
 Urban &   &   &   &   &   &   &   &   &  \\ 
@@
 \bottomrule
+
 \end{tabular}
@@
 \end{table}
-\]
-
-\hypertarget{demographic-indicators}{%
-\subsection{Demographic Indicators}\label{demographic-indicators}}
-
-We included standard demographic indicators used in New Zealand Attitudes and Values Research: \href{https://cdn.auckland.ac.nz/assets/psych/about/our-research/nzavs/NZAVSTechnicalDocuments/NZAVS-Technical-Documents-e11-Default-Statistical-Standards.pdf}{NZAVS modelling standards}
+
+\hypertarget{life-satisfaction}{%
+\subsubsection{Life Satisfaction}\label{life-satisfaction}}
+
+Life satisfaction is a measure of emotional well-being that was assessed using a 2-item version of the Satisfaction With Life Scale, which has previously been shown to correlate with aspects of religiosity (Diener, Emmons, Larsen, \& Griffin, 1985).
+Participants rated their agreement with the statements (a) \enquote{I am satisfied with my life}; and (b) \enquote{In most ways my life is close to ideal.} The items were rated on 7-point response options ranging from 1 = strongly disagree to 7 = strongly agree.
+The mean Cronbach's alpha for this 2-item scale was \(\alpha\) = 0.90.
+Higher scores on this scale indicate higher life satisfaction.
+Overall scores were means of the 2-items.
+For detailed information pertaining to means, standard deviations, and missingness, see Table 1.
+
+\hypertarget{personal-well-being}{%
+\subsubsection{Personal well-being}\label{personal-well-being}}
+
+Personal well-being is a measure of function all well-being that was assessed using a 4-item version of the Australian Unity well-being Index (Cummins, Eckerseley, Pallant, Van Vugt, \& Misajon, 2017).
+Participants rated their satisfaction with (a) \enquote{Your standard of living}; (b) \enquote{Your health}; (c) \enquote{Your future security}; (d) \enquote{Your personal relationships.} Items were rated using a 10-point response option ranging from 1 = completely dissatisfied to 10 = completely satisfied.
+The mean Cronbach's alpha for this scale was \(\alpha\) = 0.90.
+Higher scores on this scale indicate higher personal well-being.
+Overall scores were means of the 4-items.
+For detailed information pertaining to means, standard deviations, and missingness, see Table 1.
+
+\hypertarget{belief-in-a-spirit-or-a-god}{%
+\subsubsection{Belief in a Spirit or a God}\label{belief-in-a-spirit-or-a-god}}
+
+We assess belief in a spirit or a God by asking two non-affective questions: \enquote{Do you believe in a spirit or life force} and \enquote{Do you believe in a god?} Responses were coded as (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who believe in a Spirit or Life Force; (3) Believers in a God who disbelieve in a Spirit or Life force; (4) Believers in a God and Spirit or Life Force.
+These believe were developed for the NZAVS from the 2005 Eurobarometer.
+For detailed information pertaining to beliefs, see Table 1.
+
+Table 1 about here:
+
+\hypertarget{sample-demographics}{%
+\subsection{Sample Demographics}\label{sample-demographics}}
 
@@
 The baseline for this indicator is Skepticism about a Spirit/Life Forces and Skepticism about a God.
-To handle the dependencies and heterogeneity introduced from the repeated measures, we include individual ID as an effect modelled as random.
+To handle the dependencies introduced from the repeated measures, we include individual ID as an effect modeled as random, and estimated a random effect for the slope for years.
+The model equations are given below and priors are given below.
+
+\hypertarget{potential-confounders-and-missing-data}{%
+\subsection{Potential Confounders and Missing Data}\label{potential-confounders-and-missing-data}}
+
+Because missing values can result in biased estimates (Blackwell, Honaker, \& King, 2017), we multiply imputed 10 datasets using the AmeliaII package in R(Honaker et al., 2011b).We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences.
+Additionally the results were robust to filtering of participants of N \textgreater{} 3 to N = 8 waves.
+In a supplemental analysis we include standard demographic indicators used in New Zealand Attitudes and Values Research: \href{https://cdn.auckland.ac.nz/assets/psych/about/our-research/nzavs/NZAVSTechnicalDocuments/NZAVS-Technical-Documents-e11-Default-Statistical-Standards.pdf}{NZAVS modelling standards}.
 We centered and scaled the education, deprivation, and political conservativism.
 We centred age was centred at and put into decade-long intervals.
-The model equation(s) are described in Equation 1.
-
-\hypertarget{equation-1}{%
-\subsection{Equation 1}\label{equation-1}}
-
-\[
-\begin{aligned}
-\operatorname{{\small LIFESAT/PWI}}  &\sim N \left(\mu,\sigma^2 \right) \\ \mu &=\alpha_{j[i]} + \beta_{1}(\operatorname{Years}) + \beta_{2}(\operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\
-&\quad \beta_{3}(\operatorname{Beliefs}_{\operatorname{GodAndSpirit}}) + \beta_{4}(\operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) + \beta_{5}(\operatorname{Age.10yrs.C})\ + \\
-&\quad \beta_{6}(\operatorname{Deprivation.S}) + \beta_{7}(\operatorname{Edu.S}) + \beta_{8}(\operatorname{Employed})\ + \\
-&\quad \beta_{9}(\operatorname{EthnicCats}_{\operatorname{Maori}}) + \beta_{10}(\operatorname{EthnicCats}_{\operatorname{Pacific}}) + \beta_{11}(\operatorname{EthnicCats}_{\operatorname{Asian}})\ + \\
-&\quad \beta_{12}(\operatorname{Male}_{\operatorname{1}}) + \beta_{13}(\operatorname{Partner}_{\operatorname{Has\ }}) + \beta_{14}(\operatorname{Pol.Orient.S})\ + \\
-&\quad \beta_{15}(\operatorname{Urban}) + \beta_{16}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}}) + \beta_{17}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ + \\
-&\quad \beta_{18}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) \\ \alpha_{j} &\sim N \left(\mu_{\alpha_{j}},\sigma^2_{\alpha_{j}} \right) , \operatorname{ for  Id }~j~= 1, \dots~J = 20,979~\operatorname{Individuals}
-\end{aligned}
-\]
+There are increasing worries
+
+We provide the analyses of these additional analyses in the supplement. The supplement describing these analysis is located at: \url{https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing}.
 
 \hypertarget{system-and-packages}{%
-\subsubsection{System and packages}\label{system-and-packages}}
+\subsection{System and packages}\label{system-and-packages}}
 
@@
 
-\hypertarget{missing-data}{%
-\subsubsection{Missing data}\label{missing-data}}
-
-Because missing values can result in biased estimates (Blackwell, Honaker, \& King, 2017), we multiply imputed 10 datasets using the AmeliaII package in R(Honaker et al., 2011b).We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences.
-Additionally the results were robust to filtering of participants of N \textgreater3 to N = 9 waves.
-Additionally, the findings reported below were robust to modeling only the effects of years and beliefs, without adjusting for the standard NZAVS demographic covariates.
-The supplement describing these analysis is located at: \url{https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing}.
-
 \hypertarget{code}{%
@@
 
-\begin{figure}
-\centering
-\includegraphics{/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Figs/USE_2_coefficientplots-1.png}
-\caption{\label{fig:unnamed-chunk-1}Combination Coefficient Plots for personal well-being and life satisfaction longitudinal growth models}
-\end{figure}
-
 Table 2 about here:
 
-\[
 \begin{table}
 \begin{center}
-\scalebox{0.6}{
+\scalebox{0.8}{
 \begin{tabular}{l c c}
 \toprule
- & PWI & Life Satisfaction \\
+ & PWI & Life Sat \\
 \midrule
-(Intercept)                        & $\mathbf{6.63}^{***}$  & $\mathbf{4.94}^{***}$  \\
-                                   & $(0.03)$               & $(0.02)$               \\
-Years                              & $0.00$                 & $-0.00$                \\
-                                   & $(0.00)$               & $(0.00)$               \\
-Beliefs\_SpiritExcludesGod\_       & $\mathbf{-0.14}^{***}$ & $\mathbf{-0.08}^{***}$ \\
-                                   & $(0.03)$               & $(0.02)$               \\
-BeliefsGodAndSpirit                & $\mathbf{-0.16}^{***}$ & $-0.04$                \\
-                                   & $(0.03)$               & $(0.02)$               \\
-BeliefsGodExcludesSpirit           & $0.02$                 & $0.07$                 \\
-                                   & $(0.05)$               & $(0.04)$               \\
-Age.10yrs.C                        & $\mathbf{0.08}^{***}$  & $\mathbf{0.09}^{***}$  \\
-                                   & $(0.01)$               & $(0.01)$               \\
-Deprivation.S                      & $\mathbf{-0.14}^{***}$ & $\mathbf{-0.07}^{***}$ \\
-                                   & $(0.01)$               & $(0.01)$               \\
-Edu.S                              & $\mathbf{0.12}^{***}$  & $\mathbf{0.06}^{***}$  \\
-                                   & $(0.01)$               & $(0.01)$               \\
-EmployedEmployed                   & $\mathbf{0.09}^{***}$  & $\mathbf{0.08}^{***}$  \\
-                                   & $(0.01)$               & $(0.01)$               \\
-EthnicCatsMaori                    & $\mathbf{-0.30}^{***}$ & $\mathbf{-0.06}^{**}$  \\
-                                   & $(0.03)$               & $(0.02)$               \\
-EthnicCatsPacific                  & $\mathbf{-0.29}^{***}$ & $\mathbf{-0.08}^{*}$   \\
-                                   & $(0.05)$               & $(0.04)$               \\
-EthnicCatsAsian                    & $-0.08$                & $-0.04$                \\
-                                   & $(0.05)$               & $(0.03)$               \\
-Male1                              & $\mathbf{-0.11}^{***}$ & $\mathbf{-0.19}^{***}$ \\
-                                   & $(0.02)$               & $(0.01)$               \\
-PartnerHas Partner                 & $\mathbf{0.61}^{***}$  & $\mathbf{0.42}^{***}$  \\
-                                   & $(0.02)$               & $(0.01)$               \\
-Pol.Orient.S                       & $\mathbf{0.02}^{***}$  & $\mathbf{0.01}^{**}$   \\
-                                   & $(0.01)$               & $(0.00)$               \\
-UrbanUrban                         & $\mathbf{-0.02}^{*}$   & $\mathbf{-0.02}^{*}$   \\
-                                   & $(0.01)$               & $(0.01)$               \\
-Years:Beliefs\_SpiritExcludesGod\_ & $\mathbf{0.02}^{**}$   & $\mathbf{0.02}^{***}$  \\
-                                   & $(0.00)$               & $(0.00)$               \\
-Years:BeliefsGodAndSpirit          & $\mathbf{0.02}^{***}$  & $\mathbf{0.01}^{***}$  \\
-                                   & $(0.00)$               & $(0.00)$               \\
-Years:BeliefsGodExcludesSpirit     & $0.00$                 & $-0.00$                \\
-                                   & $(0.01)$               & $(0.01)$               \\
+Intercept                           & $\mathbf{7.04}$   & $\mathbf{5.19}$  \\
+                                    & $ [ 7.02;  7.07]$ & $ [ 5.17; 5.21]$ \\
+yearsC                              & $\mathbf{0.02}$   & $\mathbf{0.01}$  \\
+                                    & $ [ 0.02;  0.03]$ & $ [ 0.01; 0.02]$ \\
+Beliefs\_SpiritExcludesGod\_        & $\mathbf{-0.06}$  & $0.02$           \\
+                                    & $ [-0.09; -0.03]$ & $ [-0.00; 0.04]$ \\
+BeliefsGodAndSpirit                 & $\mathbf{-0.06}$  & $\mathbf{0.05}$  \\
+                                    & $ [-0.10; -0.03]$ & $ [ 0.03; 0.07]$ \\
+BeliefsGodExcludesSpirit            & $-0.00$           & $\mathbf{0.08}$  \\
+                                    & $ [-0.04;  0.04]$ & $ [ 0.04; 0.11]$ \\
+yearsC:Beliefs\_SpiritExcludesGod\_ & $\mathbf{0.01}$   & $\mathbf{0.01}$  \\
+                                    & $ [ 0.00;  0.02]$ & $ [ 0.01; 0.02]$ \\
+yearsC:BeliefsGodAndSpirit          & $\mathbf{0.01}$   & $\mathbf{0.01}$  \\
+                                    & $ [ 0.01;  0.02]$ & $ [ 0.00; 0.02]$ \\
+yearsC:BeliefsGodExcludesSpirit     & $0.00$            & $-0.00$          \\
+                                    & $ [-0.01;  0.02]$ & $ [-0.01; 0.01]$ \\
 \midrule
-AIC                                & $248580.18$            & $200599.16$            \\
-BIC                                & $248775.08$            & $200793.96$            \\
-Log Likelihood                     & $-124269.09$           & $-100278.58$           \\
-Num. obs.                          & $79270$                & $78888$                \\
-Num. groups: Id                    & $20979$                & $20972$                \\
-Var: Id (Intercept)                & $1.78$                 & $0.88$                 \\
-Var: Residual                      & $0.74$                 & $0.42$                 \\
+SD: Idsd(Intercept)                 & $1.44$            & $0.99$           \\
+SD: Idsd(yearsC)                    & $0.13$            & $0.10$           \\
+SD: Idcor(Intercept,yearsC)         & $-0.02$           & $0.03$           \\
 \bottomrule
-\multicolumn{3}{l}{\scriptsize{$^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$}}
 \end{tabular}
 }
-\caption{Regression Models}
-\label{table:coefficients}
+\caption{}
+\label{tab:REGRESS}
 \end{center}
 \end{table}
-\]
+
+Figure 1 about here
+
+\begin{figure}
+\includegraphics[width=\textwidth]{/Users/jbul176/The Virtues Project Dropbox/Joseph Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/coefficientplotsbayes} \caption{Coefficient Graphs}\label{fig:unnamed-chunk-1}
+\end{figure}
 
@@
 
-We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict personal well-being (PWI) with years (0-9), beliefs (four categories), Age.10yrs.C, deprivation (S), education (S), employed, ethnic categories, male, partner, political conservativism (S) and urban, with Id as a random effect.
-The model's intercept, corresponding to years = 0, beliefs = Skeptic, age = (sample mean), deprivation.S = (sample mean), education (S) =(sample mean), employed = not employed, ethnic Categories = European, male = not male, partner = no partner, political conservativism (S) = (sample mean), urban = not urban, is at 4.94 (SE = 0.02, 95\% CI {[}4.89, 4.99{]}, p \textless{} .001) is at 6.63 (SE = 0.03, 95\% CI {[}6.57, 6.70{]}, p \textless{} .001).
+We fitted a linear mixed effects model using the brms package (Bürkner, 2018) in R (R Core Team, 2020) to predict personal well-being (PWI) with years (0-9), beliefs (four categories), with Id as random intercept and years as a random slope effect (see below: model equations).
+
 Within this model:
@@
 \item
-  The effect of Years is positive is not reliable (beta = 2.66e-03, SE = 3.58e-03, std.
-beta = 3.19e-03, p = 0.458).
-We infer that personal well-being did not increase in the Skeptic population between 2010 and 2018.
-\item
-  The effect of Beliefs in \emph{Spirit Excludes God} is reliably negative (beta = -0.14, SE = 0.03, std.
-beta = -0.03, p \textless{} .001).
-We infer that in 2010, the population that believed in a Spirit but did not believe in a God had lower personal well-being than did Skeptics.
-\item
-  The effect of Beliefs in \emph{God And Spirit} is reliably negative (beta = -0.16, SE = 0.03, std.
-beta = -0.02, p \textless{} .001).
-We infer that in 2010, the population that believed in a God and a Spirit or Life Force had lower Personal Well-being than did Skeptics.
-\item
-  The effect of Beliefs in \emph{God Excludes Spirit} is not reliable (beta = 0.02, SE = 0.05, std.
-beta = 0.02, p = 0.669).
-We infer that in 2010, the population that believed in a God and doubted a Spirit or Life Force similar levels of personal well-being to Skeptics.
-\item
-  The effect of Age.10yrs.C is reliably positive (beta = 0.08, SE = 7.38e-03, std. beta = 0.07, p \textless{} .001). Older people tend to have greater personal well-being.
-\item
-  The effect of Deprivation (S) is reliably negative (beta = -0.14, SE = 7.50e-03, std. beta = -0.08, p \textless{} .001). Deprived people tend to have lower personal well-being.
-\item
-  The effect of Education (S) is reliably positive (beta = 0.12, SE = 8.74e-03, std. beta = 0.07, p \textless{} .001). Educated people tend to have greater personal well-being.
-\item
-  The effect of Employed is reliably positive (beta = 0.09, SE = 0.01, std. beta = 0.05, p \textless{} .001).
-\item
-  The effect of identifying as M=\{a\}ori is reliably negative (beta = -0.30, SE = 0.03, std. beta = -0.18, p \textless{} .001).
-\item
-  The effect of identifying as Pacific is reliably negative (beta = -0.29, SE = 0.05, std. beta = -0.17, p \textless{} .001).
-\item
-  The effect of identifying as Asian is not reliable (beta = -0.08, SE = 0.05, std. beta = -0.05, p = 0.089).
-\item
-  The effect of identifying as male is reliably negative (beta = -0.11, SE = 0.02, std. beta = -0.06, p \textless{} .001).
-\item
-  The effect of having a partner is reliably positive (beta = 0.61, SE = 0.02, std. beta = 0.37, p \textless{} .001).
-\item
-  The effect of political conservativism (S) is reliably positive (beta = 0.02, SE = 5.67e-03, std. beta = 0.01, p \textless{} .001).
-\item
-  The effect of living in an Urban area is reliably negative (beta = -0.02, SE = 0.01, std. beta = -0.01, p \textless{} .05). T
+  The effect of years is positive is reliable (posterior median = 0.025, HDI {[}0.020, 0.031{]} pd = 100.00\%).
+We infer that personal well-being increased in the skeptic population between 2010 and 2018.
+\item
+  The effect of beliefs in \emph{Spirit Excludes God} is reliably negative (posterior median = -0.058, HDI {[}-0.080, -0.035{]}, pd = 100.00\%).
+We infer that the population that believed in a Spirit but did not believe in a God was lower in personal well-being at the intercept compared to skeptics.
+\item
+  The effect of beliefs in \emph{God And Spirit} is reliably negative (posterior median = -0.064, HDI {[}-0.091, -0.037{]}, pd = 100.00\%).
+We infer that the population that believed in a God and a Spirit or Life Force was lower in personal well-being at the intercept compared to skeptics.
+\item
+  The effect of beliefs in \emph{God Excludes Spirit} is not reliable (posterior median = -0.002 HDI {[}-0.039, 0.033{]}, pd = 53.50\%).
+We infer that the population that believed in a God and doubted a Spirit or Life Force were similar were similar in personal well-being at the intercept to skeptics.
 \end{itemize}
@@
 \hypertarget{personal-well-being-years-x-beliefs}{%
-\subsection{Personal well-being: years X beliefs}\label{personal-well-being-years-x-beliefs}}
-
-\includegraphics{Figs/USEPWI_expected-1.png}
-Focusing on the interaction of time with belief-type on personal well-being we observe:
+\subsection{Personal Well-Being: years X beliefs}\label{personal-well-being-years-x-beliefs}}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{/Users/jbul176/The Virtues Project Dropbox/Joseph Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/Bayesian_pwi_predicted} \caption{Personal Well-being prediction graph}\label{fig:unnamed-chunk-2}
+\end{figure}
+
+The effect of years \(\times\) belief in a \emph{Spirit Excludes God} is reliably positive (posterior median = 0.010, HDI {[} 0.003, 0.017{]}, pd = 99.02\%).
+We infer that people with beliefs in a spirit or life force who do not believe in a God tended to grow in their personal well-being over time at a faster rate than skeptics grew.
+- The effect of years \(\times\) in \emph{God And Spirit} is reliably positive (posterior median = 0.014, HDI {[}0.008, 0.021{]}, pd = 99.98\%).
+We infer that people who believe in a spirit or life force who also believe in a God tended to grow in their personal well-being over time at a faster rate than skeptics grew.
+- The effect of years \(\times\) belief in \emph{God Excludes Spirit} is not reliable (posterior median = 0.002, HDI {[}-0.009, 0.015{]}, pd = 63.73\%).
+We infer that people who believe in a God but not a spirit or life force did not grow at a faster rate in personal well-being than skeptics grew.
+
+\hypertarget{life-satisfaction-1}{%
+\subsection{Life Satisfaction}\label{life-satisfaction-1}}
+
+We fitted a linear mixed effects model using the brms package (Bürkner, 2018) in R (R Core Team, 2020) to predict personal well-being (PWI) with years (0-9), beliefs (four categories), with Id as random intercept and years as a random slope effect (see model equations, below).
+
+Within this model:
 
@@
 \item
-  The effect of years \(\times\) belief in a \emph{Spirit Excludes God} is reliably positive (beta = 0.02, SE = 4.86e-03, std.
-beta = 0.02, p \textless{} .01).
-We infer that, unlike the skeptic population, people with beliefs in a spirit or life force who do not believe in a God tend to grow in their personal well-being over time.
-\item
-  The effect of years \(\times\) in \emph{God And Spirit} is reliably positive (beta = 0.02, SE = 4.41e-03, std.
-beta = 0.03, p \textless{} .001).
-We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their personal well-being over time
-\item
-  The effect of Years \(\times\) belief in \emph{God Excludes Spirit} is not reliable(beta = 1.96e-03, SE = 8.29e-03, std.
-beta = 2.36e-03, p = 0.813).
-Thus, God believers who doubt in a Spirit or Life Force start maintain higher personal well-being than do other groups, but they are not expected to grow in personal well-being over time.
+  The effect of years is positive is reliable (posterior median = 0.014, HDI {[} 0.010, 0.019{]}, pd = 100.00\%).
+We infer that life satisfaction increased in the skeptic population between 2010 and 2018.
+  -The effect of beliefs in \emph{Spirit Excludes God} is reliably positive (posterior median = 0.018, HDI {[} 0.001, 0.035{]}, pd = 96.05\%).
+We infer that the population that believed in a Spirit but did not believe in a God was higher in life satisfaction at the intercept compared with skeptics.
+\item
+  The effect of beliefs in \emph{God And Spirit} is reliably positive (posterior median = 0.051, HDI {[} 0.031, 0.071{]}, pd = 100.00\%).
+We infer that the population that believed in a God and a Spirit or Life Force was higher in life satisfaction at the intercept compared with skeptics.
+\item
+  The effect of beliefs in \emph{God Excludes Spirit} is reliably positive (posterior median = 0.076, HDI {[} 0.048, 0.102{]}, pd = 100.00\%).
+We infer that the population that believed in a God and doubted a Spirit or Life Force was higher in life satisfaction at the intercept compared with skeptics.
 \end{itemize}
 
-\hypertarget{life-satisfaction-1}{%
-\subsection{Life Satisfaction}\label{life-satisfaction-1}}
-
-We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict personal well-being (PWI) with years (0-9), beliefs (four categories), Age.10yrs.C, deprivation (S), education (S), employed, ethnic categories, male, partner, political conservativism (S) and urban, with Id as a random effect.
-Standardized parameters were obtained by fitting the model on a standardized version of the dataset.
-The model's total explanatory power is substantial (conditional R2 = 0.69).
-The model's intercept, corresponding to years = 0, beliefs = Skeptic, age = (sample mean), deprivation.S = (sample mean), education (S) =(sample mean), employed = not employed, ethnic Categories = European, male = not male, partner = no partner, political conservativism (S) = (sample mean), urban = not urban is at 4.94 (SE = 0.02, 95\% CI {[}4.89, 4.99{]}, p \textless{} .001).
-Within this model:
+\hypertarget{life-satisfaction-years-x-beliefs}{%
+\subsection{Life Satisfaction: years X beliefs}\label{life-satisfaction-years-x-beliefs}}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{/Users/jbul176/The Virtues Project Dropbox/Joseph Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/Bayesian_lifesat_predicted} \caption{Life satisfaction prediction graph}\label{fig:unnamed-chunk-3}
+\end{figure}
 
@@
 \item
-  The effect of years is not reliable (beta = -5.50e-04, SE = 2.69e-03, std.
-beta = -9.32e-04, p = 0.838).
-We infer that life satisfaction did not increase in the Skeptic population between 2010 and 2018.
-\item
-  The effect of beliefs in \emph{Spirit Excludes God} is reliably negative (beta = -0.08, SE = 0.02, std.
-beta = 0.01, p \textless{} .001).
-We infer that overall the population that believes in a Spirit but does not believe in a God is expected to have a lower Life Satisfaction than skeptics.
-\item
-  The effect of beliefs in \emph{God And Spirit} is not reliable (beta = -0.04, SE = 0.02, std.
-beta = 0.04, p = 0.105).
-People who believe in God but doubt a Spirit/Life force are expected to have had similar levels of life satisfaction in Wave 10 as skeptics.
-\item
-  The effect of beliefs in \emph{God Excludes Spirit} is reliably positive (beta = 0.07, SE = 0.04, std.
-beta = 0.05, p = 0.062).We infer that overall the population that believes in a God but does not believe in a spirit is not reliably different from the Skeptics population in Life Satisfaction.
-\item
-  The effect of Age.10yrs.C is reliably positive (beta = 0.09, SE = 5.26e-03, std. beta = 0.10, p \textless{} .001). Older people have greater Life Satisfaction
-\item
-  The effect of Deprivation (S) is is reliably negative (beta = -0.07, SE = 5.50e-03, std. beta = -0.06, p \textless{} .001). People who are more deprived have lower Life Satisfaction.
-\item
-  The effect of Education (S) is reliably positive (beta = 0.06, SE = 6.33e-03, std. beta = 0.05, p \textless{} .001). Educated people have greater Life Satisfaction
-\item
-  The effect of being Employed is reliably positive (beta = 0.08, SE = 9.69e-03, std. beta = 0.07, p \textless{} .001). People who are employed tend to have greater Life Satisfaction
-\item
-  The effect of identifying as Maori is reliably negative (beta = -0.06, SE = 0.02, std.
-beta = -0.05, p \textless{} .01).
-People who belong to the minority Maori group tend to have lower Life Satisfaction.
-\item
-  The effect of identifying as Pacific is reliably negative (beta = -0.08, SE = 0.04, std.
-beta = -0.07, p \textless{} .05).
-People who belong to the minority Pacific peoples group tend to have lower life satisfaction.
-\item
-  The effect of identifying as Asian is not reliably different from people of European ethnicity in life satisfaction(beta = -0.04, SE = 0.03, std. beta = -0.03, p = 0.223).
-\item
-  The effect of identifying as male is reliably negative (beta = -0.19, SE = 0.01, std. beta = -0.16, p \textless{} .001).
-\item
-  The effect of having a partner is reliably positive (beta = 0.42, SE = 0.01, std. beta = 0.35, p \textless{} .001).
-\item
-  The effect of political conservativism (S) is reliably positive (beta = 0.01, SE = 4.23e-03, std. beta = 0.01, p \textless{} .01).
-\item
-  The effect of living in an urban area is reliably negative (beta = -0.02, SE = 8.85e-03, std. beta = -0.02, p \textless{} .05).
+  The interaction of years \(\times\) beliefs in \emph{Spirit Excludes God} is reliably positive (posterior median = 0.012, HDI {[} 0.007, 0.017{]}, pd = 100.00\%).
+We infer that people with beliefs in a Spirit or Life Force who do not believe in a God tended to grow over time in their life satisfaction at a faster rate than skeptics.
+
+  \begin{itemize}
+  \tightlist
+  \item
+    The interaction of years \(\times\) \emph{God And Spirit} is reliably positive (posterior median = 0.010, HDI {[} 0.004, 0.015{]}, pd = 99.90\%).
+We infer that people who believe in a spirit or life force who also believe in a God tended to grow over time in their life satisfaction at a faster rate than skeptics.
+  \item
+    The effect of years \(\times\) beliefs in \emph{God Excludes Spirit} is not reliable (posterior median = -0.002, HDI {[}-0.011, 0.007{]}, pd = 62.80\%).
+We infer that people who believe in a God but not a spirit or life force tended to grow over time in their life satisfaction at a similar rate to skeptics.
   \end{itemize}
-
-\hypertarget{life-satisfaction-years-x-beliefs}{%
-\subsection{Life Satisfaction: years X beliefs}\label{life-satisfaction-years-x-beliefs}}
-
-\includegraphics{/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Figs/USElifsat_expected-1.png}
-- The interaction of years \(\times\) beliefs in \emph{Spirit Excludes God} ludesGod is reliably positive (beta = 0.02, SE = 3.66e-03, std.
-beta = 0.03, p \textless{} .001).
-We infer that, unlike the Skeptic population, people with beliefs in a Spirit or Life Force who do not believe in a God tend to grow in their Life Satisfaction over time.
-- The interaction of years \(\times\) \emph{God And Spirit} is reliably positive (beta = 0.01, SE = 3.32e-03, std.
-beta = 0.02, p \textless{} .001).
-We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their life satisfaction over time
-- The effect of years \(\times\) beliefs in \emph{God Excludes Spirit} is not reliable (beta = -1.59e-03, SE = 6.25e-03, std.
-beta = -2.70e-03, p = 0.799).
-Thus, God believers who doubt in a Spirit or Life Force start maintain higher Life-Satisfaction than do other groups, but they are not expected to grow in Life-Satisfaction over time.
-
-\hypertarget{imputed-data-analysis}{%
-\subsection{Imputed data analysis}\label{imputed-data-analysis}}
-
-To adjust for potential bias from missing responses, we conducted a supplementary analysis in which we imputed missing values over the the longitudinal sample.
-This analysis is describe in Supplement 1.
-Averaging over the missing responses does not change inference, however we observe growth across all four groups in both personal well-being and life satisfaction.
-Additionally we observe substantially stronger effects for the slope of Spirit Beliefs on both personal well-being and life-satisfaction.
-Figures 3 and Figures 4 describe the predicted slopes of change.
-The inference from this analysis remains the same as from the analysis in which we delete missing values row-wise, by panel wave.
-In all analyses, those who believe in a spirit or life-force experience greater growth in both personal well-being and life satisfaction over time.
-
-Figure 3 about here:
-
-\begin{figure}
-\centering
-\includegraphics{Figs/coefficient_plot_both_imputed-1.png}
-\caption{\label{fig:unnamed-chunk-4}Imputed data analysis: coefficient plot}
-\end{figure}
-
-Figure 4 about here:
-
-\begin{figure}
-\centering
-\includegraphics{/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Figs/predicted_PWI_impute-1.png}
-\caption{\label{fig:unnamed-chunk-5}Imputed data analysis: predicted personal wellbeing by years X beliefs}
-\end{figure}
-
-Figure 5 about here:
-
-\begin{figure}
-\centering
-\includegraphics{/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Figs/predicted_lifesat_imputed-1.png}
-\caption{\label{fig:unnamed-chunk-6}Imputed data analysis: predicted personal wellbeing by years X beliefs}
-\end{figure}
+\end{itemize}
 
@@
 \item
-  At the outset of the study, in 2010, the expected subjective well-being as measured using the personal well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God (\(\hat{y}\) = 6.50, se =.03) and lower among those who believed in both a Spirit/Life-Force and a God (\(\hat{y}\) = 6.48, se =.03) in comparison both to the skeptical population who neither believed in a God nor a life force (\(\hat{y}\) = 6.64,se =.03) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force (\(\hat{y}\) 6.66, se =.05).
-(See Figure 2) Similarly,the expected subjective well-being as measured using the life satisfaction Inventory is lower among those who believed in Spirit or Life Force but not a God (\(\hat{y}\) = 4.86, se =.02) and lower among those who believed in both a Spirit/Life-Force and a God (\(\hat{y}\) = 4.91, se =.02) in comparison both with the skeptical population who neither believed in a God nor a life force (\(\hat{y}\) = 4.94,se =.02) and the population that believed in a God but was skeptical about belief in a Spirit or Life Force (\(\hat{y}\) 5.02, se =.02) (see Figure 3)
+  At the outset of the study, in 2010, the expected subjective well-being as measured using the personal well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God (\(\hat{y} =\) 6.80 {[}6.76, 6.85{]}) and lower among those who believed in both a Spirit/Life-Force and a God (\(\hat{y}\) = 6.78 {[}6.74, 6.81{]}) in comparison both to the skeptical population who neither believed in a God nor a life force (\(\hat{y}\) = 6.92 {[}6.87, 6.96{]}) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force (\(\hat{y}\) 6.90 {[}6.83, 6.98{]}).
+(See Figure 2) Similarly,the expected subjective well-being as measured using the life satisfaction Inventory was lowest in 2010 among those who believed in Spirit or Life Force but not a God (\(\hat{y}\) = 5.07 {[}5.04, 5.10{]}).
+Life satisfaction was similar among those who believed in both a Spirit/Life-Force and a God (\(\hat{y}\) = 5.11 {[}5.08, 5.13{]}) in comparison both with the skeptical population who neither believed in a God nor a life force (\(\hat{y}\) = 5.12 {[}5.08, 5.15{]}).
+Life satisfaction was highest among the population that believed in a God but was skeptical about belief in a Spirit or Life Force (\(\hat{y}\) 5.19 {[}5.14, 5.25{]}) (see Figure 3)
 \item
   However, over time, belief in a Spirit or Life Force predicts a trajectory of growth in personal well-being.
-In 2018, personal well-being among who believe in Spirit or Life Force but not a God had grown to (\(\hat{y}\) = 6.67, se =.03, \(\Delta \hat{y} = 0.17\)).
-Among those who believed in both a Spirit/Life-Force and a God, personal well-being had grown to (\(\hat{y} = 6.69, se = 0.02, \Delta \hat{y} = .21\)).
-To place this effect in practical perspective, such growth in personal well-being is about twice as large as the expected benefit of being employed and about a third as large as the expected benefit from having a romantic partner.
-By contrast, personal well-being among the skeptical population who neither believed in a God nor a Life Force remained stable across the eight year interval \(\hat{y} = 6.66, se = 0.03, \Delta \hat{y} = .02\)).
-By the same token, personal well-being among those who believe in a God but are skeptical about a Spirit/Life Force was constant \(\hat{y} = 6.70, se = 0.04, \Delta \hat{y} = .04\)).
+In 2018, personal well-being among who believe in Spirit or Life Force but not a God had grown to (\(\hat{y} = 7.12\) {[}7.09, 7.16{]}, \(\Delta \hat{y} = 0.32\)).
+Among those who believed in both a Spirit/Life-Force and a God, personal well-being had grown to (\(\hat{y} = 7.13\) {[}7.10, 7.16{]}, \(\Delta \hat{y} = 0.35\)).
+By contrast, personal well-being among the skeptical population who neither believed in a God nor a Life Force grew less steeply across the eight year interval \(\hat{y} = 7.14\) {[}7.11, 7.18{]}, \(\Delta \hat{y} = 0.22\)).
+By the same token, personal well-being among those who believe in a God but are skeptical about a Spirit/Life Force grew less steeply \(\hat{y} = 7.15\), {[}7.09, 7.21{]} \(\Delta \hat{y} = 0.25\)).
 Similarly, over time, belief in a Spirit or Life Force predicts a trajectory of growth in life satisfaction.
-In 2018, life satisfaction among those who believe in Spirit or Life Force but not a God had grown to (\(\hat{y} = 5.01, se = 0.02, \Delta \hat{y} = 0.07\)).
+In 2018, the expected effect of spiritual beliefs on life satisfaction among those who believe in Spirit or Life Force but not a God had grown to (\(\hat{y} = 5.31\) {[}5.28, 5.33{]}, \(\Delta \hat{y} = 0.23\)).
 In practical terms, such growth is equivalent to the expected gain in life satisfaction from being employed and a sixth as large as the expected benefit to life satisfaction from having a romantic partner.
-Among those who believed in both a Spirit/Life-Force and a God, life satisfaction had grown to (\(\hat{y} = 5.03, se = 0.02, \Delta \hat{y} = 0.12\)).
-In practical terms, such growth is equivalent to 1.5 times the expected gain to life satisfaction from being employed and almost a third benefit to life satisfaction from having a romantic partner.
-By contrast, life satisfaction among the skeptical population who neither believed in a God nor a Life Force remained stable over the eight year interval \(\hat{y} = 4.94, se =0.02, \Delta \hat{y} = 0\)).
-Relatedly, life satisfaction in the population that believe in God but are skeptical about a Spirit/Life Force was constant \(\hat{y} = 5.00, se = 0.03, \Delta \hat{y} = -.02\)) (See Figure 3).
+In practical terms, the rate of increase in personal wellbeing among those with Spirit/Life force beliefs was about one and half times greater than it was among those who lacked such beliefs (See Figure 2).
+Among those who believed in both a Spirit/Life-Force and a God, life satisfaction had grown to (\(\hat{y} = 5.33 [5.31, 5.35]\), \(\Delta \hat{y} = 0.24\)).
+By contrast, life satisfaction among the skeptical population who neither believed in a God nor a Life Force lagged behind over the eight year interval \(\hat{y} = 5.24 [5.21, 5.27], \Delta \hat{y} = 0.13\)).
+Relatedly, life satisfaction in the population that believe in God but are skeptical about a Spirit/Life Force grew less steeply \(\hat{y} = 5.31 [5.26, 5.36], \Delta \hat{y} = 0.11\)).
+In practical terms, the rate of increase in life satisfaction among those with Spirit/Life force beliefs was twice as large (See Figure 3).
 \item
@@
 
-BH conceived of the study.
-BH wrote the first draft of the manuscript, and performed the initial data analysis with input from JB.
+BH and JB conceived of the study.
+BH wrote the first draft of the manuscript.
 EW improved the approach with BH and JB and revised the manuscript.
@@
 
+\newpage
+
+\hypertarget{appendix}{%
+\section{Appendix}\label{appendix}}
+
+\hypertarget{model-equations}{%
+\subsection{Model Equations}\label{model-equations}}
+
+\hypertarget{priors-for-the-personal-wellbeing-model}{%
+\subsubsection{Priors for the Personal Wellbeing model}\label{priors-for-the-personal-wellbeing-model}}
+
+Model priors for Personal Wellbeing including group-level ID intercept and beliefs slopes are as follows
+
+\[
+\begin{aligned}
+\operatorname{Personal Well Being}_{ij} \sim \text{Normal}(\mu_{ij},\sigma^2) \nonumber\\
+\mu_{ij}\sim \boldsymbol{\alpha +\beta} \nonumber\\
+\boldsymbol{\alpha} \sim \alpha_{0} +\alpha_j \nonumber\\
+\alpha_{0} \sim \text{StudentT}(3,7.2,2.5)\nonumber\\
+\boldsymbol{\beta}\sim \boldsymbol{b} + \beta_j \nonumber\\
+\boldsymbol{b}\sim\text{Normal}(0,2.5) \nonumber\\
+\begin{bmatrix}
+\alpha_j \\
+\beta_j
+\end{bmatrix}
+\sim 
+\text{MVNormal}
+\begin{pmatrix}
+\begin{bmatrix}
+0\\
+0
+\end{bmatrix}
+,\boldsymbol{S} \nonumber
+\end{pmatrix}\\
+\boldsymbol{S} = 
+\begin{pmatrix}
+0 & \sigma_{\beta_j} \nonumber
+\end{pmatrix} \boldsymbol{R} \begin{pmatrix}
+\sigma_{\alpha_j} & 0 \\
+0 & \sigma_{\beta_j} \nonumber
+\end{pmatrix}\\
+\sigma_{\alpha_j} \sim 
+\text{HalfCauchy}(0,2.5) \nonumber\\
+\sigma_{\beta_j} \sim 
+\text{HalfCauchy}(0,2.5) \nonumber\\
+\boldsymbol{R}\sim \text{LKJcorr}(1) \nonumber\\
+\sigma  \sim  \operatorname{HalfStudent T}^{+}(3, 0, 2.5) 
+\end{aligned}
+\]
+
+\newpage
+
+\hypertarget{priors-for-the-life-satisfaction-model}{%
+\subsubsection{Priors for the Life Satisfaction model}\label{priors-for-the-life-satisfaction-model}}
+
+Model priors for Life Satisfaction Model with including group-level ID intercept and belief slopes are as follows:
+
+\[
+\begin{aligned}
+\operatorname{Life Satisfaction}_{ij} \sim \text{Normal}(\mu_{ij},\sigma^2) \nonumber\\
+\mu_{ij}\sim \boldsymbol{\alpha +\beta} \nonumber\\
+\boldsymbol{\alpha} \sim \alpha_{0} +\alpha_j \nonumber\\
+\alpha_{0} \sim \text{StudentT}(3,5.5,2.5)\nonumber\\
+\alpha_j \sim \text{StudentT}+(3,0,2.5)\nonumber\\
+\boldsymbol{\beta}\sim \boldsymbol{b} + \beta_j \nonumber\\
+\boldsymbol{b}\sim\text{Normal}(0,2.5) \nonumber\\
+\begin{bmatrix}
+\alpha_j \\
+\beta_j
+\end{bmatrix}
+\sim 
+\text{MVNormal}
+\begin{pmatrix}
+\begin{bmatrix}
+0\\
+0
+\end{bmatrix}
+,\boldsymbol{S} \nonumber
+\end{pmatrix}\\
+\boldsymbol{S} = 
+\begin{pmatrix}
+\sigma_{\alpha_j} & 0 \\
+0 & \sigma_{\beta_j} \nonumber
+\end{pmatrix} \boldsymbol{R} \begin{pmatrix}
+\sigma_{\alpha_j} & 0 \\
+0 & \sigma_{\beta_j} \nonumber
+\end{pmatrix}\\
+\sigma_{\alpha_j} \sim 
+\text{HalfCauchy}(0,1) \nonumber\\
+\sigma_{\beta_j} \sim 
+\text{HalfCauchy}(0,1) \nonumber\\
+\boldsymbol{R}\sim \text{LKJcorr}(1) \nonumber\\
+\sigma  \sim  \operatorname{HalfStudent T}^{+}(3, 0, 2.5) 
+\end{aligned}
+\]
+\newpage
+
+\hypertarget{deterministic-model}{%
+\subsubsection{Deterministic model}\label{deterministic-model}}
+
+The deterministic portion of both the Personal Wellbeing and the Life Satisfaction models is as follows:
+\[
+\begin{aligned}
+& \quad  \boldsymbol{b} = \\
+& \quad \beta_{1}(\operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\
+&\quad \beta_{2}(\operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ +  \beta_{3}(\operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}})\ +\\
+& \quad \beta_{4}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\ 
+& \quad \beta_{4}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ + \\
+&\quad \beta_{6}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) \\ 
+\end{aligned}
+\]
 \newpage
@@
 \leavevmode\hypertarget{ref-Blackwell2017-oq}{}%
-Blackwell, M., Honaker, J., \& King, G. (2017). A unified approach to measurement error and missing data: Details and extensions.
+Blackwell, M., Honaker, J., \& King, G. (2017). A unified approach to measurement error and missing data: Details and extensions. \emph{Sociological Methods \& Research}.
+
+\leavevmode\hypertarget{ref-BRMSpackage}{}%
+Bürkner, P.-C. (2018). Advanced Bayesian multilevel modeling with the R package brms. \emph{The R Journal}, \emph{10}(1), 395--411. \url{https://doi.org/10.32614/RJ-2018-017}
 
@@
 \leavevmode\hypertarget{ref-Cummins2017-ur}{}%
-Cummins, R. A., Eckerseley, R., Pallant, J., Van Vugt, J., \& Misajon, R. (2017). Australian unity wellbeing index.
+Cummins, R. A., Eckerseley, R., Pallant, J., Van Vugt, J., \& Misajon, R. (2017). Australian unity wellbeing index. \emph{PsycTESTS Dataset}.
 
 \leavevmode\hypertarget{ref-Diener1985-xy}{}%
-Diener, E., Emmons, R. A., Larsen, R. J., \& Griffin, S. (1985). The satisfaction with life scale.
+Diener, E., Emmons, R. A., Larsen, R. J., \& Griffin, S. (1985). The satisfaction with life scale. \emph{Journal of Personality Assessment}.
 
@@
 \leavevmode\hypertarget{ref-King2013-cg}{}%
-King, M., Marston, L., McManus, S., Brugha, T., Meltzer, H., \& Bebbington, P. (2013). Religion, spirituality and mental health: Results from a national study of english households.
+King, M., Marston, L., McManus, S., Brugha, T., Meltzer, H., \& Bebbington, P.
+(2013).
+Religion, spirituality and mental health: Results from a national study of english households.
+\emph{British Journal of Psychiatry}.
 
@@
 \leavevmode\hypertarget{ref-Zaza2005-ac}{}%
-Zaza, C., Sellick, S. M., \& Hillier, L. M. (2005). Coping with cancer.
+Zaza, C., Sellick, S. M., \& Hillier, L. M. (2005). Coping with cancer. \emph{Journal of Psychosocial Oncology}.
 

### 2020-11-27 Manuscript_Bayesian.Rmd
changes to manuscript.

@@
 ---
-title             : "National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Belief"
+title             : "National Longitudinal Evidence for Growth in Subjective Well-being from Spiritual Beliefs"
 shorttitle        : "Longitudinal Wellbeing from Spiritual Beliefs"
@@
 classoption       : "man"
-output            : papaja::apa6_word
+output            : papaja::apa6_pdf
 extra_dependencies: ["rotating", "booktabs", "graphicx","longtable","amsmath","amsfonts","amssymb"]

### 2020-11-27 Manuscript_Bayesian.tex
changes to manuscript.

@@
 \hypersetup{
-  pdftitle={National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Belief},
+  pdftitle={National Longitudinal Evidence for Growth in Subjective Well-being from Spiritual Beliefs},
   pdfauthor={Benjamin Highland1, Everett L. Worthington2, Don E Davis3, Chris G. Sibley4, \& Joseph A. Bulbulia5},
@@
 
-\title{National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Belief}
+\title{National Longitudinal Evidence for Growth in Subjective Well-being from Spiritual Beliefs}
 \author{Benjamin Highland\textsuperscript{1}, Everett L. Worthington\textsuperscript{2}, Don E Davis\textsuperscript{3}, Chris G. Sibley\textsuperscript{4}, \& Joseph A. Bulbulia\textsuperscript{5}}

### 2020-11-27 Supplement_And_Analysis.Rmd
changes to manuscript.

@@
 ---
-title             : "Supplement: Analysis for National Longitudinal Evidence for Growth in Subjective well-being from Spiritual Belief"
+title             : "Supplement: Analysis for National Longitudinal Evidence for Growth in Subjective Well-being from Spiritual Beliefs"
 author: "contact: joseph.bulbulia@gmail.com on behalf of Highland et al."
@@
 ```{r global_options, echo=FALSE, include = FALSE}
-knitr::opts_chunk$set(fig.path="Figs/", 
+knitr::opts_chunk$set(fig.path="figs/", 
                       message=FALSE, 
@@
 
-```{r latex_data_table, include=FALSE}
-# you might need to install the dev verion.//remotes::install_github("tysonstanley/furniture")
-demographic_table_latex(d_3) # note the furniture package conflicts with the table1 package. 
-conflicts()
-furniture::table1(d_3, splitby = ~Wave,
-                    Age, 
-                    NZdep,
-                    Education,
-                    Employed,
-                    Ethnic_Categories,
-                    Male,Has_Partner, 
-                    Political_Conservativism,
-                    Urban, 
-                    Beliefs, 
-                    LIFESAT,
-                    PWI,
-                    overall=F, 
-                    output = "latex2", 
-                    booktabs=TRUE)
-```
+<!-- ```{r latex_data_table, include=FALSE} -->
+<!-- # you might need to install the dev verion.//remotes::install_github("tysonstanley/furniture") -->
+<!-- #demographic_table_latex(d_3) # note the furniture package conflicts with the table1 package.  -->
+<!-- ``` -->
 
@@
 
-In the the reported analyses, we included covariates to adjust for potential confounding effects from usual suspect factors associated with our two measures of well-being (Personal Well-Being, Life Satisfaction).
+In the reported analyses, we included covariates to adjust for potential confounding effects from usual suspect factors associated with our two measures of well-being (Personal Well-Being, Life Satisfaction).
 Where $Z$ is a third variable, $X$ is the predictor, and $Y$ is the outcome, there are at least four types of confounding that might arise:
@@
 Below we write the model in a notation that explicitly states the the default weakly informative priors that the BRMS package produces.
-And indicated in the model we include group-level indicator for variation in subject-level intercepts $\alpha_j$, a group-level indicator for subject-level time (or "growth")  slope  $\Beta_j$, and we model the co-variance of $\alpha_j$, Below we write the model in a notation that explicitly states the the default weakly informative priors that the BRMS package produces.
-And indicated in the model we include group-level indicator for variation in subject-level intercepts $\alpha_j$, a group-level indicator for subject-level time (or "growth")  slope  $\Beta_j$, and we model the co-variance of $\alpha_j$ and  $\Beta_j$ which for computational reasons is estimate as a correlation matrix of (${\bf S}$} and ${\bf R}$).
+And indicated in the model we include group-level indicator for variation in subject-level intercepts $\alpha_j$, a group-level indicator for subject-level time (or "growth")  slope  $\beta_j$, and we model the co-variance of $\alpha_j$, Below we write the model in a notation that explicitly states the the default weakly informative priors that the BRMS package produces.
+And indicated in the model we include group-level indicator for variation in subject-level intercepts $\alpha_j$, a group-level indicator for subject-level time (or "growth")  slope  $\beta_j$, and we model the co-variance of $\alpha_j$ and  $\beta_j$ which for computational reasons is estimate as a correlation matrix of (${\bf S}$} and ${\bf R}$).
 
@@
 
-```{r texreg_bayes_table:pwi,cache=TRUE, include=FALSE}
-#Results for publication.
-table_model_latex_pwi(mod.1_b)
-```
-
-```{r texreg_bayes_table:ls, cache=TRUE, include=FALSE}
-table_model_latex_ls(mod.2_b)
-```
+<!-- ```{r texreg_bayes_table:pwi,cache=TRUE, include=FALSE} -->
+<!-- #Results for publication. -->
+<!-- table_model_latex_pwi(mod.1_b) -->
+<!-- ``` -->
+
+<!-- ```{r texreg_bayes_table:ls, cache=TRUE, include=FALSE} -->
+<!-- table_model_latex_ls(mod.2_b) -->
+<!-- ``` -->
 
@@
 
-```{r mixing_sd_pwi}
+```{r mixing_sd_pwi, cache=TRUE}
 plot(mod.1_b, pars ="sd_") # requires stronger priors
@@
 
-```{r mixing_ls_pwi}
+```{r mixing_ls_pwi, cache=TRUE}
 plot(mod.2_b, pars ="sd_") # requires stronger priors

### 2020-11-28 Manuscript_Bayesian.Rmd
added saturated models for the bayesian estimate. Not distinguishable from the REML models.

@@
 
-We provide the analyses of these additional analyses in the supplement.  The supplement describing these analysis is located at: https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing.
+We provide the analyses of these additional analyses in the supplement.
+The supplement describing these analysis is located at: [https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing](https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing).
+The html document is located at: [https://osf.io/cu2gr/](https://osf.io/cu2gr/)
 
@@
 All code for this analysis and a full report of the missing data imputation can be found at [https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing](https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing).
+An html document decribing the same analysis is located at: [https://osf.io/cu2gr/](https://osf.io/cu2gr/).
 

### 2020-11-28 Supplement_And_Analysis.Rmd
added saturated models for the bayesian estimate. Not distinguishable from the REML models.

@@
 graph_predictions(mod.2_b, "Predicted Values of Life Satisfaction")
+```
+
+### Test of the Bayesian saturacted model
+
+```{r setmodelformulaSAT}
+forumula_bayesPWI_SATUR <- bf(PWI ~  yearsC * Beliefs +  Age_in_Decades_C + Deprivation_S + Education_S  + Employed + Ethnic_Categories + Male  + Has_Partner  + Political_Conservativism_S + Urban + (1+ yearsC|Id))
+forumula_bayesLIFESAT_SATUR <- bf(LIFESAT ~  yearsC * Beliefs +  Age_in_Decades_C + Deprivation_S + Education_S  + Employed + Ethnic_Categories + Male  + Has_Partner  + Political_Conservativism_S + Urban + (1+ yearsC|Id))
+```
+```{r pwi_bayesmodelSAT}
+# PWI
+mod.1_bSAT <-brm(forumula_bayesPWI_SATUR, data= d_3, file ="mod.1_bS.rds", cores = 8) # note these are very large files
+summary(mod.1_bSAT)
+```
+
+```{r Bayesian_pwi_predictedSAT, cache=TRUE, dependson= mod.1_b}
+graph_predictions(mod.1_bSAT, "Predicted Values of Personal Well-Being")
+```
+```{r Bayesian_lifesat_predictedSAT, cache=TRUE, dependson= mod.2_b}
+graph_predictions(mod.2_b, "Predicted Values of Life Satisfaction")
+```
+
+```{r lsbayesmodelSAT}
+# LIFE SAT
+mod.2_bSAT <-brm (forumula_bayesLIFESAT_SATUR, data= d_3,cores = 8, file ="mod.2_bS.rds")
+summary(mod.2_bSAT)
+
+###
+tab_model(mod.1_bSAT,mod.2_bSAT)
+# note these are very large files
 ```

### 2020-11-29 Manuscript_Bayesian.Rmd
near to final draft, latex table not right yet

--- 2020-11-27 Manuscript_Bayesian.Rmd
@@
 Here, we leverage nine waves of panel data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction ($N$ = 20979, New Zealand, 2010-2018).
-Contrary to previous research, we find that belief in a spirit or life force predicts lower personal well-being and life-satisfaction.
-However, in support of previous speculation, beliefs in a spirit or life force predict increasing personal well-being and life satisfaction over time relative to disbelief.
+Contrary to previous research, we find that belief in a spirit or Life Force predicts lower personal well-being and life-satisfaction.
+However, in support of previous speculation, beliefs in a spirit or Life Force predict increasing personal well-being and life satisfaction over time relative to disbelief.
 These finding are robust after to known demographic influences, and intriguingly, hold among those who believe in a God but disbelieve in a Spirit or Life Force.
@@
 keywords          : "Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being"
-wordcount         : "4685 words in text body, 1191 words in reference section"
+wordcount         : "5181 words in text body, 1191 words in reference section"
 bibliography      : ["b-references.bib"]
@@
 footnotelist      : no
-linenumbers       : yes
+linenumbers       : no
 mask              : no
@@
 classoption       : "man"
-output            : papaja::apa6_pdf
+output            : papaja::apa6_word
 extra_dependencies: ["rotating", "booktabs", "graphicx","longtable","amsmath","amsfonts","amssymb"]
@@
 First, previous research has frequently operationalised spirituality using measures of well-being.
-For example, The *Daily Spiritual Experience Scale*, which is widely used in spiritual mental health research, includes items such as "I feel thankful for my blessings" and "I feel deep inner peace or harmony" to measure spirituality [@Underwood2002-hg].
-These items are used both to measure spirituality and subjective well-being.
-As Koenig observes, it is unsurprising that spirituality is associated with well-being when the same items are used to measure both constructs.
-Koenig aptly describes this problematic approach as "tautological" [@Koenig2008-lv].
-How pervasive is this problem?
+For example, The *Daily Spiritual Experience Scale*, which is widely used in spiritual mental health research, includes items to measure spirituality such as "I feel thankful for my blessings" and "I feel deep inner peace or harmony" [@Underwood2002-hg].
+However, these items are also used to measure subjective well-being.
+Koenig describes this approach as "tautological" [@Koenig2008-lv]: it is unsurprising that spirituality is associated with well-being when the same items are used to measure both constructs.
+How pervasive is the use of tautological measures?
 A recent systematic-review of spirituality/mental health research found that nearly 45% of previous studies employ tautological scales [@Garssen2016-kb].
@@
 A second challenge is to disentangle concepts of spirituality from concepts of religion.
-The term "spirituality" was first used to describe the religious practices of ascetics and monks, however, more recently, the term has taken on a broader family of meanings [@Swinton2001-vr].
-Spirituality and health research does not converge on a unique set of meanings from the wider universe of popular meanings.
-Although it is commonplace across most areas of psychological science for researchers to explore different operationalisations of focal concepts, the psychology of spirituality has mostly been conducted with cross-sectional North American samples among participants who affiliate with traditional religion [@Ano2005-hx; @Hackney2003-rs; @Sawatzky2005-rw; @Smith2003-re;  @Visser2010-kq; @Yonker2012-zg].
-As such, the extent to which the psychology of spirituality and health has repeated or extended the psychology of religion and health remains unclear.
-Notably, researchers who have operationalised spirituality constructs as distinct from religion constructs have observed a negative association between spirituality and subjective well-being [@King2013-cg].
-Thus to infer how spirituality affects subjective well-being, it is important to disentangle the concepts of spirituality from concepts of religion, and to carefully investigate features of constructs in populations where these dimensions overlap, partially overlap, and diverge.
+The term "spirituality" was first used to describe the religious practices of ascetics and monks, however, more recently, the term has taken on a broader family of meanings that some of which apply to non-religious contexts [@Swinton2001-vr].
+Notably, spirituality and health research has not converged on a unique set of meanings.
+Although it is commonplace across most areas of psychological science for researchers to investigate different operationalisations of focal concepts (think of memory research, or emotions research), the psychology of spirituality has mostly been conducted with cross-sectional North American samples among participants who affiliate with traditional religion [@Ano2005-hx; @Hackney2003-rs; @Sawatzky2005-rw; @Smith2003-re;  @Visser2010-kq; @Yonker2012-zg].
+Whether the psychology of spirituality and health has repeated or extended the psychology of religion and health remains unclear.
+Put simply, cnceptual overlap between spirituality constructs and religion constructs creates a second line of terminological confusion that interacts with terminological confusion arising from overlap between spirituality constructs and well-being constructs.
+Indeed, researchers who have operationalised spirituality constructs as distinct from religion constructs have observed a negative association between spirituality and subjective well-being [@King2013-cg].
+Thus to infer how spirituality affects subjective well-being, it is not sufficient to disentange spirituality constructs from tautological well-being constructs, it is also important to disentangle the concept of spirituality from the concept of religion, and to carefully investigate the mechanisms of well-being in populations who overlap, partially overlap, and differ in these dimensions.
 
 A third challenge is to investigate the relationship between spirituality and subjective well-being in nationally diverse samples over individual life-spans.
-Reliable inference requires observing change within people who differ dimensions of spirituality over time, and assessing how their lives turn out [@Garssen2016-km].
-Previous investigations of spirituality and well-being, including high-quality reviews and meta-analyses, have relied on cross-sectional samples [@Ano2005-hx; @Hackney2003-rs; @Sawatzky2005-rw; @Smith2003-re; @Visser2010-kq; @Yonker2012-zg].
-However, To address the key questions in spirituality psychological health research requires longitudinal investigations.
-Unfortunately the relevant datasets are rare.
+To assess the mechanisms of well-being production requires observing change within people who differ in theoregically relevant ways and assessing how their lives turn out [@Garssen2016-km].
+However, previous investigations of spirituality and well-being, including high-quality reviews and meta-analyses, have nearly exclusively relied on cross-sectional samples [@Ano2005-hx; @Hackney2003-rs; @Sawatzky2005-rw; @Smith2003-re; @Visser2010-kq; @Yonker2012-zg].
+Although researching the key questions in spirituality psychological health research requires national longitudinal investigations, unfortunately the relevant datasets are rare.
 
 Here, we addresses these three challenges of previous research in the following ways: First, to avoid a tautological elision of "spirituality" and "well-being" we adopt a non-affective measure of spirituality as a "believe in a Spirit or Life Force." On the other side, to measure subjective well-being, we use a well-validated scale for personal well-being and a well-validated scale for Life-Satisfaction (see Method).
-Second, to disentangle spiritual beliefs from traditional religious beliefs and skeptical beliefs we adopt a four-level categorical indicator combining the binary indicator of belief in a Spirit or Life Force with the binary indicator "Do you believe in some form of a spirit or life force?" The levels of this four-level categorical indicator for spiritual, religious, and skeptical beliefs are as follows: (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who nevertheless believe in a Spirit or Life Force; (3) Those who believe in a God but do not believe in a Spirit or Life force; (4) Those who who believe in a both God and a Spirit or Life Force.
+Second, to disentangle spiritual beliefs from traditional religious beliefs and skeptical beliefs we adopt a four-level categorical indicator combining the binary indicator of belief in a Spirit or Life Force with the binary indicator "Do you believe in some form of a spirit or Life Force?" The levels of this four-level categorical indicator for spiritual, religious, and skeptical beliefs are as follows: (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life Force; (2) Skeptics about God who nevertheless believe in a Spirit or Life Force; (3) Those who believe in a God but do not believe in a Spirit or Life Force; (4) Those who who believe in a both God and a Spirit or Life Force.
 Clearly such an approach cannot simultaneously address all interests in previous spirituality/well-being research.
@@
 We analyzed data from participants who responded to our survey at least three times between Time 2 (2010) and Time 10 (2018), resulting in a sample of $N$ = 21,705 ($N$ = 20,979 with complete responses).
-
-## Measures
+Descriptive summaries for  demographic and theoretical indicators are given in Table 1.
 
@@
 \caption{}\label{}
-\scalebox{0.65}{
+\scalebox{0.6}{
 \begin{tabular}{ l c c c c c c c c c }
@@
 
+## Theoretical indicators 
 ### Life Satisfaction
@@
 Overall scores were means of the 2-items.
-For detailed information pertaining to means, standard deviations, and missingness, see Table 1.
-
-### Personal well-being 
-Personal well-being is a measure of function all well-being that was assessed using a 4-item version of the Australian Unity well-being Index [@Cummins2017-ur].
-Participants rated their satisfaction with (a) "Your standard of living"; (b) "Your health"; (c) "Your future security"; (d) "Your personal relationships." Items were rated using a 10-point response option ranging from 1 = completely dissatisfied to 10 = completely satisfied.
-The mean Cronbach’s alpha for this scale was $\alpha$ = 0.90.
-Higher scores on this scale indicate higher personal well-being.
-Overall scores were means of the 4-items.
-For detailed information pertaining to means, standard deviations, and missingness, see Table 1.
-
-### Belief in a Spirit or a God  
-We assess belief in a spirit or a God by asking two non-affective questions: "Do you believe in a spirit or life force" and "Do you believe in a god?" Responses were coded as (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who believe in a Spirit or Life Force; (3) Believers in a God who disbelieve in a Spirit or Life force; (4) Believers in a God and Spirit or Life Force.
-These believe were developed for the NZAVS from the 2005 Eurobarometer.
-For detailed information pertaining to beliefs, see Table 1.
-
-Table 1 about here:
 
@@
 Age was put into units of 10 years and centered at its mean. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 2.
-Gender.
-Gender was assessed by asking participants if they were "Male" was coded as "1" and "Female" was coded as "0." For detailed information pertaining to yearly responses and missingness, see Table 1.
+Gender. Gender was assessed by asking participants if they were "Male" was coded as "1" and "Female" was coded as "0."
 
@@
 
+### Personal well-being 
+Personal well-being is a measure of function all well-being that was assessed using a 4-item version of the Australian Unity well-being Index [@Cummins2017-ur].
+Participants rated their satisfaction with (a) "Your standard of living"; (b) "Your health"; (c) "Your future security"; (d) "Your personal relationships." Items were rated using a 10-point response option ranging from 1 = completely dissatisfied to 10 = completely satisfied.
+The mean Cronbach’s alpha for this scale was $\alpha$ = 0.90.
+Higher scores on this scale indicate higher personal well-being.
+Overall scores were means of the 4-items.
+
+### Belief in a Spirit/Life Force and/or a God  
+We assess belief in a spirit or a God by asking two non-affective questions: "Do you believe in a spirit or Life Force" and "Do you believe in a god?" Responses were coded as (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life Force; (2) Skeptics about God who believe in a Spirit or Life Force; (3) Believers in a God who disbelieve in a Spirit or Life Force; (4) Believers in a God and Spirit or Life Force.
+These believe were developed for the NZAVS from the 2005 Eurobarometer.
+
 ## Statistical Analyses
 Statistical analysis was performed using R version 4.0.2 R version 3.6.1 (2020-06-22) 2019-07-05).
-We analyze data from participants who responded to the NZAVS survey at least three times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of  21,705 (N = 20,979 with complete responses).
+We analyze data from participants who responded to the NZAVS survey at least three times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of  21,705 (*N* = 20,979 with complete responses).
 To model growth in subjective well-being we use generalized linear mixed models.
@@
 The baseline for this indicator is Skepticism about a Spirit/Life Forces and Skepticism about a God.
-To handle the dependencies introduced from the repeated measures, we include individual ID as an effect modeled as random, and estimated a random effect for the slope for years.
-The model equations are given below and priors are given below.
+To handle the dependencies introduced from the repeated measures, we include individual ID as an effect modeled as random, and we estimated a random effect for the slope for years.
+This allowed us to model correllation between individual-level intercept (starting high or low in Personal Well-Being/Life Satisfaction) and stability/growth/decline in the slope effect of years.
+We estimated all statitiscal models using both REML estimation [@R-lme4] and Bayesian estimation [@BRMSpackage] with weakly informative priors.
+We include the full reports in the Supplement.
+The REML estimates and the Bayesian estimates converged to within three decimal places.
+We include model equations and priors at the end of this article.
+Here we report the results of the Bayesian estimator.
 
-## Potential Confounders and Missing Data 
+## Potential Confounders
+
+To address confounding we first obtained the New Zealand Attitudes and Values Study standard demographic indicators recommended to address confounding: [NZAVS modelling standards](https://cdn.auckland.ac.nz/assets/psych/about/our-research/nzavs/NZAVSTechnicalDocuments/NZAVS-Technical-Documents-e11-Default-Statistical-Standards.pdf).
+Next we modelled potential relationships between the coefficients within the model using the daggity package in R, and plotting results using the daggity and ggdag packages in R.
+
+This method reveals two adjustments sets equired to avoid potential confounding: \{age, employment, male, political conservativism, and time\}, and \{age, ethnicty, male, and time \}.Here, we report the second set because employment is likely to contain more measurement error: some are not employed because they cannot find employment, others because they are not seeking employment (e.g.
+retirement).
+Results were insensitive to the choice set.
+Indeed, the inference proved robust to the simple model with only the interaction of years and beliefs, and robust to the fully saturated model in which all covariates were included (see the online supplement for details).
+In the fully saturated model, it appeared that those skeptical of a Spirit or Life Force, exhibited no growth in Persona Well-Being and Life-Satisifaction over the nine years.
+Although this result offers stronger stronger evidence for a Spirit/Life Force well-being mechanism, it is likely the effect of collider confounding and masking.
+We discuss this point in more detail in the supplement, and mention it here to caution researchers from "everything-but-the-kitchen-sink" approaches to regression modelling.
+
+## Missing Data 
 Because missing values can result in biased estimates [@Blackwell2017-oq], we multiply imputed 10 datasets using the AmeliaII package in R[@Honaker2011-yu].We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences.
 Additionally the results were robust to filtering of participants of N > 3 to N = 8 waves.
-In a supplemental analysis we include standard demographic indicators used in New Zealand Attitudes and Values Research: [NZAVS modelling standards](https://cdn.auckland.ac.nz/assets/psych/about/our-research/nzavs/NZAVSTechnicalDocuments/NZAVS-Technical-Documents-e11-Default-Statistical-Standards.pdf).
-We centered and scaled the education, deprivation, and political conservativism.
-We centred age was centred at and put into decade-long intervals.
-There are increasing worries
-
-We provide the analyses of these additional analyses in the supplement.  The supplement describing these analysis is located at: https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing.
+We include the multiple-imputation results in a supplement.
+Here, there were no differences in estimation relevant to inference, however, to improve inference, we advocate wider use of multiple-imputation or Bayesian approaches for handling missingness.
 
@@
 \begin{center}
-\scalebox{0.8}{
+\scalebox{0.5}{
 \begin{tabular}{l c c}
@@
 \midrule
-Intercept                           & $\mathbf{7.04}$   & $\mathbf{5.19}$  \\
-                                    & $ [ 7.02;  7.07]$ & $ [ 5.17; 5.21]$ \\
+Intercept                           & $\mathbf{7.14}$   & $\mathbf{5.29}$   \\
+                                    & $ [ 7.11;  7.18]$ & $ [ 5.26;  5.31]$ \\
 yearsC                              & $\mathbf{0.02}$   & $\mathbf{0.01}$   \\
-                                    & $ [ 0.02;  0.03]$ & $ [ 0.01; 0.02]$ \\
-Beliefs\_SpiritExcludesGod\_        & $\mathbf{-0.06}$  & $0.02$           \\
-                                    & $ [-0.09; -0.03]$ & $ [-0.00; 0.04]$ \\
-BeliefsGodAndSpirit                 & $\mathbf{-0.06}$  & $\mathbf{0.05}$  \\
-                                    & $ [-0.10; -0.03]$ & $ [ 0.03; 0.07]$ \\
-BeliefsGodExcludesSpirit            & $-0.00$           & $\mathbf{0.08}$  \\
-                                    & $ [-0.04;  0.04]$ & $ [ 0.04; 0.11]$ \\
+                                    & $ [ 0.01;  0.02]$ & $ [ 0.00;  0.01]$ \\
+Beliefs\_SpiritExcludesGod\_        & $\mathbf{-0.06}$  & $0.01$            \\
+                                    & $ [-0.09; -0.03]$ & $ [-0.01;  0.03]$ \\
+BeliefsGodAndSpirit                 & $\mathbf{-0.06}$  & $\mathbf{0.03}$   \\
+                                    & $ [-0.09; -0.03]$ & $ [ 0.01;  0.06]$ \\
+BeliefsGodExcludesSpirit            & $-0.00$           & $\mathbf{0.06}$   \\
+                                    & $ [-0.04;  0.04]$ & $ [ 0.03;  0.09]$ \\
+Age\_in\_Decades\_C                 & $\mathbf{0.07}$   & $\mathbf{0.08}$   \\
+                                    & $ [ 0.06;  0.09]$ & $ [ 0.07;  0.09]$ \\
+Ethnic\_CategoriesMaori             & $\mathbf{-0.38}$  & $\mathbf{-0.11}$  \\
+                                    & $ [-0.43; -0.33]$ & $ [-0.14; -0.08]$ \\
+Ethnic\_CategoriesPacific           & $\mathbf{-0.38}$  & $\mathbf{-0.15}$  \\
+                                    & $ [-0.48; -0.29]$ & $ [-0.22; -0.09]$ \\
+Ethnic\_CategoriesAsian             & $\mathbf{-0.09}$  & $-0.05$           \\
+                                    & $ [-0.18; -0.01]$ & $ [-0.11;  0.02]$ \\
+Male1                               & $\mathbf{-0.06}$  & $\mathbf{-0.16}$  \\
+                                    & $ [-0.10; -0.02]$ & $ [-0.19; -0.13]$ \\
 yearsC:Beliefs\_SpiritExcludesGod\_ & $\mathbf{0.01}$   & $\mathbf{0.01}$   \\
-                                    & $ [ 0.00;  0.02]$ & $ [ 0.01; 0.02]$ \\
+                                    & $ [ 0.00;  0.02]$ & $ [ 0.00;  0.02]$ \\
 yearsC:BeliefsGodAndSpirit          & $\mathbf{0.01}$   & $\mathbf{0.01}$   \\
@@
 \midrule
-SD: Idsd(Intercept)                 & $1.44$            & $0.99$           \\
+SD: Idsd(Intercept)                 & $1.42$            & $0.98$            \\
 SD: Idsd(yearsC)                    & $0.13$            & $0.10$            \\
-SD: Idcor(Intercept,yearsC)         & $-0.02$           & $0.03$           \\
+SD: Idcor(Intercept,yearsC)         & $-0.03$           & $0.03$            \\
 \bottomrule
@@
 \caption{}
-\label{tab:REGRESS}
+\label{tab:REGRESS_LS}
 \end{center}
@@
 ```{r fig.cap = "Coefficient Graphs", echo =FALSE,out.width="\\textwidth"}
-knitr::include_graphics("/Users/jbul176/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/coefficientplotsbayes.png")
+knitr::include_graphics("/Users/jbul176/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/coefBayesDag.png")
 ```
@@
 
-  - The effect of years is positive is reliable (posterior median = 0.025, HDI [0.020, 0.031] pd = 100.00%).
-We infer that personal well-being increased in the skeptic population between 2010 and 2018.
-  - The effect of beliefs in *Spirit Excludes God*  is reliably negative (posterior median = -0.058, HDI [-0.080, -0.035], pd = 100.00%).
+  - The intercept, representing the expected personal well-being among skeptics about a Spirit/Life Force and God who are Europeans, 50.77 years of age and do not identify as male is reliable (posterior median = 7.138 HDI [ 7.111,  7.167], pd = 100.00%).
+  - The effect of years is positive is reliable (posterior median =  7.138, HDI [ 7.111,  7.167], pd  100.00%).
+We infer that personal well-being increased in the baseline skeptic population between 2010 and 2018.
+  - The effect of beliefs in *Spirit Excludes God*  is reliably negative (posterior median =  -0.059, HDI [-0.081, -0.036], pd 100.00%).
 We infer that the population that believed in a Spirit but did not believe in a God was lower in personal well-being at the intercept compared to skeptics.
-  - The effect of beliefs in *God And Spirit*  is reliably negative (posterior median = -0.064, HDI [-0.091, -0.037], pd = 100.00%).
+  - The effect of beliefs in *God And Spirit*  is not  negative (posterior median = -0.060, HDI [-0.086, -0.034], pd = 100.00%) .
 We infer that the population that believed in a God and a Spirit or Life Force was lower in personal well-being at the intercept compared to skeptics.
-  - The effect of beliefs in *God Excludes Spirit*  is not reliable (posterior median = -0.002 HDI [-0.039,  0.033], pd = 53.50%).
+  - The effect of beliefs in *God Excludes Spirit*  is not reliable (posterior median = -0.001 HDI [-0.036,  0.035], pd = 51.10%).
 We infer that the population that believed in a God and doubted a Spirit or Life Force were similar were similar in personal well-being at the intercept to skeptics.
+  - The effect of Age in Decades (C) is   0.072 | [ 0.060,  0.083] | 100.00% |
+  - The effect of Maori Ethnicity is  -0.380 | [-0.420, -0.341] | 100.00% | 
+  - The effect of Pacific Ethnicity is  -0.384 | [-0.464, -0.312] | 100.00% |
+  - The effect of Asian Ethnicity is -0.093 | [-0.165, -0.023] |  98.28% |  
+  - The effect of Male gender is  -0.060 | [-0.094, -0.026] |  99.90% |
 
@@
 ```{r fig.cap = "Personal Well-being prediction graph", echo =FALSE, out.width="\\textwidth"}
-knitr::include_graphics("/Users/jbul176/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/Bayesian_pwi_predicted.png")
+knitr::include_graphics("/Users/jbul176/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/pwi_confounder_adjusted.png")
 ```
 
-The effect of years $\times$ belief in a *Spirit Excludes God*  is reliably positive  (posterior median = 0.010, HDI [ 0.003,  0.017], pd = 99.02%).
-We infer that people with beliefs in a spirit or life force who do not believe in a God tended to grow in their personal well-being over time at a faster rate than skeptics grew.
-  - The effect of years $\times$ in *God And Spirit*  is reliably positive  (posterior median = 0.014, HDI [0.008, 0.021], pd = 99.98%).
-We infer that people who believe in a spirit or life force who also believe in a God tended to grow in their personal well-being over time at a faster rate than skeptics grew.
-  - The effect of years $\times$ belief in *God Excludes Spirit*  is not reliable (posterior median = 0.002, HDI [-0.009,  0.015], pd =  63.73%).
-We infer that people who believe in a God but not a spirit or life force did not grow at a faster rate in personal well-being than skeptics grew.
+The effect of years $\times$ belief in a *Spirit Excludes God*  is reliably positive  (posterior median = 0.011, HDI [ 0.003,  0.018], pd = 99.02%).
+We infer that people with beliefs in a spirit or Life Force who do not believe in a God tended to grow in their personal well-being over time at a faster rate than skeptics grew.
+  - The effect of years $\times$ in *God And Spirit*  is reliably positive  (posterior median = 0.014, HDI [0.007, 0.021], pd = 99.98%).
+We infer that people who believe in a spirit or Life Force who also believe in a God tended to grow in their personal well-being over time at a faster rate than skeptics grew.
+  - The effect of years $\times$ belief in *God Excludes Spirit*  is not reliable (posterior median = 0.004, HDI [-0.009,  0.015], pd =  68.73%).
+We infer that people who believe in a God but not a spirit or Life Force did not grow at a faster rate in personal well-being than skeptics grew.
 
@@
 Within this model: 
-
-  - The effect of years is positive is reliable (posterior median =  0.014, HDI [ 0.010, 0.019], pd = 100.00%).
+  - The intercept, representing the expected life satisfaction among skeptics about a Spirit/Life Force and God who are Europeans, 50.77 years of age and do not identify as male is reliable (posterior median = 5.287, HDI [ 5.267,  5.306], pd = 100.00%).
+  - The effect of years is positive is reliable (posterior median =  0.008, HDI [ 0.003, 0.011], pd = 99.75%).
 We infer that life satisfaction increased in the skeptic population between 2010 and 2018.
-  -The effect of beliefs in *Spirit Excludes God*  is reliably positive (posterior median =   0.018, HDI [ 0.001, 0.035], pd = 96.05%).
+  -The effect of beliefs in *Spirit Excludes God*  is not reliable (posterior median =   0.007, HDI [-0.010,  0.023], pd = 75.33%).
 We infer that the population that believed in a Spirit but did not believe in a God was higher in life satisfaction at the intercept compared with skeptics.
-  - The effect of beliefs in *God And Spirit*  is reliably positive (posterior median = 0.051, HDI [ 0.031, 0.071], pd = 100.00%).
+  - The effect of beliefs in *God And Spirit*  is reliably positive (posterior median = 0.034, HDI [ 0.015, 0.053], pd = 99.83%).
 We infer that the population that believed in a God and a Spirit or Life Force was higher in life satisfaction at the intercept compared with skeptics.
-  - The effect of beliefs in *God Excludes Spirit*  is reliably positive (posterior median =  0.076, HDI [ 0.048, 0.102], pd = 100.00%).
+  - The effect of beliefs in *God Excludes Spirit*  is reliably positive (posterior median =  0.057, HDI [ 0.030, 0.082], pd = 99.95%).
 We infer that the population that believed in a God and doubted a Spirit or Life Force  was higher in life satisfaction at the intercept compared with skeptics.
+  - The effect of Age in Decades (C) is reliably positive (posterior median = 0.079, hDI [ 0.071,  0.087] pd = 100.00%).
+  - The effect of Maori Ethnicity is reliably negative (posterior median = -0.109 | [-0.137, -0.082], pd = 100.00%).
+  - The effect of Pacific Ethnicity is reliably negative (posterior median = -0.153 | [-0.210, -0.103], pd = 100.00%).
+  - The effect of Asian Ethnicity is not reliable (posterior median = -0.047, [-0.098,  0.004], pd = 93.08%).
+  - The effect of Male gender is reliabily negative (posterior median = -0.161, [-0.184, -0.138], pd = 100.00%)
 
@@
 ```{r  fig.cap = "Life satisfaction prediction graph", echo = FALSE,out.width="\\textwidth"}
-knitr::include_graphics("/Users/jbul176/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/Bayesian_lifesat_predicted.png")
+knitr::include_graphics("/Users/jbul176/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/ls_counfounder_adjusted.png")
 ```
-  - The interaction of years $\times$ beliefs in *Spirit Excludes God* is reliably positive (posterior median = 0.012, HDI [ 0.007, 0.017], pd = 100.00%).
+  - The interaction of years $\times$ beliefs in *Spirit Excludes God* is reliably positive (posterior median = 0.011, HDI [ 0.006, 0.016], pd = 99.95%).
 We infer that people with beliefs in a Spirit or Life Force who do not believe in a God tended to grow over time in their life satisfaction at a faster rate than skeptics.
-  - The interaction of years $\times$ *God And Spirit*  is reliably positive (posterior median =  0.010, HDI [ 0.004, 0.015], pd = 99.90%).
-We infer that people who believe in a spirit or life force who also believe in a God tended to grow over time in their life satisfaction at a faster rate than skeptics.
-  - The effect of years $\times$ beliefs in *God Excludes Spirit*  is not reliable (posterior median = -0.002, HDI [-0.011, 0.007], pd = 62.80%).
-We infer that people who believe in a God but not a spirit or life force tended to grow over time in their life satisfaction at a similar rate to skeptics.
+  - The interaction of years $\times$ *God And Spirit*  is reliably positive (posterior median =  0.009, HDI [ 0.004, 0.014], pd = 99.72%).
+We infer that people who believe in a spirit or Life Force who also believe in a God tended to grow over time in their life satisfaction at a faster rate than skeptics.
+  - The effect of years $\times$ beliefs in *God Excludes Spirit*  is not reliable (posterior median = -0.003, HDI [-0.012, 0.006], pd = 72.28%).
+We infer that people who believe in a God but not a spirit or Life Force tended to grow over time in their life satisfaction at a similar rate to skeptics.
 
@@
 
-The purpose of this study has been to clarify the relationship between belief in a spirit/life force and subjective well-being, addressing three challenges from previous research: (1) tautological measures of spirituality and subjective well-being; (2) insufficiently distinct measures of spirituality and religion; (3) cross-sectional samples.
+The purpose of this study has been to clarify the relationship between belief in a spirit/Life Force and subjective well-being, addressing three challenges from previous research: (1) tautological measures of spirituality and subjective well-being; (2) insufficiently distinct measures of spirituality and religion; (3) cross-sectional samples.
 Here, we ask whether beliefs in a Spirit or Life Force predict growth in personal well-being and life-satisfaction over 9 waves (8 years) in nationally diverse western populations who differ in the dimensions of spiritual and religious beliefs.
@@
 1.
-At the outset of the study, in 2010, the expected subjective well-being as measured using the personal well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God ($\hat{y} =$  6.80 [6.76, 6.85]) and lower among those who believed in both a Spirit/Life-Force and a God ($\hat{y}$ =  6.78 [6.74, 6.81]) in comparison both to the skeptical population who neither believed in a God nor a life force ($\hat{y}$ = 6.92 [6.87, 6.96]) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 6.90 [6.83, 6.98]).
-(See Figure 2)  Similarly,the expected subjective well-being as measured using the life satisfaction Inventory was lowest in 2010 among those who believed in Spirit or Life Force but not a God ($\hat{y}$ = 5.07 [5.04, 5.10]).
-Life satisfaction was similar among those who believed in both a Spirit/Life-Force and a God  ($\hat{y}$ =  5.11 [5.08, 5.13]) in comparison both with the skeptical population who neither believed in a God nor a life force ($\hat{y}$ = 5.12 [5.08, 5.15]).
-Life satisfaction was highest among the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 5.19 [5.14, 5.25]) (see Figure 3)
+At the outset of the study, in 2010, the expected subjective well-being as measured using the personal well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God ($\hat{y} =$  6.93 | [6.88, 6.98]) and lower among those who believed in both a Spirit/Life Force and a God ($\hat{y}$ = 6.91 | [6.87, 6.95]) in comparison both to the skeptical population who neither believed in a God nor Spirit/ Life Force ($\hat{y}$ = 7.04 | [6.99, 7.10]) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 7.03 | [6.95, 7.10]).
+(See Figure 2)  Similarly,the expected subjective well-being as measured using the life satisfaction Inventory was lower in 2010 among those who believed in Spirit or Life Force but not a God ($\hat{y}$ =   5.20 | [5.16, 5.23]).
+Life satisfaction was similar among those who believed in both a Spirit/Life Force and a God  ($\hat{y}$ =   5.23 | [5.21, 5.26]) in comparison both with the skeptical population who neither believed in a God nor a Life Force ($\hat{y}$ = 5.25 | [5.21, 5.29]).
+Life satisfaction was highest among the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 5.32 | [5.26, 5.38]) (see Figure 3)
 
@@
 However, over time, belief in a Spirit or Life Force predicts a trajectory of growth in personal well-being.
-In 2018, personal well-being among who believe in Spirit or Life Force but not a God had grown to ($\hat{y} =  7.12$ [7.09, 7.16], $\Delta \hat{y} = 0.32$).
-Among those who believed in both a Spirit/Life-Force and a God, personal well-being had grown to ($\hat{y} =  7.13$ [7.10, 7.16], $\Delta \hat{y} = 0.35$).
-By contrast, personal well-being among the skeptical population who neither believed in a God nor a Life Force grew less steeply across the eight year interval $\hat{y} = 7.14$ [7.11, 7.18], $\Delta \hat{y} = 0.22$).
-By the same token, personal well-being among those who believe in a God but are skeptical about a Spirit/Life Force grew less steeply $\hat{y} = 7.15$, [7.09, 7.21] $\Delta \hat{y} = 0.25$).
+In 2018, personal well-being among who believe in Spirit or Life Force but not a God had grown to ($\hat{y} =   7.19$ [7.15, 7.23], $\Delta \hat{y} = 0.26$).
+Among those who believed in both a Spirit/Life Force and a God, personal well-being had grown to ($\hat{y} = 7.20$ [7.17, 7.24], $\Delta \hat{y} = 0.29$).
+By contrast, personal well-being among the skeptical population who neither believed in a God nor a Life Force grew less steeply across the eight year interval $\hat{y} =  7.21$ [7.17, 7.25], $\Delta \hat{y} = 0.17$).
+By the same token, personal well-being among those who believe in a God but are skeptical about a Spirit/Life Force grew less steeply $\hat{y} =  7.22$ [7.16, 7.28] $\Delta \hat{y} = 0.19$).
+In practical terms, the rate of increase in personal wellbeing among those with Spirit/Life Force beliefs was about 30% greater than it was among those who lacked such Spirit/Life Force (See Figure 2).
 Similarly, over time, belief in a Spirit or Life Force predicts a trajectory of growth in life satisfaction.
-In 2018, the expected effect of spiritual beliefs on life satisfaction among those who believe in Spirit or Life Force but not a God had grown to ($\hat{y} =  5.31$ [5.28, 5.33], $\Delta \hat{y} = 0.23$).
-In practical terms, such growth is equivalent to the expected gain in life satisfaction from being employed and a sixth as large as the expected benefit to life satisfaction from having a romantic partner.
-In practical terms, the rate of increase in personal wellbeing among those with Spirit/Life force beliefs was about one and half times greater than it was among those who lacked such beliefs (See Figure 2).
-Among those who believed in both a Spirit/Life-Force and a God, life satisfaction had grown to ($\hat{y} = 5.33 [5.31, 5.35]$, $\Delta \hat{y} = 0.24$).
-By contrast, life satisfaction among the skeptical population who neither believed in a God nor a Life Force lagged behind over the eight year interval $\hat{y} =  5.24 [5.21, 5.27], \Delta \hat{y} = 0.13$).
-Relatedly, life satisfaction in the population that believe in God but are skeptical about a Spirit/Life Force grew less steeply $\hat{y} =  5.31 [5.26, 5.36], \Delta \hat{y} = 0.11$).
-In practical terms, the rate of increase in life satisfaction among those with Spirit/Life force beliefs was twice as large (See Figure 3).
+In 2018, the expected effect of spiritual beliefs on life satisfaction among those who believe in Spirit or Life Force but not a God had grown to ($\hat{y} =  5.37$ [5.34, 5.40], $\Delta \hat{y} = 0.17$).
+Among those who believed in both a Spirit/Life Force and a God, life satisfaction had grown to ($\hat{y} = 5.39$  [5.36, 5.41], $\Delta \hat{y} = 0.19$).By contrast, life satisfaction among the skeptical population who neither believed in a God nor a Life Force lagged behind over the eight year interval $\hat{y} =   5.32$ [5.29, 5.35], $\Delta \hat{y} = 0.07$).
+Relatedly, life satisfaction in the population that believe in God but are skeptical about a Spirit/Life Force grew less steeply $\hat{y} =  5.36$ [5.31, 5.41], $\Delta \hat{y} = 0.04$).
+In practical terms, the rate of increase in life satisfaction among those with Spirit/Life Force beliefs over three times greater than among those that lacked such beliefs (See Figure 3).
 
@@
 Inferred growth from spiritual beliefs is evident from 9 annual waves of data collection (8 years), and across a nationally diverse sample.
-On this basis of this longitudinal information accumulated across nationally, we are confident that growth in personal well-being and in life-satisfaction among those who believe in a Spirit and Life force are evident.
-Such effects co-occur in people who lack a belief in a God as well as among people who believe in a God as well as a Spirit or Life force.
+On this basis of this longitudinal information accumulated across nationally, we are confident that growth in personal well-being and in life-satisfaction among those who believe in a Spirit and Life Force are evident.
+Such effects co-occur in people who lack a belief in a God as well as among people who believe in a God as well as a Spirit or Life Force.
 
@@
 Because spiritual belief items were only introduced during the year of the earthquakes, we cannot here resolve whether the shifting relationship between spirit beliefs and well-being were the result of demand for spiritual beliefs among those secular and religious people who were adversely affected by the financial crisis or natural disaster.
-On the other hand, such exogenous influences might have adversely affected those with Spirit/Life force beliefs in the early phases of this study, and the patterns observed here are better understood as a regression to a population mean.
+On the other hand, such exogenous influences might have adversely affected those with Spirit/Life Force beliefs in the early phases of this study, and the patterns observed here are better understood as a regression to a population mean.
 Though beyond the scope of this study, how subjective well-being changes in response to changes in spiritual beliefs, as well as special sensitivities to distress among people with spiritual beliefs are important questions for future investigations.
-Despite the longitudinal quality of this study, a causal explanation remains elusive.
 
@@
 We have not attempted to model the diverse philosophical, social, and behavioral attributes within people who adhere to spirit beliefs.
-Moreover, people might identify as spiritual without having any accompanying belief in a Spirit or Life force.
+Moreover, people might identify as spiritual without having any accompanying belief in a Spirit or Life Force.
 The task of clarifying how distinct features across a diverse family of "spirituality" concepts affect perceptions of individual well-being remain intriguing horizons.
@@
 Western populations are increasingly identifying as "spiritual but not religious" [@Zinnbauer2005-vz; @Swinton2001-vr].
-Here, we observe that even among people who lack a traditional religious belief in God, believing in a Spirit or Life force predicts incremental improvements in well-being that sum to practically large effects.
+Here, we observe that even among people who lack a traditional religious belief in God, believing in a Spirit or Life Force predicts incremental improvements in well-being that sum to practically large effects.
 Though the contextual and psychological mechanisms that underpin such growth will require many decades of research, this preliminary finding is good news for those increasing numbers of individuals continue to report spiritual beliefs but who do not affirm traditional religious beliefs.

### 2020-11-29 Manuscript_Bayesian.tex
added saturated models for the bayesian estimate. Not distinguishable from the REML models.
near to final draft, latex table not right yet

@@
 \shorttitle{Longitudinal Wellbeing from Spiritual Beliefs}
-\keywords{Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being\newline\indent Word count: 4685 words in text body, 1191 words in reference section}
+\keywords{Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being\newline\indent Word count: 5181 words in text body, 1191 words in reference section}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 \makeatother
-\usepackage{lineno}
-
-\linenumbers
 \usepackage{csquotes}
@@
 Here, we leverage nine waves of panel data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction (\(N\) = 20979, New Zealand, 2010-2018).
-Contrary to previous research, we find that belief in a spirit or life force predicts lower personal well-being and life-satisfaction.
-However, in support of previous speculation, beliefs in a spirit or life force predict increasing personal well-being and life satisfaction over time relative to disbelief.
+Contrary to previous research, we find that belief in a spirit or Life Force predicts lower personal well-being and life-satisfaction.
+However, in support of previous speculation, beliefs in a spirit or Life Force predict increasing personal well-being and life satisfaction over time relative to disbelief.
 These finding are robust after to known demographic influences, and intriguingly, hold among those who believe in a God but disbelieve in a Spirit or Life Force.
@@
 First, previous research has frequently operationalised spirituality using measures of well-being.
-For example, The \emph{Daily Spiritual Experience Scale}, which is widely used in spiritual mental health research, includes items such as \enquote{I feel thankful for my blessings} and \enquote{I feel deep inner peace or harmony} to measure spirituality (Underwood \& Teresi, 2002).
-These items are used both to measure spirituality and subjective well-being.
-As Koenig observes, it is unsurprising that spirituality is associated with well-being when the same items are used to measure both constructs.
-Koenig aptly describes this problematic approach as \enquote{tautological} (Koenig, 2008).
-How pervasive is this problem?
+For example, The \emph{Daily Spiritual Experience Scale}, which is widely used in spiritual mental health research, includes items to measure spirituality such as \enquote{I feel thankful for my blessings} and \enquote{I feel deep inner peace or harmony} (Underwood \& Teresi, 2002).
+However, these items are also used to measure subjective well-being.
+Koenig describes this approach as \enquote{tautological} (Koenig, 2008): it is unsurprising that spirituality is associated with well-being when the same items are used to measure both constructs.
+How pervasive is the use of tautological measures?
 A recent systematic-review of spirituality/mental health research found that nearly 45\% of previous studies employ tautological scales (Garssen et al., 2016).
@@
 A second challenge is to disentangle concepts of spirituality from concepts of religion.
-The term \enquote{spirituality} was first used to describe the religious practices of ascetics and monks, however, more recently, the term has taken on a broader family of meanings (Swinton, 2001).
-Spirituality and health research does not converge on a unique set of meanings from the wider universe of popular meanings.
-Although it is commonplace across most areas of psychological science for researchers to explore different operationalisations of focal concepts, the psychology of spirituality has mostly been conducted with cross-sectional North American samples among participants who affiliate with traditional religion (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky, Ratner, \& Chiu, 2005; Smith et al., 2003; Visser, Garssen, \& Vingerhoets, 2010; Yonker, Schnabelrauch, \& Dehaan, 2012).
-As such, the extent to which the psychology of spirituality and health has repeated or extended the psychology of religion and health remains unclear.
-Notably, researchers who have operationalised spirituality constructs as distinct from religion constructs have observed a negative association between spirituality and subjective well-being (King et al., 2013).
-Thus to infer how spirituality affects subjective well-being, it is important to disentangle the concepts of spirituality from concepts of religion, and to carefully investigate features of constructs in populations where these dimensions overlap, partially overlap, and diverge.
+The term \enquote{spirituality} was first used to describe the religious practices of ascetics and monks, however, more recently, the term has taken on a broader family of meanings that some of which apply to non-religious contexts (Swinton, 2001).
+Notably, spirituality and health research has not converged on a unique set of meanings.
+Although it is commonplace across most areas of psychological science for researchers to investigate different operationalisations of focal concepts (think of memory research, or emotions research), the psychology of spirituality has mostly been conducted with cross-sectional North American samples among participants who affiliate with traditional religion (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky, Ratner, \& Chiu, 2005; Smith et al., 2003; Visser, Garssen, \& Vingerhoets, 2010; Yonker, Schnabelrauch, \& Dehaan, 2012).
+Whether the psychology of spirituality and health has repeated or extended the psychology of religion and health remains unclear.
+Put simply, cnceptual overlap between spirituality constructs and religion constructs creates a second line of terminological confusion that interacts with terminological confusion arising from overlap between spirituality constructs and well-being constructs.
+Indeed, researchers who have operationalised spirituality constructs as distinct from religion constructs have observed a negative association between spirituality and subjective well-being (King et al., 2013).
+Thus to infer how spirituality affects subjective well-being, it is not sufficient to disentange spirituality constructs from tautological well-being constructs, it is also important to disentangle the concept of spirituality from the concept of religion, and to carefully investigate the mechanisms of well-being in populations who overlap, partially overlap, and differ in these dimensions.
 
 A third challenge is to investigate the relationship between spirituality and subjective well-being in nationally diverse samples over individual life-spans.
-Reliable inference requires observing change within people who differ dimensions of spirituality over time, and assessing how their lives turn out (Garssen \& Visser, 2016).
-Previous investigations of spirituality and well-being, including high-quality reviews and meta-analyses, have relied on cross-sectional samples (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky et al., 2005; Smith et al., 2003; Visser et al., 2010; Yonker et al., 2012).
-However, To address the key questions in spirituality psychological health research requires longitudinal investigations.
-Unfortunately the relevant datasets are rare.
+To assess the mechanisms of well-being production requires observing change within people who differ in theoregically relevant ways and assessing how their lives turn out (Garssen \& Visser, 2016).
+However, previous investigations of spirituality and well-being, including high-quality reviews and meta-analyses, have nearly exclusively relied on cross-sectional samples (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky et al., 2005; Smith et al., 2003; Visser et al., 2010; Yonker et al., 2012).
+Although researching the key questions in spirituality psychological health research requires national longitudinal investigations, unfortunately the relevant datasets are rare.
 
 Here, we addresses these three challenges of previous research in the following ways: First, to avoid a tautological elision of \enquote{spirituality} and \enquote{well-being} we adopt a non-affective measure of spirituality as a \enquote{believe in a Spirit or Life Force.} On the other side, to measure subjective well-being, we use a well-validated scale for personal well-being and a well-validated scale for Life-Satisfaction (see Method).
-Second, to disentangle spiritual beliefs from traditional religious beliefs and skeptical beliefs we adopt a four-level categorical indicator combining the binary indicator of belief in a Spirit or Life Force with the binary indicator \enquote{Do you believe in some form of a spirit or life force?} The levels of this four-level categorical indicator for spiritual, religious, and skeptical beliefs are as follows: (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who nevertheless believe in a Spirit or Life Force; (3) Those who believe in a God but do not believe in a Spirit or Life force; (4) Those who who believe in a both God and a Spirit or Life Force.
+Second, to disentangle spiritual beliefs from traditional religious beliefs and skeptical beliefs we adopt a four-level categorical indicator combining the binary indicator of belief in a Spirit or Life Force with the binary indicator \enquote{Do you believe in some form of a spirit or Life Force?} The levels of this four-level categorical indicator for spiritual, religious, and skeptical beliefs are as follows: (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life Force; (2) Skeptics about God who nevertheless believe in a Spirit or Life Force; (3) Those who believe in a God but do not believe in a Spirit or Life Force; (4) Those who who believe in a both God and a Spirit or Life Force.
 Clearly such an approach cannot simultaneously address all interests in previous spirituality/well-being research.
@@
 We analyzed data from participants who responded to our survey at least three times between Time 2 (2010) and Time 10 (2018), resulting in a sample of \(N\) = 21,705 (\(N\) = 20,979 with complete responses).
-
-\hypertarget{measures}{%
-\subsection{Measures}\label{measures}}
+Descriptive summaries for demographic and theoretical indicators are given in Table 1.
 
@@
 
+\hypertarget{theoretical-indicators}{%
+\subsection{Theoretical indicators}\label{theoretical-indicators}}
+
 \hypertarget{life-satisfaction}{%
@@
 Overall scores were means of the 2-items.
-For detailed information pertaining to means, standard deviations, and missingness, see Table 1.
-
-\hypertarget{personal-well-being}{%
-\subsubsection{Personal well-being}\label{personal-well-being}}
-
-Personal well-being is a measure of function all well-being that was assessed using a 4-item version of the Australian Unity well-being Index (Cummins, Eckerseley, Pallant, Van Vugt, \& Misajon, 2017).
-Participants rated their satisfaction with (a) \enquote{Your standard of living}; (b) \enquote{Your health}; (c) \enquote{Your future security}; (d) \enquote{Your personal relationships.} Items were rated using a 10-point response option ranging from 1 = completely dissatisfied to 10 = completely satisfied.
-The mean Cronbach's alpha for this scale was \(\alpha\) = 0.90.
-Higher scores on this scale indicate higher personal well-being.
-Overall scores were means of the 4-items.
-For detailed information pertaining to means, standard deviations, and missingness, see Table 1.
-
-\hypertarget{belief-in-a-spirit-or-a-god}{%
-\subsubsection{Belief in a Spirit or a God}\label{belief-in-a-spirit-or-a-god}}
-
-We assess belief in a spirit or a God by asking two non-affective questions: \enquote{Do you believe in a spirit or life force} and \enquote{Do you believe in a god?} Responses were coded as (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who believe in a Spirit or Life Force; (3) Believers in a God who disbelieve in a Spirit or Life force; (4) Believers in a God and Spirit or Life Force.
-These believe were developed for the NZAVS from the 2005 Eurobarometer.
-For detailed information pertaining to beliefs, see Table 1.
-
-Table 1 about here:
 
@@
 Age was put into units of 10 years and centered at its mean. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 2.
-Gender.
-Gender was assessed by asking participants if they were \enquote{Male} was coded as \enquote{1} and \enquote{Female} was coded as \enquote{0.} For detailed information pertaining to yearly responses and missingness, see Table 1.
+Gender. Gender was assessed by asking participants if they were \enquote{Male} was coded as \enquote{1} and \enquote{Female} was coded as \enquote{0.}
 
@@
 
+\hypertarget{personal-well-being}{%
+\subsubsection{Personal well-being}\label{personal-well-being}}
+
+Personal well-being is a measure of function all well-being that was assessed using a 4-item version of the Australian Unity well-being Index (Cummins, Eckerseley, Pallant, Van Vugt, \& Misajon, 2017).
+Participants rated their satisfaction with (a) \enquote{Your standard of living}; (b) \enquote{Your health}; (c) \enquote{Your future security}; (d) \enquote{Your personal relationships.} Items were rated using a 10-point response option ranging from 1 = completely dissatisfied to 10 = completely satisfied.
+The mean Cronbach's alpha for this scale was \(\alpha\) = 0.90.
+Higher scores on this scale indicate higher personal well-being.
+Overall scores were means of the 4-items.
+
+\hypertarget{belief-in-a-spiritlife-force-andor-a-god}{%
+\subsubsection{Belief in a Spirit/Life Force and/or a God}\label{belief-in-a-spiritlife-force-andor-a-god}}
+
+We assess belief in a spirit or a God by asking two non-affective questions: \enquote{Do you believe in a spirit or Life Force} and \enquote{Do you believe in a god?} Responses were coded as (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life Force; (2) Skeptics about God who believe in a Spirit or Life Force; (3) Believers in a God who disbelieve in a Spirit or Life Force; (4) Believers in a God and Spirit or Life Force.
+These believe were developed for the NZAVS from the 2005 Eurobarometer.
+
 \hypertarget{statistical-analyses}{%
@@
 Statistical analysis was performed using R version 4.0.2 R version 3.6.1 (2020-06-22) 2019-07-05).
-We analyze data from participants who responded to the NZAVS survey at least three times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of 21,705 (N = 20,979 with complete responses).
+We analyze data from participants who responded to the NZAVS survey at least three times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of 21,705 (\emph{N} = 20,979 with complete responses).
 To model growth in subjective well-being we use generalized linear mixed models.
@@
 The baseline for this indicator is Skepticism about a Spirit/Life Forces and Skepticism about a God.
-To handle the dependencies introduced from the repeated measures, we include individual ID as an effect modeled as random, and estimated a random effect for the slope for years.
-The model equations are given below and priors are given below.
-
-\hypertarget{potential-confounders-and-missing-data}{%
-\subsection{Potential Confounders and Missing Data}\label{potential-confounders-and-missing-data}}
+To handle the dependencies introduced from the repeated measures, we include individual ID as an effect modeled as random, and we estimated a random effect for the slope for years.
+This allowed us to model correllation between individual-level intercept (starting high or low in Personal Well-Being/Life Satisfaction) and stability/growth/decline in the slope effect of years.
+We estimated all statitiscal models using both REML estimation (Bates, Mächler, Bolker, \& Walker, 2015) and Bayesian estimation (Bürkner, 2018) with weakly informative priors.
+We include the full reports in the Supplement.
+The REML estimates and the Bayesian estimates converged to within three decimal places.
+We include model equations and priors at the end of this article.
+Here we report the results of the Bayesian estimator.
+
+\hypertarget{potential-confounders}{%
+\subsection{Potential Confounders}\label{potential-confounders}}
+
+To address confounding we first obtained the New Zealand Attitudes and Values Study standard demographic indicators recommended to address confounding: \href{https://cdn.auckland.ac.nz/assets/psych/about/our-research/nzavs/NZAVSTechnicalDocuments/NZAVS-Technical-Documents-e11-Default-Statistical-Standards.pdf}{NZAVS modelling standards}.
+Next we modelled potential relationships between the coefficients within the model using the daggity package in R, and plotting results using the daggity and ggdag packages in R.
+
+This method reveals two adjustments sets equired to avoid potential confounding: \{age, employment, male, political conservativism, and time\}, and \{age, ethnicty, male, and time \}.Here, we report the second set because employment is likely to contain more measurement error: some are not employed because they cannot find employment, others because they are not seeking employment (e.g.~retirement).
+Results were insensitive to the choice set.
+Indeed, the inference proved robust to the simple model with only the interaction of years and beliefs, and robust to the fully saturated model in which all covariates were included (see the online supplement for details).
+In the fully saturated model, it appeared that those skeptical of a Spirit or Life Force, exhibited no growth in Persona Well-Being and Life-Satisifaction over the nine years.
+Although this result offers stronger stronger evidence for a Spirit/Life Force well-being mechanism, it is likely the effect of collider confounding and masking.
+We discuss this point in more detail in the supplement, and mention it here to caution researchers from \enquote{everything-but-the-kitchen-sink} approaches to regression modelling.
+
+\hypertarget{missing-data}{%
+\subsection{Missing Data}\label{missing-data}}
 
@@
 Additionally the results were robust to filtering of participants of N \textgreater{} 3 to N = 8 waves.
-In a supplemental analysis we include standard demographic indicators used in New Zealand Attitudes and Values Research: \href{https://cdn.auckland.ac.nz/assets/psych/about/our-research/nzavs/NZAVSTechnicalDocuments/NZAVS-Technical-Documents-e11-Default-Statistical-Standards.pdf}{NZAVS modelling standards}.
-We centered and scaled the education, deprivation, and political conservativism.
-We centred age was centred at and put into decade-long intervals.
-There are increasing worries
-
-We provide the analyses of these additional analyses in the supplement. The supplement describing these analysis is located at: \url{https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing}.
+We include the multiple-imputation results in a supplement.
+Here, there were no differences in estimation relevant to inference, however, to improve inference, we advocate wider use of multiple-imputation or Bayesian approaches for handling missingness.
 
@@
 We are greatful to the contributors and mantainers of the following packages:
-R (Version 4.0.2; R Core Team, 2020) and the R-packages \emph{Amelia} (Version 1.7.6; Honaker et al., 2011a), \emph{citr} (Version 0.3.2; Aust, 2019), \emph{dplyr} (Version 1.0.2; Wickham et al., 2020), \emph{equatiomatic} (Version 0.1.0.9000; Anderson \& Heiss, 2020), \emph{forcats} (Version 0.5.0; Wickham, 2020a), \emph{ggeffects} (Version 0.15.0; Lüdecke, 2018), \emph{gghighlight} (Version 0.3.0; Yutani, 2020), \emph{ggplot2} (Version 3.3.2; Wickham, 2016), \emph{ggsci} (Version 2.9; Xiao, 2018), \emph{lme4} (Version 1.1.23; Bates, Mächler, Bolker, \& Walker, 2015), \emph{Matrix} (Version 1.2.18; Bates \& Maechler, 2019), \emph{papaja} (Version 0.1.0.9997; Aust \& Barth, 2020), \emph{parameters} (Version 0.8.6.1; D.
+R (Version 4.0.2; R Core Team, 2020) and the R-packages \emph{Amelia} (Version 1.7.6; Honaker et al., 2011a), \emph{citr} (Version 0.3.2; Aust, 2019), \emph{dplyr} (Version 1.0.2; Wickham et al., 2020), \emph{equatiomatic} (Version 0.1.0.9000; Anderson \& Heiss, 2020), \emph{forcats} (Version 0.5.0; Wickham, 2020a), \emph{ggeffects} (Version 0.15.0; Lüdecke, 2018), \emph{gghighlight} (Version 0.3.0; Yutani, 2020), \emph{ggplot2} (Version 3.3.2; Wickham, 2016), \emph{ggsci} (Version 2.9; Xiao, 2018), \emph{lme4} (Version 1.1.23; Bates et al., 2015), \emph{Matrix} (Version 1.2.18; Bates \& Maechler, 2019), \emph{papaja} (Version 0.1.0.9997; Aust \& Barth, 2020), \emph{parameters} (Version 0.8.6.1; D.
 Lüdecke et al., 2020a), \emph{patchwork} (Version 1.0.1; Pedersen, 2020), \emph{prettycode} (Version 1.1.0; Csárdi, 2019), \emph{purrr} (Version 0.3.4; Henry \& Wickham, 2020), \emph{Rcpp} (Version 1.0.5.1; Eddelbuettel \& François, 2011; Eddelbuettel \& Balamuta, 2017), \emph{readr} (Version 1.3.1; Wickham, Hester, \& Francois, 2018), \emph{report} (Version 0.1.0; Makowski, Dominique, Lüdecke, \& Daniel, 2019), \emph{see} (Version 0.5.2.1; D.
@@
 All code for this analysis and a full report of the missing data imputation can be found at \url{https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing}.
+An html document decribing the same analysis is located at: \url{https://osf.io/cu2gr/}.
 
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{/Users/jbul176/The Virtues Project Dropbox/Joseph Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/coefficientplotsbayes} \caption{Coefficient Graphs}\label{fig:unnamed-chunk-1}
+\includegraphics[width=\textwidth]{/Users/jbul176/The Virtues Project Dropbox/Joseph Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/coefBayesDag} \caption{Coefficient Graphs}\label{fig:unnamed-chunk-1}
 \end{figure}
@@
 \item
-  The effect of years is positive is reliable (posterior median = 0.025, HDI {[}0.020, 0.031{]} pd = 100.00\%).
-We infer that personal well-being increased in the skeptic population between 2010 and 2018.
-\item
-  The effect of beliefs in \emph{Spirit Excludes God} is reliably negative (posterior median = -0.058, HDI {[}-0.080, -0.035{]}, pd = 100.00\%).
+  The intercept, representing the expected personal well-being among skeptics about a Spirit/Life Force and God who are Europeans, 50.77 years of age and do not identify as male is reliable (posterior median = 7.138 HDI {[} 7.111, 7.167{]}, pd = 100.00\%).\\
+\item
+  The effect of years is positive is reliable (posterior median = 7.138, HDI {[} 7.111, 7.167{]}, pd 100.00\%).
+We infer that personal well-being increased in the baseline skeptic population between 2010 and 2018.
+\item
+  The effect of beliefs in \emph{Spirit Excludes God} is reliably negative (posterior median = -0.059, HDI {[}-0.081, -0.036{]}, pd 100.00\%).
 We infer that the population that believed in a Spirit but did not believe in a God was lower in personal well-being at the intercept compared to skeptics.
 \item
-  The effect of beliefs in \emph{God And Spirit} is reliably negative (posterior median = -0.064, HDI {[}-0.091, -0.037{]}, pd = 100.00\%).
+  The effect of beliefs in \emph{God And Spirit} is not negative (posterior median = -0.060, HDI {[}-0.086, -0.034{]}, pd = 100.00\%) .
 We infer that the population that believed in a God and a Spirit or Life Force was lower in personal well-being at the intercept compared to skeptics.
 \item
-  The effect of beliefs in \emph{God Excludes Spirit} is not reliable (posterior median = -0.002 HDI {[}-0.039, 0.033{]}, pd = 53.50\%).
+  The effect of beliefs in \emph{God Excludes Spirit} is not reliable (posterior median = -0.001 HDI {[}-0.036, 0.035{]}, pd = 51.10\%).
 We infer that the population that believed in a God and doubted a Spirit or Life Force were similar were similar in personal well-being at the intercept to skeptics.
+\item
+  The effect of Age in Decades (C) is 0.072 \textbar{} {[} 0.060, 0.083{]} \textbar{} 100.00\% \textbar{}
+\item
+  The effect of Maori Ethnicity is -0.380 \textbar{} {[}-0.420, -0.341{]} \textbar{} 100.00\% \textbar{}
+\item
+  The effect of Pacific Ethnicity is -0.384 \textbar{} {[}-0.464, -0.312{]} \textbar{} 100.00\% \textbar{}
+\item
+  The effect of Asian Ethnicity is -0.093 \textbar{} {[}-0.165, -0.023{]} \textbar{} 98.28\% \textbar{}\\
+\item
+  The effect of Male gender is -0.060 \textbar{} {[}-0.094, -0.026{]} \textbar{} 99.90\% \textbar{}
 \end{itemize}
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{/Users/jbul176/The Virtues Project Dropbox/Joseph Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/Bayesian_pwi_predicted} \caption{Personal Well-being prediction graph}\label{fig:unnamed-chunk-2}
+\includegraphics[width=\textwidth]{/Users/jbul176/The Virtues Project Dropbox/Joseph Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/pwi_confounder_adjusted} \caption{Personal Well-being prediction graph}\label{fig:unnamed-chunk-2}
 \end{figure}
 
-The effect of years \(\times\) belief in a \emph{Spirit Excludes God} is reliably positive (posterior median = 0.010, HDI {[} 0.003, 0.017{]}, pd = 99.02\%).
-We infer that people with beliefs in a spirit or life force who do not believe in a God tended to grow in their personal well-being over time at a faster rate than skeptics grew.
-- The effect of years \(\times\) in \emph{God And Spirit} is reliably positive (posterior median = 0.014, HDI {[}0.008, 0.021{]}, pd = 99.98\%).
-We infer that people who believe in a spirit or life force who also believe in a God tended to grow in their personal well-being over time at a faster rate than skeptics grew.
-- The effect of years \(\times\) belief in \emph{God Excludes Spirit} is not reliable (posterior median = 0.002, HDI {[}-0.009, 0.015{]}, pd = 63.73\%).
-We infer that people who believe in a God but not a spirit or life force did not grow at a faster rate in personal well-being than skeptics grew.
+The effect of years \(\times\) belief in a \emph{Spirit Excludes God} is reliably positive (posterior median = 0.011, HDI {[} 0.003, 0.018{]}, pd = 99.02\%).
+We infer that people with beliefs in a spirit or Life Force who do not believe in a God tended to grow in their personal well-being over time at a faster rate than skeptics grew.
+- The effect of years \(\times\) in \emph{God And Spirit} is reliably positive (posterior median = 0.014, HDI {[}0.007, 0.021{]}, pd = 99.98\%).
+We infer that people who believe in a spirit or Life Force who also believe in a God tended to grow in their personal well-being over time at a faster rate than skeptics grew.
+- The effect of years \(\times\) belief in \emph{God Excludes Spirit} is not reliable (posterior median = 0.004, HDI {[}-0.009, 0.015{]}, pd = 68.73\%).
+We infer that people who believe in a God but not a spirit or Life Force did not grow at a faster rate in personal well-being than skeptics grew.
 
@@
 Within this model:
+- The intercept, representing the expected life satisfaction among skeptics about a Spirit/Life Force and God who are Europeans, 50.77 years of age and do not identify as male is reliable (posterior median = 5.287, HDI {[} 5.267, 5.306{]}, pd = 100.00\%).
+- The effect of years is positive is reliable (posterior median = 0.008, HDI {[} 0.003, 0.011{]}, pd = 99.75\%).
+We infer that life satisfaction increased in the skeptic population between 2010 and 2018.
+-The effect of beliefs in \emph{Spirit Excludes God} is not reliable (posterior median = 0.007, HDI {[}-0.010, 0.023{]}, pd = 75.33\%).
+We infer that the population that believed in a Spirit but did not believe in a God was higher in life satisfaction at the intercept compared with skeptics.
+- The effect of beliefs in \emph{God And Spirit} is reliably positive (posterior median = 0.034, HDI {[} 0.015, 0.053{]}, pd = 99.83\%).
+We infer that the population that believed in a God and a Spirit or Life Force was higher in life satisfaction at the intercept compared with skeptics.
+- The effect of beliefs in \emph{God Excludes Spirit} is reliably positive (posterior median = 0.057, HDI {[} 0.030, 0.082{]}, pd = 99.95\%).
+We infer that the population that believed in a God and doubted a Spirit or Life Force was higher in life satisfaction at the intercept compared with skeptics.
+- The effect of Age in Decades (C) is reliably positive (posterior median = 0.079, hDI {[} 0.071, 0.087{]} pd = 100.00\%).
+- The effect of Maori Ethnicity is reliably negative (posterior median = -0.109 \textbar{} {[}-0.137, -0.082{]}, pd = 100.00\%).
+- The effect of Pacific Ethnicity is reliably negative (posterior median = -0.153 \textbar{} {[}-0.210, -0.103{]}, pd = 100.00\%).
+- The effect of Asian Ethnicity is not reliable (posterior median = -0.047, {[}-0.098, 0.004{]}, pd = 93.08\%).
+- The effect of Male gender is reliabily negative (posterior median = -0.161, {[}-0.184, -0.138{]}, pd = 100.00\%)
+
+\hypertarget{life-satisfaction-years-x-beliefs}{%
+\subsection{Life Satisfaction: years X beliefs}\label{life-satisfaction-years-x-beliefs}}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{/Users/jbul176/The Virtues Project Dropbox/Joseph Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/ls_counfounder_adjusted} \caption{Life satisfaction prediction graph}\label{fig:unnamed-chunk-3}
+\end{figure}
 
@@
 \item
-  The effect of years is positive is reliable (posterior median = 0.014, HDI {[} 0.010, 0.019{]}, pd = 100.00\%).
-We infer that life satisfaction increased in the skeptic population between 2010 and 2018.
-  -The effect of beliefs in \emph{Spirit Excludes God} is reliably positive (posterior median = 0.018, HDI {[} 0.001, 0.035{]}, pd = 96.05\%).
-We infer that the population that believed in a Spirit but did not believe in a God was higher in life satisfaction at the intercept compared with skeptics.
-\item
-  The effect of beliefs in \emph{God And Spirit} is reliably positive (posterior median = 0.051, HDI {[} 0.031, 0.071{]}, pd = 100.00\%).
-We infer that the population that believed in a God and a Spirit or Life Force was higher in life satisfaction at the intercept compared with skeptics.
-\item
-  The effect of beliefs in \emph{God Excludes Spirit} is reliably positive (posterior median = 0.076, HDI {[} 0.048, 0.102{]}, pd = 100.00\%).
-We infer that the population that believed in a God and doubted a Spirit or Life Force was higher in life satisfaction at the intercept compared with skeptics.
-\end{itemize}
-
-\hypertarget{life-satisfaction-years-x-beliefs}{%
-\subsection{Life Satisfaction: years X beliefs}\label{life-satisfaction-years-x-beliefs}}
-
-\begin{figure}
-\includegraphics[width=\textwidth]{/Users/jbul176/The Virtues Project Dropbox/Joseph Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/Bayesian_lifesat_predicted} \caption{Life satisfaction prediction graph}\label{fig:unnamed-chunk-3}
-\end{figure}
+  The interaction of years \(\times\) beliefs in \emph{Spirit Excludes God} is reliably positive (posterior median = 0.011, HDI {[} 0.006, 0.016{]}, pd = 99.95\%).
+We infer that people with beliefs in a Spirit or Life Force who do not believe in a God tended to grow over time in their life satisfaction at a faster rate than skeptics.
 
@@
   \item
-  The interaction of years \(\times\) beliefs in \emph{Spirit Excludes God} is reliably positive (posterior median = 0.012, HDI {[} 0.007, 0.017{]}, pd = 100.00\%).
-We infer that people with beliefs in a Spirit or Life Force who do not believe in a God tended to grow over time in their life satisfaction at a faster rate than skeptics.
-
-  \begin{itemize}
-  \tightlist
-  \item
-    The interaction of years \(\times\) \emph{God And Spirit} is reliably positive (posterior median = 0.010, HDI {[} 0.004, 0.015{]}, pd = 99.90\%).
-We infer that people who believe in a spirit or life force who also believe in a God tended to grow over time in their life satisfaction at a faster rate than skeptics.
-  \item
-    The effect of years \(\times\) beliefs in \emph{God Excludes Spirit} is not reliable (posterior median = -0.002, HDI {[}-0.011, 0.007{]}, pd = 62.80\%).
-We infer that people who believe in a God but not a spirit or life force tended to grow over time in their life satisfaction at a similar rate to skeptics.
+    The interaction of years \(\times\) \emph{God And Spirit} is reliably positive (posterior median = 0.009, HDI {[} 0.004, 0.014{]}, pd = 99.72\%).
+We infer that people who believe in a spirit or Life Force who also believe in a God tended to grow over time in their life satisfaction at a faster rate than skeptics.
+  \item
+    The effect of years \(\times\) beliefs in \emph{God Excludes Spirit} is not reliable (posterior median = -0.003, HDI {[}-0.012, 0.006{]}, pd = 72.28\%).
+We infer that people who believe in a God but not a spirit or Life Force tended to grow over time in their life satisfaction at a similar rate to skeptics.
   \end{itemize}
@@
 \item
-  At the outset of the study, in 2010, the expected subjective well-being as measured using the personal well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God (\(\hat{y} =\) 6.80 {[}6.76, 6.85{]}) and lower among those who believed in both a Spirit/Life-Force and a God (\(\hat{y}\) = 6.78 {[}6.74, 6.81{]}) in comparison both to the skeptical population who neither believed in a God nor a life force (\(\hat{y}\) = 6.92 {[}6.87, 6.96{]}) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force (\(\hat{y}\) 6.90 {[}6.83, 6.98{]}).
-(See Figure 2) Similarly,the expected subjective well-being as measured using the life satisfaction Inventory was lowest in 2010 among those who believed in Spirit or Life Force but not a God (\(\hat{y}\) = 5.07 {[}5.04, 5.10{]}).
-Life satisfaction was similar among those who believed in both a Spirit/Life-Force and a God (\(\hat{y}\) = 5.11 {[}5.08, 5.13{]}) in comparison both with the skeptical population who neither believed in a God nor a life force (\(\hat{y}\) = 5.12 {[}5.08, 5.15{]}).
-Life satisfaction was highest among the population that believed in a God but was skeptical about belief in a Spirit or Life Force (\(\hat{y}\) 5.19 {[}5.14, 5.25{]}) (see Figure 3)
+  At the outset of the study, in 2010, the expected subjective well-being as measured using the personal well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God (\(\hat{y} =\) 6.93 \textbar{} {[}6.88, 6.98{]}) and lower among those who believed in both a Spirit/Life Force and a God (\(\hat{y}\) = 6.91 \textbar{} {[}6.87, 6.95{]}) in comparison both to the skeptical population who neither believed in a God nor Spirit/ Life Force (\(\hat{y}\) = 7.04 \textbar{} {[}6.99, 7.10{]}) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force (\(\hat{y}\) 7.03 \textbar{} {[}6.95, 7.10{]}).
+(See Figure 2) Similarly,the expected subjective well-being as measured using the life satisfaction Inventory was lower in 2010 among those who believed in Spirit or Life Force but not a God (\(\hat{y}\) = 5.20 \textbar{} {[}5.16, 5.23{]}).
+Life satisfaction was similar among those who believed in both a Spirit/Life Force and a God (\(\hat{y}\) = 5.23 \textbar{} {[}5.21, 5.26{]}) in comparison both with the skeptical population who neither believed in a God nor a Life Force (\(\hat{y}\) = 5.25 \textbar{} {[}5.21, 5.29{]}).
+Life satisfaction was highest among the population that believed in a God but was skeptical about belief in a Spirit or Life Force (\(\hat{y}\) 5.32 \textbar{} {[}5.26, 5.38{]}) (see Figure 3)
 \item
   However, over time, belief in a Spirit or Life Force predicts a trajectory of growth in personal well-being.
-In 2018, personal well-being among who believe in Spirit or Life Force but not a God had grown to (\(\hat{y} = 7.12\) {[}7.09, 7.16{]}, \(\Delta \hat{y} = 0.32\)).
-Among those who believed in both a Spirit/Life-Force and a God, personal well-being had grown to (\(\hat{y} = 7.13\) {[}7.10, 7.16{]}, \(\Delta \hat{y} = 0.35\)).
-By contrast, personal well-being among the skeptical population who neither believed in a God nor a Life Force grew less steeply across the eight year interval \(\hat{y} = 7.14\) {[}7.11, 7.18{]}, \(\Delta \hat{y} = 0.22\)).
-By the same token, personal well-being among those who believe in a God but are skeptical about a Spirit/Life Force grew less steeply \(\hat{y} = 7.15\), {[}7.09, 7.21{]} \(\Delta \hat{y} = 0.25\)).
+In 2018, personal well-being among who believe in Spirit or Life Force but not a God had grown to (\(\hat{y} = 7.19\) {[}7.15, 7.23{]}, \(\Delta \hat{y} = 0.26\)).
+Among those who believed in both a Spirit/Life Force and a God, personal well-being had grown to (\(\hat{y} = 7.20\) {[}7.17, 7.24{]}, \(\Delta \hat{y} = 0.29\)).
+By contrast, personal well-being among the skeptical population who neither believed in a God nor a Life Force grew less steeply across the eight year interval \(\hat{y} = 7.21\) {[}7.17, 7.25{]}, \(\Delta \hat{y} = 0.17\)).
+By the same token, personal well-being among those who believe in a God but are skeptical about a Spirit/Life Force grew less steeply \(\hat{y} = 7.22\) {[}7.16, 7.28{]} \(\Delta \hat{y} = 0.19\)).
+In practical terms, the rate of increase in personal wellbeing among those with Spirit/Life Force beliefs was about 30\% greater than it was among those who lacked such Spirit/Life Force (See Figure 2).
 Similarly, over time, belief in a Spirit or Life Force predicts a trajectory of growth in life satisfaction.
-In 2018, the expected effect of spiritual beliefs on life satisfaction among those who believe in Spirit or Life Force but not a God had grown to (\(\hat{y} = 5.31\) {[}5.28, 5.33{]}, \(\Delta \hat{y} = 0.23\)).
-In practical terms, such growth is equivalent to the expected gain in life satisfaction from being employed and a sixth as large as the expected benefit to life satisfaction from having a romantic partner.
-In practical terms, the rate of increase in personal wellbeing among those with Spirit/Life force beliefs was about one and half times greater than it was among those who lacked such beliefs (See Figure 2).
-Among those who believed in both a Spirit/Life-Force and a God, life satisfaction had grown to (\(\hat{y} = 5.33 [5.31, 5.35]\), \(\Delta \hat{y} = 0.24\)).
-By contrast, life satisfaction among the skeptical population who neither believed in a God nor a Life Force lagged behind over the eight year interval \(\hat{y} = 5.24 [5.21, 5.27], \Delta \hat{y} = 0.13\)).
-Relatedly, life satisfaction in the population that believe in God but are skeptical about a Spirit/Life Force grew less steeply \(\hat{y} = 5.31 [5.26, 5.36], \Delta \hat{y} = 0.11\)).
-In practical terms, the rate of increase in life satisfaction among those with Spirit/Life force beliefs was twice as large (See Figure 3).
+In 2018, the expected effect of spiritual beliefs on life satisfaction among those who believe in Spirit or Life Force but not a God had grown to (\(\hat{y} = 5.37\) {[}5.34, 5.40{]}, \(\Delta \hat{y} = 0.17\)).
+Among those who believed in both a Spirit/Life Force and a God, life satisfaction had grown to (\(\hat{y} = 5.39\) {[}5.36, 5.41{]}, \(\Delta \hat{y} = 0.19\)).By contrast, life satisfaction among the skeptical population who neither believed in a God nor a Life Force lagged behind over the eight year interval \(\hat{y} = 5.32\) {[}5.29, 5.35{]}, \(\Delta \hat{y} = 0.07\)).
+Relatedly, life satisfaction in the population that believe in God but are skeptical about a Spirit/Life Force grew less steeply \(\hat{y} = 5.36\) {[}5.31, 5.41{]}, \(\Delta \hat{y} = 0.04\)).
+In practical terms, the rate of increase in life satisfaction among those with Spirit/Life Force beliefs over three times greater than among those that lacked such beliefs (See Figure 3).
 \item
   Inferred growth from spiritual beliefs is evident from 9 annual waves of data collection (8 years), and across a nationally diverse sample.
-On this basis of this longitudinal information accumulated across nationally, we are confident that growth in personal well-being and in life-satisfaction among those who believe in a Spirit and Life force are evident.
-Such effects co-occur in people who lack a belief in a God as well as among people who believe in a God as well as a Spirit or Life force.
+On this basis of this longitudinal information accumulated across nationally, we are confident that growth in personal well-being and in life-satisfaction among those who believe in a Spirit and Life Force are evident.
+Such effects co-occur in people who lack a belief in a God as well as among people who believe in a God as well as a Spirit or Life Force.
 \end{enumerate}
@@
 We have not attempted to model the diverse philosophical, social, and behavioral attributes within people who adhere to spirit beliefs.
-Moreover, people might identify as spiritual without having any accompanying belief in a Spirit or Life force.
+Moreover, people might identify as spiritual without having any accompanying belief in a Spirit or Life Force.
 The task of clarifying how distinct features across a diverse family of \enquote{spirituality} concepts affect perceptions of individual well-being remain intriguing horizons.
@@
 Western populations are increasingly identifying as \enquote{spiritual but not religious} (Swinton, 2001; Zinnbauer \& Pargament, 2005).
-Here, we observe that even among people who lack a traditional religious belief in God, believing in a Spirit or Life force predicts incremental improvements in well-being that sum to practically large effects.
+Here, we observe that even among people who lack a traditional religious belief in God, believing in a Spirit or Life Force predicts incremental improvements in well-being that sum to practically large effects.
 Though the contextual and psychological mechanisms that underpin such growth will require many decades of research, this preliminary finding is good news for those increasing numbers of individuals continue to report spiritual beliefs but who do not affirm traditional religious beliefs.
@@
 \leavevmode\hypertarget{ref-R-equatiomatic}{}%
-Anderson, D., \& Heiss, A. (2020). \emph{Equatiomatic: Transform models into 'latex' equations}. Retrieved from \url{https://github.com/datalorax/equatiomatic}
+Anderson, D., \& Heiss, A. (2020). Equatiomatic: Transform models into 'latex' equations. Retrieved from \url{https://github.com/datalorax/equatiomatic}
 
@@
 \leavevmode\hypertarget{ref-R-citr}{}%
-Aust, F. (2019). \emph{Citr: 'RStudio' add-in to insert markdown citations}. Retrieved from \url{https://github.com/crsh/citr}
+Aust, F. (2019). Citr: 'RStudio' add-in to insert markdown citations. Retrieved from \url{https://github.com/crsh/citr}
 
@@
 \leavevmode\hypertarget{ref-R-Matrix}{}%
-Bates, D., \& Maechler, M. (2019). \emph{Matrix: Sparse and dense matrix classes and methods}. Retrieved from \url{https://CRAN.R-project.org/package=Matrix}
+Bates, D., \& Maechler, M. (2019). Matrix: Sparse and dense matrix classes and methods. Retrieved from \url{https://CRAN.R-project.org/package=Matrix}
 
@@
 \leavevmode\hypertarget{ref-R-prettycode}{}%
-Csárdi, G. (2019). \emph{Prettycode: Pretty print r code in the terminal}. Retrieved from \url{https://CRAN.R-project.org/package=prettycode}
+Csárdi, G. (2019). Prettycode: Pretty print r code in the terminal. Retrieved from \url{https://CRAN.R-project.org/package=prettycode}
 
@@
 \leavevmode\hypertarget{ref-R-purrr}{}%
-Henry, L., \& Wickham, H. (2020). \emph{Purrr: Functional programming tools}. Retrieved from \url{https://CRAN.R-project.org/package=purrr}
+Henry, L., \& Wickham, H. (2020). Purrr: Functional programming tools. Retrieved from \url{https://CRAN.R-project.org/package=purrr}
 
@@
 \leavevmode\hypertarget{ref-R-patchwork}{}%
-Pedersen, T. L. (2020). \emph{Patchwork: The composer of plots}. Retrieved from \url{https://CRAN.R-project.org/package=patchwork}
+Pedersen, T. L. (2020). Patchwork: The composer of plots. Retrieved from \url{https://CRAN.R-project.org/package=patchwork}
 
@@
 \leavevmode\hypertarget{ref-R-forcats}{}%
-Wickham, H. (2020a). \emph{Forcats: Tools for working with categorical variables (factors)}. Retrieved from \url{https://CRAN.R-project.org/package=forcats}
+Wickham, H. (2020a). Forcats: Tools for working with categorical variables (factors). Retrieved from \url{https://CRAN.R-project.org/package=forcats}
 
@@
 \leavevmode\hypertarget{ref-R-dplyr}{}%
-Wickham, H., François, R., Henry, L., \& Müller, K. (2020). \emph{Dplyr: A grammar of data manipulation}. Retrieved from \url{https://CRAN.R-project.org/package=dplyr}
+Wickham, H., François, R., Henry, L., \& Müller, K. (2020). Dplyr: A grammar of data manipulation. Retrieved from \url{https://CRAN.R-project.org/package=dplyr}
 
 \leavevmode\hypertarget{ref-R-readr}{}%
-Wickham, H., Hester, J., \& Francois, R. (2018). \emph{Readr: Read rectangular text data}. Retrieved from \url{https://CRAN.R-project.org/package=readr}
+Wickham, H., Hester, J., \& Francois, R. (2018). Readr: Read rectangular text data. Retrieved from \url{https://CRAN.R-project.org/package=readr}
 
 \leavevmode\hypertarget{ref-R-ggsci}{}%
-Xiao, N. (2018). \emph{Ggsci: Scientific journal and sci-fi themed color palettes for 'ggplot2'}. Retrieved from \url{https://CRAN.R-project.org/package=ggsci}
+Xiao, N. (2018). Ggsci: Scientific journal and sci-fi themed color palettes for 'ggplot2'. Retrieved from \url{https://CRAN.R-project.org/package=ggsci}
 
@@
 \leavevmode\hypertarget{ref-R-gghighlight}{}%
-Yutani, H. (2020). \emph{Gghighlight: Highlight lines and points in 'ggplot2'}. Retrieved from \url{https://CRAN.R-project.org/package=gghighlight}
+Yutani, H. (2020). Gghighlight: Highlight lines and points in 'ggplot2'. Retrieved from \url{https://CRAN.R-project.org/package=gghighlight}
 

### 2020-11-29 Supplement_And_Analysis.Rmd
near to final draft, latex table not right yet

--- 2020-11-27 Supplement_And_Analysis.Rmd
@@
 
-We wrote the publication in $\LaTeX$, which requires converting the tables to .tex files; this conversion used a custom function we wrote in the `funs.R` file.
+<!-- We wrote the publication in $\LaTeX$, which requires converting the tables to .tex files; this conversion used a custom function we wrote in the `funs.R` file.  -->
 
@@
 <!-- # you might need to install the dev verion.//remotes::install_github("tysonstanley/furniture") -->
-<!-- #demographic_table_latex(d_3) # note the furniture package conflicts with the table1 package.  -->
+<!-- demographic_table_latex() # note the furniture package conflicts with the table1 package. -->
+<!-- d_3%>% -->
+<!--   group_by(Wave)%>% -->
+<!-- furniture::table1( -->
+<!--                   Age,  -->
+<!--                   NZdep, -->
+<!--                   Education, -->
+<!--                   Employed, -->
+<!--                   Ethnic_Categories, -->
+<!--                   Male,Has_Partner,  -->
+<!--                   Political_Conservativism, -->
+<!--                   Urban,  -->
+<!--                   Beliefs,  -->
+<!--                   LIFESAT, -->
+<!--                   PWI, -->
+<!--                   overall=F) -->
+
+<!-- dat3<-d_3%>%dplyr::select( Age, NZdep, -->
+<!--                   Education, -->
+<!--                   Employed, -->
+<!--                   Ethnic_Categories, -->
+<!--                   Male,Has_Partner, -->
+<!--                   Political_Conservativism, -->
+<!--                   Urban, -->
+<!--                   Beliefs, -->
+<!--                   LIFESAT, -->
+<!--                   PWI,Wave) -->
+<!-- stargazer::stargazer(dat3) -->
+<!-- library(modelsummary) -->
+<!--   modelsummary::datasummary_df(ALL(Age) ~ Mean + SD + Histogram, data=dat3) -->
 <!-- ``` -->
@@
 
-In the reported analyses, we included covariates to adjust for potential confounding effects from usual suspect factors associated with our two measures of well-being (Personal Well-Being, Life Satisfaction).
 Where $Z$ is a third variable, $X$ is the predictor, and $Y$ is the outcome, there are at least four types of confounding that might arise:
+
 1. $Z$ causes $X$ and  $Z$ causes $Y$: $Z \leftarrow Z \rightarrow Y$
@@
 
-We report our usual-suspect analysis below, which does not change the inference in this study.
-However, the third and fourth type of confounding illustrate how conditioning on a covariates may confound by *opening* a relationship between the $X$ and $Y$ indicators in a regression analysis.
+We do not wish to exclude third variable explanations of the kind that arise from the first type of confounding.
+The second type of counfounding can mask a relationship.
+The third and fourth type of confounding illustrate how conditioning on a covariates may confound by *opening* a relationship between the $X$ and $Y$ indicators in a regression analysis.
 Such confounding is particularly worrying in the case of longitudinal research.
-The second type of counfounding can mask a relationship.
-We do not wish to exclude third variable explanations of the kind that arise from the first type of confounding.
-Indeed, we are careful to state in the main text that our study cannot infer a causal relationship between spirit beliefs and well-being.
-Rather, the contribution of the study comes as an inference about how the expected relationship between beliefs and well-being changes between those who differ in such beliefs over time.
-By contrast, a causal inference requires testing specific mechanistic theory.
-This is a matter for future research.
-Finally, the usual suspect analysis that we report below does not affect inference.
-We find that the relationships that we observe in the simple analysis also hold there.
+
+To address confounding we first obtained the New Zealand Attitudes and Values Study standard demographic indicators recommended to address confounding: [NZAVS modelling standards](https://cdn.auckland.ac.nz/assets/psych/about/our-research/nzavs/NZAVSTechnicalDocuments/NZAVS-Technical-Documents-e11-Default-Statistical-Standards.pdf).
+Next we modelled potential relationships between the coefficients within the model using the daggity package in R, and plotting results using the daggity and ggdag packages in R.
+
+```{r}
+# obtain dag
+tidy_ggdag <- dagitty::dagify(
+  WB ~ belief + years + partner + nzdep + urban + male + pols + empl,
+  WB ~~ partner + years,
+  belief ~ years + age + male + ethn,
+  partner ~ nzdep + age + belief, 
+  nzdep ~ empl,
+  pols ~ age  + empl + ethn + years,
+  empl ~ nzdep + edu + ethn,
+  exposure =  "belief",
+  outcome =   "WB")%>%
+  tidy_dagitty()
+
+# graph adjustment sets
+ggdag::ggdag_adjustment_set(tidy_ggdag, node_size = 14) + 
+  theme(legend.position = "bottom") 
+```
+
+This method reveals two adjustments sets: \{age, employment, male, political conservativism, and time\}, and \{age, ethnicty, male, and time.\}  We report the second set because employment is likely to contain more measurement error: some are not employed because they cannot find employment, others because they are not seeking employment (e.g.
+retirement).
+However we report both model in our frequentist analysis.
+Neither model choice effects the results.
+We also report the simple unadjusted model and the fully saturated model (with potential collider and masking effects).
+
+A word of caution, the fully-saturated model -- which would be typical in a longitudional regression analysis -- attenuates the growth in personal well-being and in life satisfaction among those who are skeptical in a spirit or Life-fore.
+The inference of this study is not sentitive to this difference because the slopes differ for spirit beliefs across all models.
+However, we do not wish to infer that skepticism about a spirit or life force entirely prevents growth in personal well-being or life-satisfaction.
+We advise that researchers do not introduce variables into regression mmodels without systematically.
+evaluating potential confounding.
 
@@
 ```
+
 ```{r pwi_bayesmodel}
@@
 ```
-<!-- ```{r Bayesian_model_table, cache=TRUE, include =FALSE} -->
-<!-- table_model_latex_dual() -->
-<!-- ``` -->
 
@@
 ```
+
 ```{r mixing_ls, cache=TRUE}
@@
 ```{r mixing_sd_pwi, cache=TRUE}
-plot(mod.1_b, pars ="sd_") # requires stronger priors
+plot(mod.1_b, pars ="sd_") #
 ```
@@
 ```{r mixing_ls_pwi, cache=TRUE}
-plot(mod.2_b, pars ="sd_") # requires stronger priors
+plot(mod.2_b, pars ="sd_") # 
 ```
@@
 ```
+
 ```{r Bayesian_lifesat_predicted, cache=TRUE, dependson= mod.2_b}
@@
 
-## ML estimation
+To close the "back door" on potential confounds we create a regression model
+```{r dag}
+library("dagitty")
+library("ggdag")
+
+# theoretically plausible relationships
+tidy_ggdag <- dagify(
+  WB ~ belief + time + age + partner + nzdep + urban + male + pols + empl, # what might affect wellbeing?
+  WB ~~ partner, # bi-causal path
+  belief ~ time + age + male + ethn, # predictors of beliefs
+  partner ~ nzdep + age + belief,  # predictors of having a partner
+  nzdep ~ empl,  # predictor of deprivation
+  pols ~ age  + empl + ethn, # predictor of political orientation
+  empl ~ nzdep + edu + ethn, # predictor of employment
+  exposure =  "belief",
+  outcome =   "WB")%>%
+  tidy_dagitty()
+
+#graph
+ggdag_adjustment_set(tidy_ggdag, node_size = 14) + 
+  theme(legend.position = "bottom") 
+```
+
+# Bayesian counfounder model
+
+```{r setmodelformulaSAT}
+forumula_bayesPWI_DAG <- bf(PWI ~  yearsC * Beliefs +  Age_in_Decades_C + Ethnic_Categories + Male + (1+ yearsC|Id))
+forumula_bayesLIFESAT_DAG <- bf(LIFESAT ~   yearsC * Beliefs +  Age_in_Decades_C + Ethnic_Categories + Male + (1+ yearsC|Id))
+```
+
+```{r pwi_bayesmodelPWI_DAG}
+# PWI
+mod.1_bDAG <-brm(forumula_bayesPWI_DAG, data= d_3, file ="mod.1_bDAG.rds", cores = 8) # note these are very large files
+summary(mod.1_bDAG)
+```
+
+```{r lsbayesmodelLS_DAG}
+# LIFE SAT
+mod.2_bDAG <-brm (forumula_bayesLIFESAT_DAG, data= d_3,cores = 8, file ="mod.2_bDAG.rds")
+###
+tab_model(mod.1_bDAG,mod.2_bDAG)
+# note these are very large files
+```
+
+The posterior prediction plots look reasonable.
+
+```{r posteriorpredictivechecks:pwi_BAYES_CONFOUNDER, cache=TRUE}
+# cannot run on my laptop
+pp_1<-rstanarm::pp_check(mod.1_bDAG) + ylab("Personal well-being")
+pp_2<-rstanarm::pp_check(mod.2_bDAG) + ylab("Life Satisfaction") ## life sat suggests an ordinal model
+pp_1 
+```
+
+```{r posteriorpredictivechecks:ls_BAYES_CONFOUNDER}
+pp_2
+```
+
+```{r Bayesian_model_table, cache=TRUE, include =FALSE}
+table_model_latex_dual(mod.1_bDAG,mod.2_bDAG)
+```
+
+```{r cache =TRUE}
+g1dag1<-plot(parameters::model_parameters(mod.1_bDAG))+ ggtitle("Personal Well-Being Coefficients")
+g2dag1<-plot(parameters::model_parameters(mod.2_bDAG))+ ggtitle("Life Satisfaction Coefficients")
+```
+
+And here's the graph.
+
+```{r plot_coefficientplots_BayesDAG1, cache = TRUE}
+g1dag1/g2dag1
+```
+we used `r cite_r("b-references.bib")`
+
+```{r cache =TRUE}
+g1dag1<-plot(parameters::model_parameters(mod.1dag))+ ggtitle("Personal well-being Coefficients")
+g2dag1<-plot(parameters::model_parameters(mod.2dag))+ ggtitle("Life Satisfaction Coefficients")
+```
+
+And here's the graph.
+
+```{r plot_coefficientplots_LMER_FULL, cache = TRUE}
+g1dag1/g2dag1
+```
+
+Get the predictions
+
+```{r Bayesian_pwi_predictedSAT, cache=TRUE, dependson= mod.1_b}
+get_predictions(mod.1_bDAG)
+```
+
+```{r Bayesian_lifesat_predictedSAT, cache=TRUE, dependson= mod.2_b}
+get_predictions(mod.2_bDAG)
+```
+
+```{r Bayesian_pwi_predictedSAT, cache=TRUE, dependson= mod.1_b}
+graph_predictions(mod.1_bDAG, "Predicted Values of Personal Well-Being: Bayesian Model (Confounder Adjusted)")
+```
+
+```{r Bayesian_lifesat_predictedSAT, cache=TRUE, dependson= mod.2_b}
+graph_predictions(mod.2_bDAG, "Predicted Values of Life Satisfaction: Bayesian Model (Confounder Adjusted)")
+```
+
+# describe the posteriors 
+```{r simple_table, cache =TRUE}
+bayestestR::describe_posterior(mod.1_bDAG, test = c("p_direction"),diagnostic = "ESS")
+```
+
+```{r simple_table_ls, cache =TRUE}
+bayestestR::describe_posterior(mod.2_bDAG, test = c("p_direction"), diagnostic = "ESS")
+```
+
+## Test of the Bayesian saturacted model  for the record
+
+```{r setmodelformulaSAT}
+# note that BRMS automatically centers so we do not need to center years
+forumula_bayesPWI_SATUR <- bf(PWI ~  years * Beliefs +  Age_in_Decades_C + Deprivation_S + Education_S  + Employed + Ethnic_Categories + Male  + Has_Partner  + Political_Conservativism_S + Urban + (1+ years|Id))
+
+forumula_bayesLIFESAT_SATUR <- bf(LIFESAT ~  yearsC * Beliefs +  Age_in_Decades_C + Deprivation_S + Education_S  + Employed + Ethnic_Categories + Male  + Has_Partner  + Political_Conservativism_S + Urban + (1+ years|Id))
+```
+
+```{r pwi_bayesmodelSAT}
+# PWI
+mod.1_bSAT <-brm(forumula_bayesPWI_SATUR, data= d_3, file ="mod.1_bS.rds", cores = 8) # note these are very large files
+summary(mod.1_bSAT)
+```
+
+```{r Bayesian_pwi_predictedSAT, cache=TRUE, dependson= mod.1_b}
+graph_predictions(mod.1_bSAT, "Predicted Values of Personal Well-Being")
+```
+
+```{r Bayesian_lifesat_predictedSAT, cache=TRUE, dependson= mod.2_b}
+graph_predictions(mod.2_b, "Predicted Values of Life Satisfaction")
+```
+
+```{r lsbayesmodelSAT}
+# LIFE SAT
+mod.2_bSAT <-brm (forumula_bayesLIFESAT_SATUR, data= d_3,cores = 8, file ="mod.2_bS.rds")
+summary(mod.2_bSAT)
+###
+tab_model(mod.1_bSAT,mod.2_bSAT)
+# note these are very large files
+```
+
+## REML estimation
 
@@
 
+## Adjustment using DAG
+
+### First confounder model
+
+```{r lmer_model_with_full_covariatesDAG1, cache =TRUE}
+library(lme4)
+pwi_formulaDag1 <- 'PWI ~  yearsC * Beliefs +  Age_in_Decades_C  + Ethnic_Categories + Male + (1 + yearsC|Id)'
+ls_formulaDag1 <- 'LIFESAT ~ yearsC * Beliefs +  Age_in_Decades_C  + Ethnic_Categories + Male  + (1 + yearsC|Id)'
+
+mod.1dag1 <-lmer (pwi_formulaDag1, data= d_3)
+mod.2dag1 <-lmer (ls_formulaDag1, data= d_3)
+```
+
+Here is the table.
+```{r tab_dag1, cache = TRUE, dependson=d_3}
+sjPlot::tab_model(mod.1dag1,mod.2dag1)
+```
+
+We compared complex with simple models, and we found the inference does not change. 
+```{r}
+# identical result
+sjPlot::tab_model(mod.1dag1,mod.2dag1)
+```
+
+```{r Bayesian_lifesat_predict, cache =TRUE}
+graph_predictions(mod.1dag1, "Predicted Values of Life Satisfaction: Dag 1")
+```
+```{r Bayesian_lifesat_predict, cache =TRUE}
+graph_predictions(mod.2dag1, "Predicted Values of Life Satisfaction: Dag 1")
+```
+
+## Second confounder model
+
+```{r lmer_model_with_full_covariates, cache =TRUE}
+library(lme4)
+pwi_formulaDag2 <- 'PWI ~  yearsC * Beliefs +  Age_in_Decades_C  + Employed + Male + Deprivation_S + Political_Conservativism_S + (1 + yearsC|Id)'
+ls_formulaDag2 <- 'LIFESAT ~ yearsC * Beliefs + Employed +  Age_in_Decades_C  + Employed + Male + Deprivation_S + Political_Conservativism_S + (1 + yearsC|Id)'
+
+mod.1dag2 <-lmer (pwi_formulaDag2, data= d_3)
+mod.2dag2 <-lmer (ls_formulaDag2, data= d_3)
+```
+
+Here is the table
+
+```{r tab_dag2, cache = TRUE, dependson=d_3}
+sjPlot::tab_model(mod.1dag2,mod.1dag2)
+```
+
+```{r dag2_plotcreate cache =TRUE}
+g1dag2<-plot(parameters::model_parameters(mod.1dag2))+ ggtitle("Personal well-being Coefficients")
+g2dag2<-plot(parameters::model_parameters(mod.2dag2))+ ggtitle("Life Satisfaction Coefficients")
+```
+
+And here's the graph.
+
+```{r plot_coefficientplots_LMER_FULL_DAG2, cache = TRUE}
+g1dag2/g2dag2 
+```
+
+```{r Bayesian_lifesat_predict, cache =TRUE}
+graph_predictions(mod.1dag2, "Predicted Values of Life Satisfaction: Dag 2")
+```
+
+```{r Bayesian_lifesat_predict, cache =TRUE}
+graph_predictions(mod.2dag2, "Predicted Values of Life Satisfaction: Dag 2")
+```
+
+Note we would prefer the the first counfounder model because there is less measurement error (not employed could mean mean not retired).
+
+Additionally the first counfounder model has fewer terms to estimate.
+
+### ELSE
+For the record, we report the fully saturated "everything but the kitchen sink" model.  We show that throwing all regression coefficients produces a stronger effect for beliefs.
+
 ```{r lmer model centering years}
@@
 ls_formulaC <- 'LIFESAT ~  yearsC * Beliefs +  Age_in_Decades_C + Deprivation_S + Education_S  + Employed + Ethnic_Categories + Male  + Has_Partner  + Political_Conservativism_S + Urban + (1+ yearsC|Id)'
-
+summary(mod.1a)
 mod.1a <-lmer (pwi_formulaC, data= d_3)

### 2020-11-29 funs.R

@@ -165,11 +165,37 @@ get_predictions <- function(x){
 }
 
 
+
+graph_predictions_BD <- function(x,y){ 
+  out <- ggeffects::ggpredict(model = x, terms = c("years [minmax]","Beliefs"), 
+                              ci.lvl = 0.95,
+                              type = "fe",
+                              typical = "mean",
+                              back.transform = TRUE,
+                              ppd = FALSE,
+                              interval = "confidence")
+  plot(out, facets = T) +  gghighlight::gghighlight()  +  theme_blank() + ggtitle(y) # title to be suppled
+}
+
+get_predictions_BD <- function(x){ 
+  out <- ggeffects::ggpredict(model = x, terms = c("years [minmax]","Beliefs"), 
+                              ci.lvl = 0.95,
+                              type = "fe",
+                              typical = "mean",
+                              back.transform = TRUE,
+                              ppd = FALSE,
+                              interval = "confidence")
+  return(out)
+}
+
+
+
+
 # latex model function
 table_model_latex_pwi <- function(x){ # x is a model
   xtract <-texreg::extract(
     x,
-    level = 0.95,
+    level = 0.90,
     include.random = TRUE,
     include.rsquared = F,
     include.nobs = T,
@@ -182,7 +208,7 @@ table_model_latex_pwi <- function(x){ # x is a model
          scalebox = .5,
          #fontsize= "footnotesize",
          label = "tab:REGRESS_PWI",
-         ci.force.level = 0.95, bold = 0.05,
+         ci.force.level = 0.90, bold = 0.05,
          settingstars = 0,
          booktabs = TRUE,
          custom.note ="")
@@ -191,7 +217,7 @@ table_model_latex_pwi <- function(x){ # x is a model
 table_model_latex_ls <- function(x){ # x is a model
   xtract <-texreg::extract(
     x,
-    level = 0.95,
+    level = 0.90,
     include.random = TRUE,
     include.rsquared = F,
     include.nobs = T,
@@ -204,7 +230,7 @@ table_model_latex_ls <- function(x){ # x is a model
          scalebox = .5,
          #fontsize= "footnotesize",
          label = "tab:REGRESS_LS",
-         ci.force.level = 0.95, bold = 0.05,
+         ci.force.level = 0.90, bold = 0.05,
          settingstars = 0,
          booktabs = TRUE,
[…]

### 2020-11-29 libs.R

@@ -28,6 +28,11 @@ library("brms")
 library("rstan")
 library("rstanarm")
 library("tidybayes")
+library("dagitty")
+library("ggdag")
+
+citation("ggdag")
+
 rstan_options(auto_write=TRUE)
 options(mc.cores=parallel::detectCores ())
 #theme_set(theme_few())

### 2020-11-29 Manuscript_Bayesian.Rmd
fixed equation

@@
 & \quad \beta_{1}(\operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\
-&\quad \beta_{2}(\operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ +  \beta_{3}(\operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}})\ +\\
-& \quad \beta_{4}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\ 
-& \quad \beta_{4}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ + \\
-&\quad \beta_{6}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) \\ 
+&\quad \beta_{2}(\operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\  + \\
+& \quad  \beta_{3}(\operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}})\ +\\
+&\quad \beta_{4}(\operatorname{Age in Decades (C)})\  +\\ 
+& \quad  \beta_{5}(\operatorname{Ethnicity}_{\operatorname{Maori}})\ +\\
+&\quad \beta_{6}(\operatorname{Ethnicity}_{\operatorname{Pacific}})\ +  \\ 
+& \quad  \beta_{7}(\operatorname{Ethnicity}_{\operatorname{Asian}})\ +\\
+& \quad \beta_{8}(\operatorname{Male} )+ \\ 
+& \quad \beta_{9}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\ 
+& \quad \beta_{10}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ + \\
+&\quad \beta_{11}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) \\ 
 \end{aligned}

### 2020-11-29 Supplement_And_Analysis.Rmd
fixed equation

@@
 & \quad \beta_{1}(\operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\
-&\quad \beta_{2}(\operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ +  \beta_{3}(\operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}})\ +\\
+&\quad \beta_{2}(\operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\  + \\
+& \quad  \beta_{3}(\operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}})\ +\\
+&\quad \beta_{5}(\operatorname{Age in Decades (C)})\  +\\ 
+& \quad  \beta_{6}(\operatorname{Ethnicity}_{\operatorname{Maori}})\ +\\
+&\quad \beta_{7}(\operatorname{Ethnicity}_{\operatorname{Pacific}})\ +  \\ 
+& \quad  \beta_{8}(\operatorname{Ethnicity}_{\operatorname{Asian}})\ +\\
+& \quad \beta_{4}(\operatorname{Male} )+ \\ 
 & \quad \beta_{4}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\ 
-& \quad \beta_{4}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ + \\
+& \quad \beta_{5}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ + \\
 &\quad \beta_{6}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) \\ 

### 2020-11-30 Manuscript_Bayesian.Rmd
fixed table small edits

@@
 keywords          : "Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being"
-wordcount         : "5181 words in text body, 1191 words in reference section"
+wordcount         : "5161 words in text body, 1191 words in reference section"
 bibliography      : ["b-references.bib"]
@@
 library("here")
+
 ```
@@
 Table 1 about here 
-
-\begin{table}[ ht ] 
-\centering 
-\caption{}\label{}
-\scalebox{0.6}{
-\begin{tabular}{ l c c c c c c c c c }
+\begin{table}[h]
+\begin{center}
+\scalebox{0.45}{
+\begin{tabular}[t]{llllllllllllllllllll}
 \toprule
- &   \multicolumn{ 9 }{c}{ Wave }\\ 
-  & 2010 & 2011 & 2012 & 2013 & 2014 & 2015 & 2016 & 2017 & 2018 \\ 
- & n = \emph{missing} & n = \emph{missing} & n = 8547 & n = 13472 & n = \emph{missing} & n = 11687 & n = 16048 & n = 14174 & n = 14883 \\ 
+\multicolumn{2}{c}{ } & \multicolumn{2}{c}{2010 (N=3820)} & \multicolumn{2}{c}{2011 (N=6153)} & \multicolumn{2}{c}{2012 (N=10385)} & \multicolumn{2}{c}{2013 (N=15836)} & \multicolumn{2}{c}{2014 (N=15026)} & \multicolumn{2}{c}{2015 (N=13665)} & \multicolumn{2}{c}{2016 (N=18071)} & \multicolumn{2}{c}{2017 (N=15779)} & \multicolumn{2}{c}{2018 (N=16501)} \\
+\cmidrule(l{3pt}r{3pt}){3-4} \cmidrule(l{3pt}r{3pt}){5-6} \cmidrule(l{3pt}r{3pt}){7-8} \cmidrule(l{3pt}r{3pt}){9-10} \cmidrule(l{3pt}r{3pt}){11-12} \cmidrule(l{3pt}r{3pt}){13-14} \cmidrule(l{3pt}r{3pt}){15-16} \cmidrule(l{3pt}r{3pt}){17-18} \cmidrule(l{3pt}r{3pt}){19-20}
+  &    & Mean & Std.
+Dev.
+& Mean  & Std.
+Dev.
+& Mean   & Std.
+Dev.
+& Mean    & Std.
+Dev.
+& Mean     & Std.
+Dev.
+& Mean      & Std.
+Dev.
+& Mean       & Std.
+Dev.
+& Mean        & Std.
+Dev.
+& Mean         & Std.
+Dev.
+\\
 \midrule
-Age &   &   &   &   &   &   &   &   &  \\ 
-NZdep &   &   &   &   &   &   &   &   &  \\ 
-Education &   &   &   &   &   &   &   &   &  \\ 
-Employed &   &   &   &   &   &   &   &   &  \\ 
-\hspace{6pt}    0 &  (\%) &  (\%) & 2094 (24.5\%) & 2955 (21.9\%) &  (\%) & 2525 (21.6\%) & 3486 (21.7\%) & 3223 (22.7\%) & 3404 (22.9\%)\\ 
-\hspace{6pt}    1 &  (\%) &  (\%) & 6453 (75.5\%) & 10517 (78.1\%) &  (\%) & 9162 (78.4\%) & 12562 (78.3\%) & 10951 (77.3\%) & 11479 (77.1\%)\\ 
-Ethnic\_Categories &   &   &   &   &   &   &   &   &  \\ 
-\hspace{6pt}    Euro &  (\%) &  (\%) & 6731 (78.8\%) & 11024 (81.8\%) &  (\%) & 9671 (82.8\%) & 13389 (83.4\%) & 11814 (83.3\%) & 12427 (83.5\%)\\ 
-\hspace{6pt}    Maori &  (\%) &  (\%) & 1188 (13.9\%) & 1608 (11.9\%) &  (\%) & 1341 (11.5\%) & 1749 (10.9\%) & 1620 (11.4\%) & 1677 (11.3\%)\\ 
-\hspace{6pt}    Pacific &  (\%) &  (\%) & 284 (3.3\%) & 339 (2.5\%) &  (\%) & 277 (2.4\%) & 331 (2.1\%) & 243 (1.7\%) & 261 (1.8\%)\\ 
-\hspace{6pt}    Asian &  (\%) &  (\%) & 344 (4\%) & 501 (3.7\%) &  (\%) & 398 (3.4\%) & 579 (3.6\%) & 497 (3.5\%) & 518 (3.5\%)\\ 
-Male &   &   &   &   &   &   &   &   &  \\ 
-\hspace{6pt}    0 &  (\%) &  (\%) & 5302 (62\%) & 8457 (62.8\%) &  (\%) & 7325 (62.7\%) & 10148 (63.2\%) & 9011 (63.6\%) & 9433 (63.4\%)\\ 
-\hspace{6pt}    1 &  (\%) &  (\%) & 3245 (38\%) & 5015 (37.2\%) &  (\%) & 4362 (37.3\%) & 5900 (36.8\%) & 5163 (36.4\%) & 5450 (36.6\%)\\ 
-Has\_Partner &   &   &   &   &   &   &   &   &  \\ 
-\hspace{6pt}    0 &  (\%) &  (\%) & 2408 (28.2\%) & 3546 (26.3\%) &  (\%) & 2815 (24.1\%) & 3773 (23.5\%) & 3330 (23.5\%) & 3556 (23.9\%)\\ 
-\hspace{6pt}    1 &  (\%) &  (\%) & 6139 (71.8\%) & 9926 (73.7\%) &  (\%) & 8872 (75.9\%) & 12275 (76.5\%) & 10844 (76.5\%) & 11327 (76.1\%)\\ 
-Political\_Conservativism &   &   &   &   &   &   &   &   &  \\ 
-Urban &   &   &   &   &   &   &   &   &  \\ 
-\hspace{6pt}    Not\_Urban &  (\%) &  (\%) & 2865 (33.5\%) & 4278 (31.8\%) &  (\%) & 3951 (33.8\%) & 5538 (34.5\%) & 2610 (18.4\%) & 2661 (17.9\%)\\ 
-\hspace{6pt}    Urban &  (\%) &  (\%) & 5682 (66.5\%) & 9194 (68.2\%) &  (\%) & 7736 (66.2\%) & 10510 (65.5\%) & 11564 (81.6\%) & 12222 (82.1\%)\\ 
-Beliefs &   &   &   &   &   &   &   &   &  \\ 
-\hspace{6pt}    \_Skeptic\_ &  (\%) &  (\%) & 2048 (24\%) & 3286 (24.4\%) &  (\%) & 2951 (25.3\%) & 4116 (25.6\%) & 3916 (27.6\%) & 4120 (27.7\%)\\ 
-\hspace{6pt}    \_SpiritExcludesGod\_ &  (\%) &  (\%) & 2295 (26.9\%) & 3735 (27.7\%) &  (\%) & 3169 (27.1\%) & 4501 (28\%) & 3918 (27.6\%) & 3970 (26.7\%)\\ 
-\hspace{6pt}    GodAndSpirit &  (\%) &  (\%) & 3707 (43.4\%) & 5499 (40.8\%) &  (\%) & 5119 (43.8\%) & 6453 (40.2\%) & 5464 (38.5\%) & 6210 (41.7\%)\\ 
-\hspace{6pt}    GodExcludesSpirit &  (\%) &  (\%) & 497 (5.8\%) & 952 (7.1\%) &  (\%) & 448 (3.8\%) & 978 (6.1\%) & 876 (6.2\%) & 583 (3.9\%)\\ 
-LIFESAT &   &   &   &   &   &   &   &   &  \\ 
-PWI &   &   &   &   &   &   &   &   &  \\ 
+Age &  & 51.0 & 14.7 & 50.9 & 15.5 & 50.2 & 14.8 & 48.6 & 14.0 & 49.7 & 14.0 & 51.0 & 13.9 & 50.9 & 13.6 & 51.9 & 13.6 & 52.7 & 13.5\\
+NZdep &  & 4.8 & 2.8 & 4.6 & 2.7 & 4.9 & 2.8 & 4.7 & 2.8 & 4.7 & 2.8 & 4.7 & 2.8 & 4.6 & 2.7 & 4.6 & 2.7 & 4.6 & 2.7\\
+Education &  &  &  &  &  & 5.8 & 2.8 & 6.0 & 2.8 & 6.1 & 2.8 & 6.2 & 2.8 & 6.4 & 2.7 & 6.5 & 2.7 & 6.5 & 2.7\\
+Political\_Conservativism &  & 3.9 & 1.3 & 3.7 & 1.4 & 3.6 & 1.3 & 3.6 & 1.3 & 3.6 & 1.3 & 3.6 & 1.3 & 3.6 & 1.4 & 3.6 & 1.4 & 3.6 & 1.4\\
+LIFESAT &  & 5.2 & 1.2 & 5.1 & 1.2 & 5.1 & 1.2 & 5.2 & 1.2 & 5.2 & 1.2 & 5.2 & 1.2 & 5.2 & 1.2 & 5.3 & 1.2 & 5.3 & 1.2\\
+PWI &  & 6.8 & 1.7 & 6.8 & 1.7 & 6.8 & 1.7 & 7.0 & 1.7 & 7.0 & 1.7 & 7.1 & 1.7 & 7.1 & 1.7 & 7.1 & 1.7 & 7.1 & 1.6\\
+\midrule
+ &  & N & \% & N & \% & N & \% & N & \% & N & \% & N & \% & N & \% & N & \% & N & \%\\
+Employed & 0 & 793 & 0.7 & 1276 & 1.1 & 2666 & 2.3 & 3531 & 3.1 & 3262 & 2.8 & 2972 & 2.6 & 4035 & 3.5 & 3655 & 3.2 & 3746 & 3.3\\
+ & 1 & 2977 & 2.6 & 4795 & 4.2 & 7566 & 6.6 & 11991 & 10.4 & 11626 & 10.1 & 10378 & 9.0 & 14016 & 12.2 & 12070 & 10.5 & 12250 & 10.6\\
+Ethnic\_Categories & Euro & 2947 & 2.6 & 4375 & 3.8 & 7881 & 6.8 & 12270 & 10.6 & 12051 & 10.5 & 10959 & 9.5 & 14641 & 12.7 & 12906 & 11.2 & 13475 & 11.7\\
+ & Maori & 556 & 0.5 & 638 & 0.6 & 1489 & 1.3 & 1897 & 1.6 & 1830 & 1.6 & 1608 & 1.4 & 1964 & 1.7 & 1826 & 1.6 & 1872 & 1.6\\
+ & Pacific & 110 & 0.1 & 126 & 0.1 & 343 & 0.3 & 408 & 0.4 & 400 & 0.3 & 336 & 0.3 & 385 & 0.3 & 286 & 0.2 & 296 & 0.3\\
+ & Asian & 124 & 0.1 & 190 & 0.2 & 409 & 0.4 & 579 & 0.5 & 572 & 0.5 & 473 & 0.4 & 664 & 0.6 & 574 & 0.5 & 579 & 0.5\\
+Male & 0 & 2390 & 2.1 & 3862 & 3.4 & 6542 & 5.7 & 10033 & 8.7 & 9500 & 8.2 & 8548 & 7.4 & 11377 & 9.9 & 9949 & 8.6 & 10388 & 9.0\\
+ & 1 & 1430 & 1.2 & 2291 & 2.0 & 3843 & 3.3 & 5803 & 5.0 & 5478 & 4.8 & 5072 & 4.4 & 6635 & 5.8 & 5776 & 5.0 & 6077 & 5.3\\
+Has\_Partner & 0 & 1054 & 0.9 & 1806 & 1.6 & 2988 & 2.6 & 4237 & 3.7 & 3664 & 3.2 & 3407 & 3.0 & 4274 & 3.7 & 3775 & 3.3 & 3931 & 3.4\\
+ & 1 & 2764 & 2.4 & 4325 & 3.8 & 7331 & 6.4 & 11493 & 10.0 & 11130 & 9.7 & 10176 & 8.8 & 13613 & 11.8 & 11964 & 10.4 & 12078 & 10.5\\
+Urban & Not\_Urban & 1531 & 1.3 & 2010 & 1.7 & 3512 & 3.0 & 5116 & 4.4 & 4894 & 4.2 & 4541 & 3.9 & 6223 & 5.4 & 2873 & 2.5 & 2938 & 2.5\\
+ & Urban & 2271 & 2.0 & 3695 & 3.2 & 6709 & 5.8 & 10602 & 9.2 & 9937 & 8.6 & 8815 & 7.6 & 11622 & 10.1 & 12687 & 11.0 & 13291 & 11.5\\
+Beliefs & \_Skeptic\_ & 625 & 0.5 & 1312 & 1.1 & 2235 & 1.9 & 3639 & 3.2 & 0 & 0.0 & 3278 & 2.8 & 4420 & 3.8 & 4135 & 3.6 & 4514 & 3.9\\
+ & \_SpiritExcludesGod\_ & 778 & 0.7 & 1508 & 1.3 & 2535 & 2.2 & 4206 & 3.6 & 0 & 0.0 & 3542 & 3.1 & 4808 & 4.2 & 4148 & 3.6 & 4309 & 3.7\\
+ & GodAndSpirit & 1751 & 1.5 & 2560 & 2.2 & 4181 & 3.6 & 6201 & 5.4 & 0 & 0.0 & 5780 & 5.0 & 7002 & 6.1 & 5771 & 5.0 & 6782 & 5.9\\
+ & GodExcludesSpirit & 228 & 0.2 & 274 & 0.2 & 593 & 0.5 & 1137 & 1.0 & 0 & 0.0 & 523 & 0.5 & 1094 & 0.9 & 950 & 0.8 & 666 & 0.6\\
 \bottomrule
-
 \end{tabular}
 }
+\caption{Sample statistics by year}
+\end{center}
 \end{table}
@@
 We are greatful to the contributors and mantainers of the following packages:
-R [Version 4.0.2; @R-base] and the R-packages *Amelia* [Version 1.7.6; @R-Amelia], *citr* [Version 0.3.2; @R-citr], *dplyr* [Version 1.0.2; @R-dplyr], *equatiomatic* [Version 0.1.0.9000; @R-equatiomatic], *forcats* [Version 0.5.0; @R-forcats], *ggeffects* [Version 0.15.0; @R-ggeffects], *gghighlight* [Version 0.3.0; @R-gghighlight], *ggplot2* [Version 3.3.2; @R-ggplot2], *ggsci* [Version 2.9; @R-ggsci], *lme4* [Version 1.1.23; @R-lme4], *Matrix* [Version 1.2.18; @R-Matrix], *papaja* [Version 0.1.0.9997; @R-papaja], *parameters* [Version 0.8.6.1; @R-parameters], *patchwork* [Version 1.0.1; @R-patchwork], *prettycode* [Version 1.1.0; @R-prettycode], *purrr* [Version 0.3.4; @R-purrr], *Rcpp* [Version 1.0.5.1; @R-Rcpp_a; @R-Rcpp_b], *readr* [Version 1.3.1; @R-readr], *report* [Version 0.1.0; @R-report], *see* [Version 0.5.2.1; @R-see], *sjPlot* [Version 2.8.4; @R-sjPlot], *stringr* [Version 1.4.0; @R-stringr], *styler* [Version 1.3.2; @R-styler], *table1* [Version 1.2; @R-table1], *texreg* [Version 1.37.5; @R-texreg], *tibble* [Version 3.0.4; @R-tibble], *tidyr* [Version 1.1.2; @R-tidyr], and *tidyverse* [Version 1.3.0; @R-tidyverse]"
+R [Version 4.0.3; @R-base] and the R-packages *Amelia* [Version 1.7.6; @R-Amelia], *citr* [Version 0.3.2; @R-citr], *dplyr* [Version 1.0.2; @R-dplyr], *equatiomatic* [Version 0.1.0.9000; @R-equatiomatic], *forcats* [Version 0.5.0; @R-forcats], *ggeffects* [Version 1.0.0; @R-ggeffects], *gghighlight* [Version 0.3.0; @R-gghighlight], *ggplot2* [Version 3.3.2; @R-ggplot2], *ggsci* [Version 2.9; @R-ggsci], *lme4* [Version 1.1.25; @R-lme4], *Matrix* [Version 1.2.18; @R-Matrix], *papaja* [Version 0.1.0.9997; @R-papaja], *parameters* [Version 0.9.0.1; @R-parameters], *patchwork* [Version 1.1.0; @R-patchwork], *prettycode* [@R-prettycode], *purrr* [Version 0.3.4; @R-purrr], *Rcpp* [Version 1.0.5; @R-Rcpp_a; @R-Rcpp_b], *readr* [Version 1.4.0; @R-readr], *report* [Version 0.2.0; @R-report], *see* [Version 0.6.0.1; @R-see], *sjPlot* [Version 2.8.6; @R-sjPlot], *stringr* [Version 1.4.0; @R-stringr], *styler* [Version 1.3.2; @R-styler], *table1* [Version 1.2.1; @R-table1], *texreg* [Version 1.37.5; @R-texreg], *tibble* [Version 3.0.4; @R-tibble], *tidyr* [Version 1.1.2; @R-tidyr], and *tidyverse* [Version 1.3.0; @R-tidyverse]
 
@@
 \begin{center}
-\scalebox{0.5}{
+\scalebox{0.6}{
 \begin{tabular}{l c c}
@@
 ```{r fig.cap = "Coefficient Graphs", echo =FALSE,out.width="\\textwidth"}
-knitr::include_graphics("/Users/jbul176/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/coefBayesDag.png")
+knitr::include_graphics("/Users/jbul716/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/coefBayesDag.png")
 ```
@@
 ```{r fig.cap = "Personal Well-being prediction graph", echo =FALSE, out.width="\\textwidth"}
-knitr::include_graphics("/Users/jbul176/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/pwi_confounder_adjusted.png")
+knitr::include_graphics("/Users/jbul716/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/pwi_confounder_adjusted.png")
 ```
@@
 ```{r  fig.cap = "Life satisfaction prediction graph", echo = FALSE,out.width="\\textwidth"}
-knitr::include_graphics("/Users/jbul176/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/ls_counfounder_adjusted.png")
+knitr::include_graphics("/Users/jbul716/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/ls_counfounder_adjusted.png")
 ```

### 2020-11-30 Manuscript_Bayesian.tex
fixed table small edits

@@
 \shorttitle{Longitudinal Wellbeing from Spiritual Beliefs}
-\keywords{Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being\newline\indent Word count: 5181 words in text body, 1191 words in reference section}
+\keywords{Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being\newline\indent Word count: 5161 words in text body, 1191 words in reference section}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 
-\begin{table}[ ht ] 
-\centering 
-\caption{}\label{}
-\scalebox{0.6}{
-\begin{tabular}{ l c c c c c c c c c }
+\begin{table}[h]
+\begin{center}
+\scalebox{0.45}{
+\begin{tabular}[t]{llllllllllllllllllll}
 \toprule
- &   \multicolumn{ 9 }{c}{ Wave }\\ 
-  & 2010 & 2011 & 2012 & 2013 & 2014 & 2015 & 2016 & 2017 & 2018 \\ 
- & n = \emph{missing} & n = \emph{missing} & n = 8547 & n = 13472 & n = \emph{missing} & n = 11687 & n = 16048 & n = 14174 & n = 14883 \\ 
+\multicolumn{2}{c}{ } & \multicolumn{2}{c}{2010 (N=3820)} & \multicolumn{2}{c}{2011 (N=6153)} & \multicolumn{2}{c}{2012 (N=10385)} & \multicolumn{2}{c}{2013 (N=15836)} & \multicolumn{2}{c}{2014 (N=15026)} & \multicolumn{2}{c}{2015 (N=13665)} & \multicolumn{2}{c}{2016 (N=18071)} & \multicolumn{2}{c}{2017 (N=15779)} & \multicolumn{2}{c}{2018 (N=16501)} \\
+\cmidrule(l{3pt}r{3pt}){3-4} \cmidrule(l{3pt}r{3pt}){5-6} \cmidrule(l{3pt}r{3pt}){7-8} \cmidrule(l{3pt}r{3pt}){9-10} \cmidrule(l{3pt}r{3pt}){11-12} \cmidrule(l{3pt}r{3pt}){13-14} \cmidrule(l{3pt}r{3pt}){15-16} \cmidrule(l{3pt}r{3pt}){17-18} \cmidrule(l{3pt}r{3pt}){19-20}
+  &    & Mean & Std.
+Dev.
+& Mean  & Std.
+Dev.
+& Mean   & Std.
+Dev.
+& Mean    & Std.
+Dev.
+& Mean     & Std.
+Dev.
+& Mean      & Std.
+Dev.
+& Mean       & Std.
+Dev.
+& Mean        & Std.
+Dev.
+& Mean         & Std.
+Dev.
+\\
 \midrule
-Age &   &   &   &   &   &   &   &   &  \\ 
-NZdep &   &   &   &   &   &   &   &   &  \\ 
-Education &   &   &   &   &   &   &   &   &  \\ 
-Employed &   &   &   &   &   &   &   &   &  \\ 
-\hspace{6pt}    0 &  (\%) &  (\%) & 2094 (24.5\%) & 2955 (21.9\%) &  (\%) & 2525 (21.6\%) & 3486 (21.7\%) & 3223 (22.7\%) & 3404 (22.9\%)\\ 
-\hspace{6pt}    1 &  (\%) &  (\%) & 6453 (75.5\%) & 10517 (78.1\%) &  (\%) & 9162 (78.4\%) & 12562 (78.3\%) & 10951 (77.3\%) & 11479 (77.1\%)\\ 
-Ethnic\_Categories &   &   &   &   &   &   &   &   &  \\ 
-\hspace{6pt}    Euro &  (\%) &  (\%) & 6731 (78.8\%) & 11024 (81.8\%) &  (\%) & 9671 (82.8\%) & 13389 (83.4\%) & 11814 (83.3\%) & 12427 (83.5\%)\\ 
-\hspace{6pt}    Maori &  (\%) &  (\%) & 1188 (13.9\%) & 1608 (11.9\%) &  (\%) & 1341 (11.5\%) & 1749 (10.9\%) & 1620 (11.4\%) & 1677 (11.3\%)\\ 
-\hspace{6pt}    Pacific &  (\%) &  (\%) & 284 (3.3\%) & 339 (2.5\%) &  (\%) & 277 (2.4\%) & 331 (2.1\%) & 243 (1.7\%) & 261 (1.8\%)\\ 
-\hspace{6pt}    Asian &  (\%) &  (\%) & 344 (4\%) & 501 (3.7\%) &  (\%) & 398 (3.4\%) & 579 (3.6\%) & 497 (3.5\%) & 518 (3.5\%)\\ 
-Male &   &   &   &   &   &   &   &   &  \\ 
-\hspace{6pt}    0 &  (\%) &  (\%) & 5302 (62\%) & 8457 (62.8\%) &  (\%) & 7325 (62.7\%) & 10148 (63.2\%) & 9011 (63.6\%) & 9433 (63.4\%)\\ 
-\hspace{6pt}    1 &  (\%) &  (\%) & 3245 (38\%) & 5015 (37.2\%) &  (\%) & 4362 (37.3\%) & 5900 (36.8\%) & 5163 (36.4\%) & 5450 (36.6\%)\\ 
-Has\_Partner &   &   &   &   &   &   &   &   &  \\ 
-\hspace{6pt}    0 &  (\%) &  (\%) & 2408 (28.2\%) & 3546 (26.3\%) &  (\%) & 2815 (24.1\%) & 3773 (23.5\%) & 3330 (23.5\%) & 3556 (23.9\%)\\ 
-\hspace{6pt}    1 &  (\%) &  (\%) & 6139 (71.8\%) & 9926 (73.7\%) &  (\%) & 8872 (75.9\%) & 12275 (76.5\%) & 10844 (76.5\%) & 11327 (76.1\%)\\ 
-Political\_Conservativism &   &   &   &   &   &   &   &   &  \\ 
-Urban &   &   &   &   &   &   &   &   &  \\ 
-\hspace{6pt}    Not\_Urban &  (\%) &  (\%) & 2865 (33.5\%) & 4278 (31.8\%) &  (\%) & 3951 (33.8\%) & 5538 (34.5\%) & 2610 (18.4\%) & 2661 (17.9\%)\\ 
-\hspace{6pt}    Urban &  (\%) &  (\%) & 5682 (66.5\%) & 9194 (68.2\%) &  (\%) & 7736 (66.2\%) & 10510 (65.5\%) & 11564 (81.6\%) & 12222 (82.1\%)\\ 
-Beliefs &   &   &   &   &   &   &   &   &  \\ 
-\hspace{6pt}    \_Skeptic\_ &  (\%) &  (\%) & 2048 (24\%) & 3286 (24.4\%) &  (\%) & 2951 (25.3\%) & 4116 (25.6\%) & 3916 (27.6\%) & 4120 (27.7\%)\\ 
-\hspace{6pt}    \_SpiritExcludesGod\_ &  (\%) &  (\%) & 2295 (26.9\%) & 3735 (27.7\%) &  (\%) & 3169 (27.1\%) & 4501 (28\%) & 3918 (27.6\%) & 3970 (26.7\%)\\ 
-\hspace{6pt}    GodAndSpirit &  (\%) &  (\%) & 3707 (43.4\%) & 5499 (40.8\%) &  (\%) & 5119 (43.8\%) & 6453 (40.2\%) & 5464 (38.5\%) & 6210 (41.7\%)\\ 
-\hspace{6pt}    GodExcludesSpirit &  (\%) &  (\%) & 497 (5.8\%) & 952 (7.1\%) &  (\%) & 448 (3.8\%) & 978 (6.1\%) & 876 (6.2\%) & 583 (3.9\%)\\ 
-LIFESAT &   &   &   &   &   &   &   &   &  \\ 
-PWI &   &   &   &   &   &   &   &   &  \\ 
+Age &  & 51.0 & 14.7 & 50.9 & 15.5 & 50.2 & 14.8 & 48.6 & 14.0 & 49.7 & 14.0 & 51.0 & 13.9 & 50.9 & 13.6 & 51.9 & 13.6 & 52.7 & 13.5\\
+NZdep &  & 4.8 & 2.8 & 4.6 & 2.7 & 4.9 & 2.8 & 4.7 & 2.8 & 4.7 & 2.8 & 4.7 & 2.8 & 4.6 & 2.7 & 4.6 & 2.7 & 4.6 & 2.7\\
+Education &  &  &  &  &  & 5.8 & 2.8 & 6.0 & 2.8 & 6.1 & 2.8 & 6.2 & 2.8 & 6.4 & 2.7 & 6.5 & 2.7 & 6.5 & 2.7\\
+Political\_Conservativism &  & 3.9 & 1.3 & 3.7 & 1.4 & 3.6 & 1.3 & 3.6 & 1.3 & 3.6 & 1.3 & 3.6 & 1.3 & 3.6 & 1.4 & 3.6 & 1.4 & 3.6 & 1.4\\
+LIFESAT &  & 5.2 & 1.2 & 5.1 & 1.2 & 5.1 & 1.2 & 5.2 & 1.2 & 5.2 & 1.2 & 5.2 & 1.2 & 5.2 & 1.2 & 5.3 & 1.2 & 5.3 & 1.2\\
+PWI &  & 6.8 & 1.7 & 6.8 & 1.7 & 6.8 & 1.7 & 7.0 & 1.7 & 7.0 & 1.7 & 7.1 & 1.7 & 7.1 & 1.7 & 7.1 & 1.7 & 7.1 & 1.6\\
+\midrule
+ &  & N & \% & N & \% & N & \% & N & \% & N & \% & N & \% & N & \% & N & \% & N & \%\\
+Employed & 0 & 793 & 0.7 & 1276 & 1.1 & 2666 & 2.3 & 3531 & 3.1 & 3262 & 2.8 & 2972 & 2.6 & 4035 & 3.5 & 3655 & 3.2 & 3746 & 3.3\\
+ & 1 & 2977 & 2.6 & 4795 & 4.2 & 7566 & 6.6 & 11991 & 10.4 & 11626 & 10.1 & 10378 & 9.0 & 14016 & 12.2 & 12070 & 10.5 & 12250 & 10.6\\
+Ethnic\_Categories & Euro & 2947 & 2.6 & 4375 & 3.8 & 7881 & 6.8 & 12270 & 10.6 & 12051 & 10.5 & 10959 & 9.5 & 14641 & 12.7 & 12906 & 11.2 & 13475 & 11.7\\
+ & Maori & 556 & 0.5 & 638 & 0.6 & 1489 & 1.3 & 1897 & 1.6 & 1830 & 1.6 & 1608 & 1.4 & 1964 & 1.7 & 1826 & 1.6 & 1872 & 1.6\\
+ & Pacific & 110 & 0.1 & 126 & 0.1 & 343 & 0.3 & 408 & 0.4 & 400 & 0.3 & 336 & 0.3 & 385 & 0.3 & 286 & 0.2 & 296 & 0.3\\
+ & Asian & 124 & 0.1 & 190 & 0.2 & 409 & 0.4 & 579 & 0.5 & 572 & 0.5 & 473 & 0.4 & 664 & 0.6 & 574 & 0.5 & 579 & 0.5\\
+Male & 0 & 2390 & 2.1 & 3862 & 3.4 & 6542 & 5.7 & 10033 & 8.7 & 9500 & 8.2 & 8548 & 7.4 & 11377 & 9.9 & 9949 & 8.6 & 10388 & 9.0\\
+ & 1 & 1430 & 1.2 & 2291 & 2.0 & 3843 & 3.3 & 5803 & 5.0 & 5478 & 4.8 & 5072 & 4.4 & 6635 & 5.8 & 5776 & 5.0 & 6077 & 5.3\\
+Has\_Partner & 0 & 1054 & 0.9 & 1806 & 1.6 & 2988 & 2.6 & 4237 & 3.7 & 3664 & 3.2 & 3407 & 3.0 & 4274 & 3.7 & 3775 & 3.3 & 3931 & 3.4\\
+ & 1 & 2764 & 2.4 & 4325 & 3.8 & 7331 & 6.4 & 11493 & 10.0 & 11130 & 9.7 & 10176 & 8.8 & 13613 & 11.8 & 11964 & 10.4 & 12078 & 10.5\\
+Urban & Not\_Urban & 1531 & 1.3 & 2010 & 1.7 & 3512 & 3.0 & 5116 & 4.4 & 4894 & 4.2 & 4541 & 3.9 & 6223 & 5.4 & 2873 & 2.5 & 2938 & 2.5\\
+ & Urban & 2271 & 2.0 & 3695 & 3.2 & 6709 & 5.8 & 10602 & 9.2 & 9937 & 8.6 & 8815 & 7.6 & 11622 & 10.1 & 12687 & 11.0 & 13291 & 11.5\\
+Beliefs & \_Skeptic\_ & 625 & 0.5 & 1312 & 1.1 & 2235 & 1.9 & 3639 & 3.2 & 0 & 0.0 & 3278 & 2.8 & 4420 & 3.8 & 4135 & 3.6 & 4514 & 3.9\\
+ & \_SpiritExcludesGod\_ & 778 & 0.7 & 1508 & 1.3 & 2535 & 2.2 & 4206 & 3.6 & 0 & 0.0 & 3542 & 3.1 & 4808 & 4.2 & 4148 & 3.6 & 4309 & 3.7\\
+ & GodAndSpirit & 1751 & 1.5 & 2560 & 2.2 & 4181 & 3.6 & 6201 & 5.4 & 0 & 0.0 & 5780 & 5.0 & 7002 & 6.1 & 5771 & 5.0 & 6782 & 5.9\\
+ & GodExcludesSpirit & 228 & 0.2 & 274 & 0.2 & 593 & 0.5 & 1137 & 1.0 & 0 & 0.0 & 523 & 0.5 & 1094 & 0.9 & 950 & 0.8 & 666 & 0.6\\
 \bottomrule
-
 \end{tabular}
 }
+\caption{Sample statistics by year}
+\end{center}
 \end{table}
@@
 We are greatful to the contributors and mantainers of the following packages:
-R (Version 4.0.2; R Core Team, 2020) and the R-packages \emph{Amelia} (Version 1.7.6; Honaker et al., 2011a), \emph{citr} (Version 0.3.2; Aust, 2019), \emph{dplyr} (Version 1.0.2; Wickham et al., 2020), \emph{equatiomatic} (Version 0.1.0.9000; Anderson \& Heiss, 2020), \emph{forcats} (Version 0.5.0; Wickham, 2020a), \emph{ggeffects} (Version 0.15.0; Lüdecke, 2018), \emph{gghighlight} (Version 0.3.0; Yutani, 2020), \emph{ggplot2} (Version 3.3.2; Wickham, 2016), \emph{ggsci} (Version 2.9; Xiao, 2018), \emph{lme4} (Version 1.1.23; Bates et al., 2015), \emph{Matrix} (Version 1.2.18; Bates \& Maechler, 2019), \emph{papaja} (Version 0.1.0.9997; Aust \& Barth, 2020), \emph{parameters} (Version 0.8.6.1; D.
-Lüdecke et al., 2020a), \emph{patchwork} (Version 1.0.1; Pedersen, 2020), \emph{prettycode} (Version 1.1.0; Csárdi, 2019), \emph{purrr} (Version 0.3.4; Henry \& Wickham, 2020), \emph{Rcpp} (Version 1.0.5.1; Eddelbuettel \& François, 2011; Eddelbuettel \& Balamuta, 2017), \emph{readr} (Version 1.3.1; Wickham, Hester, \& Francois, 2018), \emph{report} (Version 0.1.0; Makowski, Dominique, Lüdecke, \& Daniel, 2019), \emph{see} (Version 0.5.2.1; D.
-Lüdecke et al., 2020b), \emph{sjPlot} (Version 2.8.4; D.
-Lüdecke, 2020), \emph{stringr} (Version 1.4.0; Wickham, 2019), \emph{styler} (Version 1.3.2; Müller \& Walthert, 2020), \emph{table1} (Version 1.2; Rich, 2020), \emph{texreg} (Version 1.37.5; Leifeld, 2013), \emph{tibble} (Version 3.0.4; Müller \& Wickham, 2020), \emph{tidyr} (Version 1.1.2; Wickham, 2020b), and \emph{tidyverse} (Version 1.3.0; Wickham, Averick, et al., 2019)"
+R (Version 4.0.3; R Core Team, 2020) and the R-packages \emph{Amelia} (Version 1.7.6; Honaker et al., 2011a), \emph{citr} (Version 0.3.2; Aust, 2019), \emph{dplyr} (Version 1.0.2; Wickham et al., 2020), \emph{equatiomatic} (Version 0.1.0.9000; Anderson \& Heiss, 2020), \emph{forcats} (Version 0.5.0; Wickham, 2020a), \emph{ggeffects} (Version 1.0.0; Lüdecke, 2018), \emph{gghighlight} (Version 0.3.0; Yutani, 2020), \emph{ggplot2} (Version 3.3.2; Wickham, 2016), \emph{ggsci} (Version 2.9; Xiao, 2018), \emph{lme4} (Version 1.1.25; Bates et al., 2015), \emph{Matrix} (Version 1.2.18; Bates \& Maechler, 2019), \emph{papaja} (Version 0.1.0.9997; Aust \& Barth, 2020), \emph{parameters} (Version 0.9.0.1; D.
+Lüdecke et al., 2020a), \emph{patchwork} (Version 1.1.0; Pedersen, 2020), \emph{prettycode} (Csárdi, 2019), \emph{purrr} (Version 0.3.4; Henry \& Wickham, 2020), \emph{Rcpp} (Version 1.0.5; Eddelbuettel \& François, 2011; Eddelbuettel \& Balamuta, 2017), \emph{readr} (Version 1.4.0; Wickham, Hester, \& Francois, 2018), \emph{report} (Version 0.2.0; Makowski, Dominique, Lüdecke, \& Daniel, 2019), \emph{see} (Version 0.6.0.1; D.
+Lüdecke et al., 2020b), \emph{sjPlot} (Version 2.8.6; D.
+Lüdecke, 2020), \emph{stringr} (Version 1.4.0; Wickham, 2019), \emph{styler} (Version 1.3.2; Müller \& Walthert, 2020), \emph{table1} (Version 1.2.1; Rich, 2020), \emph{texreg} (Version 1.37.5; Leifeld, 2013), \emph{tibble} (Version 3.0.4; Müller \& Wickham, 2020), \emph{tidyr} (Version 1.1.2; Wickham, 2020b), and \emph{tidyverse} (Version 1.3.0; Wickham, Averick, et al., 2019)
 
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{/Users/jbul176/The Virtues Project Dropbox/Joseph Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/coefBayesDag} \caption{Coefficient Graphs}\label{fig:unnamed-chunk-1}
+\includegraphics[width=\textwidth]{/Users/jbul716/The Virtues Project Dropbox/Joseph Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/coefBayesDag} \caption{Coefficient Graphs}\label{fig:unnamed-chunk-1}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{/Users/jbul176/The Virtues Project Dropbox/Joseph Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/pwi_confounder_adjusted} \caption{Personal Well-being prediction graph}\label{fig:unnamed-chunk-2}
+\includegraphics[width=\textwidth]{/Users/jbul716/The Virtues Project Dropbox/Joseph Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/pwi_confounder_adjusted} \caption{Personal Well-being prediction graph}\label{fig:unnamed-chunk-2}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{/Users/jbul176/The Virtues Project Dropbox/Joseph Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/ls_counfounder_adjusted} \caption{Life satisfaction prediction graph}\label{fig:unnamed-chunk-3}
+\includegraphics[width=\textwidth]{/Users/jbul716/The Virtues Project Dropbox/Joseph Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/ls_counfounder_adjusted} \caption{Life satisfaction prediction graph}\label{fig:unnamed-chunk-3}
 \end{figure}

### 2020-11-30 Supplement_And_Analysis.Rmd
fixed table small edits

@@
 output            : "html_document"
+
 editor_options: 
@@
 ```{r setup, include = FALSE}
-# library("papaja")
-#r_refs("r-references.bib")
+library("papaja")
+r_refs("b-references.bib")
 ```
@@
 range(d_3$years)
+# html talbe
+demographic_table(d_3)
+```
+
+Sample details are as follows:
+```{r sample details}
+sample_report(d_3)
+```
+
+<!-- ```{r  badlatextable} badd tables -->
+<!-- library("arsenal") -->
+<!-- summary(arsenal::tableby(Wave ~ Age + -->
+<!--                    NZdep + -->
+<!--                    Education + -->
+<!--                    Employed + -->
+<!--                    Ethnic_Categories + -->
+<!--                    Male  + -->
+<!--                    Has_Partner  + -->
+<!--                    Political_Conservativism + -->
+<!--                    Urban + -->
+<!--                    LIFESAT + PWI + includeNA(Beliefs), data = d_3,digits = 2, total=FALSE, format.p = FALSE, digits.p = 0))%>% -->
+<!--     kable( booktabs = T, "latex") -->
+<!-- ``` -->
+
+```{r latextable, cache=TRUE}
+library(modelsummary)
+ts<-d_3 %>%
+  dplyr::select(Age, 
+                NZdep,
+                Education,
+                Employed,
+                Ethnic_Categories,
+                Male,
+                Has_Partner, 
+                Political_Conservativism,
+                Urban, 
+                Beliefs, 
+                LIFESAT,
+                PWI,
+                Wave)
+modelsummary::datasummary_balance( ~ Wave, data=ts, dinm=FALSE, output = 'table.tex')
+# https://github.com/IndrajeetPatil/ggstatsplot
+# https://www.datanovia.com/en/lessons/compute-summary-statistics-in-r/
 ```
@@
 
-```{r, include=FALSE}
+```{r demographic table, cache=TRUE}
 # must remove Furniture package before running table 1
 demographic_table(d_3)
-```
-
-```{r, include=FALSE}
-beliefs_table(d_3)
-```
+#demographic_table_latex() not wurking  d
+```
+
+<!-- ```{r, include=FALSE} -->
+<!-- beliefs_table(d_3) -->
+<!-- ``` -->
 
@@
 # obtain dag
-tidy_ggdag <- dagitty::dagify(
+tidy_ggdag <- dagify(
   WB ~ belief + years + partner + nzdep + urban + male + pols + empl,
@@
 However we report both model in our frequentist analysis.
-Neither model choice effects the results.
+Neither model choice affects the results.
 We also report the simple unadjusted model and the fully saturated model (with potential collider and masking effects).
 
-A word of caution, the fully-saturated model -- which would be typical in a longitudional regression analysis -- attenuates the growth in personal well-being and in life satisfaction among those who are skeptical in a spirit or Life-fore.
-The inference of this study is not sentitive to this difference because the slopes differ for spirit beliefs across all models.
+Importantly, in the fully-saturated model -- which would be typical in a longitudinal regression analysis -- attenuates the growth in personal well-being and in life satisfaction among those who are skeptical in a spirit or Life-fore.
+The inference of this study is not sensitive to this difference because the slopes differ for spirit beliefs across all models.
 However, we do not wish to infer that skepticism about a spirit or life force entirely prevents growth in personal well-being or life-satisfaction.
-We advise that researchers do not introduce variables into regression mmodels without systematically.
+We advise that researchers do not introduce variables into regression models without systematically.
 evaluating potential confounding.
@@
 ## Bayesian data analysis
-
-Bayesian inference require substantial computational resources.
-The cost of time is repaid with a model that efficiently pools information to produce a more accurate, and transparent, estimate of uncertainty.
-The posterior distributions of the model describe intuitive conditional probabilities.
 
@@
 
-```{r setmodelformula}
-forumula_bayesPWI <- bf(PWI ~ yearsC * Beliefs + (1 + yearsC|Id))
-forumula_bayesLIFESAT <- bf(LIFESAT ~  yearsC * Beliefs + (1 + yearsC|Id))
-```
-
-```{r pwi_bayesmodel}
-# PWI
-mod.1_b <-brm(forumula_bayesPWI, data= d_3, file ="mod.1_b.rds", cores = 8) # note these are very large files
-```
-
-```{r lsbayesmodel}
-# LIFE SAT
-mod.2_b <-brm (forumula_bayesLIFESAT, data= d_3,cores = 8, file ="mod.2_b.rds")
-# note these are very large files
-```
-
-```{r simple_table, cache =TRUE}
-bayestestR::describe_posterior(mod.1_b, ci = 0.90,  test = c("p_direction"),diagnostic = "ESS")
-```
-
-```{r simple_table_ls, cache =TRUE}
-bayestestR::describe_posterior(mod.2_b, ci = 0.90, test = c("p_direction"), diagnostic = "ESS")
-```
-
-<!-- Tables: -->
-<!-- # ```{r pwi_Bayesian_model_table_pwi, cache =TRUE} -->
-<!-- # tab_model(mod.1_b) -->
-<!-- # ``` -->
-<!-- # -->
-<!-- # ```{r lifesat_Bayesian_model_table_ls lifesat, cache =TRUE} -->
-<!-- # tab_model(mod.2_b) -->
-<!-- # ``` -->
-
-<!-- ```{r texreg_bayes_table:pwi,cache=TRUE, include=FALSE} -->
-<!-- #Results for publication. -->
-<!-- table_model_latex_pwi(mod.1_b) -->
-<!-- ``` -->
-
-<!-- ```{r texreg_bayes_table:ls, cache=TRUE, include=FALSE} -->
-<!-- table_model_latex_ls(mod.2_b) -->
-<!-- ``` -->
-
-<!-- Coefficient plots used in manuscript-->
-<!-- ```{r coefficientplotsbayes, cache=TRUE} -->
-<!-- plot_coefs(mod.1_b,mod.2_b) -->
-<!-- ``` -->
-
-```{r stancode for models, include=FALSE}
-#model equation
-#Here we extract our priors for reporting
-
-stancode(mod.1_b)
-stancode(mod.2_b)
-```
-
-```{r prior summaries, include =FALSE}
-#this does the same as the stancode
-brms::prior_summary(mod.1_b)
-brms::prior_summary(mod.2_b)
-```
-
-```{r Bayesian_model_table, cache=TRUE}
-#tab_model(mod.1_b, show.ci= .89) # we really should use something other than 95 but the team consensus is to use 95
-tab_model(mod.1_b, mod.2_b, show.ci= .95) # no difference to inference
-```
-
-We inspected the posterior distributions, which look good.
-
-```{r mixing_pwi, cache=TRUE}
-plot(mod.1_b, pars ="b_") 
-```
-
-```{r mixing_ls, cache=TRUE}
-plot(mod.2_b, pars ="b_") 
-```
-
-As do the random components
-
-```{r mixing_sd_pwi, cache=TRUE}
-plot(mod.1_b, pars ="sd_") #
-```
-
-```{r mixing_ls_pwi, cache=TRUE}
-plot(mod.2_b, pars ="sd_") # 
-```
-
-The posterior prediction plots look reasonable.
-
-```{r posteriorpredictivechecks:pwi, cache=TRUE}
-# cannot run on my laptop
-pp_1<-rstanarm::pp_check(mod.1_b) + ylab("Personal well-being")
-pp_2<-rstanarm::pp_check(mod.2_b) + ylab("Life Satisfaction") ## life sat suggests an ordinal model
-pp_1 + pp_2
-```
-
-### Predictions graphs
-
-Personal Well being predictions.
-
-```{r}
-preds_1b<-get_predictions(mod.1_b)
-preds_1b
-```
-
-Life Satisfaction predictions.
-
-```{r}
-preds_2b<-get_predictions(mod.2_b)
-preds_2b
-```
-
-Next we graph the expected effects.
-
-```{r Bayesian_pwi_predicted, cache=TRUE, dependson= mod.1_b}
-graph_predictions(mod.1_b, "Predicted Values of Personal Well-Being")
-```
-
-```{r Bayesian_lifesat_predicted, cache=TRUE, dependson= mod.2_b}
-graph_predictions(mod.2_b, "Predicted Values of Life Satisfaction")
-```
-
 To close the "back door" on potential confounds we create a regression model
@@
 
+Here we describe the posteriors for both models 
+```{r simple_table, cache =TRUE}
+bayestestR::describe_posterior(mod.1_bDAG, test = c("p_direction"),diagnostic = "ESS")
+```
+
+```{r simple_table_ls, cache =TRUE}
+bayestestR::describe_posterior(mod.2_bDAG, test = c("p_direction"), diagnostic = "ESS")
+```
+
 ```{r Bayesian_model_table, cache=TRUE, include =FALSE}
@@
 
-```{r cache =TRUE}
-g1dag1<-plot(parameters::model_parameters(mod.1dag))+ ggtitle("Personal well-being Coefficients")
-g2dag1<-plot(parameters::model_parameters(mod.2dag))+ ggtitle("Life Satisfaction Coefficients")
+<!-- ```{r} -->
+<!-- cite_r("b-references.bib") -->
+<!-- ``` -->
+
+Next we plot the graphs. First we extract the predictions
+
+```{r Bayesian_pwi_predictedSAT_DAG_GET, cache=TRUE}
+get_predictions(mod.1_bDAG)
+```
+
+```{r Bayesian_lifesat_predictedSAT_DAG_GET, cache=TRUE}
+get_predictions(mod.2_bDAG)
+```
+```{r DAGS, cache =TRUE}
+g1dag1<-plot(parameters::model_parameters(mod.1_bDAG))+ ggtitle("Personal well-being Coefficients")
+g2dag2<-plot(parameters::model_parameters(mod.2_bDAG))+ ggtitle("Life Satisfaction Coefficients")
 ```
@@
 
-```{r plot_coefficientplots_LMER_FULL, cache = TRUE}
+```{r plot_coefficientplots_LMER_FULL_DAG, cache = TRUE}
 g1dag1/g2dag1
@@
 
-Get the predictions
-
-```{r Bayesian_pwi_predictedSAT, cache=TRUE, dependson= mod.1_b}
-get_predictions(mod.1_bDAG)
-```
-
-```{r Bayesian_lifesat_predictedSAT, cache=TRUE, dependson= mod.2_b}
-get_predictions(mod.2_bDAG)
-```
-
-```{r Bayesian_pwi_predictedSAT, cache=TRUE, dependson= mod.1_b}
+Here are the coefficient plots
+
+```{r Bayesian_pwi_predictedSAT_DAG, cache=TRUE}
 graph_predictions(mod.1_bDAG, "Predicted Values of Personal Well-Being: Bayesian Model (Confounder Adjusted)")
@@
 
-```{r Bayesian_lifesat_predictedSAT, cache=TRUE, dependson= mod.2_b}
+```{r Bayesian_lifesat_predictedSAT_DAG, cache=TRUE}
 graph_predictions(mod.2_bDAG, "Predicted Values of Life Satisfaction: Bayesian Model (Confounder Adjusted)")
@@
 
-# describe the posteriors 
-```{r simple_table, cache =TRUE}
-bayestestR::describe_posterior(mod.1_bDAG, test = c("p_direction"),diagnostic = "ESS")
-```
-
-```{r simple_table_ls, cache =TRUE}
-bayestestR::describe_posterior(mod.2_bDAG, test = c("p_direction"), diagnostic = "ESS")
-```
-
-## Test of the Bayesian saturacted model  for the record
-
-```{r setmodelformulaSAT}
+## Bayesian model, no confounders addressed
+
+In the simple model, the in
+
+```{r setmodelformula}
+forumula_bayesPWI <- bf(PWI ~ yearsC * Beliefs + (1 + yearsC|Id))
+forumula_bayesLIFESAT <- bf(LIFESAT ~  yearsC * Beliefs + (1 + yearsC|Id))
+```
+
+```{r pwi_bayesmodel_DAG}
+# PWI
+mod.1_b <-brm(forumula_bayesPWI, data= d_3, file ="mod.1_b.rds", cores = 8) # note these are very large files
+```
+
+```{r lsbayesmodel_DAG}
+# LIFE SAT
+mod.2_b <-brm (forumula_bayesLIFESAT, data= d_3,cores = 8, file ="mod.2_b.rds")
+# note these are very large files
+```
+
+```{r simple_table_DAG, cache =TRUE}
+bayestestR::describe_posterior(mod.1_b, ci = 0.90,  test = c("p_direction"),diagnostic = "ESS")
+```
+
+```{r simple_table_ls_DAG, cache =TRUE}
+bayestestR::describe_posterior(mod.2_b, ci = 0.90, test = c("p_direction"), diagnostic = "ESS")
+```
+
+<!-- Tables: -->
+<!-- # ```{r pwi_Bayesian_model_table_pwi, cache =TRUE} -->
+<!-- # tab_model(mod.1_b) -->
+<!-- # ``` -->
+<!-- # -->
+<!-- # ```{r lifesat_Bayesian_model_table_ls lifesat, cache =TRUE} -->
+<!-- # tab_model(mod.2_b) -->
+<!-- # ``` -->
+
+<!-- ```{r texreg_bayes_table:pwi,cache=TRUE, include=FALSE} -->
+<!-- #Results for publication. -->
+<!-- table_model_latex_pwi(mod.1_b) -->
+<!-- ``` -->
+
+<!-- ```{r texreg_bayes_table:ls, cache=TRUE, include=FALSE} -->
+<!-- table_model_latex_ls(mod.2_b) -->
+<!-- ``` -->
+
+<!-- Coefficient plots used in manuscript-->
+<!-- ```{r coefficientplotsbayes, cache=TRUE} -->
+<!-- plot_coefs(mod.1_b,mod.2_b) -->
+<!-- ``` -->
+
+```{r stancode for models, include=FALSE}
+#model equation
+#Here we extract our priors for reporting
+
+stancode(mod.1_b)
+stancode(mod.2_b)
+```
+
+```{r prior summaries, include =FALSE}
+#this does the same as the stancode
+brms::prior_summary(mod.1_b)
+brms::prior_summary(mod.2_b)
+```
+
+```{r Bayesian_model_table_SIMPLE, cache=TRUE}
+#tab_model(mod.1_b, show.ci= .89) # we really should use something other than 95 but the team consensus is to use 95
+tab_model(mod.1_b, mod.2_b, show.ci= .95) # no difference to inference
+```
+
+We inspected the posterior distributions, which look good.
+
+```{r mixing_pwi, cache=TRUE}
+plot(mod.1_b, pars ="b_") 
+```
+
+```{r mixing_ls, cache=TRUE}
+plot(mod.2_b, pars ="b_") 
+```
+
+As do the random components
+
+```{r mixing_sd_pwi, cache=TRUE}
+plot(mod.1_b, pars ="sd_") #
+```
+
+```{r mixing_ls_pwi, cache=TRUE}
+plot(mod.2_b, pars ="sd_") # 
+```
+
+The posterior prediction plots look reasonable.
+
+```{r posteriorpredictivechecks:pwi, cache=TRUE}
+# cannot run on my laptop
+pp_1<-rstanarm::pp_check(mod.1_b) + ylab("Personal well-being")
+pp_2<-rstanarm::pp_check(mod.2_b) + ylab("Life Satisfaction") ## life sat suggests an ordinal model
+pp_1 + pp_2
+```
+
+### Predictions graphs
+
+Personal Well being predictions.
+
+```{r}
+preds_1b<-get_predictions(mod.1_b)
+preds_1b
+```
+
+Life Satisfaction predictions.
+
+```{r}
+preds_2b<-get_predictions(mod.2_b)
+preds_2b
+```
+
+Next we graph the expected effects.
+
+```{r Bayesian_pwi_predicted, cache=TRUE, dependson= mod.1_b}
+graph_predictions(mod.1_b, "Predicted Values of Personal Well-Being")
+```
+
+```{r Bayesian_lifesat_predicted, cache=TRUE, dependson= mod.2_b}
+graph_predictions(mod.2_b, "Predicted Values of Life Satisfaction")
+```
+
+## Test of the Bayesian saturated model
+
+```{r setmodelformulaSAT_FULL}
 # note that BRMS automatically centers so we do not need to center years
@@
 
-```{r pwi_bayesmodelSAT}
+```{r pwi_bayesmodelSAT_FULL}
 # PWI
@@
 summary(mod.1_bSAT)
-```
-
-```{r Bayesian_pwi_predictedSAT, cache=TRUE, dependson= mod.1_b}
-graph_predictions(mod.1_bSAT, "Predicted Values of Personal Well-Being")
-```
-
-```{r Bayesian_lifesat_predictedSAT, cache=TRUE, dependson= mod.2_b}
-graph_predictions(mod.2_b, "Predicted Values of Life Satisfaction")
 ```
@@
 
+```{r Bayesian_pwi_predictedSAT, cache=TRUE}
+graph_predictions(mod.1_bSAT, "Predicted Values of Personal Well-Being")
+```
+```{r Bayesian_lifesat_predictedSAT, cache=TRUE}
+graph_predictions(mod.2_b, "Predicted Values of Life Satisfaction")
+```
+
 ## REML estimation
@@
 Results are virtually identical to the Bayesian model.
+
+## Adjusting for confounds
+
+### First confounder model
+
+```{r lmer_model_with_full_covariatesDAG1, cache =TRUE}
+library(lme4)
+pwi_formulaDag1 <- 'PWI ~  yearsC * Beliefs +  Age_in_Decades_C  + Ethnic_Categories + Male + (1 + yearsC|Id)'
+ls_formulaDag1 <- 'LIFESAT ~ yearsC * Beliefs +  Age_in_Decades_C  + Ethnic_Categories + Male  + (1 + yearsC|Id)'
+
+mod.1dag1 <-lmer (pwi_formulaDag1, data= d_3)
+mod.2dag1 <-lmer (ls_formulaDag1, data= d_3)
+```
+
+Here is the table.
+```{r tab_dag1, cache = TRUE}
+sjPlot::tab_model(mod.1dag1,mod.2dag1)
+```
+
+We compared complex with simple models, and we found the inference does not change. 
+```{r}
+# identical result
+sjPlot::tab_model(mod.1dag1,mod.2dag1)
+```
+
+```{r grapwi_predict_DAG, cache =TRUE}
+graph_predictions(mod.1dag1, "Predicted Values of Life Satisfaction: Dag 1")
+```
+```{r gralifesat_predict_DAG, cache =TRUE}
+graph_predictions(mod.2dag1, "Predicted Values of Life Satisfaction: Dag 1")
+```
+
+## Second confounder model
+
+```{r lmer_model_with_full_covariates_DAG, cache =TRUE}
+library(lme4)
+pwi_formulaDag2 <- 'PWI ~  yearsC * Beliefs +  Age_in_Decades_C  + Employed + Male + Deprivation_S + Political_Conservativism_S + (1 + yearsC|Id)'
+ls_formulaDag2 <- 'LIFESAT ~ yearsC * Beliefs + Employed +  Age_in_Decades_C  + Employed + Male + Deprivation_S + Political_Conservativism_S + (1 + yearsC|Id)'
+
+mod.1dag2 <-lmer (pwi_formulaDag2, data= d_3)
+mod.2dag2 <-lmer (ls_formulaDag2, data= d_3)
+```
+
+Here is the table
+
+```{r tab_dag2, cache = TRUE}
+sjPlot::tab_model(mod.1dag2,mod.1dag2)
+```
+
+```{r dag2_plotcreate, cache =TRUE}
+g1dag2<-plot(parameters::model_parameters(mod.1dag2))+ ggtitle("Personal well-being Coefficients")
+g2dag2<-plot(parameters::model_parameters(mod.2dag2))+ ggtitle("Life Satisfaction Coefficients")
+```
+
+And here's the graph.
+
+```{r plot_coefficientplots_LMER_FULL_DAG2, cache = TRUE}
+g1dag2/g2dag2 
+```
+
+```{r pwi_predict_DAG, cache =TRUE}
+graph_predictions(mod.1dag2, "Predicted Values of Life Satisfaction: Dag 2")
+```
+
+```{r glifesat_predict_DAG, cache =TRUE}
+graph_predictions(mod.2dag2, "Predicted Values of Life Satisfaction: Dag 2")
+```
+
+Note we would prefer the the first counfounder model because there is less measurement error (not employed could mean mean not retired).
+
+Additionally the first counfounder model has fewer terms to estimate.
+
+### ELSE
+For the record, we report the fully saturated "everything but the kitchen sink" model.  We show that throwing all regression coefficients produces a stronger effect for beliefs.
+
+```{r lmer model centering years_SATURATED}
+library(lme4)
+pwi_formulaC <- 'PWI ~  yearsC * Beliefs +  Age_in_Decades_C + Deprivation_S + Education_S  + Employed + Ethnic_Categories + Male  + Has_Partner  + Political_Conservativism_S + Urban + (1+ yearsC|Id)'
+
+ls_formulaC <- 'LIFESAT ~  yearsC * Beliefs +  Age_in_Decades_C + Deprivation_S + Education_S  + Employed + Ethnic_Categories + Male  + Has_Partner  + Political_Conservativism_S + Urban + (1+ yearsC|Id)'
+summary(mod.1a)
+mod.1a <-lmer (pwi_formulaC, data= d_3)
+mod.2a <-lmer (ls_formulaC, data= d_3)
+
+# identical result
+performance::check_collinearity(mod.1a)
+performance::check_collinearity(mod.2a)
+```
+
+Here's the table, again, we find inference is robust.
+
+```{r table of saturated models, cached = TRUE}
+sjPlot::tab_model(mod.1a,mod.2a)
+```
+
+<!-- Her are the coefficient plots for the full models -->
+<!-- ```{r coefficientplots_LMER_FULL, cache=TRUE} -->
+<!-- plot_models(mod.2, mod.1) + ylim(-.4, .75) + theme_plain()  # Plots -->
+<!-- ``` -->
 
@@
 
-## Adjusting for Demographics.
+## Fully saturated model
 
@@
 
-## Adjustment using DAG
-
-### First confounder model
-
-```{r lmer_model_with_full_covariatesDAG1, cache =TRUE}
-library(lme4)
-pwi_formulaDag1 <- 'PWI ~  yearsC * Beliefs +  Age_in_Decades_C  + Ethnic_Categories + Male + (1 + yearsC|Id)'
-ls_formulaDag1 <- 'LIFESAT ~ yearsC * Beliefs +  Age_in_Decades_C  + Ethnic_Categories + Male  + (1 + yearsC|Id)'
-
-mod.1dag1 <-lmer (pwi_formulaDag1, data= d_3)
-mod.2dag1 <-lmer (ls_formulaDag1, data= d_3)
-```
-
-Here is the table.
-```{r tab_dag1, cache = TRUE, dependson=d_3}
-sjPlot::tab_model(mod.1dag1,mod.2dag1)
-```
-
-We compared complex with simple models, and we found the inference does not change. 
-```{r}
-# identical result
-sjPlot::tab_model(mod.1dag1,mod.2dag1)
-```
-
-```{r Bayesian_lifesat_predict, cache =TRUE}
-graph_predictions(mod.1dag1, "Predicted Values of Life Satisfaction: Dag 1")
-```
-```{r Bayesian_lifesat_predict, cache =TRUE}
-graph_predictions(mod.2dag1, "Predicted Values of Life Satisfaction: Dag 1")
-```
-
-## Second confounder model
-
-```{r lmer_model_with_full_covariates, cache =TRUE}
-library(lme4)
-pwi_formulaDag2 <- 'PWI ~  yearsC * Beliefs +  Age_in_Decades_C  + Employed + Male + Deprivation_S + Political_Conservativism_S + (1 + yearsC|Id)'
-ls_formulaDag2 <- 'LIFESAT ~ yearsC * Beliefs + Employed +  Age_in_Decades_C  + Employed + Male + Deprivation_S + Political_Conservativism_S + (1 + yearsC|Id)'
-
-mod.1dag2 <-lmer (pwi_formulaDag2, data= d_3)
-mod.2dag2 <-lmer (ls_formulaDag2, data= d_3)
-```
-
-Here is the table
-
-```{r tab_dag2, cache = TRUE, dependson=d_3}
-sjPlot::tab_model(mod.1dag2,mod.1dag2)
-```
-
-```{r dag2_plotcreate cache =TRUE}
-g1dag2<-plot(parameters::model_parameters(mod.1dag2))+ ggtitle("Personal well-being Coefficients")
-g2dag2<-plot(parameters::model_parameters(mod.2dag2))+ ggtitle("Life Satisfaction Coefficients")
-```
-
-And here's the graph.
-
-```{r plot_coefficientplots_LMER_FULL_DAG2, cache = TRUE}
-g1dag2/g2dag2 
-```
-
-```{r Bayesian_lifesat_predict, cache =TRUE}
-graph_predictions(mod.1dag2, "Predicted Values of Life Satisfaction: Dag 2")
-```
-
-```{r Bayesian_lifesat_predict, cache =TRUE}
-graph_predictions(mod.2dag2, "Predicted Values of Life Satisfaction: Dag 2")
-```
-
-Note we would prefer the the first counfounder model because there is less measurement error (not employed could mean mean not retired).
-
-Additionally the first counfounder model has fewer terms to estimate.
-
-### ELSE
-For the record, we report the fully saturated "everything but the kitchen sink" model.  We show that throwing all regression coefficients produces a stronger effect for beliefs.
-
-```{r lmer model centering years}
-library(lme4)
-pwi_formulaC <- 'PWI ~  yearsC * Beliefs +  Age_in_Decades_C + Deprivation_S + Education_S  + Employed + Ethnic_Categories + Male  + Has_Partner  + Political_Conservativism_S + Urban + (1+ yearsC|Id)'
-
-ls_formulaC <- 'LIFESAT ~  yearsC * Beliefs +  Age_in_Decades_C + Deprivation_S + Education_S  + Employed + Ethnic_Categories + Male  + Has_Partner  + Political_Conservativism_S + Urban + (1+ yearsC|Id)'
-summary(mod.1a)
-mod.1a <-lmer (pwi_formulaC, data= d_3)
-mod.2a <-lmer (ls_formulaC, data= d_3)
-
-# identical result
-performance::check_collinearity(mod.1a)
-performance::check_collinearity(mod.2a)
-```
-
-Here's the table, again, we find inference is robust.
-
-```{r table of saturated models, cached = TRUE}
-sjPlot::tab_model(mod.1a,mod.2a)
-```
-
-<!-- Her are the coefficient plots for the full models -->
-<!-- ```{r coefficientplots_LMER_FULL, cache=TRUE} -->
-<!-- plot_models(mod.2, mod.1) + ylim(-.4, .75) + theme_plain()  # Plots -->
-<!-- ``` -->
-
-Here is the code for the coefficient plots of our preferred simple model.
-
 ```{r cache =TRUE}
@@
 
-Let's plot all the models. We find that uncertainty increases, but that, again, the inference remains robust.
+The inference is robut even to the poorly specified models using REML estimation.
 
@@
 
-We reran the imputed data analysis.
+We reran the imputed data analysis. (Note we ajust for confounds)
 
@@
 # run models iterating over imputed data
-mod_imputed_pwi<-loop_lmer_model(pwi_formula_simple, imputed3)
+mod_imputed_pwi<-loop_lmer_model(pwi_formulaDag1, imputed3)
 ```
@@
 # run models iterating over imputed data
-mod_imputed_ls<-loop_lmer_model(ls_formula_simple,imputed3)
+mod_imputed_ls<-loop_lmer_model(ls_formulaDag1 ,imputed3)
 

### 2020-11-30 funs.R

@@ -114,6 +114,26 @@ demographic_table_latex <- function(x){
 }
 
 
+sample_report <-function(x){
+  x %>%
+  dplyr::select(Age, 
+                NZdep,
+                Education,
+                Employed,
+                Ethnic_Categories,
+                Male,
+                Has_Partner, 
+                Political_Conservativism,
+                Urban, 
+                Beliefs, 
+                LIFESAT,
+                PWI,
+                Wave)%>%
+  dplyr::group_by(Wave)%>%
+  report::report()%>% 
+  summary()
+}
+
 # function for predicting PWI  Not used
 
 # graph_predictions_pwi <- function(x){ 

### 2020-11-30 libs.R

@@ -9,8 +9,6 @@ library("gghighlight")
 library("Amelia")
 library("magrittr")
 library("patchwork")
-#library("furniture") # messes with table1
-library("table1")
 library("sjPlot")
 library("texreg")
 library("kableExtra")
@@ -19,8 +17,9 @@ library("ggpubr")
 #library("ggpubr")
 #library("wesanderson")
 library("ggsci")
+library("report")
 library("papaja")
-#library("here")
+library("here")
 library("citr")
 library("here")
 library("huxtable")
@@ -30,7 +29,7 @@ library("rstanarm")
 library("tidybayes")
 library("dagitty")
 library("ggdag")
-
+#library("furniture") # messes with table1
 citation("ggdag")
 
 rstan_options(auto_write=TRUE)

### 2020-11-30 Supplement_And_Analysis.Rmd
near to final supplement. needs to be proofed

@@
 
+```{r mixing_pwi_DAG, cache=TRUE}
+plot(mod.1_bDAG, pars ="b_") 
+```
+
+```{r mixing_lsDAG, cache=TRUE}
+plot(mod.2_bDAG, pars ="b_") 
+```
+
+As do the random components
+
+```{r mixing_sd_pwi, cache=TRUE}
+plot(mod.1_bDAG, pars ="sd_") #
+```
+
+```{r mixing_ls_pwi, cache=TRUE}
+plot(mod.2_bDAG, pars ="sd_") # 
+```
+
 ## Bayesian model, no confounders addressed
@@
 
-```{r mixing_pwi, cache=TRUE}
-plot(mod.1_b, pars ="b_") 
-```
-
-```{r mixing_ls, cache=TRUE}
-plot(mod.2_b, pars ="b_") 
-```
-
-As do the random components
-
-```{r mixing_sd_pwi, cache=TRUE}
-plot(mod.1_b, pars ="sd_") #
-```
-
-```{r mixing_ls_pwi, cache=TRUE}
-plot(mod.2_b, pars ="sd_") # 
-```
-
 The posterior prediction plots look reasonable.
@@
 
-### ELSE
-For the record, we report the fully saturated "everything but the kitchen sink" model.  We show that throwing all regression coefficients produces a stronger effect for beliefs.
-
-```{r lmer model centering years_SATURATED}
-library(lme4)
-pwi_formulaC <- 'PWI ~  yearsC * Beliefs +  Age_in_Decades_C + Deprivation_S + Education_S  + Employed + Ethnic_Categories + Male  + Has_Partner  + Political_Conservativism_S + Urban + (1+ yearsC|Id)'
-
-ls_formulaC <- 'LIFESAT ~  yearsC * Beliefs +  Age_in_Decades_C + Deprivation_S + Education_S  + Employed + Ethnic_Categories + Male  + Has_Partner  + Political_Conservativism_S + Urban + (1+ yearsC|Id)'
-summary(mod.1a)
-mod.1a <-lmer (pwi_formulaC, data= d_3)
-mod.2a <-lmer (ls_formulaC, data= d_3)
-
-# identical result
-performance::check_collinearity(mod.1a)
-performance::check_collinearity(mod.2a)
-```
-
-Here's the table, again, we find inference is robust.
-
-```{r table of saturated models, cached = TRUE}
-sjPlot::tab_model(mod.1a,mod.2a)
-```
-
 <!-- Her are the coefficient plots for the full models -->
@@
 
-```{r robust to years and beliefs only, cache = TRUE, dependson=d_3,cache =TRUE}
-library(lme4)
-pwi_formula_simple <- 'PWI ~  yearsC * Beliefs + (1+ yearsC|Id)'
-ls_formula_simple <- 'LIFESAT ~  yearsC * Beliefs + (1+ yearsC|Id)'
-mod.1aa <-lmer(pwi_formula_simple, data= d_3)
-mod.2aa <-lmer(ls_formula_simple, data= d_3)
-
-# table
-sjPlot::tab_model(mod.1aa,mod.2aa)
-```
-
-Here is a predictive graph for years $\times$ beliefs interaction on PWI. Rhe function for producing this graph is in the `funs.R` file.
-
-```{r Bayesian_pwi_predict, cache =TRUE}
-graph_predictions(mod.1aa, "Predicted Values of Personal well-being")
-```
-
-Here is a predictive graph for the years $\times$ beliefs interaction on life satisfaction.
-
-```{r Bayesian_lifesat_predict, cache =TRUE}
-graph_predictions(mod.2aa, "Predicted Values of Life Satisfaction")
-```
-
-## Fully saturated model
-
-Next we attempted a model with that is adjusted by "usual suspect" demographic indicator. The inference remains unchanged.
-
-```{r lmer_model_with_full_covariates, cache =TRUE}
-library(lme4)
-pwi_formula <- 'PWI ~  yearsC * Beliefs +  Age_in_Decades_C + Deprivation_S + Education_S  + Employed + Ethnic_Categories + Male  + Has_Partner  + Political_Conservativism_S + Urban + (1 + yearsC|Id)'
-ls_formula <- 'LIFESAT ~  yearsC * Beliefs +  Age_in_Decades_C + Deprivation_S + Education_S  + Employed + Ethnic_Categories + Male  + Has_Partner  + Political_Conservativism_S + Urban + (1 + yearsC|Id)'
-
-mod.1 <-lmer (pwi_formula, data= d_3)
-mod.2 <-lmer (ls_formula, data= d_3)
-```
-
-Here is the table.
-```{r saturated model, cache = TRUE, dependson=d_3}
-sjPlot::tab_model(mod.1,mod.2)
-```
-
-We compared complex with simple models, and we found the inference does not change. 
-```{r}
-# identical result
-sjPlot::tab_model(mod.1,mod.1aa)
-sjPlot::tab_model(mod.2,mod.2aa)
-```
-
-Next we checked for multi-collinearity. Multi-collinearity is low. (Note that we centre the "years" variable, had we not done this, mult-collinearity would have been high).
-
-```{r check collinearity, cache =TRUE}
-# Low VIFS from 
-performance::check_collinearity(mod.1)
-performance::check_collinearity(mod.2)
-```
-
-```{r cache =TRUE}
-g1<-plot(parameters::model_parameters(mod.1aa))+ ggtitle("Personal well-being Coefficients")
-g2<-plot(parameters::model_parameters(mod.2aa))+ ggtitle("Life Satisfaction Coefficients")
-```
-
-And here's the graph.
-
-```{r plot_coefficientplots_LMER_FULL, cache = TRUE}
-g1/g2
-```
-
 Next, recall we want to check whether inference is robustness to the researcher degree of freedom in selecting how many number of waves a participant must have responded to for inclusion.
@@
 ```{r run all the models with different ns, cache = TRUE}
-mod.1_3 <-lmer (pwi_formula_simple, data= d_4)
-mod.2_3 <-lmer (ls_formula_simple, data= d_4)
-
-mod.1_4 <-lmer (pwi_formula_simple, data= d_4)
-mod.2_4 <-lmer (ls_formula_simple, data= d_4)
-
-mod.1_5 <-lmer (pwi_formula_simple, data= d_5)
-mod.2_5 <-lmer (ls_formula_simple, data= d_5)
-
-mod.1_6 <-lmer (pwi_formula_simple, data= d_6)
-mod.2_6 <-lmer (ls_formula_simple, data= d_6)
-
-mod.1_7 <-lmer (pwi_formula_simple, data= d_7)
-mod.2_7 <-lmer (ls_formula_simple, data= d_7)
-
-mod.1_8 <-lmer (pwi_formula_simple, data= d_8)
-mod.2_8 <-lmer (ls_formula_simple, data= d_8)
+mod.1_3 <-lmer (pwi_formulaDag1, data= d_4)
+mod.2_3 <-lmer (mod.2dag2, data= d_4)
+
+mod.1_4 <-lmer (pwi_formulaDag1, data= d_4)
+mod.2_4 <-lmer (mod.2dag2, data= d_4)
+
+mod.1_5 <-lmer (pwi_formulaDag1, data= d_5)
+mod.2_5 <-lmer (mod.2dag2, data= d_5)
+
+mod.1_6 <-lmer (pwi_formulaDag1, data= d_6)
+mod.2_6 <-lmer (mod.2dag2, data= d_6)
+
+mod.1_7 <-lmer (pwi_formulaDag1, data= d_7)
+mod.2_7 <-lmer (mod.2dag2, data= d_7)
+
+mod.1_8 <-lmer (pwi_formulaDag1, data= d_8)
+mod.2_8 <-lmer (mod.2dag2, data= d_8)
 ```
@@
 g2a_3/g2a_8 +   plot_annotation(title = "Compare 3 X response model (top) with 8 X response model (bottom) : Life Satisfaction")
-```
-
-Next, we modelled the relationship between having a partner and deprivation.
-We did so to address a reviewer's worry that we might have introduced correlation of between deprivation and having a partner that could spoil inference.
-We indeed found such a relationship.
-This is an important potential confound, and another reason to favour the simple longitudinal model that we used.
-
-```{r checkforref model,cache =TRUE}
-m<-lmer (NZdep ~ Partner + (1|Id), data= d_3, na.action = na.omit)
-tab_model(m)
-model_parameters(m)
 ```

### 2020-12-16 Supplement_And_Analysis.Rmd
typos
checked within/between age estimates using parameters::demean  -- no diff.

@@
 output            : "html_document" 
-
 editor_options: 
@@
 
-## How many responses waves to select?
+## How many waves to select?
 
@@
 mod.2dag2 <-lmer (ls_formulaDag2, data= d_3)
+
+dat <- cbind(
+  d_3,
+  demean(d_3, select = c("Age", "Beliefs"), group = "Id")
+)
+
+dat<-dat%>%
+  dplyr::mutate(Age_between_C = scale(Age_between, center =TRUE, scale = FALSE))
+
+str(dat)
+head(dat$Age_between_C)
+
+# After thought -- check within and between group aging... no differences
+# dmPWI <- 'PWI ~  Age_within * Beliefs + Age_between_C  +  Employed + Male + Deprivation_S + Political_Conservativism_S + (1 + Age_within|Id)'
+# mod_dmPWI <-lmer (dmPWI, data= dat)
+# tab_model(mod_dmPWI)
+# summary(mod_dmPWI)
+# n1<-ggeffects::ggpredict(model = mod_dmPWI, terms = c("Age_within [minmax]","Beliefs"), 
+#                               ci.lvl = 0.95,
+#                               type = "fe",
+#                               typical = "mean",
+#                               back.transform = TRUE,
+#                               ppd = FALSE,
+#                               interval = "confidence")
+#   plot(n1, facets = T) +  gghighlight::gghighlight()  +  theme_blank() #+ ggtitle(y) # title to be suppled
+#   
+# graph_predictions(mod_dmPWI, "Predicted Values of PWI")
+# # 
+# lsPWI <- 'LIFESAT ~  Age_within * Beliefs + Age_between_C  +  Employed + Male + Deprivation_S + Political_Conservativism_S + (1 + Age_within|Id)'
+# mod_lsPWI <-lmer (lsPWI, data= dat)
+# tab_model(mod_lsPWI)
+# summary(mod_lsPWI)
+# head(dat$Age_within)
+# n2<-ggeffects::ggpredict(model = mod_lsPWI, terms = c("Age_within [minmax]","Beliefs"),
+#                               ci.lvl = 0.95,
+#                               type = "fe",
+#                               typical = "mean",
+#                               back.transform = TRUE,
+#                               ppd = FALSE,
+#                               interval = "confidence")
+# plot(n2, facets = T) +  gghighlight::gghighlight()  +  theme_blank() #+ ggtitle(y) # title to be suppled
+
 ```

### 2021-02-12 Manuscript_Bayesian.Rmd
fixed bib

@@
 Here, we addresses these three challenges of previous research in the following ways: First, to avoid a tautological elision of "spirituality" and "well-being" we adopt a non-affective measure of spirituality as a "believe in a Spirit or Life Force." On the other side, to measure subjective well-being, we use a well-validated scale for personal well-being and a well-validated scale for Life-Satisfaction (see Method).
-Second, to disentangle spiritual beliefs from traditional religious beliefs and skeptical beliefs we adopt a four-level categorical indicator combining the binary indicator of belief in a Spirit or Life Force with the binary indicator "Do you believe in some form of a spirit or Life Force?" The levels of this four-level categorical indicator for spiritual, religious, and skeptical beliefs are as follows: (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life Force; (2) Skeptics about God who nevertheless believe in a Spirit or Life Force; (3) Those who believe in a God but do not believe in a Spirit or Life Force; (4) Those who who believe in a both God and a Spirit or Life Force.
-Clearly such an approach cannot simultaneously address all interests in previous spirituality/well-being research.
+Second, to disentangle spiritual beliefs from traditional religious beliefs and dis0beliefs we adopt a four-level categorical indicator combining the binary indicator of belief in a Spirit or Life Force with the binary indicator "Do you believe in some form of a spirit or Life Force?" The levels of this four-level categorical indicator for spiritual and religious beliefs and dis-beliefs are as follows: (1) **God and Spirit Disbelievers**: those who neither believe in a God nor believe in a Spirit or Life Force; (2) **Spirit Believers/God Disbelievers**: those who disbelieve in a God and believe in a Spirit or Life Force; (3) **God Believers/Spirit Disbelievers**: those who believe in a God but do not believe in a Spirit or Life Force; (4) **God and Spirit Believers**: Those who believe in a both God and a Spirit or Life Force.
+Clearly such an approach to categorising religion and spirituality cannot simultaneously address many interests in previous spirituality/well-being research.
 For example we do not combine our cognitive measure of belief with behavioral measures or attitudinal measures.
 However, this cost to scope is paid for with benefits to precision.
-The spiritual indicators measure cognitive states in which distinct varieties of skeptical beliefs, traditional religious beliefs, and non-traditional spirit beliefs are straightforwardly disentangled, and measured within people.
-Third, to understand how spirit beliefs affect growth, stability, or decline in psychological well-being over time, we leverage longitudinal responses from a nationally diverse probability sample year over nine years (2010 to 2018 New Zealand Attitudes and Values Study, NZAVS).
+The spiritual indicators measure cognitive states in which distinct varieties of religious and spiritual beliefs and dis-beliefs are straightforwardly disentangled and measured within people, and their longitudinal associations with reliable well-being indicators are systematically quantified Third, to understand how spirit beliefs affect growth, stability, or decline in psychological well-being over time, we leverage longitudinal responses from a nationally diverse probability sample year over nine years (2010 to 2018 New Zealand Attitudes and Values Study, NZAVS).
 
@@
  & Urban & 2271 & 2.0 & 3695 & 3.2 & 6709 & 5.8 & 10602 & 9.2 & 9937 & 8.6 & 8815 & 7.6 & 11622 & 10.1 & 12687 & 11.0 & 13291 & 11.5\\
-Beliefs & \_Skeptic\_ & 625 & 0.5 & 1312 & 1.1 & 2235 & 1.9 & 3639 & 3.2 & 0 & 0.0 & 3278 & 2.8 & 4420 & 3.8 & 4135 & 3.6 & 4514 & 3.9\\
+Beliefs & \_Disbeliefs\_ & 625 & 0.5 & 1312 & 1.1 & 2235 & 1.9 & 3639 & 3.2 & 0 & 0.0 & 3278 & 2.8 & 4420 & 3.8 & 4135 & 3.6 & 4514 & 3.9\\
  & \_SpiritExcludesGod\_ & 778 & 0.7 & 1508 & 1.3 & 2535 & 2.2 & 4206 & 3.6 & 0 & 0.0 & 3542 & 3.1 & 4808 & 4.2 & 4148 & 3.6 & 4309 & 3.7\\
@@
 ### Belief in a Spirit/Life Force and/or a God  
-We assess belief in a spirit or a God by asking two non-affective questions: "Do you believe in a spirit or Life Force" and "Do you believe in a god?" Responses were coded as (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life Force; (2) Skeptics about God who believe in a Spirit or Life Force; (3) Believers in a God who disbelieve in a Spirit or Life Force; (4) Believers in a God and Spirit or Life Force.
-These believe were developed for the NZAVS from the 2005 Eurobarometer.
+We assess belief in a spirit or a God by asking two non-affective questions: "Do you believe in a spirit or Life Force" and "Do you believe in a god?" Responses were coded as 1) **God and Spirit Disbelievers**: those who neither believe in a God nor believe in a Spirit or Life Force; (2) **Spirit Believers/God Disbelievers**: those who disbelieve in a God and believe in a Spirit or Life Force; (3) **God Believers/Spirit Disbelievers**: those who believe in a God but do not believe in a Spirit or Life Force; (4) **God and Spirit Believers**: those who believe in a both God and a Spirit or Life Force.
 
@@
 In our models, we account for the non-linear effects of time within participants by including the interaction of years (a continuous variable) with Beliefs (a four-level categorical indicator).
-The baseline for this indicator is Skepticism about a Spirit/Life Forces and Skepticism about a God.
+The baseline for this indicator is disbelief in a Spirit/Life Force and disbelief in a God.
 To handle the dependencies introduced from the repeated measures, we include individual ID as an effect modeled as random, and we estimated a random effect for the slope for years.
@@
 Indeed, the inference proved robust to the simple model with only the interaction of years and beliefs, and robust to the fully saturated model in which all covariates were included (see the online supplement for details).
-In the fully saturated model, it appeared that those skeptical of a Spirit or Life Force, exhibited no growth in Persona Well-Being and Life-Satisifaction over the nine years.
+In the fully saturated model, it appeared that those disbelieving in a Spirit or Life Force, exhibited no growth in Persona Well-Being and Life-Satisifaction over the nine years.
 Although this result offers stronger stronger evidence for a Spirit/Life Force well-being mechanism, it is likely the effect of collider confounding and masking.
@@
 
-  - The intercept, representing the expected personal well-being among skeptics about a Spirit/Life Force and God who are Europeans, 50.77 years of age and do not identify as male is reliable (posterior median = 7.138 HDI [ 7.111,  7.167], pd = 100.00%).
+  - The intercept, representing the expected personal well-being among disbelievers in a Spirit/Life Force and God who are Europeans, 50.77 years of age and do not identify as male is reliable (posterior median = 7.138 HDI [ 7.111,  7.167], pd = 100.00%).
   - The effect of years is positive is reliable (posterior median =  7.138, HDI [ 7.111,  7.167], pd  100.00%).
-We infer that personal well-being increased in the baseline skeptic population between 2010 and 2018.
+We infer that personal well-being increased in the baseline disbelieving population between 2010 and 2018.
   - The effect of beliefs in *Spirit Excludes God*  is reliably negative (posterior median =  -0.059, HDI [-0.081, -0.036], pd 100.00%).
-We infer that the population that believed in a Spirit but did not believe in a God was lower in personal well-being at the intercept compared to skeptics.
+We infer that the population that believed in a Spirit but did not believe in a God was lower in personal well-being at the intercept compared to disbelievers.
   - The effect of beliefs in *God And Spirit*  is not  negative (posterior median = -0.060, HDI [-0.086, -0.034], pd = 100.00%) .
-We infer that the population that believed in a God and a Spirit or Life Force was lower in personal well-being at the intercept compared to skeptics.
+We infer that the population that believed in a God and a Spirit or Life Force was lower in personal well-being at the intercept compared to disbelievers.
   - The effect of beliefs in *God Excludes Spirit*  is not reliable (posterior median = -0.001 HDI [-0.036,  0.035], pd = 51.10%).
-We infer that the population that believed in a God and doubted a Spirit or Life Force were similar were similar in personal well-being at the intercept to skeptics.
+We infer that the population that believed in a God and doubted a Spirit or Life Force were similar were similar in personal well-being at the intercept to the disbelieving population:
   - The effect of Age in Decades (C) is   0.072 | [ 0.060,  0.083] | 100.00% |
@@
 The effect of years $\times$ belief in a *Spirit Excludes God*  is reliably positive  (posterior median = 0.011, HDI [ 0.003,  0.018], pd = 99.02%).
-We infer that people with beliefs in a spirit or Life Force who do not believe in a God tended to grow in their personal well-being over time at a faster rate than skeptics grew.
+We infer that people with beliefs in a spirit or Life Force who do not believe in a God tended to grow in their personal well-being over time at a faster rate than disbelievers grew.
   - The effect of years $\times$ in *God And Spirit*  is reliably positive  (posterior median = 0.014, HDI [0.007, 0.021], pd = 99.98%).
-We infer that people who believe in a spirit or Life Force who also believe in a God tended to grow in their personal well-being over time at a faster rate than skeptics grew.
+We infer that people who believe in a spirit or Life Force who also believe in a God tended to grow in their personal well-being over time at a faster rate than disbelievers grew.
   - The effect of years $\times$ belief in *God Excludes Spirit*  is not reliable (posterior median = 0.004, HDI [-0.009,  0.015], pd =  68.73%).
-We infer that people who believe in a God but not a spirit or Life Force did not grow at a faster rate in personal well-being than skeptics grew.
+We infer that people who believe in a God but not a spirit or Life Force did not grow at a faster rate in personal well-being than disbelievers grew.
 
@@
 Within this model: 
-  - The intercept, representing the expected life satisfaction among skeptics about a Spirit/Life Force and God who are Europeans, 50.77 years of age and do not identify as male is reliable (posterior median = 5.287, HDI [ 5.267,  5.306], pd = 100.00%).
+  - The intercept, representing the expected life satisfaction among disbelievers in a Spirit/Life Force and God who are Europeans, 50.77 years of age and do not identify as male is reliable (posterior median = 5.287, HDI [ 5.267,  5.306], pd = 100.00%).
   - The effect of years is positive is reliable (posterior median =  0.008, HDI [ 0.003, 0.011], pd = 99.75%).
-We infer that life satisfaction increased in the skeptic population between 2010 and 2018.
+We infer that life satisfaction increased in the disbelieving population between 2010 and 2018.
   -The effect of beliefs in *Spirit Excludes God*  is not reliable (posterior median =   0.007, HDI [-0.010,  0.023], pd = 75.33%).
-We infer that the population that believed in a Spirit but did not believe in a God was higher in life satisfaction at the intercept compared with skeptics.
+We infer that the population that believed in a Spirit but did not believe in a God was higher in life satisfaction at the intercept compared with disbelievers.
   - The effect of beliefs in *God And Spirit*  is reliably positive (posterior median = 0.034, HDI [ 0.015, 0.053], pd = 99.83%).
-We infer that the population that believed in a God and a Spirit or Life Force was higher in life satisfaction at the intercept compared with skeptics.
+We infer that the population that believed in a God and a Spirit or Life Force was higher in life satisfaction at the intercept compared with disbelievers.
   - The effect of beliefs in *God Excludes Spirit*  is reliably positive (posterior median =  0.057, HDI [ 0.030, 0.082], pd = 99.95%).
-We infer that the population that believed in a God and doubted a Spirit or Life Force  was higher in life satisfaction at the intercept compared with skeptics.
+We infer that the population that believed in a God and doubted a Spirit or Life Force  was higher in life satisfaction at the intercept compared with disbelievers.
   - The effect of Age in Decades (C) is reliably positive (posterior median = 0.079, hDI [ 0.071,  0.087] pd = 100.00%).
@@
   - The interaction of years $\times$ beliefs in *Spirit Excludes God* is reliably positive (posterior median = 0.011, HDI [ 0.006, 0.016], pd = 99.95%).
-We infer that people with beliefs in a Spirit or Life Force who do not believe in a God tended to grow over time in their life satisfaction at a faster rate than skeptics.
+We infer that people with beliefs in a Spirit or Life Force who do not believe in a God tended to grow over time in their life satisfaction at a faster rate than disbelievers.
   - The interaction of years $\times$ *God And Spirit*  is reliably positive (posterior median =  0.009, HDI [ 0.004, 0.014], pd = 99.72%).
-We infer that people who believe in a spirit or Life Force who also believe in a God tended to grow over time in their life satisfaction at a faster rate than skeptics.
+We infer that people who believe in a spirit or Life Force who also believe in a God tended to grow over time in their life satisfaction at a faster rate than disbelievers.
   - The effect of years $\times$ beliefs in *God Excludes Spirit*  is not reliable (posterior median = -0.003, HDI [-0.012, 0.006], pd = 72.28%).
-We infer that people who believe in a God but not a spirit or Life Force tended to grow over time in their life satisfaction at a similar rate to skeptics.
+We infer that people who believe in a God but not a spirit or Life Force tended to grow over time in their life satisfaction at a similar rate to disbelievers.
 
@@
 1.
-At the outset of the study, in 2010, the expected subjective well-being as measured using the personal well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God ($\hat{y} =$  6.93 | [6.88, 6.98]) and lower among those who believed in both a Spirit/Life Force and a God ($\hat{y}$ = 6.91 | [6.87, 6.95]) in comparison both to the skeptical population who neither believed in a God nor Spirit/ Life Force ($\hat{y}$ = 7.04 | [6.99, 7.10]) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 7.03 | [6.95, 7.10]).
+At the outset of the study, in 2010, the expected subjective well-being as measured using the personal well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God ($\hat{y} =$  6.93 | [6.88, 6.98]) and lower among those who believed in both a Spirit/Life Force and a God ($\hat{y}$ = 6.91 | [6.87, 6.95]) in comparison both to the population who neither believed in a God nor Spirit/ Life Force ($\hat{y}$ = 7.04 | [6.99, 7.10]) and to the population that believed in a God but was disbelieving in a Spirit or Life Force ($\hat{y}$ 7.03 | [6.95, 7.10]).
 (See Figure 2)  Similarly,the expected subjective well-being as measured using the life satisfaction Inventory was lower in 2010 among those who believed in Spirit or Life Force but not a God ($\hat{y}$ =   5.20 | [5.16, 5.23]).
-Life satisfaction was similar among those who believed in both a Spirit/Life Force and a God  ($\hat{y}$ =   5.23 | [5.21, 5.26]) in comparison both with the skeptical population who neither believed in a God nor a Life Force ($\hat{y}$ = 5.25 | [5.21, 5.29]).
-Life satisfaction was highest among the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 5.32 | [5.26, 5.38]) (see Figure 3)
+Life satisfaction was similar among those who believed in both a Spirit/Life Force and a God  ($\hat{y}$ =   5.23 | [5.21, 5.26]) in comparison both with the disbelieving population who neither believed in a God nor a Life Force ($\hat{y}$ = 5.25 | [5.21, 5.29]).
+Life satisfaction was highest among the population that believed in a God but was disbelieving in a Spirit or Life Force ($\hat{y}$ 5.32 | [5.26, 5.38]) (see Figure 3)
 
@@
 Among those who believed in both a Spirit/Life Force and a God, personal well-being had grown to ($\hat{y} = 7.20$ [7.17, 7.24], $\Delta \hat{y} = 0.29$).
-By contrast, personal well-being among the skeptical population who neither believed in a God nor a Life Force grew less steeply across the eight year interval $\hat{y} =  7.21$ [7.17, 7.25], $\Delta \hat{y} = 0.17$).
-By the same token, personal well-being among those who believe in a God but are skeptical about a Spirit/Life Force grew less steeply $\hat{y} =  7.22$ [7.16, 7.28] $\Delta \hat{y} = 0.19$).
+By contrast, personal well-being among the disbelieving population who neither believed in a God nor a Life Force grew less steeply across the eight year interval $\hat{y} =  7.21$ [7.17, 7.25], $\Delta \hat{y} = 0.17$).
+By the same token, personal well-being among those who believe in a God but are disbelieving a Spirit/Life Force grew less steeply $\hat{y} =  7.22$ [7.16, 7.28] $\Delta \hat{y} = 0.19$).
 In practical terms, the rate of increase in personal wellbeing among those with Spirit/Life Force beliefs was about 30% greater than it was among those who lacked such Spirit/Life Force (See Figure 2).
@@
 In 2018, the expected effect of spiritual beliefs on life satisfaction among those who believe in Spirit or Life Force but not a God had grown to ($\hat{y} =  5.37$ [5.34, 5.40], $\Delta \hat{y} = 0.17$).
-Among those who believed in both a Spirit/Life Force and a God, life satisfaction had grown to ($\hat{y} = 5.39$  [5.36, 5.41], $\Delta \hat{y} = 0.19$).By contrast, life satisfaction among the skeptical population who neither believed in a God nor a Life Force lagged behind over the eight year interval $\hat{y} =   5.32$ [5.29, 5.35], $\Delta \hat{y} = 0.07$).
-Relatedly, life satisfaction in the population that believe in God but are skeptical about a Spirit/Life Force grew less steeply $\hat{y} =  5.36$ [5.31, 5.41], $\Delta \hat{y} = 0.04$).
+Among those who believed in both a Spirit/Life Force and a God, life satisfaction had grown to ($\hat{y} = 5.39$  [5.36, 5.41], $\Delta \hat{y} = 0.19$).By contrast, life satisfaction among the population who neither believed in a God nor a Life Force lagged behind over the eight year interval $\hat{y} =   5.32$ [5.29, 5.35], $\Delta \hat{y} = 0.07$).
+Relatedly, life satisfaction in the population that believe in God but are disbelieving in a Spirit/Life Force grew less steeply $\hat{y} =  5.36$ [5.31, 5.41], $\Delta \hat{y} = 0.04$).
 In practical terms, the rate of increase in life satisfaction among those with Spirit/Life Force beliefs over three times greater than among those that lacked such beliefs (See Figure 3).
@@
 Two fundamental question arise from the patterns we observe in this study.
-The first question is why those who believed in Spirit or Life Force were initially lower in their personal well-being and life-satisfaction in the initial years of this study in comparison to religious and non-religious skeptics?
+The first question is why those who believed in Spirit or Life Force  were expected to be initially lower relative the populations that did not believe in a Spirit or Life Force?
 During the early years of New Zealand Attitudes and Values data collection New Zealand was emerging from the twin impacts of a global financial recession and a series of devastating earthquakes that destroyed 30% of Christchurch, New Zealand's second largest city [@greaves2015regional].
@@
 
-# Appendix 
-## Model Equations
-### Priors for the Personal Wellbeing model
-
-Model priors for Personal Wellbeing including group-level ID intercept and beliefs slopes are as follows
-
-$$
-\begin{aligned}
-\operatorname{Personal Well Being}_{ij} \sim \text{Normal}(\mu_{ij},\sigma^2) \nonumber\\
-\mu_{ij}\sim \boldsymbol{\alpha +\beta} \nonumber\\
-\boldsymbol{\alpha} \sim \alpha_{0} +\alpha_j \nonumber\\
-\alpha_{0} \sim \text{StudentT}(3,7.2,2.5)\nonumber\\
-\boldsymbol{\beta}\sim \boldsymbol{b} + \beta_j \nonumber\\
-\boldsymbol{b}\sim\text{Normal}(0,2.5) \nonumber\\
-\begin{bmatrix}
-\alpha_j \\
-\beta_j
-\end{bmatrix}
-\sim 
-\text{MVNormal}
-\begin{pmatrix}
-\begin{bmatrix}
-0\\
-0
-\end{bmatrix}
-,\boldsymbol{S} \nonumber
-\end{pmatrix}\\
-\boldsymbol{S} = 
-\begin{pmatrix}
-0 & \sigma_{\beta_j} \nonumber
-\end{pmatrix} \boldsymbol{R} \begin{pmatrix}
-\sigma_{\alpha_j} & 0 \\
-0 & \sigma_{\beta_j} \nonumber
-\end{pmatrix}\\
-\sigma_{\alpha_j} \sim 
-\text{HalfCauchy}(0,2.5) \nonumber\\
-\sigma_{\beta_j} \sim 
-\text{HalfCauchy}(0,2.5) \nonumber\\
-\boldsymbol{R}\sim \text{LKJcorr}(1) \nonumber\\
-\sigma  \sim  \operatorname{HalfStudent T}^{+}(3, 0, 2.5) 
-\end{aligned}
-$$
-
-\newpage
-
-### Priors for the Life Satisfaction model
-
-Model priors for Life Satisfaction Model with including group-level ID intercept and belief slopes are as follows:
-
-$$
-\begin{aligned}
-\operatorname{Life Satisfaction}_{ij} \sim \text{Normal}(\mu_{ij},\sigma^2) \nonumber\\
-\mu_{ij}\sim \boldsymbol{\alpha +\beta} \nonumber\\
-\boldsymbol{\alpha} \sim \alpha_{0} +\alpha_j \nonumber\\
-\alpha_{0} \sim \text{StudentT}(3,5.5,2.5)\nonumber\\
-\alpha_j \sim \text{StudentT}+(3,0,2.5)\nonumber\\
-\boldsymbol{\beta}\sim \boldsymbol{b} + \beta_j \nonumber\\
-\boldsymbol{b}\sim\text{Normal}(0,2.5) \nonumber\\
-\begin{bmatrix}
-\alpha_j \\
-\beta_j
-\end{bmatrix}
-\sim 
-\text{MVNormal}
-\begin{pmatrix}
-\begin{bmatrix}
-0\\
-0
-\end{bmatrix}
-,\boldsymbol{S} \nonumber
-\end{pmatrix}\\
-\boldsymbol{S} = 
-\begin{pmatrix}
-\sigma_{\alpha_j} & 0 \\
-0 & \sigma_{\beta_j} \nonumber
-\end{pmatrix} \boldsymbol{R} \begin{pmatrix}
-\sigma_{\alpha_j} & 0 \\
-0 & \sigma_{\beta_j} \nonumber
-\end{pmatrix}\\
-\sigma_{\alpha_j} \sim 
-\text{HalfCauchy}(0,1) \nonumber\\
-\sigma_{\beta_j} \sim 
-\text{HalfCauchy}(0,1) \nonumber\\
-\boldsymbol{R}\sim \text{LKJcorr}(1) \nonumber\\
-\sigma  \sim  \operatorname{HalfStudent T}^{+}(3, 0, 2.5) 
-\end{aligned}
-$$
-\newpage
-
-### Deterministic model
-
-The deterministic portion of both the Personal Wellbeing and the Life Satisfaction models is as follows: 
-$$
-\begin{aligned}
-& \quad  \boldsymbol{b} = \\
-& \quad \beta_{1}(\operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\
-&\quad \beta_{2}(\operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\  + \\
-& \quad  \beta_{3}(\operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}})\ +\\
-&\quad \beta_{4}(\operatorname{Age in Decades (C)})\  +\\ 
-& \quad  \beta_{5}(\operatorname{Ethnicity}_{\operatorname{Maori}})\ +\\
-&\quad \beta_{6}(\operatorname{Ethnicity}_{\operatorname{Pacific}})\ +  \\ 
-& \quad  \beta_{7}(\operatorname{Ethnicity}_{\operatorname{Asian}})\ +\\
-& \quad \beta_{8}(\operatorname{Male} )+ \\ 
-& \quad \beta_{9}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\ 
-& \quad \beta_{10}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ + \\
-&\quad \beta_{11}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) \\ 
-\end{aligned}
-$$
-\newpage
-
 # References

### 2021-02-12 Supplement_And_Analysis.Rmd
fixed bib

@@
 
-# Bayesian counfounder model
+## Priors in the Bayesian Model
+
+### Priors for the Personal Wellbeing model
+
+Model priors for Personal Wellbeing including group-level ID intercept and beliefs slopes are as follows:
+
+$$
+\begin{aligned}
+\operatorname{Personal Well Being}_{ij} \sim \text{Normal}(\mu_{ij},\sigma^2) \nonumber\\
+\mu_{ij}\sim \boldsymbol{\alpha +\beta} \nonumber\\
+\boldsymbol{\alpha} \sim \alpha_{0} +\alpha_j \nonumber\\
+\alpha_{0} \sim \text{StudentT}(3,7.2,2.5)\nonumber\\
+\boldsymbol{\beta}\sim \boldsymbol{b} + \beta_j \nonumber\\
+\boldsymbol{b}\sim\text{Normal}(0,2.5) \nonumber\\
+\begin{bmatrix}
+\alpha_j \\
+\beta_j
+\end{bmatrix}
+\sim 
+\text{MVNormal}
+\begin{pmatrix}
+\begin{bmatrix}
+0\\
+0
+\end{bmatrix}
+,\boldsymbol{S} \nonumber
+\end{pmatrix}\\
+\boldsymbol{S} = 
+\begin{pmatrix}
+0 & \sigma_{\beta_j} \nonumber
+\end{pmatrix} \boldsymbol{R} \begin{pmatrix}
+\sigma_{\alpha_j} & 0 \\
+0 & \sigma_{\beta_j} \nonumber
+\end{pmatrix}\\
+\sigma_{\alpha_j} \sim 
+\text{HalfCauchy}(0,2.5) \nonumber\\
+\sigma_{\beta_j} \sim 
+\text{HalfCauchy}(0,2.5) \nonumber\\
+\boldsymbol{R}\sim \text{LKJcorr}(1) \nonumber\\
+\sigma  \sim  \operatorname{HalfStudent T}^{+}(3, 0, 2.5) 
+\end{aligned}
+$$
+
+### Priors for the Life Satisfaction model
+
+Model priors for Life Satisfaction Model with including group-level ID intercept and belief slopes are as follows:
+
+$$
+\begin{aligned}
+\operatorname{Life Satisfaction}_{ij} \sim \text{Normal}(\mu_{ij},\sigma^2) \nonumber\\
+\mu_{ij}\sim \boldsymbol{\alpha +\beta} \nonumber\\
+\boldsymbol{\alpha} \sim \alpha_{0} +\alpha_j \nonumber\\
+\alpha_{0} \sim \text{StudentT}(3,5.5,2.5)\nonumber\\
+\alpha_j \sim \text{StudentT}+(3,0,2.5)\nonumber\\
+\boldsymbol{\beta}\sim \boldsymbol{b} + \beta_j \nonumber\\
+\boldsymbol{b}\sim\text{Normal}(0,2.5) \nonumber\\
+\begin{bmatrix}
+\alpha_j \\
+\beta_j
+\end{bmatrix}
+\sim 
+\text{MVNormal}
+\begin{pmatrix}
+\begin{bmatrix}
+0\\
+0
+\end{bmatrix}
+,\boldsymbol{S} \nonumber
+\end{pmatrix}\\
+\boldsymbol{S} = 
+\begin{pmatrix}
+\sigma_{\alpha_j} & 0 \\
+0 & \sigma_{\beta_j} \nonumber
+\end{pmatrix} \boldsymbol{R} \begin{pmatrix}
+\sigma_{\alpha_j} & 0 \\
+0 & \sigma_{\beta_j} \nonumber
+\end{pmatrix}\\
+\sigma_{\alpha_j} \sim 
+\text{HalfCauchy}(0,1) \nonumber\\
+\sigma_{\beta_j} \sim 
+\text{HalfCauchy}(0,1) \nonumber\\
+\boldsymbol{R}\sim \text{LKJcorr}(1) \nonumber\\
+\sigma  \sim  \operatorname{HalfStudent T}^{+}(3, 0, 2.5) 
+\end{aligned}
+$$
+\newpage
+
+### Deterministic model
+
+The deterministic portion of both the Personal Wellbeing and the Life Satisfaction models is as follows: 
+$$
+\begin{aligned}
+& \quad  \boldsymbol{b} = \\
+& \quad \beta_{1}(\operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\
+&\quad \beta_{2}(\operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\  + \\
+& \quad  \beta_{3}(\operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}})\ +\\
+&\quad \beta_{4}(\operatorname{Age in Decades (C)})\  +\\ 
+& \quad  \beta_{5}(\operatorname{Ethnicity}_{\operatorname{Maori}})\ +\\
+&\quad \beta_{6}(\operatorname{Ethnicity}_{\operatorname{Pacific}})\ +  \\ 
+& \quad  \beta_{7}(\operatorname{Ethnicity}_{\operatorname{Asian}})\ +\\
+& \quad \beta_{8}(\operatorname{Male} )+ \\ 
+& \quad \beta_{9}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\ 
+& \quad \beta_{10}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ + \\
+&\quad \beta_{11}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) \\ 
+\end{aligned}
+$$
+\newpage
+
+## Bayesian counfounder model
 

### 2021-02-14 Analysis.Rmd
update

@@
 ---
-title             : "National Longitudinal Evidence for Growth in Well-being From Spiritual Beliefs"
+title             : "Supplement: National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Beliefs"
 shorttitle        : "Title"
@@
 abstract: |
- A growing body of research finds an association between spirituality and subjective well-being.
-However, the use of tautological scales and a reliance on cross-sectional data compromise inference.
-Here, drawing on nine years of panel responses from a nationally diverse longitudinal study.
-We systematically test whether having spiritual beliefs improves subjective well-being (N = 21,705, New Zealand, 2010-2018).
-Supporting previous studies, we observe that belief in a spirit or life force predicts increasing life satisfaction over time.
-However, we also observe that relative to the disbelieving population, belief in a spirit or life force is initially associated with lower personal well-being.
-Such effects, which hold, after adjusting for known demographic and personal influences on subjective well-being, are consistent with findings that spiritual beliefs are compelling to those with lower well-being.
-However, despite the scale and resolution of these longitudinal data, the causal interplay between spiritual beliefs and subjective well-being remain matters for future investigations.
+ Previous research finds an association between spirituality and subjective well-being.
+However, the widespread use of tautological spirituality scales, poorly defined concepts of spirituality, and heavy reliance on cross-sectional samples cast doubts.
+Here, we leverage nine waves of panel data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction ($N$ = 20979, New Zealand, 2010-2018).
+Contrary to previous research, we find that belief in a spirit or life force predicts lower personal well-being and life-satisfaction.
+However, in support of previous speculation, beliefs in a spirit or life force predict increasing personal well-being and life satisfaction over time relative to disbelief.
+These finding are robust after to known demographic influences, and intriguingly, hold among those who believe in a God but disbelieve in a Spirit or Life Force.
+The recent growth in spiritual beliefs and decline in traditional religion across many industrial societies motivates further causal investigations of the mechanisms by which spiritual beliefs lead to growth in subjective well-being.
 
@@
 ```
-
 ```{r global_options, echo=FALSE, include = FALSE}
@@
 ```{r load_libraries, echo=FALSE, include=FALSE}
-library("styler")
-library("tidyverse")
-library("lme4")
-library("ggplot2")
-library("tidyr")
-library("ggeffects")
-library("Amelia")
-library("table1")
-library("patchwork")
-#library("easystats")
-library("parameters")
-library("prettycode")
-library("table1")
-library("gghighlight")
-library("sjPlot")
-library("texreg")
-library("see")
-#library("viridis")
-#library("ggpubr")
-#library("wesanderson")
-library("ggsci")
-library("papaja")
-#library("here")
-library(parameters) # great tables, plots
-library("citr")
-library("brms")
-library("rstan")
-rstan_options(auto_write=TRUE)
-options(mc.cores=parallel::detectCores ())
-#theme_set(theme_few())
+source("libs.R")
+source("funs.R")
+```
+```{r, sessioninfo, include=FALSE}
+#  custom function for citing packages. 
+cite_packages()
 
+# custom function for my machine to read data. you'll need to read data to your machine. tweak function in the `funs.R' script for your machine
+d<-data_read_mymachine()
+
+# function to clean data, specify number of waves as 3 
+d_3<-data_clean_spirit_wellbeing(d,3) # d = dataframe, 3 = number of waves
+# function to count uinque ides
+show_unique_id(d_3$Id)
+
+# function to count waves for participants
+count_waves_participants(d_3)
+
+# table of beliefs 
+table(d_3$Beliefs)
+
+# years
+range(d_3$years)
 ```
-
-```{r, sessioninfo, include=FALSE}
-
-citation(package = "base", lib.loc = "/Users/josephbulbulia/Dropbox/BIBS")
-toLatex(sessionInfo(), locale = FALSE)
-sapply(names(sessionInfo()$otherPkgs), function(x) print(citation(x), style = "Bibtex"))
-
-out <- sapply(names(sessionInfo()$otherPkgs),
-              function(x) print(citation(x), style = "Bibtex"))
-print(out)
-```
-
-# Methods
- <!-- 21-word solution (Simmons, Nelson & Simonsohn, 2012; retrieved from http://ssrn.com/abstract=2160588) -->
-We used `r cite_r("b-references.bib")` for all our analyses.
-
-Load data (this is sepcific to my machine)
-```{r load_data}
-ldf.5<-readRDS("~/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/00Bulbulia\ Pubs/2020/ldf.5") 
-```
-
-Processing of data.
-```{r data processing, cache = TRUE, dependson = "ldf.5"}
-# filter only those who are measured 
-dfben  <-ldf.5 %>%
-  dplyr:: filter(Wave !=2009)%>% # exclude wave 1 (no spirit measures that wave)
-  dplyr::filter(YearMeasured==1)%>%
-  dplyr::group_by(Id) %>% filter(n() > 2)%>% # select those who have responded to at least 3 waves # Also works for 5 and 7
-  dplyr::filter(n() !=0)%>%
-  dplyr::ungroup(Id)%>%
-  dplyr:: mutate(Years = as.numeric(years))%>%
-  dplyr:: mutate(yearsC = scale(years, center=TRUE,scale=FALSE))%>%
-  dplyr::mutate(Age.10yrs = (Age/10))%>%
-  dplyr::mutate(Religion.Church.Log = log(Religion.Church+ 1))%>%
-  dplyr::mutate(Pol.Orient.S = scale(Pol.Orient, center = TRUE, scale = TRUE),
-         Age.10yrs.C = scale(Age.10yrs, center = TRUE, scale = FALSE),
-         Male = factor(Gender),
-         Employed = factor(Employed),
-         Partner = factor(Partner),
-         Edu.S = scale(as.numeric(Edu) ,scale=TRUE,center=TRUE),
-         Education = as.numeric(Edu),
-         Urban = factor(Urban),
-         Believe.Spirit = as.factor(Believe.Spirit),
-         Believe.God = as.factor(Believe.God),
-         EthnicCategories = factor(EthnicCats),
-         LIFESAT.S = scale(LIFESAT, center = TRUE, scale = TRUE),
-         PWI.S = scale(PWI, center = TRUE, scale = TRUE),
-         Deprivation.S = scale(NZdep, scale=TRUE, center=TRUE),
-         Relid.C = scale(Relid, scale = TRUE,center=TRUE),
-         Religious = as.factor(Religious),
-         Religion.Church.Log.C = scale(Religion.Church.Log, center = TRUE, scale = FALSE))
-
-# nice labels
-table1::label(dfben$Age.10yrs.C) <- "Age in Decades (C)"
-table1::label(dfben$PWI) <- "Personal Wellbeing"
-table1::label(dfben$LIFESAT) <- "Life Satisfaction"
-
-table1::label(dfben$Deprivation.S) <- "Deprivation (S)"
-table1::label(dfben$Pol.Orient) <- "Political Conservative"
-table1::label(dfben$Pol.Orient.S) <- "Political Conservative (S)"
-table1::label(dfben$NZdep) <- "NZ Deprivation Index"
-dfben$Partner <- factor(dfben$Partner, labels = c("No Partner","Has Partner"))
-dfben$Employed <- factor(dfben$Employed, labels = c("Not Employed","Employed"))
-table1::label(dfben$Edu.S) <- "Education (S)"
-
-# beliefs indicator
-dfben$Beliefs<- factor(ifelse(dfben$Believe.God == "Not Believe God" & dfben$Believe.Spirit == "Not Believe Spirit", "_Skeptic_", 
-                                        ifelse(dfben$Believe.God == "Not Believe God" & dfben$Believe.Spirit == "Believe Spirit", "_SpiritExcludesGod_",
-                                               ifelse(dfben$Believe.God == "Believe God" & dfben$Believe.Spirit == "Believe Spirit","GodAndSpirit","GodExcludesSpirit"))))
-```
-
-Table of belief types
-```{r}
-table(dfben$Beliefs)
-# years
-range(dfben$years)
-```
-
-```{r echo=FALSE}
-# nice labels
-#dfben$GodSpiritBeliefs <- factor(dfben$GodSpiritBeliefs, labels = c("Skeptic","SpiritExcludesGod","GodAndSpirit","GodExcludesSpirit"))
-# check to make sure labels are right
-#table(dfben$GodSpiritBeliefs) == table(dfben$GodSpiritBeliefs)
-```
-
-There are `r  length(unique(dfben$Id))` participants.
-
-This is the number of times participants returned responses: 
-```{r}
-dplyr::count(tally(group_by(dfben, Id), sort = TRUE, name="number_waves"),number_waves)
-```
-
-Table.
 
@@
 # must remove Furniture package before running table 1
-table1::table1(~  Age + NZdep + Education + Employed + EthnicCategories + Male  + Partner  + Pol.Orient + Urban +  Beliefs + LIFESAT + PWI|Wave, data = dfben, overall = F)
+demographic_table(d_3)
 ```
@@
 ```{r}
-table1::table1(~ Beliefs|Wave, data = dfben, overall = T)
+beliefs_table(d_3)
 ```
 
-```{r, include =FALSE}
-#furniture::table1(dfben, Age, NZdep,Education,Employed,EthnicCategories,Male,Partner, Pol.Orient,Urban, Beliefs, LIFESAT,PWI,
-                  #  splitby = ~Wave, overall=F, output = "latex2", booktabs=TRUE)
+```{r latex_data_table}
+# you might need to install the dev verion.//remotes::install_github("tysonstanley/furniture")
+demographic_table_latex(d_3)
 ```
 
-<!-- ```{r, include=FALSE} -->
-<!-- r<-easystats::report_participants( -->
-<!--   dfben, -->
-<!--   age = "Age", -->
-<!--   sex = "Male", -->
-<!--   education = "Education", -->
-<!--   participants = "Id", -->
-<!--   group = "Wave", -->
-<!--   spell_n = FALSE, -->
-<!-- ) -->
-<!-- ??report_participants -->
-<!-- ``` -->
-
-We demean variables following https://easystats.github.io/parameters/articles/demean.html
+# Bayesian estimation
 
 ```{r}
-dfben <- cbind(
-  dfben,
-  parameters::demean(dfben, select = c("Beliefs"), group = "Id")
-)
+forumula_bayesPWI <- bf(PWI ~  1 +  years * Beliefs + (1|Id))
+forumula_bayesLIFESAT <- bf(PWI ~  1 +  years * Beliefs + (1|Id))
 
+# PWI
+mod.1aaB <-brm(forumula_bayesPWI, data= d_3, file ="mod.1aaB")
+
+# LIFE SAT
+mod.2aaB <-brm (forumula_bayesLIFESAT, data= d_3,data= dfben,
+                file ="mod.2aaB")
 ```
 
-# Results
-
-```{r models, cache = TRUE, dependson = "dfben"}
-### Spirit Belief Models ### 
-mod01<-lmer (PWI ~ Years +
-                Beliefs__Skeptic__between  +  
-                Beliefs__SpiritExcludesGod__between + 
-                Beliefs_GodAndSpirit_between  + 
-                Beliefs_GodExcludesSpirit_between +
-                Beliefs__Skeptic__within  +  
-                Beliefs__SpiritExcludesGod__within + 
-                Beliefs_GodAndSpirit_within  + 
-                Beliefs_GodExcludesSpirit_within +
-                (1|Id), data= dfben, na.action = na.omit)
-
-parameters::check_heterogeneity(mod01, group = "ID")
-parameters::check_heterogeneity(mod01, group = "ID")
-
-mod02<-lmer (LIFESAT ~ Years +
-                Beliefs__Skeptic__between  +  
-                Beliefs__SpiritExcludesGod__between + 
-                Beliefs_GodAndSpirit_between  + 
-                Beliefs_GodExcludesSpirit_between +
-                Beliefs__Skeptic__within  +  
-                Beliefs__SpiritExcludesGod__within + 
-                Beliefs_GodAndSpirit_within  + 
-                Beliefs_GodExcludesSpirit_within +
-                (1|Id), data= dfben, na.action = na.omit)
-
-parameters::check_heterogeneity(mod02, group = "ID")
-
-# inference holds, but can we use this method with factors? What do the contrasts mean? 
-tab_model(mod01)
-tab_model(mod02)
-plot_models(mod01,mod02)
+```{r bayesian_pwi}
+conditional_effects(mod.1aaB)[[3]]
 ```
-
-```{r models, cache = TRUE, dependson = "dfben"}
-### Spirit Belief Models ### 
-mod01<-lmer (PWI ~ yearsC +
-                Beliefs__Skeptic__between  +  
-                Beliefs__SpiritExcludesGod__between + 
-                Beliefs_GodAndSpirit_between  + 
-                Beliefs_GodExcludesSpirit_between +
-                Beliefs__Skeptic__within  +  
-                Beliefs__SpiritExcludesGod__within + 
-                Beliefs_GodAndSpirit_within  + 
-                Beliefs_GodExcludesSpirit_within +
-                (1|Id), data= dfben, na.action = na.omit)
-
-parameters::check_heterogeneity(mod01, group = "ID")
-
-mod02<- lmer(LIFESAT ~ yearsC +
-                Beliefs__Skeptic__between  +  
-                Beliefs__SpiritExcludesGod__between + 
-                Beliefs_GodAndSpirit_between  + 
-                Beliefs_GodExcludesSpirit_between +
-                Beliefs__Skeptic__within  +  
-                Beliefs__SpiritExcludesGod__within + 
-                Beliefs_GodAndSpirit_within  + 
-                Beliefs_GodExcludesSpirit_within +
-                (1|Id), data= dfben, na.action = na.omit)
-
-parameters::check_heterogeneity(mod02, group = "ID")
-
-# inference holds, but can we use this method with factors? What do the contrasts mean? 
-tab_model(mod01)
-tab_model(mod02)
-plot_models(mod01,mod02)
-
-performance::check_collinearity(mod01)
-performance::check_collinearity(mod02)
+```{r bayesian_lifesat}
+conditional_effects(mod.2aaB)[[3]]
 
@@
 ```
-
 Indeed, models are robust without control covariates
@@
 tab_model(mod.2, mod.1) # Table
-```
-
-```{r}
-# check model with bayesian estimator (preferred)
-mod.1aaB <-brm(PWI ~  1 +  years * Beliefs + (1|Id), data= dfben, file ="mod.1aaB")
-mod.2aaB <-brm (LIFESAT ~ 1 + years * Beliefs + (1|Id), data= dfben,
-                file ="mod.2aaB")
-```
-
-note that the bayesian model recovers the time slope
-
-```{r bayesian_pwi}
-conditional_effects(mod.1aaB)[[3]]
-```
-
-```{r bayesian_lifesat}
-conditional_effects(mod.2aaB)[[3]]
-
 ```
@@
 # complete cases
-
 dfbenTS<-dfben%>%
@@
 
+<!-- We might want to look at within/between indivual change (i.e.
+do people who change in degree of spirit beliefs manifest different wellbeing.
+I ran this as below, however, what we really want is a markov model, where we compare state changes between the different types of belief.
+So this model is not useful.
+For more on demeaning see:  -->
+<!-- https://easystats.github.io/parameters/articles/demean.html -->
+
+<!-- ```{r} -->
+<!-- dfben <- cbind( -->
+<!--   dfben, -->
+<!--   parameters::demean(dfben, select = c("Beliefs"), group = "Id") -->
+<!-- ) -->
+<!-- ``` -->
+<!-- ```{r models, cache = TRUE, dependson = "dfben"} -->
+<!-- ### Spirit Belief Models ###  -->
+<!-- mod01<-lmer (PWI ~ Years + -->
+<!--                 Beliefs__Skeptic__between  +   -->
+<!--                 Beliefs__SpiritExcludesGod__between +  -->
+<!--                 Beliefs_GodAndSpirit_between  +  -->
+<!--                 Beliefs_GodExcludesSpirit_between + -->
+<!--                 Beliefs__Skeptic__within  +   -->
+<!--                 Beliefs__SpiritExcludesGod__within +  -->
+<!--                 Beliefs_GodAndSpirit_within  +  -->
+<!--                 Beliefs_GodExcludesSpirit_within + -->
+<!--                 (1|Id), data= dfben, na.action = na.omit) -->
+
+<!-- parameters::check_heterogeneity(mod01, group = "ID") -->
+<!-- parameters::check_heterogeneity(mod01, group = "ID") -->
+
+<!-- mod02<-lmer (LIFESAT ~ Years + -->
+<!--                 Beliefs__Skeptic__between  +   -->
+<!--                 Beliefs__SpiritExcludesGod__between +  -->
+<!--                 Beliefs_GodAndSpirit_between  +  -->
+<!--                 Beliefs_GodExcludesSpirit_between + -->
+<!--                 Beliefs__Skeptic__within  +   -->
+<!--                 Beliefs__SpiritExcludesGod__within +  -->
+<!--                 Beliefs_GodAndSpirit_within  +  -->
+<!--                 Beliefs_GodExcludesSpirit_within + -->
+<!--                 (1|Id), data= dfben, na.action = na.omit) -->
+<!-- parameters::check_heterogeneity(mod02, group = "ID") -->
+<!-- # inference holds, but can we use this method with factors? What do the contrasts mean?  -->
+<!-- tab_model(mod01) -->
+<!-- tab_model(mod02) -->
+<!-- plot_models(mod01,mod02) -->
+<!-- ``` -->
+
 # Discussion

### 2021-02-14 dag.R

@@ -1,0 +1,26 @@
+
+library("tidyverse")
+library("dagitty")
+library("ggdag")
+
+
+tidy_ggdag <- dagify(
+  WB ~ belief + years + partner + nzdep + urban + male + pols + empl,
+  WB ~~ partner + years,
+  belief ~ years + age + male + ethn,
+  partner ~ nzdep + age + belief, 
+  nzdep ~ empl,
+  pols ~ age  + empl + ethn + years,
+  empl ~ nzdep + edu + ethn,
+  exposure =  "belief",
+  outcome =   "WB")%>%
+  tidy_dagitty()
+
+ggdag(tidy_ggdag) +
+  theme_dag()
+
+ggdag_adjustment_set(tidy_ggdag, node_size = 14) + 
+  theme(legend.position = "bottom") 
+# Male
+# Ethn
+# AgeC

### 2021-02-24 funs.R
Update funs.R

@@ -4,7 +4,7 @@ data_read_mymachine <- function(){
   out <-readRDS("~/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/00Bulbulia\ Pubs/2020/ldf.5")
 }
 
-# this code is for citing packages 
+# this code is for citing packages
 
 cite_packages <- function() {
   citation(package = "base", lib.loc = "/Users/josephbulbulia/Dropbox/BIBS")
@@ -17,17 +17,18 @@ cite_packages <- function() {
 
 
 # data clean
-data_clean_spirit_wellbeing <- function(df,y) {  
+data_clean_spirit_wellbeing <- function(df,y) {
   #function for cleaning this dataset for any arbitrary number of waves
   n_waves =  as.numeric(y)
   out <- df %>%
-    dplyr::mutate(Beliefs = factor(ifelse(df$Believe.God == "Not Believe God" & df$Believe.Spirit == "Not Believe Spirit", "_Skeptic_", 
+    dplyr::mutate(Beliefs = factor(ifelse(df$Believe.God == "Not Believe God" & df$Believe.Spirit == "Not Believe Spirit", "_Skeptic_",
                                           ifelse(df$Believe.God == "Not Believe God" & df$Believe.Spirit == "Believe Spirit", "_SpiritExcludesGod_",
                                                  ifelse(df$Believe.God == "Believe God" & df$Believe.Spirit == "Believe Spirit","GodAndSpirit","GodExcludesSpirit")))))%>%
     dplyr::filter(Wave !=2009)%>% # exclude wave 1 (no spirit measures that wave)
+    droplevels() %>% dropp unused waves
     dplyr::filter(YearMeasured==1)%>%
-    dplyr::group_by(Id) %>% 
-    dplyr::filter(n() > n_waves-1)%>% # select those who have responded to at least 3 waves # 
+    dplyr::group_by(Id) %>%
+    dplyr::filter(n() > n_waves-1)%>% # select those who have responded to at least 3 waves #
     dplyr::filter(n() !=0)%>%
     dplyr::ungroup(Id)%>%
     dplyr:: mutate(Years = as.numeric(years))%>%
@@ -52,7 +53,7 @@ data_clean_spirit_wellbeing <- function(df,y) {
                   Religious = as.factor(Religious),
                   Religion_Church_Log_C = scale(Religion.Church.Log, center = TRUE, scale = FALSE))%>%
[…]

### 2021-02-26 Manuscript_Bayesian.Rmd
fixes

@@
 However, the widespread use of tautological spirituality scales, poorly defined concepts of spirituality, and heavy reliance on cross-sectional samples cast doubts.
-Here, we leverage nine waves of panel data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction ($N$ = 20979, New Zealand, 2010-2018).
-Contrary to previous research, we find that belief in a spirit or Life Force predicts lower personal well-being and life-satisfaction.
+Here, we leverage ten waves of panel data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction ($N$ = 3,257, New Zealand, 2010-2020).
+Contrary to previous research, we find that belief in a Spirit or Life Force predicts lower personal well-being and life-satisfaction.
 However, in support of previous speculation, beliefs in a spirit or Life Force predict increasing personal well-being and life satisfaction over time relative to disbelief.
@@
 keywords          : "Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being"
-wordcount         : "5161 words in text body, 1191 words in reference section"
+wordcount         : "4760 words in text body, 1191 words in reference section"
 bibliography      : ["b-references.bib"]
@@
 extra_dependencies: ["rotating", "booktabs", "graphicx","longtable","amsmath","amsfonts","amssymb"]
+---
 
----
 ```{r libraries}
@@
 ```
+
 ```{r global_options, echo=FALSE, include = FALSE}
@@
 The NZAVS is an annual, longitudinal national probability sample of registered New Zealand voters, which was started in 2009.
-The Time 10 wave of the NZAVS contained responses from 47,951 participants (17,981 retained from one or more previous waves and 29,970 new additions from booster sampling and/or unmatched participants or unsolicited opt-ins).
-Participants who provided an email address were also emailed and invited to complete an online version if they preferred.
-We offered a prize draw for participation, non-respondents were emailed and phoned multiple times, and all participants were mailed a Season’s Greetings card from the NZAVS research team and informed that they had been automatically entered into a bonus seasonal grocery voucher prize draw.
-We also mailed our yearly pamphlet summarizing key research findings published during the current wave of the study.
+The NZAVS samples size by waves is as follows: Wave 1 (2009/10) = 6518; Wave 2 (2010/11) = 4441; Wave 3 (2011/12) = 6884; Wave 4 (2012/13) =  12179; Wave 5 (2013/14) =  18,261; Wave 6 (2014/15) = 15,820; Wave 7 (2015/16) = 13,942; Wave 8 (2016/17) = 21,936; Wave 9 (2017/18) =  17,072; Wave 10 (2018/19) = 47951; Wave 11(2019/20) = 36,524.
+
+The first NZAVS measures of belief in God and belief in a spirit or life force appear in Wave 2 (2010/11) NZAVS contained responses from 4,441 participants.
+The Wave 2 (2010/11) NZAVS retained 4,425 from the initial Time 1 (2009/10) NZAVS sample of 6,518 participants, and included an additional 16 respondents who could not be matched to the Time 1 participant database (a retention rate of 67.9% over one year).
+Participants in the initial Time 1 (2009) sample were randomly selected from the New Zealand electoral roll (a national registry of registered voters).
+The response rate in the initial Time 1 sample, adjusting for the accuracy of the electoral roll and including anonymous responses was 16.6%.
+Participants were posted a copy of the questionnaire, with a second postal follow-up two months later.
+Participants who provided an email address were also emailed and invited to complete an online questionnaire if they preferred.
+Full details for the NZAVS sampling procedure can be found in the supplement to this study can be found on our OSF link: <https://osf.io/cu2gr/>
 
@@
 
-The Time 10 (2018) wave of the NZAVS included 47,951 respondents.
-We analyzed data from participants who responded to our survey at least three times between Time 2 (2010) and Time 10 (2018), resulting in a sample of $N$ = 21,705 ($N$ = 20,979 with complete responses).
+We analyzed data from participants who responded to our survey in at least 9 of the 10 waves between Time 2 (2010/2011) and Time 11 (2019/2020), resulting in a sample of $N$ = 3,257 individuals.
 Descriptive summaries for demographic and theoretical indicators are given in Table 1.
 
-Table 1 about here 
-\begin{table}[h]
-\begin{center}
-\scalebox{0.45}{
-\begin{tabular}[t]{llllllllllllllllllll}
-\toprule
-\multicolumn{2}{c}{ } & \multicolumn{2}{c}{2010 (N=3820)} & \multicolumn{2}{c}{2011 (N=6153)} & \multicolumn{2}{c}{2012 (N=10385)} & \multicolumn{2}{c}{2013 (N=15836)} & \multicolumn{2}{c}{2014 (N=15026)} & \multicolumn{2}{c}{2015 (N=13665)} & \multicolumn{2}{c}{2016 (N=18071)} & \multicolumn{2}{c}{2017 (N=15779)} & \multicolumn{2}{c}{2018 (N=16501)} \\
-\cmidrule(l{3pt}r{3pt}){3-4} \cmidrule(l{3pt}r{3pt}){5-6} \cmidrule(l{3pt}r{3pt}){7-8} \cmidrule(l{3pt}r{3pt}){9-10} \cmidrule(l{3pt}r{3pt}){11-12} \cmidrule(l{3pt}r{3pt}){13-14} \cmidrule(l{3pt}r{3pt}){15-16} \cmidrule(l{3pt}r{3pt}){17-18} \cmidrule(l{3pt}r{3pt}){19-20}
-  &    & Mean & Std.
-Dev.
-& Mean  & Std.
-Dev.
-& Mean   & Std.
-Dev.
-& Mean    & Std.
-Dev.
-& Mean     & Std.
-Dev.
-& Mean      & Std.
-Dev.
-& Mean       & Std.
-Dev.
-& Mean        & Std.
-Dev.
-& Mean         & Std.
-Dev.
-\\
-\midrule
-Age &  & 51.0 & 14.7 & 50.9 & 15.5 & 50.2 & 14.8 & 48.6 & 14.0 & 49.7 & 14.0 & 51.0 & 13.9 & 50.9 & 13.6 & 51.9 & 13.6 & 52.7 & 13.5\\
-NZdep &  & 4.8 & 2.8 & 4.6 & 2.7 & 4.9 & 2.8 & 4.7 & 2.8 & 4.7 & 2.8 & 4.7 & 2.8 & 4.6 & 2.7 & 4.6 & 2.7 & 4.6 & 2.7\\
-Education &  &  &  &  &  & 5.8 & 2.8 & 6.0 & 2.8 & 6.1 & 2.8 & 6.2 & 2.8 & 6.4 & 2.7 & 6.5 & 2.7 & 6.5 & 2.7\\
-Political\_Conservativism &  & 3.9 & 1.3 & 3.7 & 1.4 & 3.6 & 1.3 & 3.6 & 1.3 & 3.6 & 1.3 & 3.6 & 1.3 & 3.6 & 1.4 & 3.6 & 1.4 & 3.6 & 1.4\\
-LIFESAT &  & 5.2 & 1.2 & 5.1 & 1.2 & 5.1 & 1.2 & 5.2 & 1.2 & 5.2 & 1.2 & 5.2 & 1.2 & 5.2 & 1.2 & 5.3 & 1.2 & 5.3 & 1.2\\
-PWI &  & 6.8 & 1.7 & 6.8 & 1.7 & 6.8 & 1.7 & 7.0 & 1.7 & 7.0 & 1.7 & 7.1 & 1.7 & 7.1 & 1.7 & 7.1 & 1.7 & 7.1 & 1.6\\
-\midrule
- &  & N & \% & N & \% & N & \% & N & \% & N & \% & N & \% & N & \% & N & \% & N & \%\\
-Employed & 0 & 793 & 0.7 & 1276 & 1.1 & 2666 & 2.3 & 3531 & 3.1 & 3262 & 2.8 & 2972 & 2.6 & 4035 & 3.5 & 3655 & 3.2 & 3746 & 3.3\\
- & 1 & 2977 & 2.6 & 4795 & 4.2 & 7566 & 6.6 & 11991 & 10.4 & 11626 & 10.1 & 10378 & 9.0 & 14016 & 12.2 & 12070 & 10.5 & 12250 & 10.6\\
-Ethnic\_Categories & Euro & 2947 & 2.6 & 4375 & 3.8 & 7881 & 6.8 & 12270 & 10.6 & 12051 & 10.5 & 10959 & 9.5 & 14641 & 12.7 & 12906 & 11.2 & 13475 & 11.7\\
- & Maori & 556 & 0.5 & 638 & 0.6 & 1489 & 1.3 & 1897 & 1.6 & 1830 & 1.6 & 1608 & 1.4 & 1964 & 1.7 & 1826 & 1.6 & 1872 & 1.6\\
- & Pacific & 110 & 0.1 & 126 & 0.1 & 343 & 0.3 & 408 & 0.4 & 400 & 0.3 & 336 & 0.3 & 385 & 0.3 & 286 & 0.2 & 296 & 0.3\\
- & Asian & 124 & 0.1 & 190 & 0.2 & 409 & 0.4 & 579 & 0.5 & 572 & 0.5 & 473 & 0.4 & 664 & 0.6 & 574 & 0.5 & 579 & 0.5\\
-Male & 0 & 2390 & 2.1 & 3862 & 3.4 & 6542 & 5.7 & 10033 & 8.7 & 9500 & 8.2 & 8548 & 7.4 & 11377 & 9.9 & 9949 & 8.6 & 10388 & 9.0\\
- & 1 & 1430 & 1.2 & 2291 & 2.0 & 3843 & 3.3 & 5803 & 5.0 & 5478 & 4.8 & 5072 & 4.4 & 6635 & 5.8 & 5776 & 5.0 & 6077 & 5.3\\
-Has\_Partner & 0 & 1054 & 0.9 & 1806 & 1.6 & 2988 & 2.6 & 4237 & 3.7 & 3664 & 3.2 & 3407 & 3.0 & 4274 & 3.7 & 3775 & 3.3 & 3931 & 3.4\\
- & 1 & 2764 & 2.4 & 4325 & 3.8 & 7331 & 6.4 & 11493 & 10.0 & 11130 & 9.7 & 10176 & 8.8 & 13613 & 11.8 & 11964 & 10.4 & 12078 & 10.5\\
-Urban & Not\_Urban & 1531 & 1.3 & 2010 & 1.7 & 3512 & 3.0 & 5116 & 4.4 & 4894 & 4.2 & 4541 & 3.9 & 6223 & 5.4 & 2873 & 2.5 & 2938 & 2.5\\
- & Urban & 2271 & 2.0 & 3695 & 3.2 & 6709 & 5.8 & 10602 & 9.2 & 9937 & 8.6 & 8815 & 7.6 & 11622 & 10.1 & 12687 & 11.0 & 13291 & 11.5\\
-Beliefs & \_Disbeliefs\_ & 625 & 0.5 & 1312 & 1.1 & 2235 & 1.9 & 3639 & 3.2 & 0 & 0.0 & 3278 & 2.8 & 4420 & 3.8 & 4135 & 3.6 & 4514 & 3.9\\
- & \_SpiritExcludesGod\_ & 778 & 0.7 & 1508 & 1.3 & 2535 & 2.2 & 4206 & 3.6 & 0 & 0.0 & 3542 & 3.1 & 4808 & 4.2 & 4148 & 3.6 & 4309 & 3.7\\
- & GodAndSpirit & 1751 & 1.5 & 2560 & 2.2 & 4181 & 3.6 & 6201 & 5.4 & 0 & 0.0 & 5780 & 5.0 & 7002 & 6.1 & 5771 & 5.0 & 6782 & 5.9\\
- & GodExcludesSpirit & 228 & 0.2 & 274 & 0.2 & 593 & 0.5 & 1137 & 1.0 & 0 & 0.0 & 523 & 0.5 & 1094 & 0.9 & 950 & 0.8 & 666 & 0.6\\
-\bottomrule
-\end{tabular}
-}
-\caption{Sample statistics by year}
-\end{center}
-\end{table}
+Table 1 about here.
 
-## Theoretical indicators 
-### Life Satisfaction
-Life satisfaction is a measure of emotional well-being that was assessed using a 2-item version of the Satisfaction With Life Scale, which has previously been shown to correlate with aspects of religiosity [@Diener1985-xy].
-Participants rated their agreement with the statements (a) "I am satisfied with my life"; and (b) "In most ways my life is close to ideal." The items were rated on 7-point response options ranging from 1 = strongly disagree to 7 = strongly agree.
-The mean Cronbach’s alpha for this 2-item scale was $\alpha$ = 0.90.
-Higher scores on this scale indicate higher life satisfaction.
-Overall scores were means of the 2-items.
-
-## Sample Demographics
-
-### Age 
-Age was put into units of 10 years and centered at its mean. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 2.
-Gender. Gender was assessed by asking participants if they were "Male" was coded as "1" and "Female" was coded as "0."
-
-### Education
-Education level was measured using an 11-point rating developed by the New Zealand Qualification Authority known as the New Zealand Qualification Framework (NZQF; 0 = no qualification, 10 = doctoral degree).
-Education was centered at its mean and standardized.
-For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 1.
-
-### Deprivation
-We measured the socioeconomic status of participants’ immediate (small area) neighborhood using the 2013 New Zealand Deprivation Index, which uses aggregate census information about the residents of small neighborhood-type units to assign a decile-rank index from 1 (most affluent) to 10 (most impoverished) [@Atkinson2014-ex].
-The index is based on a Principal Components Analysis of the following nine variables (in weighted order): the proportion of adults who received a means-tested benefit, household income, proportion not owning own home, proportion single-parent families, proportion unemployed, proportion lacking qualifications, proportion household crowding, proportion no telephone access, and proportion no car access.
-Thus, the index reflects the average level of deprivation for small neighborhood-type units (or small community areas of about 80–90 people each) across the entire country.
-Our sample had a mean deprivation index of 4.80 (SD = 2.79).
-Deprivation was centered at its mean and standardized.
-For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 1.
-
-### Employed
-Employment status was assessed by asking participants if they were currently working, "yes" was coded as "1" and "no" was coded as "0." For detailed information pertaining to yearly responses and missingness, see Table 1.
-
-### Partner
-Participants were asked if they were in a relationship, "yes" was coded as "1" and no was coded as "0." For detailed information pertaining to yearly responses and missingness, see Table 1.
-
-### Ethnicity
-Ethnicity was assessed using four basic categories: (1) New Zealand European/Pakeha, (2) Maori, (3) Pacific Islander, and (4) Asian.
-For detailed information pertaining to yearly responses and missingness, see Table 1.
-
-### Urban
-People were coded as either residing in an urban "1" or rural "0" area based on New Zealand census data. For detailed information pertaining to yearly responses and missingness, see Table 1.
-
-### Political Orientation
-To assess political orientation, we asked people to rate their political orientation using seven-point response options (1 = Liberal; 7 = Conservative).
-Higher values indicate more conservative political beliefs.
-Political Orientation was standardized and centered at its mean.
-For detailed information pertaining to means, standard deviations, and missingness, see Table 1.
-
-### Personal well-being 
-Personal well-being is a measure of function all well-being that was assessed using a 4-item version of the Australian Unity well-being Index [@Cummins2017-ur].
-Participants rated their satisfaction with (a) "Your standard of living"; (b) "Your health"; (c) "Your future security"; (d) "Your personal relationships." Items were rated using a 10-point response option ranging from 1 = completely dissatisfied to 10 = completely satisfied.
-The mean Cronbach’s alpha for this scale was $\alpha$ = 0.90.
-Higher scores on this scale indicate higher personal well-being.
-Overall scores were means of the 4-items.
-
-### Belief in a Spirit/Life Force and/or a God  
-We assess belief in a spirit or a God by asking two non-affective questions: "Do you believe in a spirit or Life Force" and "Do you believe in a god?" Responses were coded as 1) **God and Spirit Disbelievers**: those who neither believe in a God nor believe in a Spirit or Life Force; (2) **Spirit Believers/God Disbelievers**: those who disbelieve in a God and believe in a Spirit or Life Force; (3) **God Believers/Spirit Disbelievers**: those who believe in a God but do not believe in a Spirit or Life Force; (4) **God and Spirit Believers**: those who believe in a both God and a Spirit or Life Force.
-
-## Statistical Analyses
-Statistical analysis was performed using R version 4.0.2 R version 3.6.1 (2020-06-22) 2019-07-05).
-We analyze data from participants who responded to the NZAVS survey at least three times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of  21,705 (*N* = 20,979 with complete responses).
-To model growth in subjective well-being we use generalized linear mixed models.
-In our models, we account for the non-linear effects of time within participants by including the interaction of years (a continuous variable) with Beliefs (a four-level categorical indicator).
-The baseline for this indicator is disbelief in a Spirit/Life Force and disbelief in a God.
-To handle the dependencies introduced from the repeated measures, we include individual ID as an effect modeled as random, and we estimated a random effect for the slope for years.
-This allowed us to model correllation between individual-level intercept (starting high or low in Personal Well-Being/Life Satisfaction) and stability/growth/decline in the slope effect of years.
-We estimated all statitiscal models using both REML estimation [@R-lme4] and Bayesian estimation [@BRMSpackage] with weakly informative priors.
-We include the full reports in the Supplement.
-The REML estimates and the Bayesian estimates converged to within three decimal places.
-We include model equations and priors at the end of this article.
-Here we report the results of the Bayesian estimator.
-
-## Potential Confounders
-
-To address confounding we first obtained the New Zealand Attitudes and Values Study standard demographic indicators recommended to address confounding: [NZAVS modelling standards](https://cdn.auckland.ac.nz/assets/psych/about/our-research/nzavs/NZAVSTechnicalDocuments/NZAVS-Technical-Documents-e11-Default-Statistical-Standards.pdf).
-Next we modelled potential relationships between the coefficients within the model using the daggity package in R, and plotting results using the daggity and ggdag packages in R.
-
-This method reveals two adjustments sets equired to avoid potential confounding: \{age, employment, male, political conservativism, and time\}, and \{age, ethnicty, male, and time \}.Here, we report the second set because employment is likely to contain more measurement error: some are not employed because they cannot find employment, others because they are not seeking employment (e.g.
-retirement).
-Results were insensitive to the choice set.
-Indeed, the inference proved robust to the simple model with only the interaction of years and beliefs, and robust to the fully saturated model in which all covariates were included (see the online supplement for details).
-In the fully saturated model, it appeared that those disbelieving in a Spirit or Life Force, exhibited no growth in Persona Well-Being and Life-Satisifaction over the nine years.
-Although this result offers stronger stronger evidence for a Spirit/Life Force well-being mechanism, it is likely the effect of collider confounding and masking.
-We discuss this point in more detail in the supplement, and mention it here to caution researchers from "everything-but-the-kitchen-sink" approaches to regression modelling.
-
-## Missing Data 
-Because missing values can result in biased estimates [@Blackwell2017-oq], we multiply imputed 10 datasets using the AmeliaII package in R[@Honaker2011-yu].We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences.
-Additionally the results were robust to filtering of participants of N > 3 to N = 8 waves.
-We include the multiple-imputation results in a supplement.
-Here, there were no differences in estimation relevant to inference, however, to improve inference, we advocate wider use of multiple-imputation or Bayesian approaches for handling missingness.
-
-## System and packages
-The analysis was performed using R version 4.0.2 (2020-06-22). The Platform was x86_64-apple-darwin17.0 (64-bit) Running under: macOS Catalina 10.15.6
-We are greatful to the contributors and mantainers of the following packages:
-R [Version 4.0.3; @R-base] and the R-packages *Amelia* [Version 1.7.6; @R-Amelia], *citr* [Version 0.3.2; @R-citr], *dplyr* [Version 1.0.2; @R-dplyr], *equatiomatic* [Version 0.1.0.9000; @R-equatiomatic], *forcats* [Version 0.5.0; @R-forcats], *ggeffects* [Version 1.0.0; @R-ggeffects], *gghighlight* [Version 0.3.0; @R-gghighlight], *ggplot2* [Version 3.3.2; @R-ggplot2], *ggsci* [Version 2.9; @R-ggsci], *lme4* [Version 1.1.25; @R-lme4], *Matrix* [Version 1.2.18; @R-Matrix], *papaja* [Version 0.1.0.9997; @R-papaja], *parameters* [Version 0.9.0.1; @R-parameters], *patchwork* [Version 1.1.0; @R-patchwork], *prettycode* [@R-prettycode], *purrr* [Version 0.3.4; @R-purrr], *Rcpp* [Version 1.0.5; @R-Rcpp_a; @R-Rcpp_b], *readr* [Version 1.4.0; @R-readr], *report* [Version 0.2.0; @R-report], *see* [Version 0.6.0.1; @R-see], *sjPlot* [Version 2.8.6; @R-sjPlot], *stringr* [Version 1.4.0; @R-stringr], *styler* [Version 1.3.2; @R-styler], *table1* [Version 1.2.1; @R-table1], *texreg* [Version 1.37.5; @R-texreg], *tibble* [Version 3.0.4; @R-tibble], *tidyr* [Version 1.1.2; @R-tidyr], and *tidyverse* [Version 1.3.0; @R-tidyverse]
-
-## Code
-All code for this analysis and a full report of the missing data imputation can be found at [https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing](https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing).
-An html document decribing the same analysis is located at: [https://osf.io/cu2gr/](https://osf.io/cu2gr/).
-
-# Results
-
-Table 2 about here:
-
+```{=tex}
 \begin{table}
 \begin{center}
-\scalebox{0.6}{
+\scalebox{0.8}{
 \begin{tabular}{l c c}
@@
 \midrule
-Intercept                           & $\mathbf{7.14}$   & $\mathbf{5.29}$   \\
-                                    & $ [ 7.11;  7.18]$ & $ [ 5.26;  5.31]$ \\
-yearsC                              & $\mathbf{0.02}$   & $\mathbf{0.01}$   \\
-                                    & $ [ 0.01;  0.02]$ & $ [ 0.00;  0.01]$ \\
-Beliefs\_SpiritExcludesGod\_        & $\mathbf{-0.06}$  & $0.01$            \\
-                                    & $ [-0.09; -0.03]$ & $ [-0.01;  0.03]$ \\
-BeliefsGodAndSpirit                 & $\mathbf{-0.06}$  & $\mathbf{0.03}$   \\
-                                    & $ [-0.09; -0.03]$ & $ [ 0.01;  0.06]$ \\
-BeliefsGodExcludesSpirit            & $-0.00$           & $\mathbf{0.06}$   \\
-                                    & $ [-0.04;  0.04]$ & $ [ 0.03;  0.09]$ \\
-Age\_in\_Decades\_C                 & $\mathbf{0.07}$   & $\mathbf{0.08}$   \\
-                                    & $ [ 0.06;  0.09]$ & $ [ 0.07;  0.09]$ \\
-Ethnic\_CategoriesMaori             & $\mathbf{-0.38}$  & $\mathbf{-0.11}$  \\
-                                    & $ [-0.43; -0.33]$ & $ [-0.14; -0.08]$ \\
-Ethnic\_CategoriesPacific           & $\mathbf{-0.38}$  & $\mathbf{-0.15}$  \\
-                                    & $ [-0.48; -0.29]$ & $ [-0.22; -0.09]$ \\
-Ethnic\_CategoriesAsian             & $\mathbf{-0.09}$  & $-0.05$           \\
-                                    & $ [-0.18; -0.01]$ & $ [-0.11;  0.02]$ \\
-Male1                               & $\mathbf{-0.06}$  & $\mathbf{-0.16}$  \\
-                                    & $ [-0.10; -0.02]$ & $ [-0.19; -0.13]$ \\
-yearsC:Beliefs\_SpiritExcludesGod\_ & $\mathbf{0.01}$   & $\mathbf{0.01}$   \\
-                                    & $ [ 0.00;  0.02]$ & $ [ 0.00;  0.02]$ \\
-yearsC:BeliefsGodAndSpirit          & $\mathbf{0.01}$   & $\mathbf{0.01}$   \\
-                                    & $ [ 0.01;  0.02]$ & $ [ 0.00;  0.02]$ \\
-yearsC:BeliefsGodExcludesSpirit     & $0.00$            & $-0.00$           \\
-                                    & $ [-0.01;  0.02]$ & $ [-0.01;  0.01]$ \\
+Intercept                            & $\mathbf{7.24}$   & $\mathbf{5.36}$   \\
+                                     & $ [ 7.18;  7.30]$ & $ [ 5.31;  5.40]$ \\
+BeliefsGodAndSpirit                  & $\mathbf{-0.08}$  & $-0.01$           \\
+                                     & $ [-0.13; -0.03]$ & $ [-0.05;  0.02]$ \\
+BeliefsGodExcludesSpirit             & $-0.04$           & $0.01$            \\
+                                     & $ [-0.10;  0.03]$ & $ [-0.04;  0.06]$ \\
+BeliefsSpiritExcludesGod             & $\mathbf{-0.06}$  & $-0.02$           \\
+                                     & $ [-0.10; -0.01]$ & $ [-0.05;  0.01]$ \\
+Age\_within                          & $\mathbf{0.02}$   & $\mathbf{0.01}$   \\
+                                     & $ [ 0.01;  0.03]$ & $ [ 0.01;  0.02]$ \\
+Age\_betweenC                        & $\mathbf{0.01}$   & $\mathbf{0.01}$   \\
+                                     & $ [ 0.01;  0.01]$ & $ [ 0.01;  0.01]$ \\
+Ethnic\_CategoriesMaori              & $-0.08$           & $0.04$            \\
+                                     & $ [-0.16;  0.00]$ & $ [-0.02;  0.10]$ \\
+Ethnic\_CategoriesPacific            & $\mathbf{-0.39}$  & $\mathbf{-0.29}$  \\
+                                     & $ [-0.54; -0.24]$ & $ [-0.39; -0.18]$ \\
+Ethnic\_CategoriesAsian              & $-0.05$           & $-0.04$           \\
+                                     & $ [-0.20;  0.13]$ & $ [-0.15;  0.09]$ \\
+Male1                                & $-0.05$           & $\mathbf{-0.11}$  \\
+                                     & $ [-0.13;  0.04]$ & $ [-0.16; -0.05]$ \\
+BeliefsGodAndSpirit:Age\_within      & $\mathbf{0.02}$   & $\mathbf{0.01}$   \\
+                                     & $ [ 0.01;  0.03]$ & $ [ 0.01;  0.02]$ \\
+BeliefsGodExcludesSpirit:Age\_within & $0.01$            & $0.00$            \\
+                                     & $ [-0.01;  0.03]$ & $ [-0.01;  0.02]$ \\
+BeliefsSpiritExcludesGod:Age\_within & $\mathbf{0.02}$   & $\mathbf{0.01}$   \\
+                                     & $ [ 0.01;  0.03]$ & $ [ 0.00;  0.02]$ \\
 \midrule
-SD: Idsd(Intercept)                 & $1.42$            & $0.98$            \\
-SD: Idsd(yearsC)                    & $0.13$            & $0.10$            \\
-SD: Idcor(Intercept,yearsC)         & $-0.03$           & $0.03$            \\
+SD: Idsd(Intercept)                  & $1.37$            & $0.96$            \\
+SD: Idsd(Age\_within)                & $0.10$            & $0.08$            \\
+SD: Idcor(Intercept,Age\_within)     & $-0.05$           & $-0.01$           \\
 \bottomrule
@@
 \end{table}
+```
+## Theoretical indicators
 
+### Life Satisfaction
+
+Life satisfaction is a measure of emotional well-being that was assessed using a 2-item version of the Satisfaction With Life Scale, which has previously been shown to correlate with aspects of religiosity [@Diener1985-xy].
+Participants rated their agreement with the statements (a) "I am satisfied with my life"; and (b) "In most ways my life is close to ideal." The items were rated on 7-point response options ranging from 1 = strongly disagree to 7 = strongly agree.
+The mean Cronbach's alpha for this 2-item scale was $\alpha$ = 0.90.
+Higher scores on this scale indicate higher life satisfaction.
+Overall scores were means of the 2-items.
+
+## Sample Demographics
+
+### Age
+
+Age was put into units of 10 years and centered at its mean.
+For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 2.
+Gender.
+Gender was assessed by asking participants if they were "Male" was coded as "1" and "Female" was coded as "0."
+
+### Education
+
+Education level was measured using an 11-point rating developed by the New Zealand Qualification Authority known as the New Zealand Qualification Framework (NZQF; 0 = no qualification, 10 = doctoral degree).
+Education was centered at its mean and standardized.
+For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 1.
+
+### Deprivation
+
+We measured the socioeconomic status of participants' immediate (small area) neighborhood using the 2013 New Zealand Deprivation Index, which uses aggregate census information about the residents of small neighborhood-type units to assign a decile-rank index from 1 (most affluent) to 10 (most impoverished) [@Atkinson2014-ex].
+The index is based on a Principal Components Analysis of the following nine variables (in weighted order): the proportion of adults who received a means-tested benefit, household income, proportion not owning own home, proportion single-parent families, proportion unemployed, proportion lacking qualifications, proportion household crowding, proportion no telephone access, and proportion no car access.
+Thus, the index reflects the average level of deprivation for small neighborhood-type units (or small community areas of about 80--90 people each) across the entire country.
+Our sample had a mean deprivation index of 4.80 (SD = 2.79).
+Deprivation was centered at its mean and standardized.
+For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 1.
+
+### Employed
+
+Employment status was assessed by asking participants if they were currently working, "yes" was coded as "1" and "no" was coded as "0." For detailed information pertaining to yearly responses and missingness, see Table 1.
+
+### Partner
+
+Participants were asked if they were in a relationship, "yes" was coded as "1" and no was coded as "0." For detailed information pertaining to yearly responses and missingness, see Table 1.
+
+### Ethnicity
+
+Ethnicity was assessed using four basic categories: (1) New Zealand European/Pakeha, (2) Maori, (3) Pacific Islander, and (4) Asian.
+For detailed information pertaining to yearly responses and missingness, see Table 1.
+
+### Urban
+
+People were coded as either residing in an urban "1" or rural "0" area based on New Zealand census data. For detailed information pertaining to yearly responses and missingness, see Table 1.
+
+### Political Orientation
+
+To assess political orientation, we asked people to rate their political orientation using seven-point response options (1 = Liberal; 7 = Conservative).
+Higher values indicate more conservative political beliefs.
+Political Orientation was standardized and centered at its mean.
+For detailed information pertaining to means, standard deviations, and missingness, see Table 1.
+
+### Personal well-being
+
+Personal well-being is a measure of function all well-being that was assessed using a 4-item version of the Australian Unity well-being Index [@Cummins2017-ur].
+Participants rated their satisfaction with (a) "Your standard of living"; (b) "Your health"; (c) "Your future security"; (d) "Your personal relationships." Items were rated using a 10-point response option ranging from 1 = completely dissatisfied to 10 = completely satisfied.
+The mean Cronbach's alpha for this scale was $\alpha$ = 0.90.
+Higher scores on this scale indicate higher personal well-being.
+Overall scores were means of the 4-items.
+
+### Belief in a Spirit/Life Force and/or a God
+
+We assess belief in a spirit or a God by asking two non-affective questions: "Do you believe in a spirit or Life Force" and "Do you believe in a god?" Responses were coded as 1) **God and Spirit Disbelievers**: those who neither believe in a God nor believe in a Spirit or Life Force; (2) **Spirit Believers/God Disbelievers**: those who disbelieve in a God and believe in a Spirit or Life Force; (3) **God Believers/Spirit Disbelievers**: those who believe in a God but do not believe in a Spirit or Life Force; (4) **God and Spirit Believers**: those who believe in a both God and a Spirit or Life Force.
+
+## Statistical Analyses
+
+Statistical analysis was performed using R version 4.0.2 R version 3.6.1 (2020-06-22) 2019-07-05).
+We analyze data from participants who responded to the NZAVS survey at least nine times between 2010/11 (Time 2) and 2019/20 (Time 11), resulting in a sample of 21,705 ($N$ = 3,257 with complete responses).
+To model growth in subjective well-being we use generalized linear mixed models.
+To avoid heterogeneity bias, we "demeaned" the variable "age" separating this variable into the within and between person components.
+Heterogeneity bias arises when an indicators varies within and between groups, which will case fixed effects estimates to correlate with group-level effects [@bell2015explaining], which is a common feature of panel data (in which a participant is a level-2 indicator)[@R-parameters].
+We modelled the interaction of within participant aging and belief by including the interaction of aging within (a continuous variable, centered) and belief-state, a four-level categorical indicator.
+The baseline for this indicator is disbelief in a Spirit/Life Force and disbelief in a God -- or "disbeliever." To handle the dependencies introduced from the repeated measures, we include individual ID as an effect modeled as random, and we estimated a random effect for the slope for aging (within) This approach allowed us to model correlation between individual-level intercept (starting high or low in Personal Well-Being/Life Satisfaction) and stability/growth/decline in the slope effect of years.
+We estimated all statistical models using both REML estimation [@R-lme4] and Bayesian estimation [@BRMSpackage] with weakly informative priors, reported in the on-line supplement.
+The REML estimates, which we report in the supplement, converged to within three decimal places.
+The model equations and priors are in the on-line supplement.
+
+## Potential Confounders
+
+To address confounding we first obtained the New Zealand Attitudes and Values Study standard demographic indicators recommended to address confounding: [NZAVS modelling standards](https://cdn.auckland.ac.nz/assets/psych/about/our-research/nzavs/NZAVSTechnicalDocuments/NZAVS-Technical-Documents-e11-Default-Statistical-Standards.pdf).
+Next we modelled potential relationships between the coefficients within the model using the `ggdag` package in R.
+
+This method reveals two adjustments sets required to avoid potential confounding: \{age_within, age_between, employment, male, political conservativism, and time\}, and \{age_within, age_between, ethnicty, male, and time\}.
+Here, we used the second adjustment set because employment is likely to contain more measurement error: some are not employed because they cannot find employment, others because they are not seeking employment (e.g.
+retirement).
+Results were insensitive to the choice set.
+
+## Missing Data
+
+Because missing values can result in biased estimates [@Blackwell2017-oq], we multiply imputed 10 datasets using the AmeliaII package in R[@Honaker2011-yu].We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences.
+
+## System and packages
+
+The analysis was performed using R version 4.0.2 (2020-06-22).
+The Platform was x86_64-apple-darwin17.0 (64-bit) Running under: macOS Catalina 10.15.6 We are greatful to the contributors and mantainers of the following packages: R [Version 4.0.3; @R-base] and the R-packages *Amelia* [Version 1.7.6; @R-Amelia], *citr* [Version 0.3.2; @R-citr], *dplyr* [Version 1.0.2; @R-dplyr], *equatiomatic* [Version 0.1.0.9000; @R-equatiomatic], *forcats* [Version 0.5.0; @R-forcats], *ggeffects* [Version 1.0.0; @R-ggeffects], *gghighlight* [Version 0.3.0; @R-gghighlight], *ggplot2* [Version 3.3.2; @R-ggplot2], *ggsci* [Version 2.9; @R-ggsci], *lme4* [Version 1.1.25; @R-lme4], *Matrix* [Version 1.2.18; @R-Matrix], *papaja* [Version 0.1.0.9997; @R-papaja], *parameters* [Version 0.9.0.1; @R-parameters], *patchwork* [Version 1.1.0; @R-patchwork], *prettycode* [@R-prettycode], *purrr* [Version 0.3.4; @R-purrr], *Rcpp* [Version 1.0.5; @R-Rcpp_a; @R-Rcpp_b], *readr* [Version 1.4.0; @R-readr], *report* [Version 0.2.0; @R-report], *see* [Version 0.6.0.1; @R-see], *sjPlot* [Version 2.8.6; @R-sjPlot], *stringr* [Version 1.4.0; @R-stringr], *styler* [Version 1.3.2; @R-styler], *table1* [Version 1.2.1; @R-table1], *texreg* [Version 1.37.5; @R-texreg], *tibble* [Version 3.0.4; @R-tibble], *tidyr* [Version 1.1.2; @R-tidyr], and *tidyverse* [Version 1.3.0; @R-tidyverse]
+
+## Code
+
+All code for this analysis and a full report of the missing data imputation can be found at [https://github.com/go-bayes/nzavs_spiritbeliefWellbeing](https://github.com/go-bayes/nzavs_spiritbeliefWellbeing){.uri}.
+An html document decribing the same analysis is located at: <https://osf.io/cu2gr/>.
+
+# Results
+
+Table 2 about here:
+
+```{=tex}
+\begin{table}
+\begin{center}
+\scalebox{0.8}{
+\begin{tabular}{l c c}
+\toprule
+ & PWI & Life Sat \\
+\midrule
+Intercept                                & $\mathbf{6.61}$   & $\mathbf{4.85}$   \\
+                                         & $ [ 6.44;  6.79]$ & $ [ 4.73;  4.97]$ \\
+Beliefs\_SpiritExcludesGod\_             & $\mathbf{-0.06}$  & $-0.02$           \\
+                                         & $ [-0.10; -0.01]$ & $ [-0.05;  0.01]$ \\
+BeliefsGodAndSpirit                      & $\mathbf{-0.07}$  & $-0.01$           \\
+                                         & $ [-0.12; -0.02]$ & $ [-0.05;  0.02]$ \\
+BeliefsGodExcludesSpirit                 & $-0.03$           & $0.01$            \\
+                                         & $ [-0.11;  0.03]$ & $ [-0.04;  0.06]$ \\
+Age\_within                              & $\mathbf{0.02}$   & $\mathbf{0.01}$   \\
+                                         & $ [ 0.01;  0.03]$ & $ [ 0.01;  0.02]$ \\
+Age\_between                             & $\mathbf{0.01}$   & $\mathbf{0.01}$   \\
+                                         & $ [ 0.01;  0.01]$ & $ [ 0.01;  0.01]$ \\
+Ethnic\_CategoriesMaori                  & $-0.08$           & $0.04$            \\
+                                         & $ [-0.16;  0.00]$ & $ [-0.02;  0.09]$ \\
+Ethnic\_CategoriesPacific                & $\mathbf{-0.39}$  & $\mathbf{-0.29}$  \\
+                                         & $ [-0.53; -0.23]$ & $ [-0.40; -0.18]$ \\
+Ethnic\_CategoriesAsian                  & $-0.06$           & $-0.03$           \\
+                                         & $ [-0.23;  0.12]$ & $ [-0.16;  0.08]$ \\
+Male1                                    & $-0.05$           & $\mathbf{-0.10}$  \\
+                                         & $ [-0.13;  0.03]$ & $ [-0.17; -0.05]$ \\
+Beliefs\_SpiritExcludesGod\_:Age\_within & $\mathbf{0.02}$   & $\mathbf{0.01}$   \\
+                                         & $ [ 0.01;  0.03]$ & $ [ 0.00;  0.02]$ \\
+BeliefsGodAndSpirit:Age\_within          & $\mathbf{0.02}$   & $\mathbf{0.01}$   \\
+                                         & $ [ 0.01;  0.03]$ & $ [ 0.01;  0.02]$ \\
+BeliefsGodExcludesSpirit:Age\_within     & $0.01$            & $0.00$            \\
+                                         & $ [-0.01;  0.03]$ & $ [-0.01;  0.02]$ \\
+\midrule
+SD: Idsd(Intercept)                      & $1.37$            & $0.96$            \\
+SD: Idsd(Age\_within)                    & $0.10$            & $0.08$            \\
+SD: Idcor(Intercept,Age\_within)         & $-0.05$           & $-0.01$           \\
+\bottomrule
+\end{tabular}
+}
+\caption{}
+\label{tab:REGRESS_LS}
+\end{center}
+\end{table}
+```
 Figure 1 about here
+
 ```{r fig.cap = "Coefficient Graphs", echo =FALSE,out.width="\\textwidth"}
-knitr::include_graphics("/Users/jbul716/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/coefBayesDag.png")
+knitr::include_graphics("/Users/jbul176/GIT/nzavs_spiritbeliefWellbeing/coef.png")
 ```
 
-## Personal Well-Being
-We fitted a linear mixed effects model using the brms package [@BRMSpackage] in R [@R-base] to predict personal well-being (PWI) with years (0-9), beliefs (four categories), with Id as random intercept and years as a random slope effect (see below: model equations).
+## Personal well-being
+
+We fitted a linear mixed effects model using the brms package [@BRMSpackage] in R [@R-base] to predict personal well-being (PWI) with age_within centered (range = -5.79, 6.83), beliefs (four categories), with Id as random intercept and years as a random slope effect (see below: model equations).
 
@@
 
-  - The intercept, representing the expected personal well-being among disbelievers in a Spirit/Life Force and God who are Europeans, 50.77 years of age and do not identify as male is reliable (posterior median = 7.138 HDI [ 7.111,  7.167], pd = 100.00%).
-  - The effect of years is positive is reliable (posterior median =  7.138, HDI [ 7.111,  7.167], pd  100.00%).
-We infer that personal well-being increased in the baseline disbelieving population between 2010 and 2018.
-  - The effect of beliefs in *Spirit Excludes God*  is reliably negative (posterior median =  -0.059, HDI [-0.081, -0.036], pd 100.00%).
+  - The intercept, representing the expected personal well-being among disbelievers in a Spirit/Life Force and God who are Europeans, 50.77 years of age and do not identify as male is reliable (posterior median = 6.611  HDI [ 6.440,  6.780], pd = 100.00%).\
+  - The effect of aging (years) is positive is reliable (posterior median =  0.021, HDI [ 0.014,  0.029], pd 100.00%).
+We infer that personal well-being increased in the baseline disbelieving population between 2010 and 2020.
+  - The effect of beliefs in *Spirit Excludes God* is reliably negative (posterior median = -0.058, HDI [-0.097, -0.015], pd  98.90%).
 We infer that the population that believed in a Spirit but did not believe in a God was lower in personal well-being at the intercept compared to disbelievers.
-  - The effect of beliefs in *God And Spirit*  is not  negative (posterior median = -0.060, HDI [-0.086, -0.034], pd = 100.00%) .
+  - The effect of beliefs in *God And Spirit* is negative (posterior median =-0.075 , HDI [-0.124, -0.024], pd =  99.35%) .
 We infer that the population that believed in a God and a Spirit or Life Force was lower in personal well-being at the intercept compared to disbelievers.
-  - The effect of beliefs in *God Excludes Spirit*  is not reliable (posterior median = -0.001 HDI [-0.036,  0.035], pd = 51.10%).
-We infer that the population that believed in a God and doubted a Spirit or Life Force were similar were similar in personal well-being at the intercept to the disbelieving population:
-  - The effect of Age in Decades (C) is   0.072 | [ 0.060,  0.083] | 100.00% |
-  - The effect of Maori Ethnicity is  -0.380 | [-0.420, -0.341] | 100.00% | 
-  - The effect of Pacific Ethnicity is  -0.384 | [-0.464, -0.312] | 100.00% |
-  - The effect of Asian Ethnicity is -0.093 | [-0.165, -0.023] |  98.28% |  
-  - The effect of Male gender is  -0.060 | [-0.094, -0.026] |  99.90% |
+  - The effect of beliefs in *God Excludes Spirit* is not reliable (posterior median = -0.035 HDI [-0.107,  0.029], pd = 78.50%).
+We infer that the population that believed in a God and doubted a Spirit or Life Force were similar in personal well-being at the intercept to the disbelieving population:
 
-## Personal Well-Being: years X beliefs
+## Personal well-being: aging (years) X beliefs
 
 ```{r fig.cap = "Personal Well-being prediction graph", echo =FALSE, out.width="\\textwidth"}
-knitr::include_graphics("/Users/jbul716/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/pwi_confounder_adjusted.png")
+knitr::include_graphics("/Users/jbul176/GIT/nzavs_spiritbeliefWellbeing/pwi.png")
 ```
 
-The effect of years $\times$ belief in a *Spirit Excludes God*  is reliably positive  (posterior median = 0.011, HDI [ 0.003,  0.018], pd = 99.02%).
+  - The effect of aging (years) $\times$ belief in a *Spirit Excludes God* is reliably positive (posterior median = 0.020, HDI [  0.010,  0.030], pd =  99.90%).
 We infer that people with beliefs in a spirit or Life Force who do not believe in a God tended to grow in their personal well-being over time at a faster rate than disbelievers grew.
-  - The effect of years $\times$ in *God And Spirit*  is reliably positive  (posterior median = 0.014, HDI [0.007, 0.021], pd = 99.98%).
+  - The effect of aging (years) $\times$ in *God And Spirit* is reliably positive (posterior median =0.022, HDI [0.012,  0.031], pd = 100.00%).
 We infer that people who believe in a spirit or Life Force who also believe in a God tended to grow in their personal well-being over time at a faster rate than disbelievers grew.
-  - The effect of years $\times$ belief in *God Excludes Spirit*  is not reliable (posterior median = 0.004, HDI [-0.009,  0.015], pd =  68.73%).
+  - The effect of aging (years) $\times$ belief in *God Excludes Spirit* is not reliable (posterior median = 0.010, HDI [-0.007,  0.026], pd = 83.40%).
 We infer that people who believe in a God but not a spirit or Life Force did not grow at a faster rate in personal well-being than disbelievers grew.
@@
 
-We fitted a linear mixed effects model using the brms package [@BRMSpackage] in R [@R-base] to predict personal well-being (PWI) with years (0-9), beliefs (four categories), with Id as random intercept and years as a random slope effect (see model equations, below).
+We fitted a linear mixed effects model using the brms package [@BRMSpackage] in R [@R-base] to predict Life Satisfaction (LS) with age_within centered (range = -5.79, 6.83), beliefs (four categories), with Id as random intercept and years as a random slope effect (see below: model equations).
 
 Within this model: 
-  - The intercept, representing the expected life satisfaction among disbelievers in a Spirit/Life Force and God who are Europeans, 50.77 years of age and do not identify as male is reliable (posterior median = 5.287, HDI [ 5.267,  5.306], pd = 100.00%).
-  - The effect of years is positive is reliable (posterior median =  0.008, HDI [ 0.003, 0.011], pd = 99.75%).
+  - The intercept, representing the expected life satisfaction among disbelievers in a Spirit/Life Force and God who are Europeans, 50.77 years of age and do not identify as male is reliable (posterior median = 6.611, HDI [6.440,  6.780], pd = 100.00%).
+  - The effect of aging (years) is positive is reliable (posterior median = 0.021, HDI [0.014, 0.029], pd = 100.00%).
 We infer that life satisfaction increased in the disbelieving population between 2010 and 2018.
-  -The effect of beliefs in *Spirit Excludes God*  is not reliable (posterior median =   0.007, HDI [-0.010,  0.023], pd = 75.33%).
-We infer that the population that believed in a Spirit but did not believe in a God was higher in life satisfaction at the intercept compared with disbelievers.
-  - The effect of beliefs in *God And Spirit*  is reliably positive (posterior median = 0.034, HDI [ 0.015, 0.053], pd = 99.83%).
-We infer that the population that believed in a God and a Spirit or Life Force was higher in life satisfaction at the intercept compared with disbelievers.
-  - The effect of beliefs in *God Excludes Spirit*  is reliably positive (posterior median =  0.057, HDI [ 0.030, 0.082], pd = 99.95%).
-We infer that the population that believed in a God and doubted a Spirit or Life Force  was higher in life satisfaction at the intercept compared with disbelievers.
-  - The effect of Age in Decades (C) is reliably positive (posterior median = 0.079, hDI [ 0.071,  0.087] pd = 100.00%).
-  - The effect of Maori Ethnicity is reliably negative (posterior median = -0.109 | [-0.137, -0.082], pd = 100.00%).
-  - The effect of Pacific Ethnicity is reliably negative (posterior median = -0.153 | [-0.210, -0.103], pd = 100.00%).
-  - The effect of Asian Ethnicity is not reliable (posterior median = -0.047, [-0.098,  0.004], pd = 93.08%).
-  - The effect of Male gender is reliabily negative (posterior median = -0.161, [-0.184, -0.138], pd = 100.00%)
+  -The effect of beliefs in *Spirit Excludes God* is negative  (posterior median = -0.058, HDI [-0.097, -0.015], pd = 98.90%).
+We infer that the population that believed in a Spirit but did not believe in a God was lower in life satisfaction at the intercept compared with disbelievers.
+  - The effect of beliefs in *God And Spirit* is reliably negative (posterior median = -0.075, HDI [ -0.124, -0.024], pd =  99.35%).
+We infer that the population that believed in a God and a Spirit or Life Force was lower in life satisfaction at the intercept compared with disbelievers.
+  - The effect of beliefs in *God Excludes Spirit* is not reliable  (posterior median = -0.035, HDI [-0.107,  0.029], pd = 78.50%).
+We infer that the population that believed in a God and doubted a Spirit or Life Force not reliably different in life satisfaction at the intercept compared with disbelievers.
 
-## Life Satisfaction: years X beliefs
+## Life Satisfaction: aging (years) X beliefs
 
 ```{r  fig.cap = "Life satisfaction prediction graph", echo = FALSE,out.width="\\textwidth"}
-knitr::include_graphics("/Users/jbul716/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/ls_counfounder_adjusted.png")
+knitr::include_graphics("/Users/jbul176/GIT/nzavs_spiritbeliefWellbeing/ls.png")
 ```
-  - The interaction of years $\times$ beliefs in *Spirit Excludes God* is reliably positive (posterior median = 0.011, HDI [ 0.006, 0.016], pd = 99.95%).
+
+  - The interaction of aging (years) $\times$ beliefs in *Spirit Excludes God* is reliably positive (posterior median =  0.020 , HDI [0.010,  0.030], pd = 99.90%).
 We infer that people with beliefs in a Spirit or Life Force who do not believe in a God tended to grow over time in their life satisfaction at a faster rate than disbelievers.
-  - The interaction of years $\times$ *God And Spirit*  is reliably positive (posterior median =  0.009, HDI [ 0.004, 0.014], pd = 99.72%).
+  - The interaction of aging (years) $\times$ *God And Spirit* is reliably positive (posterior median = 0.022, HDI [0.012,  0.031], pd = 100.00%).
 We infer that people who believe in a spirit or Life Force who also believe in a God tended to grow over time in their life satisfaction at a faster rate than disbelievers.
-  - The effect of years $\times$ beliefs in *God Excludes Spirit*  is not reliable (posterior median = -0.003, HDI [-0.012, 0.006], pd = 72.28%).
-We infer that people who believe in a God but not a spirit or Life Force tended to grow over time in their life satisfaction at a similar rate to disbelievers.
+  - The effect of years $\times$ beliefs in *God Excludes Spirit* is not reliable (posterior median = 0.010, HDI [-0.007,  0.026], pd = 83.40%).
+We infer that people who believe in a God but not a spirit or Life Force did no experience reliably different growth in life satisfaction compared with disbelievers.
 
@@
 1.
-At the outset of the study, in 2010, the expected subjective well-being as measured using the personal well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God ($\hat{y} =$  6.93 | [6.88, 6.98]) and lower among those who believed in both a Spirit/Life Force and a God ($\hat{y}$ = 6.91 | [6.87, 6.95]) in comparison both to the population who neither believed in a God nor Spirit/ Life Force ($\hat{y}$ = 7.04 | [6.99, 7.10]) and to the population that believed in a God but was disbelieving in a Spirit or Life Force ($\hat{y}$ 7.03 | [6.95, 7.10]).
-(See Figure 2)  Similarly,the expected subjective well-being as measured using the life satisfaction Inventory was lower in 2010 among those who believed in Spirit or Life Force but not a God ($\hat{y}$ =   5.20 | [5.16, 5.23]).
-Life satisfaction was similar among those who believed in both a Spirit/Life Force and a God  ($\hat{y}$ =   5.23 | [5.21, 5.26]) in comparison both with the disbelieving population who neither believed in a God nor a Life Force ($\hat{y}$ = 5.25 | [5.21, 5.29]).
-Life satisfaction was highest among the population that believed in a God but was disbelieving in a Spirit or Life Force ($\hat{y}$ 5.32 | [5.26, 5.38]) (see Figure 3)
+In 2010 the expected subjective well-being as measured using the personal well-being inventory was expected to be lower among those who believed in Spirit or Life Force but not a God ($\hat{y} =$  6.95 \| [6.86, 7.03]), lower among those who believed in both a Spirit/Life Force and a God ($\hat{y}$ = 6.91 \| [6.83, 6.99]), and lower among the population that believed in a God but was disbelieving in a Spirit or Life Force 6.95 \| [6.86, 7.03] when compared with the population of disbelievers ($\hat{y}$ = 7.12 \| [7.03, 7.21]) and to ($\hat{y}$ 7.03 \| [6.95, 7.10]).
+(See Figure 2).
+
+Focusing on life satisfaction, people who believed in Spirit, either with a belief in God ($\hat{y}$ = 5.18 \| [5.13, 5.24]) or without a belief in God (5.18 \| [5.12, 5.24]) tended to be lower in life satsifaction than disbelievers in both a God and a Spirit (5.27 \| [5.20, 5.33]) and disbelievers in a spirit who also believed in a God (5.26 \| [5.15, 5.36]) (see Figure 3)
 
 2.
-However, over time, belief in a Spirit or Life Force predicts a trajectory of growth in personal well-being.
-In 2018, personal well-being among who believe in Spirit or Life Force but not a God had grown to ($\hat{y} =   7.19$ [7.15, 7.23], $\Delta \hat{y} = 0.26$).
-Among those who believed in both a Spirit/Life Force and a God, personal well-being had grown to ($\hat{y} = 7.20$ [7.17, 7.24], $\Delta \hat{y} = 0.29$).
-By contrast, personal well-being among the disbelieving population who neither believed in a God nor a Life Force grew less steeply across the eight year interval $\hat{y} =  7.21$ [7.17, 7.25], $\Delta \hat{y} = 0.17$).
-By the same token, personal well-being among those who believe in a God but are disbelieving a Spirit/Life Force grew less steeply $\hat{y} =  7.22$ [7.16, 7.28] $\Delta \hat{y} = 0.19$).
-In practical terms, the rate of increase in personal wellbeing among those with Spirit/Life Force beliefs was about 30% greater than it was among those who lacked such Spirit/Life Force (See Figure 2).
-Similarly, over time, belief in a Spirit or Life Force predicts a trajectory of growth in life satisfaction.
-In 2018, the expected effect of spiritual beliefs on life satisfaction among those who believe in Spirit or Life Force but not a God had grown to ($\hat{y} =  5.37$ [5.34, 5.40], $\Delta \hat{y} = 0.17$).
-Among those who believed in both a Spirit/Life Force and a God, life satisfaction had grown to ($\hat{y} = 5.39$  [5.36, 5.41], $\Delta \hat{y} = 0.19$).By contrast, life satisfaction among the population who neither believed in a God nor a Life Force lagged behind over the eight year interval $\hat{y} =   5.32$ [5.29, 5.35], $\Delta \hat{y} = 0.07$).
-Relatedly, life satisfaction in the population that believe in God but are disbelieving in a Spirit/Life Force grew less steeply $\hat{y} =  5.36$ [5.31, 5.41], $\Delta \hat{y} = 0.04$).
-In practical terms, the rate of increase in life satisfaction among those with Spirit/Life Force beliefs over three times greater than among those that lacked such beliefs (See Figure 3).
+Over a decade, people who believed in a God and Spirit ($\Delta \hat{y}$ = 0.55), people who believed in God and disbelieved in a Spirit ($\Delta \hat{y}$ = 0.40), and people who believed in a spirit and disbelieved in a God ($\Delta \hat{y}$ = 0.51) grew at up to twice the rate in personal well-being that disbelievers grew ($\Delta \hat{y}$ = 0.26).
 
-3.
-Inferred growth from spiritual beliefs is evident from 9 annual waves of data collection (8 years), and across a nationally diverse sample.
-On this basis of this longitudinal information accumulated across nationally, we are confident that growth in personal well-being and in life-satisfaction among those who believe in a Spirit and Life Force are evident.
-Such effects co-occur in people who lack a belief in a God as well as among people who believe in a God as well as a Spirit or Life Force.
+Similarly, belief in a Spirit or Life Force predicts a trajectory of growth in life satisfaction over the decade (2010-2020); 
+Over a decade, people who believed in a God and Spirit ($\Delta \hat{y}$ = .35), people who believed in God and disbelieved in a Spirit ($\Delta \hat{y}$ = .24), and people who believed in a spirit and disbelieved in a God ($\Delta \hat{y}$ = 0.33) grew at up to twice the rate in personal well-being that disbelievers grew ($\Delta \hat{y}$ = 0.18).
+Here to we find a growth rate that is twice as large for people who believe in a spirit or life force.
+
+Because These inferred growth rates for personal well being and life satisfaction are evident from repeated measures across 10 annual waves of nationally diverse data collection, we are confident that in New Zealand, spiritual beliefs predict strong increases in personal well-being and life satisfaction
 
@@
 ## Importance
+
 Despite its limitations, our study is important both for its methods and its findings.
@@
 ### Authors contributions
-BH and JB conceived of the study.
-BH wrote the first draft of the manuscript.
-EW improved the approach with BH and JB and revised the manuscript.
-DD revised the manuscript.
-CS was responsible for NZAVS data collection and revised the manuscript.
-JB performed the data re-analysis and revised the manuscript.
+
+BH and JB conceived of the study. BH wrote the first draft of the manuscript. EW improved the approach. JB did the reanalysis. BH, DD, CS, EW, and JB revised the manuscript.
 
 ### Funding statement
+
 The New Zealand Attitudes and Values Study is supported by a grant from the Templeton Religion Trust (TRT0196). The funders had no role in the preparation of the article or the decision to publish.
@@
 
+```{=tex}
 \begingroup
@@
 \setlength{\leftskip}{0.5in}
+```
+::: {#refs custom-style="Bibliography"}
+:::
 
-<div id="refs" custom-style="Bibliography"></div>
 \endgroup

### 2021-02-26 Manuscript_Bayesian.tex
fixes

--- 2020-11-29 Manuscript_Bayesian.tex
@@
 \usepackage{lmodern}
-\usepackage{amssymb,amsmath}
+\usepackage{amsmath}
 \usepackage{ifxetex,ifluatex}
@@
   \usepackage{textcomp} % provide euro and other symbols
+  \usepackage{amssymb}
 \else % if luatex or xetex
@@
 \urlstyle{same} % disable monospaced font for URLs
-\usepackage{graphicx,grffile}
+\usepackage{graphicx}
 \makeatletter
@@
 \shorttitle{Longitudinal Wellbeing from Spiritual Beliefs}
-\keywords{Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being\newline\indent Word count: 5181 words in text body, 1191 words in reference section}
+\keywords{Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being\newline\indent Word count: 4760 words in text body, 1191 words in reference section}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 \fi
+\ifluatex
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
+  % set entry spacing
+  \ifnum #2 > 0
+  \setlength{\parskip}{#2\baselineskip}
+  \fi
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 However, the widespread use of tautological spirituality scales, poorly defined concepts of spirituality, and heavy reliance on cross-sectional samples cast doubts.
-Here, we leverage nine waves of panel data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction (\(N\) = 20979, New Zealand, 2010-2018).
-Contrary to previous research, we find that belief in a spirit or Life Force predicts lower personal well-being and life-satisfaction.
+Here, we leverage ten waves of panel data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction (\(N\) = 3,257, New Zealand, 2010-2020).
+Contrary to previous research, we find that belief in a Spirit or Life Force predicts lower personal well-being and life-satisfaction.
 However, in support of previous speculation, beliefs in a spirit or Life Force predict increasing personal well-being and life satisfaction over time relative to disbelief.
@@
 The relationship between spirituality and psychological well-being is a matter of enduring fascination (Ellsworth \& Ellsworth, 2010).
-Most previous attempts to quantify the relationship between these two domains find a positive association: spirituality predicts well-being (Ginsburg, Quirt, Ginsburg, \& MacKillop, 1995; Koenig, 2010; Koenig, Associate Professor of Psychology and Religious Studies Michael E McCullough, PhD, McCullough, Larson, \& Adjunct Professor of Psychiatry David B Larson, 2001; Russinova, Wewiorski, \& Cash, 2002; Smith, McCullough, \& Poll, 2003; Zaza, Sellick, \& Hillier, 2005).
+Most previous attempts to quantify the relationship between these two domains find a positive association: spirituality predicts well-being (Ginsburg, Quirt, Ginsburg, \& MacKillop, 1995; Koenig, 2010; Koenig, McCullough, \& Larson, 2001; Russinova, Wewiorski, \& Cash, 2002; Smith, McCullough, \& Poll, 2003; Zaza, Sellick, \& Hillier, 2005).
 However, critics have raised three credible challenges.
@@
 First, previous research has frequently operationalised spirituality using measures of well-being.
-For example, The \emph{Daily Spiritual Experience Scale}, which is widely used in spiritual mental health research, includes items to measure spirituality such as \enquote{I feel thankful for my blessings} and \enquote{I feel deep inner peace or harmony} (Underwood \& Teresi, 2002).
+For example, The \emph{Daily Spiritual Experience Scale}, which is widely used in spiritual mental health research, includes items to measure spirituality such as ``I feel thankful for my blessings'' and ``I feel deep inner peace or harmony'' (Underwood \& Teresi, 2002).
 However, these items are also used to measure subjective well-being.
-Koenig describes this approach as \enquote{tautological} (Koenig, 2008): it is unsurprising that spirituality is associated with well-being when the same items are used to measure both constructs.
+Koenig describes this approach as ``tautological'' (Koenig, 2008): it is unsurprising that spirituality is associated with well-being when the same items are used to measure both constructs.
 How pervasive is the use of tautological measures?
-A recent systematic-review of spirituality/mental health research found that nearly 45\% of previous studies employ tautological scales (Garssen et al., 2016).
+A recent systematic-review of spirituality/mental health research found that nearly 45\% of previous studies employ tautological scales (Garssen, Visser, \& Jager Meezenbroek, 2016).
 The first challenge, then, is to define measures of spirituality that are not well-being measures dressed up in other words.
@@
 A second challenge is to disentangle concepts of spirituality from concepts of religion.
-The term \enquote{spirituality} was first used to describe the religious practices of ascetics and monks, however, more recently, the term has taken on a broader family of meanings that some of which apply to non-religious contexts (Swinton, 2001).
+The term ``spirituality'' was first used to describe the religious practices of ascetics and monks, however, more recently, the term has taken on a broader family of meanings that some of which apply to non-religious contexts (Swinton, 2001).
 Notably, spirituality and health research has not converged on a unique set of meanings.
-Although it is commonplace across most areas of psychological science for researchers to investigate different operationalisations of focal concepts (think of memory research, or emotions research), the psychology of spirituality has mostly been conducted with cross-sectional North American samples among participants who affiliate with traditional religion (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky, Ratner, \& Chiu, 2005; Smith et al., 2003; Visser, Garssen, \& Vingerhoets, 2010; Yonker, Schnabelrauch, \& Dehaan, 2012).
+Although it is commonplace across most areas of psychological science for researchers to investigate different operationalisations of focal concepts (think of memory research, or emotions research), the psychology of spirituality has mostly been conducted with cross-sectional North American samples among participants who affiliate with traditional religion (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky, Ratner, \& Chiu, 2005; Smith, McCullough, \& Poll, 2003; Visser, Garssen, \& Vingerhoets, 2010; Yonker, Schnabelrauch, \& Dehaan, 2012).
 Whether the psychology of spirituality and health has repeated or extended the psychology of religion and health remains unclear.
@@
 To assess the mechanisms of well-being production requires observing change within people who differ in theoregically relevant ways and assessing how their lives turn out (Garssen \& Visser, 2016).
-However, previous investigations of spirituality and well-being, including high-quality reviews and meta-analyses, have nearly exclusively relied on cross-sectional samples (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky et al., 2005; Smith et al., 2003; Visser et al., 2010; Yonker et al., 2012).
+However, previous investigations of spirituality and well-being, including high-quality reviews and meta-analyses, have nearly exclusively relied on cross-sectional samples (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky, Ratner, \& Chiu, 2005; Smith, McCullough, \& Poll, 2003; Visser, Garssen, \& Vingerhoets, 2010; Yonker, Schnabelrauch, \& Dehaan, 2012).
 Although researching the key questions in spirituality psychological health research requires national longitudinal investigations, unfortunately the relevant datasets are rare.
 
-Here, we addresses these three challenges of previous research in the following ways: First, to avoid a tautological elision of \enquote{spirituality} and \enquote{well-being} we adopt a non-affective measure of spirituality as a \enquote{believe in a Spirit or Life Force.} On the other side, to measure subjective well-being, we use a well-validated scale for personal well-being and a well-validated scale for Life-Satisfaction (see Method).
-Second, to disentangle spiritual beliefs from traditional religious beliefs and skeptical beliefs we adopt a four-level categorical indicator combining the binary indicator of belief in a Spirit or Life Force with the binary indicator \enquote{Do you believe in some form of a spirit or Life Force?} The levels of this four-level categorical indicator for spiritual, religious, and skeptical beliefs are as follows: (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life Force; (2) Skeptics about God who nevertheless believe in a Spirit or Life Force; (3) Those who believe in a God but do not believe in a Spirit or Life Force; (4) Those who who believe in a both God and a Spirit or Life Force.
-Clearly such an approach cannot simultaneously address all interests in previous spirituality/well-being research.
+Here, we addresses these three challenges of previous research in the following ways: First, to avoid a tautological elision of ``spirituality'' and ``well-being'' we adopt a non-affective measure of spirituality as a ``believe in a Spirit or Life Force.'' On the other side, to measure subjective well-being, we use a well-validated scale for personal well-being and a well-validated scale for Life-Satisfaction (see Method).
+Second, to disentangle spiritual beliefs from traditional religious beliefs and dis0beliefs we adopt a four-level categorical indicator combining the binary indicator of belief in a Spirit or Life Force with the binary indicator ``Do you believe in some form of a spirit or Life Force?'' The levels of this four-level categorical indicator for spiritual and religious beliefs and dis-beliefs are as follows: (1) \textbf{God and Spirit Disbelievers}: those who neither believe in a God nor believe in a Spirit or Life Force; (2) \textbf{Spirit Believers/God Disbelievers}: those who disbelieve in a God and believe in a Spirit or Life Force; (3) \textbf{God Believers/Spirit Disbelievers}: those who believe in a God but do not believe in a Spirit or Life Force; (4) \textbf{God and Spirit Believers}: Those who believe in a both God and a Spirit or Life Force.
+Clearly such an approach to categorising religion and spirituality cannot simultaneously address many interests in previous spirituality/well-being research.
 For example we do not combine our cognitive measure of belief with behavioral measures or attitudinal measures.
 However, this cost to scope is paid for with benefits to precision.
-The spiritual indicators measure cognitive states in which distinct varieties of skeptical beliefs, traditional religious beliefs, and non-traditional spirit beliefs are straightforwardly disentangled, and measured within people.
-Third, to understand how spirit beliefs affect growth, stability, or decline in psychological well-being over time, we leverage longitudinal responses from a nationally diverse probability sample year over nine years (2010 to 2018 New Zealand Attitudes and Values Study, NZAVS).
+The spiritual indicators measure cognitive states in which distinct varieties of religious and spiritual beliefs and dis-beliefs are straightforwardly disentangled and measured within people, and their longitudinal associations with reliable well-being indicators are systematically quantified Third, to understand how spirit beliefs affect growth, stability, or decline in psychological well-being over time, we leverage longitudinal responses from a nationally diverse probability sample year over nine years (2010 to 2018 New Zealand Attitudes and Values Study, NZAVS).
 
@@
 The NZAVS is an annual, longitudinal national probability sample of registered New Zealand voters, which was started in 2009.
-The Time 10 wave of the NZAVS contained responses from 47,951 participants (17,981 retained from one or more previous waves and 29,970 new additions from booster sampling and/or unmatched participants or unsolicited opt-ins).
-Participants who provided an email address were also emailed and invited to complete an online version if they preferred.
-We offered a prize draw for participation, non-respondents were emailed and phoned multiple times, and all participants were mailed a Season's Greetings card from the NZAVS research team and informed that they had been automatically entered into a bonus seasonal grocery voucher prize draw.
-We also mailed our yearly pamphlet summarizing key research findings published during the current wave of the study.
+The NZAVS samples size by waves is as follows: Wave 1 (2009/10) = 6518; Wave 2 (2010/11) = 4441; Wave 3 (2011/12) = 6884; Wave 4 (2012/13) = 12179; Wave 5 (2013/14) = 18,261; Wave 6 (2014/15) = 15,820; Wave 7 (2015/16) = 13,942; Wave 8 (2016/17) = 21,936; Wave 9 (2017/18) = 17,072; Wave 10 (2018/19) = 47951; Wave 11(2019/20) = 36,524.
+
+The first NZAVS measures of belief in God and belief in a spirit or life force appear in Wave 2 (2010/11) NZAVS contained responses from 4,441 participants.
+The Wave 2 (2010/11) NZAVS retained 4,425 from the initial Time 1 (2009/10) NZAVS sample of 6,518 participants, and included an additional 16 respondents who could not be matched to the Time 1 participant database (a retention rate of 67.9\% over one year).
+Participants in the initial Time 1 (2009) sample were randomly selected from the New Zealand electoral roll (a national registry of registered voters).
+The response rate in the initial Time 1 sample, adjusting for the accuracy of the electoral roll and including anonymous responses was 16.6\%.
+Participants were posted a copy of the questionnaire, with a second postal follow-up two months later.
+Participants who provided an email address were also emailed and invited to complete an online questionnaire if they preferred.
+Full details for the NZAVS sampling procedure can be found in the supplement to this study can be found on our OSF link: \url{https://osf.io/cu2gr/}
 
@@
 
-The Time 10 (2018) wave of the NZAVS included 47,951 respondents.
-We analyzed data from participants who responded to our survey at least three times between Time 2 (2010) and Time 10 (2018), resulting in a sample of \(N\) = 21,705 (\(N\) = 20,979 with complete responses).
+We analyzed data from participants who responded to our survey in at least 9 of the 10 waves between Time 2 (2010/2011) and Time 11 (2019/2020), resulting in a sample of \(N\) = 3,257 individuals.
 Descriptive summaries for demographic and theoretical indicators are given in Table 1.
 
-Table 1 about here
+Table 1 about here.
 
-\begin{table}[ ht ] 
-\centering 
-\caption{}\label{}
-\scalebox{0.6}{
-\begin{tabular}{ l c c c c c c c c c }
+\begin{table}
+\begin{center}
+\scalebox{0.8}{
+\begin{tabular}{l c c}
 \toprule
- &   \multicolumn{ 9 }{c}{ Wave }\\ 
-  & 2010 & 2011 & 2012 & 2013 & 2014 & 2015 & 2016 & 2017 & 2018 \\ 
- & n = \emph{missing} & n = \emph{missing} & n = 8547 & n = 13472 & n = \emph{missing} & n = 11687 & n = 16048 & n = 14174 & n = 14883 \\ 
+ & PWI & Life Sat \\
 \midrule
-Age &   &   &   &   &   &   &   &   &  \\ 
-NZdep &   &   &   &   &   &   &   &   &  \\ 
-Education &   &   &   &   &   &   &   &   &  \\ 
-Employed &   &   &   &   &   &   &   &   &  \\ 
-\hspace{6pt}    0 &  (\%) &  (\%) & 2094 (24.5\%) & 2955 (21.9\%) &  (\%) & 2525 (21.6\%) & 3486 (21.7\%) & 3223 (22.7\%) & 3404 (22.9\%)\\ 
-\hspace{6pt}    1 &  (\%) &  (\%) & 6453 (75.5\%) & 10517 (78.1\%) &  (\%) & 9162 (78.4\%) & 12562 (78.3\%) & 10951 (77.3\%) & 11479 (77.1\%)\\ 
-Ethnic\_Categories &   &   &   &   &   &   &   &   &  \\ 
-\hspace{6pt}    Euro &  (\%) &  (\%) & 6731 (78.8\%) & 11024 (81.8\%) &  (\%) & 9671 (82.8\%) & 13389 (83.4\%) & 11814 (83.3\%) & 12427 (83.5\%)\\ 
-\hspace{6pt}    Maori &  (\%) &  (\%) & 1188 (13.9\%) & 1608 (11.9\%) &  (\%) & 1341 (11.5\%) & 1749 (10.9\%) & 1620 (11.4\%) & 1677 (11.3\%)\\ 
-\hspace{6pt}    Pacific &  (\%) &  (\%) & 284 (3.3\%) & 339 (2.5\%) &  (\%) & 277 (2.4\%) & 331 (2.1\%) & 243 (1.7\%) & 261 (1.8\%)\\ 
-\hspace{6pt}    Asian &  (\%) &  (\%) & 344 (4\%) & 501 (3.7\%) &  (\%) & 398 (3.4\%) & 579 (3.6\%) & 497 (3.5\%) & 518 (3.5\%)\\ 
-Male &   &   &   &   &   &   &   &   &  \\ 
-\hspace{6pt}    0 &  (\%) &  (\%) & 5302 (62\%) & 8457 (62.8\%) &  (\%) & 7325 (62.7\%) & 10148 (63.2\%) & 9011 (63.6\%) & 9433 (63.4\%)\\ 
-\hspace{6pt}    1 &  (\%) &  (\%) & 3245 (38\%) & 5015 (37.2\%) &  (\%) & 4362 (37.3\%) & 5900 (36.8\%) & 5163 (36.4\%) & 5450 (36.6\%)\\ 
-Has\_Partner &   &   &   &   &   &   &   &   &  \\ 
-\hspace{6pt}    0 &  (\%) &  (\%) & 2408 (28.2\%) & 3546 (26.3\%) &  (\%) & 2815 (24.1\%) & 3773 (23.5\%) & 3330 (23.5\%) & 3556 (23.9\%)\\ 
-\hspace{6pt}    1 &  (\%) &  (\%) & 6139 (71.8\%) & 9926 (73.7\%) &  (\%) & 8872 (75.9\%) & 12275 (76.5\%) & 10844 (76.5\%) & 11327 (76.1\%)\\ 
-Political\_Conservativism &   &   &   &   &   &   &   &   &  \\ 
-Urban &   &   &   &   &   &   &   &   &  \\ 
-\hspace{6pt}    Not\_Urban &  (\%) &  (\%) & 2865 (33.5\%) & 4278 (31.8\%) &  (\%) & 3951 (33.8\%) & 5538 (34.5\%) & 2610 (18.4\%) & 2661 (17.9\%)\\ 
-\hspace{6pt}    Urban &  (\%) &  (\%) & 5682 (66.5\%) & 9194 (68.2\%) &  (\%) & 7736 (66.2\%) & 10510 (65.5\%) & 11564 (81.6\%) & 12222 (82.1\%)\\ 
-Beliefs &   &   &   &   &   &   &   &   &  \\ 
-\hspace{6pt}    \_Skeptic\_ &  (\%) &  (\%) & 2048 (24\%) & 3286 (24.4\%) &  (\%) & 2951 (25.3\%) & 4116 (25.6\%) & 3916 (27.6\%) & 4120 (27.7\%)\\ 
-\hspace{6pt}    \_SpiritExcludesGod\_ &  (\%) &  (\%) & 2295 (26.9\%) & 3735 (27.7\%) &  (\%) & 3169 (27.1\%) & 4501 (28\%) & 3918 (27.6\%) & 3970 (26.7\%)\\ 
-\hspace{6pt}    GodAndSpirit &  (\%) &  (\%) & 3707 (43.4\%) & 5499 (40.8\%) &  (\%) & 5119 (43.8\%) & 6453 (40.2\%) & 5464 (38.5\%) & 6210 (41.7\%)\\ 
-\hspace{6pt}    GodExcludesSpirit &  (\%) &  (\%) & 497 (5.8\%) & 952 (7.1\%) &  (\%) & 448 (3.8\%) & 978 (6.1\%) & 876 (6.2\%) & 583 (3.9\%)\\ 
-LIFESAT &   &   &   &   &   &   &   &   &  \\ 
-PWI &   &   &   &   &   &   &   &   &  \\ 
+Intercept                            & $\mathbf{7.24}$   & $\mathbf{5.36}$   \\
+                                     & $ [ 7.18;  7.30]$ & $ [ 5.31;  5.40]$ \\
+BeliefsGodAndSpirit                  & $\mathbf{-0.08}$  & $-0.01$           \\
+                                     & $ [-0.13; -0.03]$ & $ [-0.05;  0.02]$ \\
+BeliefsGodExcludesSpirit             & $-0.04$           & $0.01$            \\
+                                     & $ [-0.10;  0.03]$ & $ [-0.04;  0.06]$ \\
+BeliefsSpiritExcludesGod             & $\mathbf{-0.06}$  & $-0.02$           \\
+                                     & $ [-0.10; -0.01]$ & $ [-0.05;  0.01]$ \\
+Age\_within                          & $\mathbf{0.02}$   & $\mathbf{0.01}$   \\
+                                     & $ [ 0.01;  0.03]$ & $ [ 0.01;  0.02]$ \\
+Age\_betweenC                        & $\mathbf{0.01}$   & $\mathbf{0.01}$   \\
+                                     & $ [ 0.01;  0.01]$ & $ [ 0.01;  0.01]$ \\
+Ethnic\_CategoriesMaori              & $-0.08$           & $0.04$            \\
+                                     & $ [-0.16;  0.00]$ & $ [-0.02;  0.10]$ \\
+Ethnic\_CategoriesPacific            & $\mathbf{-0.39}$  & $\mathbf{-0.29}$  \\
+                                     & $ [-0.54; -0.24]$ & $ [-0.39; -0.18]$ \\
+Ethnic\_CategoriesAsian              & $-0.05$           & $-0.04$           \\
+                                     & $ [-0.20;  0.13]$ & $ [-0.15;  0.09]$ \\
+Male1                                & $-0.05$           & $\mathbf{-0.11}$  \\
+                                     & $ [-0.13;  0.04]$ & $ [-0.16; -0.05]$ \\
+BeliefsGodAndSpirit:Age\_within      & $\mathbf{0.02}$   & $\mathbf{0.01}$   \\
+                                     & $ [ 0.01;  0.03]$ & $ [ 0.01;  0.02]$ \\
+BeliefsGodExcludesSpirit:Age\_within & $0.01$            & $0.00$            \\
+                                     & $ [-0.01;  0.03]$ & $ [-0.01;  0.02]$ \\
+BeliefsSpiritExcludesGod:Age\_within & $\mathbf{0.02}$   & $\mathbf{0.01}$   \\
+                                     & $ [ 0.01;  0.03]$ & $ [ 0.00;  0.02]$ \\
+\midrule
+SD: Idsd(Intercept)                  & $1.37$            & $0.96$            \\
+SD: Idsd(Age\_within)                & $0.10$            & $0.08$            \\
+SD: Idcor(Intercept,Age\_within)     & $-0.05$           & $-0.01$           \\
 \bottomrule
-
 \end{tabular}
 }
+\caption{}
+\label{tab:REGRESS_LS}
+\end{center}
 \end{table}
@@
 Life satisfaction is a measure of emotional well-being that was assessed using a 2-item version of the Satisfaction With Life Scale, which has previously been shown to correlate with aspects of religiosity (Diener, Emmons, Larsen, \& Griffin, 1985).
-Participants rated their agreement with the statements (a) \enquote{I am satisfied with my life}; and (b) \enquote{In most ways my life is close to ideal.} The items were rated on 7-point response options ranging from 1 = strongly disagree to 7 = strongly agree.
+Participants rated their agreement with the statements (a) ``I am satisfied with my life''; and (b) ``In most ways my life is close to ideal.'' The items were rated on 7-point response options ranging from 1 = strongly disagree to 7 = strongly agree.
 The mean Cronbach's alpha for this 2-item scale was \(\alpha\) = 0.90.
@@
 
-Age was put into units of 10 years and centered at its mean. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 2.
-Gender. Gender was assessed by asking participants if they were \enquote{Male} was coded as \enquote{1} and \enquote{Female} was coded as \enquote{0.}
+Age was put into units of 10 years and centered at its mean.
+For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 2.
+Gender.
+Gender was assessed by asking participants if they were ``Male'' was coded as ``1'' and ``Female'' was coded as ``0.''
 
@@
 
-Employment status was assessed by asking participants if they were currently working, \enquote{yes} was coded as \enquote{1} and \enquote{no} was coded as \enquote{0.} For detailed information pertaining to yearly responses and missingness, see Table 1.
+Employment status was assessed by asking participants if they were currently working, ``yes'' was coded as ``1'' and ``no'' was coded as ``0.'' For detailed information pertaining to yearly responses and missingness, see Table 1.
 
@@
 
-Participants were asked if they were in a relationship, \enquote{yes} was coded as \enquote{1} and no was coded as \enquote{0.} For detailed information pertaining to yearly responses and missingness, see Table 1.
+Participants were asked if they were in a relationship, ``yes'' was coded as ``1'' and no was coded as ``0.'' For detailed information pertaining to yearly responses and missingness, see Table 1.
 
@@
 
-People were coded as either residing in an urban \enquote{1} or rural \enquote{0} area based on New Zealand census data.
-For detailed information pertaining to yearly responses and missingness, see Table 1.
+People were coded as either residing in an urban ``1'' or rural ``0'' area based on New Zealand census data. For detailed information pertaining to yearly responses and missingness, see Table 1.
 
@@
 Personal well-being is a measure of function all well-being that was assessed using a 4-item version of the Australian Unity well-being Index (Cummins, Eckerseley, Pallant, Van Vugt, \& Misajon, 2017).
-Participants rated their satisfaction with (a) \enquote{Your standard of living}; (b) \enquote{Your health}; (c) \enquote{Your future security}; (d) \enquote{Your personal relationships.} Items were rated using a 10-point response option ranging from 1 = completely dissatisfied to 10 = completely satisfied.
+Participants rated their satisfaction with (a) ``Your standard of living''; (b) ``Your health''; (c) ``Your future security''; (d) ``Your personal relationships.'' Items were rated using a 10-point response option ranging from 1 = completely dissatisfied to 10 = completely satisfied.
 The mean Cronbach's alpha for this scale was \(\alpha\) = 0.90.
@@
 
-We assess belief in a spirit or a God by asking two non-affective questions: \enquote{Do you believe in a spirit or Life Force} and \enquote{Do you believe in a god?} Responses were coded as (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life Force; (2) Skeptics about God who believe in a Spirit or Life Force; (3) Believers in a God who disbelieve in a Spirit or Life Force; (4) Believers in a God and Spirit or Life Force.
-These believe were developed for the NZAVS from the 2005 Eurobarometer.
+We assess belief in a spirit or a God by asking two non-affective questions: ``Do you believe in a spirit or Life Force'' and ``Do you believe in a god?'' Responses were coded as 1) \textbf{God and Spirit Disbelievers}: those who neither believe in a God nor believe in a Spirit or Life Force; (2) \textbf{Spirit Believers/God Disbelievers}: those who disbelieve in a God and believe in a Spirit or Life Force; (3) \textbf{God Believers/Spirit Disbelievers}: those who believe in a God but do not believe in a Spirit or Life Force; (4) \textbf{God and Spirit Believers}: those who believe in a both God and a Spirit or Life Force.
 
@@
 Statistical analysis was performed using R version 4.0.2 R version 3.6.1 (2020-06-22) 2019-07-05).
-We analyze data from participants who responded to the NZAVS survey at least three times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of 21,705 (\emph{N} = 20,979 with complete responses).
+We analyze data from participants who responded to the NZAVS survey at least nine times between 2010/11 (Time 2) and 2019/20 (Time 11), resulting in a sample of 21,705 (\(N\) = 3,257 with complete responses).
 To model growth in subjective well-being we use generalized linear mixed models.
-In our models, we account for the non-linear effects of time within participants by including the interaction of years (a continuous variable) with Beliefs (a four-level categorical indicator).
-The baseline for this indicator is Skepticism about a Spirit/Life Forces and Skepticism about a God.
-To handle the dependencies introduced from the repeated measures, we include individual ID as an effect modeled as random, and we estimated a random effect for the slope for years.
-This allowed us to model correllation between individual-level intercept (starting high or low in Personal Well-Being/Life Satisfaction) and stability/growth/decline in the slope effect of years.
-We estimated all statitiscal models using both REML estimation (Bates, Mächler, Bolker, \& Walker, 2015) and Bayesian estimation (Bürkner, 2018) with weakly informative priors.
-We include the full reports in the Supplement.
-The REML estimates and the Bayesian estimates converged to within three decimal places.
-We include model equations and priors at the end of this article.
-Here we report the results of the Bayesian estimator.
+To avoid heterogeneity bias, we ``demeaned'' the variable ``age'' separating this variable into the within and between person components.
+Heterogeneity bias arises when an indicators varies within and between groups, which will case fixed effects estimates to correlate with group-level effects (Bell \& Jones, 2015), which is a common feature of panel data (in which a participant is a level-2 indicator)(Lüdecke, Ben-Shachar, Patil, \& Makowski, 2020).
+We modelled the interaction of within participant aging and belief by including the interaction of aging within (a continuous variable, centered) and belief-state, a four-level categorical indicator.
+The baseline for this indicator is disbelief in a Spirit/Life Force and disbelief in a God -- or ``disbeliever.'' To handle the dependencies introduced from the repeated measures, we include individual ID as an effect modeled as random, and we estimated a random effect for the slope for aging (within) This approach allowed us to model correlation between individual-level intercept (starting high or low in Personal Well-Being/Life Satisfaction) and stability/growth/decline in the slope effect of years.
+We estimated all statistical models using both REML estimation (Bates, Mächler, Bolker, \& Walker, 2015) and Bayesian estimation (Bürkner, 2018) with weakly informative priors, reported in the on-line supplement.
+The REML estimates, which we report in the supplement, converged to within three decimal places.
+The model equations and priors are in the on-line supplement.
 
@@
 To address confounding we first obtained the New Zealand Attitudes and Values Study standard demographic indicators recommended to address confounding: \href{https://cdn.auckland.ac.nz/assets/psych/about/our-research/nzavs/NZAVSTechnicalDocuments/NZAVS-Technical-Documents-e11-Default-Statistical-Standards.pdf}{NZAVS modelling standards}.
-Next we modelled potential relationships between the coefficients within the model using the daggity package in R, and plotting results using the daggity and ggdag packages in R.
+Next we modelled potential relationships between the coefficients within the model using the \texttt{ggdag} package in R.
 
-This method reveals two adjustments sets equired to avoid potential confounding: \{age, employment, male, political conservativism, and time\}, and \{age, ethnicty, male, and time \}.Here, we report the second set because employment is likely to contain more measurement error: some are not employed because they cannot find employment, others because they are not seeking employment (e.g.~retirement).
+This method reveals two adjustments sets required to avoid potential confounding: \{age\_within, age\_between, employment, male, political conservativism, and time\}, and \{age\_within, age\_between, ethnicty, male, and time\}.
+Here, we used the second adjustment set because employment is likely to contain more measurement error: some are not employed because they cannot find employment, others because they are not seeking employment (e.g.~retirement).
 Results were insensitive to the choice set.
-Indeed, the inference proved robust to the simple model with only the interaction of years and beliefs, and robust to the fully saturated model in which all covariates were included (see the online supplement for details).
-In the fully saturated model, it appeared that those skeptical of a Spirit or Life Force, exhibited no growth in Persona Well-Being and Life-Satisifaction over the nine years.
-Although this result offers stronger stronger evidence for a Spirit/Life Force well-being mechanism, it is likely the effect of collider confounding and masking.
-We discuss this point in more detail in the supplement, and mention it here to caution researchers from \enquote{everything-but-the-kitchen-sink} approaches to regression modelling.
 
@@
 
-Because missing values can result in biased estimates (Blackwell, Honaker, \& King, 2017), we multiply imputed 10 datasets using the AmeliaII package in R(Honaker et al., 2011b).We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences.
-Additionally the results were robust to filtering of participants of N \textgreater{} 3 to N = 8 waves.
-We include the multiple-imputation results in a supplement.
-Here, there were no differences in estimation relevant to inference, however, to improve inference, we advocate wider use of multiple-imputation or Bayesian approaches for handling missingness.
+Because missing values can result in biased estimates (Blackwell, Honaker, \& King, 2017), we multiply imputed 10 datasets using the AmeliaII package in R(Honaker, King, Blackwell, \& Others, 2011).We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences.
 
@@
 
-The analysis was performed using R version 4.0.2 (2020-06-22). The Platform was x86\_64-apple-darwin17.0 (64-bit) Running under: macOS Catalina 10.15.6
-We are greatful to the contributors and mantainers of the following packages:
-R (Version 4.0.2; R Core Team, 2020) and the R-packages \emph{Amelia} (Version 1.7.6; Honaker et al., 2011a), \emph{citr} (Version 0.3.2; Aust, 2019), \emph{dplyr} (Version 1.0.2; Wickham et al., 2020), \emph{equatiomatic} (Version 0.1.0.9000; Anderson \& Heiss, 2020), \emph{forcats} (Version 0.5.0; Wickham, 2020a), \emph{ggeffects} (Version 0.15.0; Lüdecke, 2018), \emph{gghighlight} (Version 0.3.0; Yutani, 2020), \emph{ggplot2} (Version 3.3.2; Wickham, 2016), \emph{ggsci} (Version 2.9; Xiao, 2018), \emph{lme4} (Version 1.1.23; Bates et al., 2015), \emph{Matrix} (Version 1.2.18; Bates \& Maechler, 2019), \emph{papaja} (Version 0.1.0.9997; Aust \& Barth, 2020), \emph{parameters} (Version 0.8.6.1; D.
-Lüdecke et al., 2020a), \emph{patchwork} (Version 1.0.1; Pedersen, 2020), \emph{prettycode} (Version 1.1.0; Csárdi, 2019), \emph{purrr} (Version 0.3.4; Henry \& Wickham, 2020), \emph{Rcpp} (Version 1.0.5.1; Eddelbuettel \& François, 2011; Eddelbuettel \& Balamuta, 2017), \emph{readr} (Version 1.3.1; Wickham, Hester, \& Francois, 2018), \emph{report} (Version 0.1.0; Makowski, Dominique, Lüdecke, \& Daniel, 2019), \emph{see} (Version 0.5.2.1; D.
-Lüdecke et al., 2020b), \emph{sjPlot} (Version 2.8.4; D.
-Lüdecke, 2020), \emph{stringr} (Version 1.4.0; Wickham, 2019), \emph{styler} (Version 1.3.2; Müller \& Walthert, 2020), \emph{table1} (Version 1.2; Rich, 2020), \emph{texreg} (Version 1.37.5; Leifeld, 2013), \emph{tibble} (Version 3.0.4; Müller \& Wickham, 2020), \emph{tidyr} (Version 1.1.2; Wickham, 2020b), and \emph{tidyverse} (Version 1.3.0; Wickham, Averick, et al., 2019)"
+The analysis was performed using R version 4.0.2 (2020-06-22).
+The Platform was x86\_64-apple-darwin17.0 (64-bit) Running under: macOS Catalina 10.15.6 We are greatful to the contributors and mantainers of the following packages: R {[}Version 4.0.3; R Core Team (2020){]} and the R-packages \emph{Amelia} {[}Version 1.7.6; Honaker, King, and Blackwell (2011){]}, \emph{citr} {[}Version 0.3.2; Aust (2019){]}, \emph{dplyr} {[}Version 1.0.2; Wickham, François, Henry, and Müller (2020){]}, \emph{equatiomatic} {[}Version 0.1.0.9000; Anderson and Heiss (2020){]}, \emph{forcats} {[}Version 0.5.0; Wickham (2020a){]}, \emph{ggeffects} {[}Version 1.0.0; Lüdecke (2018){]}, \emph{gghighlight} {[}Version 0.3.0; Yutani (2020){]}, \emph{ggplot2} {[}Version 3.3.2; Wickham (2016){]}, \emph{ggsci} {[}Version 2.9; Xiao (2018){]}, \emph{lme4} {[}Version 1.1.25; Bates, Mächler, Bolker, and Walker (2015){]}, \emph{Matrix} {[}Version 1.2.18; Bates and Maechler (2019){]}, \emph{papaja} {[}Version 0.1.0.9997; Aust and Barth (2020){]}, \emph{parameters} {[}Version 0.9.0.1; Lüdecke, Ben-Shachar, Patil, and Makowski (2020){]}, \emph{patchwork} {[}Version 1.1.0; Pedersen (2020){]}, \emph{prettycode} (Csárdi, 2019), \emph{purrr} {[}Version 0.3.4; Henry and Wickham (2020){]}, \emph{Rcpp} {[}Version 1.0.5; Eddelbuettel and François (2011); Eddelbuettel and Balamuta (2017){]}, \emph{readr} {[}Version 1.4.0; Wickham, Hester, and Francois (2018){]}, \emph{report} {[}Version 0.2.0; Makowski, Dominique, Lüdecke, and Daniel (2019){]}, \emph{see} {[}Version 0.6.0.1; Lüdecke, Ben-Shachar, Waggoner, and Makowski (2020){]}, \emph{sjPlot} {[}Version 2.8.6; Lüdecke (2020){]}, \emph{stringr} {[}Version 1.4.0; Wickham (2019){]}, \emph{styler} {[}Version 1.3.2; Müller and Walthert (2020){]}, \emph{table1} {[}Version 1.2.1; Rich (2020){]}, \emph{texreg} {[}Version 1.37.5; Leifeld (2013){]}, \emph{tibble} {[}Version 3.0.4; Müller and Wickham (2020){]}, \emph{tidyr} {[}Version 1.1.2; Wickham (2020b){]}, and \emph{tidyverse} {[}Version 1.3.0; Wickham et al.
+(2019){]}
 
@@
 
-All code for this analysis and a full report of the missing data imputation can be found at \url{https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing}.
+All code for this analysis and a full report of the missing data imputation can be found at \url{https://github.com/go-bayes/nzavs_spiritbeliefWellbeing}.
 An html document decribing the same analysis is located at: \url{https://osf.io/cu2gr/}.
@@
 \midrule
-Intercept                           & $\mathbf{7.14}$   & $\mathbf{5.29}$   \\
-                                    & $ [ 7.11;  7.18]$ & $ [ 5.26;  5.31]$ \\
-yearsC                              & $\mathbf{0.02}$   & $\mathbf{0.01}$   \\
-                                    & $ [ 0.01;  0.02]$ & $ [ 0.00;  0.01]$ \\
-Beliefs\_SpiritExcludesGod\_        & $\mathbf{-0.06}$  & $0.01$            \\
-                                    & $ [-0.09; -0.03]$ & $ [-0.01;  0.03]$ \\
-BeliefsGodAndSpirit                 & $\mathbf{-0.06}$  & $\mathbf{0.03}$   \\
-                                    & $ [-0.09; -0.03]$ & $ [ 0.01;  0.06]$ \\
-BeliefsGodExcludesSpirit            & $-0.00$           & $\mathbf{0.06}$   \\
-                                    & $ [-0.04;  0.04]$ & $ [ 0.03;  0.09]$ \\
-Age\_in\_Decades\_C                 & $\mathbf{0.07}$   & $\mathbf{0.08}$   \\
-                                    & $ [ 0.06;  0.09]$ & $ [ 0.07;  0.09]$ \\
-Ethnic\_CategoriesMaori             & $\mathbf{-0.38}$  & $\mathbf{-0.11}$  \\
-                                    & $ [-0.43; -0.33]$ & $ [-0.14; -0.08]$ \\
-Ethnic\_CategoriesPacific           & $\mathbf{-0.38}$  & $\mathbf{-0.15}$  \\
-                                    & $ [-0.48; -0.29]$ & $ [-0.22; -0.09]$ \\
-Ethnic\_CategoriesAsian             & $\mathbf{-0.09}$  & $-0.05$           \\
-                                    & $ [-0.18; -0.01]$ & $ [-0.11;  0.02]$ \\
-Male1                               & $\mathbf{-0.06}$  & $\mathbf{-0.16}$  \\
-                                    & $ [-0.10; -0.02]$ & $ [-0.19; -0.13]$ \\
-yearsC:Beliefs\_SpiritExcludesGod\_ & $\mathbf{0.01}$   & $\mathbf{0.01}$   \\
-                                    & $ [ 0.00;  0.02]$ & $ [ 0.00;  0.02]$ \\
-yearsC:BeliefsGodAndSpirit          & $\mathbf{0.01}$   & $\mathbf{0.01}$   \\
-                                    & $ [ 0.01;  0.02]$ & $ [ 0.00;  0.02]$ \\
-yearsC:BeliefsGodExcludesSpirit     & $0.00$            & $-0.00$           \\
-                                    & $ [-0.01;  0.02]$ & $ [-0.01;  0.01]$ \\
+Intercept                                & $\mathbf{6.61}$   & $\mathbf{4.85}$   \\
+                                         & $ [ 6.44;  6.79]$ & $ [ 4.73;  4.97]$ \\
+Beliefs\_SpiritExcludesGod\_             & $\mathbf{-0.06}$  & $-0.02$           \\
+                                         & $ [-0.10; -0.01]$ & $ [-0.05;  0.01]$ \\
+BeliefsGodAndSpirit                      & $\mathbf{-0.07}$  & $-0.01$           \\
+                                         & $ [-0.12; -0.02]$ & $ [-0.05;  0.02]$ \\
+BeliefsGodExcludesSpirit                 & $-0.03$           & $0.01$            \\
+                                         & $ [-0.11;  0.03]$ & $ [-0.04;  0.06]$ \\
+Age\_within                              & $\mathbf{0.02}$   & $\mathbf{0.01}$   \\
+                                         & $ [ 0.01;  0.03]$ & $ [ 0.01;  0.02]$ \\
+Age\_between                             & $\mathbf{0.01}$   & $\mathbf{0.01}$   \\
+                                         & $ [ 0.01;  0.01]$ & $ [ 0.01;  0.01]$ \\
+Ethnic\_CategoriesMaori                  & $-0.08$           & $0.04$            \\
+                                         & $ [-0.16;  0.00]$ & $ [-0.02;  0.09]$ \\
+Ethnic\_CategoriesPacific                & $\mathbf{-0.39}$  & $\mathbf{-0.29}$  \\
+                                         & $ [-0.53; -0.23]$ & $ [-0.40; -0.18]$ \\
+Ethnic\_CategoriesAsian                  & $-0.06$           & $-0.03$           \\
+                                         & $ [-0.23;  0.12]$ & $ [-0.16;  0.08]$ \\
+Male1                                    & $-0.05$           & $\mathbf{-0.10}$  \\
+                                         & $ [-0.13;  0.03]$ & $ [-0.17; -0.05]$ \\
+Beliefs\_SpiritExcludesGod\_:Age\_within & $\mathbf{0.02}$   & $\mathbf{0.01}$   \\
+                                         & $ [ 0.01;  0.03]$ & $ [ 0.00;  0.02]$ \\
+BeliefsGodAndSpirit:Age\_within          & $\mathbf{0.02}$   & $\mathbf{0.01}$   \\
+                                         & $ [ 0.01;  0.03]$ & $ [ 0.01;  0.02]$ \\
+BeliefsGodExcludesSpirit:Age\_within     & $0.01$            & $0.00$            \\
+                                         & $ [-0.01;  0.03]$ & $ [-0.01;  0.02]$ \\
 \midrule
-SD: Idsd(Intercept)                 & $1.42$            & $0.98$            \\
-SD: Idsd(yearsC)                    & $0.13$            & $0.10$            \\
-SD: Idcor(Intercept,yearsC)         & $-0.03$           & $0.03$            \\
+SD: Idsd(Intercept)                      & $1.37$            & $0.96$            \\
+SD: Idsd(Age\_within)                    & $0.10$            & $0.08$            \\
+SD: Idcor(Intercept,Age\_within)         & $-0.05$           & $-0.01$           \\
 \bottomrule
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{/Users/jbul176/The Virtues Project Dropbox/Joseph Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/coefBayesDag} \caption{Coefficient Graphs}\label{fig:unnamed-chunk-1}
+\includegraphics[width=\textwidth]{/Users/jbul176/GIT/nzavs_spiritbeliefWellbeing/coef} \caption{Coefficient Graphs}\label{fig:unnamed-chunk-1}
 \end{figure}
@@
 \hypertarget{personal-well-being-1}{%
-\subsection{Personal Well-Being}\label{personal-well-being-1}}
+\subsection{Personal well-being}\label{personal-well-being-1}}
 
-We fitted a linear mixed effects model using the brms package (Bürkner, 2018) in R (R Core Team, 2020) to predict personal well-being (PWI) with years (0-9), beliefs (four categories), with Id as random intercept and years as a random slope effect (see below: model equations).
+We fitted a linear mixed effects model using the brms package (Bürkner, 2018) in R (R Core Team, 2020) to predict personal well-being (PWI) with age\_within centered (range = -5.79, 6.83), beliefs (four categories), with Id as random intercept and years as a random slope effect (see below: model equations).
 
@@
 \item
-  The intercept, representing the expected personal well-being among skeptics about a Spirit/Life Force and God who are Europeans, 50.77 years of age and do not identify as male is reliable (posterior median = 7.138 HDI {[} 7.111, 7.167{]}, pd = 100.00\%).\\
+  The intercept, representing the expected personal well-being among disbelievers in a Spirit/Life Force and God who are Europeans, 50.77 years of age and do not identify as male is reliable (posterior median = 6.611 HDI {[} 6.440, 6.780{]}, pd = 100.00\%).\\
 \item
-  The effect of years is positive is reliable (posterior median = 7.138, HDI {[} 7.111, 7.167{]}, pd 100.00\%).
-We infer that personal well-being increased in the baseline skeptic population between 2010 and 2018.
+  The effect of aging (years) is positive is reliable (posterior median = 0.021, HDI {[} 0.014, 0.029{]}, pd 100.00\%).
+We infer that personal well-being increased in the baseline disbelieving population between 2010 and 2020.
 \item
-  The effect of beliefs in \emph{Spirit Excludes God} is reliably negative (posterior median = -0.059, HDI {[}-0.081, -0.036{]}, pd 100.00\%).
-We infer that the population that believed in a Spirit but did not believe in a God was lower in personal well-being at the intercept compared to skeptics.
+  The effect of beliefs in \emph{Spirit Excludes God} is reliably negative (posterior median = -0.058, HDI {[}-0.097, -0.015{]}, pd 98.90\%).
+We infer that the population that believed in a Spirit but did not believe in a God was lower in personal well-being at the intercept compared to disbelievers.
 \item
-  The effect of beliefs in \emph{God And Spirit} is not negative (posterior median = -0.060, HDI {[}-0.086, -0.034{]}, pd = 100.00\%) .
-We infer that the population that believed in a God and a Spirit or Life Force was lower in personal well-being at the intercept compared to skeptics.
+  The effect of beliefs in \emph{God And Spirit} is negative (posterior median =-0.075 , HDI {[}-0.124, -0.024{]}, pd = 99.35\%) .
+We infer that the population that believed in a God and a Spirit or Life Force was lower in personal well-being at the intercept compared to disbelievers.
 \item
-  The effect of beliefs in \emph{God Excludes Spirit} is not reliable (posterior median = -0.001 HDI {[}-0.036, 0.035{]}, pd = 51.10\%).
-We infer that the population that believed in a God and doubted a Spirit or Life Force were similar were similar in personal well-being at the intercept to skeptics.
-\item
-  The effect of Age in Decades (C) is 0.072 \textbar{} {[} 0.060, 0.083{]} \textbar{} 100.00\% \textbar{}
-\item
-  The effect of Maori Ethnicity is -0.380 \textbar{} {[}-0.420, -0.341{]} \textbar{} 100.00\% \textbar{}
-\item
-  The effect of Pacific Ethnicity is -0.384 \textbar{} {[}-0.464, -0.312{]} \textbar{} 100.00\% \textbar{}
-\item
-  The effect of Asian Ethnicity is -0.093 \textbar{} {[}-0.165, -0.023{]} \textbar{} 98.28\% \textbar{}\\
-\item
-  The effect of Male gender is -0.060 \textbar{} {[}-0.094, -0.026{]} \textbar{} 99.90\% \textbar{}
+  The effect of beliefs in \emph{God Excludes Spirit} is not reliable (posterior median = -0.035 HDI {[}-0.107, 0.029{]}, pd = 78.50\%).
+We infer that the population that believed in a God and doubted a Spirit or Life Force were similar in personal well-being at the intercept to the disbelieving population:
 \end{itemize}
 
-\hypertarget{personal-well-being-years-x-beliefs}{%
-\subsection{Personal Well-Being: years X beliefs}\label{personal-well-being-years-x-beliefs}}
+\hypertarget{personal-well-being-aging-years-x-beliefs}{%
+\subsection{Personal well-being: aging (years) X beliefs}\label{personal-well-being-aging-years-x-beliefs}}
 
 \begin{figure}
-\includegraphics[width=\textwidth]{/Users/jbul176/The Virtues Project Dropbox/Joseph Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/pwi_confounder_adjusted} \caption{Personal Well-being prediction graph}\label{fig:unnamed-chunk-2}
-\end{figure}
-
-The effect of years \(\times\) belief in a \emph{Spirit Excludes God} is reliably positive (posterior median = 0.011, HDI {[} 0.003, 0.018{]}, pd = 99.02\%).
-We infer that people with beliefs in a spirit or Life Force who do not believe in a God tended to grow in their personal well-being over time at a faster rate than skeptics grew.
-- The effect of years \(\times\) in \emph{God And Spirit} is reliably positive (posterior median = 0.014, HDI {[}0.007, 0.021{]}, pd = 99.98\%).
-We infer that people who believe in a spirit or Life Force who also believe in a God tended to grow in their personal well-being over time at a faster rate than skeptics grew.
-- The effect of years \(\times\) belief in \emph{God Excludes Spirit} is not reliable (posterior median = 0.004, HDI {[}-0.009, 0.015{]}, pd = 68.73\%).
-We infer that people who believe in a God but not a spirit or Life Force did not grow at a faster rate in personal well-being than skeptics grew.
-
-\hypertarget{life-satisfaction-1}{%
-\subsection{Life Satisfaction}\label{life-satisfaction-1}}
-
-We fitted a linear mixed effects model using the brms package (Bürkner, 2018) in R (R Core Team, 2020) to predict personal well-being (PWI) with years (0-9), beliefs (four categories), with Id as random intercept and years as a random slope effect (see model equations, below).
-
-Within this model:
-- The intercept, representing the expected life satisfaction among skeptics about a Spirit/Life Force and God who are Europeans, 50.77 years of age and do not identify as male is reliable (posterior median = 5.287, HDI {[} 5.267, 5.306{]}, pd = 100.00\%).
-- The effect of years is positive is reliable (posterior median = 0.008, HDI {[} 0.003, 0.011{]}, pd = 99.75\%).
-We infer that life satisfaction increased in the skeptic population between 2010 and 2018.
--The effect of beliefs in \emph{Spirit Excludes God} is not reliable (posterior median = 0.007, HDI {[}-0.010, 0.023{]}, pd = 75.33\%).
-We infer that the population that believed in a Spirit but did not believe in a God was higher in life satisfaction at the intercept compared with skeptics.
-- The effect of beliefs in \emph{God And Spirit} is reliably positive (posterior median = 0.034, HDI {[} 0.015, 0.053{]}, pd = 99.83\%).
-We infer that the population that believed in a God and a Spirit or Life Force was higher in life satisfaction at the intercept compared with skeptics.
-- The effect of beliefs in \emph{God Excludes Spirit} is reliably positive (posterior median = 0.057, HDI {[} 0.030, 0.082{]}, pd = 99.95\%).
-We infer that the population that believed in a God and doubted a Spirit or Life Force was higher in life satisfaction at the intercept compared with skeptics.
-- The effect of Age in Decades (C) is reliably positive (posterior median = 0.079, hDI {[} 0.071, 0.087{]} pd = 100.00\%).
-- The effect of Maori Ethnicity is reliably negative (posterior median = -0.109 \textbar{} {[}-0.137, -0.082{]}, pd = 100.00\%).
-- The effect of Pacific Ethnicity is reliably negative (posterior median = -0.153 \textbar{} {[}-0.210, -0.103{]}, pd = 100.00\%).
-- The effect of Asian Ethnicity is not reliable (posterior median = -0.047, {[}-0.098, 0.004{]}, pd = 93.08\%).
-- The effect of Male gender is reliabily negative (posterior median = -0.161, {[}-0.184, -0.138{]}, pd = 100.00\%)
-
-\hypertarget{life-satisfaction-years-x-beliefs}{%
-\subsection{Life Satisfaction: years X beliefs}\label{life-satisfaction-years-x-beliefs}}
-
-\begin{figure}
-\includegraphics[width=\textwidth]{/Users/jbul176/The Virtues Project Dropbox/Joseph Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/ls_counfounder_adjusted} \caption{Life satisfaction prediction graph}\label{fig:unnamed-chunk-3}
+\includegraphics[width=\textwidth]{/Users/jbul176/GIT/nzavs_spiritbeliefWellbeing/pwi} \caption{Personal Well-being prediction graph}\label{fig:unnamed-chunk-2}
 \end{figure}
@@
 \item
-  The interaction of years \(\times\) beliefs in \emph{Spirit Excludes God} is reliably positive (posterior median = 0.011, HDI {[} 0.006, 0.016{]}, pd = 99.95\%).
-We infer that people with beliefs in a Spirit or Life Force who do not believe in a God tended to grow over time in their life satisfaction at a faster rate than skeptics.
+  The effect of aging (years) \(\times\) belief in a \emph{Spirit Excludes God} is reliably positive (posterior median = 0.020, HDI {[} 0.010, 0.030{]}, pd = 99.90\%).
+We infer that people with beliefs in a spirit or Life Force who do not believe in a God tended to grow in their personal well-being over time at a faster rate than disbelievers grew.
+\item
+  The effect of aging (years) \(\times\) in \emph{God And Spirit} is reliably positive (posterior median =0.022, HDI {[}0.012, 0.031{]}, pd = 100.00\%).
+We infer that people who believe in a spirit or Life Force who also believe in a God tended to grow in their personal well-being over time at a faster rate than disbelievers grew.
+\item
+  The effect of aging (years) \(\times\) belief in \emph{God Excludes Spirit} is not reliable (posterior median = 0.010, HDI {[}-0.007, 0.026{]}, pd = 83.40\%).
+We infer that people who believe in a God but not a spirit or Life Force did not grow at a faster rate in personal well-being than disbelievers grew.
+\end{itemize}
+
+\hypertarget{life-satisfaction-1}{%
+\subsection{Life Satisfaction}\label{life-satisfaction-1}}
+
+We fitted a linear mixed effects model using the brms package (Bürkner, 2018) in R (R Core Team, 2020) to predict Life Satisfaction (LS) with age\_within centered (range = -5.79, 6.83), beliefs (four categories), with Id as random intercept and years as a random slope effect (see below: model equations).
+
+Within this model:
+- The intercept, representing the expected life satisfaction among disbelievers in a Spirit/Life Force and God who are Europeans, 50.77 years of age and do not identify as male is reliable (posterior median = 6.611, HDI {[}6.440, 6.780{]}, pd = 100.00\%).
+- The effect of aging (years) is positive is reliable (posterior median = 0.021, HDI {[}0.014, 0.029{]}, pd = 100.00\%).
+We infer that life satisfaction increased in the disbelieving population between 2010 and 2018.
+-The effect of beliefs in \emph{Spirit Excludes God} is negative (posterior median = -0.058, HDI {[}-0.097, -0.015{]}, pd = 98.90\%).
+We infer that the population that believed in a Spirit but did not believe in a God was lower in life satisfaction at the intercept compared with disbelievers.
+- The effect of beliefs in \emph{God And Spirit} is reliably negative (posterior median = -0.075, HDI {[} -0.124, -0.024{]}, pd = 99.35\%).
+We infer that the population that believed in a God and a Spirit or Life Force was lower in life satisfaction at the intercept compared with disbelievers.
+- The effect of beliefs in \emph{God Excludes Spirit} is not reliable (posterior median = -0.035, HDI {[}-0.107, 0.029{]}, pd = 78.50\%).
+We infer that the population that believed in a God and doubted a Spirit or Life Force not reliably different in life satisfaction at the intercept compared with disbelievers.
+
+\hypertarget{life-satisfaction-aging-years-x-beliefs}{%
+\subsection{Life Satisfaction: aging (years) X beliefs}\label{life-satisfaction-aging-years-x-beliefs}}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{/Users/jbul176/GIT/nzavs_spiritbeliefWellbeing/ls} \caption{Life satisfaction prediction graph}\label{fig:unnamed-chunk-3}
+\end{figure}
 
@@
 \item
-    The interaction of years \(\times\) \emph{God And Spirit} is reliably positive (posterior median = 0.009, HDI {[} 0.004, 0.014{]}, pd = 99.72\%).
-We infer that people who believe in a spirit or Life Force who also believe in a God tended to grow over time in their life satisfaction at a faster rate than skeptics.
+  The interaction of aging (years) \(\times\) beliefs in \emph{Spirit Excludes God} is reliably positive (posterior median = 0.020 , HDI {[}0.010, 0.030{]}, pd = 99.90\%).
+We infer that people with beliefs in a Spirit or Life Force who do not believe in a God tended to grow over time in their life satisfaction at a faster rate than disbelievers.
 \item
-    The effect of years \(\times\) beliefs in \emph{God Excludes Spirit} is not reliable (posterior median = -0.003, HDI {[}-0.012, 0.006{]}, pd = 72.28\%).
-We infer that people who believe in a God but not a spirit or Life Force tended to grow over time in their life satisfaction at a similar rate to skeptics.
-  \end{itemize}
+  The interaction of aging (years) \(\times\) \emph{God And Spirit} is reliably positive (posterior median = 0.022, HDI {[}0.012, 0.031{]}, pd = 100.00\%).
+We infer that people who believe in a spirit or Life Force who also believe in a God tended to grow over time in their life satisfaction at a faster rate than disbelievers.
+\item
+  The effect of years \(\times\) beliefs in \emph{God Excludes Spirit} is not reliable (posterior median = 0.010, HDI {[}-0.007, 0.026{]}, pd = 83.40\%).
+We infer that people who believe in a God but not a spirit or Life Force did no experience reliably different growth in life satisfaction compared with disbelievers.
 \end{itemize}
@@
 \def\labelenumi{\arabic{enumi}.}
+\tightlist
 \item
-  At the outset of the study, in 2010, the expected subjective well-being as measured using the personal well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God (\(\hat{y} =\) 6.93 \textbar{} {[}6.88, 6.98{]}) and lower among those who believed in both a Spirit/Life Force and a God (\(\hat{y}\) = 6.91 \textbar{} {[}6.87, 6.95{]}) in comparison both to the skeptical population who neither believed in a God nor Spirit/ Life Force (\(\hat{y}\) = 7.04 \textbar{} {[}6.99, 7.10{]}) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force (\(\hat{y}\) 7.03 \textbar{} {[}6.95, 7.10{]}).
-(See Figure 2) Similarly,the expected subjective well-being as measured using the life satisfaction Inventory was lower in 2010 among those who believed in Spirit or Life Force but not a God (\(\hat{y}\) = 5.20 \textbar{} {[}5.16, 5.23{]}).
-Life satisfaction was similar among those who believed in both a Spirit/Life Force and a God (\(\hat{y}\) = 5.23 \textbar{} {[}5.21, 5.26{]}) in comparison both with the skeptical population who neither believed in a God nor a Life Force (\(\hat{y}\) = 5.25 \textbar{} {[}5.21, 5.29{]}).
-Life satisfaction was highest among the population that believed in a God but was skeptical about belief in a Spirit or Life Force (\(\hat{y}\) 5.32 \textbar{} {[}5.26, 5.38{]}) (see Figure 3)
+  In 2010 the expected subjective well-being as measured using the personal well-being inventory was expected to be lower among those who believed in Spirit or Life Force but not a God (\(\hat{y} =\) 6.95 \textbar{} {[}6.86, 7.03{]}), lower among those who believed in both a Spirit/Life Force and a God (\(\hat{y}\) = 6.91 \textbar{} {[}6.83, 6.99{]}), and lower among the population that believed in a God but was disbelieving in a Spirit or Life Force 6.95 \textbar{} {[}6.86, 7.03{]} when compared with the population of disbelievers (\(\hat{y}\) = 7.12 \textbar{} {[}7.03, 7.21{]}) and to (\(\hat{y}\) 7.03 \textbar{} {[}6.95, 7.10{]}).
+(See Figure 2).
+\end{enumerate}
+
+Focusing on life satisfaction, people who believed in Spirit, either with a belief in God (\(\hat{y}\) = 5.18 \textbar{} {[}5.13, 5.24{]}) or without a belief in God (5.18 \textbar{} {[}5.12, 5.24{]}) tended to be lower in life satsifaction than disbelievers in both a God and a Spirit (5.27 \textbar{} {[}5.20, 5.33{]}) and disbelievers in a spirit who also believed in a God (5.26 \textbar{} {[}5.15, 5.36{]}) (see Figure 3)
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{1}
+\tightlist
 \item
-  However, over time, belief in a Spirit or Life Force predicts a trajectory of growth in personal well-being.
-In 2018, personal well-being among who believe in Spirit or Life Force but not a God had grown to (\(\hat{y} = 7.19\) {[}7.15, 7.23{]}, \(\Delta \hat{y} = 0.26\)).
-Among those who believed in both a Spirit/Life Force and a God, personal well-being had grown to (\(\hat{y} = 7.20\) {[}7.17, 7.24{]}, \(\Delta \hat{y} = 0.29\)).
-By contrast, personal well-being among the skeptical population who neither believed in a God nor a Life Force grew less steeply across the eight year interval \(\hat{y} = 7.21\) {[}7.17, 7.25{]}, \(\Delta \hat{y} = 0.17\)).
-By the same token, personal well-being among those who believe in a God but are skeptical about a Spirit/Life Force grew less steeply \(\hat{y} = 7.22\) {[}7.16, 7.28{]} \(\Delta \hat{y} = 0.19\)).
-In practical terms, the rate of increase in personal wellbeing among those with Spirit/Life Force beliefs was about 30\% greater than it was among those who lacked such Spirit/Life Force (See Figure 2).
-Similarly, over time, belief in a Spirit or Life Force predicts a trajectory of growth in life satisfaction.
-In 2018, the expected effect of spiritual beliefs on life satisfaction among those who believe in Spirit or Life Force but not a God had grown to (\(\hat{y} = 5.37\) {[}5.34, 5.40{]}, \(\Delta \hat{y} = 0.17\)).
-Among those who believed in both a Spirit/Life Force and a God, life satisfaction had grown to (\(\hat{y} = 5.39\) {[}5.36, 5.41{]}, \(\Delta \hat{y} = 0.19\)).By contrast, life satisfaction among the skeptical population who neither believed in a God nor a Life Force lagged behind over the eight year interval \(\hat{y} = 5.32\) {[}5.29, 5.35{]}, \(\Delta \hat{y} = 0.07\)).
-Relatedly, life satisfaction in the population that believe in God but are skeptical about a Spirit/Life Force grew less steeply \(\hat{y} = 5.36\) {[}5.31, 5.41{]}, \(\Delta \hat{y} = 0.04\)).
-In practical terms, the rate of increase in life satisfaction among those with Spirit/Life Force beliefs over three times greater than among those that lacked such beliefs (See Figure 3).
-\item
-  Inferred growth from spiritual beliefs is evident from 9 annual waves of data collection (8 years), and across a nationally diverse sample.
-On this basis of this longitudinal information accumulated across nationally, we are confident that growth in personal well-being and in life-satisfaction among those who believe in a Spirit and Life Force are evident.
-Such effects co-occur in people who lack a belief in a God as well as among people who believe in a God as well as a Spirit or Life Force.
+  Over a decade, people who believed in a God and Spirit (\(\Delta \hat{y}\) = 0.55), people who believed in God and disbelieved in a Spirit (\(\Delta \hat{y}\) = 0.40), and people who believed in a spirit and disbelieved in a God (\(\Delta \hat{y}\) = 0.51) grew at up to twice the rate in personal well-being that disbelievers grew (\(\Delta \hat{y}\) = 0.26).
 \end{enumerate}
+
+Similarly, belief in a Spirit or Life Force predicts a trajectory of growth in life satisfaction over the decade (2010-2020);
+Over a decade, people who believed in a God and Spirit (\(\Delta \hat{y}\) = .35), people who believed in God and disbelieved in a Spirit (\(\Delta \hat{y}\) = .24), and people who believed in a spirit and disbelieved in a God (\(\Delta \hat{y}\) = 0.33) grew at up to twice the rate in personal well-being that disbelievers grew (\(\Delta \hat{y}\) = 0.18).
+Here to we find a growth rate that is twice as large for people who believe in a spirit or life force.
+
+Because These inferred growth rates for personal well being and life satisfaction are evident from repeated measures across 10 annual waves of nationally diverse data collection, we are confident that in New Zealand, spiritual beliefs predict strong increases in personal well-being and life satisfaction
 
@@
 Two fundamental question arise from the patterns we observe in this study.
-The first question is why those who believed in Spirit or Life Force were initially lower in their personal well-being and life-satisfaction in the initial years of this study in comparison to religious and non-religious skeptics?
+The first question is why those who believed in Spirit or Life Force were expected to be initially lower relative the populations that did not believe in a Spirit or Life Force?
 During the early years of New Zealand Attitudes and Values data collection New Zealand was emerging from the twin impacts of a global financial recession and a series of devastating earthquakes that destroyed 30\% of Christchurch, New Zealand's second largest city (Greaves et al., 2015).
@@
 Second, our task here has been to clarify whether beliefs predict growth in well-being.
-However, as we have indicated at the outset, the term \enquote{spirituality} applies to a family of concepts.
+However, as we have indicated at the outset, the term ``spirituality'' applies to a family of concepts.
 We have not attempted to model the diverse philosophical, social, and behavioral attributes within people who adhere to spirit beliefs.
 Moreover, people might identify as spiritual without having any accompanying belief in a Spirit or Life Force.
-The task of clarifying how distinct features across a diverse family of \enquote{spirituality} concepts affect perceptions of individual well-being remain intriguing horizons.
+The task of clarifying how distinct features across a diverse family of ``spirituality'' concepts affect perceptions of individual well-being remain intriguing horizons.
 
@@
 Our finding that, indeed, spiritual beliefs predict growth in perceived well-being and life satisfaction may hold practical importance.
-Previous research indicates that religious beliefs are associated with psychological well-being (Smith et al., 2003).
-Western populations are increasingly identifying as \enquote{spiritual but not religious} (Swinton, 2001; Zinnbauer \& Pargament, 2005).
+Previous research indicates that religious beliefs are associated with psychological well-being (Smith, McCullough, \& Poll, 2003).
+Western populations are increasingly identifying as ``spiritual but not religious'' (Swinton, 2001; Zinnbauer \& Pargament, 2005).
 Here, we observe that even among people who lack a traditional religious belief in God, believing in a Spirit or Life Force predicts incremental improvements in well-being that sum to practically large effects.
@@
 
-BH and JB conceived of the study.
-BH wrote the first draft of the manuscript.
-EW improved the approach with BH and JB and revised the manuscript.
-DD revised the manuscript.
-CS was responsible for NZAVS data collection and revised the manuscript.
-JB performed the data re-analysis and revised the manuscript.
+BH and JB conceived of the study. BH wrote the first draft of the manuscript. EW improved the approach. JB did the reanalysis. BH, DD, CS, EW, and JB revised the manuscript.
 
@@
 
-\newpage
-
-\hypertarget{appendix}{%
-\section{Appendix}\label{appendix}}
-
-\hypertarget{model-equations}{%
-\subsection{Model Equations}\label{model-equations}}
-
-\hypertarget{priors-for-the-personal-wellbeing-model}{%
-\subsubsection{Priors for the Personal Wellbeing model}\label{priors-for-the-personal-wellbeing-model}}
-
-Model priors for Personal Wellbeing including group-level ID intercept and beliefs slopes are as follows
-
-\[
-\begin{aligned}
-\operatorname{Personal Well Being}_{ij} \sim \text{Normal}(\mu_{ij},\sigma^2) \nonumber\\
-\mu_{ij}\sim \boldsymbol{\alpha +\beta} \nonumber\\
-\boldsymbol{\alpha} \sim \alpha_{0} +\alpha_j \nonumber\\
-\alpha_{0} \sim \text{StudentT}(3,7.2,2.5)\nonumber\\
-\boldsymbol{\beta}\sim \boldsymbol{b} + \beta_j \nonumber\\
-\boldsymbol{b}\sim\text{Normal}(0,2.5) \nonumber\\
-\begin{bmatrix}
-\alpha_j \\
-\beta_j
-\end{bmatrix}
-\sim 
-\text{MVNormal}
-\begin{pmatrix}
-\begin{bmatrix}
-0\\
-0
-\end{bmatrix}
-,\boldsymbol{S} \nonumber
-\end{pmatrix}\\
-\boldsymbol{S} = 
-\begin{pmatrix}
-0 & \sigma_{\beta_j} \nonumber
-\end{pmatrix} \boldsymbol{R} \begin{pmatrix}
-\sigma_{\alpha_j} & 0 \\
-0 & \sigma_{\beta_j} \nonumber
-\end{pmatrix}\\
-\sigma_{\alpha_j} \sim 
-\text{HalfCauchy}(0,2.5) \nonumber\\
-\sigma_{\beta_j} \sim 
-\text{HalfCauchy}(0,2.5) \nonumber\\
-\boldsymbol{R}\sim \text{LKJcorr}(1) \nonumber\\
-\sigma  \sim  \operatorname{HalfStudent T}^{+}(3, 0, 2.5) 
-\end{aligned}
-\]
-
-\newpage
-
-\hypertarget{priors-for-the-life-satisfaction-model}{%
-\subsubsection{Priors for the Life Satisfaction model}\label{priors-for-the-life-satisfaction-model}}
-
-Model priors for Life Satisfaction Model with including group-level ID intercept and belief slopes are as follows:
-
-\[
-\begin{aligned}
-\operatorname{Life Satisfaction}_{ij} \sim \text{Normal}(\mu_{ij},\sigma^2) \nonumber\\
-\mu_{ij}\sim \boldsymbol{\alpha +\beta} \nonumber\\
-\boldsymbol{\alpha} \sim \alpha_{0} +\alpha_j \nonumber\\
-\alpha_{0} \sim \text{StudentT}(3,5.5,2.5)\nonumber\\
-\alpha_j \sim \text{StudentT}+(3,0,2.5)\nonumber\\
-\boldsymbol{\beta}\sim \boldsymbol{b} + \beta_j \nonumber\\
-\boldsymbol{b}\sim\text{Normal}(0,2.5) \nonumber\\
-\begin{bmatrix}
-\alpha_j \\
-\beta_j
-\end{bmatrix}
-\sim 
-\text{MVNormal}
-\begin{pmatrix}
-\begin{bmatrix}
-0\\
-0
-\end{bmatrix}
-,\boldsymbol{S} \nonumber
-\end{pmatrix}\\
-\boldsymbol{S} = 
-\begin{pmatrix}
-\sigma_{\alpha_j} & 0 \\
-0 & \sigma_{\beta_j} \nonumber
-\end{pmatrix} \boldsymbol{R} \begin{pmatrix}
-\sigma_{\alpha_j} & 0 \\
-0 & \sigma_{\beta_j} \nonumber
-\end{pmatrix}\\
-\sigma_{\alpha_j} \sim 
-\text{HalfCauchy}(0,1) \nonumber\\
-\sigma_{\beta_j} \sim 
-\text{HalfCauchy}(0,1) \nonumber\\
-\boldsymbol{R}\sim \text{LKJcorr}(1) \nonumber\\
-\sigma  \sim  \operatorname{HalfStudent T}^{+}(3, 0, 2.5) 
-\end{aligned}
-\]
-\newpage
-
-\hypertarget{deterministic-model}{%
-\subsubsection{Deterministic model}\label{deterministic-model}}
-
-The deterministic portion of both the Personal Wellbeing and the Life Satisfaction models is as follows:
-\[
-\begin{aligned}
-& \quad  \boldsymbol{b} = \\
-& \quad \beta_{1}(\operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\
-&\quad \beta_{2}(\operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ +  \beta_{3}(\operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}})\ +\\
-& \quad \beta_{4}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\ 
-& \quad \beta_{4}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ + \\
-&\quad \beta_{6}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) \\ 
-\end{aligned}
-\]
 \newpage
@@
 \hypertarget{refs}{}
+\begin{CSLReferences}{1}{0}
 \leavevmode\hypertarget{ref-R-equatiomatic}{}%
-Anderson, D., \& Heiss, A. (2020). Equatiomatic: Transform models into 'latex' equations. Retrieved from \url{https://github.com/datalorax/equatiomatic}
+Anderson, D., \& Heiss, A. (2020). Equatiomatic: Transform models into 'LaTeX' equations. Retrieved from \url{https://github.com/datalorax/equatiomatic}
 
@@
 \leavevmode\hypertarget{ref-Atkinson2014-ex}{}%
-Atkinson, J., Salmond, C., \& Crampton, P. (2014). NZDep2013 index of deprivation. \emph{Wellington: Department of Public Health, University of Otago}.
+Atkinson, J., Salmond, C., \& Crampton, P. (2014). {NZDep2013} index of deprivation. \emph{Wellington: Department of Public Health, University of Otago}.
 
@@
 \leavevmode\hypertarget{ref-R-papaja}{}%
-Aust, F., \& Barth, M. (2020). \emph{papaja: Create APA manuscripts with R Markdown}. Retrieved from \url{https://github.com/crsh/papaja}
+Aust, F., \& Barth, M. (2020). \emph{{papaja}: {Create} {APA} manuscripts with {R Markdown}}. Retrieved from \url{https://github.com/crsh/papaja}
 
@@
 (2015).
-Fitting linear mixed-effects models using lme4.
+Fitting linear mixed-effects models using {lme4}.
 \emph{Journal of Statistical Software}, \emph{67}(1), 1--48.
 \url{https://doi.org/10.18637/jss.v067.i01}
+
+\leavevmode\hypertarget{ref-bell2015explaining}{}%
+Bell, A., \& Jones, K. (2015). Explaining fixed effects: Random effects modeling of time-series cross-sectional and panel data. \emph{Political Science Research and Methods}, \emph{3}(1), 133--153.
 
@@
 \leavevmode\hypertarget{ref-BRMSpackage}{}%
-Bürkner, P.-C. (2018). Advanced Bayesian multilevel modeling with the R package brms. \emph{The R Journal}, \emph{10}(1), 395--411. \url{https://doi.org/10.32614/RJ-2018-017}
+Bürkner, P.-C. (2018). Advanced {Bayesian} multilevel modeling with the {R} package {brms}. \emph{The R Journal}, \emph{10}(1), 395--411. \url{https://doi.org/10.32614/RJ-2018-017}
 
@@
 (2017).
-Extending extitR with extitC++: A Brief Introduction to extitRcpp.
+{Extending extit{R} with extit{C++}: A Brief Introduction to extit{Rcpp}}.
 \emph{PeerJ Preprints}, \emph{5}, e3188v1.
@@
 \leavevmode\hypertarget{ref-R-Rcpp_a}{}%
-Eddelbuettel, D., \& François, R. (2011). Rcpp: Seamless R and C++ integration. \emph{Journal of Statistical Software}, \emph{40}(8), 1--18. \url{https://doi.org/10.18637/jss.v040.i08}
+Eddelbuettel, D., \& François, R. (2011). {Rcpp}: Seamless {R} and {C++} integration. \emph{Journal of Statistical Software}, \emph{40}(8), 1--18. \url{https://doi.org/10.18637/jss.v040.i08}
 
@@
 Honaker, J., King, G., \& Blackwell, M.
-(2011a).
-Amelia II: A program for missing data.
+(2011).
+{Amelia II}: A program for missing data.
 \emph{Journal of Statistical Software}, \emph{45}(7), 1--47.
@@
 \leavevmode\hypertarget{ref-Honaker2011-yu}{}%
-Honaker, J., King, G., Blackwell, M., \& Others. (2011b). Amelia II: A program for missing data. \emph{J. Stat. Softw.}, \emph{45}(7), 1--47.
+Honaker, J., King, G., Blackwell, M., \& Others. (2011). Amelia {II}: A program for missing data. \emph{J. Stat. Softw.}, \emph{45}(7), 1--47.
 
@@
 \leavevmode\hypertarget{ref-Koenig2008-lv}{}%
-Koenig, H. G. (2008). Concerns about measuring ``spirituality'' in research. \emph{J. Nerv. Ment. Dis.}, \emph{196}(5), 349.
+Koenig, H. G. (2008). Concerns about measuring {``spirituality''} in research. \emph{J. Nerv. Ment. Dis.}, \emph{196}(5), 349.
 
@@
 \leavevmode\hypertarget{ref-Koenig2001-ow}{}%
-Koenig, H.
-G., Associate Professor of Psychology and Religious Studies Michael E McCullough, PhD, McCullough, M.
-E., Larson, D.
-B., \& Adjunct Professor of Psychiatry David B Larson.
-(2001).
-\emph{Handbook of religion and health}.
-Oxford University Press.
+Koenig, H. G., McCullough, Michael E, \& Larson, D. B. (2001). \emph{Handbook of religion and health}. Oxford University Press.
 
@@
 (2013).
-texreg: Conversion of statistical model output in R to LaTeX and HTML tables.
+{texreg}: Conversion of statistical model output in {R} to {LaTeX} and {HTML} tables.
 \emph{Journal of Statistical Software}, \emph{55}(8), 1--24.
@@
 \leavevmode\hypertarget{ref-R-sjPlot}{}%
-Lüdecke, D. (2020). \emph{SjPlot: Data visualization for statistics in social science}. Retrieved from \url{https://CRAN.R-project.org/package=sjPlot}
+Lüdecke, D. (2020). \emph{sjPlot: Data visualization for statistics in social science}. Retrieved from \url{https://CRAN.R-project.org/package=sjPlot}
 
@@
 S., Patil, I., \& Makowski, D.
-(2020a).
-Parameters: Extracting, computing and exploring the parameters of statistical models using R.
+(2020).
+Parameters: Extracting, computing and exploring the parameters of statistical models using {R}.
 \emph{Journal of Open Source Software}, \emph{5}(53), 2445.
@@
 S., Waggoner, P., \& Makowski, D.
-(2020b).
+(2020).
 See: Visualisation toolbox for 'easystats' and extra geoms, themes and color palettes for 'ggplot2'.
@@
 \leavevmode\hypertarget{ref-R-table1}{}%
-Rich, B. (2020). \emph{Table1: Tables of descriptive statistics in html}. Retrieved from \url{https://CRAN.R-project.org/package=table1}
+Rich, B. (2020). \emph{table1: Tables of descriptive statistics in HTML}. Retrieved from \url{https://CRAN.R-project.org/package=table1}
 
@@
 \leavevmode\hypertarget{ref-Sawatzky2005-rw}{}%
-Sawatzky, R., Ratner, P. A., \& Chiu, L. (2005). A Meta-Analysis of the relationship between spirituality and quality of life. \emph{Soc. Indic. Res.}, \emph{72}(2), 153--188.
+Sawatzky, R., Ratner, P. A., \& Chiu, L. (2005). A {Meta-Analysis} of the relationship between spirituality and quality of life. \emph{Soc. Indic. Res.}, \emph{72}(2), 153--188.
 
@@
 \leavevmode\hypertarget{ref-R-ggplot2}{}%
-Wickham, H. (2016). \emph{Ggplot2: Elegant graphics for data analysis}. Springer-Verlag New York. Retrieved from \url{https://ggplot2.tidyverse.org}
+Wickham, H. (2016). \emph{ggplot2: Elegant graphics for data analysis}. Springer-Verlag New York. Retrieved from \url{https://ggplot2.tidyverse.org}
 
@@
 (2019).
-Welcome to the tidyverse.
+Welcome to the {tidyverse}.
 \emph{Journal of Open Source Software}, \emph{4}(43), 1686.
@@
 
+\end{CSLReferences}
+
 \endgroup

### 2021-02-26 Supplement_And_Analysis.Rmd
fixes

--- 2020-11-30 Supplement_And_Analysis.Rmd
@@
 output            : "html_document" 
-
 editor_options: 
   chunk_output_type: console
+  markdown: 
+    wrap: 72
 ---
@@
 
-This supplement offers a step-by-step analysis for the "Longitudinal Evidence for Growth in Subjective Well-being from Spiritual Beliefs." The functions and other materials for this analysis can be found at the following GitHub link: [https://github.com/jacopastorius/nzavs_spiritbeliefwell-being](https://github.com/jacopastorius/nzavs_spiritbeliefwell-being)
+## Sample
+
+### Sampling Procedure -- NZAVS Time 1 (2009; conducted from 01.07.2009-14.10.2010)
+The Time 1 (2009) NZAVS contained responses from 6,518 participants sampled from the 2009 New Zealand electoral roll, who were currently residing in New Zealand (one can be registered to vote in New Zealand but living overseas).
+The electoral roll is publicly available for scientific research and in 2009 contained 2,986,546 registered voters.
+This represented all citizens over 18 years of age who were eligible to vote regardless of whether they chose to vote, barring people who had their contact details removed due to specific case-by-case concerns about privacy.
+The sample frame was spilt into three parts.
+Sample Frame 1 constituted a sample of 25,000 people from the electoral roll (4,060 respondents).
+Sample Frame 2 constituted a further 10,000 people from the electoral roll (1,609 respondents).
+Sample Frame 3 constituted 5,500 people randomly selected from meshblock area units of the country with a high proportion of Māori, Pacific Nations and Asian peoples (671 respondents).
+All three samples were drawn using a stratified random procedure in which 25,000, 10,000 and 5,500 unique households (postal addresses) were first drawn, and then one person per household was randomly selected.
+A further 178 people responded but did not provide contact details and so could not be matched to a sample frame.
+In sum, postal questionnaires were sent to 40,500 registered voters or roughly 1.36% of all registered voters in New Zealand.
+The overall response rate (adjusting for the address accuracy of the electoral roll and including anonymous responses) was 16.6%.
+
+### Sampling Procedure -- NZAVS Time 2 (2010; conducted from 15.10.2010-26.08.2011)
+
+The Time 2 (2010) NZAVS contained responses from 4,441 participants.
+The Time 2 (2010) NZAVS retained 4,425 from the initial Time 1 (2009) NZAVS sample of 6,518 participants, and included an additional 16 respondents who could not be matched to the Time 1 participant database (a retention rate of 67.9% over one year).
+Participants in the initial Time 1 (2009) sample were randomly selected from the New Zealand electoral roll (a national registry of registered voters).
+The response rate in the initial Time 1 sample, adjusting for the accuracy of the electoral roll and including anonymous responses was 16.6%.
+Participants were posted a copy of the questionnaire, with a second postal follow-up two months later.
+Participants who provided an email address were also emailed and invited to complete an online questionnaire if they preferred.
+
+### Sampling Procedure -- NZAVS Time 3 (2011; conducted from 27.08.2011-18.09.2012)
+
+The Time 3 (2011) NZAVS contained responses from 6,884 participants (3,918 retained from one or more previous wave, 2,966 new additions from booster sampling, and 4 unmatched participants or unsolicited opt-ins).
+The Time 3 (2011) NZAVS retained 3,918 from the initial Time 1 national probability sample (a 60.1% retention rate over two years).
+A further three participants who joined at Time 2 were also retained.
+Participants were posted a copy of the questionnaire, with a second postal follow-up two months later.
+Participants who provided an email address were also emailed and invited to complete an online version if they preferred.
+
+To boost sample size at Time 3 and compensate for sample attrition, a booster sample was recruited through an unrelated survey posted on the website of a major New Zealand newspaper in 2011.
+A total of 3,208 participants registered an initial expression of interest in being contacted to participate in the NZAVS via this survey.
+Participants in this non-random booster sample were emailed an invitation to participate in an online version of the NZAVS, and those who did not respond to the email were also sent a postal version of the questionnaire.
+A total of 2,966 participants completed the questionnaire when subsequently contacted (92.5%).
+This yielded a total sample size for the Time 3 (2012) NZAVS of 6,884 (3,918 retained from Time 1, 3 additions retained from opt-ins at Time 2, 2,966 recruited from the newspaper website at Time 3, and 4 opt-ins at Time 3).
+
+### Sampling Procedure -- NZAVS Time 3.5 (2012 mid-year)
+
+The Time 3.5 (2012 mid-year) NZAVS contained responses from 4,514 participants who completed a follow-up online-only questionnaire administered roughly six months following the full Time 3 (2011) questionnaire.
+The Time 3.5 sample was supplementary to the full Time 3 (2011) NZAVS.
+The sample frame included those participants who had provided an email address when completing the full Time 3 questionnaire earlier that year, as well as approximately 400-450 Pacific participants who were recruited informally via Pacifika networks.
+The sample included 1,977 retained participants from the initial Time 1 (2009) NZAVS random electoral roll sample, 2,113 participants from the non-random online newspaper website sample collected as part of the Time 3 (2011) NZAVS, a further 50 participants online newspaper website sample who has initially indicated but had not responded initially at T3 (but did responded when contacted again at Time 3.5), 197 participants who self-selected into the study or who were unable to be matched to a sample frame, and 177 additional Pacific participants recruited via Pacifika networks who complete a different version of the questionnaire focusing on Pacific identity and wellbeing.
+Sampling Procedure -- NZAVS Time 4 (2012; conducted from 19.09.2012-16.09.2013)
+
+### The Time 4 (2012) NZAVS contained responses from 12,179 participants (6,807 retained from one or more previous wave, 5,107 new additions from booster sampling, and 265 unmatched participants or unsolicited opt-ins).
+
+Informal analysis indicates that unsolicited opt-ins were often the partners of existing participants.
+The sample retained 4,053 participants from the initial Time 1 (2009) NZAVS of 6,518 participants (a retention rate of 62.2% over three years).
+The sample retained 5,762 participants from the full Time 3 (2011) sample (a retention rate of 83.7% from the previous year).
+Participants were posted a copy of the questionnaire, with a second postal follow-up two months later.
+Participants who provided an email address were also emailed and invited to complete an online version if they preferred.
+Non-respondents were emailed a follow-up reminder email approximately two months later.
+Three attempts were then made using each provided phone number (typically home and cell phone) to contact non-respondents to encourage participation.
+These attempts were made on separate days, approximately one week apart.
+When possible, a phone message was left for each phone number after the third attempt.
+Participants were also posted a pamphlet outlining recent findings from the study mid-way through the year.
+Finally, participants were posted a Season’s Greetings from the NZAVS research team, and informed that they had been automatically entered into a bonus seasonal grocery voucher prize draw for a total pool of $NZ 1,000).
+Participants were informed that the draw would happen automatically and winners contacted.
+The Season’s Greetings card also asked participants to contact us (online, email or phone) to let us know if any of their contact details had changed before the prize draw was conducted.
+These additional materials are presented by Huang, Greaves, and Sibley (2014) in an online NZAVS technical report.
+
+To boost sample size at Time 4 and increase sample diversity for subsequent waves, five independent booster samples using different sample frames were also conducted.
+Booster sampling was conducted without replacement (i.e., all people included in previous sample frames were identified and removed from the electoral roll before generation of the new sample frames).
+The first sample frame consisted of a randomly selected sample of 20,000 people from the 2012 New Zealand Electoral, and who were currently residing in New Zealand (one can be registered to vote in New Zealand but living overseas).
+A total of 2,429 participants responded to this booster sample (response rate = 12.33% when adjusting for the 98.5% accuracy of the 2012 electoral roll).
+The second sample frame consisted of a regional booster of 10,000 people randomly selected from people listed in the 2012 Electoral Roll who lived in the Auckland region.
+A total of 890 participants responded to this booster sample (adjusted response rate = 9.04).
+The Auckland region was oversampled because it is the fastest growing and most ethnically diverse region of the country with an increasing number of Asian and Pacific peoples in particular.
+The questionnaire used for this Auckland sample was longer than the standard NZAVS questionnaire, and contained additional unrelated questions that are not included in the NZAVS dataset (these related to the use of community facilities).
+Exit interviews conducted during
+
+Time 5 indicated that the longer length of this questionnaire may have contributed to the low response rate in this case. 
+The third sample frame consisted of 3,000 people randomly selected from the 2012 Electoral Roll who lived in the Christchurch region.
+A total of 332 participants responded to this booster sample (adjusted response rate = 11.24%).
+The Christchurch region was oversampled because it has experienced significant hardship and change due to the Christchurch earthquakes of 2010 and 2011 with many people moving out of the region (Statistics New Zealand, 2013) and problems with mail delivery with some city zones being placed under restricted entry due to safety concerns and considerable infrastructure destroyed).
+The fourth sample frame consisted of 9,000 respondents selected from meshblock area units across the country that were moderate-to-high in deprivation according to the index developed by Salmond, Crampton and Atkinson (2007).
+Regions with levels of deprivation were selected using scores on the decile-ranked NZ Deprivation index from 6-10, with 10 being the most deprived).
+This sample frame used scaled weighting so that people in increasingly deprived regions were increasingly more likely to be selected (with random sampling of people within regions that had a given level of deprivation).
+The scaling factor was as follows: ni = nbase * weighti, where nbase = 600, and weighti, ranged from 1 to 5 and increased by 1 for each one-unit increased in deprivation score.
+Thus, 600 people were randomly selected from regions with a deprivation score of 6, 1,200 people were randomly selected from regions with a deprivation score of 7, and so on.
+This sampling strategy was designed to increase the representativeness of the sample across regions with different levels of deprivation, as the NZAVS showed increased attrition in increasingly more deprived regions over the first three years of the study.
+A total of 767 participants responded to this booster sample (adjusted response rate = 8.65%).
+The fifth sample frame consisted of 9,000 people randomly selected from those who indicated on the 2012 Electoral Roll that they were of Māori ethnicity (ethnic affiliation as Māori is listed on the roll, but other ethnic affiliations are not).
+A total of 689 participants responded to this booster sample (adjusted response rate = 7.78%).
+The questionnaire administered to the Māori booster sample included questions specifically designed for Māori.
+
+### Sampling Procedure -- NZAVS Time 5 (2013; conducted from 17.09.2013-20.10.2014)
+
+The Time 5 (2013) NZAVS contained responses from 18,261 participants (10,502 retained from one or more previous wave, 7,579 new additions from booster sampling, and 179 unmatched participants or unsolicited opt-ins).
+Informal analysis indicates that unsolicited opt-ins were often the partners of existing participants.
+The sample retained 3,934 participants from the initial Time 1 (2009) NZAVS of 6,518 participants (a retention rate of 60.4% over four years).
+The sample retained 9,844 participants from the full Time 4 (2011) sample (a retention rate of 80.8% from the previous year).
+Participants were posted a copy of the questionnaire, with a second postal follow-up two months later.
+Participants who provided an email address were also emailed and invited to complete an online version if they preferred.
+As described in the Time 4 procedure, we offered a prize draw for participation, non-respondents were emailed and phoned multiple times, and all participants were posted a Season’s Greetings card from the NZAVS research team and informed that they had been automatically entered into a bonus seasonal grocery voucher prize draw.
+We also posted our yearly pamphlet summarizing key research findings published during the current wave of the study.
+
+To boost sample size and increase sample diversity for subsequent waves, two booster samples were also conducted by selecting people from the New Zealand electoral roll.
+As with previous booster samples, sampling was conducted without replacement (i.e., all people included in previous sample frames were identified and removed from the 2014 roll).
+The first sample frame consisted of 70,000 people aged from 18-60 randomly selected from the 2014 New Zealand Electoral Roll, who were currently residing in New Zealand (one can be registered to vote in New Zealand but living overseas).
+The New Zealand Electoral Roll contains participants’ date of birth (within a one-year window), and we limited our frame to people who 60 or younger, due to our aim of retaining participants for the following 15 years.
+A total of 7487 participants responded to this booster sample (response rate = 10.9% when adjusting for the 98.6% accuracy of the 2014 electoral roll).
+The second sample frame consisted of 1,500 people who were listed on the Electoral Roll as being of Maori ancestry, aged between 18-60 years of age, and currently residing in New Zealand.
+A total of 92 participants responded to this booster sample (response rate = 6.2% adjusting for electoral roll accuracy).
+
+### Sampling Procedure -- NZAVS Time 6 (2014; conducted from 21.10.2014-19.08.2015)
+
+The Time 6 (2014) NZAVS contained responses from 15,820 participants (15,740 retained from one or more previous wave, and 82 unmatched participants or unsolicited opt-ins).
+The sample retained 3,728 participants from the initial Time 1 (2009) NZAVS of 6,518 participants (a retention rate of 57.2% over five years).
+The sample retained 14,878 participants from the full Time 5 (2013) sample (a retention rate of 81.5% from the previous year).
+Participants who provided an email address were first emailed and invited to complete an online version if they preferred.
+Participants who did not complete the online version (or did not provide an email) were then posted a copy of the questionnaire, with a second postal follow-up two months later.
+We staggered the time of contact, so that participants who had completed the previous wave were contacted approximately one year after they last completed the questionnaire.
+As described in the Time 5 procedure, we offered a prize draw for participation, non-respondents were emailed and phoned multiple times, and all participants were posted a Season’s Greetings card from the NZAVS research team and informed that they had been automatically entered into a bonus seasonal grocery voucher prize draw.
+We also emailed participants an online pamphlet containing a series of video interviews with the researchers summarizing different research findings.
+
+### Sampling Procedure -- NZAVS Time 7 (2015; conducted from 20.08.2015-10.09.2016)
+
+The Time 7 (2015) NZAVS contained responses from 13,942 participants (13,941 retained from one or more previous wave, and 1 unmatched participant or unsolicited opt-ins).
+The sample retained 3,344 participants from the initial Time 1 (2009) NZAVS of 6,518 participants (a retention rate of 51.3% over five years).
+The sample retained 12,550 participants from the full Time 6 (2014) sample (a retention rate of 79.3% from the previous year).
+Participants who provided an email address were first emailed and invited to complete an online version if they preferred.
+Participants who did not complete the online version (or did not provide an email) were then posted a copy of the questionnaire, with a second postal follow-up two months later.
+We staggered the time of contact, so that participants who had completed the previous wave were contacted approximately one year after they last completed the questionnaire.
+As described in the Time 5 procedure, we offered a prize draw for participation, non-respondents were emailed and phoned multiple times, and all participants were posted a Season’s Greetings card from the NZAVS research team and informed that they had been automatically entered into a bonus seasonal grocery voucher prize draw.
+
+### Sampling Procedure -- NZAVS Time 8 (2016; conducted from 11.09.2016-12.08.2017)
+
+The Time 8 (2016) NZAVS contained responses from 21,936 participants (13,781 retained from one or more previous wave, 7,667 new additions from booster sampling, and 488 unmatched participants or unsolicited opt-ins).
+The sample retained 3,347 participants from the initial Time 1 (2009) NZAVS of 6,518 participants (a retention rate of 51.4%).
+The sample retained 11,933 participants from the full Time 7 (2015) sample (a retention rate of 85.6% from the previous year).
+Participants who provided an email address were first emailed and invited to complete an online version if they preferred.
+Participants who did not complete the online version (or did not provide an email) were then posted a copy of the questionnaire, with a second postal follow-up two months later.
+We staggered the time of contact, so that participants who had completed the previous wave were contacted approximately one year after they last completed the questionnaire.
+As described in the Time 4 procedure, we offered a prize draw for participation, non-respondents were emailed and phoned multiple times, and all participants were posted a Season’s Greetings card from the NZAVS research team and informed that they had been automatically entered into a bonus seasonal grocery voucher prize draw.
+To boost sample size and increase sample diversity for subsequent waves, a booster sample was also conducted by selecting people from the New Zealand electoral roll.
+As with previous booster samples, sampling was conducted without replacement (i.e., all people included in previous sample frames were identified and removed from the 2016 roll).
+The sample frame consisted of 80,000 people aged from 18-65 randomly selected from the 2016 New Zealand Electoral Roll, who were currently residing in New Zealand (one can be registered to vote in New Zealand but living overseas).
+The New Zealand Electoral Roll contains participants’ date of birth (within a one-year window), and we limited our frame to people who 65 or younger, due to our aim of retaining participants longitudinally.
+A total of 7667 participants responded to this booster sample (response rate = 9.7% when adjusting for the 98.6% accuracy of the 2016 electoral roll).
+
+### Sampling Procedure -- NZAVS Time 9 (2017; conducted from 13.08.2017-17.06.2018)
+
+The Time 9 (2017) NZAVS contained responses from 17,072 participants (16,931 retained from one or more previous wave, and 141 unmatched participants or unsolicited opt-ins).
+The sample retained 2,771 participants from the initial Time 1 (2009) NZAVS of 6,518 participants (a retention rate of 42.5% over five years).
+The sample retained 15,784 participants from the full Time 8 (2016) sample (a retention rate of 72.0% from the previous year).
+Participants who provided an email address were first emailed and invited to complete an online version if they preferred.
+Participants who did not complete the online version (or did not provide an email) were then posted a copy of the questionnaire, with a second postal follow-up two months later.
+We staggered the time of contact, so that participants who had completed the previous wave were contacted approximately one year after they last completed the questionnaire.
+We offered a prize draw for participation (five draws each for $1000 grocery vouchers, $5000 total prize pool).
+All participants were posted a Season’s Greetings card from the NZAVS research team and informed that they had been automatically entered into a bonus seasonal grocery voucher prize draw.
+
+The retention rate from Time 8 to Time 9 of 72% was notably lower than the ~80% achieved in previous few years.
+We had opted not to phone non-respondents during the Time 9 wave of data collection, and instead decided to let one-year past and then intensify phoning of non-respondents the following year.
+We decided on this approach in the hope that it might reduce ‘contact fatigue’, and hence increase the recovery rate and give more time (two years) for intermittent and non-respondents to become re-enthused about participating in the study.
+We thus decided to risk a lower retention rate in Time 9 with the hopes of getting a bounce back in the recovery rate during the following wave of data collection amongst these most hard-to-retain participants (see Satherley et al.
+2015, for a detailed analysis of the demographic and individual difference factors predicting retention, non-response and intermittent response).
+Our decision to reduce the risk of contact fatigue during Time 9 was also compounded by a major telecommunications carrier in New Zealand opting to discontinue their email servers in November 2017.
+Many of our participants had used these email services, and hence a large number of the emails in our database were rendered invalid.
+This increased the subsequent load on attempting to contact participants via postal mail and phone until contact could be made and our email database could be updated with participants’ new email addresses.
+
+### Sampling Procedure -- NZAVS Time 10 (2018; conducted from 18.06.2018-28.09.2019)
+
+The Time 10 (2018) NZAVS contained responses from 47,951 participants (18,010 retained from one or more previous wave.
+The sample retained 2,964 participants from the Time 1 (2009) sample (a retention rate of 45.5%).
+The sample retained 14,049 participants from Time 9 (2017; a retention rate of 82.3% from the previous year).
+Participants who provided an email address were first emailed and invited to complete an online version if they preferred.
+Participants who did not complete the online version (or did not provide an email) were then posted a copy of the questionnaire, with a second postal follow-up two months later.
+We staggered the time of contact, so that participants who had completed the previous wave were contacted approximately one year after they last completed the questionnaire.
+We offered a prize draw for participation (five draws each for \$1000 grocery vouchers, \$5000 total prize pool).
+All participants were posted a Season’s Greetings card from the NZAVS research team and informed that they had been automatically entered into a bonus seasonal grocery voucher prize draw.
+Participants were also emailed an eight-page newsletter about the study.
+
+To boost sample size and increase sample diversity for subsequent waves, a booster sample was conducted by selecting people from the New Zealand electoral roll.
+As with previous booster samples, sampling was conducted without replacement (i.e., people included in previous sample frames were identified and removed from the 2018 roll).
+The sample frame consisted of 325,000 people aged from 18-65 randomly selected from the 2018 New Zealand Electoral Roll, who were currently residing in New Zealand (one can be registered to vote in New Zealand but living overseas).
+The electoral roll contained ~3,250,000 registered voters.
+The New Zealand Electoral Roll contains participants’ date of birth (within a one-year window), and we limited our frame to people who 65 or younger, due to our aim of retaining participants longitudinally.
+We concurrently advertised the survey on Facebook via a $5000 paid promotion of a link to a YouTube video describing the NZAVS and the large booster sample we were conducting.
+The advertisement targeted men and women aged 18-65+ who lived in New Zealand and ran for 14 days.
+This paid promotion reached 147,296 people, with 4,721 link clicks (i.e., clicking to watch the video), according to Facebook.
+The goal of the paid promotion was twofold: (a) to increase name recognition of the NZAVS during the period in which questionnaires were being posted, and (b) to help improve retention by potentially reaching previous participants who happened to see the advertisement.
+A total of 29,293 participants who were contained in our sample frame completed the questionnaire (response rate = 9.2% when adjusting for the 98.2% accuracy of the 2018 electoral roll).
+A further 648 participants completed the questionnaire, but were unable to be matched to our sample frame (for example, due to a lack of contact information) or were unsolicited opt-ins.
+Informal analysis indicates that unsolicited opt-ins were often the partners of existing participants.
+
+### Sampling Procedure -- NZAVS Time 11 (2019; conducted from 29.09.2019-17.10.2020)
+
+The Time 11 wave was conducted during COVID-19 pandemic.
+Procedures thus differed in that there was an increased focus on online deliver using email reminders and extensive Facebook advertising, no Christmas card, and incomplete phoning of non-respondents.
+
+The Time 11 (2019) NZAVS contained responses from 42,684 participants (36,522 retained from one or more previous wave.
+The sample retained 2,506 participants from the Time 1 (2009) sample (a retention rate of 38.4%).
+The sample retained 34,782 participants from Time 10 (2018; a retention rate of 72.5% from the previous year).
+Participants who provided an email address were first emailed and invited to complete an online version if they preferred.
+Participants who did not complete the online version (or did not provide an email) were then posted a copy of the questionnaire, with a second postal follow-up two months later.
+We staggered the time of contact, so that participants who had completed the previous wave were contacted approximately one year after they last completed the questionnaire.
+A second reminder email was sent approximately four months following the first email attempt.
+We offered a prize draw for participation (five draws each for $1000 grocery vouchers, $5000 total prize pool).
+Participants were also emailed an eight-page newsletter about the study.
+As in past years, three attempts were made to phone non-respondents using each available cell and landline number.
+However, due to the university closure during COVID-19 lockdowns, phoning attempts were made for only 54% of the phoning pool (11,687 from a total of 21,636 non-respondents who provided at least one phone number).
+
+Two additional forms of recruitment were also introduced during Time 11.
+The first was a large  information box in the questionnaire (taking a full page on the paper version), which asked people: ‘Do you have a partner who would also like to join the NZAVS?’ with additional details for how partners might join the study (see questionnaire for the full text).
+The second was a Facebook advertisement.
+The advertisement targeted men and women aged 18-65+ who lived in New Zealand and ran from and 4th April 2020 – 4th July 2020 (overlapping with New Zeeland’s first lockdown period and recovery), and again from 18th August 2020 – 4th September (during the second Auckland lockdown).
+Given the unprecedented nature of the COVID-19 lockdowns, we thought it important to maximise sampling during these periods.
+The goal of the Facebook advertisement was threefold: (a) to increase name recognition of the NZAVS and remind people to complete the paper/online version already posted/emailed to them, (b) to help improve retention by potentially reaching previous lost participants who happened to see the advertisement, and (c) to recruit new participants (and also the partners of existing participants) while people were at home with some possibly having more free time during lockdown.
+This last goal was indirect and not explicitly stated it in the advertisement.
+
+The Facebook advertisement read as follows: “Participate in the New Zealand Attitudes and Values Study.
+Complete the 2020 Questionnaire online” with the body of text: “If you are part of the NZAVS, but have not heard from us in the last year, then please consider completing the 2020 questionnaire online.
+The study is more important than ever as we aim to understand the impact of COVID-19 on mental health, wellbeing and resilience in our communities.
+We wish you all the best at this time and hope you keep well and stay safe.” This paid promotion reached 883,969 people, with 37,850 link clicks (i.e., clicking the link for the Qualtrics survey) according to Facebook.
+A total of 6106 people continued complete the questionnaire and provide full contact details, and were thus included in the dataset (4734 were new participants opting in to the study, and 1372 were previously ‘lost’ participants).
+
+### References
+Huang, Y., Greaves, L. M., & Sibley, C. G., (2014). Archive of additional (non-questionnaire) materials posted to participants. NZAVS Technical Documents, e07. Available at: www.nzavs.auckland.ac.nz
+
+Salmond, C., Crampton, P., & Atkinson, J. (2007). NZDep2006 Index of Deprivation User’s Manual. University of Otago.
+
+Satherley, N., Milojev, P., Greaves, L.
+M., Huang, Y., Osborne, D., Bulbulia, J., & Sibley, C.
+G.
+(2015).
+Demographic and psychological predictors of panel attrition: evidence from the New Zealand Attitudes and Values Study.
+PLoS ONE, 10, e0121950.
+
+Statistics New Zealand (2013). Glossary of Terms. Statistics New Zealand.
+
+Statistics New Zealand (2014). Quickstats about great Christchurch. Statistics New Zealand.
+
+## Analysis
+
+This supplement offers a step-by-step analysis for the "Longitudinal
+Evidence for Growth in Subjective Well-being from Spiritual Beliefs."
+The functions and other materials for this analysis can be found at the
+following GitHub link:
+<https://github.com/jacopastorius/nzavs_spiritbeliefwell-being>
 
@@
 
-We use functions to automate repetition in coding, to avoid errors from such repetition, and to make the data analysis easier to follow.
-
-This function reads our data into my machine. 
-```{r, read_data, cache=TRUE}
+Next we read the data:
+
+```{r read_data, cache=TRUE}
 # custom function for my machine to read data. you'll need to read data to your machine. tweak function in the `funs.R` script for your machine
@@
 
+# pa<-d%>%
+#  dplyr::filter(YearMeasured==1)%>%
+#     dplyr::group_by(Id) %>%
+#     dplyr::filter(n() !=0)%>%
+#     dplyr::ungroup(Id)
+# length(unique(pa$Id))
+# table(pa$Wave)
+
+# d2<-data_clean_spirit_wellbeing(pa,0) # d = dataframe, 3 = number of waves
+# table(d2$Wave)
+
 # function to clean data, specify number of waves as 3 
-d_3<-data_clean_spirit_wellbeing(d,3) # d = dataframe, 3 = number of waves
+d_3<-data_clean_spirit_wellbeing(d,9) # d = dataframe, 3 = number of waves
 # function to count uinque ides
@@
 table(d_3$Beliefs)
-# years Note that we use a continuous response measure particpants. 
+# years Note that we use a continuous response measure participants. 
 range(d_3$years)
 # html talbe
-demographic_table(d_3)
+#demographic_table(d_3)
+
+d_3 <- cbind(
+  d_3,
+  demean(d_3, select = c("Age"), group = "Id"
+))
+
+d_3<-d_3%>%
+  dplyr::mutate(Age_betweenC = scale(Age_between, scale = FALSE))
+
+```
+
+Breakdown of participants by wave
+```{r}
+table(d_3$Wave)
 ```
@@
 Sample details are as follows:
+
 ```{r sample details}
@@
 <!-- ```{r  badlatextable} badd tables -->
+
 <!-- library("arsenal") -->
+
 <!-- summary(arsenal::tableby(Wave ~ Age + -->
+
 <!--                    NZdep + -->
+
 <!--                    Education + -->
+
 <!--                    Employed + -->
+
 <!--                    Ethnic_Categories + -->
+
 <!--                    Male  + -->
+
 <!--                    Has_Partner  + -->
+
 <!--                    Political_Conservativism + -->
+
 <!--                    Urban + -->
+
 <!--                    LIFESAT + PWI + includeNA(Beliefs), data = d_3,digits = 2, total=FALSE, format.p = FALSE, digits.p = 0))%>% -->
@@
 # https://www.datanovia.com/en/lessons/compute-summary-statistics-in-r/
+
 ```
@@
 <!-- ```{r, include=FALSE} -->
+
 <!-- beliefs_table(d_3) -->
+
 <!-- ``` -->
@@
 <!-- ```{r latex_data_table, include=FALSE} -->
+
 <!-- # you might need to install the dev verion.//remotes::install_github("tysonstanley/furniture") -->
+
 <!-- demographic_table_latex() # note the furniture package conflicts with the table1 package. -->
+
 <!-- d_3%>% -->
+
 <!--   group_by(Wave)%>% -->
+
 <!-- furniture::table1( -->
+
 <!--                   Age,  -->
+
 <!--                   NZdep, -->
+
 <!--                   Education, -->
+
 <!--                   Employed, -->
+
 <!--                   Ethnic_Categories, -->
+
 <!--                   Male,Has_Partner,  -->
+
 <!--                   Political_Conservativism, -->
+
 <!--                   Urban,  -->
+
 <!--                   Beliefs,  -->
+
 <!--                   LIFESAT, -->
+
 <!--                   PWI, -->
+
 <!--                   overall=F) -->
@@
 <!-- dat3<-d_3%>%dplyr::select( Age, NZdep, -->
+
 <!--                   Education, -->
+
 <!--                   Employed, -->
+
 <!--                   Ethnic_Categories, -->
+
 <!--                   Male,Has_Partner, -->
+
 <!--                   Political_Conservativism, -->
+
 <!--                   Urban, -->
+
 <!--                   Beliefs, -->
+
 <!--                   LIFESAT, -->
+
 <!--                   PWI,Wave) -->
+
 <!-- stargazer::stargazer(dat3) -->
+
 <!-- library(modelsummary) -->
+
 <!--   modelsummary::datasummary_df(ALL(Age) ~ Mean + SD + Histogram, data=dat3) -->
-<!-- ``` -->
-
-## How many responses waves to select?
-
-We selected a minimum of three waves because we want to recover a longitudinal sample in which a slope for the effect of the years $\times$ beliefs effect might be recovered.
-However, we needed to consider whether our selection of three waves was sensitive to that degree of researcher freedom.
-So we also prepared 4-8 years of data frames for separate analyses.
-
-```{r, include = TRUE}
-# create datasets to test robustness to years
-d_4<-data_clean_spirit_wellbeing(d,4) # 4 years
-d_5<-data_clean_spirit_wellbeing(d,5) # 5 years
-d_6<-data_clean_spirit_wellbeing(d,6) # 6 years
-d_7<-data_clean_spirit_wellbeing(d,7) # 7 years
-d_8<-data_clean_spirit_wellbeing(d,8) # 7 years
-
-# function to count uinque number of ID's
-show_unique_id(d_3$Id) 
-show_unique_id(d_4$Id)
-show_unique_id(d_5$Id) 
-show_unique_id(d_6$Id) 
-show_unique_id(d_7$Id) 
-show_unique_id(d_8$Id)
-
-# function to count waves for participants
-count_waves_participants(d_3)
-count_waves_participants(d_4)
-count_waves_participants(d_5)
-count_waves_participants(d_6)
-count_waves_participants(d_7)
-count_waves_participants(d_8)
-```
-
-The analyses for these waves is described below.
-These analyses reveal that the inference of the paper is not sensitive to the threshold number of waves for participant selection.
-Which is to say, we recover the same inference with an 8-wave selection criterion.
+
+<!-- ``` -->
+
+<!-- ## How many waves to select? We are using 10 years -->
+
+<!-- We selected a minimum of three waves because we want to recover a -->
+
+<!-- longitudinal sample in which a slope for the effect of the years -->
+
+<!-- $\times$ beliefs effect might be recovered. However, we needed to -->
+
+<!-- consider whether our selection of three waves was sensitive to that -->
+
+<!-- degree of researcher freedom. So we also prepared 4-8 years of data -->
+
+<!-- frames for separate analyses. -->
+
+<!-- ```{r, include = TRUE} -->
+
+<!-- # create datasets to test robustness to years -->
+
+<!-- # d_4<-data_clean_spirit_wellbeing(d,4) # 4 years -->
+
+<!-- # d_5<-data_clean_spirit_wellbeing(d,5) # 5 years -->
+
+<!-- # d_6<-data_clean_spirit_wellbeing(d,6) # 6 years -->
+
+<!-- # d_7<-data_clean_spirit_wellbeing(d,7) # 7 years -->
+
+<!-- # d_8<-data_clean_spirit_wellbeing(d,8) # 7 years -->
+
+<!-- #  -->
+
+<!-- # # function to count uinque number of ID's -->
+
+<!-- # show_unique_id(d_3$Id)  -->
+
+<!-- # show_unique_id(d_4$Id) -->
+
+<!-- # show_unique_id(d_5$Id)  -->
+
+<!-- # show_unique_id(d_6$Id)  -->
+
+<!-- # show_unique_id(d_7$Id)  -->
+
+<!-- # show_unique_id(d_8$Id)  -->
+
+<!-- #  -->
+
+<!-- # # function to count waves for participants -->
+
+<!-- # count_waves_participants(d_3) -->
+
+<!-- # count_waves_participants(d_4) -->
+
+<!-- # count_waves_participants(d_5) -->
+
+<!-- # count_waves_participants(d_6) -->
+
+<!-- # count_waves_participants(d_7) -->
+
+<!-- # count_waves_participants(d_8) -->
+
+<!-- ``` -->
+
+<!-- The analyses for these waves is described below. These analyses reveal -->
+
+<!-- that the inference of the paper is not sensitive to the threshold number -->
+
+<!-- of waves for participant selection. Which is to say, we recover the same -->
+
+<!-- inference with an -wave selection criterion. -->
 
@@
 
-Where $Z$ is a third variable, $X$ is the predictor, and $Y$ is the outcome, there are at least four types of confounding that might arise:
+Where $Z$ is a third variable, $X$ is the predictor, and $Y$ is the
+outcome, there are at least four types of confounding that might arise:
 
@@
 2.  $X$ causes $Z$ and $Z$ causes $Y$: $X \rightarrow Z \rightarrow Y$
-3. $X$ causes $Z$ and $Y$ causes $Z$: $X \rightarrow Z; Y \rightarrow Z$
-4. $X$ causes $Z$; $Z$ causes $D$; and we condition on D (the descendent of $Z$: $Y$ causes $Z$: $Z$ \rightarrow D; $X \rightarrow Z; Y \rightarrow Z$ (see: [@mcelreath2020statistical], Chapter 6).
-
-We do not wish to exclude third variable explanations of the kind that arise from the first type of confounding.
-The second type of counfounding can mask a relationship.
-The third and fourth type of confounding illustrate how conditioning on a covariates may confound by *opening* a relationship between the $X$ and $Y$ indicators in a regression analysis.
-Such confounding is particularly worrying in the case of longitudinal research.
-
-To address confounding we first obtained the New Zealand Attitudes and Values Study standard demographic indicators recommended to address confounding: [NZAVS modelling standards](https://cdn.auckland.ac.nz/assets/psych/about/our-research/nzavs/NZAVSTechnicalDocuments/NZAVS-Technical-Documents-e11-Default-Statistical-Standards.pdf).
-Next we modelled potential relationships between the coefficients within the model using the daggity package in R, and plotting results using the daggity and ggdag packages in R.
+3.  $X$ causes $Z$ and $Y$ causes $Z$:
+    $X \rightarrow Z; Y \rightarrow Z$
+4.  $X$ causes $Z$; $Z$ causes $D$; and we condition on D (the
+    descendent of $Z$: $Y$ causes $Z$: $Z$ \rightarrow D;
+    $X \rightarrow Z; Y \rightarrow Z$ (see:
+    [@mcelreath2020statistical], Chapter 6).
+
+We do not wish to exclude third variable explanations of the kind that
+arise from the first type of confounding. The second type of
+counfounding can mask a relationship. The third and fourth type of
+confounding illustrate how conditioning on a covariates may confound by
+*opening* a relationship between the $X$ and $Y$ indicators in a
+regression analysis. Such confounding is particularly worrying in the
+case of longitudinal research.
+
+To address confounding we first obtained the New Zealand Attitudes and
+Values Study standard demographic indicators recommended to address
+confounding: [NZAVS modelling
+standards](https://cdn.auckland.ac.nz/assets/psych/about/our-research/nzavs/NZAVSTechnicalDocuments/NZAVS-Technical-Documents-e11-Default-Statistical-Standards.pdf).
+Next we modelled potential relationships between the coefficients within
+the model using the daggity package in R, and plotting results using the
+daggity and ggdag packages in R.
 
@@
 # obtain dag
+range(d_3$Age_within)
 tidy_ggdag <- dagify(
-  WB ~ belief + years + partner + nzdep + urban + male + pols + empl,
-  WB ~~ partner + years,
-  belief ~ years + age + male + ethn,
-  partner ~ nzdep + age + belief, 
-  nzdep ~ empl,
-  pols ~ age  + empl + ethn + years,
-  empl ~ nzdep + edu + ethn,
+  WB ~ belief + age_within + age_between + partner + nzdep + urban + male + pols + empl,
+  WB ~~ partner,
+  belief ~ age_within + age_between + male + ethn,
+  partner ~ nzdep + age_within + age_between + belief, 
+  nzdep ~ empl + age_within + age_between,
+  pols ~ age_within + age_between + empl + ethn + years,
+  empl ~  edu + ethn + age_within + age_between,
   exposure =  "belief",
@@
 ggdag::ggdag_adjustment_set(tidy_ggdag, node_size = 14) + 
-  theme(legend.position = "bottom") 
-```
-
-This method reveals two adjustments sets: \{age, employment, male, political conservativism, and time\}, and \{age, ethnicty, male, and time.\}  We report the second set because employment is likely to contain more measurement error: some are not employed because they cannot find employment, others because they are not seeking employment (e.g.
-retirement).
-However we report both model in our frequentist analysis.
-Neither model choice affects the results.
-We also report the simple unadjusted model and the fully saturated model (with potential collider and masking effects).
-
-Importantly, in the fully-saturated model -- which would be typical in a longitudinal regression analysis -- attenuates the growth in personal well-being and in life satisfaction among those who are skeptical in a spirit or Life-fore.
-The inference of this study is not sensitive to this difference because the slopes differ for spirit beliefs across all models.
-However, we do not wish to infer that skepticism about a spirit or life force entirely prevents growth in personal well-being or life-satisfaction.
-We advise that researchers do not introduce variables into regression models without systematically.
-evaluating potential confounding.
+  theme(legend.position = "bottom")  + theme_dag_gray()
+```
+
+This method reveals two adjustments sets: {age, employment, male,
+political conservativism, and time}, and {age, ethnicty, male, and
+time.} We report the second set because employment is likely to contain
+more measurement error: some are not employed because they cannot find
+employment, others because they are not seeking employment (e.g.
+retirement). However we report both model in our frequentist analysis.
+Neither model choice affects the results. We also report the simple
+unadjusted model and the fully saturated model (with potential collider
+and masking effects).
+
+Importantly, in the fully-saturated model -- which would be typical in a
+longitudinal regression analysis -- attenuates the growth in personal
+well-being and in life satisfaction among those who do not believe in a
+spirit or Life-fore. The inference of this study is not sensitive to
+this difference because the slopes differ for spirit beliefs across all
+models. However, we do not wish to infer that skepticism about a spirit
+or life force entirely prevents growth in personal well-being or
+life-satisfaction. We advise that researchers do not introduce variables
+into regression models without systematically. evaluating potential
+confounding.
 
@@
 
-Below we write the model in a notation that explicitly states the the default weakly informative priors that the BRMS package produces.
-And indicated in the model we include group-level indicator for variation in subject-level intercepts $\alpha_j$, a group-level indicator for subject-level time (or "growth")  slope  $\beta_j$, and we model the co-variance of $\alpha_j$, Below we write the model in a notation that explicitly states the the default weakly informative priors that the BRMS package produces.
-And indicated in the model we include group-level indicator for variation in subject-level intercepts $\alpha_j$, a group-level indicator for subject-level time (or "growth")  slope  $\beta_j$, and we model the co-variance of $\alpha_j$ and  $\beta_j$ which for computational reasons is estimate as a correlation matrix of (${\bf S}$} and ${\bf R}$).
+Below we write the model in a notation that explicitly states the the
+default weakly informative priors that the BRMS package produces. And
+indicated in the model we include group-level indicator for variation in
+subject-level intercepts $\alpha_j$, a group-level indicator for
+subject-level time (or "growth") slope $\beta_j$, and we model the
+co-variance of $\alpha_j$, Below we write the model in a notation that
+explicitly states the the default weakly informative priors that the
+BRMS package produces. And indicated in the model we include group-level
+indicator for variation in subject-level intercepts $\alpha_j$, a
+group-level indicator for subject-level time (or "growth") slope
+$\beta_j$, and we model the co-variance of $\alpha_j$ and $\beta_j$
+which for computational reasons is estimate as a correlation matrix of
+(${\bf S}$} and ${\bf R}$).
 
@@
 
-First we report priors for the Personal Well-Being model. Note that we model the correlation of the id-level intercept and id-level year-slope.
-
-\[
+First we report priors for the Personal Well-Being model. Note that we
+model the correlation of the id-level intercept and id-level year-slope.
+
+$$
 \begin{aligned}
@@
 \end{aligned}
-\]
+$$
 
@@
 
-Model priors for the  Life Satisfaction model. Here, again, we model the correlation of the id-level intercept and id-level year-slope.
-
-\[
+Model priors for the Life Satisfaction model. Here, again, we model the
+correlation of the id-level intercept and id-level year-slope.
+
+$$
 \begin{aligned}
@@
 
-The deterministic part of both the Personal Well-Being and the Life Satisfaction models is as follows.
-
-\[
+The deterministic part of both the Personal Well-Being and the Life
+Satisfaction models is as follows.
+
+$$
 \begin{aligned}
 & \quad  \boldsymbol{b} = \\
-& \quad \beta_{1}(\operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\
-&\quad \beta_{2}(\operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\  + \\
-& \quad  \beta_{3}(\operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}})\ +\\
-&\quad \beta_{5}(\operatorname{Age in Decades (C)})\  +\\ 
-& \quad  \beta_{6}(\operatorname{Ethnicity}_{\operatorname{Maori}})\ +\\
-&\quad \beta_{7}(\operatorname{Ethnicity}_{\operatorname{Pacific}})\ +  \\ 
-& \quad  \beta_{8}(\operatorname{Ethnicity}_{\operatorname{Asian}})\ +\\
-& \quad \beta_{4}(\operatorname{Male} )+ \\ 
-& \quad \beta_{4}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\ 
-& \quad \beta_{5}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ + \\
-&\quad \beta_{6}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) \\ 
+& \quad \beta_{1}(\operatorname{Beliefs}_{\operatorname{Spirit~excludes~God}})\ + \\
+&\quad \beta_{2}(\operatorname{Beliefs}_{\operatorname{God~and~Spirit}})\  + \\
+& \quad  \beta_{3}(\operatorname{Beliefs}_{\operatorname{God~excludes~Spirit}})\ +\\
+&\quad \beta_{4}(\operatorname{Aging-within-subjects (centered~within)})\  +\\ 
+&\quad \beta_{5}(\operatorname{Age-between-subjects (centered~between)})\  +\\ 
+& \quad  \beta_{5}(\operatorname{Ethnicity}_{\operatorname{Maori}})\ +\\
+&\quad \beta_{6}(\operatorname{Ethnicity}_{\operatorname{Pacific}})\ +  \\ 
+& \quad  \beta_{7}(\operatorname{Ethnicity}_{\operatorname{Asian}})\ +\\
+& \quad \beta_{8}(\operatorname{Male} )+ \\ 
+& \quad \beta_{9}(\operatorname{Aging-within-subjects} \times \operatorname{Beliefs}_{\operatorname{Spirit~excludes~God}})\ + \\ 
+& \quad \beta_{10}(\operatorname{Aging-within-subjects} \times \operatorname{Beliefs}_{\operatorname{God~and~Spirit}})\ + \\
+&\quad \beta_{11}(\operatorname{Aging-within-subjects} \times \operatorname{Beliefs}_{\operatorname{God~Excludes~Spirit}}) \\ 
 \end{aligned}
-\]
-
-To close the "back door" on potential confounds we create a regression model
+$$
+
+To assess potential confounding we write a DAG for our theoretically
+postulated model.
+
 ```{r dag}
@@
 tidy_ggdag <- dagify(
-  WB ~ belief + time + age + partner + nzdep + urban + male + pols + empl, # what might affect wellbeing?
+  WB ~ belief + age_within + age_between + partner + nzdep + urban + male + pols + empl, # what might affect wellbeing?
   WB ~~ partner, # bi-causal path
-  belief ~ time + age + male + ethn, # predictors of beliefs
-  partner ~ nzdep + age + belief,  # predictors of having a partner
+  belief ~ age_within  + age_between  + male + ethn, # predictors of beliefs
+  partner ~ nzdep + age_within  + age_between + belief,  # predictors of having a partner
   nzdep ~ empl,  # predictor of deprivation
-  pols ~ age  + empl + ethn, # predictor of political orientation
-  empl ~ nzdep + edu + ethn, # predictor of employment
+  pols ~ age_within  + age_between  + empl + ethn, # predictor of political orientation
+  empl ~ nzdep + edu + ethn + age_within  + age_between, # predictor of employment
   exposure =  "belief",
   outcome =   "WB")%>%
-  tidy_dagitty()
+  tidy_dagitty() + theme_dag_gray()
 
@@
 
-# Bayesian counfounder model
+## Priors in the Bayesian Model
+
+### Priors for the Personal Wellbeing model
+
+Model priors for Personal Wellbeing including group-level ID intercept
+and beliefs slopes are as follows:
+
+$$
+\begin{aligned}
+\operatorname{Personal Well Being}_{ij} \sim \text{Normal}(\mu_{ij},\sigma^2) \nonumber\\
+\mu_{ij}\sim \boldsymbol{\alpha +\beta} \nonumber\\
+\boldsymbol{\alpha} \sim \alpha_{0} +\alpha_j \nonumber\\
+\alpha_{0} \sim \text{StudentT}(3,7.2,2.5)\nonumber\\
+\boldsymbol{\beta}\sim \boldsymbol{b} + \beta_j \nonumber\\
+\boldsymbol{b}\sim\text{Normal}(0,2.5) \nonumber\\
+\begin{bmatrix}
+\alpha_j \\
+\beta_j
+\end{bmatrix}
+\sim 
+\text{MVNormal}
+\begin{pmatrix}
+\begin{bmatrix}
+0\\
+0
+\end{bmatrix}
+,\boldsymbol{S} \nonumber
+\end{pmatrix}\\
+\boldsymbol{S} = 
+\begin{pmatrix}
+0 & \sigma_{\beta_j} \nonumber
+\end{pmatrix} \boldsymbol{R} \begin{pmatrix}
+\sigma_{\alpha_j} & 0 \\
+0 & \sigma_{\beta_j} \nonumber
+\end{pmatrix}\\
+\sigma_{\alpha_j} \sim 
+\text{HalfCauchy}(0,2.5) \nonumber\\
+\sigma_{\beta_j} \sim 
+\text{HalfCauchy}(0,2.5) \nonumber\\
+\boldsymbol{R}\sim \text{LKJcorr}(1) \nonumber\\
+\sigma  \sim  \operatorname{HalfStudent T}^{+}(3, 0, 2.5) 
+\end{aligned}
+$$
+
+### Priors for the Life Satisfaction model
+
+Model priors for Life Satisfaction Model with including group-level ID
+intercept and belief slopes are as follows:
+
+$$\begin{aligned}
+\operatorname{Life Satisfaction}_{ij} \sim \text{Normal}(\mu_{ij},\sigma^2) \nonumber\\
+\mu_{ij}\sim \boldsymbol{\alpha +\beta} \nonumber\\
+\boldsymbol{\alpha} \sim \alpha_{0} +\alpha_j \nonumber\\
+\alpha_{0} \sim \text{StudentT}(3,5.5,2.5)\nonumber\\
+\alpha_j \sim \text{StudentT}+(3,0,2.5)\nonumber\\
+\boldsymbol{\beta}\sim \boldsymbol{b} + \beta_j \nonumber\\
+\boldsymbol{b}\sim\text{Normal}(0,2.5) \nonumber\\
+\begin{bmatrix}
+\alpha_j \\
+\beta_j
+\end{bmatrix}
+\sim 
+\text{MVNormal}
+\begin{pmatrix}
+\begin{bmatrix}
+0\\
+0
+\end{bmatrix}
+,\boldsymbol{S} \nonumber
+\end{pmatrix}\\
+\boldsymbol{S} = 
+\begin{pmatrix}
+\sigma_{\alpha_j} & 0 \\
+0 & \sigma_{\beta_j} \nonumber
+\end{pmatrix} \boldsymbol{R} \begin{pmatrix}
+\sigma_{\alpha_j} & 0 \\
+0 & \sigma_{\beta_j} \nonumber
+\end{pmatrix}\\
+\sigma_{\alpha_j} \sim 
+\text{HalfCauchy}(0,1) \nonumber\\
+\sigma_{\beta_j} \sim 
+\text{HalfCauchy}(0,1) \nonumber\\
+\boldsymbol{R}\sim \text{LKJcorr}(1) \nonumber\\
+\sigma  \sim  \operatorname{HalfStudent T}^{+}(3, 0, 2.5) 
+\end{aligned}$$
+
+### Deterministic model
+
+The deterministic portion of both the Personal Wellbeing and the Life
+Satisfaction models is as follows:
+
+$$\begin{aligned}
+& \quad  \boldsymbol{b} = \\
+& \quad \beta_{1}(\operatorname{Beliefs}_{\operatorname{Spirit~excludes~God}})\ + \\
+&\quad \beta_{2}(\operatorname{Beliefs}_{\operatorname{God~and~Spirit}})\  + \\
+& \quad  \beta_{3}(\operatorname{Beliefs}_{\operatorname{God~excludes~Spirit}})\ +\\
+&\quad \beta_{4}(\operatorname{Aging-within-subjects (centered~within)})\  +\\ 
+&\quad \beta_{5}(\operatorname{Age-between-subjects (centered~between)})\  +\\ 
+& \quad  \beta_{5}(\operatorname{Ethnicity}_{\operatorname{Maori}})\ +\\
+&\quad \beta_{6}(\operatorname{Ethnicity}_{\operatorname{Pacific}})\ +  \\ 
+& \quad  \beta_{7}(\operatorname{Ethnicity}_{\operatorname{Asian}})\ +\\
+& \quad \beta_{8}(\operatorname{Male} )+ \\ 
+& \quad \beta_{9}(\operatorname{Aging-within-subjects} \times \operatorname{Beliefs}_{\operatorname{Spirit~excludes~God}})\ + \\ 
+& \quad \beta_{10}(\operatorname{Aging-within-subjects} \times \operatorname{Beliefs}_{\operatorname{God~and~Spirit}})\ + \\
+&\quad \beta_{11}(\operatorname{Aging-within-subjects} \times \operatorname{Beliefs}_{\operatorname{God~Excludes~Spirit}}) \\ 
+\end{aligned}$$
+
+## Bayesian causal model
 
 ```{r setmodelformulaSAT}
-forumula_bayesPWI_DAG <- bf(PWI ~  yearsC * Beliefs +  Age_in_Decades_C + Ethnic_Categories + Male + (1+ yearsC|Id))
-forumula_bayesLIFESAT_DAG <- bf(LIFESAT ~   yearsC * Beliefs +  Age_in_Decades_C + Ethnic_Categories + Male + (1+ yearsC|Id))
+forumula_bayesPWI_DAG <- bf(PWI ~  Beliefs *  Age_within +  Age_betweenC  + Ethnic_Categories + Male + (1 + Age_within|Id))
+forumula_bayesLIFESAT_DAG <- bf(LIFESAT ~   Beliefs *  Age_within +  Age_betweenC  + Ethnic_Categories + Male + (1 + Age_within|Id))
 ```
@@
 # PWI
-mod.1_bDAG <-brm(forumula_bayesPWI_DAG, data= d_3, file ="mod.1_bDAG.rds", cores = 8) # note these are very large files
-summary(mod.1_bDAG)
+# mod.1_bDAG <-brm(forumula_bayesPWI_DAG, data= d_3, file ="mod.1_bDAG.rds", cores = 8) # note these are very large files
+# summary(mod.1_bDAG)
+
+        # REDO
+mod.1_bDAGR <-brm(forumula_bayesPWI_DAG, data= d_3, file ="mod.1_bDAGR.rds", cores = 8) # note these are very large files
+#summary(mod.1_bDAGR)
 ```
@@
 # LIFE SAT
-mod.2_bDAG <-brm (forumula_bayesLIFESAT_DAG, data= d_3,cores = 8, file ="mod.2_bDAG.rds")
+mod.2_bDAGR <-brm (forumula_bayesLIFESAT_DAG, data= d_3,cores = 8, file ="mod.2_bDAGR.rds")
+summary(mod.2_bDAGR)
 ###
-tab_model(mod.1_bDAG,mod.2_bDAG)
 # note these are very large files
@@
 # cannot run on my laptop
-pp_1<-rstanarm::pp_check(mod.1_bDAG) + ylab("Personal well-being")
-pp_2<-rstanarm::pp_check(mod.2_bDAG) + ylab("Life Satisfaction") ## life sat suggests an ordinal model
+pp_1<-rstanarm::pp_check(mod.1_bDAGR) + ylab("Personal well-being")
+pp_2<-rstanarm::pp_check(mod.2_bDAGR) + ylab("Life Satisfaction") ## life sat suggests an ordinal model
 pp_1 
@@
 Here we describe the posteriors for both models
+
 ```{r simple_table, cache =TRUE}
-bayestestR::describe_posterior(mod.1_bDAG, test = c("p_direction"),diagnostic = "ESS")
+bayestestR::describe_posterior(mod.1_bDAGR, test = c("p_direction"),diagnostic = "ESS")
 ```
@@
 ```{r simple_table_ls, cache =TRUE}
-bayestestR::describe_posterior(mod.2_bDAG, test = c("p_direction"), diagnostic = "ESS")
+bayestestR::describe_posterior(mod.2_bDAGR, test = c("p_direction"), diagnostic = "ESS")
 ```
@@
 ```{r Bayesian_model_table, cache=TRUE, include =FALSE}
-table_model_latex_dual(mod.1_bDAG,mod.2_bDAG)
+table_model_latex_dual(mod.1_bDAGR,mod.2_bDAGR)
 ```
@@
 ```{r cache =TRUE}
-g1dag1<-plot(parameters::model_parameters(mod.1_bDAG))+ ggtitle("Personal Well-Being Coefficients")
-g2dag1<-plot(parameters::model_parameters(mod.2_bDAG))+ ggtitle("Life Satisfaction Coefficients")
+g1dag1<-plot(parameters::model_parameters(mod.1_bDAGR))+ ggtitle("Personal Well-Being Coefficients")
+g2dag1<-plot(parameters::model_parameters(mod.2_bDAGR))+ ggtitle("Life Satisfaction Coefficients")
 ```
@@
 
-```{r plot_coefficientplots_BayesDAG1, cache = TRUE}
+```{r coef, cache = TRUE}
 g1dag1/g2dag1
 ```
-we used `r cite_r("b-references.bib")`
+
+Next we plot the graphs. First we extract the predictions
+
+```{r Bayesian_pwi_predictedSAT_DAG_GET, cache=TRUE}
+get_predictions(mod.1_bDAGR)
+```
+
+```{r Bayesian_lifesat_predictedSAT_DAG_GET, cache=TRUE}
+get_predictions(mod.2_bDAGR)
+```
+
+<!-- ```{r DAGS, cache =TRUE} old graphs -->
+<!-- g1dag1<-plot(parameters::model_parameters(mod.1_bDAG))+ ggtitle("Personal well-being Coefficients") -->
+<!-- g2dag2<-plot(parameters::model_parameters(mod.2_bDAG))+ ggtitle("Life Satisfaction Coefficients") -->
+<!-- ``` -->
+
+# Personal wellbeing graph
+
+```{r}
+p0 <- brms::conditional_effects(
+  mod.1_bDAGR,
+  c("Age_within:Beliefs" ),
+  spaghetti = TRUE,
+  categorical = FALSE,
+  nsamples = 200,
+  prob = .95,
+  method = 'fitted'
+)
+
+plot(p0, plot = FALSE)[[1]] +   facet_wrap("Beliefs",nrow = 1) + xlab("Years of Aging (centered)") + ylab ("Personal Wellbeing 0-10")  + theme_blank() + labs(title = "Predicted effects of beliefs states for personal wellbeing")+ theme(legend.position="top") + scale_color_viridis_d(alpha = .2)
+# pwi
+```
+
+# Life Satisfaction graph 
+```{r}
+p2 <- brms::conditional_effects(
+  mod.2_bDAGR,
+  c("Age_within:Beliefs" ),
+  spaghetti = TRUE,
+  categorical = FALSE,
+  nsamples = 200,
+  prob = .95,
+  method = 'fitted'
+)
+
+plot(p2, plot = FALSE)[[1]] +   facet_wrap("Beliefs",nrow = 1) + xlab("Years of Aging (centered)") + ylab ("Life Satisfaction 1-7")  + theme_blank() + labs(title = "Predicted effects of beliefs states for life satisfaction") + theme(legend.position="top") + scale_color_viridis_d(alpha = .2)
+# graph called LS
+```
+
+<!-- ```{r Bayesian_pwi_predictedSAT_DAG, cache=TRUE} -->
+<!-- graph_predictions(mod.1_bDAGR, "Predicted Values of Personal Well-Being: Bayesian Model (Confounder Adjusted)") -->
+<!-- ``` -->
+
+<!-- ```{r Bayesian_lifesat_predictedSAT_DAG, cache=TRUE} -->
+<!-- graph_predictions(mod.2_bDAGR, "Predicted Values of Life Satisfaction: Bayesian Model (Confounder Adjusted)") -->
+<!-- ``` -->
+
+Check mixing
+
+```{r mixing_pwi_DAG, cache=TRUE}
+plot(mod.1_bDAGR, pars ="b_") 
+```
+
+```{r mixing_lsDAG, cache=TRUE}
+plot(mod.2_bDAGR, pars ="b_") 
+```
+
+```{r mixing_sd_pwi, cache=TRUE}
+plot(mod.1_bDAGR, pars ="sd_") #
+```
+
+```{r mixing_ls_pwi, cache=TRUE}
+plot(mod.2_bDAGR, pars ="sd_") # 
+```
+
+The packages we used are: `r cite_r("b-references.bib")`
+
+<!-- ### EXTRA DETAIL  PREVIOUS MODELS -->
+
+<!-- ## Bayesian model, no confounders addressed -->
+
+<!-- In the simple model, the in -->
+
+<!-- ```{r setmodelformula} -->
+
+<!-- forumula_bayesPWI <- bf(PWI ~ yearsC * Beliefs + (1 + yearsC|Id)) -->
+
+<!-- forumula_bayesLIFESAT <- bf(LIFESAT ~  yearsC * Beliefs + (1 + yearsC|Id)) -->
+
+<!-- ``` -->
+
+<!-- ```{r pwi_bayesmodel_DAG} -->
+
+<!-- # PWI -->
+
+<!-- mod.1_b <-brm(forumula_bayesPWI, data= d_3, file ="mod.1_b.rds", cores = 8) # note these are very large files -->
+
+<!-- ``` -->
+
+<!-- ```{r lsbayesmodel_DAG} -->
+
+<!-- # LIFE SAT -->
+
+<!-- mod.2_b <-brm (forumula_bayesLIFESAT, data= d_3,cores = 8, file ="mod.2_b.rds") -->
+
+<!-- # note these are very large files -->
+
+<!-- ``` -->
+
+<!-- ```{r simple_table_DAG, cache =TRUE} -->
+
+<!-- bayestestR::describe_posterior(mod.1_b, ci = 0.90,  test = c("p_direction"),diagnostic = "ESS") -->
+
+<!-- ``` -->
+
+<!-- ```{r simple_table_ls_DAG, cache =TRUE} -->
+
+<!-- bayestestR::describe_posterior(mod.2_b, ci = 0.90, test = c("p_direction"), diagnostic = "ESS") -->
+
+<!-- ``` -->
+
+<!-- <!-- Tables: -->
+
+<!-- <!-- # ```{r pwi_Bayesian_model_table_pwi, cache =TRUE} -->
+
+<!-- <!-- # tab_model(mod.1_b) -->
+
+<!-- <!-- # ``` -->
+
+<!-- <!-- # -->
+
+<!-- <!-- # ```{r lifesat_Bayesian_model_table_ls lifesat, cache =TRUE} -->
+
+<!-- <!-- # tab_model(mod.2_b) -->
+
+<!-- <!-- # ``` -->
+
+<!-- <!-- ```{r texreg_bayes_table:pwi,cache=TRUE, include=FALSE} -->
+
+<!-- <!-- #Results for publication. -->
+
+<!-- <!-- table_model_latex_pwi(mod.1_b) -->
+
+<!-- <!-- ``` -->
+
+<!-- <!-- ```{r texreg_bayes_table:ls, cache=TRUE, include=FALSE} -->
+
+<!-- <!-- table_model_latex_ls(mod.2_b) -->
+
+<!-- <!-- ``` -->
+
+<!-- <!-- Coefficient plots used in manuscript-->
+
+<!-- <!-- ```{r coefficientplotsbayes, cache=TRUE} -->
+
+<!-- <!-- plot_coefs(mod.1_b,mod.2_b) -->
+
+<!-- <!-- ``` -->
+
+<!-- ```{r stancode for models, include=FALSE} -->
+
+<!-- #model equation -->
+
+<!-- #Here we extract our priors for reporting -->
+
+<!-- stancode(mod.1_b) -->
+
+<!-- stancode(mod.2_b) -->
+
+<!-- ``` -->
+
+<!-- ```{r prior summaries, include =FALSE} -->
+
+<!-- #this does the same as the stancode -->
+
+<!-- brms::prior_summary(mod.1_b) -->
+
+<!-- brms::prior_summary(mod.2_b) -->
+
+<!-- ``` -->
+
+<!-- ```{r Bayesian_model_table_SIMPLE, cache=TRUE} -->
+
+<!-- #tab_model(mod.1_b, show.ci= .89) # we really should use something other than 95 but the team consensus is to use 95 -->
+
+<!-- tab_model(mod.1_b, mod.2_b, show.ci= .95) # no difference to inference -->
+
+<!-- ``` -->
+
+<!-- We inspected the posterior distributions, which look good. -->
+
+<!-- The posterior prediction plots look reasonable.   -->
+
+<!-- ```{r posteriorpredictivechecks:pwi, cache=TRUE} -->
+
+<!-- # cannot run on my laptop -->
+
+<!-- pp_1<-rstanarm::pp_check(mod.1_b) + ylab("Personal well-being") -->
+
+<!-- pp_2<-rstanarm::pp_check(mod.2_b) + ylab("Life Satisfaction") ## life sat suggests an ordinal model -->
+
+<!-- pp_1 + pp_2 -->
+
+<!-- ``` -->
+
+<!-- ### Predictions graphs -->
+
+<!-- Personal Well being predictions. -->
 
 <!-- ```{r} -->
-<!-- cite_r("b-references.bib") -->
-<!-- ``` -->
-
-Next we plot the graphs. First we extract the predictions
-
-```{r Bayesian_pwi_predictedSAT_DAG_GET, cache=TRUE}
-get_predictions(mod.1_bDAG)
-```
-
-```{r Bayesian_lifesat_predictedSAT_DAG_GET, cache=TRUE}
-get_predictions(mod.2_bDAG)
-```
-```{r DAGS, cache =TRUE}
-g1dag1<-plot(parameters::model_parameters(mod.1_bDAG))+ ggtitle("Personal well-being Coefficients")
-g2dag2<-plot(parameters::model_parameters(mod.2_bDAG))+ ggtitle("Life Satisfaction Coefficients")
-```
-
-And here's the graph.
-
-```{r plot_coefficientplots_LMER_FULL_DAG, cache = TRUE}
-g1dag1/g2dag1
-```
-
-Here are the coefficient plots
-
-```{r Bayesian_pwi_predictedSAT_DAG, cache=TRUE}
-graph_predictions(mod.1_bDAG, "Predicted Values of Personal Well-Being: Bayesian Model (Confounder Adjusted)")
-```
-
-```{r Bayesian_lifesat_predictedSAT_DAG, cache=TRUE}
-graph_predictions(mod.2_bDAG, "Predicted Values of Life Satisfaction: Bayesian Model (Confounder Adjusted)")
-```
-
-```{r mixing_pwi_DAG, cache=TRUE}
-plot(mod.1_bDAG, pars ="b_") 
-```
-
-```{r mixing_lsDAG, cache=TRUE}
-plot(mod.2_bDAG, pars ="b_") 
-```
-
-As do the random components
-
-```{r mixing_sd_pwi, cache=TRUE}
-plot(mod.1_bDAG, pars ="sd_") #
-```
-
-```{r mixing_ls_pwi, cache=TRUE}
-plot(mod.2_bDAG, pars ="sd_") # 
-```
-
-## Bayesian model, no confounders addressed
-
-In the simple model, the in
-
-```{r setmodelformula}
-forumula_bayesPWI <- bf(PWI ~ yearsC * Beliefs + (1 + yearsC|Id))
-forumula_bayesLIFESAT <- bf(LIFESAT ~  yearsC * Beliefs + (1 + yearsC|Id))
-```
-
-```{r pwi_bayesmodel_DAG}
-# PWI
-mod.1_b <-brm(forumula_bayesPWI, data= d_3, file ="mod.1_b.rds", cores = 8) # note these are very large files
-```
-
-```{r lsbayesmodel_DAG}
-# LIFE SAT
-mod.2_b <-brm (forumula_bayesLIFESAT, data= d_3,cores = 8, file ="mod.2_b.rds")
-# note these are very large files
-```
-
-```{r simple_table_DAG, cache =TRUE}
-bayestestR::describe_posterior(mod.1_b, ci = 0.90,  test = c("p_direction"),diagnostic = "ESS")
-```
-
-```{r simple_table_ls_DAG, cache =TRUE}
-bayestestR::describe_posterior(mod.2_b, ci = 0.90, test = c("p_direction"), diagnostic = "ESS")
-```
-
-<!-- Tables: -->
-<!-- # ```{r pwi_Bayesian_model_table_pwi, cache =TRUE} -->
-<!-- # tab_model(mod.1_b) -->
-<!-- # ``` -->
-<!-- # -->
-<!-- # ```{r lifesat_Bayesian_model_table_ls lifesat, cache =TRUE} -->
-<!-- # tab_model(mod.2_b) -->
-<!-- # ``` -->
-
-<!-- ```{r texreg_bayes_table:pwi,cache=TRUE, include=FALSE} -->
-<!-- #Results for publication. -->
-<!-- table_model_latex_pwi(mod.1_b) -->
-<!-- ``` -->
-
-<!-- ```{r texreg_bayes_table:ls, cache=TRUE, include=FALSE} -->
-<!-- table_model_latex_ls(mod.2_b) -->
-<!-- ``` -->
-
-<!-- Coefficient plots used in manuscript-->
-<!-- ```{r coefficientplotsbayes, cache=TRUE} -->
-<!-- plot_coefs(mod.1_b,mod.2_b) -->
-<!-- ``` -->
-
-```{r stancode for models, include=FALSE}
-#model equation
-#Here we extract our priors for reporting
-
-stancode(mod.1_b)
-stancode(mod.2_b)
-```
-
-```{r prior summaries, include =FALSE}
-#this does the same as the stancode
-brms::prior_summary(mod.1_b)
-brms::prior_summary(mod.2_b)
-```
-
-```{r Bayesian_model_table_SIMPLE, cache=TRUE}
-#tab_model(mod.1_b, show.ci= .89) # we really should use something other than 95 but the team consensus is to use 95
-tab_model(mod.1_b, mod.2_b, show.ci= .95) # no difference to inference
-```
-
-We inspected the posterior distributions, which look good.
-
-The posterior prediction plots look reasonable.
-
-```{r posteriorpredictivechecks:pwi, cache=TRUE}
-# cannot run on my laptop
-pp_1<-rstanarm::pp_check(mod.1_b) + ylab("Personal well-being")
-pp_2<-rstanarm::pp_check(mod.2_b) + ylab("Life Satisfaction") ## life sat suggests an ordinal model
-pp_1 + pp_2
-```
-
-### Predictions graphs
-
-Personal Well being predictions.
-
-```{r}
-preds_1b<-get_predictions(mod.1_b)
-preds_1b
-```
-
-Life Satisfaction predictions.
-
-```{r}
-preds_2b<-get_predictions(mod.2_b)
-preds_2b
-```
-
-Next we graph the expected effects.
-
-```{r Bayesian_pwi_predicted, cache=TRUE, dependson= mod.1_b}
-graph_predictions(mod.1_b, "Predicted Values of Personal Well-Being")
-```
-
-```{r Bayesian_lifesat_predicted, cache=TRUE, dependson= mod.2_b}
-graph_predictions(mod.2_b, "Predicted Values of Life Satisfaction")
-```
-
-## Test of the Bayesian saturated model
-
-```{r setmodelformulaSAT_FULL}
-# note that BRMS automatically centers so we do not need to center years
-forumula_bayesPWI_SATUR <- bf(PWI ~  years * Beliefs +  Age_in_Decades_C + Deprivation_S + Education_S  + Employed + Ethnic_Categories + Male  + Has_Partner  + Political_Conservativism_S + Urban + (1+ years|Id))
-
-forumula_bayesLIFESAT_SATUR <- bf(LIFESAT ~  yearsC * Beliefs +  Age_in_Decades_C + Deprivation_S + Education_S  + Employed + Ethnic_Categories + Male  + Has_Partner  + Political_Conservativism_S + Urban + (1+ years|Id))
-```
-
-```{r pwi_bayesmodelSAT_FULL}
-# PWI
-mod.1_bSAT <-brm(forumula_bayesPWI_SATUR, data= d_3, file ="mod.1_bS.rds", cores = 8) # note these are very large files
-summary(mod.1_bSAT)
-```
-
-```{r lsbayesmodelSAT}
-# LIFE SAT
-mod.2_bSAT <-brm (forumula_bayesLIFESAT_SATUR, data= d_3,cores = 8, file ="mod.2_bS.rds")
-summary(mod.2_bSAT)
-###
-tab_model(mod.1_bSAT,mod.2_bSAT)
-# note these are very large files
-```
-
-```{r Bayesian_pwi_predictedSAT, cache=TRUE}
-graph_predictions(mod.1_bSAT, "Predicted Values of Personal Well-Being")
-```
-```{r Bayesian_lifesat_predictedSAT, cache=TRUE}
-graph_predictions(mod.2_b, "Predicted Values of Life Satisfaction")
-```
-
-## REML estimation
-
-Many researchers have yet to familiarise themselves with Bayesian estimation.
+
+<!-- preds_1b<-get_predictions(mod.1_b) -->
+
+<!-- preds_1b -->
+
+<!-- ``` -->
+
+<!-- Life Satisfaction predictions. -->
+
+<!-- ```{r} -->
+
+<!-- preds_2b<-get_predictions(mod.2_b) -->
+
+<!-- preds_2b -->
+
+<!-- ``` -->
+
+<!-- Next we graph the expected effects.  -->
+
+<!-- ```{r Bayesian_pwi_predicted, cache=TRUE, dependson= mod.1_b} -->
+
+<!-- graph_predictions(mod.1_b, "Predicted Values of Personal Well-Being") -->
+
+<!-- ``` -->
+
+<!-- ```{r Bayesian_lifesat_predicted, cache=TRUE, dependson= mod.2_b} -->
+
+<!-- graph_predictions(mod.2_b, "Predicted Values of Life Satisfaction") -->
+
+<!-- ``` -->
+
+<!-- ## Test of the Bayesian saturated model:  do not do this -->
+
+<!-- ```{r setmodelformulaSAT_FULL} -->
+
+<!-- # note that BRMS automatically centers so we do not need to center years -->
+
+<!-- forumula_bayesPWI_SATUR <- bf(PWI ~  years * Beliefs +  Age_in_Decades_C + Deprivation_S + Education_S  + Employed + Ethnic_Categories + Male  + Has_Partner  + Political_Conservativism_S + Urban + (1+ years|Id)) -->
+
+<!-- forumula_bayesLIFESAT_SATUR <- bf(LIFESAT ~  yearsC * Beliefs +  Age_in_Decades_C + Deprivation_S + Education_S  + Employed + Ethnic_Categories + Male  + Has_Partner  + Political_Conservativism_S + Urban + (1+ years|Id)) -->
+
+```{=html}
+<!-- ``` --
+<!-- ```{r pwi_bayesmodelSAT_FULL} -->
+```
+<!-- # PWI -->
+
+<!-- mod.1_bSAT <-brm(forumula_bayesPWI_SATUR, data= d_3, file ="mod.1_bS.rds", cores = 8) # note these are very large files -->
+
+<!-- summary(mod.1_bSAT) -->
+
+<!-- ``` -->
+
+<!-- ```{r lsbayesmodelSAT} -->
+
+<!-- # LIFE SAT -->
+
+<!-- mod.2_bSAT <-brm (forumula_bayesLIFESAT_SATUR, data= d_3,cores = 8, file ="mod.2_bS.rds") -->
+
+<!-- summary(mod.2_bSAT) -->
+
+<!-- ### -->
+
+<!-- tab_model(mod.1_bSAT,mod.2_bSAT) -->
+
+<!-- # note these are very large files -->
+
+<!-- ``` -->
+
+<!-- ```{r Bayesian_pwi_predictedSAT, cache=TRUE} -->
+
+<!-- graph_predictions(mod.1_bSAT, "Predicted Values of Personal Well-Being") -->
+
+<!-- ``` -->
+
+<!-- ```{r Bayesian_lifesat_predictedSAT, cache=TRUE} -->
+
+<!-- graph_predictions(mod.2_b, "Predicted Values of Life Satisfaction") -->
+
+<!-- ``` -->
+
+<!-- ## REML estimation -->
+
+<!-- Many researchers have yet to familiarise themselves with Bayesian estimation.
 We therefore report results from models using a Maximum Likelihood estimator.
@@
 ML estimatation is also faster than our preferred Bayesian estimator.
-
-Results are virtually identical to the Bayesian model.
-
-## Adjusting for confounds
-
-### First confounder model
-
-```{r lmer_model_with_full_covariatesDAG1, cache =TRUE}
-library(lme4)
-pwi_formulaDag1 <- 'PWI ~  yearsC * Beliefs +  Age_in_Decades_C  + Ethnic_Categories + Male + (1 + yearsC|Id)'
-ls_formulaDag1 <- 'LIFESAT ~ yearsC * Beliefs +  Age_in_Decades_C  + Ethnic_Categories + Male  + (1 + yearsC|Id)'
-
-mod.1dag1 <-lmer (pwi_formulaDag1, data= d_3)
-mod.2dag1 <-lmer (ls_formulaDag1, data= d_3)
-```
-
-Here is the table.
-```{r tab_dag1, cache = TRUE}
-sjPlot::tab_model(mod.1dag1,mod.2dag1)
-```
-
-We compared complex with simple models, and we found the inference does not change. 
-```{r}
-# identical result
-sjPlot::tab_model(mod.1dag1,mod.2dag1)
-```
-
-```{r grapwi_predict_DAG, cache =TRUE}
-graph_predictions(mod.1dag1, "Predicted Values of Life Satisfaction: Dag 1")
-```
-```{r gralifesat_predict_DAG, cache =TRUE}
-graph_predictions(mod.2dag1, "Predicted Values of Life Satisfaction: Dag 1")
-```
-
-## Second confounder model
-
-```{r lmer_model_with_full_covariates_DAG, cache =TRUE}
-library(lme4)
-pwi_formulaDag2 <- 'PWI ~  yearsC * Beliefs +  Age_in_Decades_C  + Employed + Male + Deprivation_S + Political_Conservativism_S + (1 + yearsC|Id)'
-ls_formulaDag2 <- 'LIFESAT ~ yearsC * Beliefs + Employed +  Age_in_Decades_C  + Employed + Male + Deprivation_S + Political_Conservativism_S + (1 + yearsC|Id)'
-
-mod.1dag2 <-lmer (pwi_formulaDag2, data= d_3)
-mod.2dag2 <-lmer (ls_formulaDag2, data= d_3)
-```
-
-Here is the table
-
-```{r tab_dag2, cache = TRUE}
-sjPlot::tab_model(mod.1dag2,mod.1dag2)
-```
-
-```{r dag2_plotcreate, cache =TRUE}
-g1dag2<-plot(parameters::model_parameters(mod.1dag2))+ ggtitle("Personal well-being Coefficients")
-g2dag2<-plot(parameters::model_parameters(mod.2dag2))+ ggtitle("Life Satisfaction Coefficients")
-```
-
-And here's the graph.
-
-```{r plot_coefficientplots_LMER_FULL_DAG2, cache = TRUE}
-g1dag2/g2dag2 
-```
-
-```{r pwi_predict_DAG, cache =TRUE}
-graph_predictions(mod.1dag2, "Predicted Values of Life Satisfaction: Dag 2")
-```
-
-```{r glifesat_predict_DAG, cache =TRUE}
-graph_predictions(mod.2dag2, "Predicted Values of Life Satisfaction: Dag 2")
-```
-
-Note we would prefer the the first counfounder model because there is less measurement error (not employed could mean mean not retired).
-
-Additionally the first counfounder model has fewer terms to estimate.
-
-<!-- Her are the coefficient plots for the full models -->
-<!-- ```{r coefficientplots_LMER_FULL, cache=TRUE} -->
-<!-- plot_models(mod.2, mod.1) + ylim(-.4, .75) + theme_plain()  # Plots -->
-<!-- ``` -->
-
-Next, recall we want to check whether inference is robustness to the researcher degree of freedom in selecting how many number of waves a participant must have responded to for inclusion.
-
-```{r run all the models with different ns, cache = TRUE}
-mod.1_3 <-lmer (pwi_formulaDag1, data= d_4)
-mod.2_3 <-lmer (mod.2dag2, data= d_4)
-
-mod.1_4 <-lmer (pwi_formulaDag1, data= d_4)
-mod.2_4 <-lmer (mod.2dag2, data= d_4)
-
-mod.1_5 <-lmer (pwi_formulaDag1, data= d_5)
-mod.2_5 <-lmer (mod.2dag2, data= d_5)
-
-mod.1_6 <-lmer (pwi_formulaDag1, data= d_6)
-mod.2_6 <-lmer (mod.2dag2, data= d_6)
-
-mod.1_7 <-lmer (pwi_formulaDag1, data= d_7)
-mod.2_7 <-lmer (mod.2dag2, data= d_7)
-
-mod.1_8 <-lmer (pwi_formulaDag1, data= d_8)
-mod.2_8 <-lmer (mod.2dag2, data= d_8)
-```
-
-The inference is robut even to the poorly specified models using REML estimation.
-
-```{r extremes_of_researcher_degree_of_freedom_test, cache =TRUE}
-library(parameters)
-g1a_3<-plot(parameters::model_parameters(mod.1_3))+ ggtitle("Personal well-being Coefficients: 8 Waves")
-g2a_3<-plot(parameters::model_parameters(mod.2_3))+ ggtitle("Life Satisfaction Coefficients: 8 Waves")
-
-g1a_8<-plot(parameters::model_parameters(mod.1_8))+ ggtitle("Personal well-being Coefficients: 8 Waves")
-g2a_8<-plot(parameters::model_parameters(mod.2_8))+ ggtitle("Life Satisfaction Coefficients: 8 Waves")
-```
-
-```{r USE_2_coefficientplots_LMER_FULL_pwi,cache =TRUE}
-library(patchwork)
-g1a_3/g1a_8 +   plot_annotation(title = "Compare 3 X response model (top) with 8 X response model (bottom): Personal Well-Being")
-```
-
-```{r USE_2_coefficientplots_LMER_FULL_ls,cache =TRUE}
-library(patchwork)
-g2a_3/g2a_8 +   plot_annotation(title = "Compare 3 X response model (top) with 8 X response model (bottom) : Life Satisfaction")
-```
-
-# Missing Data Analysis
-
-Finally, we ran our analyses on 10 multiply imputed datasets.
+-->
+
+<!-- Results are virtually identical to the Bayesian model. -->
+
+<!-- ## Adjusting for confounds -->
+
+<!-- ### First confounder model  -->
+
+<!-- ```{r lmer_model_with_full_covariatesDAG1, cache =TRUE} -->
+
+<!-- library(lme4) -->
+
+<!-- pwi_formulaDag1 <- 'PWI ~  Beliefs *  Age_within +  Age_between  + Ethnic_Categories + Male + (1 + Age_within|Id)' -->
+
+<!-- ls_formulaDag1 <- 'LIFESAT ~ Beliefs *  Age_within +  Age_between  + Ethnic_Categories + Male + (1 + Age_within|Id' -->
+
+<!-- mod.1dag1 <-lmer (pwi_formulaDag1, data= d_3) -->
+
+<!-- mod.2dag1 <-lmer (ls_formulaDag1, data= d_3) -->
+
+<!-- ``` -->
+
+<!-- Here is the table. -->
+
+<!-- ```{r tab_dag1, cache = TRUE} -->
+
+<!-- sjPlot::tab_model(mod.1dag1,mod.2dag1) -->
+
+<!-- ``` -->
+
+<!-- We compared complex with simple models, and we found the inference does not change.  -->
+
+<!-- ```{r} -->
+
+<!-- # identical result -->
+
+<!-- sjPlot::tab_model(mod.1dag1,mod.2dag1) -->
+
+<!-- ``` -->
+
+<!-- ```{r grapwi_predict_DAG, cache =TRUE} -->
+
+<!-- graph_predictions(mod.1dag1, "Predicted Values of Life Satisfaction: Dag 1") -->
+
+<!-- ``` -->
+
+<!-- ```{r gralifesat_predict_DAG, cache =TRUE} -->
+
+<!-- graph_predictions(mod.2dag1, "Predicted Values of Life Satisfaction: Dag 1") -->
+
+<!-- ``` -->
+
+<!-- ## Second confounder model  -->
+
+<!-- ```{r lmer_model_with_full_covariates_DAG, cache =TRUE} -->
+
+<!-- library(lme4) -->
+
+<!-- pwi_formulaDag2 <- 'PWI ~  yearsC * Beliefs +  Age_in_Decades_C  + Employed + Male + Deprivation_S + Political_Conservativism_S + (1 + yearsC|Id)' -->
+
+<!-- ls_formulaDag2 <- 'LIFESAT ~ yearsC * Beliefs + Employed +  Age_in_Decades_C  + Employed + Male + Deprivation_S + Political_Conservativism_S + (1 + yearsC|Id)' -->
+
+<!-- mod.1dag2 <-lmer (pwi_formulaDag2, data= d_3) -->
+
+<!-- mod.2dag2 <-lmer (ls_formulaDag2, data= d_3) -->
+
+<!-- dat <- cbind( -->
+
+<!--   d_3, -->
+
+<!--   demean(d_3, select = c("Age", "Beliefs"), group = "Id") -->
+
+<!-- ) -->
+
+<!-- dat<-dat%>% -->
+
+<!--   dplyr::mutate(Age_between_C = scale(Age_between, center =TRUE, scale = FALSE)) -->
+
+<!-- str(dat) -->
+
+<!-- head(dat$Age_between_C) -->
+
+<!-- # After thought -- check within and between group aging... no differences -->
+
+<!-- # dmPWI <- 'PWI ~  Age_within * Beliefs + Age_between_C  +  Employed + Male + Deprivation_S + Political_Conservativism_S + (1 + Age_within|Id)' -->
+
+<!-- # mod_dmPWI <-lmer (dmPWI, data= dat) -->
+
+<!-- # tab_model(mod_dmPWI) -->
+
+<!-- # summary(mod_dmPWI) -->
+
+<!-- # n1<-ggeffects::ggpredict(model = mod_dmPWI, terms = c("Age_within [minmax]","Beliefs"),  -->
+
+<!-- #                               ci.lvl = 0.95, -->
+
+<!-- #                               type = "fe", -->
+
+<!-- #                               typical = "mean", -->
+
+<!-- #                               back.transform = TRUE, -->
+
+<!-- #                               ppd = FALSE, -->
+
+<!-- #                               interval = "confidence") -->
+
+<!-- #   plot(n1, facets = T) +  gghighlight::gghighlight()  +  theme_blank() #+ ggtitle(y) # title to be suppled -->
+
+<!-- #    -->
+
+<!-- # graph_predictions(mod_dmPWI, "Predicted Values of PWI") -->
+
+<!-- # #  -->
+
+<!-- # lsPWI <- 'LIFESAT ~  Age_within * Beliefs + Age_between_C  +  Employed + Male + Deprivation_S + Political_Conservativism_S + (1 + Age_within|Id)' -->
+
+<!-- # mod_lsPWI <-lmer (lsPWI, data= dat) -->
+
+<!-- # tab_model(mod_lsPWI) -->
+
+<!-- # summary(mod_lsPWI) -->
+
+<!-- # head(dat$Age_within) -->
+
+<!-- # n2<-ggeffects::ggpredict(model = mod_lsPWI, terms = c("Age_within [minmax]","Beliefs"), -->
+
+<!-- #                               ci.lvl = 0.95, -->
+
+<!-- #                               type = "fe", -->
+
+<!-- #                               typical = "mean", -->
+
+<!-- #                               back.transform = TRUE, -->
+
+<!-- #                               ppd = FALSE, -->
+
+<!-- #                               interval = "confidence") -->
+
+<!-- # plot(n2, facets = T) +  gghighlight::gghighlight()  +  theme_blank() #+ ggtitle(y) # title to be suppled -->
+
+<!-- ``` -->
+
+<!-- Here is the table -->
+
+<!-- ```{r tab_dag2, cache = TRUE} -->
+
+<!-- sjPlot::tab_model(mod.1dag2,mod.1dag2) -->
+
+<!-- ``` -->
+
+<!-- ```{r dag2_plotcreate, cache =TRUE} -->
+
+<!-- g1dag2<-plot(parameters::model_parameters(mod.1dag2))+ ggtitle("Personal well-being Coefficients") -->
+
+<!-- g2dag2<-plot(parameters::model_parameters(mod.2dag2))+ ggtitle("Life Satisfaction Coefficients") -->
+
+<!-- ``` -->
+
+<!-- And here's the graph. -->
+
+<!-- ```{r plot_coefficientplots_LMER_FULL_DAG2, cache = TRUE} -->
+
+<!-- g1dag2/g2dag2  -->
+
+<!-- ``` -->
+
+<!-- ```{r pwi_predict_DAG, cache =TRUE} -->
+
+<!-- graph_predictions(mod.1dag2, "Predicted Values of Life Satisfaction: Dag 2") -->
+
+<!-- ``` -->
+
+<!-- ```{r glifesat_predict_DAG, cache =TRUE} -->
+
+<!-- graph_predictions(mod.2dag2, "Predicted Values of Life Satisfaction: Dag 2") -->
+
+<!-- ``` -->
+
+<!-- Note we would prefer the the first counfounder model because there is less measurement error (not employed could mean mean not retired).  -->
+
+<!-- Additionally the first counfounder model has fewer terms to estimate.  -->
+
+<!-- <!-- Her are the coefficient plots for the full models -->
+
+--\> <!-- <!-- ```{r coefficientplots_LMER_FULL, cache=TRUE} --> --\>
+<!-- <!-- plot_models(mod.2, mod.1) + ylim(-.4, .75) + theme_plain()  # Plots -->
+--\> <!-- <!-- ``` --> --\>
+
+<!-- Next, recall we want to check whether inference is robustness to the researcher degree of freedom in selecting how many number of waves a participant must have responded to for inclusion. -->
+
+<!-- ```{r run all the models with different ns, cache = TRUE} -->
+
+<!-- mod.1_3 <-lmer (pwi_formulaDag1, data= d_4) -->
+
+<!-- mod.2_3 <-lmer (mod.2dag2, data= d_4) -->
+
+<!-- mod.1_4 <-lmer (pwi_formulaDag1, data= d_4) -->
+
+<!-- mod.2_4 <-lmer (mod.2dag2, data= d_4) -->
+
+<!-- mod.1_5 <-lmer (pwi_formulaDag1, data= d_5) -->
+
+<!-- mod.2_5 <-lmer (mod.2dag2, data= d_5) -->
+
+<!-- mod.1_6 <-lmer (pwi_formulaDag1, data= d_6) -->
+
+<!-- mod.2_6 <-lmer (mod.2dag2, data= d_6) -->
+
+<!-- mod.1_7 <-lmer (pwi_formulaDag1, data= d_7) -->
+
+<!-- mod.2_7 <-lmer (mod.2dag2, data= d_7) -->
+
+<!-- mod.1_8 <-lmer (pwi_formulaDag1, data= d_8) -->
+
+<!-- mod.2_8 <-lmer (mod.2dag2, data= d_8) -->
+
+<!-- ``` -->
+
+<!-- The inference is robut even to the poorly specified models using REML estimation. -->
+
+<!-- ```{r extremes_of_researcher_degree_of_freedom_test, cache =TRUE} -->
+
+<!-- library(parameters) -->
+
+<!-- g1a_3<-plot(parameters::model_parameters(mod.1_3))+ ggtitle("Personal well-being Coefficients: 8 Waves") -->
+
+<!-- g2a_3<-plot(parameters::model_parameters(mod.2_3))+ ggtitle("Life Satisfaction Coefficients: 8 Waves") -->
+
+<!-- g1a_8<-plot(parameters::model_parameters(mod.1_8))+ ggtitle("Personal well-being Coefficients: 8 Waves") -->
+
+<!-- g2a_8<-plot(parameters::model_parameters(mod.2_8))+ ggtitle("Life Satisfaction Coefficients: 8 Waves") -->
+
+<!-- ``` -->
+
+<!-- ```{r USE_2_coefficientplots_LMER_FULL_pwi,cache =TRUE} -->
+
+<!-- library(patchwork) -->
+
+<!-- g1a_3/g1a_8 +   plot_annotation(title = "Compare 3 X response model (top) with 8 X response model (bottom): Personal Well-Being") -->
+
+<!-- ``` -->
+
+<!-- ```{r USE_2_coefficientplots_LMER_FULL_ls,cache =TRUE} -->
+
+<!-- library(patchwork) -->
+
+<!-- g2a_3/g2a_8 +   plot_annotation(title = "Compare 3 X response model (top) with 8 X response model (bottom) : Life Satisfaction") -->
+
+<!-- ``` -->
+
+<!-- # Missing Data Analysis -->
+
+<!-- Finally, we ran our analyses on 10 multiply imputed datasets.
 Missing responses can be an important source of bias in a longitudinal study.
 Again we wrote some functions to deal with all the work, you can find these in `funs.R` file.
-
-```{r echo = TRUE,  cache = TRUE}
-# Imputed data
-imputed3<-amelia_imputation_clean(d_3)
-# saveRDS(imputed3,"imputed3")
-# imputed3<-readRDS(imputed3)
-```
-
-We reran the imputed data analysis. (Note we ajust for confounds)
-
-```{r cache = TRUE, imputedsimplePWI, cache=TRUE}
-# run models iterating over imputed data
-mod_imputed_pwi<-loop_lmer_model(pwi_formulaDag1, imputed3)
-```
-
-```{r imputed pwi table, cache=TRUE}
-tab_mods_pwi<-loop_lmer_model_tab(mod_imputed_pwi)
-#create_hux_table(tab_mods_pwi)
-```
-
-Next a graph of the coefficients.
-
-```{r coef_PWI_imputed_coefplot, cache =TRUE}
-# coeff plot
-imp_pwi <- plot(tab_mods_pwi) + ggtitle("Personal well-being Coefficients (Imputed)") # pick your  own title
-```
-
-The inference is robust in the imputation model.
-
-The following code returns the analysis for the life satisfaction model (again the functions are in the `funs.R` file).
-
-```{r cache = TRUE}
-# run models iterating over imputed data
-mod_imputed_ls<-loop_lmer_model(ls_formulaDag1 ,imputed3)
-
-```{r table_ls_imputed,cache=TRUE}
-# table
-tab_mods_ls<-loop_lmer_model_tab(mod_imputed_ls)
-#create_hux_table(tab_mods_ls)
-```
-
-Here's the graph:
-
-```{r coef_imputed_ls, cache = TRUE}
-# coeff plot
-imp_ls<-plot(tab_mods_ls) + ggtitle("Life Satisfaction Coefficients (Imputed)") 
-```
-We found that the model is robust to multiple imputations over the missing data, though with greater uncertainty for the Personal Well-being effect.
-
-```{r coefficient_plot_both_imputed,  cache = TRUE}
- imp_pwi/imp_ls +  plot_annotation(tag_levels = 'i')
-```
-
-Here are the prediction graphs.
-
-```{r  coef_PWI_imputed coefplot, cache=TRUE}
-# function for prediction
-graph_predictions_imputed(mod_imputed_pwi,"Predicted Values of Personal well-being (imputed)")
-```
-
-```{r pred_LS_imputed,  cache = TRUE}
-# function for prediction
-graph_predictions_imputed(mod_imputed_ls,"Predicted Values of Life Satisfaction (imputed)")
-```
+-->
+
+<!-- ```{r echo = TRUE,  cache = TRUE} -->
+
+<!-- # Imputed data -->
+
+<!-- imputed3<-amelia_imputation_clean(d_3) -->
+
+<!-- # saveRDS(imputed3,"imputed3") -->
+
+<!-- # imputed3<-readRDS(imputed3) -->
+
+<!-- ``` -->
+
+<!-- We reran the imputed data analysis. (Note we ajust for confounds) -->
+
+<!-- ```{r cache = TRUE, imputedsimplePWI, cache=TRUE} -->
+
+<!-- # run models iterating over imputed data -->
+
+<!-- mod_imputed_pwi<-loop_lmer_model(pwi_formulaDag1, imputed3) -->
+
+<!-- ``` -->
+
+<!-- ```{r imputed pwi table, cache=TRUE} -->
+
+<!-- tab_mods_pwi<-loop_lmer_model_tab(mod_imputed_pwi) -->
+
+<!-- #create_hux_table(tab_mods_pwi) -->
+
+<!-- ``` -->
+
+<!-- Next a graph of the coefficients. -->
+
+<!-- ```{r coef_PWI_imputed_coefplot, cache =TRUE} -->
+
+<!-- # coeff plot -->
+
+<!-- imp_pwi <- plot(tab_mods_pwi) + ggtitle("Personal well-being Coefficients (Imputed)") # pick your  own title -->
+
+<!-- ``` -->
+
+<!-- The inference is robust in the imputation model. -->
+
+<!-- The following code returns the analysis for the life satisfaction model (again the functions are in the `funs.R` file). -->
+
+<!-- ```{r cache = TRUE} -->
+
+<!-- # run models iterating over imputed data -->
+
+<!-- mod_imputed_ls<-loop_lmer_model(ls_formulaDag1 ,imputed3) -->
+
+<!-- ```{r table_ls_imputed,cache=TRUE} -->
+
+<!-- # table -->
+
+<!-- tab_mods_ls<-loop_lmer_model_tab(mod_imputed_ls) -->
+
+<!-- #create_hux_table(tab_mods_ls) -->
+
+<!-- ``` -->
+
+<!-- Here's the graph: -->
+
+<!-- ```{r coef_imputed_ls, cache = TRUE} -->
+
+<!-- # coeff plot -->
+
+<!-- imp_ls<-plot(tab_mods_ls) + ggtitle("Life Satisfaction Coefficients (Imputed)")  -->
+
+<!-- ``` -->
+
+<!-- We found that the model is robust to multiple imputations over the missing data, though with greater uncertainty for the Personal Well-being effect.  -->
+
+<!-- ```{r coefficient_plot_both_imputed,  cache = TRUE} -->
+
+<!--  imp_pwi/imp_ls +  plot_annotation(tag_levels = 'i') -->
+
+<!-- ``` -->
+
+<!-- Here are the prediction graphs. -->
+
+<!-- ```{r  coef_PWI_imputed coefplot, cache=TRUE} -->
+
+<!-- # function for prediction -->
+
+<!-- graph_predictions_imputed(mod_imputed_pwi,"Predicted Values of Personal well-being (imputed)") -->
+
+<!-- ``` -->
+
+<!-- ```{r pred_LS_imputed,  cache = TRUE} -->
+
+<!-- # function for prediction -->
+
+<!-- graph_predictions_imputed(mod_imputed_ls,"Predicted Values of Life Satisfaction (imputed)") -->
+
+<!-- ``` -->
 
 <!-- ## Bayesian Multiple imputation models -->
+
 <!-- We wanted to run a Bayesian MI model. However, we did this, we blew up the RAM in JB's machine. So we must satisfy ourselves with ML MI models.  -->
+
 <!-- ```{r} won't work for missing categorical predictors-->
+
 <!-- form_pwi <- bf(PWI | mi() ~ years * mi(Beliefs)) + -->
+
 <!--   bf(Beliefs | mi() ~ years) + set_rescor(FALSE) -->
+
 <!-- imp_bayes_pwi <- brm(form_pwi, data = d_3) -->
-<!-- ``` -->
+
+<!-- ``` -->
+
 <!-- ```{r create data frame of lists for BRMS imputation} -->
+
 <!-- # brms requires a list of the imputations -->
+
 <!-- brmdat <-list(imputed3$imputations$imp1, -->
+
 <!--               imputed3$imputations$imp2, -->
+
 <!--               imputed3$imputations$imp3, -->
+
 <!--               imputed3$imputations$imp4, -->
+
 <!--               imputed3$imputations$imp5, -->
+
 <!--               imputed3$imputations$imp6,   -->
+
 <!--               imputed3$imputations$imp7, -->
+
 <!--               imputed3$imputations$imp8, -->
+
 <!--               imputed3$imputations$imp9, -->
+
 <!--               imputed3$imputations$imp10) -->
-<!-- ``` -->
+
+<!-- ``` -->
+
 <!-- ```{r Bayesian imputed model_pwi} -->
+
 <!-- # not run -->
+
 <!-- #fit_imp1 <- brm_multiple(PWI ~ years * Beliefs + (1 + years|Id), data = brmdat) -->
+
 <!-- #summary(fit_imp1) -->
+
 <!-- #save(fit_imp1) -->
-<!-- ``` -->
+
+<!-- ``` -->
+
 <!-- ```{r Bayesian imputation model_ls} -->
+
 <!-- # not run -->
+
 <!-- #fit_imp2 <- brm_multiple(LIFESAT ~ years * Beliefs + (1+years|Id), data = brmdat, file = "fit_imp2") -->
-<!-- ``` -->
-
-There are further analyses that we ran (e.g. within/between subject effects). These can be found in the GitHub link, posted above. The inferences remain the same.
+
+<!-- ``` -->
+
+<!-- There are further analyses that we ran (e.g. within/between subject effects). These can be found in the GitHub link, posted above. The inferences remain the same. -->
 
@@
 
-The NZAVS is an ongoing longitudinal study.
-Our ethics requires us to protect the privacy of our participants.
-For this reason, we cannot post the raw data, even anonymous data.
-However, we provide data by request to qualified researchers for the purposes of replication.
-We areare interested in collaboration with qualified researchers: please contact Chris G.
-Sibley [c.sibley@auckland.ac.nz](c.sibley@auckland.ac.nz).
+The NZAVS is an ongoing longitudinal study. Our ethics requires us to
+protect the privacy of our participants. For this reason, we cannot post
+the raw data, even anonymous data. However, we provide data by request
+to qualified researchers for the purposes of replication. We areare
+interested in collaboration with qualified researchers: please contact
+Chris G. Sibley [c.sibley\@auckland.ac.nz](c.sibley@auckland.ac.nz).
 
@@
 \setlength{\leftskip}{0.5in}
-
-<div id="refs" custom-style="Bibliography"></div>
+```
+::: {#refs custom-style="Bibliography"}
+:::
+
 \endgroup
+
 <!-- ## model equations -->
+
 <!-- ```{r} -->
+
 <!-- stancode(mod.1_b) -->
-<!-- ``` -->
+
+<!-- ``` -->
+
 <!-- <!-- Frequentist model: -->
+
 <!-- ```{r eq1} -->
+
 <!-- library(equatiomatic) -->
+
 <!-- summary(mod.1aa) -->
+
 <!-- print(extract_eq(mod.1aa, wrap = TRUE)) -->
-<!-- ``` -->
+
+<!-- ``` -->
+
 <!-- ```{r eq2} -->
+
 <!-- library(equatiomatic) -->
+
 <!-- print(extract_eq(mod.1aa, wrap = TRUE, use_coefs=TRUE,terms_per_line = 3)) -->
-<!-- ``` -->
+
+<!-- ``` -->
+
 <!-- We might want to look at within/between indivual change (i.e.
@@
 For more on demeaning see:  -->
+
 <!-- https://easystats.github.io/parameters/articles/demean.html -->
+
 <!-- ```{r} -->
+
 <!-- dfben <- cbind( -->
+
 <!--   dfben, -->
+
 <!--   parameters::demean(dfben, select = c("Beliefs"), group = "Id") -->
+
 <!-- ) -->
-<!-- ``` -->
+
+<!-- ``` -->
+
 <!-- ```{r models, cache = TRUE, dependson = "dfben"} -->
+
 <!-- ### Spirit Belief Models ###  -->
+
 <!-- mod01<-lmer (PWI ~ Years + -->
+
 <!--                 Beliefs__Skeptic__between  +   -->
+
 <!--                 Beliefs__SpiritExcludesGod__between +  -->
+
 <!--                 Beliefs_GodAndSpirit_between  +  -->
+
 <!--                 Beliefs_GodExcludesSpirit_between + -->
+
 <!--                 Beliefs__Skeptic__within  +   -->
+
 <!--                 Beliefs__SpiritExcludesGod__within +  -->
+
 <!--                 Beliefs_GodAndSpirit_within  +  -->
+
 <!--                 Beliefs_GodExcludesSpirit_within + -->
+
 <!--                 (1|Id), data= dfben, na.action = na.omit) -->
+
 <!-- parameters::check_heterogeneity(mod01, group = "ID") -->
+
 <!-- parameters::check_heterogeneity(mod01, group = "ID") -->
+
 <!-- mod02<-lmer (LIFESAT ~ Years + -->
+
 <!--                 Beliefs__Skeptic__between  +   -->
+
 <!--                 Beliefs__SpiritExcludesGod__between +  -->
+
 <!--                 Beliefs_GodAndSpirit_between  +  -->
+
 <!--                 Beliefs_GodExcludesSpirit_between + -->
+
 <!--                 Beliefs__Skeptic__within  +   -->
+
 <!--                 Beliefs__SpiritExcludesGod__within +  -->
+
 <!--                 Beliefs_GodAndSpirit_within  +  -->
+
 <!--                 Beliefs_GodExcludesSpirit_within + -->
+
 <!--                 (1|Id), data= dfben, na.action = na.omit) -->
+
 <!-- parameters::check_heterogeneity(mod02, group = "ID") -->
+
 <!-- # inference holds, but can we use this method with factors? What do the contrasts mean?  -->
+
 <!-- tab_model(mod01) -->
+
 <!-- tab_model(mod02) -->
+
 <!-- plot_models(mod01,mod02) -->
-<!-- ``` -->
+
+<!-- ``` -->
+
 <!-- ```{r eq1} -->
+
 <!-- #for model equation -->
+
 <!-- library(equatiomatic) -->
+
 <!-- print(extract_eq(mod.1, wrap = TRUE,terms_per_line =3)) -->
-<!-- ``` -->
+
+<!-- ``` -->
+
 <!-- PWI Model 2 equation -->
+
 <!-- ```{r eq2} -->
+
 <!-- library(equatiomatic) -->
+
 <!-- print(extract_eq(mod.2, wrap = TRUE, use_coefs=TRUE,terms_per_line = 3)) -->
-<!-- ``` -->
+
+<!-- ``` -->
+
 <!-- ## report -->
+
 <!-- ### Model 1 -->
+
 <!-- model 1 report: for publication change all "significants" to "reliable" -->
+
 <!-- short report: -->
+
 <!-- ```{r, report1, cache =TRUE, dependson=mod.1,include=FALSE} -->
+
 <!-- easystats::model_text(mod.1) -->
-<!-- ``` -->
+
+<!-- ``` -->
+
 <!-- short report -->
+
 <!-- ```{r, cache=TRUE, include=FALSE } -->
+
 <!-- library(report) -->
+
 <!-- r1<- easystats::report(mod.1) -->
+
 <!-- easystats::table_short(r1) -->
+
 <!-- easystats::text_short(r1) -->
-<!-- ``` -->
+
+<!-- ``` -->
+
 <!-- ```{r, cache =TRUE, dependson=mod.1, include=FALSE} -->
+
 <!-- r1<- easystats::report(mod.1) -->
+
 <!-- easystats::text_long(r1) -->
-<!-- ``` -->
+
+<!-- ``` -->
+
 <!-- short report -->
+
 <!-- ```{r, cache =TRUE, dependson=mod.2,include=FALSE} -->
+
 <!-- # Model 2 -->
+
 <!-- # For publication change all "significants" to "reliable" -->
+
 <!-- easystats::model_text(mod.2) -->
-<!-- ``` -->
+
+<!-- ``` -->
+
 <!-- short report -->
+
 <!-- ```{r,include=FALSE} -->
+
 <!-- r2 <- easystats::report(mod.2) -->
+
 <!-- easystats::text_short(r2) -->
+
 <!-- easystats::table_short(r2) -->
-<!-- ``` -->
+
+<!-- ``` -->
+
 <!-- long report -->
+
 <!-- ```{r,include=FALSE} -->
+
 <!-- r2<- easystats::report(mod.2) -->
+
 <!-- easystats::text_long(r2) -->
-<!-- ``` -->
+
+<!-- ``` -->
+
 <!-- ## Equation 1: Personal Well-being  -->
+
 <!-- model equations for models without random intercept slope correlations; not correct, or reported\[ -->
+
 <!-- \begin{aligned} -->
+
 <!-- \operatorname{{\small PWI}}  &\sim N \left(\mu,\sigma^2 \right) \\  -->
+
 <!-- \mu & =  \boldsymbol{a} +  \boldsymbol{b} \\ -->
+
 <!-- & \quad \boldsymbol{a} =  \alpha_{0} + \alpha_{j[i]} \\ -->
+
 <!-- & \quad \alpha_{0} \sim \operatorname{Student T}^{+}(3,7.2,2.5) \\ -->
+
 <!-- & \quad  \alpha_{j[i]}  \sim \operatorname{Normal}(0, \sigma_j) \\  -->
+
 <!-- & \quad \sigma_j \sim \operatorname{HalfStudent T}^{+}(3, 0, 2.5) \\ -->
+
 <!-- & \quad   \boldsymbol{b}   \sim \operatorname{Student T}^{+}(0, 2.5) \\  -->
+
 <!-- & \quad \sigma  \sim  \operatorname{HalfStudent T}^{+}(3, 0, 2.5) \\ -->
+
 <!-- & \quad  \boldsymbol{b} = \\ -->
+
 <!-- & \quad \beta_{2}(\operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\ -->
+
 <!-- &\quad \beta_{3}(\operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ +  \beta_{4}(\operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}})\ +\\ -->
+
 <!-- & \quad \beta_{5}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\  -->
+
 <!-- & \quad \beta_{17}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ + \\ -->
+
 <!-- &\quad \beta_{18}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) \\  -->
+
 <!-- \end{aligned} -->
+
 <!-- \] -->
+
 <!-- ## Equation 2: Life Satisfaction. -->
+
 <!-- \[ -->
+
 <!-- \begin{aligned} -->
+
 <!-- \operatorname{{\small LifeSatisfaction}}  &\sim N \left(\mu,\sigma^2 \right) \\  -->
+
 <!-- \mu & =  \boldsymbol{a} +  \boldsymbol{b} \\ -->
+
 <!-- & \quad \boldsymbol{a} =  \alpha_{0} + \alpha_{j[i]} \\ -->
+
 <!-- & \quad \alpha_{0} \sim \operatorname{Student T}^{+}(3,5.5, 2.5) \\ -->
+
 <!-- & \quad  \alpha_{j[i]}  \sim \operatorname{Normal}(0, \sigma_j) \\  -->
+
 <!-- & \quad \sigma_j \sim \operatorname{HalfStudent T}^{+}(3, 0, 2.5) \\ -->
+
 <!-- & \quad   \boldsymbol{b}   \sim \operatorname{Student T}^{+}(0, 2.5) \\  -->
+
 <!-- & \quad \sigma  \sim  \operatorname{HalfStudent T}^{+}(3, 0, 2.5) \\ -->
+
 <!-- & \quad  \boldsymbol{b} = \\ -->
+
 <!-- & \quad \beta_{2}(\operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\ -->
+
 <!-- &\quad \beta_{3}(\operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ +  \beta_{4}(\operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}})\ +\\ -->
+
 <!-- & \quad \beta_{5}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\  -->
+
 <!-- & \quad \beta_{17}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ + \\ -->
+
 <!-- &\quad \beta_{18}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) \\  -->
+
 <!-- \end{aligned} -->
+
 <!-- \] -->
+
 <!-- ## Equation 2 -->
+
 <!-- \[ -->
+
 <!-- \begin{aligned} -->
+
 <!-- \operatorname{{\small LIFESAT/PWI}}  &\sim N \left(\mu,\sigma^2 \right) \\ \mu &=\alpha_{j[i]} + \beta_{1}(\operatorname{Years}) + \beta_{2}(\operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\ -->
+
 <!-- &\quad \beta_{3}(\operatorname{Beliefs}_{\operatorname{GodAndSpirit}}) + \beta_{4}(\operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) + \beta_{5}(\operatorname{Age.10yrs.C})\ + \\ -->
+
 <!-- &\quad \beta_{6}(\operatorname{Deprivation.S}) + \beta_{7}(\operatorname{Edu.S}) + \beta_{8}(\operatorname{Employed})\ + \\ -->
+
 <!-- &\quad \beta_{9}(\operatorname{EthnicCats}_{\operatorname{Maori}}) + \beta_{10}(\operatorname{EthnicCats}_{\operatorname{Pacific}}) + \beta_{11}(\operatorname{EthnicCats}_{\operatorname{Asian}})\ + \\ -->
+
 <!-- &\quad \beta_{12}(\operatorname{Male}_{\operatorname{1}}) + \beta_{13}(\operatorname{Partner}_{\operatorname{Has\ }}) + \beta_{14}(\operatorname{Pol.Orient.S})\ + \\ -->
+
 <!-- &\quad \beta_{15}(\operatorname{Urban}) + \beta_{16}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}}) + \beta_{17}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ + \\ -->
+
 <!-- &\quad \beta_{18}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) \\ \alpha_{j} &\sim N \left(\mu_{\alpha_{j}},\sigma^2_{\alpha_{j}} \right) , \operatorname{ for  Id }~j~= 1, \dots~J = 20,979~\operatorname{Individuals} -->
+
 <!-- \end{aligned} -->
+
 <!-- \] -->
+
 <!-- \[ -->
+
 <!-- \begin{table}[ ht ]  -->
+
 <!-- \centering  -->
+
 <!-- \caption{}\label{} -->
+
 <!-- \scalebox{.6}{ -->
+
 <!-- \begin{tabular}{ l c c c c c c c c c } -->
+
 <!-- \toprule -->
+
 <!--  &   \multicolumn{ 9 }{c}{ Wave }\\  -->
+
 <!--   & 2010 & 2011 & 2012 & 2013 & 2014 & 2015 & 2016 & 2017 & 2018 \\  -->
+
 <!--  & n = \emph{missing} & n = \emph{missing} & n = 8547 & n = 13472 & n = \emph{missing} & n = 11687 & n = 16048 & n = 14174 & n = 14883 \\  -->
+
 <!--  \midrule -->
+
 <!-- Age &   &   &   &   &   &   &   &   &  \\  -->
+
 <!-- NZdep &   &   &   &   &   &   &   &   &  \\  -->
+
 <!-- Education &   &   &   &   &   &   &   &   &  \\  -->
+
 <!-- Employed &   &   &   &   &   &   &   &   &  \\  -->
+
 <!-- \hspace{6pt}    0 &  (\%) &  (\%) & 2094 (24.5\%) & 2955 (21.9\%) &  (\%) & 2525 (21.6\%) & 3486 (21.7\%) & 3223 (22.7\%) & 3404 (22.9\%)\\  -->
+
 <!-- \hspace{6pt}    1 &  (\%) &  (\%) & 6453 (75.5\%) & 10517 (78.1\%) &  (\%) & 9162 (78.4\%) & 12562 (78.3\%) & 10951 (77.3\%) & 11479 (77.1\%)\\  -->
+
 <!-- EthnicCategories &   &   &   &   &   &   &   &   &  \\  -->
+
 <!-- \hspace{6pt}    Euro &  (\%) &  (\%) & 6731 (78.8\%) & 11024 (81.8\%) &  (\%) & 9671 (82.8\%) & 13389 (83.4\%) & 11814 (83.3\%) & 12427 (83.5\%)\\  -->
+
 <!-- \hspace{6pt}    Maori &  (\%) &  (\%) & 1188 (13.9\%) & 1608 (11.9\%) &  (\%) & 1341 (11.5\%) & 1749 (10.9\%) & 1620 (11.4\%) & 1677 (11.3\%)\\  -->
+
 <!-- \hspace{6pt}    Pacific &  (\%) &  (\%) & 284 (3.3\%) & 339 (2.5\%) &  (\%) & 277 (2.4\%) & 331 (2.1\%) & 243 (1.7\%) & 261 (1.8\%)\\  -->
+
 <!-- \hspace{6pt}    Asian &  (\%) &  (\%) & 344 (4\%) & 501 (3.7\%) &  (\%) & 398 (3.4\%) & 579 (3.6\%) & 497 (3.5\%) & 518 (3.5\%)\\  -->
+
 <!-- Male &   &   &   &   &   &   &   &   &  \\  -->
+
 <!-- \hspace{6pt}    0 &  (\%) &  (\%) & 5302 (62\%) & 8457 (62.8\%) &  (\%) & 7325 (62.7\%) & 10148 (63.2\%) & 9011 (63.6\%) & 9433 (63.4\%)\\  -->
+
 <!-- \hspace{6pt}    1 &  (\%) &  (\%) & 3245 (38\%) & 5015 (37.2\%) &  (\%) & 4362 (37.3\%) & 5900 (36.8\%) & 5163 (36.4\%) & 5450 (36.6\%)\\  -->
+
 <!-- Partner &   &   &   &   &   &   &   &   &  \\  -->
+
 <!-- \hspace{6pt}    0 &  (\%) &  (\%) & 2408 (28.2\%) & 3546 (26.3\%) &  (\%) & 2815 (24.1\%) & 3773 (23.5\%) & 3330 (23.5\%) & 3556 (23.9\%)\\  -->
+
 <!-- \hspace{6pt}    1 &  (\%) &  (\%) & 6139 (71.8\%) & 9926 (73.7\%) &  (\%) & 8872 (75.9\%) & 12275 (76.5\%) & 10844 (76.5\%) & 11327 (76.1\%)\\  -->
+
 <!-- Pol Orient &   &   &   &   &   &   &   &   &  \\  -->
+
 <!-- Urban &   &   &   &   &   &   &   &   &  \\  -->
+
 <!-- \hspace{6pt}    Not\_Urban &  (\%) &  (\%) & 2865 (33.5\%) & 4278 (31.8\%) &  (\%) & 3951 (33.8\%) & 5538 (34.5\%) & 2610 (18.4\%) & 2661 (17.9\%)\\  -->
+
 <!-- \hspace{6pt}    Urban &  (\%) &  (\%) & 5682 (66.5\%) & 9194 (68.2\%) &  (\%) & 7736 (66.2\%) & 10510 (65.5\%) & 11564 (81.6\%) & 12222 (82.1\%)\\  -->
+
 <!-- Beliefs &   &   &   &   &   &   &   &   &  \\  -->
+
 <!-- \hspace{6pt}    \_Skeptic\_ &  (\%) &  (\%) & 2048 (24\%) & 3286 (24.4\%) &  (\%) & 2951 (25.3\%) & 4116 (25.6\%) & 3916 (27.6\%) & 4120 (27.7\%)\\  -->
+
 <!-- \hspace{6pt}    \_SpiritExcludesGod\_ &  (\%) &  (\%) & 2295 (26.9\%) & 3735 (27.7\%) &  (\%) & 3169 (27.1\%) & 4501 (28\%) & 3918 (27.6\%) & 3970 (26.7\%)\\  -->
+
 <!-- \hspace{6pt}    GodAndSpirit &  (\%) &  (\%) & 3707 (43.4\%) & 5499 (40.8\%) &  (\%) & 5119 (43.8\%) & 6453 (40.2\%) & 5464 (38.5\%) & 6210 (41.7\%)\\  -->
+
 <!-- \hspace{6pt}    GodExcludesSpirit &  (\%) &  (\%) & 497 (5.8\%) & 952 (7.1\%) &  (\%) & 448 (3.8\%) & 978 (6.1\%) & 876 (6.2\%) & 583 (3.9\%)\\  -->
+
 <!-- LIFESAT &   &   &   &   &   &   &   &   &  \\  -->
+
 <!-- PWI &   &   &   &   &   &   &   &   &  \\  -->
+
 <!-- \bottomrule -->
+
 <!-- \end{tabular} -->
+
 <!-- } -->
+
 <!-- \end{table} -->
+
 <!-- \] -->
+
 <!-- \[ -->
+
 <!-- \begin{table} -->
+
 <!-- \begin{center} -->
+
 <!-- \scalebox{0.6}{ -->
+
 <!-- \begin{tabular}{l c c} -->
+
 <!-- \toprule -->
+
 <!--  & PWI & Life Satisfaction \\ -->
+
 <!-- \midrule -->
+
 <!-- (Intercept)                        & $\mathbf{6.63}^{***}$  & $\mathbf{4.94}^{***}$  \\ -->
+
 <!--                                    & $(0.03)$               & $(0.02)$               \\ -->
+
 <!-- Years                              & $0.00$                 & $-0.00$                \\ -->
+
 <!--                                    & $(0.00)$               & $(0.00)$               \\ -->
+
 <!-- Beliefs\_SpiritExcludesGod\_       & $\mathbf{-0.14}^{***}$ & $\mathbf{-0.08}^{***}$ \\ -->
+
 <!--                                    & $(0.03)$               & $(0.02)$               \\ -->
+
 <!-- BeliefsGodAndSpirit                & $\mathbf{-0.16}^{***}$ & $-0.04$                \\ -->
+
 <!--                                    & $(0.03)$               & $(0.02)$               \\ -->
+
 <!-- BeliefsGodExcludesSpirit           & $0.02$                 & $0.07$                 \\ -->
+
 <!--                                    & $(0.05)$               & $(0.04)$               \\ -->
+
 <!-- Age.10yrs.C                        & $\mathbf{0.08}^{***}$  & $\mathbf{0.09}^{***}$  \\ -->
+
 <!--                                    & $(0.01)$               & $(0.01)$               \\ -->
+
 <!-- Deprivation.S                      & $\mathbf{-0.14}^{***}$ & $\mathbf{-0.07}^{***}$ \\ -->
+
 <!--                                    & $(0.01)$               & $(0.01)$               \\ -->
+
 <!-- Edu.S                              & $\mathbf{0.12}^{***}$  & $\mathbf{0.06}^{***}$  \\ -->
+
 <!--                                    & $(0.01)$               & $(0.01)$               \\ -->
+
 <!-- EmployedEmployed                   & $\mathbf{0.09}^{***}$  & $\mathbf{0.08}^{***}$  \\ -->
+
 <!--                                    & $(0.01)$               & $(0.01)$               \\ -->
+
 <!-- EthnicCatsMaori                    & $\mathbf{-0.30}^{***}$ & $\mathbf{-0.06}^{**}$  \\ -->
+
 <!--                                    & $(0.03)$               & $(0.02)$               \\ -->
+
 <!-- EthnicCatsPacific                  & $\mathbf{-0.29}^{***}$ & $\mathbf{-0.08}^{*}$   \\ -->
+
 <!--                                    & $(0.05)$               & $(0.04)$               \\ -->
+
 <!-- EthnicCatsAsian                    & $-0.08$                & $-0.04$                \\ -->
+
 <!--                                    & $(0.05)$               & $(0.03)$               \\ -->
+
 <!-- Male1                              & $\mathbf{-0.11}^{***}$ & $\mathbf{-0.19}^{***}$ \\ -->
+
 <!--                                    & $(0.02)$               & $(0.01)$               \\ -->
+
 <!-- PartnerHas Partner                 & $\mathbf{0.61}^{***}$  & $\mathbf{0.42}^{***}$  \\ -->
+
 <!--                                    & $(0.02)$               & $(0.01)$               \\ -->
+
 <!-- Pol.Orient.S                       & $\mathbf{0.02}^{***}$  & $\mathbf{0.01}^{**}$   \\ -->
+
 <!--                                    & $(0.01)$               & $(0.00)$               \\ -->
+
 <!-- UrbanUrban                         & $\mathbf{-0.02}^{*}$   & $\mathbf{-0.02}^{*}$   \\ -->
+
 <!--                                    & $(0.01)$               & $(0.01)$               \\ -->
+
 <!-- Years:Beliefs\_SpiritExcludesGod\_ & $\mathbf{0.02}^{**}$   & $\mathbf{0.02}^{***}$  \\ -->
+
 <!--                                    & $(0.00)$               & $(0.00)$               \\ -->
+
 <!-- Years:BeliefsGodAndSpirit          & $\mathbf{0.02}^{***}$  & $\mathbf{0.01}^{***}$  \\ -->
+
 <!--                                    & $(0.00)$               & $(0.00)$               \\ -->
+
 <!-- Years:BeliefsGodExcludesSpirit     & $0.00$                 & $-0.00$                \\ -->
+
 <!--                                    & $(0.01)$               & $(0.01)$               \\ -->
+
 <!-- \midrule -->
+
 <!-- AIC                                & $248580.18$            & $200599.16$            \\ -->
+
 <!-- BIC                                & $248775.08$            & $200793.96$            \\ -->
+
 <!-- Log Likelihood                     & $-124269.09$           & $-100278.58$           \\ -->
+
 <!-- Num. obs.                          & $79270$                & $78888$                \\ -->
+
 <!-- Num. groups: Id                    & $20979$                & $20972$                \\ -->
+
 <!-- Var: Id (Intercept)                & $1.78$                 & $0.88$                 \\ -->
+
 <!-- Var: Residual                      & $0.74$                 & $0.42$                 \\ -->
+
 <!-- \bottomrule -->
+
 <!-- \multicolumn{3}{l}{\scriptsize{$^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$}} -->
+
 <!-- \end{tabular} -->
+
 <!-- } -->
+
 <!-- \caption{Regression Models} -->
+
 <!-- \label{table:coefficients} -->
+
 <!-- \end{center} -->
+
 <!-- \end{table} -->
+
 <!-- \] -->
+
 <!-- \newpage -->
+
 <!-- ## Personal Well-Being -->
+
 <!-- We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict personal well-being (PWI) with years (0-9), beliefs (four categories), Age.10yrs.C, deprivation (S), education (S), employed, ethnic categories, male, partner, political conservativism (S) and urban, with Id as a random effect.
@@
 Within this model: -->
+
 <!--   - The effect of Years is positive is not reliable (beta = 2.66e-03, SE = 3.58e-03, std.
@@
 -->
+
 <!--   - The effect of Beliefs in *Spirit Excludes God*  is reliably negative (beta = -0.14, SE = 0.03, std.
@@
 -->
+
 <!--   - The effect of Beliefs in *God And Spirit*  is reliably negative (beta = -0.16, SE = 0.03, std.
@@
 -->
+
 <!--   - The effect of Beliefs in *God Excludes Spirit*  is not reliable (beta = 0.02, SE = 0.05, std.
@@
 -->
+
 <!--   - The effect of Age.10yrs.C is reliably positive  (beta = 0.08, SE = 7.38e-03, std. beta = 0.07, p < .001). Older people tend to have greater personal well-being. -->
+
 <!--   - The effect of Deprivation (S) is reliably negative (beta = -0.14, SE = 7.50e-03, std. beta = -0.08, p < .001). Deprived people tend to have lower personal well-being. -->
+
 <!--   - The effect of Education (S) is reliably positive  (beta = 0.12, SE = 8.74e-03, std. beta = 0.07, p < .001). Educated people tend to have greater personal well-being. -->
+
 <!--   - The effect of Employed is reliably positive  (beta = 0.09, SE = 0.01, std. beta = 0.05, p < .001). -->
+
 <!--   - The effect of identifying as M\={a}ori is reliably negative (beta = -0.30, SE = 0.03, std. beta = -0.18, p < .001). -->
+
 <!--   - The effect of identifying as Pacific is reliably negative (beta = -0.29, SE = 0.05, std. beta = -0.17, p < .001).  -->
+
 <!--   - The effect of identifying as Asian is not reliable (beta = -0.08, SE = 0.05, std. beta = -0.05, p = 0.089). -->
+
 <!--   - The effect of identifying as male is reliably negative (beta = -0.11, SE = 0.02, std. beta = -0.06, p < .001).  -->
+
 <!--   - The effect of having a partner is reliably positive (beta = 0.61, SE = 0.02, std. beta = 0.37, p < .001). -->
+
 <!--   - The effect of political conservativism (S) is reliably positive  (beta = 0.02, SE = 5.67e-03, std. beta = 0.01, p < .001).  -->
+
 <!--   - The effect of living in an Urban area is reliably negative (beta = -0.02, SE = 0.01, std. beta = -0.01, p < .05). T -->
+
 <!-- ##  Personal well-being: years X beliefs  -->
+
 <!-- ```{r fig.cap = "Combination Coefficient plots for personal well-being and life satisfaction longitudinal growth models"} -->
+
 <!-- knitr::include_graphics("Figs/USEPWI_expected-1.png") -->
-<!-- ``` -->
+
+<!-- ``` -->
+
 <!-- Focusing on the interaction of time with belief-type on personal well-being we observe:  -->
@@
 -->
+
 <!--   - The effect of years $\times$ in *God And Spirit*  is reliably positive  (beta = 0.02, SE = 4.41e-03, std.
@@
 We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their personal well-being over time -->
+
 <!--   - The effect of Years $\times$ belief in *God Excludes Spirit*  is not reliable(beta = 1.96e-03, SE = 8.29e-03, std.
@@
 -->
+
 <!-- ## Life Satisfaction  -->
+
 <!-- We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict personal well-being (PWI) with years (0-9), beliefs (four categories), Age.10yrs.C, deprivation (S), education (S), employed, ethnic categories, male, partner, political conservativism (S) and urban, with Id as a random effect.
@@
 Within this model: -->
+
 <!--   - The effect of years is not reliable (beta = -5.50e-04, SE = 2.69e-03, std.
@@
 -->
+
 <!--   - The effect of beliefs in *Spirit Excludes God* is reliably negative (beta = -0.08, SE = 0.02, std.
@@
 -->
+
 <!--   - The effect of beliefs in *God And Spirit* is not reliable (beta = -0.04, SE = 0.02, std.
@@
 -->
+
 <!--   - The effect of beliefs in *God Excludes Spirit* is reliably positive (beta = 0.07, SE = 0.04, std.
@@
 -->
+
 <!--   - The effect of Age.10yrs.C is reliably positive (beta = 0.09, SE = 5.26e-03, std. beta = 0.10, p < .001). Older people have greater Life Satisfaction -->
+
 <!--   - The effect of Deprivation (S) is  is reliably negative (beta = -0.07, SE = 5.50e-03, std. beta = -0.06, p < .001). People who are more deprived have lower Life Satisfaction. -->
+
 <!--   - The effect of Education (S) is reliably positive (beta = 0.06, SE = 6.33e-03, std. beta = 0.05, p < .001). Educated people have greater Life Satisfaction -->
+
 <!--   - The effect of being Employed is reliably positive (beta = 0.08, SE = 9.69e-03, std. beta = 0.07, p < .001). People who are employed tend to have greater Life Satisfaction -->
+
 <!--   - The effect of identifying as Maori is reliably negative (beta = -0.06, SE = 0.02, std.
@@
 -->
+
 <!--   - The effect of identifying as Pacific is reliably negative (beta = -0.08, SE = 0.04, std.
@@
 -->
+
 <!--   - The effect of identifying as Asian is not reliably different from people of European ethnicity in life satisfaction(beta = -0.04, SE = 0.03, std. beta = -0.03, p = 0.223). -->
+
 <!--   - The effect of identifying as male is reliably negative (beta = -0.19, SE = 0.01, std. beta = -0.16, p < .001). -->
+
 <!--   - The effect of having a partner is reliably positive (beta = 0.42, SE = 0.01, std. beta = 0.35, p < .001). -->
+
 <!--   - The effect of political conservativism (S) is reliably positive (beta = 0.01, SE = 4.23e-03, std. beta = 0.01, p < .01). -->
+
 <!--   - The effect of living in an urban area is reliably negative (beta = -0.02, SE = 8.85e-03, std. beta = -0.02, p < .05). -->
+
 <!-- ## Life Satisfaction: years X beliefs  -->
+
 <!-- ```{r fig.cap = "Prediction Plot Life Satisfaction Spirit Belief Longitudinal Models"} -->
+
 <!-- knitr::include_graphics("/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Figs/USElifsat_expected-1.png") -->
-<!-- ``` -->
+
+<!-- ``` -->
+
 <!--   - The interaction of years $\times$ beliefs in *Spirit Excludes God* ludesGod is reliably positive (beta = 0.02, SE = 3.66e-03, std.
@@
 -->
+
 <!--   - The interaction of years $\times$ *God And Spirit*  is reliably positive (beta = 0.01, SE = 3.32e-03, std.
@@
 We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their life satisfaction over time -->
+
 <!--   - The effect of years $\times$ beliefs in *God Excludes Spirit*  is not reliable (beta = -1.59e-03, SE = 6.25e-03, std.

### 2021-02-26 funs.R

@@ -1,9 +1,9 @@
 # funs.R
 # functions used in this analysis
 data_read_mymachine <- function(){
-  out <-readRDS("~/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/00Bulbulia\ Pubs/2020/ldf.5")
+  out <-readRDS("/Users/jbul176/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/00Bulbulia\ Pubs/2021/DATA/ldf.5")
 }
-
+#  #out <-readRDS("~/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/00Bulbulia\ Pubs/2020/ldf.5") # previous submission did not have 2019/20 waves
 # this code is for citing packages
 
 cite_packages <- function() {
@@ -21,14 +21,14 @@ data_clean_spirit_wellbeing <- function(df,y) {
   #function for cleaning this dataset for any arbitrary number of waves
   n_waves =  as.numeric(y)
   out <- df %>%
-    dplyr::mutate(Beliefs = factor(ifelse(df$Believe.God == "Not Believe God" & df$Believe.Spirit == "Not Believe Spirit", "_Skeptic_",
-                                          ifelse(df$Believe.God == "Not Believe God" & df$Believe.Spirit == "Believe Spirit", "_SpiritExcludesGod_",
+    dplyr::mutate(Beliefs = factor(ifelse(df$Believe.God == "Not Believe God" & df$Believe.Spirit == "Not Believe Spirit", "Disbeliever",
+                                          ifelse(df$Believe.God == "Not Believe God" & df$Believe.Spirit == "Believe Spirit", "SpiritExcludesGod",
                                                  ifelse(df$Believe.God == "Believe God" & df$Believe.Spirit == "Believe Spirit","GodAndSpirit","GodExcludesSpirit")))))%>%
     dplyr::filter(Wave !=2009)%>% # exclude wave 1 (no spirit measures that wave)
-    droplevels() %>% dropp unused waves
+    droplevels() %>% #dropp unused waves
     dplyr::filter(YearMeasured==1)%>%
     dplyr::group_by(Id) %>%
-    dplyr::filter(n() > n_waves-1)%>% # select those who have responded to at least 3 waves #
+    dplyr::filter(n() > n_waves-1)%>% # select those who have responded to at least n waves #
     dplyr::filter(n() !=0)%>%
[…]

### 2021-03-11 Manuscript_Bayesian.Rmd
updated manuscript

@@
 Notably, spirituality and health research has not converged on a unique set of meanings.
-Although it is commonplace across most areas of psychological science for researchers to investigate different operationalisations of focal concepts (think of memory research, or emotions research), the psychology of spirituality has mostly been conducted with cross-sectional North American samples among participants who affiliate with traditional religion [@Ano2005-hx; @Hackney2003-rs; @Sawatzky2005-rw; @Smith2003-re; @Visser2010-kq; @Yonker2012-zg].
+Although it is commonplace across most areas of psychological science for researchers to investigate different operationalisations of focal concepts (think of memory research, or emotions research), the psychology of spirituality has mostly been conducted with cross-sectional North American samples among participants who affiliate with traditional religion [@Ano2005-hx; @Hackney2003-rs; @Sawatzky2005-rw; @Smith2003-re; @Visser2010-kq; @Yonker2012-zg; @holt2003spiritual].
 Whether the psychology of spirituality and health has repeated or extended the psychology of religion and health remains unclear.
-Put simply, cnceptual overlap between spirituality constructs and religion constructs creates a second line of terminological confusion that interacts with terminological confusion arising from overlap between spirituality constructs and well-being constructs.
+Put simply, conceptual overlap between spirituality constructs and religion constructs creates a second line of terminological confusion that interacts with terminological confusion arising from overlap between spirituality constructs and well-being constructs.
 Indeed, researchers who have operationalised spirituality constructs as distinct from religion constructs have observed a negative association between spirituality and subjective well-being [@King2013-cg].
-Thus to infer how spirituality affects subjective well-being, it is not sufficient to disentange spirituality constructs from tautological well-being constructs, it is also important to disentangle the concept of spirituality from the concept of religion, and to carefully investigate the mechanisms of well-being in populations who overlap, partially overlap, and differ in these dimensions.
+Thus to infer how spirituality affects subjective well-being, it is not sufficient to disentangle spirituality constructs from tautological well-being constructs, it is also important to disentangle the concept of spirituality from the concept of religion, and to carefully investigate the mechanisms of well-being in populations who overlap, partially overlap, and differ in these dimensions.
 
 A third challenge is to investigate the relationship between spirituality and subjective well-being in nationally diverse samples over individual life-spans.
-To assess the mechanisms of well-being production requires observing change within people who differ in theoregically relevant ways and assessing how their lives turn out [@Garssen2016-km].
-However, previous investigations of spirituality and well-being, including high-quality reviews and meta-analyses, have nearly exclusively relied on cross-sectional samples [@Ano2005-hx; @Hackney2003-rs; @Sawatzky2005-rw; @Smith2003-re; @Visser2010-kq; @Yonker2012-zg].
-Although researching the key questions in spirituality psychological health research requires national longitudinal investigations, unfortunately the relevant datasets are rare.
+To assess the mechanisms of well-being production requires observing change within people who differ in theoretically relevant ways and assessing how their lives turn out [@Garssen2016-km].
+However, previous investigations of spirituality and well-being, including high-quality reviews and meta-analyses, have nearly exclusively relied on cross-sectional samples [@Ano2005-hx; @Hackney2003-rs; @Sawatzky2005-rw; @Smith2003-re; @Visser2010-kq; @Yonker2012-zg; @januu2020associations], and sampled from small sub-populations such as older adults [@lawler2009role], urban African American women [@holt2003spiritual], rural Kenyan men [@goodman2020mental], and Ghanaian parents of children with special needs [@dey2019spirituality].
+Such limitations have been evident over 20 years (see for example: [@thoresen1999spirituality].
+However, until recently, national-scale longitudinal time-series data required to investigate the long term effects of spirituality have been lacking.
 
@@
 
-Because missing values can result in biased estimates [@Blackwell2017-oq], we multiply imputed 10 datasets using the AmeliaII package in R[@Honaker2011-yu].We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences.
+Because missing values can result in biased estimates [@Blackwell2017-oq], we multiply imputed 10 datasets using the Amelia II package in R[@Honaker2011-yu].We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences.
 
@@
 The analysis was performed using R version 4.0.2 (2020-06-22).
-The Platform was x86_64-apple-darwin17.0 (64-bit) Running under: macOS Catalina 10.15.6 We are greatful to the contributors and mantainers of the following packages: R [Version 4.0.3; @R-base] and the R-packages *Amelia* [Version 1.7.6; @R-Amelia], *citr* [Version 0.3.2; @R-citr], *dplyr* [Version 1.0.2; @R-dplyr], *equatiomatic* [Version 0.1.0.9000; @R-equatiomatic], *forcats* [Version 0.5.0; @R-forcats], *ggeffects* [Version 1.0.0; @R-ggeffects], *gghighlight* [Version 0.3.0; @R-gghighlight], *ggplot2* [Version 3.3.2; @R-ggplot2], *ggsci* [Version 2.9; @R-ggsci], *lme4* [Version 1.1.25; @R-lme4], *Matrix* [Version 1.2.18; @R-Matrix], *papaja* [Version 0.1.0.9997; @R-papaja], *parameters* [Version 0.9.0.1; @R-parameters], *patchwork* [Version 1.1.0; @R-patchwork], *prettycode* [@R-prettycode], *purrr* [Version 0.3.4; @R-purrr], *Rcpp* [Version 1.0.5; @R-Rcpp_a; @R-Rcpp_b], *readr* [Version 1.4.0; @R-readr], *report* [Version 0.2.0; @R-report], *see* [Version 0.6.0.1; @R-see], *sjPlot* [Version 2.8.6; @R-sjPlot], *stringr* [Version 1.4.0; @R-stringr], *styler* [Version 1.3.2; @R-styler], *table1* [Version 1.2.1; @R-table1], *texreg* [Version 1.37.5; @R-texreg], *tibble* [Version 3.0.4; @R-tibble], *tidyr* [Version 1.1.2; @R-tidyr], and *tidyverse* [Version 1.3.0; @R-tidyverse]
+The Platform was x86_64-apple-darwin17.0 (64-bit) Running under: macOS Catalina 10.15.6 We are grateful to the contributors and maintainers of the following packages: R [Version 4.0.3; @R-base] and the R-packages *Amelia* [Version 1.7.6; @R-Amelia], *citr* [Version 0.3.2; @R-citr], *dplyr* [Version 1.0.2; @R-dplyr], *equatiomatic* [Version 0.1.0.9000; @R-equatiomatic], *forcats* [Version 0.5.0; @R-forcats], *ggeffects* [Version 1.0.0; @R-ggeffects], *gghighlight* [Version 0.3.0; @R-gghighlight], *ggplot2* [Version 3.3.2; @R-ggplot2], *ggsci* [Version 2.9; @R-ggsci], *lme4* [Version 1.1.25; @R-lme4], *Matrix* [Version 1.2.18; @R-Matrix], *papaja* [Version 0.1.0.9997; @R-papaja], *parameters* [Version 0.9.0.1; @R-parameters], *patchwork* [Version 1.1.0; @R-patchwork], *prettycode* [@R-prettycode], *purrr* [Version 0.3.4; @R-purrr], *Rcpp* [Version 1.0.5; @R-Rcpp_a; @R-Rcpp_b], *readr* [Version 1.4.0; @R-readr], *report* [Version 0.2.0; @R-report], *see* [Version 0.6.0.1; @R-see], *sjPlot* [Version 2.8.6; @R-sjPlot], *stringr* [Version 1.4.0; @R-stringr], *styler* [Version 1.3.2; @R-styler], *table1* [Version 1.2.1; @R-table1], *texreg* [Version 1.37.5; @R-texreg], *tibble* [Version 3.0.4; @R-tibble], *tidyr* [Version 1.1.2; @R-tidyr], and *tidyverse* [Version 1.3.0; @R-tidyverse]
 
@@
 All code for this analysis and a full report of the missing data imputation can be found at [https://github.com/go-bayes/nzavs_spiritbeliefWellbeing](https://github.com/go-bayes/nzavs_spiritbeliefWellbeing){.uri}.
-An html document decribing the same analysis is located at: <https://osf.io/cu2gr/>.
+An html document describing the same analysis is located at: <https://osf.io/cu2gr/>.
 
@@
 
-Focusing on life satisfaction, people who believed in Spirit, either with a belief in God ($\hat{y}$ = 5.18 \| [5.13, 5.24]) or without a belief in God (5.18 \| [5.12, 5.24]) tended to be lower in life satsifaction than disbelievers in both a God and a Spirit (5.27 \| [5.20, 5.33]) and disbelievers in a spirit who also believed in a God (5.26 \| [5.15, 5.36]) (see Figure 3)
+Focusing on life satisfaction, people who believed in Spirit, either with a belief in God ($\hat{y}$ = 5.18 \| [5.13, 5.24]) or without a belief in God (5.18 \| [5.12, 5.24]) tended to be lower in life satisfaction than disbelievers in both a God and a Spirit (5.27 \| [5.20, 5.33]) and disbelievers in a spirit who also believed in a God (5.26 \| [5.15, 5.36]) (see Figure 3)
 
@@
 
-Because These inferred growth rates for personal well being and life satisfaction are evident from repeated measures across 10 annual waves of nationally diverse data collection, we are confident that in New Zealand, spiritual beliefs predict strong increases in personal well-being and life satisfaction
+Because these inferred growth rates for personal well being and life satisfaction are evident from repeated measures across 10 annual waves of nationally diverse data collection, we are confident that in New Zealand, spiritual beliefs predict strong increases in personal well-being and life satisfaction
 

### 2021-03-11 Manuscript_Bayesian.tex
updated manuscript

@@
 Notably, spirituality and health research has not converged on a unique set of meanings.
-Although it is commonplace across most areas of psychological science for researchers to investigate different operationalisations of focal concepts (think of memory research, or emotions research), the psychology of spirituality has mostly been conducted with cross-sectional North American samples among participants who affiliate with traditional religion (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky, Ratner, \& Chiu, 2005; Smith, McCullough, \& Poll, 2003; Visser, Garssen, \& Vingerhoets, 2010; Yonker, Schnabelrauch, \& Dehaan, 2012).
+Although it is commonplace across most areas of psychological science for researchers to investigate different operationalisations of focal concepts (think of memory research, or emotions research), the psychology of spirituality has mostly been conducted with cross-sectional North American samples among participants who affiliate with traditional religion (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Holt, Clark, Kreuter, \& Rubio, 2003; Sawatzky, Ratner, \& Chiu, 2005; Smith, McCullough, \& Poll, 2003; Visser, Garssen, \& Vingerhoets, 2010; Yonker, Schnabelrauch, \& Dehaan, 2012).
 Whether the psychology of spirituality and health has repeated or extended the psychology of religion and health remains unclear.
-Put simply, cnceptual overlap between spirituality constructs and religion constructs creates a second line of terminological confusion that interacts with terminological confusion arising from overlap between spirituality constructs and well-being constructs.
+Put simply, conceptual overlap between spirituality constructs and religion constructs creates a second line of terminological confusion that interacts with terminological confusion arising from overlap between spirituality constructs and well-being constructs.
 Indeed, researchers who have operationalised spirituality constructs as distinct from religion constructs have observed a negative association between spirituality and subjective well-being (King et al., 2013).
-Thus to infer how spirituality affects subjective well-being, it is not sufficient to disentange spirituality constructs from tautological well-being constructs, it is also important to disentangle the concept of spirituality from the concept of religion, and to carefully investigate the mechanisms of well-being in populations who overlap, partially overlap, and differ in these dimensions.
+Thus to infer how spirituality affects subjective well-being, it is not sufficient to disentangle spirituality constructs from tautological well-being constructs, it is also important to disentangle the concept of spirituality from the concept of religion, and to carefully investigate the mechanisms of well-being in populations who overlap, partially overlap, and differ in these dimensions.
 
 A third challenge is to investigate the relationship between spirituality and subjective well-being in nationally diverse samples over individual life-spans.
-To assess the mechanisms of well-being production requires observing change within people who differ in theoregically relevant ways and assessing how their lives turn out (Garssen \& Visser, 2016).
-However, previous investigations of spirituality and well-being, including high-quality reviews and meta-analyses, have nearly exclusively relied on cross-sectional samples (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky, Ratner, \& Chiu, 2005; Smith, McCullough, \& Poll, 2003; Visser, Garssen, \& Vingerhoets, 2010; Yonker, Schnabelrauch, \& Dehaan, 2012).
-Although researching the key questions in spirituality psychological health research requires national longitudinal investigations, unfortunately the relevant datasets are rare.
+To assess the mechanisms of well-being production requires observing change within people who differ in theoretically relevant ways and assessing how their lives turn out (Garssen \& Visser, 2016).
+However, previous investigations of spirituality and well-being, including high-quality reviews and meta-analyses, have nearly exclusively relied on cross-sectional samples (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Jan, Malinakova, Kosarkova, \& Tavel, 2020; Sawatzky, Ratner, \& Chiu, 2005; Smith, McCullough, \& Poll, 2003; Visser, Garssen, \& Vingerhoets, 2010; Yonker, Schnabelrauch, \& Dehaan, 2012), and sampled from small sub-populations such as older adults (Lawler-Row \& Elliott, 2009), urban African American women (Holt, Clark, Kreuter, \& Rubio, 2003), rural Kenyan men (Goodman, Gitari, Keiser, Elliott, \& Seidel, 2020), and Ghanaian parents of children with special needs (Dey, Amponsah, \& Wiafe-Akenteng, 2019).
+Such limitations have been evident over 20 years (see for example: (Thoresen, 1999).
+However, until recently, national-scale longitudinal time-series data required to investigate the long term effects of spirituality have been lacking.
 
@@
 
-Because missing values can result in biased estimates (Blackwell, Honaker, \& King, 2017), we multiply imputed 10 datasets using the AmeliaII package in R(Honaker, King, Blackwell, \& Others, 2011).We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences.
+Because missing values can result in biased estimates (Blackwell, Honaker, \& King, 2017), we multiply imputed 10 datasets using the Amelia II package in R(Honaker, King, Blackwell, \& Others, 2011).We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences.
 
@@
 The analysis was performed using R version 4.0.2 (2020-06-22).
-The Platform was x86\_64-apple-darwin17.0 (64-bit) Running under: macOS Catalina 10.15.6 We are greatful to the contributors and mantainers of the following packages: R {[}Version 4.0.3; R Core Team (2020){]} and the R-packages \emph{Amelia} {[}Version 1.7.6; Honaker, King, and Blackwell (2011){]}, \emph{citr} {[}Version 0.3.2; Aust (2019){]}, \emph{dplyr} {[}Version 1.0.2; Wickham, François, Henry, and Müller (2020){]}, \emph{equatiomatic} {[}Version 0.1.0.9000; Anderson and Heiss (2020){]}, \emph{forcats} {[}Version 0.5.0; Wickham (2020a){]}, \emph{ggeffects} {[}Version 1.0.0; Lüdecke (2018){]}, \emph{gghighlight} {[}Version 0.3.0; Yutani (2020){]}, \emph{ggplot2} {[}Version 3.3.2; Wickham (2016){]}, \emph{ggsci} {[}Version 2.9; Xiao (2018){]}, \emph{lme4} {[}Version 1.1.25; Bates, Mächler, Bolker, and Walker (2015){]}, \emph{Matrix} {[}Version 1.2.18; Bates and Maechler (2019){]}, \emph{papaja} {[}Version 0.1.0.9997; Aust and Barth (2020){]}, \emph{parameters} {[}Version 0.9.0.1; Lüdecke, Ben-Shachar, Patil, and Makowski (2020){]}, \emph{patchwork} {[}Version 1.1.0; Pedersen (2020){]}, \emph{prettycode} (Csárdi, 2019), \emph{purrr} {[}Version 0.3.4; Henry and Wickham (2020){]}, \emph{Rcpp} {[}Version 1.0.5; Eddelbuettel and François (2011); Eddelbuettel and Balamuta (2017){]}, \emph{readr} {[}Version 1.4.0; Wickham, Hester, and Francois (2018){]}, \emph{report} {[}Version 0.2.0; Makowski, Dominique, Lüdecke, and Daniel (2019){]}, \emph{see} {[}Version 0.6.0.1; Lüdecke, Ben-Shachar, Waggoner, and Makowski (2020){]}, \emph{sjPlot} {[}Version 2.8.6; Lüdecke (2020){]}, \emph{stringr} {[}Version 1.4.0; Wickham (2019){]}, \emph{styler} {[}Version 1.3.2; Müller and Walthert (2020){]}, \emph{table1} {[}Version 1.2.1; Rich (2020){]}, \emph{texreg} {[}Version 1.37.5; Leifeld (2013){]}, \emph{tibble} {[}Version 3.0.4; Müller and Wickham (2020){]}, \emph{tidyr} {[}Version 1.1.2; Wickham (2020b){]}, and \emph{tidyverse} {[}Version 1.3.0; Wickham et al.
+The Platform was x86\_64-apple-darwin17.0 (64-bit) Running under: macOS Catalina 10.15.6 We are grateful to the contributors and maintainers of the following packages: R {[}Version 4.0.3; R Core Team (2020){]} and the R-packages \emph{Amelia} {[}Version 1.7.6; Honaker, King, and Blackwell (2011){]}, \emph{citr} {[}Version 0.3.2; Aust (2019){]}, \emph{dplyr} {[}Version 1.0.2; Wickham, François, Henry, and Müller (2020){]}, \emph{equatiomatic} {[}Version 0.1.0.9000; Anderson and Heiss (2020){]}, \emph{forcats} {[}Version 0.5.0; Wickham (2020a){]}, \emph{ggeffects} {[}Version 1.0.0; Lüdecke (2018){]}, \emph{gghighlight} {[}Version 0.3.0; Yutani (2020){]}, \emph{ggplot2} {[}Version 3.3.2; Wickham (2016){]}, \emph{ggsci} {[}Version 2.9; Xiao (2018){]}, \emph{lme4} {[}Version 1.1.25; Bates, Mächler, Bolker, and Walker (2015){]}, \emph{Matrix} {[}Version 1.2.18; Bates and Maechler (2019){]}, \emph{papaja} {[}Version 0.1.0.9997; Aust and Barth (2020){]}, \emph{parameters} {[}Version 0.9.0.1; Lüdecke, Ben-Shachar, Patil, and Makowski (2020){]}, \emph{patchwork} {[}Version 1.1.0; Pedersen (2020){]}, \emph{prettycode} (Csárdi, 2019), \emph{purrr} {[}Version 0.3.4; Henry and Wickham (2020){]}, \emph{Rcpp} {[}Version 1.0.5; Eddelbuettel and François (2011); Eddelbuettel and Balamuta (2017){]}, \emph{readr} {[}Version 1.4.0; Wickham, Hester, and Francois (2018){]}, \emph{report} {[}Version 0.2.0; Makowski, Dominique, Lüdecke, and Daniel (2019){]}, \emph{see} {[}Version 0.6.0.1; Lüdecke, Ben-Shachar, Waggoner, and Makowski (2020){]}, \emph{sjPlot} {[}Version 2.8.6; Lüdecke (2020){]}, \emph{stringr} {[}Version 1.4.0; Wickham (2019){]}, \emph{styler} {[}Version 1.3.2; Müller and Walthert (2020){]}, \emph{table1} {[}Version 1.2.1; Rich (2020){]}, \emph{texreg} {[}Version 1.37.5; Leifeld (2013){]}, \emph{tibble} {[}Version 3.0.4; Müller and Wickham (2020){]}, \emph{tidyr} {[}Version 1.1.2; Wickham (2020b){]}, and \emph{tidyverse} {[}Version 1.3.0; Wickham et al.
 (2019){]}
@@
 All code for this analysis and a full report of the missing data imputation can be found at \url{https://github.com/go-bayes/nzavs_spiritbeliefWellbeing}.
-An html document decribing the same analysis is located at: \url{https://osf.io/cu2gr/}.
+An html document describing the same analysis is located at: \url{https://osf.io/cu2gr/}.
 
@@
 
-Focusing on life satisfaction, people who believed in Spirit, either with a belief in God (\(\hat{y}\) = 5.18 \textbar{} {[}5.13, 5.24{]}) or without a belief in God (5.18 \textbar{} {[}5.12, 5.24{]}) tended to be lower in life satsifaction than disbelievers in both a God and a Spirit (5.27 \textbar{} {[}5.20, 5.33{]}) and disbelievers in a spirit who also believed in a God (5.26 \textbar{} {[}5.15, 5.36{]}) (see Figure 3)
+Focusing on life satisfaction, people who believed in Spirit, either with a belief in God (\(\hat{y}\) = 5.18 \textbar{} {[}5.13, 5.24{]}) or without a belief in God (5.18 \textbar{} {[}5.12, 5.24{]}) tended to be lower in life satisfaction than disbelievers in both a God and a Spirit (5.27 \textbar{} {[}5.20, 5.33{]}) and disbelievers in a spirit who also believed in a God (5.26 \textbar{} {[}5.15, 5.36{]}) (see Figure 3)
 
@@
 
+\leavevmode\hypertarget{ref-dey2019spirituality}{}%
+Dey, N.
+E.
+Y., Amponsah, B., \& Wiafe-Akenteng, C.
+B.
+(2019).
+Spirituality and subjective well-being of ghanaian parents of children with special needs: The mediating role of resilience.
+\emph{Journal of Health Psychology}, 1359105319873956.
+
 \leavevmode\hypertarget{ref-Diener1985-xy}{}%
@@
 
+\leavevmode\hypertarget{ref-goodman2020mental}{}%
+Goodman, M.
+L., Gitari, S., Keiser, P., Elliott, A., \& Seidel, S.
+(2020).
+Mental health and childhood memories among rural kenyan men: Considering the role of spirituality in life-course pathways.
+\emph{Journal of Health Psychology}, 1359105320944984.
+
 \leavevmode\hypertarget{ref-greaves2015regional}{}%
@@
 
+\leavevmode\hypertarget{ref-holt2003spiritual}{}%
+Holt, C.
+L., Clark, E.
+M., Kreuter, M.
+W., \& Rubio, D.
+M.
+(2003).
+Spiritual health locus of control and breast cancer beliefs among urban african american women.
+\emph{Health Psychology}, \emph{22}(3), 294.
+
 \leavevmode\hypertarget{ref-R-Amelia}{}%
@@
 Honaker, J., King, G., Blackwell, M., \& Others. (2011). Amelia {II}: A program for missing data. \emph{J. Stat. Softw.}, \emph{45}(7), 1--47.
+
+\leavevmode\hypertarget{ref-januu2020associations}{}%
+Jan, A., Malinakova, K., Kosarkova, A., \& Tavel, P. (2020). Associations of childhood trauma experiences with religious and spiritual struggles. \emph{Journal of Health Psychology}, 1359105320950793.
 
@@
 Koenig, H. G., McCullough, Michael E, \& Larson, D. B. (2001). \emph{Handbook of religion and health}. Oxford University Press.
+
+\leavevmode\hypertarget{ref-lawler2009role}{}%
+Lawler-Row, K. A., \& Elliott, J. (2009). The role of religious activity and spirituality in the health and well-being of older adults. \emph{Journal of Health Psychology}, \emph{14}(1), 43--52.
 
@@
 
+\leavevmode\hypertarget{ref-thoresen1999spirituality}{}%
+Thoresen, C. E. (1999). Spirituality and health: Is there a relationship? \emph{Journal of Health Psychology}, \emph{4}(3), 291--300.
+
 \leavevmode\hypertarget{ref-Underwood2002-hg}{}%

### 2021-03-11 Supplement_And_Analysis.Rmd
updated manuscript

@@
 Participants who provided an email address were also emailed and invited to complete an online version if they preferred.
+
 Non-respondents were emailed a follow-up reminder email approximately two months later.
@@
 This sample frame used scaled weighting so that people in increasingly deprived regions were increasingly more likely to be selected (with random sampling of people within regions that had a given level of deprivation).
-The scaling factor was as follows: ni = nbase * weighti, where nbase = 600, and weighti, ranged from 1 to 5 and increased by 1 for each one-unit increased in deprivation score.
+The scaling factor was as follows: n_i = n_base * weigh_i, where n_base = 600, and weight_i, ranged from 1 to 5 and increased by 1 for each one-unit increased in deprivation score.
 Thus, 600 people were randomly selected from regions with a deprivation score of 6, 1,200 people were randomly selected from regions with a deprivation score of 7, and so on.
@@
 A second reminder email was sent approximately four months following the first email attempt.
-We offered a prize draw for participation (five draws each for $1000 grocery vouchers, $5000 total prize pool).
+We offered a prize draw for participation (five draws each for \$1000 grocery vouchers, \$5000 total prize pool).
 Participants were also emailed an eight-page newsletter about the study.
@@
 The Facebook advertisement read as follows: “Participate in the New Zealand Attitudes and Values Study.
-Complete the 2020 Questionnaire online” with the body of text: “If you are part of the NZAVS, but have not heard from us in the last year, then please consider completing the 2020 questionnaire online.
+Complete the 2020 Questionnaire online” with the body of text: "If you are part of the NZAVS, but have not heard from us in the last year, then please consider completing the 2020 questionnaire online.
 The study is more important than ever as we aim to understand the impact of COVID-19 on mental health, wellbeing and resilience in our communities.
-We wish you all the best at this time and hope you keep well and stay safe.” This paid promotion reached 883,969 people, with 37,850 link clicks (i.e., clicking the link for the Qualtrics survey) according to Facebook.
+We wish you all the best at this time and hope you keep well and stay safe." This paid promotion reached 883,969 people, with 37,850 link clicks (i.e., clicking the link for the Qualtrics survey) according to Facebook.
 A total of 6106 people continued complete the questionnaire and provide full contact details, and were thus included in the dataset (4734 were new participants opting in to the study, and 1372 were previously ‘lost’ participants).
@@
 
-This supplement offers a step-by-step analysis for the "Longitudinal
+This supplement offers a step-by-step analysis for the manuscript: "Longitudinal
 Evidence for Growth in Subjective Well-being from Spiritual Beliefs."
-The functions and other materials for this analysis can be found at the
+
+The functions and figures for this analysis can be found at the
 following GitHub link:
-<https://github.com/jacopastorius/nzavs_spiritbeliefwell-being>
+<https://github.com/go-bayes/nzavs_spiritbeliefwell-being>
 
@@
 
-# pa<-d%>%
-#  dplyr::filter(YearMeasured==1)%>%
-#     dplyr::group_by(Id) %>%
-#     dplyr::filter(n() !=0)%>%
-#     dplyr::ungroup(Id)
-# length(unique(pa$Id))
-# table(pa$Wave)
-
-# d2<-data_clean_spirit_wellbeing(pa,0) # d = dataframe, 3 = number of waves
-# table(d2$Wave)
-
 # function to clean data, specify number of waves as 3 
-d_3<-data_clean_spirit_wellbeing(d,9) # d = dataframe, 3 = number of waves
+d_3<-data_clean_spirit_wellbeing(d,9) # d = dataframe, 9 = number of waves
+d_3a<-data_clean_spirit_wellbeing(d,3) # d = dataframe, 3 = number of waves
+
 # function to count uinque ides
@@
 
+## For data with min 9 x waves
 d_3 <- cbind(
@@
 
+# data with 3 x waves.
+d_3a <- cbind(
+  d_3a,
+  demean(d_3a, select = c("Age"), group = "Id"
+))
+
+d_3a<-d_3a%>%
+  dplyr::mutate(Age_betweenC = scale(Age_between, scale = FALSE))
 ```
@@
 
+# Unique participants
+
+```{r}
+length(unique(d_3$Id)) #3257
+```
+
 Sample details are as follows:
@@
 ```
-
-<!-- ```{r  badlatextable} badd tables -->
-
-<!-- library("arsenal") -->
-
-<!-- summary(arsenal::tableby(Wave ~ Age + -->
-
-<!--                    NZdep + -->
-
-<!--                    Education + -->
-
-<!--                    Employed + -->
-
-<!--                    Ethnic_Categories + -->
-
-<!--                    Male  + -->
-
-<!--                    Has_Partner  + -->
-
-<!--                    Political_Conservativism + -->
-
-<!--                    Urban + -->
-
-<!--                    LIFESAT + PWI + includeNA(Beliefs), data = d_3,digits = 2, total=FALSE, format.p = FALSE, digits.p = 0))%>% -->
-<!--     kable( booktabs = T, "latex") -->
-<!-- ``` -->
 
@@
 ```
-
-<!-- ```{r, include=FALSE} -->
-
-<!-- beliefs_table(d_3) -->
-
-<!-- ``` -->
-
-<!-- We wrote the publication in $\LaTeX$, which requires converting the tables to .tex files; this conversion used a custom function we wrote in the `funs.R` file.  -->
-
-<!-- ```{r latex_data_table, include=FALSE} -->
-
-<!-- # you might need to install the dev verion.//remotes::install_github("tysonstanley/furniture") -->
-
-<!-- demographic_table_latex() # note the furniture package conflicts with the table1 package. -->
-
-<!-- d_3%>% -->
-
-<!--   group_by(Wave)%>% -->
-
-<!-- furniture::table1( -->
-
-<!--                   Age,  -->
-
-<!--                   NZdep, -->
-
-<!--                   Education, -->
-
-<!--                   Employed, -->
-
-<!--                   Ethnic_Categories, -->
-
-<!--                   Male,Has_Partner,  -->
-
-<!--                   Political_Conservativism, -->
-
-<!--                   Urban,  -->
-
-<!--                   Beliefs,  -->
-
-<!--                   LIFESAT, -->
-
-<!--                   PWI, -->
-
-<!--                   overall=F) -->
-
-<!-- dat3<-d_3%>%dplyr::select( Age, NZdep, -->
-
-<!--                   Education, -->
-
-<!--                   Employed, -->
-
-<!--                   Ethnic_Categories, -->
-
-<!--                   Male,Has_Partner, -->
-
-<!--                   Political_Conservativism, -->
-
-<!--                   Urban, -->
-
-<!--                   Beliefs, -->
-
-<!--                   LIFESAT, -->
-
-<!--                   PWI,Wave) -->
-
-<!-- stargazer::stargazer(dat3) -->
-
-<!-- library(modelsummary) -->
-
-<!--   modelsummary::datasummary_df(ALL(Age) ~ Mean + SD + Histogram, data=dat3) -->
-
-<!-- ``` -->
-
-<!-- ## How many waves to select? We are using 10 years -->
-
-<!-- We selected a minimum of three waves because we want to recover a -->
-
-<!-- longitudinal sample in which a slope for the effect of the years -->
-
-<!-- $\times$ beliefs effect might be recovered. However, we needed to -->
-
-<!-- consider whether our selection of three waves was sensitive to that -->
-
-<!-- degree of researcher freedom. So we also prepared 4-8 years of data -->
-
-<!-- frames for separate analyses. -->
-
-<!-- ```{r, include = TRUE} -->
-
-<!-- # create datasets to test robustness to years -->
-
-<!-- # d_4<-data_clean_spirit_wellbeing(d,4) # 4 years -->
-
-<!-- # d_5<-data_clean_spirit_wellbeing(d,5) # 5 years -->
-
-<!-- # d_6<-data_clean_spirit_wellbeing(d,6) # 6 years -->
-
-<!-- # d_7<-data_clean_spirit_wellbeing(d,7) # 7 years -->
-
-<!-- # d_8<-data_clean_spirit_wellbeing(d,8) # 7 years -->
-
-<!-- #  -->
-
-<!-- # # function to count uinque number of ID's -->
-
-<!-- # show_unique_id(d_3$Id)  -->
-
-<!-- # show_unique_id(d_4$Id) -->
-
-<!-- # show_unique_id(d_5$Id)  -->
-
-<!-- # show_unique_id(d_6$Id)  -->
-
-<!-- # show_unique_id(d_7$Id)  -->
-
-<!-- # show_unique_id(d_8$Id)  -->
-
-<!-- #  -->
-
-<!-- # # function to count waves for participants -->
-
-<!-- # count_waves_participants(d_3) -->
-
-<!-- # count_waves_participants(d_4) -->
-
-<!-- # count_waves_participants(d_5) -->
-
-<!-- # count_waves_participants(d_6) -->
-
-<!-- # count_waves_participants(d_7) -->
-
-<!-- # count_waves_participants(d_8) -->
-
-<!-- ``` -->
-
-<!-- The analyses for these waves is described below. These analyses reveal -->
-
-<!-- that the inference of the paper is not sensitive to the threshold number -->
-
-<!-- of waves for participant selection. Which is to say, we recover the same -->
-
-<!-- inference with an -wave selection criterion. -->
 
@@
 # theoretically plausible relationships
-tidy_ggdag <- dagify(
-  WB ~ belief + age_within + age_between + partner + nzdep + urban + male + pols + empl, # what might affect wellbeing?
-  WB ~~ partner, # bi-causal path
-  belief ~ age_within  + age_between  + male + ethn, # predictors of beliefs
-  partner ~ nzdep + age_within  + age_between + belief,  # predictors of having a partner
-  nzdep ~ empl,  # predictor of deprivation
-  pols ~ age_within  + age_between  + empl + ethn, # predictor of political orientation
-  empl ~ nzdep + edu + ethn + age_within  + age_between, # predictor of employment
+tidy_ggdag <- ggdag::dagify(
+  WB ~ belief + age_within + age_between + partner + nzdep + urban + male + pols + empl,
+  WB ~~ partner,
+  belief ~ age_within  + age_between  + male + ethn,
+  partner ~ nzdep + age_within  + age_between + belief,
+  nzdep ~ empl,
+  pols ~ age_within  + age_between  + empl + ethn,
+  empl ~ nzdep + edu + ethn + age_within  + age_between,
   exposure =  "belief",
-  outcome =   "WB")%>%
-  tidy_dagitty() + theme_dag_gray()
-
-#graph
+  outcome =   "WB"
+) %>%
+  tidy_dagitty()
+
 ggdag_adjustment_set(tidy_ggdag, node_size = 14) + 
-  theme(legend.position = "bottom") 
+  theme(legend.position = "bottom")  + theme_dag_gray()
 ```
@@
 
+## Models
+
+```{r}
+# REDO
+m1<-brm(forumula_bayesPWI_DAG, data= d_3, file ="mod.1_bDAGR.rds", cores = 8) # note these are very large files
+summary(m1)
+```
+
+```{r}
+conditions <- data.frame(Age_within = c(-5:5))
+p0 <- brms::conditional_effects(
+  m1,
+  c("Age_within:Beliefs" ),
+  spaghetti = TRUE,
+  categorical = FALSE,
+  nsamples = 200,
+  prob = .95,
+  method = 'fitted',
+  conditions = conditions
+)
+
+plot(p0, plot = FALSE)[[1]] +   facet_wrap("Beliefs",nrow = 1) + xlab("Years of Aging (centered)") + ylab ("Personal Wellbeing 0-10")  + theme_blank() + labs(title = "Predicted effects of beliefs states for personal wellbeing")+ theme(legend.position="top") + scale_color_viridis_d(alpha = .2)
+# pwi
+
+```
+
+```{r lsbayesmodelLS_DAG2}
+# LIFE SAT
+m2 <-brm (forumula_bayesLIFESAT_DAG, data= d_3,cores = 8, file ="mod.2_bDAGR.rds")
+summary(m2)
+###
+# note these are very large files
+```
+
 ```{r pwi_bayesmodelPWI_DAG}
@@
 
+```{r}
+
+```
+
 ```{r cache =TRUE}
@@
 
-Next we plot the graphs. First we extract the predictions
-
+Next we plot the graphs. First we extract the predictions:
+
+```{r Bayesian_pwi_predictedSAT_DAG, cache=TRUE}
+get_predictions(m1)
+```
+
+```{r Bayesian_lifesat_predictedSAT_DAG_GET2, cache=TRUE}
+get_predictions(mod.2_bDAGR)
+```
 ```{r Bayesian_pwi_predictedSAT_DAG_GET, cache=TRUE}
@@
 
-<!-- We might want to look at within/between indivual change (i.e.
+<!-- We might want to look at within/between individual change (i.e.
 do people who change in degree of spirit beliefs manifest different well-being.

### 2021-03-11 funs.R

@@ -163,24 +163,30 @@ sample_report <-function(x){
 # }
 
 
-graph_predictions <- function(x,y){
-  out <- ggeffects::ggpredict(model = x, terms = c("Age_within [minmax]","Beliefs"),
-                              ci.lvl = 0.95,
-                              type = "fe",
-                              typical = "mean",
-                              back.transform = TRUE,
-                              ppd = FALSE,
-  plot(out, facets = T) +  gghighlight::gghighlight()  +  theme_blank() + ggtitle(y) # title to be suppled
+
+graph_predictions <- function(x, y) {
+  out <- ggeffects::ggpredict(
+      model = x,
+      terms = c("Age_within [minmax]", "Beliefs"),
+      ci.lvl = 0.95,
+      type = "fe",
+      typical = "mean",
+      back.transform = TRUE,
+      ppd = FALSE,
+      plot(out, facets = T) +  gghighlight::gghighlight()  +  theme_blank() + ggtitle(y))
 }
 
-get_predictions <- function(x){
-  out <- ggeffects::ggpredict(model = x, terms = c("Age_within [minmax]","Beliefs"),
-                              ci.lvl = 0.95,
-                              type = "fe",
-                              typical = "mean",
-                              back.transform = TRUE,
-                              ppd = FALSE,
-                              interval = "confidence")
+get_predictions <- function(x) {
+  out <- ggeffects::ggpredict(
+      model = x,
+      terms = c("Age_within [minmax]", "Beliefs"),
+      ci.lvl = 0.95,
+      type = "fe",
+      typical = "mean",
+      back.transform = TRUE,
+      ppd = FALSE,
+      interval = "confidence"
+    )
   return(out)
 }
 

### 2021-03-11 Manuscript.tex
Updates from Overleaf

@@
 Here, we leverage nine waves of panel data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction (\(N\) = 20979, New Zealand, 2010-2018).
-Contrary to previous research, we find that belief in a spirit or life force predicts lower personal well-being and life-satisfaction.
+Contrary to previous research, we find that belief in a spirit or life force predicts lower personal well-being and life satisfaction.
 However, in support of previous speculation, beliefs in a spirit or life force predict increasing personal well-being and life satisfaction over time relative to disbelief.
-These finding are robust after to known demographic influences, and intriguingly, hold among those who believe in a God but disbelieve in a Spirit or Life Force.
+These finding are robust after for known demographic influences, and intriguingly, hold among those who believe in a God but disbelieve in a Spirit or Life Force.
 The recent growth in spiritual beliefs and decline in traditional religion across many industrial societies motivates further causal investigations of the mechanisms by which spiritual beliefs lead to growth in subjective well-being.
@@
 For example, The \emph{Daily Spiritual Experience Scale}, which is widely used in spiritual mental health research, includes items such as \enquote{I feel thankful for my blessings} and \enquote{I feel deep inner peace or harmony} to measure spirituality (Underwood \& Teresi, 2002).
-As Koenig observes, it is unsurprising that spirituality predict well-being when the same items are used to measure both constructs.
+As Koenig observes, it is unsurprising that spirituality predicts well-being when the same items are used to measure both constructs.
 Koenig aptly describes this approach as \enquote{tautological} (Koenig, 2008).
-Indeed, recent systematic-review of spirituality/mental health research found that nearly 45\% of previous studies employ tautological scales (Garssen et al., 2016).
+Indeed, recent systematic review of spirituality/mental health research found that nearly 45\% of previous studies employ tautological scales (Garssen et al., 2016).
 The first challenge, then, is to define measures of spirituality that are not well-being measures dressed up in other words.
@@
 Although it is commonplace across most areas of psychological science for researchers to explore different operationalisations of focal concepts, the psychology of spirituality has mostly been conducted with cross-sectional North American samples among participants who affiliate with traditional religion (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky, Ratner, \& Chiu, 2005; Smith et al., 2003; Visser, Garssen, \& Vingerhoets, 2010; Yonker, Schnabelrauch, \& Dehaan, 2012).
-As such, the extent to which the psychology of spirituality and health has repeated or extended the psychology of religion and health remains unclear.
-Notably, researchers who have operationalised spirituality constructs as distinct from religion constructs have observed a negative association between spirituality and subjective well-being (King et al., 2013).
+Whether the psychology of spirituality and health has repeated or extended the psychology of religion and health remains unclear.
+Put simply, conceptual overlap between spirituality constructs and religion constructs creates a second line of terminological confusion that interacts with terminological confusion arising from overlap between spirituality constructs and well-being constructs.
+Indeed, researchers who have operationalised spirituality constructs as distinct from religion constructs have observed a negative association between spirituality and subjective well-being (King et al., 2013).
 Thus to infer how spirituality affects subjective well-being, it is important to disentangle the concepts of spirituality from concepts of religion, and to carefully investigate features of constructs in populations where these dimensions overlap, partially overlap, and diverge.
@@
 Previous investigations of spirituality and well-being, including high-quality reviews and meta-analyses, have relied on cross-sectional samples (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky et al., 2005; Smith et al., 2003; Visser et al., 2010; Yonker et al., 2012).
-However, To address the key questions in spirituality psychological health research requires longitudinal investigations.
-Unfortunately the relevant datasets are rare.
-
-Here, we addresses the three challenges of previous in the following ways: First, to avoid a tautological elision of \enquote{spirituality} and \enquote{well-being} we adopt a non-affective measure of spirituality as a \enquote{believe in a Spirit or Life Force.} On the other side, to measure subjective well-being, we use a well-validated scale for personal well-being and a well-validated scale for Life-Satisfaction (see Method).
-Second, to disentangle spiritual beliefs from traditional religious beliefs and skeptical beliefs we adopt a four-level categorical indicator combining the binary indicator of belief in a Spirit or Life Force with the binary indicator \enquote{Do you believe in some form of a spirit or life force?} The levels of this four-level categorical indicator for spiritual, religious, and skeptical beliefs are as follows: (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who nevertheless believe in a Spirit or Life Force; (3) Those who believe in a God but do not believe in a Spirit or Life force; (4) Those who who believe in a both God and a Spirit or Life Force.
+Although researching the key questions in spirituality psychological health research requires national longitudinal investigations, the relevant datasets are unfortunately rare.
+
+Here, we address the three challenges of previous in the following ways: First, to avoid a tautological elision of \enquote{spirituality} and \enquote{well-being} we adopt a non-affective measure of spirituality as a \enquote{believe in a Spirit or Life Force.} On the other side, to measure subjective well-being, we use a well-validated scale for personal well-being and a well-validated scale for Life-Satisfaction (see Method).
+Second, to disentangle spiritual beliefs from traditional religious beliefs and skeptical beliefs we adopt a four-level categorical indicator combining the binary indicator of belief in a Spirit or Life Force with the binary indicator \enquote{Do you believe in a God?} The levels of this four-level categorical indicator for spiritual, religious, and skeptical beliefs are as follows: (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who nevertheless believe in a Spirit or Life Force; (3) Those who believe in a God but do not believe in a Spirit or Life force; (4) Those who who believe in a both God and a Spirit or Life Force.
 Clearly such an approach cannot simultaneously address all interests in previous spirituality/well-being research.

### 2021-03-11 Manuscript_Bayesian.Rmd
Updates from Overleaf

--- 2021-02-12 Manuscript_Bayesian.Rmd
@@
 keywords          : "Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being"
-wordcount         : "5161 words in text body, 1191 words in reference section"
+wordcount         : "5181 words in text body, 1191 words in reference section"
 bibliography      : ["b-references.bib"]
@@
 library("here")
-
 ```
@@
 Here, we addresses these three challenges of previous research in the following ways: First, to avoid a tautological elision of "spirituality" and "well-being" we adopt a non-affective measure of spirituality as a "believe in a Spirit or Life Force." On the other side, to measure subjective well-being, we use a well-validated scale for personal well-being and a well-validated scale for Life-Satisfaction (see Method).
-Second, to disentangle spiritual beliefs from traditional religious beliefs and dis0beliefs we adopt a four-level categorical indicator combining the binary indicator of belief in a Spirit or Life Force with the binary indicator "Do you believe in some form of a spirit or Life Force?" The levels of this four-level categorical indicator for spiritual and religious beliefs and dis-beliefs are as follows: (1) **God and Spirit Disbelievers**: those who neither believe in a God nor believe in a Spirit or Life Force; (2) **Spirit Believers/God Disbelievers**: those who disbelieve in a God and believe in a Spirit or Life Force; (3) **God Believers/Spirit Disbelievers**: those who believe in a God but do not believe in a Spirit or Life Force; (4) **God and Spirit Believers**: Those who believe in a both God and a Spirit or Life Force.
-Clearly such an approach to categorising religion and spirituality cannot simultaneously address many interests in previous spirituality/well-being research.
+Second, to disentangle spiritual beliefs from traditional religious beliefs and skeptical beliefs we adopt a four-level categorical indicator combining the binary indicator of belief in a Spirit or Life Force with the binary indicator "Do you believe in some form of a spirit or Life Force?" The levels of this four-level categorical indicator for spiritual, religious, and skeptical beliefs are as follows: (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life Force; (2) Skeptics about God who nevertheless believe in a Spirit or Life Force; (3) Those who believe in a God but do not believe in a Spirit or Life Force; (4) Those who who believe in a both God and a Spirit or Life Force.
+Clearly such an approach cannot simultaneously address all interests in previous spirituality/well-being research.
 For example we do not combine our cognitive measure of belief with behavioral measures or attitudinal measures.
 However, this cost to scope is paid for with benefits to precision.
-The spiritual indicators measure cognitive states in which distinct varieties of religious and spiritual beliefs and dis-beliefs are straightforwardly disentangled and measured within people, and their longitudinal associations with reliable well-being indicators are systematically quantified Third, to understand how spirit beliefs affect growth, stability, or decline in psychological well-being over time, we leverage longitudinal responses from a nationally diverse probability sample year over nine years (2010 to 2018 New Zealand Attitudes and Values Study, NZAVS).
+The spiritual indicators measure cognitive states in which distinct varieties of skeptical beliefs, traditional religious beliefs, and non-traditional spirit beliefs are straightforwardly disentangled, and measured within people.
+Third, to understand how spirit beliefs affect growth, stability, or decline in psychological well-being over time, we leverage longitudinal responses from a nationally diverse probability sample year over nine years (2010 to 2018 New Zealand Attitudes and Values Study, NZAVS).
 
@@
 Table 1 about here
-\begin{table}[h]
-\begin{center}
-\scalebox{0.45}{
-\begin{tabular}[t]{llllllllllllllllllll}
+
+\begin{table}[ ht ] 
+\centering 
+\caption{}\label{}
+\scalebox{0.6}{
+\begin{tabular}{ l c c c c c c c c c }
 \toprule
-\multicolumn{2}{c}{ } & \multicolumn{2}{c}{2010 (N=3820)} & \multicolumn{2}{c}{2011 (N=6153)} & \multicolumn{2}{c}{2012 (N=10385)} & \multicolumn{2}{c}{2013 (N=15836)} & \multicolumn{2}{c}{2014 (N=15026)} & \multicolumn{2}{c}{2015 (N=13665)} & \multicolumn{2}{c}{2016 (N=18071)} & \multicolumn{2}{c}{2017 (N=15779)} & \multicolumn{2}{c}{2018 (N=16501)} \\
-\cmidrule(l{3pt}r{3pt}){3-4} \cmidrule(l{3pt}r{3pt}){5-6} \cmidrule(l{3pt}r{3pt}){7-8} \cmidrule(l{3pt}r{3pt}){9-10} \cmidrule(l{3pt}r{3pt}){11-12} \cmidrule(l{3pt}r{3pt}){13-14} \cmidrule(l{3pt}r{3pt}){15-16} \cmidrule(l{3pt}r{3pt}){17-18} \cmidrule(l{3pt}r{3pt}){19-20}
-  &    & Mean & Std.
-Dev.
-& Mean  & Std.
-Dev.
-& Mean   & Std.
-Dev.
-& Mean    & Std.
-Dev.
-& Mean     & Std.
-Dev.
-& Mean      & Std.
-Dev.
-& Mean       & Std.
-Dev.
-& Mean        & Std.
-Dev.
-& Mean         & Std.
-Dev.
-\\
+ &   \multicolumn{ 9 }{c}{ Wave }\\ 
+  & 2010 & 2011 & 2012 & 2013 & 2014 & 2015 & 2016 & 2017 & 2018 \\ 
+ & n = \emph{missing} & n = \emph{missing} & n = 8547 & n = 13472 & n = \emph{missing} & n = 11687 & n = 16048 & n = 14174 & n = 14883 \\ 
  \midrule
-Age &  & 51.0 & 14.7 & 50.9 & 15.5 & 50.2 & 14.8 & 48.6 & 14.0 & 49.7 & 14.0 & 51.0 & 13.9 & 50.9 & 13.6 & 51.9 & 13.6 & 52.7 & 13.5\\
-NZdep &  & 4.8 & 2.8 & 4.6 & 2.7 & 4.9 & 2.8 & 4.7 & 2.8 & 4.7 & 2.8 & 4.7 & 2.8 & 4.6 & 2.7 & 4.6 & 2.7 & 4.6 & 2.7\\
-Education &  &  &  &  &  & 5.8 & 2.8 & 6.0 & 2.8 & 6.1 & 2.8 & 6.2 & 2.8 & 6.4 & 2.7 & 6.5 & 2.7 & 6.5 & 2.7\\
-Political\_Conservativism &  & 3.9 & 1.3 & 3.7 & 1.4 & 3.6 & 1.3 & 3.6 & 1.3 & 3.6 & 1.3 & 3.6 & 1.3 & 3.6 & 1.4 & 3.6 & 1.4 & 3.6 & 1.4\\
-LIFESAT &  & 5.2 & 1.2 & 5.1 & 1.2 & 5.1 & 1.2 & 5.2 & 1.2 & 5.2 & 1.2 & 5.2 & 1.2 & 5.2 & 1.2 & 5.3 & 1.2 & 5.3 & 1.2\\
-PWI &  & 6.8 & 1.7 & 6.8 & 1.7 & 6.8 & 1.7 & 7.0 & 1.7 & 7.0 & 1.7 & 7.1 & 1.7 & 7.1 & 1.7 & 7.1 & 1.7 & 7.1 & 1.6\\
-\midrule
- &  & N & \% & N & \% & N & \% & N & \% & N & \% & N & \% & N & \% & N & \% & N & \%\\
-Employed & 0 & 793 & 0.7 & 1276 & 1.1 & 2666 & 2.3 & 3531 & 3.1 & 3262 & 2.8 & 2972 & 2.6 & 4035 & 3.5 & 3655 & 3.2 & 3746 & 3.3\\
- & 1 & 2977 & 2.6 & 4795 & 4.2 & 7566 & 6.6 & 11991 & 10.4 & 11626 & 10.1 & 10378 & 9.0 & 14016 & 12.2 & 12070 & 10.5 & 12250 & 10.6\\
-Ethnic\_Categories & Euro & 2947 & 2.6 & 4375 & 3.8 & 7881 & 6.8 & 12270 & 10.6 & 12051 & 10.5 & 10959 & 9.5 & 14641 & 12.7 & 12906 & 11.2 & 13475 & 11.7\\
- & Maori & 556 & 0.5 & 638 & 0.6 & 1489 & 1.3 & 1897 & 1.6 & 1830 & 1.6 & 1608 & 1.4 & 1964 & 1.7 & 1826 & 1.6 & 1872 & 1.6\\
- & Pacific & 110 & 0.1 & 126 & 0.1 & 343 & 0.3 & 408 & 0.4 & 400 & 0.3 & 336 & 0.3 & 385 & 0.3 & 286 & 0.2 & 296 & 0.3\\
- & Asian & 124 & 0.1 & 190 & 0.2 & 409 & 0.4 & 579 & 0.5 & 572 & 0.5 & 473 & 0.4 & 664 & 0.6 & 574 & 0.5 & 579 & 0.5\\
-Male & 0 & 2390 & 2.1 & 3862 & 3.4 & 6542 & 5.7 & 10033 & 8.7 & 9500 & 8.2 & 8548 & 7.4 & 11377 & 9.9 & 9949 & 8.6 & 10388 & 9.0\\
- & 1 & 1430 & 1.2 & 2291 & 2.0 & 3843 & 3.3 & 5803 & 5.0 & 5478 & 4.8 & 5072 & 4.4 & 6635 & 5.8 & 5776 & 5.0 & 6077 & 5.3\\
-Has\_Partner & 0 & 1054 & 0.9 & 1806 & 1.6 & 2988 & 2.6 & 4237 & 3.7 & 3664 & 3.2 & 3407 & 3.0 & 4274 & 3.7 & 3775 & 3.3 & 3931 & 3.4\\
- & 1 & 2764 & 2.4 & 4325 & 3.8 & 7331 & 6.4 & 11493 & 10.0 & 11130 & 9.7 & 10176 & 8.8 & 13613 & 11.8 & 11964 & 10.4 & 12078 & 10.5\\
-Urban & Not\_Urban & 1531 & 1.3 & 2010 & 1.7 & 3512 & 3.0 & 5116 & 4.4 & 4894 & 4.2 & 4541 & 3.9 & 6223 & 5.4 & 2873 & 2.5 & 2938 & 2.5\\
- & Urban & 2271 & 2.0 & 3695 & 3.2 & 6709 & 5.8 & 10602 & 9.2 & 9937 & 8.6 & 8815 & 7.6 & 11622 & 10.1 & 12687 & 11.0 & 13291 & 11.5\\
-Beliefs & \_Disbeliefs\_ & 625 & 0.5 & 1312 & 1.1 & 2235 & 1.9 & 3639 & 3.2 & 0 & 0.0 & 3278 & 2.8 & 4420 & 3.8 & 4135 & 3.6 & 4514 & 3.9\\
- & \_SpiritExcludesGod\_ & 778 & 0.7 & 1508 & 1.3 & 2535 & 2.2 & 4206 & 3.6 & 0 & 0.0 & 3542 & 3.1 & 4808 & 4.2 & 4148 & 3.6 & 4309 & 3.7\\
- & GodAndSpirit & 1751 & 1.5 & 2560 & 2.2 & 4181 & 3.6 & 6201 & 5.4 & 0 & 0.0 & 5780 & 5.0 & 7002 & 6.1 & 5771 & 5.0 & 6782 & 5.9\\
- & GodExcludesSpirit & 228 & 0.2 & 274 & 0.2 & 593 & 0.5 & 1137 & 1.0 & 0 & 0.0 & 523 & 0.5 & 1094 & 0.9 & 950 & 0.8 & 666 & 0.6\\
+Age &   &   &   &   &   &   &   &   &  \\ 
+NZdep &   &   &   &   &   &   &   &   &  \\ 
+Education &   &   &   &   &   &   &   &   &  \\ 
+Employed &   &   &   &   &   &   &   &   &  \\ 
+\hspace{6pt}    0 &  (\%) &  (\%) & 2094 (24.5\%) & 2955 (21.9\%) &  (\%) & 2525 (21.6\%) & 3486 (21.7\%) & 3223 (22.7\%) & 3404 (22.9\%)\\ 
+\hspace{6pt}    1 &  (\%) &  (\%) & 6453 (75.5\%) & 10517 (78.1\%) &  (\%) & 9162 (78.4\%) & 12562 (78.3\%) & 10951 (77.3\%) & 11479 (77.1\%)\\ 
+Ethnic\_Categories &   &   &   &   &   &   &   &   &  \\ 
+\hspace{6pt}    Euro &  (\%) &  (\%) & 6731 (78.8\%) & 11024 (81.8\%) &  (\%) & 9671 (82.8\%) & 13389 (83.4\%) & 11814 (83.3\%) & 12427 (83.5\%)\\ 
+\hspace{6pt}    Maori &  (\%) &  (\%) & 1188 (13.9\%) & 1608 (11.9\%) &  (\%) & 1341 (11.5\%) & 1749 (10.9\%) & 1620 (11.4\%) & 1677 (11.3\%)\\ 
+\hspace{6pt}    Pacific &  (\%) &  (\%) & 284 (3.3\%) & 339 (2.5\%) &  (\%) & 277 (2.4\%) & 331 (2.1\%) & 243 (1.7\%) & 261 (1.8\%)\\ 
+\hspace{6pt}    Asian &  (\%) &  (\%) & 344 (4\%) & 501 (3.7\%) &  (\%) & 398 (3.4\%) & 579 (3.6\%) & 497 (3.5\%) & 518 (3.5\%)\\ 
+Male &   &   &   &   &   &   &   &   &  \\ 
+\hspace{6pt}    0 &  (\%) &  (\%) & 5302 (62\%) & 8457 (62.8\%) &  (\%) & 7325 (62.7\%) & 10148 (63.2\%) & 9011 (63.6\%) & 9433 (63.4\%)\\ 
+\hspace{6pt}    1 &  (\%) &  (\%) & 3245 (38\%) & 5015 (37.2\%) &  (\%) & 4362 (37.3\%) & 5900 (36.8\%) & 5163 (36.4\%) & 5450 (36.6\%)\\ 
+Has\_Partner &   &   &   &   &   &   &   &   &  \\ 
+\hspace{6pt}    0 &  (\%) &  (\%) & 2408 (28.2\%) & 3546 (26.3\%) &  (\%) & 2815 (24.1\%) & 3773 (23.5\%) & 3330 (23.5\%) & 3556 (23.9\%)\\ 
+\hspace{6pt}    1 &  (\%) &  (\%) & 6139 (71.8\%) & 9926 (73.7\%) &  (\%) & 8872 (75.9\%) & 12275 (76.5\%) & 10844 (76.5\%) & 11327 (76.1\%)\\ 
+Political\_Conservativism &   &   &   &   &   &   &   &   &  \\ 
+Urban &   &   &   &   &   &   &   &   &  \\ 
+\hspace{6pt}    Not\_Urban &  (\%) &  (\%) & 2865 (33.5\%) & 4278 (31.8\%) &  (\%) & 3951 (33.8\%) & 5538 (34.5\%) & 2610 (18.4\%) & 2661 (17.9\%)\\ 
+\hspace{6pt}    Urban &  (\%) &  (\%) & 5682 (66.5\%) & 9194 (68.2\%) &  (\%) & 7736 (66.2\%) & 10510 (65.5\%) & 11564 (81.6\%) & 12222 (82.1\%)\\ 
+Beliefs &   &   &   &   &   &   &   &   &  \\ 
+\hspace{6pt}    \_Skeptic\_ &  (\%) &  (\%) & 2048 (24\%) & 3286 (24.4\%) &  (\%) & 2951 (25.3\%) & 4116 (25.6\%) & 3916 (27.6\%) & 4120 (27.7\%)\\ 
+\hspace{6pt}    \_SpiritExcludesGod\_ &  (\%) &  (\%) & 2295 (26.9\%) & 3735 (27.7\%) &  (\%) & 3169 (27.1\%) & 4501 (28\%) & 3918 (27.6\%) & 3970 (26.7\%)\\ 
+\hspace{6pt}    GodAndSpirit &  (\%) &  (\%) & 3707 (43.4\%) & 5499 (40.8\%) &  (\%) & 5119 (43.8\%) & 6453 (40.2\%) & 5464 (38.5\%) & 6210 (41.7\%)\\ 
+\hspace{6pt}    GodExcludesSpirit &  (\%) &  (\%) & 497 (5.8\%) & 952 (7.1\%) &  (\%) & 448 (3.8\%) & 978 (6.1\%) & 876 (6.2\%) & 583 (3.9\%)\\ 
+LIFESAT &   &   &   &   &   &   &   &   &  \\ 
+PWI &   &   &   &   &   &   &   &   &  \\ 
 \bottomrule
+
 \end{tabular}
 }
-\caption{Sample statistics by year}
-\end{center}
 \end{table}
@@
 ### Belief in a Spirit/Life Force and/or a God  
-We assess belief in a spirit or a God by asking two non-affective questions: "Do you believe in a spirit or Life Force" and "Do you believe in a god?" Responses were coded as 1) **God and Spirit Disbelievers**: those who neither believe in a God nor believe in a Spirit or Life Force; (2) **Spirit Believers/God Disbelievers**: those who disbelieve in a God and believe in a Spirit or Life Force; (3) **God Believers/Spirit Disbelievers**: those who believe in a God but do not believe in a Spirit or Life Force; (4) **God and Spirit Believers**: those who believe in a both God and a Spirit or Life Force.
+We assess belief in a spirit or a God by asking two non-affective questions: "Do you believe in a spirit or Life Force" and "Do you believe in a god?" Responses were coded as (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life Force; (2) Skeptics about God who believe in a Spirit or Life Force; (3) Believers in a God who disbelieve in a Spirit or Life Force; (4) Believers in a God and Spirit or Life Force.
+These believe were developed for the NZAVS from the 2005 Eurobarometer.
 
@@
 In our models, we account for the non-linear effects of time within participants by including the interaction of years (a continuous variable) with Beliefs (a four-level categorical indicator).
-The baseline for this indicator is disbelief in a Spirit/Life Force and disbelief in a God.
+The baseline for this indicator is Skepticism about a Spirit/Life Forces and Skepticism about a God.
 To handle the dependencies introduced from the repeated measures, we include individual ID as an effect modeled as random, and we estimated a random effect for the slope for years.
@@
 Indeed, the inference proved robust to the simple model with only the interaction of years and beliefs, and robust to the fully saturated model in which all covariates were included (see the online supplement for details).
-In the fully saturated model, it appeared that those disbelieving in a Spirit or Life Force, exhibited no growth in Persona Well-Being and Life-Satisifaction over the nine years.
+In the fully saturated model, it appeared that those skeptical of a Spirit or Life Force, exhibited no growth in Persona Well-Being and Life-Satisifaction over the nine years.
 Although this result offers stronger stronger evidence for a Spirit/Life Force well-being mechanism, it is likely the effect of collider confounding and masking.
@@
 We are greatful to the contributors and mantainers of the following packages:
-R [Version 4.0.3; @R-base] and the R-packages *Amelia* [Version 1.7.6; @R-Amelia], *citr* [Version 0.3.2; @R-citr], *dplyr* [Version 1.0.2; @R-dplyr], *equatiomatic* [Version 0.1.0.9000; @R-equatiomatic], *forcats* [Version 0.5.0; @R-forcats], *ggeffects* [Version 1.0.0; @R-ggeffects], *gghighlight* [Version 0.3.0; @R-gghighlight], *ggplot2* [Version 3.3.2; @R-ggplot2], *ggsci* [Version 2.9; @R-ggsci], *lme4* [Version 1.1.25; @R-lme4], *Matrix* [Version 1.2.18; @R-Matrix], *papaja* [Version 0.1.0.9997; @R-papaja], *parameters* [Version 0.9.0.1; @R-parameters], *patchwork* [Version 1.1.0; @R-patchwork], *prettycode* [@R-prettycode], *purrr* [Version 0.3.4; @R-purrr], *Rcpp* [Version 1.0.5; @R-Rcpp_a; @R-Rcpp_b], *readr* [Version 1.4.0; @R-readr], *report* [Version 0.2.0; @R-report], *see* [Version 0.6.0.1; @R-see], *sjPlot* [Version 2.8.6; @R-sjPlot], *stringr* [Version 1.4.0; @R-stringr], *styler* [Version 1.3.2; @R-styler], *table1* [Version 1.2.1; @R-table1], *texreg* [Version 1.37.5; @R-texreg], *tibble* [Version 3.0.4; @R-tibble], *tidyr* [Version 1.1.2; @R-tidyr], and *tidyverse* [Version 1.3.0; @R-tidyverse]
+R [Version 4.0.2; @R-base] and the R-packages *Amelia* [Version 1.7.6; @R-Amelia], *citr* [Version 0.3.2; @R-citr], *dplyr* [Version 1.0.2; @R-dplyr], *equatiomatic* [Version 0.1.0.9000; @R-equatiomatic], *forcats* [Version 0.5.0; @R-forcats], *ggeffects* [Version 0.15.0; @R-ggeffects], *gghighlight* [Version 0.3.0; @R-gghighlight], *ggplot2* [Version 3.3.2; @R-ggplot2], *ggsci* [Version 2.9; @R-ggsci], *lme4* [Version 1.1.23; @R-lme4], *Matrix* [Version 1.2.18; @R-Matrix], *papaja* [Version 0.1.0.9997; @R-papaja], *parameters* [Version 0.8.6.1; @R-parameters], *patchwork* [Version 1.0.1; @R-patchwork], *prettycode* [Version 1.1.0; @R-prettycode], *purrr* [Version 0.3.4; @R-purrr], *Rcpp* [Version 1.0.5.1; @R-Rcpp_a; @R-Rcpp_b], *readr* [Version 1.3.1; @R-readr], *report* [Version 0.1.0; @R-report], *see* [Version 0.5.2.1; @R-see], *sjPlot* [Version 2.8.4; @R-sjPlot], *stringr* [Version 1.4.0; @R-stringr], *styler* [Version 1.3.2; @R-styler], *table1* [Version 1.2; @R-table1], *texreg* [Version 1.37.5; @R-texreg], *tibble* [Version 3.0.4; @R-tibble], *tidyr* [Version 1.1.2; @R-tidyr], and *tidyverse* [Version 1.3.0; @R-tidyverse]"
 
@@
 \begin{center}
-\scalebox{0.6}{
+\scalebox{0.5}{
 \begin{tabular}{l c c}
@@
 ```{r fig.cap = "Coefficient Graphs", echo =FALSE,out.width="\\textwidth"}
-knitr::include_graphics("/Users/jbul716/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/coefBayesDag.png")
+knitr::include_graphics("/Users/jbul176/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/coefBayesDag.png")
 ```
@@
 
-  - The intercept, representing the expected personal well-being among disbelievers in a Spirit/Life Force and God who are Europeans, 50.77 years of age and do not identify as male is reliable (posterior median = 7.138 HDI [ 7.111,  7.167], pd = 100.00%).
+  - The intercept, representing the expected personal well-being among skeptics about a Spirit/Life Force and God who are Europeans, 50.77 years of age and do not identify as male is reliable (posterior median = 7.138 HDI [ 7.111,  7.167], pd = 100.00%).
   - The effect of years is positive is reliable (posterior median =  7.138, HDI [ 7.111,  7.167], pd  100.00%).
-We infer that personal well-being increased in the baseline disbelieving population between 2010 and 2018.
+We infer that personal well-being increased in the baseline skeptic population between 2010 and 2018.
   - The effect of beliefs in *Spirit Excludes God*  is reliably negative (posterior median =  -0.059, HDI [-0.081, -0.036], pd 100.00%).
-We infer that the population that believed in a Spirit but did not believe in a God was lower in personal well-being at the intercept compared to disbelievers.
+We infer that the population that believed in a Spirit but did not believe in a God was lower in personal well-being at the intercept compared to skeptics.
   - The effect of beliefs in *God And Spirit*  is not  negative (posterior median = -0.060, HDI [-0.086, -0.034], pd = 100.00%) .
-We infer that the population that believed in a God and a Spirit or Life Force was lower in personal well-being at the intercept compared to disbelievers.
+We infer that the population that believed in a God and a Spirit or Life Force was lower in personal well-being at the intercept compared to skeptics.
   - The effect of beliefs in *God Excludes Spirit*  is not reliable (posterior median = -0.001 HDI [-0.036,  0.035], pd = 51.10%).
-We infer that the population that believed in a God and doubted a Spirit or Life Force were similar were similar in personal well-being at the intercept to the disbelieving population:
+We infer that the population that believed in a God and doubted a Spirit or Life Force were similar were similar in personal well-being at the intercept to skeptics.
   - The effect of Age in Decades (C) is   0.072 | [ 0.060,  0.083] | 100.00% |
@@
 ```{r fig.cap = "Personal Well-being prediction graph", echo =FALSE, out.width="\\textwidth"}
-knitr::include_graphics("/Users/jbul716/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/pwi_confounder_adjusted.png")
+knitr::include_graphics("/Users/jbul176/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/pwi_confounder_adjusted.png")
 ```
@@
 The effect of years $\times$ belief in a *Spirit Excludes God*  is reliably positive  (posterior median = 0.011, HDI [ 0.003,  0.018], pd = 99.02%).
-We infer that people with beliefs in a spirit or Life Force who do not believe in a God tended to grow in their personal well-being over time at a faster rate than disbelievers grew.
+We infer that people with beliefs in a spirit or Life Force who do not believe in a God tended to grow in their personal well-being over time at a faster rate than skeptics grew.
   - The effect of years $\times$ in *God And Spirit*  is reliably positive  (posterior median = 0.014, HDI [0.007, 0.021], pd = 99.98%).
-We infer that people who believe in a spirit or Life Force who also believe in a God tended to grow in their personal well-being over time at a faster rate than disbelievers grew.
+We infer that people who believe in a spirit or Life Force who also believe in a God tended to grow in their personal well-being over time at a faster rate than skeptics grew.
   - The effect of years $\times$ belief in *God Excludes Spirit*  is not reliable (posterior median = 0.004, HDI [-0.009,  0.015], pd =  68.73%).
-We infer that people who believe in a God but not a spirit or Life Force did not grow at a faster rate in personal well-being than disbelievers grew.
+We infer that people who believe in a God but not a spirit or Life Force did not grow at a faster rate in personal well-being than skeptics grew.
 
@@
 Within this model: 
-  - The intercept, representing the expected life satisfaction among disbelievers in a Spirit/Life Force and God who are Europeans, 50.77 years of age and do not identify as male is reliable (posterior median = 5.287, HDI [ 5.267,  5.306], pd = 100.00%).
+  - The intercept, representing the expected life satisfaction among skeptics about a Spirit/Life Force and God who are Europeans, 50.77 years of age and do not identify as male is reliable (posterior median = 5.287, HDI [ 5.267,  5.306], pd = 100.00%).
   - The effect of years is positive is reliable (posterior median =  0.008, HDI [ 0.003, 0.011], pd = 99.75%).
-We infer that life satisfaction increased in the disbelieving population between 2010 and 2018.
+We infer that life satisfaction increased in the skeptic population between 2010 and 2018.
   -The effect of beliefs in *Spirit Excludes God*  is not reliable (posterior median =   0.007, HDI [-0.010,  0.023], pd = 75.33%).
-We infer that the population that believed in a Spirit but did not believe in a God was higher in life satisfaction at the intercept compared with disbelievers.
+We infer that the population that believed in a Spirit but did not believe in a God was higher in life satisfaction at the intercept compared with skeptics.
   - The effect of beliefs in *God And Spirit*  is reliably positive (posterior median = 0.034, HDI [ 0.015, 0.053], pd = 99.83%).
-We infer that the population that believed in a God and a Spirit or Life Force was higher in life satisfaction at the intercept compared with disbelievers.
+We infer that the population that believed in a God and a Spirit or Life Force was higher in life satisfaction at the intercept compared with skeptics.
   - The effect of beliefs in *God Excludes Spirit*  is reliably positive (posterior median =  0.057, HDI [ 0.030, 0.082], pd = 99.95%).
-We infer that the population that believed in a God and doubted a Spirit or Life Force  was higher in life satisfaction at the intercept compared with disbelievers.
+We infer that the population that believed in a God and doubted a Spirit or Life Force  was higher in life satisfaction at the intercept compared with skeptics.
   - The effect of Age in Decades (C) is reliably positive (posterior median = 0.079, hDI [ 0.071,  0.087] pd = 100.00%).
@@
 ```{r  fig.cap = "Life satisfaction prediction graph", echo = FALSE,out.width="\\textwidth"}
-knitr::include_graphics("/Users/jbul716/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/ls_counfounder_adjusted.png")
+knitr::include_graphics("/Users/jbul176/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/ls_counfounder_adjusted.png")
 ```
   - The interaction of years $\times$ beliefs in *Spirit Excludes God* is reliably positive (posterior median = 0.011, HDI [ 0.006, 0.016], pd = 99.95%).
-We infer that people with beliefs in a Spirit or Life Force who do not believe in a God tended to grow over time in their life satisfaction at a faster rate than disbelievers.
+We infer that people with beliefs in a Spirit or Life Force who do not believe in a God tended to grow over time in their life satisfaction at a faster rate than skeptics.
   - The interaction of years $\times$ *God And Spirit*  is reliably positive (posterior median =  0.009, HDI [ 0.004, 0.014], pd = 99.72%).
-We infer that people who believe in a spirit or Life Force who also believe in a God tended to grow over time in their life satisfaction at a faster rate than disbelievers.
+We infer that people who believe in a spirit or Life Force who also believe in a God tended to grow over time in their life satisfaction at a faster rate than skeptics.
   - The effect of years $\times$ beliefs in *God Excludes Spirit*  is not reliable (posterior median = -0.003, HDI [-0.012, 0.006], pd = 72.28%).
-We infer that people who believe in a God but not a spirit or Life Force tended to grow over time in their life satisfaction at a similar rate to disbelievers.
+We infer that people who believe in a God but not a spirit or Life Force tended to grow over time in their life satisfaction at a similar rate to skeptics.
 
@@
 1.
-At the outset of the study, in 2010, the expected subjective well-being as measured using the personal well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God ($\hat{y} =$  6.93 | [6.88, 6.98]) and lower among those who believed in both a Spirit/Life Force and a God ($\hat{y}$ = 6.91 | [6.87, 6.95]) in comparison both to the population who neither believed in a God nor Spirit/ Life Force ($\hat{y}$ = 7.04 | [6.99, 7.10]) and to the population that believed in a God but was disbelieving in a Spirit or Life Force ($\hat{y}$ 7.03 | [6.95, 7.10]).
-(See Figure 2)  Similarly,the expected subjective well-being as measured using the life satisfaction Inventory was lower in 2010 among those who believed in Spirit or Life Force but not a God ($\hat{y}$ =   5.20 | [5.16, 5.23]).
-Life satisfaction was similar among those who believed in both a Spirit/Life Force and a God  ($\hat{y}$ =   5.23 | [5.21, 5.26]) in comparison both with the disbelieving population who neither believed in a God nor a Life Force ($\hat{y}$ = 5.25 | [5.21, 5.29]).
-Life satisfaction was highest among the population that believed in a God but was disbelieving in a Spirit or Life Force ($\hat{y}$ 5.32 | [5.26, 5.38]) (see Figure 3)
+At the outset of the study, in 2010, the expected subjective well-being as measured using the personal well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God ($\hat{y} =$  6.93 | [6.88, 6.98]) and lower among those who believed in both a Spirit/Life Force and a God ($\hat{y}$ = 6.91 | [6.87, 6.95]) in comparison both to the skeptical population who neither believed in a God nor Spirit/Life Force ($\hat{y}$ = 7.04 | [6.99, 7.10]) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 7.03 | [6.95, 7.10]) (See Figure 2).
+Similarly, the expected subjective well-being as measured using the life satisfaction Inventory was lower in 2010 among those who believed in Spirit or Life Force but not a God ($\hat{y}$ =   5.20 | [5.16, 5.23]).
+Life satisfaction was similar among those who believed in both a Spirit/Life Force and a God ($\hat{y}$ =   5.23 | [5.21, 5.26]) in comparison both with the skeptical population who neither believed in a God nor a Life Force ($\hat{y}$ = 5.25 | [5.21, 5.29]).
+Life satisfaction was highest among the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 5.32 | [5.26, 5.38]) (see Figure 3).
 
@@
 In 2018, personal well-being among who believe in Spirit or Life Force but not a God had grown to ($\hat{y} =   7.19$ [7.15, 7.23], $\Delta \hat{y} = 0.26$).
-Among those who believed in both a Spirit/Life Force and a God, personal well-being had grown to ($\hat{y} = 7.20$ [7.17, 7.24], $\Delta \hat{y} = 0.29$).
-By contrast, personal well-being among the disbelieving population who neither believed in a God nor a Life Force grew less steeply across the eight year interval $\hat{y} =  7.21$ [7.17, 7.25], $\Delta \hat{y} = 0.17$).
-By the same token, personal well-being among those who believe in a God but are disbelieving a Spirit/Life Force grew less steeply $\hat{y} =  7.22$ [7.16, 7.28] $\Delta \hat{y} = 0.19$).
+Among those who believed in both a Spirit/Life Force and a God, personal well-being had grown ($\hat{y} = 7.20$ [7.17, 7.24], $\Delta \hat{y} = 0.29$).
+By contrast, personal well-being among the skeptical population who neither believed in a God nor a Life Force grew less steeply across the eight year interval $\hat{y} =  7.21$ [7.17, 7.25], $\Delta \hat{y} = 0.17$).
+By the same token, personal well-being among those who believe in a God but are skeptical about a Spirit/Life Force grew less steeply $\hat{y} =  7.22$ [7.16, 7.28] $\Delta \hat{y} = 0.19$).
 In practical terms, the rate of increase in personal wellbeing among those with Spirit/Life Force beliefs was about 30% greater than it was among those who lacked such Spirit/Life Force (See Figure 2).
@@
 In 2018, the expected effect of spiritual beliefs on life satisfaction among those who believe in Spirit or Life Force but not a God had grown to ($\hat{y} =  5.37$ [5.34, 5.40], $\Delta \hat{y} = 0.17$).
-Among those who believed in both a Spirit/Life Force and a God, life satisfaction had grown to ($\hat{y} = 5.39$  [5.36, 5.41], $\Delta \hat{y} = 0.19$).By contrast, life satisfaction among the population who neither believed in a God nor a Life Force lagged behind over the eight year interval $\hat{y} =   5.32$ [5.29, 5.35], $\Delta \hat{y} = 0.07$).
-Relatedly, life satisfaction in the population that believe in God but are disbelieving in a Spirit/Life Force grew less steeply $\hat{y} =  5.36$ [5.31, 5.41], $\Delta \hat{y} = 0.04$).
+Among those who believed in both a Spirit/Life Force and a God, life satisfaction had grown to ($\hat{y} = 5.39$  [5.36, 5.41], $\Delta \hat{y} = 0.19$).By contrast, life satisfaction among the skeptical population who neither believed in a God nor a Life Force lagged behind over the eight year interval $\hat{y} =   5.32$ [5.29, 5.35], $\Delta \hat{y} = 0.07$).
+Relatedly, life satisfaction in the population that believe in God but are skeptical about a Spirit/Life Force grew less steeply $\hat{y} =  5.36$ [5.31, 5.41], $\Delta \hat{y} = 0.04$).
 In practical terms, the rate of increase in life satisfaction among those with Spirit/Life Force beliefs over three times greater than among those that lacked such beliefs (See Figure 3).
@@
 Two fundamental question arise from the patterns we observe in this study.
-The first question is why those who believed in Spirit or Life Force  were expected to be initially lower relative the populations that did not believe in a Spirit or Life Force?
+The first question is why those who believed in Spirit or Life Force were initially lower in their personal well-being and life-satisfaction in the initial years of this study in comparison to religious and non-religious skeptics?
 During the early years of New Zealand Attitudes and Values data collection New Zealand was emerging from the twin impacts of a global financial recession and a series of devastating earthquakes that destroyed 30% of Christchurch, New Zealand's second largest city [@greaves2015regional].
@@
 
+# Appendix 
+## Model Equations
+### Priors for the Personal Wellbeing model
+
+Model priors for Personal Wellbeing including group-level ID intercept and beliefs slopes are as follows
+
+$$
+\begin{aligned}
+\operatorname{Personal Well Being}_{ij} \sim \text{Normal}(\mu_{ij},\sigma^2) \nonumber\\
+\mu_{ij}\sim \boldsymbol{\alpha +\beta} \nonumber\\
+\boldsymbol{\alpha} \sim \alpha_{0} +\alpha_j \nonumber\\
+\alpha_{0} \sim \text{StudentT}(3,7.2,2.5)\nonumber\\
+\boldsymbol{\beta}\sim \boldsymbol{b} + \beta_j \nonumber\\
+\boldsymbol{b}\sim\text{Normal}(0,2.5) \nonumber\\
+\begin{bmatrix}
+\alpha_j \\
+\beta_j
+\end{bmatrix}
+\sim 
+\text{MVNormal}
+\begin{pmatrix}
+\begin{bmatrix}
+0\\
+0
+\end{bmatrix}
+,\boldsymbol{S} \nonumber
+\end{pmatrix}\\
+\boldsymbol{S} = 
+\begin{pmatrix}
+0 & \sigma_{\beta_j} \nonumber
+\end{pmatrix} \boldsymbol{R} \begin{pmatrix}
+\sigma_{\alpha_j} & 0 \\
+0 & \sigma_{\beta_j} \nonumber
+\end{pmatrix}\\
+\sigma_{\alpha_j} \sim 
+\text{HalfCauchy}(0,2.5) \nonumber\\
+\sigma_{\beta_j} \sim 
+\text{HalfCauchy}(0,2.5) \nonumber\\
+\boldsymbol{R}\sim \text{LKJcorr}(1) \nonumber\\
+\sigma  \sim  \operatorname{HalfStudent T}^{+}(3, 0, 2.5) 
+\end{aligned}
+$$
+
+\newpage
+
+### Priors for the Life Satisfaction model
+
+Model priors for Life Satisfaction Model with including group-level ID intercept and belief slopes are as follows:
+
+$$
+\begin{aligned}
+\operatorname{Life Satisfaction}_{ij} \sim \text{Normal}(\mu_{ij},\sigma^2) \nonumber\\
+\mu_{ij}\sim \boldsymbol{\alpha +\beta} \nonumber\\
+\boldsymbol{\alpha} \sim \alpha_{0} +\alpha_j \nonumber\\
+\alpha_{0} \sim \text{StudentT}(3,5.5,2.5)\nonumber\\
+\alpha_j \sim \text{StudentT}+(3,0,2.5)\nonumber\\
+\boldsymbol{\beta}\sim \boldsymbol{b} + \beta_j \nonumber\\
+\boldsymbol{b}\sim\text{Normal}(0,2.5) \nonumber\\
+\begin{bmatrix}
+\alpha_j \\
+\beta_j
+\end{bmatrix}
+\sim 
+\text{MVNormal}
+\begin{pmatrix}
+\begin{bmatrix}
+0\\
+0
+\end{bmatrix}
+,\boldsymbol{S} \nonumber
+\end{pmatrix}\\
+\boldsymbol{S} = 
+\begin{pmatrix}
+\sigma_{\alpha_j} & 0 \\
+0 & \sigma_{\beta_j} \nonumber
+\end{pmatrix} \boldsymbol{R} \begin{pmatrix}
+\sigma_{\alpha_j} & 0 \\
+0 & \sigma_{\beta_j} \nonumber
+\end{pmatrix}\\
+\sigma_{\alpha_j} \sim 
+\text{HalfCauchy}(0,1) \nonumber\\
+\sigma_{\beta_j} \sim 
+\text{HalfCauchy}(0,1) \nonumber\\
+\boldsymbol{R}\sim \text{LKJcorr}(1) \nonumber\\
+\sigma  \sim  \operatorname{HalfStudent T}^{+}(3, 0, 2.5) 
+\end{aligned}
+$$
+\newpage
+
+### Deterministic model
+
+The deterministic portion of both the Personal Wellbeing and the Life Satisfaction models is as follows: 
+$$
+\begin{aligned}
+& \quad  \boldsymbol{b} = \\
+& \quad \beta_{1}(\operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\
+&\quad \beta_{2}(\operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\  + \\
+& \quad  \beta_{3}(\operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}})\ +\\
+&\quad \beta_{4}(\operatorname{Age in Decades (C)})\  +\\ 
+& \quad  \beta_{5}(\operatorname{Ethnicity}_{\operatorname{Maori}})\ +\\
+&\quad \beta_{6}(\operatorname{Ethnicity}_{\operatorname{Pacific}})\ +  \\ 
+& \quad  \beta_{7}(\operatorname{Ethnicity}_{\operatorname{Asian}})\ +\\
+& \quad \beta_{8}(\operatorname{Male} )+ \\ 
+& \quad \beta_{9}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\ 
+& \quad \beta_{10}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ + \\
+&\quad \beta_{11}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) \\ 
+\end{aligned}
+$$
+\newpage
+
 # References

### 2021-03-11 Manuscript_Bayesian.tex
Updates from Overleaf

--- 2020-11-30 Manuscript_Bayesian.tex
@@
   pdftitle={National Longitudinal Evidence for Growth in Subjective Well-being from Spiritual Beliefs},
-  pdfauthor={Benjamin Highland1, Everett L. Worthington2, Don E Davis3, Chris G. Sibley4, \& Joseph A. Bulbulia5},
+  pdfauthor={Benjamin Highland1, Everett L. Worthington2, Don E. Davis3, Chris G. Sibley4, \& Joseph A. Bulbulia5},
   pdflang={en-EN},
@@
 \shorttitle{Longitudinal Wellbeing from Spiritual Beliefs}
-\keywords{Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being\newline\indent Word count: 5161 words in text body, 1191 words in reference section}
+\keywords{Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being\newline\indent Word count: 5181 words in text body, 1191 words in reference section}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 
-\begin{table}[h]
-\begin{center}
-\scalebox{0.45}{
-\begin{tabular}[t]{llllllllllllllllllll}
+\begin{table}[ ht ] 
+\centering 
+\caption{}\label{}
+\scalebox{0.6}{
+\begin{tabular}{ l c c c c c c c c c }
 \toprule
-\multicolumn{2}{c}{ } & \multicolumn{2}{c}{2010 (N=3820)} & \multicolumn{2}{c}{2011 (N=6153)} & \multicolumn{2}{c}{2012 (N=10385)} & \multicolumn{2}{c}{2013 (N=15836)} & \multicolumn{2}{c}{2014 (N=15026)} & \multicolumn{2}{c}{2015 (N=13665)} & \multicolumn{2}{c}{2016 (N=18071)} & \multicolumn{2}{c}{2017 (N=15779)} & \multicolumn{2}{c}{2018 (N=16501)} \\
-\cmidrule(l{3pt}r{3pt}){3-4} \cmidrule(l{3pt}r{3pt}){5-6} \cmidrule(l{3pt}r{3pt}){7-8} \cmidrule(l{3pt}r{3pt}){9-10} \cmidrule(l{3pt}r{3pt}){11-12} \cmidrule(l{3pt}r{3pt}){13-14} \cmidrule(l{3pt}r{3pt}){15-16} \cmidrule(l{3pt}r{3pt}){17-18} \cmidrule(l{3pt}r{3pt}){19-20}
-  &    & Mean & Std.
-Dev.
-& Mean  & Std.
-Dev.
-& Mean   & Std.
-Dev.
-& Mean    & Std.
-Dev.
-& Mean     & Std.
-Dev.
-& Mean      & Std.
-Dev.
-& Mean       & Std.
-Dev.
-& Mean        & Std.
-Dev.
-& Mean         & Std.
-Dev.
-\\
+ &   \multicolumn{ 9 }{c}{ Wave }\\ 
+  & 2010 & 2011 & 2012 & 2013 & 2014 & 2015 & 2016 & 2017 & 2018 \\ 
+ & n = \emph{missing} & n = \emph{missing} & n = 8547 & n = 13472 & n = \emph{missing} & n = 11687 & n = 16048 & n = 14174 & n = 14883 \\ 
  \midrule
-Age &  & 51.0 & 14.7 & 50.9 & 15.5 & 50.2 & 14.8 & 48.6 & 14.0 & 49.7 & 14.0 & 51.0 & 13.9 & 50.9 & 13.6 & 51.9 & 13.6 & 52.7 & 13.5\\
-NZdep &  & 4.8 & 2.8 & 4.6 & 2.7 & 4.9 & 2.8 & 4.7 & 2.8 & 4.7 & 2.8 & 4.7 & 2.8 & 4.6 & 2.7 & 4.6 & 2.7 & 4.6 & 2.7\\
-Education &  &  &  &  &  & 5.8 & 2.8 & 6.0 & 2.8 & 6.1 & 2.8 & 6.2 & 2.8 & 6.4 & 2.7 & 6.5 & 2.7 & 6.5 & 2.7\\
-Political\_Conservativism &  & 3.9 & 1.3 & 3.7 & 1.4 & 3.6 & 1.3 & 3.6 & 1.3 & 3.6 & 1.3 & 3.6 & 1.3 & 3.6 & 1.4 & 3.6 & 1.4 & 3.6 & 1.4\\
-LIFESAT &  & 5.2 & 1.2 & 5.1 & 1.2 & 5.1 & 1.2 & 5.2 & 1.2 & 5.2 & 1.2 & 5.2 & 1.2 & 5.2 & 1.2 & 5.3 & 1.2 & 5.3 & 1.2\\
-PWI &  & 6.8 & 1.7 & 6.8 & 1.7 & 6.8 & 1.7 & 7.0 & 1.7 & 7.0 & 1.7 & 7.1 & 1.7 & 7.1 & 1.7 & 7.1 & 1.7 & 7.1 & 1.6\\
-\midrule
- &  & N & \% & N & \% & N & \% & N & \% & N & \% & N & \% & N & \% & N & \% & N & \%\\
-Employed & 0 & 793 & 0.7 & 1276 & 1.1 & 2666 & 2.3 & 3531 & 3.1 & 3262 & 2.8 & 2972 & 2.6 & 4035 & 3.5 & 3655 & 3.2 & 3746 & 3.3\\
- & 1 & 2977 & 2.6 & 4795 & 4.2 & 7566 & 6.6 & 11991 & 10.4 & 11626 & 10.1 & 10378 & 9.0 & 14016 & 12.2 & 12070 & 10.5 & 12250 & 10.6\\
-Ethnic\_Categories & Euro & 2947 & 2.6 & 4375 & 3.8 & 7881 & 6.8 & 12270 & 10.6 & 12051 & 10.5 & 10959 & 9.5 & 14641 & 12.7 & 12906 & 11.2 & 13475 & 11.7\\
- & Maori & 556 & 0.5 & 638 & 0.6 & 1489 & 1.3 & 1897 & 1.6 & 1830 & 1.6 & 1608 & 1.4 & 1964 & 1.7 & 1826 & 1.6 & 1872 & 1.6\\
- & Pacific & 110 & 0.1 & 126 & 0.1 & 343 & 0.3 & 408 & 0.4 & 400 & 0.3 & 336 & 0.3 & 385 & 0.3 & 286 & 0.2 & 296 & 0.3\\
- & Asian & 124 & 0.1 & 190 & 0.2 & 409 & 0.4 & 579 & 0.5 & 572 & 0.5 & 473 & 0.4 & 664 & 0.6 & 574 & 0.5 & 579 & 0.5\\
-Male & 0 & 2390 & 2.1 & 3862 & 3.4 & 6542 & 5.7 & 10033 & 8.7 & 9500 & 8.2 & 8548 & 7.4 & 11377 & 9.9 & 9949 & 8.6 & 10388 & 9.0\\
- & 1 & 1430 & 1.2 & 2291 & 2.0 & 3843 & 3.3 & 5803 & 5.0 & 5478 & 4.8 & 5072 & 4.4 & 6635 & 5.8 & 5776 & 5.0 & 6077 & 5.3\\
-Has\_Partner & 0 & 1054 & 0.9 & 1806 & 1.6 & 2988 & 2.6 & 4237 & 3.7 & 3664 & 3.2 & 3407 & 3.0 & 4274 & 3.7 & 3775 & 3.3 & 3931 & 3.4\\
- & 1 & 2764 & 2.4 & 4325 & 3.8 & 7331 & 6.4 & 11493 & 10.0 & 11130 & 9.7 & 10176 & 8.8 & 13613 & 11.8 & 11964 & 10.4 & 12078 & 10.5\\
-Urban & Not\_Urban & 1531 & 1.3 & 2010 & 1.7 & 3512 & 3.0 & 5116 & 4.4 & 4894 & 4.2 & 4541 & 3.9 & 6223 & 5.4 & 2873 & 2.5 & 2938 & 2.5\\
- & Urban & 2271 & 2.0 & 3695 & 3.2 & 6709 & 5.8 & 10602 & 9.2 & 9937 & 8.6 & 8815 & 7.6 & 11622 & 10.1 & 12687 & 11.0 & 13291 & 11.5\\
-Beliefs & \_Skeptic\_ & 625 & 0.5 & 1312 & 1.1 & 2235 & 1.9 & 3639 & 3.2 & 0 & 0.0 & 3278 & 2.8 & 4420 & 3.8 & 4135 & 3.6 & 4514 & 3.9\\
- & \_SpiritExcludesGod\_ & 778 & 0.7 & 1508 & 1.3 & 2535 & 2.2 & 4206 & 3.6 & 0 & 0.0 & 3542 & 3.1 & 4808 & 4.2 & 4148 & 3.6 & 4309 & 3.7\\
- & GodAndSpirit & 1751 & 1.5 & 2560 & 2.2 & 4181 & 3.6 & 6201 & 5.4 & 0 & 0.0 & 5780 & 5.0 & 7002 & 6.1 & 5771 & 5.0 & 6782 & 5.9\\
- & GodExcludesSpirit & 228 & 0.2 & 274 & 0.2 & 593 & 0.5 & 1137 & 1.0 & 0 & 0.0 & 523 & 0.5 & 1094 & 0.9 & 950 & 0.8 & 666 & 0.6\\
+Age &   &   &   &   &   &   &   &   &  \\ 
+NZdep &   &   &   &   &   &   &   &   &  \\ 
+Education &   &   &   &   &   &   &   &   &  \\ 
+Employed &   &   &   &   &   &   &   &   &  \\ 
+\hspace{6pt}    0 &  (\%) &  (\%) & 2094 (24.5\%) & 2955 (21.9\%) &  (\%) & 2525 (21.6\%) & 3486 (21.7\%) & 3223 (22.7\%) & 3404 (22.9\%)\\ 
+\hspace{6pt}    1 &  (\%) &  (\%) & 6453 (75.5\%) & 10517 (78.1\%) &  (\%) & 9162 (78.4\%) & 12562 (78.3\%) & 10951 (77.3\%) & 11479 (77.1\%)\\ 
+Ethnic\_Categories &   &   &   &   &   &   &   &   &  \\ 
+\hspace{6pt}    Euro &  (\%) &  (\%) & 6731 (78.8\%) & 11024 (81.8\%) &  (\%) & 9671 (82.8\%) & 13389 (83.4\%) & 11814 (83.3\%) & 12427 (83.5\%)\\ 
+\hspace{6pt}    Maori &  (\%) &  (\%) & 1188 (13.9\%) & 1608 (11.9\%) &  (\%) & 1341 (11.5\%) & 1749 (10.9\%) & 1620 (11.4\%) & 1677 (11.3\%)\\ 
+\hspace{6pt}    Pacific &  (\%) &  (\%) & 284 (3.3\%) & 339 (2.5\%) &  (\%) & 277 (2.4\%) & 331 (2.1\%) & 243 (1.7\%) & 261 (1.8\%)\\ 
+\hspace{6pt}    Asian &  (\%) &  (\%) & 344 (4\%) & 501 (3.7\%) &  (\%) & 398 (3.4\%) & 579 (3.6\%) & 497 (3.5\%) & 518 (3.5\%)\\ 
+Male &   &   &   &   &   &   &   &   &  \\ 
+\hspace{6pt}    0 &  (\%) &  (\%) & 5302 (62\%) & 8457 (62.8\%) &  (\%) & 7325 (62.7\%) & 10148 (63.2\%) & 9011 (63.6\%) & 9433 (63.4\%)\\ 
+\hspace{6pt}    1 &  (\%) &  (\%) & 3245 (38\%) & 5015 (37.2\%) &  (\%) & 4362 (37.3\%) & 5900 (36.8\%) & 5163 (36.4\%) & 5450 (36.6\%)\\ 
+Has\_Partner &   &   &   &   &   &   &   &   &  \\ 
+\hspace{6pt}    0 &  (\%) &  (\%) & 2408 (28.2\%) & 3546 (26.3\%) &  (\%) & 2815 (24.1\%) & 3773 (23.5\%) & 3330 (23.5\%) & 3556 (23.9\%)\\ 
+\hspace{6pt}    1 &  (\%) &  (\%) & 6139 (71.8\%) & 9926 (73.7\%) &  (\%) & 8872 (75.9\%) & 12275 (76.5\%) & 10844 (76.5\%) & 11327 (76.1\%)\\ 
+Political\_Conservativism &   &   &   &   &   &   &   &   &  \\ 
+Urban &   &   &   &   &   &   &   &   &  \\ 
+\hspace{6pt}    Not\_Urban &  (\%) &  (\%) & 2865 (33.5\%) & 4278 (31.8\%) &  (\%) & 3951 (33.8\%) & 5538 (34.5\%) & 2610 (18.4\%) & 2661 (17.9\%)\\ 
+\hspace{6pt}    Urban &  (\%) &  (\%) & 5682 (66.5\%) & 9194 (68.2\%) &  (\%) & 7736 (66.2\%) & 10510 (65.5\%) & 11564 (81.6\%) & 12222 (82.1\%)\\ 
+Beliefs &   &   &   &   &   &   &   &   &  \\ 
+\hspace{6pt}    \_Skeptic\_ &  (\%) &  (\%) & 2048 (24\%) & 3286 (24.4\%) &  (\%) & 2951 (25.3\%) & 4116 (25.6\%) & 3916 (27.6\%) & 4120 (27.7\%)\\ 
+\hspace{6pt}    \_SpiritExcludesGod\_ &  (\%) &  (\%) & 2295 (26.9\%) & 3735 (27.7\%) &  (\%) & 3169 (27.1\%) & 4501 (28\%) & 3918 (27.6\%) & 3970 (26.7\%)\\ 
+\hspace{6pt}    GodAndSpirit &  (\%) &  (\%) & 3707 (43.4\%) & 5499 (40.8\%) &  (\%) & 5119 (43.8\%) & 6453 (40.2\%) & 5464 (38.5\%) & 6210 (41.7\%)\\ 
+\hspace{6pt}    GodExcludesSpirit &  (\%) &  (\%) & 497 (5.8\%) & 952 (7.1\%) &  (\%) & 448 (3.8\%) & 978 (6.1\%) & 876 (6.2\%) & 583 (3.9\%)\\ 
+LIFESAT &   &   &   &   &   &   &   &   &  \\ 
+PWI &   &   &   &   &   &   &   &   &  \\ 
 \bottomrule
+
 \end{tabular}
 }
-\caption{Sample statistics by year}
-\end{center}
 \end{table}
@@
 We are greatful to the contributors and mantainers of the following packages:
-R (Version 4.0.3; R Core Team, 2020) and the R-packages \emph{Amelia} (Version 1.7.6; Honaker et al., 2011a), \emph{citr} (Version 0.3.2; Aust, 2019), \emph{dplyr} (Version 1.0.2; Wickham et al., 2020), \emph{equatiomatic} (Version 0.1.0.9000; Anderson \& Heiss, 2020), \emph{forcats} (Version 0.5.0; Wickham, 2020a), \emph{ggeffects} (Version 1.0.0; Lüdecke, 2018), \emph{gghighlight} (Version 0.3.0; Yutani, 2020), \emph{ggplot2} (Version 3.3.2; Wickham, 2016), \emph{ggsci} (Version 2.9; Xiao, 2018), \emph{lme4} (Version 1.1.25; Bates et al., 2015), \emph{Matrix} (Version 1.2.18; Bates \& Maechler, 2019), \emph{papaja} (Version 0.1.0.9997; Aust \& Barth, 2020), \emph{parameters} (Version 0.9.0.1; D.
-Lüdecke et al., 2020a), \emph{patchwork} (Version 1.1.0; Pedersen, 2020), \emph{prettycode} (Csárdi, 2019), \emph{purrr} (Version 0.3.4; Henry \& Wickham, 2020), \emph{Rcpp} (Version 1.0.5; Eddelbuettel \& François, 2011; Eddelbuettel \& Balamuta, 2017), \emph{readr} (Version 1.4.0; Wickham, Hester, \& Francois, 2018), \emph{report} (Version 0.2.0; Makowski, Dominique, Lüdecke, \& Daniel, 2019), \emph{see} (Version 0.6.0.1; D.
-Lüdecke et al., 2020b), \emph{sjPlot} (Version 2.8.6; D.
-Lüdecke, 2020), \emph{stringr} (Version 1.4.0; Wickham, 2019), \emph{styler} (Version 1.3.2; Müller \& Walthert, 2020), \emph{table1} (Version 1.2.1; Rich, 2020), \emph{texreg} (Version 1.37.5; Leifeld, 2013), \emph{tibble} (Version 3.0.4; Müller \& Wickham, 2020), \emph{tidyr} (Version 1.1.2; Wickham, 2020b), and \emph{tidyverse} (Version 1.3.0; Wickham, Averick, et al., 2019)
+R (Version 4.0.2; R Core Team, 2020) and the R-packages \emph{Amelia} (Version 1.7.6; Honaker et al., 2011a), \emph{citr} (Version 0.3.2; Aust, 2019), \emph{dplyr} (Version 1.0.2; Wickham et al., 2020), \emph{equatiomatic} (Version 0.1.0.9000; Anderson \& Heiss, 2020), \emph{forcats} (Version 0.5.0; Wickham, 2020a), \emph{ggeffects} (Version 0.15.0; Lüdecke, 2018), \emph{gghighlight} (Version 0.3.0; Yutani, 2020), \emph{ggplot2} (Version 3.3.2; Wickham, 2016), \emph{ggsci} (Version 2.9; Xiao, 2018), \emph{lme4} (Version 1.1.23; Bates et al., 2015), \emph{Matrix} (Version 1.2.18; Bates \& Maechler, 2019), \emph{papaja} (Version 0.1.0.9997; Aust \& Barth, 2020), \emph{parameters} (Version 0.8.6.1; D.
+Lüdecke et al., 2020a), \emph{patchwork} (Version 1.0.1; Pedersen, 2020), \emph{prettycode} (Version 1.1.0; Csárdi, 2019), \emph{purrr} (Version 0.3.4; Henry \& Wickham, 2020), \emph{Rcpp} (Version 1.0.5.1; Eddelbuettel \& François, 2011; Eddelbuettel \& Balamuta, 2017), \emph{readr} (Version 1.3.1; Wickham, Hester, \& Francois, 2018), \emph{report} (Version 0.1.0; Makowski, Dominique, Lüdecke, \& Daniel, 2019), \emph{see} (Version 0.5.2.1; D.
+Lüdecke et al., 2020b), \emph{sjPlot} (Version 2.8.4; D.
+Lüdecke, 2020), \emph{stringr} (Version 1.4.0; Wickham, 2019), \emph{styler} (Version 1.3.2; Müller \& Walthert, 2020), \emph{table1} (Version 1.2; Rich, 2020), \emph{texreg} (Version 1.37.5; Leifeld, 2013), \emph{tibble} (Version 3.0.4; Müller \& Wickham, 2020), \emph{tidyr} (Version 1.1.2; Wickham, 2020b), and \emph{tidyverse} (Version 1.3.0; Wickham, Averick, et al., 2019)"
 
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{/Users/jbul716/The Virtues Project Dropbox/Joseph Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/coefBayesDag} \caption{Coefficient Graphs}\label{fig:unnamed-chunk-1}
+\includegraphics[width=\textwidth]{/Users/jbul176/The Virtues Project Dropbox/Joseph Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/coefBayesDag} \caption{Coefficient Graphs}\label{fig:unnamed-chunk-1}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{/Users/jbul716/The Virtues Project Dropbox/Joseph Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/pwi_confounder_adjusted} \caption{Personal Well-being prediction graph}\label{fig:unnamed-chunk-2}
+\includegraphics[width=\textwidth]{/Users/jbul176/The Virtues Project Dropbox/Joseph Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/pwi_confounder_adjusted} \caption{Personal Well-being prediction graph}\label{fig:unnamed-chunk-2}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{/Users/jbul716/The Virtues Project Dropbox/Joseph Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/ls_counfounder_adjusted} \caption{Life satisfaction prediction graph}\label{fig:unnamed-chunk-3}
+\includegraphics[width=\textwidth]{/Users/jbul176/The Virtues Project Dropbox/Joseph Bulbulia/000GIT_STUFF/nzavs_spiritbeliefWellbeing/ls_counfounder_adjusted} \caption{Life satisfaction prediction graph}\label{fig:unnamed-chunk-3}
 \end{figure}

### 2021-03-12 Manuscript_Bayesian.Rmd
update

--- 2021-03-11 Manuscript_Bayesian.Rmd
@@
 output            : papaja::apa6_pdf
+keep_tex: true
 extra_dependencies: ["rotating", "booktabs", "graphicx","longtable","amsmath","amsfonts","amssymb"]

### 2021-04-05 Supplement_And_Analysis.Rmd
updated supplement

@@
 # custom function for my machine to read data. you'll need to read data to your machine. tweak function in the `funs.R` script for your machine
-d<-data_read_mymachine()
+d <- data_read_mymachine()
 
@@
 
-# Unique participants
+Unique participants
 
@@
 modelsummary::datasummary_balance( ~ Wave, data=ts, dinm=FALSE, output = 'table.tex')
+modelsummary::datasummary_balance( ~ Wave, data=ts, dinm=FALSE)
 # https://github.com/IndrajeetPatil/ggstatsplot
@@
 ```{r demographic table, cache=TRUE}
-# must remove Furniture package before running table 1
+# must remove this table is is for the manuscript
 demographic_table(d_3)
